### 2008-04-20 papers/OTSPDE/optimal_time_steps_for_pdes.tex
OTSPDE: - checked in incomplete draft of first paper on OTSPDE

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First draft: 2008/02/??, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[oneeqnum,onefignum,onetabnum,onethmnum]{siamltex}
\usepackage{graphics}
\usepackage{multirow}

\include{macros}

\title{Using Optimal Time Step Selection to Boost the Accuracy of
       Finite Difference Methods for a Class of 
       Time Dependent Partial Differential Equations}

\author{
Kevin T. Chu\footnotemark[2]
}

\begin{document}
\bibliographystyle{unsrt}
\maketitle

\renewcommand{\thefootnote}{\fnsymbol{footnote}}
\footnotetext[2]{Vitamin D, Inc., Menlo Park, CA 94025}

\renewcommand{\thefootnote}{\arabic{footnote}}

\begin{abstract}
In this article, we discuss a novel technique based on optimal time step 
selection for boosting low-order finite difference schemes
into higher-order numerical methods for time dependent partial differential
equations (PDEs).  We demonstrate the utility of this technique on several 
classical hyperbolic and parabolic problems in one and more space dimensions 
and explain the observed orders of convergence through straightforward numerical
analysis arguments.
** HIGH-ORDER PDES AND IRREGULAR DOMAINS ***
\end{abstract}

\begin{keywords}
optimal time step, finite difference methods, high-order accuracy, 
time dependent PDEs, initial-value problems
\end{keywords}

\begin{AMS}
65-02, 65M06, 65M12, 65M20
\end{AMS}

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{KEVIN T. CHU}
         {HIGH-ORDER ACCURATE FD SCHEMES VIA OPTIMAL TIME STEP SELECTION}

\section*{Introduction}
High-order numerical methods for partial differential equations (PDEs) will 
always be valuable for increasing the computational efficiency of numerical 
simulations.  Thus, it is not at all surprising that a great deal of effort in 
numerical PDEs continues to be focused on the development of high-order 
numerical schemes~\cite{gibour_2005,??}.  
Typically, high-order accuracy is achieved by constructing
schemes that have high formal orders of accuracy.  However, high-order 
accurate numerical solutions can also be obtained by using formally low-order 
schemes in clever ways.  When possible, the latter approach can be a powerful 
way to boost the accuracy of a numerical method \emph{without} introducing too 
much additional algorithmic (and programming) complexity.

High-order accurate methods that are simple combinations or modifications of 
low-order schemes abound.  A few well-known examples include Richardson 
extrapolation~\cite{richardson_1927,heath_book}, the modified nine-point 
formula for the Poisson equation~\cite{iserles_book}, and the Lax-Wendroff 
method for the advection equation~\cite{leveque_book_1992, leveque_book_2002,
gko_book}.  Cancellation of higher-order errors (whether fortuitous or 
intentional) is a key to principle underlying the superior accuracy of these 
each of methods.

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This standard approach for developing numerical methods for PDEs 
has its roots in the classical numerical analysis of PDEs, which tends to 
individually analyze the discretization of the terms in the PDE.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for \emph{all} 
independent variables.  Unfortunately, this common approach for developing 
and analyzing numerical methods is often suboptimal because it neglects the 
potential for achieving high order of accuracy using only simple finite 
difference schemes based on compact, low-order stencils.

Perhaps the greatest weakness of the standard method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides all 
of the details about higher-order errors.  The information lost in the ``big-O" 
notation eliminates the possibility of reducing numerical error by taking 
advantage of relationships between the terms in the error.  The modified 
nine-point formula for the Poisson equation (also known under the name 
Mehrstellen)~\cite{iserles_book} and the Lax-Wendroff scheme for the 
advection equation~\cite{leveque_book_1992, leveque_book_2002, gko_book} 
demonstrate the power of explicitly retaining higher-order terms in the 
discretization error when analyzing numerical schemes.  For both of these 
methods, examination of higher-order terms allows low-order methods to be 
transformed into high-order methods with only minor modifications to the 
numerical scheme.

The usual approach for developing numerical methods has another significant
shortcoming -- it places no value on optimization of numerical parameters 
(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
somewhat to blame for this situation because it gives the false impression 
that the error depends on each numerical parameter independently.  Aside
from constraints needed for numerical stability, ``big-O" notation makes 
it appear as though any combination of numerical parameters is equally good 
as long as they are all ``small enough.''  For some problems, however, 
optimizing the numerical parameters can significantly increase the accuracy
of the numerical solution.  For example, in~\cite{zhao_2006} Zhao showed 
that appropriately choosing the value of $\theta$ as a function of the time 
step, $\dt$, in the theta-method\footnote{In~\cite{zhao_2006} refers 
to $\theta$ as the weight factor and the theta-method as the single-step 
trapezoidal method.}~\cite{iserles_book, gko_book} can decrease the error 
of finite element solutions to the diffusion equation.

In this paper, we explore a novel technique based on optimal time step (OTS)
selection that can be used to construct high-order finite difference schemes 
for time dependent PDEs using only formally low-order stencils.  The essential 
idea is that a carefully chosen time step (and possibly the addition of
a few correction terms) can eliminate low-order terms in the discretization 
error, allowing a formally low-order method to deliver high-order accuracy.  
The optimal time step is calculated through a straightforward analysis of the 
discretization error that uses information from the PDE to relate the 
leading-order terms.  Optimal time step selection combines the well-known 
approach of using information provided by the PDE to reduce numerical error 
with the less-common notion that optimal choice of numerical parameters can 
boost the effectiveness of algorithmically simple and computationally 
low-cost methods.

Using the PDE being solved to reduce discretization error has a long 
tradition in numerical analysis.  Indeed, this idea underlies both the 
modified nine-point formula for the Poisson equation~\cite{iserles_book} and 
the Lax-Wendroff method for the linear advection 
equation~\cite{leveque_book_1992, leveque_book_2002, gko_book}.  
In the case of the modified nine-point formula for the Poisson equation, a 
simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
shows that it is only second-order, which might lead to the conclusion that 
the nine-point stencil has no advantage over the standard five-point stencil.  
However, a more careful analysis of the discretization error in light of the 
PDE reveals that the leading-order error, which is proportional to the 
bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
related to the Laplacian of the source term in the Poisson equation.  Thus, 
the numerical scheme is easily boosted to fourth-order by simply adding a 
correction term that cancels out the leading-order error.   
Similarly, use of the PDE to eliminate higher-order discretization errors also 
plays a critical role in the Lax-Wendroff method because it allows the second 
derivative of the solution in time to be written in terms of the second 
derivative in space.  This observation transforms an unstable, first-order 
scheme based on a central difference approximation of 
$\partial/\partial x$ into a stable, second-order 
scheme~\cite{leveque_book_1992, leveque_book_2002}.

Optimal time step selection as a technique for improving the overall accuracy 
of numerical schemes for time dependent PDEs appears to be a novel idea.
Typically, the primary concern when choosing a time step is numerical 
stability with accuracy being a secondary consideration.  The fact that a 
clever choice of time step can be used to dramatically reduce the error is 
rarely considered.  In general, little research seems to have been done on 
optimizing the parameters of numerical methods (\eg time step and grid 
spacing) to improve accuracy.  To the best of the authors knowledge, Zhao's 
method for choosing an optimal value of $\theta$ in the theta-method is the 
only work that attempts to improve accuracy by modifying numerical parameters.
While he focuses solely on the optimal choice of $\theta$ for a fixed time 
step, his approach can be viewed as a form of optimal time step selection by 
reversing the roles of $\theta$ and the time step.

Before moving on, it is worth spending a few moments to draw the distinction
between optimal time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book,??} that are used in the context of the method 
of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
selection may appear to be just a crude version of adaptive time step control.
However, the two methods are actually very different.  While both methods seek 
to reduce numerical error by controlling the time step, the numerical errors 
they control are fundamentally different.  Adaptive time stepping can only 
be used to reduce temporal discretization errors because it controls the 
errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Thus, optimal time 
step selection has the potential to produce superior results compared with 
traditional adaptive time stepping approaches for integrating the 
semi-discrete equations associated with time dependent PDEs.  Another 
important distinction between the two techniques is that optimal time step 
selection uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

As a technique for constructing numerical methods, optimal time step selection 
has several desireable features.  Like any technique that leads to high-order 
methods, optimal time step selection produces schemes that greatly reduce the 
computational cost required to obtain a numerical solution of a desired 
accuracy.  However, its real power comes from the fact that it achieves
high-order accuracy while being based solely on very simple finite difference 
schemes.  For instance, we will see how optimal time step selection makes 
it possible to achieve 4th-order accuracy for the diffusion equation using only 
a second-order stencil for the Laplacian and forward Euler time integration.  
Moreover, high-order accuracy is not restricted to problems on simple, 
rectangular domains; irregular domains are straightforward to handle by
appropriately setting the values in ghost cells~\cite{fedkiw_1999, 
osher_fedkiw_book, gibou_2005}.  Optimal time step selection is even 
beneficial when using finite difference schemes to solve some nonlinear PDEs.  
While there are certainly limitations to optimal time step selection, we will 
see that there are a many problems where it is useful.

The remainder of this paper is organized as follows.  
In Section~\ref{sec:optimal_time_step_selection}, we begin by 
discussing the technique of optimal time step selection in detail and 
presenting the analysis that shows how it boosts the order of accuracy 
for formally low-order numerical schemes.  In 
Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
optimal time selection to problems in one space dimension by applying it to 
finite difference schemes for several classical linear and nonlinear PDEs. 
In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
to improve the accuracy of simple finite difference schemes for the linear 
advection equation and diffusion equation in multiple space dimensions.  
To illustrate the ease with which high-order accuracy can be achieved for
problems on arbitrary domains, the 2D diffusion equation is solved on both 
the regular and irregular domains.  
For each PDE in Sections~\ref{sec:applications_1d} and 
\ref{sec:applications_multidim}, we provide the numerical analysis required 
to calculate the optimal time step size and order of accuracy.  
Finally, in Section~\ref{sec:summary}, we provide some concluding remarks on 
the applicability and limitations of optimal time step selection for the 
numerical solution of general time dependent PDEs.

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section we present and analyze the 
technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
$\frac{\partial u}{\partial t} = \ldots$, where the right hand side only 
involves functions of the independent variable and its spatial derivatives.  
Because this class of time dependent PDEs is so prevalent and important in 
many problems of applied mathematics, restricting our attention to problems 
of this form is not a serious limitation.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation to a 
time dependent PDE of the form
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $u$ is the solution to the PDE, $n$ is the spatial order of the 
PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the FD 
approximation by optimizing the time step, there is no reason to explicitly
construct the scheme to be high-order.  In fact, using a high-order FD scheme 
only complicates the analysis required to calculate the optimal time 
step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the FD approximation, explicitly keeping track 
of the leading-order terms.  In contrast to standard error analysis, we do 
not sweep all of the errors under ``big-O'' notation.  Rather, we take 
advantage of the fact that derivation of the leading-order terms in the 
error is straightforward for FD schemes through the use of Taylor series 
expansions.  Because each term in the error is proportional to a partial 
derivative of the solution, we can attempt to use the PDE to relate the 
leading-order terms to each other.  If we are fortunate (which in many cases 
we are), this careful analysis yields a direct relationship between the 
leading-order spatial and temporal errors that allows them to be combined as 
a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $P$ is a polynomial in the numerical parameters, $L$ is a differential 
operator, $u$ is the solution to the PDE, and we have assumed that the 
grid spacing is equal in all spatial directions\footnote{In general, the grid 
spacing need not be equal in all spatial directions.  For some problems, it is 
only possible to calculate an optimal time step if the grid spacing is
unequal between the different spatial directions.}.  By selecting the 
numerical parameters so that $P = 0$, we can completely eliminate the 
leading-order term in the discretization error.  Because time dependent PDEs 
are often solved by stepping in time, it is natural to choose the time step 
to be a function of the grid spacing.  The time step selected in this way 
is the \emph{optimal time step} for the finite difference scheme.

An important class of FD schemes for time dependent PDEs are those based
on the method of lines.   When the method of lines approach is used for 
(\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
errors, and $\alpha$ and $\beta$ are positive constants that depend on the 
details of the PDE and FD scheme.  Setting $P = 0$, we find that the optimal 
time step is given by\footnote{The $\dt = 0$ root of $P$ is discarded 
because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Note that the optimal time step 
has the same functional dependence on the grid spacing as a time step 
that is satisfies a stability constraint.

The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the FD approximation of the spatial derivative 
operator acting on ${\mathbf u}$.  The second term in the local error arises 
directly from temporal discretization errors associated with the numerical 
scheme used to integrate the system of ODEs in time.  The first term arises 
because $F$ is only an approximation of the spatial derivative operators in 
the PDE.  As a result, the system of ODEs~(\ref{eq:method_of_lines}) has an 
error due to the spatial discretization at \emph{every} point in time.  For 
a spatial discretization error of order $r$, the error in 
(\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the error 
accumulated during each time step due to spatial discretization error is 
$O(\dt \dx^r)$ regardless of how the system of ODEs is solved 
numerically.  Because the method of lines is so frequently used to construct FD 
approximations for time dependent PDEs, we will focus solely on FD schemes 
of this type for the remainder of the paper in order to keep the discussion 
simple.  It is straightforward to extend the results we obtain to more general 
FD schemes.

So far in our discussion, we have completely ignored the question of whether 
the FD scheme is stable when the time step is set to $\dto$.  When
the optimal time step is greater than the smallest stable time step, the 
method breaks down.  Fortunately, when stability issues arise for explicit 
time integration schemes, we can usually switch to an implicit (or 
semi-implicit) time integration scheme with the same order of accuracy for 
which the optimal time step is stable.  Note that the depending on how
the original FD scheme is modified, a new optimal time step may need to be 
calculated.

\subsubsection*{PDEs in Higher Spatial Dimensions 
                \label{sec:ots_higher_spatial_dims}}
It is important to emphasize that optimal time selection is applicable to 
PDEs in any number of spatial dimensions.  While problems in higher dimensions 
generally require a more careful choice of the finite difference stencil used 
to approximate spatial derivatives in order to be able to combine 
leading-order terms in the discretization error, there are no fundamental 
barriers that prevent the use of optimal time step selection.

Domain geometry of the domain, which primarily impacts the way boundary 
conditions are imposed, is an important feature of problems in more than one 
spatial dimension.  While it is usually straightforward to handle problems on 
regular domains, problems on irregular domains can pose a greater challenge 
because the location of the domain boundary may not be explicitly specified
and grid points near the boundary require special consideration.  Fortunately, 
because optimal time step selection is based on simple, localized finite 
difference stencils, irregular domains do not create any serious difficulties.  
Assuming that a `ghost cell' or `ghost value' 
approach~\cite{fedkiw_1999, osher_fedkiw_book, gibou_2005} 
is used to impose boundary conditions, the construction and analysis of finite 
difference schemes and calculation of optimal time steps is identical for 
problems on both regular and irregular domains.  For problems on 
irregular domains, the key issue is how values at ghost cell points should 
be calculated.  Once this issue has been addressed, devising high-order 
accurate methods for PDEs on irregular domains is simply a matter of applying 
optimal time step selection to the finite difference scheme used for problems 
on regular domains.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
To analyze the accuracy of the FD scheme for~(\ref{eq:model_PDE}) when the 
optimal time step is used (for a given grid spacing), we examine the remaining 
terms of the discretization error in the standard way (\ie ``big-O'' notation).
Elimination of the leading term in the discretization error leaves a local 
error of the form $O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ 
and $(q+1) > (s+1)$ are the orders of the spatial and temporal discretization 
errors, respectively, after the leading-order terms in the error have been 
removed.  Like the orders of the leading-order errors in 
(\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
error is a direct consequence of the method of lines approach to discretizing 
the PDE.

From the local error, we can compute the global error of the numerical 
solution over extended intervals of time by using the heuristic argument
that the global error is equal to the local discretization 
error divided by $\dt$.  Using heuristic argument, which is rigorously 
justified in many texts on numerical methods such as~\cite{gko_book}, we 
find that the global error of the FD scheme using an optimal time step is 
\beq
O(\dx^p) + O(\dt^q).
\label{eq:global_error_ots}
\eeq
Observe that the global error for the FD scheme when an arbitrary time
step is used has exactly the same form as~(\ref{eq:global_error_ots}) with
$p$ and $q$ replaced by $r$ and $s$, respectively.  Thus, we see that 
using an optimal time step boosts the accuracy of the FD scheme.

When interpreting~(\ref{eq:global_error_ots}), it is important to recognize
that $\dx$ and $\dt$ are \emph{not} independent because the time 
step has been set to $\dto$; there is really only one parameter, 
$\dx$, that controls the numerical accuracy of the FD scheme.  This is an 
important point that we will address further in the next section.

\subsubsection*{Formal vs. Practical Accuracy}
For time dependent PDEs, evaluating the order of accuracy for a numerical
method is subtle because there are formally two separate orders of accuracy 
to consider\footnote{The situation is more straightforward for 
time-independent PDEs because there is usually only one order of accuracy to 
be concerned with.}:  temporal and spatial.  While it is theoretically
interesting and important to understand how the error depends on both the 
grid spacing $\dx$ and the time step $\dt$, in practice the accuracy 
of a numerical scheme is always controlled by only one of the two numerical
parameters.

Even though spatial and temporal orders of accuracy for a numerical method
are formally separate, one will alway dominate for any given choice of 
$\dx$ and $\dt$.  For example, when solving the diffusion  equation 
using the backward Euler method for time integration with a second-order 
central difference stencil for the Laplacian, formal analysis shows that 
the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
stability constraints on the numerical parameters, the time
step and grid spacing are free to vary independently.  In this situation, the 
practical error for the method depends on the relative sizes of the time step 
and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
$O(\dt)$ which means that the error in the numerical solution is 
primarily controlled by the time step.  Similarly, when 
$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
the error is controlled by the grid spacing.  Finally, when 
$\dt  = O(\dx^2)$, the practical error is 
$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
primarily controlled by one of the two numerical parameters, and varying
the subdominant parameter while holding the dominant parameter fixed does 
not significantly affect the error.

When there are constraints on the numerical parameters, there is often 
less freedom in choosing the controlling parameter.  For instance, if we 
solve the diffusion equation using a forward Euler time integration scheme 
with a second-order central difference stencil for the Laplacian, stability
considerations require that we choose $\dt = O(\dx^2)$.  
Combining this stability constraint with the formal error for the scheme
shows that the practical error is 
$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
method is completely controlled by the grid spacing; the temporal error can 
never dominate the spatial error.

Numerical methods for time dependent PDEs are commonly designed so that 
the global error is primarily controlled by the grid spacing.  The numerical
errors introduced by the time integration scheme are typically \emph{forced}
to be subdominant to the spatial discretization errors manually or as a result
of stability constraints.  In other words, the operational procedure used 
when developing numerical methods is usually to first choose a spatial 
discretization scheme and grid spacing so that a desired level of accuracy 
is achieved and then select a time integration scheme and time step which 
ensures that the temporal errors do not overcome the spatial errors.  
This methodology explains the tendency to evaluate numerical schemes for
time dependent PDEs predominantly on their order of accuracy with respect 
to the grid spacing.

Optimal time step selection fits into this general framework for developing
numerical methods for time dependent PDEs.  Because the optimal time step 
is calculated explicitly as a function of the grid spacing, the technique
of optimal time step selection places an even more restrictive constraint
on $\dt$ than stability considerations.  Thus, like numerical
methods with stability constraints, the accuracy of an FD scheme using 
the optimal time step is completely controlled by the grid spacing.  The 
order of accuracy of a FD scheme advanced in time using $\dto$ 
depends on the functional dependence the optimal time step on $\dx$.
For the method of lines, we know that $\dto = O(\dx^{r/s})$, 
so the global error derived in the previous section reduces to 
\beq
O \left( \dx^{\min(p,rq/s)} \right).
\label{eq:global_error_ots_simplified}
\eeq

\subsubsection*{\label{sec:computational_performance} 
                Computational Performance}
In general, $r/s > 1$ and yields an optimal time step that is on the order
of the time step allowed by the stability constraint for an explicit method.
As a result, using the optimal time step to advance the FD scheme in time
may seem to be overly restrictive and lead to poor computational performance.
However, these apparent drawbacks are more than compensated by the boost in 
the order of accuracy.  For example, when solving the 1D diffusion equation 
using a forward Euler time integration scheme with a second-order central 
difference stencil for the Laplacian, we obtain a fourth-order scheme 
when $\dt$ is set equal to the optimal time step.  As
Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
scheme that is computationally cheaper than traditional methods that use 
the same low-order FD stencils.  The performance gain for problems in higher 
spatial dimensions is even more impressive.  As we can see in 
Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
simple forward Euler scheme with the optimal time step and the Crank-Nicholson
method grows with the spatial dimension of the problem.

\begin{table}[tbh]
\caption{\label{tab:comp_perf_vs_err} 
   Computational cost as a function of the global numerical error $E$
   for various FD schemes that numerically solve the 1D diffusion equation.  
   For all FD schemes, the standard second-order central difference 
   stencil is used to discretize the Laplacian, and the numerical 
   solution is assumed to be computed over a fixed time interval.  
A less negative power of $E$ 
   for the memory and compute time means that a FD scheme is more 
   computationally efficient.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The 
   time step for the forward/backward Euler and Crank-Nicholson schemes 
   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
   time step for the backward Euler scheme is chosen so that the temporal 
   error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the implicit methods.}  \\
  \hline 
  Forward Euler    & $O\left( E^{1/2} \right)$ 
                   & $O\left (E \right)$ 
                   & $O\left( E^{-1/2} \right)$ 
                   & $O\left( E^{-3/2} \right)$ \\
  Backward Euler   & $O\left( E^{1/2} \right)$ 
                   & $O\left( E \right)$ 
                   & $O\left( E^{-1/2} \right)$ 
                   & $O\left( E^{-3/2} \right)$ \\
  Crank-Nicholson  & $O\left( E^{1/2} \right)$ 
                   & $O\left( E^{1/2} \right)$ 
                   & $O\left( E^{-1/2} \right)$ 
                   & $O\left( E^{-1} \right)$ \\
  Forward Euler with OTS  & $O\left( E^{1/4} \right)$ 
                   & $O\left( E^{1/2} \right)$ 
                   & $O\left( E^{-1/4} \right)$ 
                   & $O\left( E^{-3/4} \right)$ \\ 
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\begin{table}[tbh]
\caption{\label{tab:comp_perf_vs_dim}
   Comparison of the computational cost as a function of the global numerical 
   error $E$ for the forward Euler scheme with optimal time step and the 
   Crank-Nicholson method.  Both FD schemes are used to solve the diffusion 
   equation and use a second-order, isotropic stencil is used to discretize 
   the Laplacian.  For a problem in $d$ spatial dimensions, the memory 
   requirement for each scheme is estimated to be $O \left(1/\dx^d \right)$, 
   and the computation time is approximated as 
   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, which assumes that we 
   are using an efficient $O(N) = O \left( 1/\dx^d \right)$ algorithm to 
   invert the matrices that arise for the Crank-Nicholson method.
   A less negative power of $E$ for the memory and compute time
   means that a FD scheme is more computationally efficient.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{|c|c|c|c|c|}
  \hline
  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
  \cline{2-3} \cline{4-5} 
    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
  & {\bf Memory} & {\bf Compute Time} \\
  \hline 
  $1$ & $O\left( E^{-1/4} \right)$ 
      & $O\left( E^{-3/4} \right)$ 
      & $O\left( E^{-1/2} \right)$ 
      & $O\left( E^{-1} \right)$ \\ 
  $2$ & $O\left( E^{-1/2} \right)$ 
      & $O\left( E^{-1} \right)$ 
      & $O\left( E^{-1} \right)$ 
      & $O\left( E^{-3/2} \right)$ \\ 
  $3$ & $O\left( E^{-3/4} \right)$ 
      & $O\left( E^{-5/4} \right)$ 
      & $O\left( E^{-3/2} \right)$ 
      & $O\left( E^{-2} \right)$ \\
  $d$ & $O\left( E^{-d/4} \right)$ 
      & $O\left( E^{-(d+2)/4} \right)$ 
      & $O\left( E^{-d/2} \right)$ 
      & $O\left( E^{-(d+1)/2} \right)$ \\ 
  \hline 
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\subsection{\label{sec:ots_linear_pde} 
            General Constant Coefficient, Linear PDEs} 
In Section~\ref{sec:ots_model_pde}, we presented the technique of optimal 
time step selection in the context of a simple model PDE with only two terms.  
To extend the technique to general constant coefficient, linear PDEs of the 
form
\beq
  \frac{\partial u}{\partial t} = 
  \left( \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} \right) + f(x,t),
  \label{eq:linear_PDE}
\eeq
the procedure must be augmented in order to achieve any boost in the accuracy.
Because it is unlikely that all of the dominant terms in the error can be
combined into a single term, the leading-order discretization error is
a sum of several terms:
\bea
  (L u) (\alpha \dx^r - \beta \dt^s) \dt
  + \sum_k (L_k u) \dx^r \dt 
  + \sum_k (L'_k u) \dt^{s+1} 
  + (G f) \dt^{s+1}
  \label{eq:leading_order_error_linear_PDE_MOL}
\eea
where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
(\ref{eq:leading_order_error_model_PDE_general}) and
(\ref{eq:leading_order_error_model_PDE_MOL}), 
$L_k$ and $L'_k$ are the spatial differential operators associated with the 
leading-order terms in the discretization error that are not involved in the 
choice of the optimal time step, and $G$ is a spatio-temporal differential 
operator that acts on source term $f$.  
Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
operators because any that arise can be eliminated by using the PDE.  Also, 
there is no explicit spatial discretization error of order $r$ that originates 
from the source term because it is not differentiated with respect to space 
when the semi-discrete equations~(\ref{eq:method_of_lines}) are derived from 
the PDE.

From~(\ref{eq:leading_order_error_linear_PDE_MOL}), it is clear that setting 
$\dt = \dto$ is not enough to completely eliminate the 
leading-order error.  To eliminate the vestigial errors, we need to add a few 
correction terms to the numerical scheme in exactly the same spirit used in 
the design of the modified nine-point formula for the Poisson 
equation~\cite{iserles_book}.  The required correction terms are precisely 
those that cancel out the second, third, and fourth terms in 
(\ref{eq:leading_order_error_linear_PDE_MOL}).  Fortunately, it is 
straightforward to compute and incorporate these terms into the numerical 
scheme.  At each time step, the second and third terms in 
(\ref{eq:leading_order_error_linear_PDE_MOL}) can be approximated using
finite differences applied to the solution $u$ at the current time.
The last term could also be approximated using finite differences, or,
alternatively, it could be calculated analytically because the source terms 
are usually known functions.  With the addition of these correction terms, 
the leading-order discretizaton error is completely eliminated, and the
global error can be determined using the analysis in 
Section~\ref{sec:error_analysis}.

It is important to point out that the finite difference stencils used to 
approximate the vestigial errors must be chosen carefully.  Special attention 
is required to ensure that the FD stencils are of sufficiently high-order; 
otherwise, we will lose the accuracy gained from using an optimal time step 
for the FD scheme.  Since the specific form of the correction terms are 
highly PDE-dependent, we defer further discussion to 
Section~\ref{sec:applications_1d} and~\ref{sec:applications_multidim} where 
we apply optimal time step selection to several specific PDEs.

\subsection{Semilinear PDEs}
Semilinear PDEs are an important class of PDEs that have the form
\beq
  \frac{\partial u}{\partial t} = A(x,t) \frac{\partial^n u}{\partial x^n}
  + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}},
      \frac{\partial^{n-2} u}{\partial x^{n-2}}, \ldots,
      \frac{\partial u}{\partial x}, u \right)
  + f(x,t)
  \label{eq:semilinear_PDE}
\eeq
where $F$ is an arbitrary function of the lower-order spatial derivatives
of $u$ and $f(x,t)$ is a source term.  While, the coefficient on the 
leading-order spatial derivative is generally allowed to be function of 
space and time, optimal time step selection is only applicable to PDEs where
$A(x,t)$ is a constant.  Because the PDE is linear in the leading-order 
spatial derivative, it is natural to construct a finite difference scheme that 
treats $F$ in an explicit manner and lets stability considerations determine 
whether the $A \frac{\partial^n u}{\partial x^n}$ term is handled explicitly 
or implicitly.  Assuming that sufficiently high-order stencils are used to 
compute $F$, the leading-order discretization error for such a finite 
difference scheme is given by
\bea
  (L u) (\alpha \dx^r - \beta \dt^s) \dt
  + \left( H u \right) \dx^r \dt 
  + \left( H' u \right) \dt^{s+1}
  + (G f) \dt^{s+1}
  \label{eq:leading_order_error_semilinear_PDE_MOL}
\eea
where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
(\ref{eq:leading_order_error_model_PDE_general}) and
(\ref{eq:leading_order_error_model_PDE_MOL}), $H$ and $H'$ are spatial 
differential operators associated with the leading-order terms in 
the discretization error that arise from the nonlinear term 
in~(\ref{eq:semilinear_PDE}), and $G$ is a spatio-temporal differential 
operator that acts on $f$.  To compute the optimal time step and boost
the accuracy of the finite difference scheme, proceed in exactly
the same manner as for general constant coefficient, linear PDEs.

\subsection{Imposing Boundary Conditions}
*** CHOICE OF DISCRETIZATION FOR BOUNDARIES IS IMPORTANT***

\subsection{Limitations of Optimal Time Step Selection}
As with all methods, optimal time step selection has its limitations.
A critical assumption that we have made throughout our analysis is that
the leading-order spatial derivative in the PDE does not have a spatially
varying coefficient.  Without this assumption, the first term in the
leading-order discretization error would be of the form 
$(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
where $\alpha$ and $\beta$ are now functions of space and time.  This
small change to the error makes it impossible to derive a single optimal
time step.  In a sense, each grid point would a different optimal time
step that also depends on the time.  For a similar reason, optimal time step 
selection is fails for quasilinear and fully nonlinear PDEs.  The issue for 
these types of PDEs is that the coefficients in $P(\dx, \dt)$ become 
functions of the solution $u$ which also varies in space and time.

\section{\label{sec:applications_1d} 
         Application to PDEs in One Space Dimension}
Because there are no geometry or anistropy considerations, optimal time step 
selection is very useful for boosting the accuracy of many finite difference 
schemes in one spatial dimension.  In this section, we demonstrate its 
utility in designing finite difference schemes to solve several classical 
linear and semilinear PDEs where analytical solutions are available for 
comparison with numerical solutions.

\subsection{Linear Advection Equation}
For our first application of optimal time step selection, we consider 
finite difference schemes for the linear advection equation
~\cite{leveque_book_1992, leveque_book_2002, gko_book}:
\beq
  \frac{\partial u}{\partial t} + a \frac{\partial u}{\partial x} = 0,
  \label{eq:advection_eqn_1d}
\eeq
where $a$ is the flow speed, which we take to be positive for convenience.  
For an initial condition of the form $u(x,0) = f(x)$, it is easy to use the 
method of characteristics to show that the analytical solution is $f(x-at)$.  
While it is unlikely that one would actually use a numerical method to 
solve~(\ref{eq:advection_eqn_1d}), it provides an interesting example where 
optimal time step selection dramatically improves the accuracy of the 
numerical solution.

There are many finite difference schemes that can be used to 
solve~(\ref{eq:advection_eqn_1d}).  Several examples are given 
in~\cite{leveque_book_1992}.  Perhaps the simplest, stable method uses 
forward Euler for time integration with a first-order upwind discretization 
for the spatial derivative term:
\beq
  u^{n+1}_j = u^{n}_j 
  - a \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right),
  \label{eq:advection_eqn_1d_FD_scheme}
\eeq
where superscripts and subscripts denote the time and space indices, 
respectively.  Using standard error analysis, it is straightforward to show
that method is formally first-order in both space and time.

To compute the optimal time step, we begin by analyzing the discretization
error for~(\ref{eq:advection_eqn_1d_FD_scheme}).  In general, we only need to 
keep track of the leading-order terms in the discretization error.  However,
for the advection equation, it is instructive to retain all of them.  
Using Taylor series to expand the true solution of the the PDE at time step
$(n+1)$, we find that
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\pt^k} \right|_j
  \label{eq:advection_eqn_1d_time_err}
\eea
where $\tu$ denotes the true solution of the PDE and 
$\tu^n_j \equiv \tilde{u}(j\dx, n\dt)$.  Similarly, Taylor series expansion 
of the upwind finite difference approximation to the spatial derivative 
yields
\bea
  \frac{\tu^{n}_j - \tu^{n}_{j-1}}{\dx} = 
  \left. \frac{\partial \tu}{\px} \right|_j
  - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\px^k} \right|_j.
  \label{eq:advection_eqn_1d_space_err}
\eea
Next, we use the PDE~(\ref{eq:advection_eqn_1d_FD_scheme}) to replace all of 
the time derivatives in~(\ref{eq:advection_eqn_1d_time_err}) with spatial
derivatives to obtain
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  + \sum_{k=1}^\infty \frac{\left( -a \dt \right)^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\px^k} \right|_j.
  \label{eq:advection_eqn_1d_time_err_modified}
\eea
Finally, by combining~(\ref{eq:advection_eqn_1d_time_err_modified}) and
and~(\ref{eq:advection_eqn_1d_space_err}) with the finite difference 
scheme~(\ref{eq:advection_eqn_1d_FD_scheme}), we obtain the evolution
equation for the error $e \equiv u - \tu$:
\bea
  e^{n+1}_j = e^{n}_j 
    &-& a \dt \left( \frac{e^{n}_j - e^{n}_{j-1}}{\dx} \right) \nonumber \\
    &+& a \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
        \left. \frac{\partial^k \tu^n}{\px^k} \right|_j
        \left[ \dx^{k-1} - \left( a \dt \right)^{k-1} \right].
  \label{eq:advection_eqn_1d_err_eqn}
\eea
As expected, the error $e$ satifies the same finite difference approximation 
as the solution with an additional source term which is the discretization 
error for the scheme.

Extracting the leading-order term of the discretization error, we find that 
it is given by
\beq
  \left( \left. \frac{a}{2} \frac{\partial^2 \tu^n}{\px^2} \right|_j \right)
  \left[ \dx - a \dt \right] \dt.
\eeq
Thus, the optimal time step is simply $\Delta t_{opt} = \Delta x / a$.
For the linear advection equation, setting the time step equal to the optimal 
time step has an amazing consequence -- not only does the leading-order term
in the discretization error vanish, the \emph{entire} local truncation 
error at each time step is completely eliminated!  In other words, the error 
in the numerical solution is due solely to errors in the initial conditions.
Assuming that the initial conditions are set exactly, there is absolutely no 
error in the numerical solution for all times!

This surprising result, which is unique to the advection equation, is 
well-known in the numerical methods community and is referred to as the unit 
CFL condition~\cite{leveque_book_2002}.  It is usually proved by examining 
how characteristic lines of the advection equation pass through grid points.  
It is interesting to see that it also naturally arises through an analysis of 
the discretization errors and optimal time step selection.

\subsection{Diffusion Equation}
For an example where optimal time step selection yields a more typical boost 
in the order of accuracy, we turn our attention to the diffusion equation:
\beq
  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2},
  \label{eq:diffusion_eqn_1d}
\eeq
where $D$ is the diffusion constant.  As with the advection equation,
there are many finite difference schemes that can be used to solve the 
diffusion equation.  Keeping with the philosophy of using simple, low-order 
methods to construct finite difference schemes, we apply optimal time step 
selection to a scheme that uses forward Euler for time integration and the 
standard second-order central difference approximation for the Laplacian:
\beq
  u^{n+1}_j = u^{n}_j 
  + D \dt 
    \left( \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right),
  \label{eq:diffusion_eqn_1d_FD_scheme}
\eeq
Standard error analysis shows that scheme is formally first-order in time and 
second-order in space.  Due to the stability constraint 
$\dt \le \dx^2/2D$, the scheme is $O(\dx^2)$ accurate overall.

As before, we begin by analyzing the leading-order terms of the 
discretization error for the finite difference scheme.  Employing Taylor
series expansions, we find that the true solution satisfies
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  + \dt \frac{\partial \tu^n}{\pt} 
  + \frac{\dt^2}{2} \frac{\partial^2 \tu^n}{\pt^2} + O \left( \dt^3 \right)
  \label{eq:diffusion_eqn_1d_time_err}
\eea
and the central difference for the Laplacian satisfies
\bea
  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
  \left. \frac{\partial^2 \tu^n}{\px^2} \right|_j
  + \frac{\dx^2}{12} \left. \frac{\partial^4 \tu^n}{\px^4} \right|_j
  + O(\dx^4)
  \label{eq:diffusion_eqn_1d_space_err}
\eea
Making use of the PDE~(\ref{eq:diffusion_eqn_1d}), 
(\ref{eq:diffusion_eqn_1d_time_err}) becomes
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  + D \dt \frac{\partial^2 \tu^n}{\px^2} 
  + \frac{(D \dt)^2}{2} \frac{\partial^4 \tu^n}{\px^4} 
  + O \left( \dt^3 \right)
  \label{eq:diffusion_eqn_1d_time_err_modified}
\eea
Combining these results with the finite difference 
scheme~(\ref{eq:diffusion_eqn_1d_FD_scheme}), we arrive at the evolution 
equation for the error $e \equiv u - \tu$:
\bea
  e^{n+1}_j = e^{n}_j 
  &+& D \dt 
    \left( \frac{e^{n}_{j+1} -2 e^{n}_j + e^{n}_{j-1}}{\dx^2} \right)
  \nonumber \\
  &+& \frac{\partial^4 \tu^n}{\px^4} 
      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
      + O(\dt \dx^4) + O(\dt^3)
  \label{eq:diffusion_eqn_1d_err_eqn}
\eea
Again, we see that the discretization error acts as a source term for the 
error, which satisfies the same evolution equation as the numerical solution 
$u$.

From~(\ref{eq:diffusion_eqn_1d_err_eqn}), we see that the optimal time step 
is given by $\dto = \dx^2/6D$.  Using~(\ref{eq:global_error_ots}),
the order of the scheme with this choice of time step is 
$O(\dx^4) + O(\dt^2) = O(\dx^4)$; we have boosted the accuracy of original 
scheme from second to fourth order.  Note that $\dto$ is stable, so there
is no need to switch to an implicit time integration scheme.

To demonstrate the improvement in accuracy of the forward Euler scheme when 
the optimal time step is used, we compare the numerical errors of three finite 
difference schemes for the diffusion equation: forward Euler with optimal
time step, forward Euler with suboptimal time step, and Crank-Nicholson.  
As we can see in Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward 
Euler with optimal time step is fourth-order accurate while both forward Euler 
with suboptimal time step and Crank-Nicholson are only second-order accurate.  
Using the forward Euler scheme with an optimal time step has a clear 
advantage over both of the two other schemes.  Even with as few as 20 grid 
points, the error in solution computed by the forward Euler scheme with an 
optimal time step is orders of magnitude smaller.

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_no_src_error_vs_N}} 
\caption{Error analysis for three finite difference schemes for the 
diffusion equation: forward Euler with optimal time step (circles), forward 
Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
Using each method, we solve the diffusion equation with $D = 1$ on the 
domain $0 < x < 1$ subject to the boundary conditions 
$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
and the initial condition
$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
- 4 \sin(11 \pi x/2)$.  
The exact solution for this problem is
$u(x,t) = 1 + x/2 
        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 Dt/4) 
        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 Dt/4)
        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 Dt/4)$.
The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
method.  Note that the errors for the forward Euler with suboptimal time 
step and Crank-Nicholson schemes lie almost directly on top of each other
at the resolution of the figures.
}
\label{fig:diffusion_eqn_1d_no_src_error}
\end{center}
\end{figure}

As mentioned in Section~\ref{sec:computational_performance}, one of the 
advantages of using optimal time step selection is the superior scaling of 
the computational cost with respect to the numerical error.  It is worth 
mentioning that another way to achieve the same computational performance
would be to use a Crank-Nicholson time integration scheme with a fourth-order 
discretization of the Laplacian~\cite{gibou_2005}.  By using an 
$O\left(\dx^2 \right)$ time step and a linear solver that scales linearly 
with the size of the matrix, we achieve an overall fourth-order accurate 
method that has the same computational complexity as the forward Euler scheme
described above with an optimal time step.  However, because the forward Euler 
scheme with optimal time step will still outperform the Crank-Nicholson scheme 
because the latter requires the solution of at least one linear system of
equation for each time step.

\subsubsection*{Diffusion Equation with Source Term} 
As discussed in Section~(\ref{sec:ots_linear_pde}), the procedure for optimal 
time step selection needs to be slightly modified when source terms are 
present.  If we add a source term, $f(x,t)$, to the right-hand side 
of~(\ref{eq:diffusion_eqn_1d}), the finite difference 
scheme~(\ref{eq:diffusion_eqn_1d_FD_scheme}) picks up an additional term
\beq
  u^{n+1}_j = u^{n}_j 
  + D\dt 
    \left( \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} + f^n_j \right).
  \label{eq:diffusion_eqn_1d_src_FD_scheme}
\eeq
and~(\ref{eq:diffusion_eqn_1d_time_err_modified}) becomes
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  &+& D\dt \left( \frac{\partial^2 \tu^n}{\px^2} + f^n_j \right)
  \nonumber \\
  &+& \frac{(D\dt)^2}{2} \frac{\partial^4 \tu^n}{\px^4} 
  + \frac{\dt^2}{2} \left( D\frac{\partial^2 f}{\px^2}
                         + \frac{\partial f}{\pt} \right)
  + O \left( \dt^3 \right)
  \label{eq:diffusion_eqn_1d_src_time_err_modified}
\eea
Therefore, the leading-order discretization error is given by
\beq
  \frac{\partial^4 \tu^n}{\px^4} 
    \left[ \frac{\dx^2}{12} - \frac{D \dt}{2} \right] (D \dt)
    - \frac{\dt^2}{2} \left( D \frac{\partial^2 f}{\px^2} 
                           + \frac{\partial f}{\pt} \right).
  \label{eq:diffusion_eqn_1d_src_err}
\eeq
To eliminate the low-order errors introduced by the source term, we 
simply add the following correction term to the right-hand side of 
the finite difference scheme~(\ref{eq:diffusion_eqn_1d_src_FD_scheme}):
\beq
\frac{\dt^2}{2} \left( D \left( f_{xx} \right)^n_j 
  + \left( f_t \right)^n_j \right),
\eeq
where the derivatives of $f$ may be evaluated analytically or using 
sufficiently high order finite difference stencils.

In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of the
OTS forward Euler scheme with and without the correction terms.  As we can 
see, OTS forward Euler with correction terms is fourth-order accurate while 
OTS forward Euler without correction terms is only second-order accurate.  
These results clearly demonstrate the importance of the correction terms for
achieving high-order accuracy.

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_error_vs_N}} 
\caption{Error analysis for four finite difference schemes for the 
diffusion equation: forward Euler with optimal time step (circles), forward 
Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
Using each method, we solve the diffusion equation with $D = 1$ on the 
domain $0 < x < 1$ subject to the boundary conditions 
$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
and the initial condition
$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
- 4 \sin(11 \pi x/2)$.  
The exact solution for this problem is
$u(x,t) = 1 + x/2 
        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 D t/4) 
        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 D t/4)
        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 D t/4)$.
The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
method.  Note that the errors for the forward Euler with suboptimal time 
step and Crank-Nicholson schemes lie almost directly on top of each other
at the resolution of the figures.
}
\label{fig:diffusion_eqn_1d_src_error}
\end{center}
\end{figure}

\subsection{Advection-Diffusion Equation}
Next, we use optimal time step selection in the context of the 
advection-diffusion equation, a more general linear, constant coefficient PDE:
\beq
  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2} 
  + a \frac{\partial u}{\px},
  \label{eq:ADE_1d}
\eeq
where $D$ is the diffusivity and $a$ is the flow speed.  For this problem,
we use forward Euler for time integration and the second-order central 
difference approximations for the diffusion and advection terms:
\beq
  u^{n+1}_j = u^{n}_j 
  + \dt 
    \left( D \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} 
         + a \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right).
  \label{eq:ADE_1d_FD_scheme}
\eeq
This scheme is formally first-order in time and second-order in space.  
Due to the stability constraint $\dt \le \dx^2/(2 D + a \dx)$, the 
scheme is $O(\dx^2)$ accurate overall (in the limit $\dx \rightarrow 0$).

To analyze the discretization error, we can reuse 
results~(\ref{eq:diffusion_eqn_1d_time_err}) 
and~(\ref{eq:diffusion_eqn_1d_space_err})
from our analysis of the diffusion equation.  In addition, we need the 
leading-order discretization error for the central difference approximation 
of $u_x$, which is given by
\bea
  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
  \left. \frac{\partial \tu^n}{\px} \right|_j
  + \frac{\dx^2}{6} \left. \frac{\partial^3 \tu^n}{\px^3} \right|_j
  + O(\dx^4).
  \label{eq:ADE_1d_ux_err}
\eea
Using the advection-diffusion equation~(\ref{eq:ADE_1d})
to convert time derivatives into spatial derivatives 
in~(\ref{eq:diffusion_eqn_1d_time_err}), we find that 
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} 
               + a \frac{\partial \tu^n}{\px} \right)
  \nonumber \\
  &+& \frac{\dt^2}{2} 
      \left( 
        D^2 \frac{\partial^4 \tu^n}{\px^4} 
      + 2 D a \frac{\partial^3 \tu^n}{\px^3} 
      + a^2 \frac{\partial^2 \tu^n}{\px^2} 
      \right) 
  + O \left( \dt^3 \right)
  \label{eq:ADE_1d_time_err_modified}
\eea
We then obtain the local truncation error, $\tau^n_j$, for the finite 
difference scheme~(\ref{eq:ADE_1d_FD_scheme}) using 
the usual procedure of deriving the evolution equation for the error 
and extracting the source term:
\bea
  \tau^n_j &=&
      \left( \frac{\partial^4 \tu^n}{\px^4} 
           + \frac{2a}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
  \nonumber \\
  &-& \frac{a^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
      + O(\dt \dx^4) + O(\dt^3)
  \label{eq:ADE_1d_err_eqn}
\eea

As we discussed in Section~\ref{sec:ots_linear_pde}, boosting the order
of accuracy for general linear, constant coefficient PDEs requires both 
use of an optimal time step and addition of correction terms.  
In this case, the optimal time step is $\dto = \dx^2/6D$ and the
correction term is 
\beq
  \frac{a^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
  \label{eq:ADE_1d_corr_term}
\eeq 
Note that using the standard second-order central difference stencil to 
compute the Laplacian in the correction term is sufficient to ensure that no 
errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into $\tau^n_j$.  
With the optimal time step and correction term, the finite difference 
scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
Figure~\ref{fig:ADE_1d_error} demonstrates the 
improved accuracy of the finite difference 
scheme~(\ref{eq:ADE_1d_FD_scheme}) when the optimal
time step and correction term are used.

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{\includegraphics{figures/adv_diff_eqn_1d_error_vs_N}} 
\caption{Error analysis for three finite difference schemes for the 
diffusion equation: forward Euler with optimal time step (circles), forward 
Euler with suboptimal time step (squares), and Crank-Nicholson (crosses).  
Using each method, we solve the diffusion equation with $D = 1$ on the 
domain $0 < x < 1$ subject to the boundary conditions 
$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
and the initial condition
$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
- 4 \sin(11 \pi x/2)$.  
The exact solution for this problem is
$u(x,t) = 1 + x/2 
        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 D t/4) 
        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 D t/4)
        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 D t/4)$.
The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
method.  Note that the errors for the forward Euler with suboptimal time 
step and Crank-Nicholson schemes lie almost directly on top of each other
at the resolution of the figures.
}
\label{fig:ADE_1d_error}
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
\ \ \ \ \
\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
\caption{Comparison of numerical solutions for the 1D advection-diffusion
equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
optimal time step and correction terms (left) and forward Euler time 
stepping with suboptimal time step (right).  The latter scheme uses a 
time step $\Delta t = \Delta x^2 / 3 D$.  Both figures show the numerical
solution at $t = 1$ computed on the domain $-10 < x < 10$ using 50 grid points.
The diffusion constant $D$ and the flow speed $A$ are set to $2$ and $5$,
respectively.
The initial conditions are taken to be $u(x,0) = \exp \left( -x^2/4 \right)$
and the boundary conditions are imposed exactly by using the analytical 
solution $u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
where $T = Dt+1$.
}
\label{fig:burgers_1d_solns}
\end{center}
\end{figure}

\subsubsection*{First-order Upwind Discretization of Advection Term}
It is common to use upwind discretizations as 
in~(\ref{eq:advection_eqn_1d_FD_scheme}) for advection terms that arise
in PDEs.  If instead of a central difference approximation, we use a
first-order upwind stencil\footnote{The appropriate upwind stencil depends
on the sign of $a$.} for the advection term in~(\ref{eq:ADE_1d}), the 
discretization error for the finite difference scheme is dominated by the 
first-order error introduced by the stencil for the advection term:
\bea
  \tau^n_j &=&
      \frac{\partial^2 \tu^n}{\px^2} 
      \left[ \frac{\dx}{2} - \frac{|a| \dt}{2}  \right] (|a| \dt)
  \nonumber \\
   & & 
    - \frac{\dt^2}{2} 
      \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
           + 2 a D \frac{\partial^3 \tu^n}{\px^3}
      \right)
      + O(\dt \dx^2) + O(\dt^3)
  \label{eq:ADE_1d_upwind_err_eqn}.
\eea
Thus, we can see that for a finite difference scheme based a first-order
upwind discretization of the advection term, we optimal time step is 
$\dto = \dx/|a|$ (as for the pure advection equation) and the required
correction term is
\beq
    \frac{\dt^2}{2} 
      \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
           + 2 a D \frac{\partial^3 \tu^n}{\px^3}
      \right)
  \label{eq:ADE_1d_upwind_corr_term}.
\eeq
This scheme is second-order overall as long as the finite difference 
approximations to all of the derivatives in~(\ref{eq:ADE_1d_upwind_corr_term})
are at least first-order accurate.

Comparing the accuracies achieved using a second-order versus a 
first-order discretization of the advection term, it is clear that 
the power of optimal time step selection is highly dependent on the 
the original finite difference scheme and underscores the importance of
carefully choosing the finite difference scheme even when optimal time
step selection is used.

\subsection{Viscous Burgers Equation}
For an example of optimal time step selection applied to a semilinear equation,
we consider the viscous Burgers equation~\cite{whitham_book}:
\beq
  \frac{\partial u}{\pt} + u \frac{\partial u}{\px} = 
     \nu \frac{\partial^2 u}{\px^2} 
  \label{eq:burgers_1d}
\eeq
where $\nu$ is viscosity.  Notice that this problem has the same structure
to the advection-diffusion equation, except that the advection velocity now 
depends explicitly on the solution $u$.  To solve this equation numerically,
we will use forward Euler for time integration and second-order central 
difference approximations for the diffusion and advection terms:
\beq
  u^{n+1}_j = u^{n}_j 
  + \dt 
    \left( \nu \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} 
         - u^n_j 
      \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] 
    \right).
  \label{eq:burgers_1d_FD_scheme}
\eeq
This scheme is formally first-order in time and second-order in space.  
Due to the stability constraint $\dt \le \dx^2/(2 \nu + a \dx)$, the 
scheme is $O(\dx^2)$ accurate overall (in the limit $\dx \rightarrow 0$).

Because the finite difference scheme for the viscous Burgers equation is 
very similar to the formally second-order advection-diffusion 
equation~(\ref{eq:ADE_1d_FD_scheme}), the analysis for the error analysis
is almost the same as in the previous section.  Following the usual 
procedure, we find that the local truncation error takes the form
\bea
  \tau^n_j &=&
      \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
           - 2 \tu \frac{\partial^3 \tu^n}{\px^3} \right)
      \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
  \nonumber \\
  &+& \frac{\dt^2}{2} 
      \left( 
           4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
      \right)
      + O(\dt \dx^4) + O(\dt^3).
  \label{eq:burgers_1d_err_eqn}
\eea
Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
for the finite difference scheme is 
\beq
  - \frac{\dt^2}{2} 
      \left( 
           4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
      \right)
  \label{eq:burgers_1d_corr_term}
\eeq 
Note that using the standard second-order central difference stencils to 
compute the correction term is sufficient to ensure that the overall error
for the scheme is $O(\dx^4)$.  Figure~\ref{fig:burgers_1d_error} shows that
using an optimal time step and correction terms improves the accuracy of the 
finite difference scheme~(\ref{eq:burgers_1d_FD_scheme}).

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{\includegraphics{figures/burgers_eqn_1d_error_vs_N}} 
\caption{Error analysis for two finite difference schemes for the 
viscous Burgers equation: forward Euler with optimal time step (circles)
and forward Euler with suboptimal time step (squares).
Using each method, we solve the viscous Burgers equation ??
}
\label{fig:burgers_1d_error}
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
\ \ \ \ \ 
\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
\caption{Comparison of numerical solutions for the 1D viscous Burgers 
equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
optimal time step and correction terms (left) and forward Euler time 
stepping with suboptimal time step (right).  The latter scheme uses a 
time step $\Delta t = \Delta x^2 / 3 \nu$.  Both figures show the numerical
solution at $t = 2$ computed on the domain $0 < x < 10$ using 100 grid points.
The viscosity $\nu$ is set to $0.1$.  The initial conditions are taken to be 
$u(x,0) = 1 + \gamma \sqrt{\frac{\nu}{\pi}} \exp\left( -(x-1)^2/4 \nu \right) 
            / \left[ 1 + \frac{\gamma}{2} \ 
                         \mathtt{erfc}\left((x-1)/\sqrt{4\nu} \right) 
    \right]$,
where $\ln(1+\gamma)$ is the effective Reynolds number (set equal to 10) and 
the wave speed is set equal to $1$.  The boundary conditions are imposed 
exactly by using
the analytical solution
$u(x,t) = 1 + \gamma \sqrt{\frac{\nu}{\pi T}} 
              \exp\left( -(x-T)^2/4 \nu T \right) 
            / \left[ 1 + \frac{\gamma}{2} \ 
                         \mathtt{erfc}\left((x-T)/\sqrt{4\nu T} \right) 
    \right]$,
where $T = t+1$.
}
\label{fig:burgers_1d_solns}
\end{center}
\end{figure}

\subsection{Fourth-Order Parabolic Equation}
As a final example in one space dimension, we consider a PDE with high-order
spatial derivatives -- a fourth-order parabolic equation 
REPLACE WITH A MORE PHYSICAL NAME:
\beq
  \frac{\partial u}{\pt} = -\kappa \frac{\partial^4 u}{\px^4} + f(x,t),
  \label{eq:4th_order_parabolic_eqn_1d}
\eeq
where $\kappa$ is the ??.  From our previous examples, we know that 
optimal time step selection requires that the coefficient on leading-order 
spatial discretization error must be directly related to the second 
derivative of the solution $u$ with respect to time.  
From~(\ref{eq:4th_order_parabolic_eqn_1d}), we see that 
\beq
  \frac{\partial^2 u}{\pt^2} = 
    \kappa^2 \frac{\partial^8 u}{\px^8} 
  - \kappa \frac{\partial^4 f}{\px^4} 
  + \frac{\partial f}{\pt}
  \label{eq:4th_order_parabolic_eqn_1d_second_time_derivative},
\eeq
which indicates that the leading-order term in the spatial discretization 
error should involve the eighth-derivative of $u$.  In other words, the
finite difference approximation to the bilaplacian needs to be at least 
fourth-order accurate.  With this insight, let us consider a finite 
difference scheme for~(\ref{eq:4th_order_parabolic_eqn_1d}) that 
uses forward Euler for time integration and the a fourth-order central 
difference approximation for the bilaplacian:
\bea
  u^{n+1}_j = u^{n}_j - \kappa \dt B u^n, 
  \label{eq:4th_order_parabolic_eqn_1d_FD_scheme}
\eea
where the fourth-order discrete bilaplacian is given by
\beq
  B u^n = \frac{ -u^{n}_{j+3} + 12 u^{n}_{j+2} - 39 u^{n}_{j+1}
               + 56 u^{n}_j
               - 39 u^{n}_{j-1} + 12 u^{n}_{j-2} -u^{n}_{j-3} }
               {6 \dx^4}.
\eeq
This scheme is formally first-order in time and fourth-order in space.  
Due to the stability constraint $\dt \le \dx^4/? \kappa$, the scheme is 
$O(\dx^4)$ accurate overall.

Proceeding in the usual way, we use Taylor series expansions and the PDE to 
express the true solution, $\tu$, at time step $(n+1)$ in terms of the true
solution and its spatial derivatives at time step $n$: 
\bea
  \tu^{n+1}_j = \tu^{n}_j 
  &+& \dt \left( -\kappa \frac{\partial^4 \tu^n}{\px^4}  
             + f^n_j
        \right)
  \nonumber \\
  &+& \frac{\dt^2}{2} 
    \left(
      \kappa^2 \frac{\partial^8 \tu^n}{\px^8} 
    - \kappa \frac{\partial^4 f}{\px^4} 
    + \frac{\partial f}{\pt}
    \right)
  + O \left( \dt^3 \right)
  \label{eq:4th_order_parabolic_eqn_1d_time_err_modified}.
\eea
Next, we substitute $\tu$ into central difference approximation to the 
bilaplacian and obtain
\bea
  B \tu^n = \left. \frac{\partial^4 \tu}{\px^4} \right|_j 
  - \frac{7 \dx^4}{240} 
    \left. \frac{\partial^8 \tu}{\px^8} \right|_j 
  + O(\dx^6)
  \label{eq:4th_order_parabolic_eqn_1d_space_err}
\eea
Combining these results with the finite difference 
scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}), we find that the
local truncation error is given by
\bea
  \tau^n_j &=& 
     \frac{\partial^8 \tu^n}{\px^8} 
    \left[ \frac{7 \dx^4}{240} - \frac{\kappa \dt}{2}  \right] (\kappa \dt)
    \nonumber \\
    &+& \frac{\dt^2}{2} 
        \left( \kappa \frac{\partial^4 f}{\px^4} 
             - \frac{\partial f}{\pt}
        \right)
    + O(\dt \dx^6) + O(\dt^3)
  \label{eq:4th_order_parabolic_eqn_1d_err_eqn}
\eea
Thus, the optimal time step is $\dto = 7 \dx^4/120\kappa$ and the correction
term is 
\beq
  - \frac{\dt^2}{2} 
      \left( \kappa \frac{\partial^4 f}{\px^4} 
           - \frac{\partial f}{\pt}
      \right).
\eeq
Together, these modifications lead to a finite difference scheme that is 
sixth-order accurate.  As with the previous examples, the correction term 
can be computed analytically or numerically.

Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} demonstrates the 
accuracy of the finite difference 
scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) when the optimal
time step and correction terms are used.  Observe that the choice of the
finite difference approximation used to compute the correction terms
impacts the overall accuracy of the scheme.  Surprisingly, the
second-order central difference approximation for $f_{xxxx}$ yields a
more accurate solution than the fourth-order scheme.  Fortuitous cancellation 
of error is the origin of the superior results obtained by using the 
second-order scheme.  A cursory analysis shows that using a second-order 
stencil for $f_{xxxx}$ yields a $O(\dx^{10})$ error involving 
$\left(\partial_x\right)^{10}u$ and $\left(\partial_x\right)^{6}f$ with 
opposite signs.  Because these two terms are related to each other through the 
PDE~(\ref{eq:4th_order_parabolic_eqn_1d}), they can partially anihilate
each other.  On the other hand, when the fourth-order stencil is used
for $f_{xxxx}$, there is no chance for cancellation because the $O(\dx^{10})$ 
error only involves $\left(\partial_x\right)^{10}u$.
Again, we see that blindly using a higher-order scheme does not
necessarily lead to a more accurate numerical solution.

\begin{figure}[htb]
\begin{center}
\scalebox{0.33}{
  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
\ \ 
\scalebox{0.33}{
  \includegraphics{figures/4th_order_parabolic_eqn_1d_comp_time}} 
\caption{Error analysis for three finite difference schemes for the 
diffusion equation: forward Euler with optimal time step (circles), forward 
Euler with suboptimal time step (squares), and Crank-Nicholson (crosses).  
Using each method, we solve the diffusion equation with $\kappa = 1$ on the 
domain $0 < x < 1$ subject to the boundary conditions 
$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
and the initial condition
$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
- 4 \sin(11 \pi x/2)$.  
The exact solution for this problem is
$u(x,t) = 1 + x/2 
        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 \kappa t/4) 
        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 \kappa t/4)
        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 \kappa t/4)$.
The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
method.  Note that the errors for the forward Euler with suboptimal time 
step and Crank-Nicholson schemes lie almost directly on top of each other
at the resolution of the figures.
}
\label{fig:4th_order_parabolic_eqn_1d_error}
\end{center}
\end{figure}

As suggested by the discussion in this section, applying optimal time step 
selection to high-order PDEs requires the use high-order discretizations
of the spatial derivatives.  While the need to derive high-order finite 
difference stencils may seem like an unnecessary burden, it is important
to remember that high-order finite difference schemes for spatial derivatives
are almost always necessary to maximize the efficiency of explicit time
integration schemes for high-order PDEs\footnote{Because restrictive time 
step constraints are always a major limitation of explicit schemes for 
time dependent PDEs with high-order spatial 
derivatives~\cite{gko_book,greer_2006}, 
the overall error of finite difference approximations for high-order PDEs
is usually dominated by the error introduced by the spatial discretization.
As a result, it is crucial that numerical schemes for high-order PDEs use 
high-order discretizations for the spatial derivatives.  Using high-order 
discretizations increases the computational efficiency of numerical schemes 
and makes it possible to obtain accurate solutions on relatively coarse 
grids.  For a PDE of order $n$, explicit schemes typically require 
the time step to be $O(\dx^n)$.  Thus, it makes sense to choose the order of 
the finite difference stencils for the spatial derivatives to be at least 
$n/s$ where $s$ is the order of accuracy for the time integration scheme.  
This choice yields a finite difference scheme that with an error that is
$O(\dx^n)$ overall.}.  
Optimal time step selection merely provides a valuable way to leverage the 
effort spent deriving high-order finite difference schemes in order to further 
boost the overall accuracy of the numerical method.

Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} shows the computational 
performance of various finite difference methods for solving 
equation~(\ref{eq:4th_order_parabolic_eqn_1d_second_time_derivative}).
Observe the dramatic decrease in computational cost as a function of 
error by simply replacing a second-order discretization for the bilaplacian 
operator with a fourth-order discretization.  For the forward Euler scheme,
using an optimal time step drives the computational cost down even more.
Comparing the best forward Euler and Crank-Nicholson schemes shows that
the optimal forward Euler scheme more efficiently utilizes memory at the
cost of requiring more computation time.  For the fourth-order parabolic 
equation, it is important to note that this tradeoff between memory and 
computation only exists in one space dimension.  In two-dimensions, both 
schemes require the same amount of compute time; in greater than three space 
dimensions, the forward Euler scheme requires less compute time than the
Crank-Nicholson scheme.  In higher dimensions, the origin of the improved 
performance for the forward Euler scheme with optimal time step is the 
significantly smaller size of the computational grid required to achieve a 
desired error level.

\begin{table}[tbh]
\caption{\label{tab:comp_perf_vs_err_4th_order_parabolic} 
   Computational cost as a function of the global numerical error $E$
   for various FD schemes that numerically solve the 1D fourth-order
   parabolic equation.  For all FD schemes, the numerical solution is 
   assumed to be computed over a fixed time interval.  A less negative 
   power of $E$ for the memory and compute time means that a FD scheme is 
   more computationally efficient.
}
\begin{minipage}{\textwidth}
\begin{center} \footnotesize
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{|l|c|c|c|c|}
  \hline
  {\bf Numerical Scheme} & $\dx$ 
  & $\dt$\footnote{The time step for the forward Euler scheme is $O(\dx^4)$
   as a result of stability considerations.  The time step for the 
   Crank-Nicholson schemes with second- and fourth-order spatial 
   discretizations are $O(\dx)$ and $O(\dx^2)$, respectively, so that the 
   temporal error is subdominant to the spatial error.}
  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
    to be $O(1/\dx)$.} 
  & {\bf Compute Time}\footnote{The computation time is approximated as 
    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
    arise for the Crank-Nicholson method.}  \\
  \hline 
  Forward Euler with 
    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
    & \multirow{2}{*}{$O\left (E^{2} \right)$}
    & \multirow{2}{*}{$O\left( E^{-1/2} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-5/2} \right)$} \\
  \ \ 2nd-Order Bilaplacian & & & & \\
  \hline 
  Forward Euler with 
    & \multirow{2}{*}{$O\left( E^{1/4} \right)$}
    & \multirow{2}{*}{$O\left (E \right)$}
    & \multirow{2}{*}{$O\left( E^{-1/4} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-{5/4}} \right)$} \\
  \ \ 4th-Order Bilaplacian & & & & \\
  \hline 
  Forward Euler with 
    & \multirow{2}{*}{$O\left( E^{1/6} \right)$} 
    & \multirow{2}{*}{$O\left( E^{2/3} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-1/6} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-5/6} \right)$} \\ 
  \ \ 4th-Order Bilaplacian and OTS & & & & \\
  \hline 
  Crank-Nicholson with 
    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-1/2} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-1} \right)$} \\
  \ \ 2nd-Order Bilaplacian & & & & \\
  \hline 
  Crank-Nicholson with 
    & \multirow{2}{*}{$O\left( E^{1/4} \right)$} 
    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-1/4} \right)$} 
    & \multirow{2}{*}{$O\left( E^{-3/4} \right)$} \\
  \ \ 4th-Order Bilaplacian & & & & \\
  \hline
\end{tabular}
\end{center}
\end{minipage}
\end{table}

\section{\label{sec:applications_multidim}
         Application to PDEs in Multiple Space Dimensions}
A key distinction between the error for FD schemes in one space dimension
and multiple space dimensions is the presence of cross-terms in the 
discretization error.  In the context of a standard analysis for a finite 
difference scheme, it is usually safe to ignore the precise form of the 
cross-terms because they only appear as higher-order corrections in the 
error.  As a result, the choice of discretization for the spatial derivative 
operators is not critical to achieve a desired accuracy for the overal FD 
scheme.  In contrast, knowledge of the form of the cross-terms is critical 
when attempting to use optimal time step selection.  Without this knowledge, 
it is impossible to ensure cancellation of higher-order errors.

Recall that optimal time step selection requires that higher-order temporal 
errors be directly related to higher-order spatial errors through the use 
of the PDE.  This requirement can be satisfied in one of two ways:
\begin{itemize}
\item by carefully designing the spatial discretization so that 
      higher-order spatial errors are directly related to the spatial 
      derivative operators in the PDE; or
\item by explicitly adding temporal correction terms to cancel out cross-terms 
      that are not eliminated  by the original finite difference scheme.
\end{itemize}
While these two approaches may appear to be different, they are actually 
equivalent because the optimal time step is \emph{not independent} of the 
grid size.  Thus, the temporal correction terms can be viewed as part of
an expanded spatial discretization scheme and vice-versa.

In the remainder of this section, we apply optimal time step selection 
to the construction of finite difference schemes for two classical linear,
time dependent PDEs in two space dimensions, including an example that
demonstrates how optimal time step selection can be used to design simple
schemes for problems on irregular domains.  It is straightforward to 
generalize the approach presented in this section to more space dimensions 
and to construct methods for semilinear PDEs by incorporating ideas discussed 
in Sections~\ref{sec:optimal_time_step_selection} 
and~\ref{sec:applications_1d}.

\subsection{Linear Advection Equation}
For our first application of optimal time step selection in multiple
space dimensions, we consider the two-dimensional transport 
equation~\cite{evans_book}
\beq
  \frac{\partial u}{\partial t} + \vec{A} \cdot \nabla u = 0,
  \label{eq:advection_eqn_2d}
\eeq
where $\vec{A} = (A_x, A_y)$ is the advection velocity.  For convenience, we 
take $A_x, A_y > 0$.  This choice allows us to avoid unnecessary complexity 
associated with the dependence of the upwind direction on the sign of $A_x$ 
and $A_y$.  It is straightforward to generalize our discussion to $A_x$ and 
$A_y$ of arbitrary sign.

Recall that in one space dimension, using an optimal time step for
a simple forward Euler scheme that uses a first-order upwind discretization
of the spatial derivative leads to a FD scheme for the advection equation
with no local truncation error.  The same result can be achieved in 
two space dimensions.  However, because there are a total of three
numerical parameters (\ie two grid spacings and the time step), it is 
insufficient to only optimize the time step.  We need to optimally select
the time step \emph{and} the ratio of the grid spacings.

For the purposes of illustration, we consider a simple forward Euler scheme 
with first-order upwind discretizations of the advection terms:
\beq
  u^{n+1}_{i,j} = u^{n}_{i,j}
  - A_x \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} \right)
  - A_y \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} \right)
  \label{eq:advection_eqn_2d_FD_scheme}
\eeq
where superscripts and subscripts denote the time and space indices, 
respectively.  Using standard error analysis, it is straightforward to show
that this method is formally first-order in both space and time.

To compute the optimal time step and grid spacing ratio, we begin, as
usual, by analyzing the discretization error 
for~(\ref{eq:advection_eqn_2d_FD_scheme}).  As in our analysis of the FD
scheme for the one-dimensional advection equation, we retain all of the terms 
in the discretization error.  Using Taylor series to expand the true solution 
of the the PDE at time step $(n+1)$, we find that
\bea
  \tu^{n+1}_{i,j} = \tu^{n}_{i,j }
  + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\pt^k} \right|_{i,j}
  \label{eq:advection_eqn_2d_time_err}.
\eea
where $\tu$ denotes the true solution of the PDE. 
Similarly, Taylor series expansion of the upwind finite difference 
approximation to the spatial derivative yields
\bea
  \frac{\tu^{n}_{i,j} - \tu^{n}_{i-1,j}}{\dx} = 
  \left. \frac{\partial \tu}{\px} \right|_{i,j}
  - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\px^k} \right|_{i,j} 
  \\
  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dx} = 
  \left. \frac{\partial \tu}{\py} \right|_{i,j}
  - \frac{1}{\dy} \sum_{k=2}^\infty \frac{\left( -\dy \right)^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\py^k} \right|_{i,j} 
  \label{eq:advection_eqn_2d_space_err}
\eea
Next, we use the PDE~(\ref{eq:advection_eqn_2d}) to replace all of 
the time derivatives in~(\ref{eq:advection_eqn_2d_time_err}) with spatial
derivatives to obtain
\bea
  \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
  + \sum_{k=1}^\infty \frac{\left( -\dt \right)^k}{k!} 
       \left. \left( A_x \partial_x
                   + A_y \partial_y
              \right)^k \tu^{n} \right|_{i,j}.
  \label{eq:advection_eqn_2d_time_err_modified}
\eea
Combining~(\ref{eq:advection_eqn_2d_time_err_modified}) 
and~(\ref{eq:advection_eqn_2d_space_err}) with the finite difference 
scheme~(\ref{eq:advection_eqn_2d_FD_scheme}), we find that the local
truncation error is given by:
\bea
  \tau^{n}_{i,j} &=&
      \frac{A_x \dt}{2} \frac{\partial^2 \tu}{\px^2}
      \left( \dx - A_x \dt \right)
    + \frac{A_y \dt}{2} \frac{\partial^2 \tu}{\py^2}
      \left( \dy - A_y \dt \right)
    - A_x A_y \dt^2 \frac{\partial^2 \tu}{\px \py}
   \nonumber \\
   &+& \frac{A_x\dt}{\dx} \sum_{k=3}^\infty \frac{\left( -\dx \right)^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\px^k} \right|_{i,j} 
   + \frac{A_y\dt}{\dy} \sum_{k=3}^\infty \frac{\left( -\dy \right)^k}{k!} 
       \left. \frac{\partial^k \tu^n}{\py^k} \right|_{i,j} 
   \nonumber \\
   &-& \sum_{k=3}^\infty \frac{\left( -\dt \right)^k}{k!} 
       \left. \left( A_x \partial_x
                   + A_y \partial_y
              \right)^k \tu^{n} \right|_{i,j} 
  \label{eq:advection_eqn_2d_trunc_err}
\eea
From this expression, we see that choosing the ratio of the grid spacings to
be $\dx/\dy = A_x/A_y$ and setting the time step to be 
$\dt_{opt} = \dx/A_x = \dy/A_y$ eliminates two of the three leading-order 
terms in the discretization error.   The last leading-order term can be 
eliminated by adding the correction term
\bea
   A_x A_y \dt^2 \frac{\partial^2 \tu}{\px \py}.
  \label{eq:advection_eqn_2d_corr_term}
\eea
With these choices, we have already boosted the order of accuracy for the 
FD scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from 
first- to second-order.

With a careful choice of discretization for the correction term, we can
boost the order of accuracy even further.  Using the first-order upwind 
discretization for $\frac{\partial^2 \tu}{\px \py}$
\beq
   \frac{u_{i,j} - u_{i-1,j} - u_{i,j-1} + u_{i-1,j-1}}
        {\dx \dy} 
  \label{eq:advection_eqn_2d_upwind_corr_approx}
\eeq
the truncation error~(\ref{eq:advection_eqn_2d_trunc_err}) becomes
\bea
  \tau^{n}_{i,j} &=&
      \frac{A_x \dt}{2} \frac{\partial^2 \tu}{\px^2}
      \left( \dx - A_x \dt \right)
    + \frac{A_y \dt}{2} \frac{\partial^2 \tu}{\py^2}
      \left( \dy - A_y \dt \right)
   \nonumber \\
   &+& \frac{A_x\dt}{\dx \dy} ( \dy - A_y \dt )
       \sum_{k=3}^\infty \frac{\left( -\dx \right)^k}{k!} 
              \left. \frac{\partial^k \tu^n}{\px^k} \right|_{i,j} 
   \nonumber \\
   &+& \frac{A_y\dt}{\dx \dy} ( \dx - A_x \dt )
       \sum_{k=3}^\infty \frac{\left( -\dy \right)^k}{k!} 
              \left. \frac{\partial^k \tu^n}{\py^k} \right|_{i,j} 
   \nonumber \\
   &+& \frac{A_x A_y\dt^2}{\dx \dy} 
       \sum_{k=3}^\infty \sum_{l=0}^{k}
              (-1)^k {k \choose l}
              \frac{\dx^l \dy^{k-l}}{k!} 
              \left. \frac{\partial^k \tu^n}{\px^l \py^{k-l}} \right|_{i,j} 
   \nonumber \\
   &-& \sum_{k=3}^\infty \frac{\left( -\dt \right)^k}{k!} 
       \left. \left( A_x \partial_x
                   + A_y \partial_y
              \right)^k \tu^{n} \right|_{i,j} 
  \label{eq:advection_eqn_2d_trunc_err_mod},
\eea
which vanishes when the optimal time step and grid spacing ratio are used.
Thus, when the simple first-order upwind approximation to the correction 
term based on (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used in
conjunction with the optimal time step and grid spacing ratio, we obtain
a FD scheme that introduces no truncation error at each time step.
Figure~\ref{fig:advection_eqn_2d_soln} demonstrates the superior accuracy 
of this scheme over the first-order forward Euler 
scheme~(\ref{eq:advection_eqn_2d_FD_scheme}) without any modifications.
Notice how the sharp corners of the initial conditions are perfectly 
preserved when the optimal time step, optimal grid spacing ratio, and 
correction terms are used.  In contrast, the sharp corners are
completely smoothed out in the solution computed using the standard 
first-order upwind scheme.

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{\includegraphics{figures/adv_eqn_2d_FE_OTS_soln}} 
\scalebox{0.35}{\includegraphics{figures/adv_eqn_2d_FE_soln}} 
\caption{Comparison of numerical solutions for the 2D advection
equation~(\ref{eq:advection_eqn_2d}) using forward Euler time stepping 
with optimal time step and correction terms (left) and forward Euler time 
stepping with suboptimal time step (right).  The latter scheme uses a 
time step $\Delta t = \Delta x / 2 (|A_x|+|A_y|)$.  
Both figures show the numerical solution at $t = 3$ computed on the domain 
$-10 < x,y < 10$ using 100 grid points in each direction.
The advection velocity $(A_x, A_y)$ is set to $(-1, -2)$.  The initial 
conditions are taken to be a 2D square function centered at the
origin: $u(x,y,0) = 1$ if $|x| + |y| \le 2$ and 
$u(x,y,0) = 0$ if $|x| + |y| > 2$. 
For the time interval considered, the $u$ is identically zero 
on the boundary.
}
\label{fig:advection_eqn_2d_soln}
\end{center}
\end{figure}

To see that the temporal correction term can be considered as part of the 
spatial discretization scheme, we can write out the complete FD scheme 
including the correction term:
\bea
  u^{n+1}_{i,j} &=& u^{n}_{i,j}
  - A_x \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} \right)
  - A_y \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} \right)
  \nonumber \\
  & & + A_x A_y \dt^2 
        \left( \frac{u_{i,j} - u_{i-1,j} - u_{i,j-1} + u_{i-1,j-1}}
                    {\dx \dy} 
        \right).
  \label{eq:advection_eqn_2d_FD_scheme_complete}
\eea
Using the relationship between the optimal time step and the grid spacings, 
this scheme can be rewritten as
\bea
  u^{n+1}_{i,j} = u^{n}_{i,j}
  &-& \frac{A_x \dt}{2}
    \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} 
         + \frac{u^{n}_{i,j-1} - u^{n}_{i-1,j-1}}{\dx} 
    \right)
  \nonumber \\
  &-& \frac{A_y \dt}{2}
    \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} 
         + \frac{u^{n}_{i-1,j} - u^{n}_{i-1,j-1}}{\dy} 
    \right).
  \label{eq:advection_eqn_2d_FD_scheme_simplified}
\eea
Thus, we see that the FD scheme (\ref{eq:advection_eqn_2d_FD_scheme_complete})
is essentially forward Euler time integration with a first-order upwind 
discretization of the spatial derivatives that is computed using all four 
upwind neighbors.

In higher space dimensions, it becomes tedious to carry out the analysis 
to derive the infinite order accurate FD scheme for the advection equation
because many high-order correction terms must be retained.  However, it is 
straighforward to generalize 
(\ref{eq:advection_eqn_2d_FD_scheme_simplified}) to $d$ space dimensions by 
using a first-order discretization for the spatial derivatives that is an 
average over the $2^{d-1}$ first-order upwind discretizations in each 
coordinate direction.   As with the two-dimensional problem, the optimal 
grid spacings satisfy $\dx_k/A_{x_k} = C$ for $1 \le k \le d$, where $C$ is 
a constant equal to the optimal time step.

Like the unit CFL condition for the one-dimensional advection equation, the 
analogue of (\ref{eq:advection_eqn_2d_FD_scheme_simplified}) in any number
of space dimensions could also have been derived by examining how 
characteristic lines of the $d$-dimensional advection equation pass 
through grid points.  For the two-dimensional problem, this perspective 
becomes apparent when we fully simplify 
(\ref{eq:advection_eqn_2d_FD_scheme_simplified}) to obtain
$u^{n+1}_{i,j} = u^{n}_{i-1,j-1}$.

\subsection{Diffusion Equation}
As for PDEs in one space dimension, achieving infinite order accuracy 
through optimal parameter selection is unique to the advection equation.
In general, optimal parameter selection boosts the accuracy by only a few 
orders.  In this section, we use the 2D diffusion equation to illustrate the 
typical accuracy boost gained through optimal time step selection.  We also 
demonstrate the utility of optimal time step selection for designing simple 
FD schemes that yield high-order accuracy for problems on irregular domains.

In two space dimensions, the diffusion equation may be written as
\beq
  \frac{\partial u}{\pt} = D \nabla^2 u,
  \label{eq:diffusion_eqn_2d}
\eeq
where $\nabla^2$ is the Laplacian operator and $D$ is the diffusion constant. 
While there are many finite difference schemes that are used to solve the 
2D diffusion equation, the most common ones employ a five-point, second-order 
central difference stencil for the Laplacian~\cite{iserles_book}:
\beq
  L^{5pt} u = \frac{u_{i,j+1} + u_{i,j-1}
                   +u_{i+1,j} + u_{i-1,j} - 4u_{i,j}}{\dx^2},
  \label{eq:laplacian_2d_5pt_stencil}
\eeq
which assumes that the grid spacing in the $x$ and $y$ directions are
equal.  Common choices for the time integration method include forward Euler
and the Crank-Nicholson method.  Each scheme has its advantages and
disadvantages.  The forward Euler scheme is very easy to implement but is 
only first-order accurate and suffers from stability constraints on the time 
step; the Crank-Nicholson scheme is second-order accurate and unconditionally 
stable but requires the solution of a linear system of equations at each time 
step.  The forward Euler scheme using the five-point stencil is given by
\bea
  u^{n+1}_j &=& u^{n}_j 
  + D \dt 
    \left( \frac{u^{n}_{i,j+1} + u^{n}_{i,j-1}
         +u^{n}_{i+1,j} + u^{n}_{i-1,j} - 4u^{n}_{i,j}}{\dx^2}
    \right),
  \label{eq:diffusion_eqn_2d_5pt_scheme}
\eea
Standard error analysis shows that this scheme is formally first-order in 
time and second-order in space.

While it is certainly possible to directly apply optimal time step selection 
to (\ref{eq:diffusion_eqn_2d_5pt_scheme}), we would be forced to use additional
correction terms to deal with the fact that the leading-order error for 
(\ref{eq:laplacian_2d_5pt_stencil}), 
$\dx^2 \left(u_{xxxx} + u_{yyyy}\right)/12$, cannot be directly related 
to the spatial derivative operator in the diffusion equation.  
The analysis is more straightforward straightforward if we construct our
finite difference scheme using the nine-point stencil for the 
Laplacian~\cite{iserles_book,patra_2005,??}, which possesses a leading-order 
error directly related to the spatial derivative operator in the PDE 
(\ref{eq:diffusion_eqn_2d}):
\bea
  L^{9pt} u = \frac{1}{6 \dx^2} \left( 
      \begin{array}{l}
         \ \ \ u_{i+1,j+1} + u_{i+1,j-1}
            + u_{i-1,j+1} + u_{i-1,j-1} \\
         +\ 4 u_{i+1,j} + 4 u_{i-1,j}
           + 4 u_{i,j+1} + 4 u_{i,j-1}
           -20 u_{i,j} 
      \end{array}
    \right). 
  \label{eq:laplacian_2d_9pt_stencil}
\eea
Observe that the nine-point stencil is simply the standard five-point 
stencil plus $\dx^2/6$ times a second-order, central difference 
approximation to $u_{xxyy}$.  Therefore, we can obtain the leading-order 
error in the nine-point stencil by combining the leading-order error for the 
five-point stencil with the extra contribution of 
$\dx^2 u_{xxyy}/6$
\bea
L^{9pt} u &=& \nabla^2 u 
              + \frac{\dx^2}{12} \left(u_{xxxx} + 2 u_{xxyy} + u_{yyyy}\right)
              + O(\dx^4)
          \nonumber \\
          &=& \nabla^2 u + \frac{\dx^2}{12} \nabla^4 u + O(\dx^4).
  \label{eq:laplacian_2d_9pt_stencil_error}
\eea
Like, (\ref{eq:diffusion_eqn_2d_5pt_scheme}), the forward Euler scheme that 
uses the nine-point discretization of the Laplacian 
\beq
  u^{n+1}_{i,j} = u^{n}_{i,j}
  + D \dt L^{9pt} u
  \label{eq:diffusion_eqn_2d_9pt_scheme}
\eeq
is $O(\dx^2)$ accurate overall because it has a stability constraint 
$\dt \le ??\dx^2/2D = O(\dx^2)$.
It is interesting to note that use of a nine-point stencil for the Laplacian 
naturally arises when directly applying optimal time step selection to the
five-point scheme through reinterpretion of the temporal correction term
as a modification to the spatial discretization.

To apply optimal time step selection to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we 
proceed in the usual way by first analyzing the leading-order terms of the 
discretization error for the finite difference scheme.  Taylor series 
expansions show that the true solution satisfies the following equations:
\bea
  \tu^{n+1}_{i,j} &=& \tu^{n}_{i,j}
  + \dt \frac{\partial \tu^n}{\pt} 
  + \frac{\dt^2}{2} \frac{\partial^2 \tu^n}{\pt^2} + O \left( \dt^3 \right)
  \label{eq:diffusion_eqn_2d_time_err} 
  \\
  L^{9pt} \tu &=&
  \left. \nabla^2 \tu \right|_{i,j}
  + \frac{\dx^2}{12} \left. \nabla^4 \tu \right|_{i,j}
  + O(\dx^4)
  \label{eq:diffusion_eqn_2d_space_err}
\eea
Using the PDE~(\ref{eq:diffusion_eqn_2d}) to replace time derivatives
with space derivatives, (\ref{eq:diffusion_eqn_2d_time_err}) becomes
\bea
  \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
  + D \dt \nabla^2 \tu^{n}_{i,j}
  + \frac{(D \dt)^2}{2} \nabla^4 \tu^{n}_{i,j}
  + O \left( \dt^3 \right)
  \label{eq:diffusion_eqn_2d_time_err_modified}
\eea
Combining these results, we find that the local truncation error for the
finite difference scheme (\ref{eq:diffusion_eqn_2d_9pt_scheme}) is given
by 
\bea
  \tau^{n}_j = 
  \nabla^4 \tu
      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
      + O(\dt \dx^4) + O(\dt^3).
  \label{eq:diffusion_eqn_2d_trunc_err}
\eea

As for the 1D diffusion equation, we see that the optimal time step is given 
by $\dto = \dx^2/6D$ and that this choice of time step yields a scheme is 
that is $O(\dx^4) + O(\dt^2) = O(\dx^4)$ overall.  Note that there is no need
to consider an implicit scheme because $\dto$ is stable for the forward Euler 
time integration scheme.

The improvement in accuracy of the forward Euler scheme when the optimal time 
step is used is clear in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
The numerical errors of the following three finite difference schemes for 
the diffusion equation show the expected orders of accuracy: forward Euler 
with optimal time step, forward Euler with suboptimal time step, and 
Crank-Nicholson.  As with our other examples, the forward Euler scheme 
with an optimal time step is much more efficient that the two other schemes.  
Even with as few as 25 grid points in each space direction, the error in 
solution computed by the forward Euler scheme with an optimal time step is 
orders of magnitude smaller.

\begin{figure}[htb]
\begin{center}
\scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_error_vs_N}} 
\ \ 
\scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_comp_time}} 
\caption{Error and performance analysis for three finite difference schemes 
for the 2D diffusion equation: forward Euler with optimal time step (circles), 
forward Euler with suboptimal time step (squares), and Crank-Nicholson 
(diamonds).  Using each method, we solve the 2D diffusion equation with 
$D = 1$ on the domain $0 < x < 1$ subject to the boundary conditions 
and the initial condition
$??$
The exact solution for this problem is
$u(x,t) = ??$
The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
method.  Note that the errors for the forward Euler with suboptimal time 
step and Crank-Nicholson schemes lie almost directly on top of each other
at the resolution of the figures.
}
\label{fig:diffusion_eqn_2d_src_analysis}
\end{center}
\end{figure}

Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also compares the computational
performance of the three schemes.  For all three schemes, notice that the 
scaling of the computational time with error is in good agreement with the 
theoretical expectations (see Table~\ref{tab:comp_perf_vs_dim}).  Again, the 
forward Euler with optimal time step demonstrates much greater efficiency 
than the other schemes.  In addition to having better scaling properties, 
the line for the forward Euler scheme with optimal time step appears to be 
significantly shifted to the left (\ie smaller errors).

\subsubsection{Irregular Domains}
All of the previous examples share an important feature: they are all
defined on simple, rectangular domains.  Thus, any of several methods 
could have been used to achieve similar or even higher accuracy solutions.
For example, a fourth-order accurate solution to the diffusion equation
could be obtained by using a pseudospectral spatial
discretization~\cite{trefethen_spectral_book, boyd_book} of the Laplacian
and a fourth-order time integration scheme (\eg Runge-Kutta).
When the domain is irregularly shaped, high-order methods are harder to design 
and implement.  For instance, spectral and pseudospectral methods are no 
longer appropriate because the domain is not a simple cartesian product of 
intervals\footnote{Clever phase field approaches have been used successfully 
to make these methods applicable to problems with irregular 
boundaries~\cite{liu??}.}.

As discussed at the end of Section~\ref{sec:ots_higher_spatial_dims}, we can 
apply optimal time step selection to construct simple finite difference schemes 
that yield high-order accurate solutions for problems on irregular domains.  
In this section, we extend the OTS finite difference scheme for the diffusion
equation on regular domains to irregular domains by describing a method for 
computing the values of the solution in ghost cells that ensures that 
imposing boundary conditions does not decrease the accuracy of the solution.

One important issue that arises when devising numerical schemes for problems 
on irregular domains is how the domain and boundary should be represented.
The various ways for representing irregular boundaries can be divided 
into two categories: explicit and implicit.  A full comparison of the many 
representations of irregular boundaries is beyond the scope of this paper, 
so we refer interested readers to~\cite{??} for more detailed discussion 
about the advantages and disadvantages of several popular methods.  For our 
purposes, we choose to implicitly represent the irregular boundary as the 
zero level set of a level set function $\phi$ which satisfies:
\bea
  \begin{array}{ll}
  \phi < 0 & \mathrm{for} \ x \in \Omega \\
  \phi > 0 & \mathrm{for} \ x \notin \Omega \\
  \phi = 0 & \mathrm{for} \ x \in \partial \Omega 
  \end{array}
\eea
where $\Omega$ and $\partial \Omega$ are the interior and boundary of the 
irregular domain.

Devising a strategy for accurately imposing boundary conditions is perhaps 
the most critical aspect of designing finite difference schemes for problems 
on irregular domains.  To avoid the need for special stencils at grid points
near the irregular boundary, we adopt the `ghost cell' or `ghost value'
approach to imposing boundary conditions~\cite{fedkiw_1999, osher_fedkiw_book, 
gibou_2005}.  The basic 
idea is to set the value of grid points that lie outside of the domain 
\emph{and} are required by the spatial discretization operator for grid 
points inside the domain by extrapolating from interior grid points using
boundary condition information.  Figure~\ref{fig:ghost_cells} shows examples 
of ghost cells that lie in the vicinity of a irregular boundary.  Notice that
there are two types of ghost cells:  edge and corner.  Edge and corner
ghost cells are distinguished by the relative position of the nearest 
interior grid point.  An edge ghost cell is linked to its nearest interior
neighbor by an edge of the computational grid whereas a corner ghost cell
and its nearest interior neighbor lie on the opposite corners of a grid cell.

\begin{figure}[htb]
\begin{center}
\scalebox{0.25}{\includegraphics{figures/ghostcell_edge}} 
\ \ \ \ \ \ \ \ \ \ \ \ \ 
\scalebox{0.25}{\includegraphics{figures/ghostcell_corner}}
\caption{Illustrations of edge (left) and corner (right) ghostcells.
The edge ghostcell, $u_E$, is filled by using cubic extrapolation of 
the value on the boundary, $u_B$, and values at three interior grid points.  
The corner ghostcells, $u_C$, is filled by using quadratic extrapolation 
of the values at interior grid points and neighboring edge ghost cells.
}
\label{fig:ghost_cells}
\end{center}
\end{figure}

The value in each edge ghost cell is set by using 1D, cubic extrapolation of 
the values of $u$ across the interface.  A cubic Lagrange extrapolant is 
constructed for each edge ghost cell using the point on the domain boundary 
where the zero level set of $\phi$ intersects the edge connecting the ghost 
cell to its nearest interior neighbor and three interior grid points along 
the line parallel to the same edge (see Figure~\ref{fig:ghost_cells}).  
The location of the point $u_B$ can be calculated by inverting the linear 
interpolant of $\phi$ that passes through the ghost cell and its nearest
interior neighbor.
Because the quality of the extrapolant deteriorates if the boundary point is 
too close to any of the interior grid 
points used to construct the Lagrange extrapolant, we follow 
\cite{gibou_2005} and choose the interior grid points so that the nearest
interior grid point is sufficiently far from the boundary point.  
Gibou~\etal suggest shifting the interpolation points by one grid point 
when the distance between the boundary point and the nearest interior grid 
point is less than $\dx^2$.  For the numerical scheme presented in this paper, 
thie threshold was too low.  We found that better results were obtained if 
we used a $O(\dx)$ threshold.

The values in corner ghost cells are set using 2D quadratic extrapolation
of the values of $u$ across the domain boundary \emph{and} the neighboring
edge ghost cells (see Figure~\ref{fig:ghost_cells}).  Notice that unlike
edge ghost cells, corner ghost cells only need to be 3rd-order accurate 
in order to obtain a 4th-order accurate solution.  The reason for this is ??.
*** KTC - FILL ME IN ***

To derive quadratic extrapolant at each corner ghost cell, we use a 
second-order Taylor series expansion around the nearest interior grid point,
$x_{0,0}$, and use finite differences to approximate all required partial 
derivatives.  For example, using central differences centered at $x_{0,0}$ for 
all partial derivatives, the quadratic extrapolant for the corner ghost cell 
shown in Figure~\ref{fig:ghost_cells} is
\bea
  u_C &=&  -u_{0,0} + \frac{1}{2} u_{-1,-1} 
      \nonumber \\
      &+& \frac{5}{4} u_{1,0} + \frac{5}{4} u_{0,1}
      - \frac{1}{4} u_{1,-1} - \frac{1}{4} u_{-1,1}
      - \frac{1}{4} u_{0,-1} - \frac{1}{4} u_{-1,0}
\eea
where $u_C$ and $u_{i,j}$ are as indicated in the figure.  Note that the 
stencil for the quadratic interpolant depends on the choice of finite 
difference approximation for the partial derivatives.  For the results 
presented in this paper, we use the above stencil (and its analogues for 
other positions of the ghost cell relative to the domain) to define the
quadratic extrapolant.  However, we obtained the same qualitative results 
when stencils based on other choices of finite difference approximations were
used.

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_2d_starfish_domain_soln}} 
\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_2d_starfish_domain_error}} 
\caption{Numerical solutions (left) and dominant error (right) for the 2D 
diffusion equation~(\ref{eq:??}) a starfish-shaped domain.  The numerical 
solution is computed at time $t = 0.1$ using forward Euler time stepping 
with an optimal time step and correction terms.  In the error plot, the dark 
regions represent points where the error in the solution is larger than 
$10$\% of the $L^\infty$ error of the solution.  Notice that the largest 
errors are concentrated near the boundaries of the mathematical domain.  To 
generate these results, the computational domain is taken to be 
$-1 < x,y < 1$ with 400 grid points in each direction.  The diffusion 
constant $D$ is set equal to $1/4$.  The source term is given by
$f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
            \exp\left(-5 D \pi^2 t\right)
          - 24 D \pi^2 \sin(5 \pi x) \sin(\pi y) 
            \exp \left(-2 D \pi^2 t \right).
$
The initial conditions are taken to be 
$u(x,y,0) = 1 + y/3 + x/2 + xy/4 
          + \sin(2 \pi x) \sin(3 \pi y) 
          - \sin(5 \pi x) \sin(\pi y)
$
and boundary conditions are imposed by using cubic and quadratic 
extrapolation of interior and boundary values to to fill edge and corner 
ghost cells, respectively.  Boundary values are obtained from the 
analytical solution
$u(x,y,t) = 1 + y/3 + x/2 + xy/4 
              + \sin(2 \pi x) \sin(3 \pi y) \exp(-5 D \pi^2 t) 
              - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)$.
The boundaries of the domains are represented by the zero-level 
set of the function
$\sqrt{x^2 + y^2} - \frac{3}{5} \left(1 + \frac{1}{2}\sin(5 \theta) \right)$, 
where $\theta = \mathtt{atan}\left( y/x \right)$.
}
\label{fig:diffusion_eqn_2d_starfish_domain}
\end{center}
\end{figure}

\begin{figure}[htb]
\begin{center}
\scalebox{0.35}{
  \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
\caption{Error analysis for four finite difference schemes 
for the 2D diffusion equation: forward Euler with optimal time step (circles), 
forward Euler with suboptimal time step (squares), 
(diamonds).  Using each method, we solve the 2D diffusion equation with 
$D = 1$ on the domain $0 < x < 1$ subject to the boundary conditions 
and the initial condition
$??$
The exact solution for this problem is
$u(x,t) = ??$
The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
method.  Note that the errors for the forward Euler with suboptimal time 
step and Crank-Nicholson schemes lie almost directly on top of each other
at the resolution of the figures.
}
\label{fig:diffusion_eqn_2d_starfish_error}
\end{center}
\end{figure}

\subsubsection{Generalization to Higher Dimensional Problems}
** COMPLEXITY OF EXTRAPOLATION DEPENDS ON CHOICE OF DISCRETIZATION FOR LAPLACIAN **

** PATRA AND KARTTUNEN GIVE ISOTROPIC STENCILS FOR 3D **

\subsubsection{Weakly Anisotropic Diffusivity Tensor}
can use unequal grid spacing when the diffusivity tensor is diagonlizable 
(i.e. normal) and the eigenvalues are unequal.  first transform to coordinate
system of principle axes then rescale problem.

\section{\label{sec:summary} Summary} 
While somewhat tedious, analyzing higher-order errors can suggest simple 
modifications to numerical methods that greatly improve their accuracy.  
Symbolic mathematics tools can be helpful to make this easier.

Optimal time step selection is not applicable to all PDEs.  ...

% ACKNOWLEDGMENTS
\section*{Acknowledgments}
The author gratefully acknowledges the support of the Institute for 
High-Performance Computing (IHPC) in Singapore and Vitamin D, Inc.
The author would like to thanks P. Fok and ?? for enlightening discussions 
and suggestions on the manuscript.

\bibliography{optimal_time_steps_for_pdes}

\end{document}

### 2008-05-10 papers/OTSPDE/optimal_time_steps_for_pdes.tex
OTSPDE:  first completed draft of paper

@@
 % Manuscript started: 2007/09/01
-% First draft: 2008/02/??, Vitamin D, Inc.
+% First draft: 2008/05/??, Vitamin D, Inc.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \author{
-Kevin T. Chu\footnotemark[2]
+Kevin T. Chu\footnotemark[2] \footnotemark[3]
 }
@@
 
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\footnotetext[3]{Institute of High Performance Computing, A*STAR, Singapore, Singapore}
+
 \renewcommand{\thefootnote}{\arabic{footnote}}
@@
 \begin{abstract}
-In this article, we discuss a novel technique based on optimal time step 
-selection for boosting low-order finite difference schemes
-into higher-order numerical methods for time dependent partial differential
-equations (PDEs).  We demonstrate the utility of this technique on several 
-classical hyperbolic and parabolic problems in one and more space dimensions 
-and explain the observed orders of convergence through straightforward numerical
-analysis arguments.
-** HIGH-ORDER PDES AND IRREGULAR DOMAINS ***
+In this article, we present a novel technique for obtaining high-order
+accurate numerical solutions to time dependent partial differential 
+equations using only formally low-order finite difference stencils and
+time integration schemes.  By using an optimally chosen the time step for a 
+finite difference scheme (and possibly adding a few correction terms), we 
+are able to boost the order of accuracy of low-order finite difference 
+schemes.  Through simple numerical analysis arguments, we explain the origin 
+of the boost in accuracy and calculate the computational cost of the
+high-order methods obtained via optimal time step (OTS) selection.  We 
+demonstrate the utility of OTS selection on a wide array of classical PDEs
+in one and more space dimensions.   We find that OTS selection is applicable 
+to linear and semilinear PDEs and that it is easily extended to problems 
+on irregular domains. 
 \end{abstract}
@@
 \begin{keywords}
-optimal time step, finite difference methods, high-order accuracy, 
-time dependent PDEs, initial-value problems
+optimal time step, finite difference methods, high-order accurate numerical 
+methods, time dependent PDEs
 \end{keywords}
@@
 numerical PDEs continues to be focused on the development of high-order 
-numerical schemes~\cite{gibour_2005,??}.  
+numerical schemes~\cite{bruger_2005, gibou_2005, ito_2005, shukla_2005, 
+shukla_2007}.  
 Typically, high-order accuracy is achieved by constructing
@@
 constraint is certainly sufficient for high-order accuracy, it is by no means 
-necessary.  This standard approach for developing numerical methods for PDEs 
-has its roots in the classical numerical analysis of PDEs, which tends to 
+necessary.  This conventional approach for developing numerical methods for 
+PDEs has its roots in the classical numerical analysis of PDEs, which tends to 
 individually analyze the discretization of the terms in the PDE.  Within this 
@@
 
-Perhaps the greatest weakness of the standard method for constructing 
+Perhaps the greatest weakness of the conventional method for constructing 
 high-order numerical methods is the use of ``big-O" notation, which hides all 
@@
 to calculate the optimal time step size and order of accuracy.  
-Finally, in Section~\ref{sec:summary}, we provide some concluding remarks on 
-the applicability and limitations of optimal time step selection for the 
-numerical solution of general time dependent PDEs.
+Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.
 
@@
 constructed to solve evolution equations of the form 
-$\frac{\partial u}{\partial t} = \ldots$, where the right hand side only 
-involves functions of the independent variable and its spatial derivatives.  
+\beq
+  \frac{\partial u}{\partial t} = 
+    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
+\eeq
+where the right hand side is an arbitrary function of the solution $u$, its 
+spatial derivatives, and the independent variable $x$.
 Because this class of time dependent PDEs is so prevalent and important in 
@@
             Optimal Time Step for a Model PDE}
-Let us begin by considering a finite difference approximation to a 
-time dependent PDE of the form
+Let us begin by considering a finite difference approximation for the
+following time dependent PDE in one spatial dimension: 
 \beq
@@
 (both spatial and temporal) for the FD approximation, explicitly keeping track 
-of the leading-order terms.  In contrast to standard error analysis, we do 
-not sweep all of the errors under ``big-O'' notation.  Rather, we take 
+of the leading-order terms.  In contrast to conventional error analysis, we 
+do not sweep all of the errors under ``big-O'' notation.  Rather, we take 
 advantage of the fact that derivation of the leading-order terms in the 
@@
 derivative of the solution, we can attempt to use the PDE to relate the 
-leading-order terms to each other.  If we are fortunate (which in many cases 
-we are), this careful analysis yields a direct relationship between the 
-leading-order spatial and temporal errors that allows them to be combined as 
-a single term of the form:
+leading-order terms to each other.  If we are fortunate (which we are
+for many finite difference schemes, this careful analysis yields a direct 
+relationship between the leading-order spatial and temporal errors that allows 
+them to be combined as a single term of the form:
 \beq
@@
 \eeq
-where $P$ is a polynomial in the numerical parameters, $L$ is a differential 
-operator, $u$ is the solution to the PDE, and we have assumed that the 
-grid spacing is equal in all spatial directions\footnote{In general, the grid 
-spacing need not be equal in all spatial directions.  For some problems, it is 
-only possible to calculate an optimal time step if the grid spacing is
-unequal between the different spatial directions.}.  By selecting the 
-numerical parameters so that $P = 0$, we can completely eliminate the 
-leading-order term in the discretization error.  Because time dependent PDEs 
-are often solved by stepping in time, it is natural to choose the time step 
-to be a function of the grid spacing.  The time step selected in this way 
-is the \emph{optimal time step} for the finite difference scheme.
+where $L$ is a differential operator and $P$ is a polynomial in the numerical 
+parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
+$P = 0$, we can completely eliminate the leading-order term in the 
+discretization error.  Because time dependent PDEs are often solved by 
+stepping in time, it is natural to choose the time step to be a function of 
+the grid spacing.  The time step selected in this way is the
+\emph{optimal time step} for the finite difference scheme.  Using the optimal
+time step for time stepping yields a numerical solution with the maximal order 
+of accuracy for a given FD scheme.
 
@@
 \eeq
-where $\gamma = (\alpha/\beta)^{1/s}$.  Note that the optimal time step 
-has the same functional dependence on the grid spacing as a time step 
-that is satisfies a stability constraint.
-
+where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
+has a similar functional dependence on the grid spacing as a time step that 
+satisfies a stability constraint.
+
+Thus far, we have completely ignored the question of whether the FD scheme is 
+stable when the time step is set to $\dto$.  Not surprisingly, OTS selection 
+loses its value, when the optimal time step is greater than the smallest 
+stable time step for an explicit scheme.  Fortunately, we can usually switch 
+to an implicit (or semi-implicit) time integration scheme with the same order 
+of accuracy for which the optimal time step is stable.  Note that the 
+depending on how the original FD scheme is modified, a new optimal time step 
+may need to be calculated.
+
+\subsubsection*{\label{sec:error_analysis} 
+            Error Analysis}
 The form of the local error in 
@@
 accumulated during each time step due to spatial discretization error is 
-$O(\dt \dx^r)$ regardless of how the system of ODEs is solved 
-numerically.  Because the method of lines is so frequently used to construct FD 
-approximations for time dependent PDEs, we will focus solely on FD schemes 
-of this type for the remainder of the paper in order to keep the discussion 
+$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
+is solved numerically.  Because the method of lines is so frequently used to 
+construct FD approximations for time dependent PDEs, we will focus solely on 
+FD schemes of this type for the remainder of the paper to keep the discussion 
 simple.  It is straightforward to extend the results we obtain to more general 
@@
 
-So far in our discussion, we have completely ignored the question of whether 
-the FD scheme is stable when the time step is set to $\dto$.  When
-the optimal time step is greater than the smallest stable time step, the 
-method breaks down.  Fortunately, when stability issues arise for explicit 
-time integration schemes, we can usually switch to an implicit (or 
-semi-implicit) time integration scheme with the same order of accuracy for 
-which the optimal time step is stable.  Note that the depending on how
-the original FD scheme is modified, a new optimal time step may need to be 
-calculated.
-
-\subsubsection*{PDEs in Higher Spatial Dimensions 
-                \label{sec:ots_higher_spatial_dims}}
-It is important to emphasize that optimal time selection is applicable to 
-PDEs in any number of spatial dimensions.  While problems in higher dimensions 
-generally require a more careful choice of the finite difference stencil used 
-to approximate spatial derivatives in order to be able to combine 
-leading-order terms in the discretization error, there are no fundamental 
-barriers that prevent the use of optimal time step selection.
-
-Domain geometry of the domain, which primarily impacts the way boundary 
-conditions are imposed, is an important feature of problems in more than one 
-spatial dimension.  While it is usually straightforward to handle problems on 
-regular domains, problems on irregular domains can pose a greater challenge 
-because the location of the domain boundary may not be explicitly specified
-and grid points near the boundary require special consideration.  Fortunately, 
-because optimal time step selection is based on simple, localized finite 
-difference stencils, irregular domains do not create any serious difficulties.  
-Assuming that a `ghost cell' or `ghost value' 
-approach~\cite{fedkiw_1999, osher_fedkiw_book, gibou_2005} 
-is used to impose boundary conditions, the construction and analysis of finite 
-difference schemes and calculation of optimal time steps is identical for 
-problems on both regular and irregular domains.  For problems on 
-irregular domains, the key issue is how values at ghost cell points should 
-be calculated.  Once this issue has been addressed, devising high-order 
-accurate methods for PDEs on irregular domains is simply a matter of applying 
-optimal time step selection to the finite difference scheme used for problems 
-on regular domains.
-
-\subsubsection*{\label{sec:error_analysis} 
-            Error Analysis}
 To analyze the accuracy of the FD scheme for~(\ref{eq:model_PDE}) when the 
-optimal time step is used (for a given grid spacing), we examine the remaining 
-terms of the discretization error in the standard way (\ie ``big-O'' notation).
-Elimination of the leading term in the discretization error leaves a local 
-error of the form $O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ 
-and $(q+1) > (s+1)$ are the orders of the spatial and temporal discretization 
-errors, respectively, after the leading-order terms in the error have been 
-removed.  Like the orders of the leading-order errors in 
-(\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
+optimal time step is used (for a given grid spacing), we examine the 
+higher-order terms in the discretization error following the conventional 
+approach (\ie ``big-O'' notation).  Elimination of the leading term in the 
+discretization error leaves a local error of the form 
+$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
+of the spatial and temporal discretization errors, respectively, after the 
+leading-order terms in the error have been removed.  As for the leading-order 
+error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
 error is a direct consequence of the method of lines approach to discretizing 
@@
 
-When interpreting~(\ref{eq:global_error_ots}), it is important to recognize
-that $\dx$ and $\dt$ are \emph{not} independent because the time 
-step has been set to $\dto$; there is really only one parameter, 
-$\dx$, that controls the numerical accuracy of the FD scheme.  This is an 
-important point that we will address further in the next section.
+It is important to emphasize that $\dx$ and $\dt$ are \emph{not} independent 
+in~(\ref{eq:global_error_ots}) because the time step must be set to $\dto$ 
+in order to derive the error estimate.  In other words, there is really only 
+one parameter, $\dx$, that controls the numerical accuracy of the FD scheme.  
+This important point will be discuss in further detail in the next section.
 
@@
 
-Optimal time step selection fits into this general framework for developing
-numerical methods for time dependent PDEs.  Because the optimal time step 
-is calculated explicitly as a function of the grid spacing, the technique
-of optimal time step selection places an even more restrictive constraint
-on $\dt$ than stability considerations.  Thus, like numerical
+Optimal time step selection fits into this general framework for controlling
+the error of numerical methods for time dependent PDEs.  Because the optimal 
+time step is calculated explicitly as a function of the grid spacing, the 
+technique of optimal time step selection places an even more restrictive 
+constraint on $\dt$ than stability considerations.  Thus, like numerical
 methods with stability constraints, the accuracy of an FD scheme using 
@@
                 Computational Performance}
-In general, $r/s > 1$ and yields an optimal time step that is on the order
-of the time step allowed by the stability constraint for an explicit method.
-As a result, using the optimal time step to advance the FD scheme in time
-may seem to be overly restrictive and lead to poor computational performance.
-However, these apparent drawbacks are more than compensated by the boost in 
-the order of accuracy.  For example, when solving the 1D diffusion equation 
-using a forward Euler time integration scheme with a second-order central 
-difference stencil for the Laplacian, we obtain a fourth-order scheme 
-when $\dt$ is set equal to the optimal time step.  As
+For many problems, $r/s > 1$ and yields an optimal time step that is on the 
+order of the time step allowed by the stability constraint for an explicit 
+method.  As a result, using the optimal time step to advance the FD scheme in 
+time may seem to be overly restrictive and lead to poor computational 
+performance.  However, these apparent drawbacks are more than compensated by 
+the boost in the order of accuracy.  For example, when solving the 1D 
+diffusion equation using a forward Euler time integration scheme with a 
+second-order central difference stencil for the Laplacian, we obtain a 
+fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
 Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
@@
 \caption{\label{tab:comp_perf_vs_err} 
-   Computational cost as a function of the global numerical error $E$
-   for various FD schemes that numerically solve the 1D diffusion equation.  
+   Computational cost of various FD schemes for numerically solving the 1D 
+   diffusion equation as a function of the global numerical error $e$.
    For all FD schemes, the standard second-order central difference 
@@
    solution is assumed to be computed over a fixed time interval.  
-A less negative power of $E$ 
-   for the memory and compute time means that a FD scheme is more 
-   computationally efficient.
 }
@@
   \hline 
-  Forward Euler    & $O\left( E^{1/2} \right)$ 
-                   & $O\left (E \right)$ 
-                   & $O\left( E^{-1/2} \right)$ 
-                   & $O\left( E^{-3/2} \right)$ \\
-  Backward Euler   & $O\left( E^{1/2} \right)$ 
-                   & $O\left( E \right)$ 
-                   & $O\left( E^{-1/2} \right)$ 
-                   & $O\left( E^{-3/2} \right)$ \\
-  Crank-Nicholson  & $O\left( E^{1/2} \right)$ 
-                   & $O\left( E^{1/2} \right)$ 
-                   & $O\left( E^{-1/2} \right)$ 
-                   & $O\left( E^{-1} \right)$ \\
-  Forward Euler with OTS  & $O\left( E^{1/4} \right)$ 
-                   & $O\left( E^{1/2} \right)$ 
-                   & $O\left( E^{-1/4} \right)$ 
-                   & $O\left( E^{-3/4} \right)$ \\ 
+  Forward Euler    & $O\left( e^{1/2} \right)$ 
+                   & $O\left( e \right)$ 
+                   & $O\left( e^{-1/2} \right)$ 
+                   & $O\left( e^{-3/2} \right)$ \\
+  Backward Euler   & $O\left( e^{1/2} \right)$ 
+                   & $O\left( e \right)$ 
+                   & $O\left( e^{-1/2} \right)$ 
+                   & $O\left( e^{-3/2} \right)$ \\
+  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
+                   & $O\left( e^{1/2} \right)$ 
+                   & $O\left( e^{-1/2} \right)$ 
+                   & $O\left( e^{-1} \right)$ \\
+  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
+                   & $O\left( e^{1/2} \right)$ 
+                   & $O\left( e^{-1/4} \right)$ 
+                   & $O\left( e^{-3/4} \right)$ \\ 
   \hline
@@
 \caption{\label{tab:comp_perf_vs_dim}
-   Comparison of the computational cost as a function of the global numerical 
-   error $E$ for the forward Euler scheme with optimal time step and the 
-   Crank-Nicholson method.  Both FD schemes are used to solve the diffusion 
-   equation and use a second-order, isotropic stencil is used to discretize 
+   Computational cost of the forward Euler scheme with optimal time step and 
+   the Crank-Nicholson method for solving the diffusion equation as a function 
+   of the global numerical error $e$.
+   For both FD schemes, second-order, isotropic stencils are used to discretize 
    the Laplacian.  For a problem in $d$ spatial dimensions, the memory 
@@
    invert the matrices that arise for the Crank-Nicholson method.
-   A less negative power of $E$ for the memory and compute time
-   means that a FD scheme is more computationally efficient.
 }
@@
   \hline 
-  $1$ & $O\left( E^{-1/4} \right)$ 
-      & $O\left( E^{-3/4} \right)$ 
-      & $O\left( E^{-1/2} \right)$ 
-      & $O\left( E^{-1} \right)$ \\ 
-  $2$ & $O\left( E^{-1/2} \right)$ 
-      & $O\left( E^{-1} \right)$ 
-      & $O\left( E^{-1} \right)$ 
-      & $O\left( E^{-3/2} \right)$ \\ 
-  $3$ & $O\left( E^{-3/4} \right)$ 
-      & $O\left( E^{-5/4} \right)$ 
-      & $O\left( E^{-3/2} \right)$ 
-      & $O\left( E^{-2} \right)$ \\
-  $d$ & $O\left( E^{-d/4} \right)$ 
-      & $O\left( E^{-(d+2)/4} \right)$ 
-      & $O\left( E^{-d/2} \right)$ 
-      & $O\left( E^{-(d+1)/2} \right)$ \\ 
+  $1$ & $O\left( e^{-1/4} \right)$ 
+      & $O\left( e^{-3/4} \right)$ 
+      & $O\left( e^{-1/2} \right)$ 
+      & $O\left( e^{-1} \right)$ \\ 
+  $2$ & $O\left( e^{-1/2} \right)$ 
+      & $O\left( e^{-1} \right)$ 
+      & $O\left( e^{-1} \right)$ 
+      & $O\left( e^{-3/2} \right)$ \\ 
+  $3$ & $O\left( e^{-3/4} \right)$ 
+      & $O\left( e^{-5/4} \right)$ 
+      & $O\left( e^{-3/2} \right)$ 
+      & $O\left( e^{-2} \right)$ \\
+  $d$ & $O\left( e^{-d/4} \right)$ 
+      & $O\left( e^{-(d+2)/4} \right)$ 
+      & $O\left( e^{-d/2} \right)$ 
+      & $O\left( e^{-(d+1)/2} \right)$ \\ 
   \hline 
@@
 
-\subsection{\label{sec:ots_linear_pde} 
+\subsection{\label{sec:ots_general_1d_pdes} 
+            OTS Selection for General PDEs in One Space Dimension} 
+In the previous section, we analyzed the technique of optimal time step 
+selection in the context of a simple model PDE in one space dimension.  
+For more general PDEs, the approach of carefully choosing the time step to 
+eliminate leading order terms in the error is still useful, but it needs to 
+be augmented to achieve a boost in the order of accuracy.  In addition to 
+using the optimal time step, correction terms may need to be added to the 
+finite difference scheme.  In this section, we discuss these correction terms 
+in the context of general constant coefficient, linear PDEs and semilinear
+PDEs.
+
+\subsubsection*{\label{sec:ots_linear_pde} 
             General Constant Coefficient, Linear PDEs} 
-In Section~\ref{sec:ots_model_pde}, we presented the technique of optimal 
-time step selection in the context of a simple model PDE with only two terms.  
-To extend the technique to general constant coefficient, linear PDEs of the 
-form
+Let us consider a general constant coefficient, linear PDE of the form
 \beq
   \frac{\partial u}{\partial t} = 
-  \left( \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} \right) + f(x,t),
-  \label{eq:linear_PDE}
-\eeq
-the procedure must be augmented in order to achieve any boost in the accuracy.
+  \left( \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} \right) + f(x,t)
+  \label{eq:linear_PDE}.
+\eeq
 Because it is unlikely that all of the dominant terms in the error can be
@@
 \bea
-  (L u) (\alpha \dx^r - \beta \dt^s) \dt
-  + \sum_k (L_k u) \dx^r \dt 
-  + \sum_k (L'_k u) \dt^{s+1} 
-  + (G f) \dt^{s+1}
+  \dt\left[ (L u) (\alpha \dx^r - \beta \dt^s) 
+  + \sum_k (L_k u) \dx^r 
+  + \sum_k (L'_k u) \dt^s 
+  + (G f) \dt^s \right]
   \label{eq:leading_order_error_linear_PDE_MOL}
@@
 (\ref{eq:leading_order_error_model_PDE_MOL}), 
-$L_k$ and $L'_k$ are the spatial differential operators associated with the 
-leading-order terms in the discretization error that are not involved in the 
-choice of the optimal time step, and $G$ is a spatio-temporal differential 
-operator that acts on source term $f$.  
+$L_k$ and $L'_k$ are the linear spatial differential operators associated 
+with the leading-order terms in the discretization error that are not involved 
+in the choice of the optimal time step, and $G$ is a spatio-temporal 
+differential operator that acts on source term $f$.  
 Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
@@
 for the FD scheme.  Since the specific form of the correction terms are 
-highly PDE-dependent, we defer further discussion to 
-Section~\ref{sec:applications_1d} and~\ref{sec:applications_multidim} where 
+PDE-dependent, we defer further discussion to 
+Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim} where 
 we apply optimal time step selection to several specific PDEs.
 
-\subsection{Semilinear PDEs}
+\subsubsection*{Semilinear PDEs}
 Semilinear PDEs are an important class of PDEs that have the form
@@
 leading-order spatial derivative is generally allowed to be function of 
-space and time, optimal time step selection is only applicable to PDEs where
+space and time, optimal time step selection can only be applied to PDEs where
 $A(x,t)$ is a constant.  Because the PDE is linear in the leading-order 
@@
 \bea
-  (L u) (\alpha \dx^r - \beta \dt^s) \dt
-  + \left( H u \right) \dx^r \dt 
-  + \left( H' u \right) \dt^{s+1}
-  + (G f) \dt^{s+1}
+  \dt \left[ (L u) (\alpha \dx^r - \beta \dt^s) 
+  + \left( H u \right) \dx^r 
+  + \left( H' u \right) \dt^s
+  + (G f) \dt^s \right]
   \label{eq:leading_order_error_semilinear_PDE_MOL}
@@
 (\ref{eq:leading_order_error_model_PDE_MOL}), $H$ and $H'$ are spatial 
-differential operators associated with the leading-order terms in 
-the discretization error that arise from the nonlinear term 
-in~(\ref{eq:semilinear_PDE}), and $G$ is a spatio-temporal differential 
+differential operators (potentially nonlinear) associated with the 
+leading-order terms in the discretization error that arise from the nonlinear 
+term in~(\ref{eq:semilinear_PDE}), and $G$ is a spatio-temporal differential 
 operator that acts on $f$.  To compute the optimal time step and boost
-the accuracy of the finite difference scheme, proceed in exactly
-the same manner as for general constant coefficient, linear PDEs.
-
-\subsection{Imposing Boundary Conditions}
-*** CHOICE OF DISCRETIZATION FOR BOUNDARIES IS IMPORTANT***
+the accuracy of the finite difference scheme, we proceed in the
+same manner as for general constant coefficient, linear PDEs.
+
+\subsection{Boundary Conditions}
+In our analysis and discussion thus far, we have completely ignored the issue 
+of boundary conditions.  The choice of discretization for boundary conditions
+can have a significant impact on the accuracy of the numerical solution.  In 
+general, high-order discretizations of boundary conditions are required to 
+ensure that errors introduced at the domain boundary do not pollute the 
+numerical solution in the domain interior.  For Dirichlet boundaries, it
+is sufficient to impose boundary conditions by using boundary (or ghost cell) 
+values that have the same order of accuracy as the solution.  
+More effort is required to ensure that Neumann boundary conditions do not
+reduce the order of accuracy of the solution.  In general, Neumann boundary
+conditions require that the accuracy of ghost cells slightly exceed the 
+desired order of accuracy for the solution.  A detailed discussion of 
+appropriate methods for imposing boundary conditions is beyond the scope of 
+this paper.  We refer interested readers to the thorough analysis presented
+in~\cite{gko_book}.  To avoid complicating the discussion in 
+Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim}, 
+Neumann boundary conditions are imposed by taking advantage of knowledge of 
+the analytical solution.
+
+\subsection{OTS Selection in Multiple Space Dimensions
+                \label{sec:ots_higher_spatial_dims}}
+Optimal time selection is applicable to PDEs in any number of spatial 
+dimensions.  It is straightforward to generalize the approach outlined in 
+the previous sections to problems in multiple space dimensions.  While 
+problems in higher dimensions generally require a more careful choice of 
+the finite difference stencil used to approximate spatial derivatives in 
+order to be able to combine leading-order terms in the discretization error, 
+there are no fundamental barriers that prevent the use of optimal time step 
+selection.
+
+The geometry of the domain is an important feature of problems in more than 
+one spatial dimension because it can significantly impact the way boundary 
+conditions are imposed.  For problems on regular domains, boundary conditions 
+can usually be handled by simply using 1D boundary condition discretizations 
+in each coordinate direction.  Problems on irregular domains pose a greater 
+challenge because the location of the domain boundary may not be explicitly 
+specified and grid points near the boundary require special consideration.  
+Fortunately, because optimal time step selection is based on simple, localized 
+finite difference stencils, irregular domains do not create any serious 
+difficulties.  Assuming that a `ghost cell' or `ghost value' 
+approach~\cite{fedkiw_1999, osher_fedkiw_book, gibou_2005} 
+is used to impose boundary conditions, the construction and analysis of finite 
+difference schemes and calculation of optimal time steps is identical for 
+problems on both regular and irregular domains.  For problems on 
+irregular domains, the key issue is how values at ghost cell points should 
+be calculated.  Once this issue has been addressed, devising high-order 
+accurate methods for PDEs on irregular domains is simply a matter of applying 
+optimal time step selection to the finite difference scheme used for problems 
+on regular domains.
+
+Another important issue that arises in multiple space dimensions is the choice
+of grid spacing.  In general, the grid spacing need not be equal in all 
+spatial directions.  For some problems, the ratio of the grid spacing in
+different coordinate directions may need to be carefully chosen to even 
+be able to calculate an optimal time step.
 
 \subsection{Limitations of Optimal Time Step Selection}
-As with all methods, optimal time step selection has its limitations.
-A critical assumption that we have made throughout our analysis is that
-the leading-order spatial derivative in the PDE does not have a spatially
-varying coefficient.  Without this assumption, the first term in the
-leading-order discretization error would be of the form 
+As with any numerical method, optimal time step selection has its limitations.
+The most serious limitation is the requirement that the coefficient on the 
+leading-order spatial derivative in the PDE be constant in time and space.  
+Without this critical assumption, the first term in the leading-order 
+discretization error would be of the form 
 $(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
 where $\alpha$ and $\beta$ are now functions of space and time.  This
-small change to the error makes it impossible to derive a single optimal
-time step.  In a sense, each grid point would a different optimal time
-step that also depends on the time.  For a similar reason, optimal time step 
-selection is fails for quasilinear and fully nonlinear PDEs.  The issue for 
-these types of PDEs is that the coefficients in $P(\dx, \dt)$ become 
-functions of the solution $u$ which also varies in space and time.
+small change in the form of the error makes it impossible to derive a single 
+optimal time step.  In a sense, each grid point would possess a different 
+optimal time step that could also depend on the time.  For a similar reason, 
+optimal time step selection also fails for quasilinear and fully nonlinear 
+PDEs.  The issue for these types of PDEs is that the coefficients in 
+$P(\dx, \dt)$ become functions of the solution $u$.
+
+Another limitation of optimal time step selection is that it does not 
+generally apply to finite difference schemes for systems of PDEs.  The 
+key issue is that each PDE might have its own optimal time step.  While it 
+would be possible to boost the accuracy of finite difference schemes for 
+systems of PDEs where the optimal time step is the same across the individual 
+PDEs, it is not clear that the method yields any benefit in general.
 
@@
 \beq
-  \frac{\partial u}{\partial t} + a \frac{\partial u}{\partial x} = 0,
+  \frac{\partial u}{\partial t} + A \frac{\partial u}{\partial x} = 0,
   \label{eq:advection_eqn_1d}
 \eeq
-where $a$ is the flow speed, which we take to be positive for convenience.  
+where $A$ is the flow speed, which we take to be positive for convenience.  
 For an initial condition of the form $u(x,0) = f(x)$, it is easy to use the 
-method of characteristics to show that the analytical solution is $f(x-at)$.  
+method of characteristics to show that the analytical solution is $f(x-At)$.  
 While it is unlikely that one would actually use a numerical method to 
@@
 solve~(\ref{eq:advection_eqn_1d}).  Several examples are given 
-in~\cite{leveque_book_1992}.  Perhaps the simplest, stable method uses 
-forward Euler for time integration with a first-order upwind discretization 
-for the spatial derivative term:
+in~\cite{leveque_book_1992}.  Perhaps the simplest, stable method is a 
+forward Euler for time integration scheme that uses a first-order upwind 
+discretization for the spatial derivative:
 \beq
   u^{n+1}_j = u^{n}_j 
-  - a \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right),
+  - A \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right),
   \label{eq:advection_eqn_1d_FD_scheme}
@@
 where superscripts and subscripts denote the time and space indices, 
-respectively.  Using standard error analysis, it is straightforward to show
-that method is formally first-order in both space and time.
+respectively.  It is straightforward to show that method is formally 
+first-order in both space and time.
 
@@
   \tu^{n+1}_j = \tu^{n}_j 
-  + \sum_{k=1}^\infty \frac{\left( -a \dt \right)^k}{k!} 
+  + \sum_{k=1}^\infty \frac{\left( -A \dt \right)^k}{k!} 
        \left. \frac{\partial^k \tu^n}{\px^k} \right|_j.
@@
   e^{n+1}_j = e^{n}_j 
-    &-& a \dt \left( \frac{e^{n}_j - e^{n}_{j-1}}{\dx} \right) \nonumber \\
-    &+& a \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
+    &-& A \dt \left( \frac{e^{n}_j - e^{n}_{j-1}}{\dx} \right) \nonumber \\
+    &+& A \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
         \left. \frac{\partial^k \tu^n}{\px^k} \right|_j
-        \left[ \dx^{k-1} - \left( a \dt \right)^{k-1} \right].
+        \left[ \dx^{k-1} - \left( A \dt \right)^{k-1} \right].
   \label{eq:advection_eqn_1d_err_eqn}
@@
 
-Extracting the leading-order term of the discretization error, we find that 
-it is given by
-\beq
-  \left( \left. \frac{a}{2} \frac{\partial^2 \tu^n}{\px^2} \right|_j \right)
-  \left[ \dx - a \dt \right] \dt.
-\eeq
-Thus, the optimal time step is simply $\Delta t_{opt} = \Delta x / a$.
+Extracting the leading-order term of the discretization error
+\beq
+  \left( \left. \frac{A}{2} \frac{\partial^2 \tu^n}{\px^2} \right|_j \right)
+  \left[ \dx - A \dt \right] \dt,
+\eeq
+we find that the optimal time step is $\Delta t_{opt} = \Delta x / A$.
 For the linear advection equation, setting the time step equal to the optimal 
@@
 error in the numerical solution for all times!  
-
 This surprising result, which is unique to the advection equation, is 
@@
 there are many finite difference schemes that can be used to solve the 
-diffusion equation.  Keeping with the philosophy of using simple, low-order 
+diffusion equation.  Keeping with our philosophy of using simple, low-order 
 methods to construct finite difference schemes, we apply optimal time step 
@@
 \eeq
-Standard error analysis shows that scheme is formally first-order in time and 
-second-order in space.  Due to the stability constraint 
+Conventional error analysis shows that scheme is formally first-order in time 
+and second-order in space.  Due to the stability constraint 
 $\dt \le \dx^2/2D$, the scheme is $O(\dx^2)$ accurate overall.
@@
 \eea
-Again, we see that the discretization error acts as a source term for the 
-error, which satisfies the same evolution equation as the numerical solution 
-$u$.
+Again, we see that the error satisfies the same evolution equation
+as the numerical solution and that the discretization error acts as a 
+source term.
 
@@
 \scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_no_src_error_vs_N}} 
-\caption{Error analysis for three finite difference schemes for the 
-diffusion equation: forward Euler with optimal time step (circles), forward 
-Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
-Using each method, we solve the diffusion equation with $D = 1$ on the 
-domain $0 < x < 1$ subject to the boundary conditions 
+\caption{$L^\infty$ error as a function of number of grid points for three 
+finite difference schemes that solve the diffusion equation: forward Euler 
+with optimal time step (circles), forward Euler with suboptimal time step 
+(squares), and Crank-Nicholson (diamonds).  All three schemes use the 
+standard second-order, central finite difference approximation for the 
+Laplacian.  Using each method, we solve the diffusion equation with $D = 0.5$ 
+on the domain $0 < x < 1$ subject to the boundary conditions 
 $u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
@@
 - 4 \sin(11 \pi x/2)$.  
-The exact solution for this problem is
-$u(x,t) = 1 + x/2 
-        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 Dt/4) 
-        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 Dt/4)
-        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 Dt/4)$.
-The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
-method.  Note that the errors for the forward Euler with suboptimal time 
+Note that the errors for the forward Euler with suboptimal time 
 step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figures.
+at the resolution of the figure.
 }
@@
 method that has the same computational complexity as the forward Euler scheme
-described above with an optimal time step.  However, because the forward Euler 
-scheme with optimal time step will still outperform the Crank-Nicholson scheme 
+described above with an optimal time step.  However, the forward Euler scheme 
+with optimal time step will still outperform the Crank-Nicholson scheme 
 because the latter requires the solution of at least one linear system of
@@
 
-In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of the
-OTS forward Euler scheme with and without the correction terms.  As we can 
+In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of 
+the OTS forward Euler scheme with and without the correction terms.  As we can 
 see, OTS forward Euler with correction terms is fourth-order accurate while 
@@
 \scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_error_vs_N}} 
-\caption{Error analysis for four finite difference schemes for the 
-diffusion equation: forward Euler with optimal time step (circles), forward 
-Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
-Using each method, we solve the diffusion equation with $D = 1$ on the 
-domain $0 < x < 1$ subject to the boundary conditions 
+\caption{$L^\infty$ error as a function of number of grid points for the 
+two finite difference schemes that solve the diffusion equation: 
+forward Euler with optimal time step and correction terms (circles) and 
+forward Euler with optimal time step but \emph{without} correction terms 
+(triangles).  Both schemes use the standard second-order, central finite 
+difference approximation for the Laplacian.
+Using each method, we solve the diffusion equation with $D = 0.5$ and
+source term 
+$f(x) = 5 \sin(3 \pi x/2) - 7 \sin(15 \pi x/2) + 10 \sin(21 \pi x/2)$
+on the domain $0 < x < 1$ subject to the boundary conditions 
 $u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
@@
 - 4 \sin(11 \pi x/2)$.  
-The exact solution for this problem is
-$u(x,t) = 1 + x/2 
-        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 D t/4) 
-        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 D t/4)
-        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 D t/4)$.
-The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
-method.  Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figures.
 }
@@
 \subsection{Advection-Diffusion Equation}
-Next, we use optimal time step selection in the context of the 
-advection-diffusion equation, a more general linear, constant coefficient PDE:
+We now demonstrate the use of optimal time step selection in the context of 
+a general linear, constant coefficient PDE -- the advection-diffusion 
+equation:
 \beq
   \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2} 
-  + a \frac{\partial u}{\px},
+  + A \frac{\partial u}{\px},
   \label{eq:ADE_1d}
 \eeq
-where $D$ is the diffusivity and $a$ is the flow speed.  For this problem,
-we use forward Euler for time integration and the second-order central 
+where $D$ is the diffusivity and $A$ is the flow speed.  For this problem,
+we use forward Euler for time integration and second-order central 
 difference approximations for the diffusion and advection terms:
@@
     \left( D \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} 
-         + a \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right).
+         + A \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right).
   \label{eq:ADE_1d_FD_scheme}
@@
 This scheme is formally first-order in time and second-order in space.  
-Due to the stability constraint $\dt \le \dx^2/(2 D + a \dx)$, the 
-scheme is $O(\dx^2)$ accurate overall (in the limit $\dx \rightarrow 0$).
+Due to the stability constraint $\dt < \min \{2D/A^2, \dx^2/2D \}$ 
+\cite{chan_1984}, the scheme is $O(\dx^2)$ accurate overall (in the limit 
+$\dx \rightarrow 0$).
 
@@
   &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} 
-               + a \frac{\partial \tu^n}{\px} \right)
+               + A \frac{\partial \tu^n}{\px} \right)
   \nonumber \\
@@
         D^2 \frac{\partial^4 \tu^n}{\px^4} 
-      + 2 D a \frac{\partial^3 \tu^n}{\px^3} 
-      + a^2 \frac{\partial^2 \tu^n}{\px^2} 
+      + 2 D A \frac{\partial^3 \tu^n}{\px^3} 
+      + A^2 \frac{\partial^2 \tu^n}{\px^2} 
       \right) 
-  + O \left( \dt^3 \right)
+  + O \left( \dt^3 \right).
   \label{eq:ADE_1d_time_err_modified}
 \eea
-We then obtain the local truncation error, $\tau^n_j$, for the finite 
-difference scheme~(\ref{eq:ADE_1d_FD_scheme}) using 
-the usual procedure of deriving the evolution equation for the error 
-and extracting the source term:
+The local truncation error $\tau^n_j$ for the finite difference 
+scheme~(\ref{eq:ADE_1d_FD_scheme}) can be obtained by the usual procedure 
+of deriving the evolution equation for the error and extracting the source 
+term:
 \bea
@@
       \left( \frac{\partial^4 \tu^n}{\px^4} 
-           + \frac{2a}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
+           + \frac{2A}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
       \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
   \nonumber \\
-  &-& \frac{a^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
+  &-& \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
       + O(\dt \dx^4) + O(\dt^3)
@@
 \beq
-  \frac{a^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
+  \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
   \label{eq:ADE_1d_corr_term}
 \eeq 
-Note that using the standard second-order central difference stencil to 
-compute the Laplacian in the correction term is sufficient to ensure that no 
-errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into $\tau^n_j$.  
+Note that using the standard second-order, central difference stencil to 
+compute the Laplacian in the correction term only makes an 
+$O(\dt^2 \dx^2)$ contribution to $\tau_j^n$, which ensures that no 
+errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into the 
+truncation error.
 With the optimal time step and correction term, the finite difference 
 scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
-Figure~\ref{fig:ADE_1d_error} demonstrates the 
-improved accuracy of the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) when the optimal
-time step and correction term are used.
+Figure~\ref{fig:ADE_1d_error} illustrates the improved accuracy of the finite 
+difference scheme~(\ref{eq:ADE_1d_FD_scheme}) when the optimal time step and 
+correction term are used.
 
@@
 \scalebox{0.35}{\includegraphics{figures/adv_diff_eqn_1d_error_vs_N}} 
-\caption{Error analysis for three finite difference schemes for the 
-diffusion equation: forward Euler with optimal time step (circles), forward 
-Euler with suboptimal time step (squares), and Crank-Nicholson (crosses).  
-Using each method, we solve the diffusion equation with $D = 1$ on the 
-domain $0 < x < 1$ subject to the boundary conditions 
-$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
-and the initial condition
-$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
-- 4 \sin(11 \pi x/2)$.  
-The exact solution for this problem is
-$u(x,t) = 1 + x/2 
-        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 D t/4) 
-        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 D t/4)
-        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 D t/4)$.
-The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
-method.  Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figures.
+\caption{$L^\infty$ error as a function of number of grid points for 
+two finite difference schemes that solve the advection-diffusion 
+equation: forward Euler with optimal time step and correction term (circles) 
+and forward Euler with suboptimal time step and no correction terms (squares).  
+Both schemes use second-order, central finite difference approximations for 
+the advection and diffusion terms.
+Using both methods, we solve the advection-diffusion equation with $D = 2$
+and $A = 5$ on the domain $-10 < x < 10$ subject to the initial condition
+$u(x,0) = \exp(-x^2/4)$.  The boundary conditions are set using the
+analytical solution 
+$u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
+where $T = Dt+1$.
 }
@@
 first-order upwind stencil\footnote{The appropriate upwind stencil depends
-on the sign of $a$.} for the advection term in~(\ref{eq:ADE_1d}), the 
+on the sign of $A$.} for the advection term in~(\ref{eq:ADE_1d}), the 
 discretization error for the finite difference scheme is dominated by the 
-first-order error introduced by the stencil for the advection term:
+first-order error introduced by the upwind stencil for the advection term:
 \bea
@@
       \frac{\partial^2 \tu^n}{\px^2} 
-      \left[ \frac{\dx}{2} - \frac{|a| \dt}{2}  \right] (|a| \dt)
+      \left[ \frac{\dx}{2} - \frac{|A| \dt}{2}  \right] (|A| \dt)
   \nonumber \\
@@
       \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
-           + 2 a D \frac{\partial^3 \tu^n}{\px^3}
+           + 2 A D \frac{\partial^3 \tu^n}{\px^3}
       \right)
@@
 Thus, we can see that for a finite difference scheme based a first-order
-upwind discretization of the advection term, we optimal time step is 
-$\dto = \dx/|a|$ (as for the pure advection equation) and the required
+upwind discretization of the advection term, the optimal time step is 
+$\dto = \dx/|A|$ (as for the pure advection equation) and the required
 correction term is
@@
 the power of optimal time step selection is highly dependent on the 
-the original finite difference scheme and underscores the importance of
-carefully choosing the finite difference scheme even when optimal time
+the original choice of finite difference scheme and underscores the importance 
+of carefully choosing the finite difference scheme even when optimal time
 step selection is used.
@@
 to the advection-diffusion equation, except that the advection velocity now 
-depends explicitly on the solution $u$.  To solve this equation numerically,
+depends on the solution $u$.  To solve this equation numerically,
 we will use forward Euler for time integration and second-order central 
@@
   + \dt 
-    \left( \nu \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} 
+    \left( \nu 
+      \left [ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
          - u^n_j 
@@
 This scheme is formally first-order in time and second-order in space.  
-Due to the stability constraint $\dt \le \dx^2/(2 \nu + a \dx)$, the 
-scheme is $O(\dx^2)$ accurate overall (in the limit $\dx \rightarrow 0$).
+Using the stability condition for the advection-diffusion equation as a 
+heuristic guide, we expect a stability constraint of the form 
+$\dt \le \dx^2/2 \nu$ in the limit $\dx \rightarrow 0$ (assuming that 
+the solution $u$ remains bounded).  Therefore, the scheme 
+(\ref{eq:burgers_1d_FD_scheme}) is $O(\dx^2)$ accurate overall.
 
@@
 equation~(\ref{eq:ADE_1d_FD_scheme}), the analysis for the error analysis
-is almost the same as in the previous section.  Following the usual 
-procedure, we find that the local truncation error takes the form
+is almost the same as in the previous section.  Following the procedure
+described for the advection-diffusion equation, we find that the 
+local truncation error has the form
 \bea
@@
 \eeq 
-Note that using the standard second-order central difference stencils to 
-compute the correction term is sufficient to ensure that the overall error
-for the scheme is $O(\dx^4)$.  Figure~\ref{fig:burgers_1d_error} shows that
-using an optimal time step and correction terms improves the accuracy of the 
-finite difference scheme~(\ref{eq:burgers_1d_FD_scheme}).
+As for the advection-diffusion, using the standard second-order central 
+difference stencils to compute the correction term is sufficient to ensure 
+that the overall error for the scheme is $O(\dx^4)$.  
+Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate
+the boost in accuracy obtained by using the optimal time step and 
+adding correction terms to the finite difference 
+scheme~(\ref{eq:burgers_1d_FD_scheme}).
 
@@
 \scalebox{0.35}{\includegraphics{figures/burgers_eqn_1d_error_vs_N}} 
-\caption{Error analysis for two finite difference schemes for the 
-viscous Burgers equation: forward Euler with optimal time step (circles)
-and forward Euler with suboptimal time step (squares).
-Using each method, we solve the viscous Burgers equation ??
-}
-\label{fig:burgers_1d_error}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \ 
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D viscous Burgers 
-equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step (right).  The latter scheme uses a 
-time step $\Delta t = \Delta x^2 / 3 \nu$.  Both figures show the numerical
-solution at $t = 2$ computed on the domain $0 < x < 10$ using 100 grid points.
-The viscosity $\nu$ is set to $0.1$.  The initial conditions are taken to be 
+\caption{$L^\infty$ error as a function of number of grid points for
+two finite difference schemes that solve the viscous Burgers equation:
+forward Euler with optimal time step and correction term (circles)
+and forward Euler with suboptimal time step $\dt = \dx^2/4\nu$ and no 
+correction terms (squares).
+Both schemes use second-order, central finite difference approximations for
+the advection and diffusion terms.
+The viscous Burgers equation is solved on the domain $0 < x < 10$ to 
+time $t=2$ with the viscosity $\nu$ set to $0.1$.  The initial conditions are 
+taken to be 
 $u(x,0) = 1 + \gamma \sqrt{\frac{\nu}{\pi}} \exp\left( -(x-1)^2/4 \nu \right) 
@@
 }
+\label{fig:burgers_1d_error}
+\end{center}
+\end{figure}
+
+\begin{figure}[htb]
+\begin{center}
+\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
+\ \ \ \ \ 
+\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
+\caption{Comparison of numerical solutions for the 1D viscous Burgers 
+equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
+optimal time step and correction terms (left) and forward Euler time 
+stepping with suboptimal time step (right).  Both figures show the 
+numerical solution at $t = 2$ computed using $100$ grid points.
+For details about the initial/boundary conditions and physical parameters 
+used to compute these solutions, see the caption for 
+Figure~\ref{fig:burgers_1d_error}. 
+}
 \label{fig:burgers_1d_solns}
@@
 As a final example in one space dimension, we consider a PDE with high-order
-spatial derivatives -- a fourth-order parabolic equation 
-REPLACE WITH A MORE PHYSICAL NAME:
-\beq
-  \frac{\partial u}{\pt} = -\kappa \frac{\partial^4 u}{\px^4} + f(x,t),
+spatial derivatives -- a fourth-order parabolic equation: 
+\beq
+  \frac{\partial u}{\pt} = -\kappa \frac{\partial^4 u}{\px^4} + f(x,t).
   \label{eq:4th_order_parabolic_eqn_1d}
 \eeq
-where $\kappa$ is the ??.  From our previous examples, we know that 
-optimal time step selection requires that the coefficient on leading-order 
-spatial discretization error must be directly related to the second 
-derivative of the solution $u$ with respect to time.  
+From our previous examples, we know that optimal time step selection requires 
+the coefficient on leading-order spatial discretization error to be directly 
+related to the second derivative of the solution $u$ with respect to time.  
 From~(\ref{eq:4th_order_parabolic_eqn_1d}), we see that 
@@
 This scheme is formally first-order in time and fourth-order in space.  
-Due to the stability constraint $\dt \le \dx^4/? \kappa$, the scheme is 
+Due to the stability constraint $\dt \le 3\dx^4/40 \kappa$, the scheme is 
 $O(\dx^4)$ accurate overall.
@@
 sixth-order accurate.  As with the previous examples, the correction term 
-can be computed analytically or numerically.
-
-Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} demonstrates the 
-accuracy of the finite difference 
-scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) when the optimal
-time step and correction terms are used.  Observe that the choice of the
-finite difference approximation used to compute the correction terms
-impacts the overall accuracy of the scheme.  Surprisingly, the
-second-order central difference approximation for $f_{xxxx}$ yields a
-more accurate solution than the fourth-order scheme.  Fortuitous cancellation 
-of error is the origin of the superior results obtained by using the 
-second-order scheme.  A cursory analysis shows that using a second-order 
-stencil for $f_{xxxx}$ yields a $O(\dx^{10})$ error involving 
-$\left(\partial_x\right)^{10}u$ and $\left(\partial_x\right)^{6}f$ with 
-opposite signs.  Because these two terms are related to each other through the 
-PDE~(\ref{eq:4th_order_parabolic_eqn_1d}), they can partially anihilate
-each other.  On the other hand, when the fourth-order stencil is used
-for $f_{xxxx}$, there is no chance for cancellation because the $O(\dx^{10})$ 
-error only involves $\left(\partial_x\right)^{10}u$.
-Again, we see that blindly using a higher-order scheme does not
-necessarily lead to a more accurate numerical solution.
+can be computed analytically or numerically.  The improved accuracy of 
+(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) when the optimal time step 
+and correction terms are used is illustrated in 
+Figure~\ref{fig:4th_order_parabolic_eqn_1d_error}.
 
@@
   \includegraphics{figures/4th_order_parabolic_eqn_1d_comp_time}} 
-\caption{Error analysis for three finite difference schemes for the 
-diffusion equation: forward Euler with optimal time step (circles), forward 
-Euler with suboptimal time step (squares), and Crank-Nicholson (crosses).  
-Using each method, we solve the diffusion equation with $\kappa = 1$ on the 
-domain $0 < x < 1$ subject to the boundary conditions 
-$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
-and the initial condition
-$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
-- 4 \sin(11 \pi x/2)$.  
-The exact solution for this problem is
-$u(x,t) = 1 + x/2 
-        + 2 \sin(5 \pi x/2)  \exp(-25 \pi^2 \kappa t/4) 
-        + 3 \sin(7 \pi x/2)  \exp(-49 \pi^2 \kappa t/4)
-        - 4 \sin(11 \pi x/2) \exp(-121 \pi^2 \kappa t/4)$.
-The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
-method.  Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figures.
+\caption{$L^\infty$ error as a function of number of grid points (left)
+and computation time as a function of $L^\infty$ error (right) for two 
+finite difference schemes that solve the fourth-order parabolic equation 
+(\ref{eq:4th_order_parabolic_eqn_1d_space_err}):
+forward Euler with optimal time step and correction term (circles)
+and forward Euler with suboptimal time step $\dt = \dx^2/16$ and no 
+correction terms (squares).
+Both schemes use a fourth-order, central finite difference approximation for
+the bilaplacian.  Using each method, we solve the fourth-order parabolic
+equation with $\kappa = 1$ on the domain $0 < x < 1$ subject to the boundary 
+conditions $u(0,t) = u(1,t) = 0$ and the initial condition
+$u(x,0) = 2\sin(2 \pi x) + \sin(5 \pi x)$.  The value of the solution
+in the ghost cells (required as a result of the wide stencil for the 
+discrete bilaplacian) are filled by taking values from the exact solution: 
+$u(x,t) = 2\sin(2 \pi x)\cos(10000 \pi t) 
+        + \sin(5 \pi x) \left[ 2-\exp(-100\pi^2 t) \right]$.
 }
@@
 and makes it possible to obtain accurate solutions on relatively coarse 
-grids.  For a PDE of order $n$, explicit schemes typically require 
-the time step to be $O(\dx^n)$.  Thus, it makes sense to choose the order of 
-the finite difference stencils for the spatial derivatives to be at least 
-$n/s$ where $s$ is the order of accuracy for the time integration scheme.  
-This choice yields a finite difference scheme that with an error that is
-$O(\dx^n)$ overall.}.  
+grids.}. 
 Optimal time step selection merely provides a valuable way to leverage the 
@@
 equation~(\ref{eq:4th_order_parabolic_eqn_1d_second_time_derivative}).
-Observe the dramatic decrease in computational cost as a function of 
+Observe the dramatic decrease in computational cost as a function of the
 error by simply replacing a second-order discretization for the bilaplacian 
@@
 using an optimal time step drives the computational cost down even more.
+The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} 
+shows that the theoretical estimates of compute time are in good agreement 
+with timings from actual computational experiments.
+
 Comparing the best forward Euler and Crank-Nicholson schemes shows that
@@
 \caption{\label{tab:comp_perf_vs_err_4th_order_parabolic} 
-   Computational cost as a function of the global numerical error $E$
+   Computational cost as a function of the global numerical error $e$
    for various FD schemes that numerically solve the 1D fourth-order
    parabolic equation.  For all FD schemes, the numerical solution is 
-   assumed to be computed over a fixed time interval.  A less negative 
-   power of $E$ for the memory and compute time means that a FD scheme is 
-   more computationally efficient.
+   assumed to be computed over a fixed time interval.
 }
@@
    Crank-Nicholson schemes with second- and fourth-order spatial 
-   discretizations are $O(\dx)$ and $O(\dx^2)$, respectively, so that the 
-   temporal error is subdominant to the spatial error.}
+   discretizations are taken to be $O(\dx)$ and $O(\dx^2)$, respectively, so 
+   that the temporal error is subdominant to the spatial error.}
   & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
@@
   Forward Euler with 
-    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left (E^{2} \right)$}
-    & \multirow{2}{*}{$O\left( E^{-1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-5/2} \right)$} \\
+    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{2} \right)$}
+    & \multirow{2}{*}{$O\left( e^{-1/2} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-5/2} \right)$} \\
   \ \ 2nd-Order Bilaplacian & & & & \\
@@
   Forward Euler with 
-    & \multirow{2}{*}{$O\left( E^{1/4} \right)$}
-    & \multirow{2}{*}{$O\left (E \right)$}
-    & \multirow{2}{*}{$O\left( E^{-1/4} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-{5/4}} \right)$} \\
+    & \multirow{2}{*}{$O\left( e^{1/4} \right)$}
+    & \multirow{2}{*}{$O\left( e \right)$}
+    & \multirow{2}{*}{$O\left( e^{-1/4} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-{5/4}} \right)$} \\
   \ \ 4th-Order Bilaplacian & & & & \\
@@
   Forward Euler with 
-    & \multirow{2}{*}{$O\left( E^{1/6} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{2/3} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-1/6} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-5/6} \right)$} \\ 
+    & \multirow{2}{*}{$O\left( e^{1/6} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{2/3} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-1/6} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-5/6} \right)$} \\ 
   \ \ 4th-Order Bilaplacian and OTS & & & & \\
@@
   Crank-Nicholson with 
-    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-1} \right)$} \\
+    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-1/2} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-1} \right)$} \\
   \ \ 2nd-Order Bilaplacian & & & & \\
@@
   Crank-Nicholson with 
-    & \multirow{2}{*}{$O\left( E^{1/4} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-1/4} \right)$} 
-    & \multirow{2}{*}{$O\left( E^{-3/4} \right)$} \\
+    & \multirow{2}{*}{$O\left( e^{1/4} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-1/4} \right)$} 
+    & \multirow{2}{*}{$O\left( e^{-3/4} \right)$} \\
   \ \ 4th-Order Bilaplacian & & & & \\
@@
          Application to PDEs in Multiple Space Dimensions}
-A key distinction between the error for FD schemes in one space dimension
-and multiple space dimensions is the presence of cross-terms in the 
-discretization error.  In the context of a standard analysis for a finite 
-difference scheme, it is usually safe to ignore the precise form of the 
-cross-terms because they only appear as higher-order corrections in the 
-error.  As a result, the choice of discretization for the spatial derivative 
-operators is not critical to achieve a desired accuracy for the overal FD 
-scheme.  In contrast, knowledge of the form of the cross-terms is critical 
-when attempting to use optimal time step selection.  Without this knowledge, 
-it is impossible to ensure cancellation of higher-order errors.
-
-Recall that optimal time step selection requires that higher-order temporal 
-errors be directly related to higher-order spatial errors through the use 
-of the PDE.  This requirement can be satisfied in one of two ways:
+Optimal time step selection can be used to boost the order of accuracy for 
+finite difference schemes in any number of space dimensions.  However, 
+several important issues arise for problems in more than one space dimension.  
+These issues are primarily related to the presence of cross-terms in the 
+discretization error, multiple grid spacing parameters,  and the domain 
+geometry.  In this section, we address these issues and demonstrate the use 
+of optimal time step selection in the construction of finite difference 
+schemes for two linear, time dependent PDEs in two space dimensions on both 
+regular and irregular domains.  It is straightforward to generalize the 
+approach presented in this section to more space dimensions and to construct 
+methods for semilinear PDEs by incorporating ideas discussed in 
+Sections~\ref{sec:optimal_time_step_selection} and~\ref{sec:applications_1d}.
+
+\subsection{\label{sec:ots_multidim_crossterms}
+            Discretization of Spatial Derivatives}
+A significant distinction between the error for FD schemes in one space 
+dimension and multiple space dimensions is the presence of cross-terms in 
+the discretization error.  When analyzing finite difference schemes using 
+conventional numerical analysis techniques, it is usually safe to ignore the 
+precise form of the cross-terms because they only appear as higher-order 
+corrections in the error.  As a result, the order of accuracy for the FD 
+scheme is not sensitive to the particular choice of discretization for the 
+spatial derivative operators.  In contrast, knowledge of the form of the 
+cross-terms is critical when attempting to use optimal time step selection.  
+Without this knowledge, it is impossible to ensure cancellation of 
+higher-order errors.
+
+Recall that optimal time step selection requires the higher-order temporal 
+errors to be directly related to higher-order spatial errors via the PDE.  
+This requirement can be satisfied in one of two ways:
 \begin{itemize}
@@
       derivative operators in the PDE; or
-\item by explicitly adding temporal correction terms to cancel out cross-terms 
-      that are not eliminated  by the original finite difference scheme.
+\item by explicitly adding correction terms to cancel out any cross-terms 
+      in the original finite difference scheme that are cannot be eliminated 
+      by using the PDE.
 \end{itemize}
 While these two approaches may appear to be different, they are actually 
-equivalent because the optimal time step is \emph{not independent} of the 
-grid size.  Thus, the temporal correction terms can be viewed as part of
-an expanded spatial discretization scheme and vice-versa.
-
-In the remainder of this section, we apply optimal time step selection 
-to the construction of finite difference schemes for two classical linear,
-time dependent PDEs in two space dimensions, including an example that
-demonstrates how optimal time step selection can be used to design simple
-schemes for problems on irregular domains.  It is straightforward to 
-generalize the approach presented in this section to more space dimensions 
-and to construct methods for semilinear PDEs by incorporating ideas discussed 
-in Sections~\ref{sec:optimal_time_step_selection} 
-and~\ref{sec:applications_1d}.
+equivalent because the correction terms can be combined with the original 
+FD schemes.  In essense, the correction terms can be viewed as part of an 
+expanded spatial discretization scheme and vice-versa.
 
@@
 where superscripts and subscripts denote the time and space indices, 
-respectively.  Using standard error analysis, it is straightforward to show
-that this method is formally first-order in both space and time.
+respectively.  Using conventional error analysis, it is straightforward to 
+show that this method is formally first-order in both space and time.
 
@@
 \eea
-With these choices, we have already boosted the order of accuracy for the 
-FD scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from 
-first- to second-order.
+With these choices, we are able to boost the order of accuracy for the 
+FD scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from first- to second-order.
 
@@
 preserved when the optimal time step, optimal grid spacing ratio, and 
-correction terms are used.  In contrast, the sharp corners are
+correction term are used.  In contrast, the sharp corners are
 completely smoothed out in the solution computed using the standard 
@@
 $u(x,y,0) = 0$ if $|x| + |y| > 2$. 
-For the time interval considered, the $u$ is identically zero 
+For the time interval considered, $u$ is identically zero 
 on the boundary.
@@
 
-To see that the temporal correction term can be considered as part of the 
-spatial discretization scheme, we can write out the complete FD scheme 
-including the correction term:
+To see that the correction term can be considered as part of the spatial 
+discretization scheme, we can write out the complete FD scheme including 
+the correction term:
 \bea
@@
 coordinate direction.   As with the two-dimensional problem, the optimal 
-grid spacings satisfy $\dx_k/A_{x_k} = C$ for $1 \le k \le d$, where $C$ is 
-a constant equal to the optimal time step.
+grid spacings satisfy $\dx_k/A_{x_k} = \dt_{opt}$ for $1 \le k \le d$.
 
@@
 becomes apparent when we fully simplify 
-(\ref{eq:advection_eqn_2d_FD_scheme_simplified}) to obtain
+(\ref{eq:advection_eqn_2d_FD_scheme_simplified}) using the 
+optimal time step and grid spacing ratio to obtain
 $u^{n+1}_{i,j} = u^{n}_{i-1,j-1}$.
@@
 \eea
-Standard error analysis shows that this scheme is formally first-order in 
+Conventional error analysis shows that this scheme is formally first-order in 
 time and second-order in space.
@@
 to the spatial derivative operator in the diffusion equation.  
-The analysis is more straightforward straightforward if we construct our
-finite difference scheme using the nine-point stencil for the 
-Laplacian~\cite{iserles_book,patra_2005,??}, which possesses a leading-order 
+The analysis is more straightforward if we construct our finite difference 
+scheme using the nine-point stencil for the 
+Laplacian~\cite{iserles_book,patra_2005}, which possesses a leading-order 
 error directly related to the spatial derivative operator in the PDE 
@@
 is $O(\dx^2)$ accurate overall because it has a stability constraint 
-$\dt \le ??\dx^2/2D = O(\dx^2)$.
+$\dt \le 3\dx^2/8D = O(\dx^2)$.
 It is interesting to note that use of a nine-point stencil for the Laplacian 
 naturally arises when directly applying optimal time step selection to the
-five-point scheme through reinterpretion of the temporal correction term
-as a modification to the spatial discretization.
+five-point scheme through reinterpretion of the correction term as a 
+modification to the spatial discretization.
 
@@
 
-The improvement in accuracy of the forward Euler scheme when the optimal time 
-step is used is clear in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
-The numerical errors of the following three finite difference schemes for 
-the diffusion equation show the expected orders of accuracy: forward Euler 
-with optimal time step, forward Euler with suboptimal time step, and 
-Crank-Nicholson.  As with our other examples, the forward Euler scheme 
-with an optimal time step is much more efficient that the two other schemes.  
-Even with as few as 25 grid points in each space direction, the error in 
-solution computed by the forward Euler scheme with an optimal time step is 
-orders of magnitude smaller.
-
 \begin{figure}[htb]
@@
 \scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_comp_time}} 
-\caption{Error and performance analysis for three finite difference schemes 
-for the 2D diffusion equation: forward Euler with optimal time step (circles), 
-forward Euler with suboptimal time step (squares), and Crank-Nicholson 
-(diamonds).  Using each method, we solve the 2D diffusion equation with 
-$D = 1$ on the domain $0 < x < 1$ subject to the boundary conditions 
-and the initial condition
-$??$
-The exact solution for this problem is
-$u(x,t) = ??$
-The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
-method.  Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figures.
+\caption{$L^\infty$ error as a function of number of grid points (left)
+and computation time as a function of $L^\infty$ error (right) for three
+finite difference schemes that solve the 2D diffusion equation:
+forward Euler with optimal time step and correction terms (circles), forward 
+Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
+The forward Euler schemes use a nine-point discretization of the Laplacian;
+the Crank-Nicholson scheme uses the standard five-point discretization of
+the Laplacian.  Using each method, we solve the 2D diffusion equation with 
+$D = 0.25$ on the domain $0 < x, y < 1$ subject to the initial condition
+$u(x,y,0) =  1 + y/3 + x/2 + xy/4 
+          + \sin(2 \pi x) \sin(3 \pi y) - \sin(5 \pi x) \sin(\pi y)$
+and boundary conditions taken from the analytical solution 
+$u(x,t) = 1 + y/3 + x/2 + xy/4 
+          + \sin(2 \pi x) \sin(3 \pi y) \exp(-5 D \pi^2 t) 
+          - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)
+$.
+Note that the errors for the forward Euler with suboptimal time step and 
+Crank-Nicholson schemes lie almost directly on top of each other at the 
+resolution of the figures.
 }
@@
 
+The improvement in the accuracy of the forward Euler scheme when the optimal 
+time step is used is clear in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
+The numerical errors of the following three finite difference schemes for 
+the 2D diffusion equation show the expected orders of accuracy: forward Euler 
+with optimal time step and correction terms, forward Euler with suboptimal 
+time step, and Crank-Nicholson.  As with our other examples, the forward 
+Euler scheme with an optimal time step is much more efficient that the two 
+other schemes.  With as few as 25 grid points in each space direction, the 
+error in solution computed by the forward Euler scheme with an optimal time 
+step is orders of magnitude smaller.
+
 Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also compares the computational
-performance of the three schemes.  For all three schemes, notice that the 
-scaling of the computational time with error is in good agreement with the 
-theoretical expectations (see Table~\ref{tab:comp_perf_vs_dim}).  Again, the 
+performance of the three schemes.  For all three schemes, the scaling of the 
+computational time with error is in good agreement with the theoretical 
+expectations presented in Table~\ref{tab:comp_perf_vs_dim}.  Again, the 
 forward Euler with optimal time step demonstrates much greater efficiency 
@@
 defined on simple, rectangular domains.  Thus, any of several methods 
-could have been used to achieve similar or even higher accuracy solutions.
+could have been used to obtain solutions with similar or even higher accuracy.
 For example, a fourth-order accurate solution to the diffusion equation
 could be obtained by using a pseudospectral spatial
-discretization~\cite{trefethen_spectral_book, boyd_book} of the Laplacian
-and a fourth-order time integration scheme (\eg Runge-Kutta).
+discretization~\cite{trefethen_spectral_book, boyd_spectral_book} of the 
+Laplacian and a fourth-order time integration scheme (\eg Runge-Kutta).
 When the domain is irregularly shaped, high-order methods are harder to design 
-and implement.  For instance, spectral and pseudospectral methods are no 
-longer appropriate because the domain is not a simple cartesian product of 
-intervals\footnote{Clever phase field approaches have been used successfully 
-to make these methods applicable to problems with irregular 
-boundaries~\cite{liu??}.}.
+and implement.  For instance, spectral and pseudospectral methods are not 
+usually appropriate because the domain is not a simple cartesian product of 
+intervals.
 
@@
 that yield high-order accurate solutions for problems on irregular domains.  
-In this section, we extend the OTS finite difference scheme for the diffusion
-equation on regular domains to irregular domains by describing a method for 
-computing the values of the solution in ghost cells that ensures that 
-imposing boundary conditions does not decrease the accuracy of the solution.
-
-One important issue that arises when devising numerical schemes for problems 
-on irregular domains is how the domain and boundary should be represented.
-The various ways for representing irregular boundaries can be divided 
-into two categories: explicit and implicit.  A full comparison of the many 
-representations of irregular boundaries is beyond the scope of this paper, 
-so we refer interested readers to~\cite{??} for more detailed discussion 
-about the advantages and disadvantages of several popular methods.  For our 
-purposes, we choose to implicitly represent the irregular boundary as the 
-zero level set of a level set function $\phi$ which satisfies:
+In this section, we extend the OTS finite difference scheme for the 2D
+diffusion equation on regular domains to irregular domains by describing a 
+method for computing the values of the solution in ghost cells that ensures 
+that imposing boundary conditions does not decrease the accuracy of the 
+solution.  For simplicity, we choose to represent the irregular boundary 
+implicitly as the zero level set of a level set function $\phi$ which 
+satisfies:
 \bea
@@
 idea is to set the value of grid points that lie outside of the domain 
-\emph{and} are required by the spatial discretization operator for grid 
-points inside the domain by extrapolating from interior grid points using
+and are required by the spatial discretization operator for grid points 
+inside the domain by extrapolating from interior grid points using
 boundary condition information.  Figure~\ref{fig:ghost_cells} shows examples 
@@
 
-The value in each edge ghost cell is set by using 1D, cubic extrapolation of 
+To set the value in each edge ghost cell, we use 1D, cubic extrapolation of 
 the values of $u$ across the interface.  A cubic Lagrange extrapolant is 
-constructed for each edge ghost cell using the point on the domain boundary 
-where the zero level set of $\phi$ intersects the edge connecting the ghost 
-cell to its nearest interior neighbor and three interior grid points along 
-the line parallel to the same edge (see Figure~\ref{fig:ghost_cells}).  
+constructed for each edge ghost cell using a point on the domain boundary 
+and three interior grid points that lie along the line parallel to the edge 
+connecting the ghost cell to its nearest interior neighbor (see 
+Figure~\ref{fig:ghost_cells}).  
 The location of the point $u_B$ can be calculated by inverting the linear 
 interpolant of $\phi$ that passes through the ghost cell and its nearest
-interior neighbor.
+interior neighbor\footnote{We found that using higher-order interpolants to 
+locate boundary points was unnecessary to achieve high-order accuracy in the
+computed solution.}.
 Because the quality of the extrapolant deteriorates if the boundary point is 
@@
 interior grid point is sufficiently far from the boundary point.  
-Gibou~\etal suggest shifting the interpolation points by one grid point 
+Gibou~\etal~suggest shifting the interpolation points by one grid point 
 when the distance between the boundary point and the nearest interior grid 
 point is less than $\dx^2$.  For the numerical scheme presented in this paper, 
-thie threshold was too low.  We found that better results were obtained if 
-we used a $O(\dx)$ threshold.
-
-The values in corner ghost cells are set using 2D quadratic extrapolation
-of the values of $u$ across the domain boundary \emph{and} the neighboring
-edge ghost cells (see Figure~\ref{fig:ghost_cells}).  Notice that unlike
-edge ghost cells, corner ghost cells only need to be 3rd-order accurate 
-in order to obtain a 4th-order accurate solution.  The reason for this is ??.
-*** KTC - FILL ME IN ***
-
-To derive quadratic extrapolant at each corner ghost cell, we use a 
-second-order Taylor series expansion around the nearest interior grid point,
-$x_{0,0}$, and use finite differences to approximate all required partial 
-derivatives.  For example, using central differences centered at $x_{0,0}$ for 
-all partial derivatives, the quadratic extrapolant for the corner ghost cell 
-shown in Figure~\ref{fig:ghost_cells} is
-\bea
-  u_C &=&  -u_{0,0} + \frac{1}{2} u_{-1,-1} 
-      \nonumber \\
-      &+& \frac{5}{4} u_{1,0} + \frac{5}{4} u_{0,1}
-      - \frac{1}{4} u_{1,-1} - \frac{1}{4} u_{-1,1}
-      - \frac{1}{4} u_{0,-1} - \frac{1}{4} u_{-1,0}
-\eea
-where $u_C$ and $u_{i,j}$ are as indicated in the figure.  Note that the 
-stencil for the quadratic interpolant depends on the choice of finite 
-difference approximation for the partial derivatives.  For the results 
-presented in this paper, we use the above stencil (and its analogues for 
-other positions of the ghost cell relative to the domain) to define the
-quadratic extrapolant.  However, we obtained the same qualitative results 
-when stencils based on other choices of finite difference approximations were
-used.
+this threshold was too low.  We found that better results were obtained if 
+we used an $O(\dx)$ threshold.
+
+To set the values in corner ghost cells, we use a fourth-order accurate
+Taylor series expansion of $u$ centered at the nearest interior 
+neighbor.  The partial derivatives in the Taylor series expansion are 
+approximated using finite differences constructed from the values of $u$ 
+across the domain boundary and the neighboring edge ghost cells (see 
+Figure~\ref{fig:ghost_cells}).  To obtain a fourth-order accurate value for 
+the corner ghost cell, we need to carefully choose the finite difference 
+stencils used for the partial derivatives.  For the corner ghost cell shown 
+in Figure~\ref{fig:ghost_cells}, the extrapolation stencil is given by
+\bea
+  u_C &=&  -4 u_{0,0} - u_{-1,-1} + 2 u_{1,0} + 2 u_{0,1}
+      - u_{1,-1} - u_{-1,1} + 2 u_{0,-1} + 2 u_{-1,0}
+\eea
+where $u_C$ and $u_{i,j}$ are as indicated in the figure.  The stencils for
+corner ghost cells that are in other positions relative to the interior of 
+the domain are obtained by rotations of Figure~\ref{fig:ghost_cells} and 
+remapping of the indicies in the above stencil.
+
+Using these methods for filling edge and corner ghost cells, it is 
+straightforward to compute high-order accurate solutions of the 2D diffusion 
+equation on irregular domains.  We simply use the OTS forward euler scheme 
+designed for regular domains for grid points in the interior of the domain
+after filling the ghost cells with high-order accurate values.  
+Figure~\ref{fig:diffusion_eqn_2d_starfish_domain} shows an example of 
+a solution computed on a starfish shaped domain.  It also shows that the 
+error in the solution is concentrated near the boundaries of the irregular
+domain.  Thus, we see that it is necessary to use high-order accurate 
+extrapolants for the ghost cells to prevent errors at the boundary from 
+destroying high-order accuracy throughout the domain.  The boosted accuracy 
+of the OTS solution compared to a simple forward Euler scheme with a 
+suboptimal time step is illustrated in 
+Figure~\ref{fig:diffusion_eqn_2d_starfish_error}.
 
@@
 \caption{Numerical solutions (left) and dominant error (right) for the 2D 
-diffusion equation~(\ref{eq:??}) a starfish-shaped domain.  The numerical 
-solution is computed at time $t = 0.1$ using forward Euler time stepping 
-with an optimal time step and correction terms.  In the error plot, the dark 
-regions represent points where the error in the solution is larger than 
-$10$\% of the $L^\infty$ error of the solution.  Notice that the largest 
-errors are concentrated near the boundaries of the mathematical domain.  To 
-generate these results, the computational domain is taken to be 
-$-1 < x,y < 1$ with 400 grid points in each direction.  The diffusion 
-constant $D$ is set equal to $1/4$.  The source term is given by
+diffusion equation on a starfish-shaped domain.  The solution is computed at 
+time $t = 0.1$ using forward Euler time stepping with an optimal time step 
+and correction terms.  In the error plot, the dark regions represent points 
+where the error in the solution is larger than $25$\% of the $L^\infty$ error 
+of the solution.  For these figures, the diffusion constant $D$ is set to 
+$1/4$ and the source term is taken to be 
 $f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
@@
 $
-The initial conditions are taken to be 
+The initial conditions for the problem are 
 $u(x,y,0) = 1 + y/3 + x/2 + xy/4 
@@
 $
-and boundary conditions are imposed by using cubic and quadratic 
-extrapolation of interior and boundary values to to fill edge and corner 
-ghost cells, respectively.  Boundary values are obtained from the 
-analytical solution
+and the boundary conditions are obtained from the analytical solution
 $u(x,y,t) = 1 + y/3 + x/2 + xy/4 
               + \sin(2 \pi x) \sin(3 \pi y) \exp(-5 D \pi^2 t) 
-              - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)$.
-The boundaries of the domains are represented by the zero-level 
-set of the function
+              - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)
+$.
+The boundary of the starfish domain is represented by the zero-level set of 
+the function
 $\sqrt{x^2 + y^2} - \frac{3}{5} \left(1 + \frac{1}{2}\sin(5 \theta) \right)$, 
 where $\theta = \mathtt{atan}\left( y/x \right)$.
+The numerical solution is calculated using the square computational domain 
+$-1 < x,y < 1$ which is discretized using 800 grid points in each coordinate 
+direction (the plot of the solution has been downsampled for visualization 
+purposes).  Boundary conditions are imposed by using fourth-order accurate
+approximations of the value of the solution in edge and corner ghost cells
+(described in the main text).
 }
@@
   \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
-\caption{Error analysis for four finite difference schemes 
-for the 2D diffusion equation: forward Euler with optimal time step (circles), 
-forward Euler with suboptimal time step (squares), 
-(diamonds).  Using each method, we solve the 2D diffusion equation with 
-$D = 1$ on the domain $0 < x < 1$ subject to the boundary conditions 
-and the initial condition
-$??$
-The exact solution for this problem is
-$u(x,t) = ??$
-The $L^\infty$ error is plotted against the grid spacing $\dx$ for each 
-method.  Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figures.
+\caption{$L^\infty$ error as a function of number of grid points for two
+finite difference schemes that solve the 2D diffusion equation on an 
+irregular domain: forward Euler with optimal time step and correction 
+terms (circles) and forward Euler with suboptimal time step (squares).
+Both methods solve the diffusion equation described in the caption for
+Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
 }
@@
 
-\subsubsection{Generalization to Higher Dimensional Problems}
-** COMPLEXITY OF EXTRAPOLATION DEPENDS ON CHOICE OF DISCRETIZATION FOR LAPLACIAN **
-
-** PATRA AND KARTTUNEN GIVE ISOTROPIC STENCILS FOR 3D **
-
-\subsubsection{Weakly Anisotropic Diffusivity Tensor}
-can use unequal grid spacing when the diffusivity tensor is diagonlizable 
-(i.e. normal) and the eigenvalues are unequal.  first transform to coordinate
-system of principle axes then rescale problem.
+\subsubsection{Diffusion Equation in Higher Dimensions}
+Using optimal time step selection to boost the accuracy of FD schemes for 
+the diffusion equation in higher space dimensions is straightforward.  
+As for the 2D problem, the key step is identification of a finite difference 
+stencil for the Laplacian that has an \emph{isotropic} discretization error.
+For 3D problems, a catalog of several such stencils is available 
+in~\cite{patra_2005}.  For higher dimensional problems, using symbolic 
+algebra software to derive FD stencils with desired discretization error 
+properties can be helpful~\cite{patra_2005, gupta_1998}.  Once an acceptable
+FD approximation for the Laplacian has been chosen, the usual OTS analysis 
+yields an optimal time step of $\dt_{opt} = \dx^2/6D$ for forward Euler
+time integration and a boost of the order of accuracy from two to four.
+
+Problems on irregular domains can be handled using the ghost cell approach. 
+While the derivation of high accuracy stencils for ghost cells becomes more
+tedious, the general approach is the same as for the 2D problem.  The ghost 
+cells can first be grouped into separate classes based on the number of 
+grid cell edges that must be traversed to reach the nearest interior neighbor.
+The values in the ghost cells can then be set in order starting with those 
+directly connected to their nearest interior neighbor and ending with those 
+whose nearest interior neighbor is on the opposite corner of a grid cell.
+Symbolic algebra programs can be helpful when deriving methods for 
+extrapolating interior and boundary values to ghost cells.
 
 \section{\label{sec:summary} Summary} 
-While somewhat tedious, analyzing higher-order errors can suggest simple 
-modifications to numerical methods that greatly improve their accuracy.  
-Symbolic mathematics tools can be helpful to make this easier.
-
-Optimal time step selection is not applicable to all PDEs.  ...
+In this paper, we have presented a novel approach for constructing high-order
+finite difference methods for time dependent PDEs \emph{without} requiring
+the use of high-order stencils or high-order time integration schemes.  
+The basic idea is that the leading-order terms in the discretization error
+can be eliminated by carefully choosing the time step and possibly adding
+a few simple correction terms.  It is straightforward to derive both the 
+optimal time step and the correction terms by (1) taking advantage of the 
+regular structure of finite difference schemes and (2) making use of the
+PDE to relate terms in the discretization error.  The boost in the order
+of accuracy achieved when the optimal time step and correction terms is well
+worth the extra effort required to explicitly retain and analyze the 
+leading-order terms in the discretization error.
+
+Through many examples, we have demonstrated the utility of optimal time step
+selection to a wide range of problems.  We showed how OTS selection can be 
+used to very easily to obtain high-order accurate solutions to most linear 
+and semilinear PDEs in any number of space dimensions on both regular
+and irregular domains.  The examples illustrate the surprisingly high-order
+accuracy that can be achieved using only low-order discretizations for 
+spatial derivatives and simple forward Euler time integration.
+
+Optimal time step selection is a simple example of the more general problem
+of optimally selecting numerical parameters to boost the accuracy of 
+numerical methods.  Little research appears to have been done in this area,
+but, as OTS selection demonstrates, optimal selection of numerical parameters 
+can significantly impact on the accuracy of numerical methods.  To realize
+the full potential of numerical methods, both optimal time step selection 
+and optimal numerical parameter selection deserve further investigation.
 

### 2008-05-12 src/1d/diffusion_eqn/genFigs_DiffusionEqn1d_Src.m
OTSPDE: * For 1d diffusion equation, added support for comparing the error obtained   with and without correction term when OTS is used. * Minor modification to demo for fourth-order parabolic equation to test for   maximum stable time step.

@@ -62,9 +62,10 @@ t_final = 0.5;
 grid_sizes = [25 50 100 200 400 800];
 
 % allocate memory for errors
-err_FE_OTS = zeros(size(grid_sizes));
-err_FE     = zeros(size(grid_sizes));
-err_CN     = zeros(size(grid_sizes));
+err_FE_OTS         = zeros(size(grid_sizes));
+err_FE_OTS_no_corr = zeros(size(grid_sizes));
+err_FE             = zeros(size(grid_sizes));
+err_CN             = zeros(size(grid_sizes));
 
 % start clock for timing plot generation time
 t_start = cputime;
@@ -114,6 +115,17 @@ for i = 1:length(grid_sizes)
                                           t_init, t_final, ...
                                           debug_on, timing_on);
 
+    % solve diffusion equation using forward Euler with OTS but no
+    % correction term
+    disp('Forward Euler OTS (no correction)');
+    [u_FE_OTS_no_corr, u_exact, x, timing_data_FE_OTS_no_corr] = ...
+       solveDiffusionEqnForwardEulerOTSNoCorr1d(D, ...
+                                                source_term_type, ...
+                                                u_0, dudx_1, ...
+                                                dx, ...
+                                                t_init, t_final, ...
+                                                debug_on, timing_on);
+
     % solve diffusion equation using forward Euler
     disp('Forward Euler');
     [u_FE, u_exact, x, timing_data_FE] = ...
@@ -137,15 +149,17 @@ for i = 1:length(grid_sizes)
     % save solutions and timing data to MAT-files
     filename = sprintf('data_%d', N);
     save([data_dir, '/', filename], ... 
-         'u_FE_OTS', 'u_FE', 'u_CN', 'u_exact', 'x', ...
-         'timing_data_FE_OTS', 'timing_data_FE', ...
[…]

### 2008-05-12 src/1d/diffusion_eqn/solveDiffusionEqnForwardEulerOTSNoCorr1d.m

@@ -1,0 +1,224 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveDiffusionEqnForwardEulerOTSNoCorr1d() computes the solutions of the 
+% 1d diffusion equation 
+%
+%   u_t = D u_xx + f(x)
+%
+% on the domain 0 < x < 1 with a Dirichlet boundary condition imposed at
+% x = 0 and a Neumann boundary condition imposed at x = 1.  The numerical
+% solution is computed on a node-centered grid using forward Euler time 
+% integration with a second-order central difference approximation to the 
+% Laplacian.  The optimal time step of dx^2/(6D) is used but no correction
+% term is used.
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveDiffusionEqnForwardEulerOTSNoCorr1d(D, ...
+%                                              source_term_type, ...
+%                                              u_0, dudx_1, ...
+%                                              dx, ...
+%                                              t_init, t_final, ...
+%                                              debug_on, timing_on)
+%
+% Arguments:
+% - D:                   diffusion coefficient 
+% - source_term_type:    type of source term to use in computation.
+%                        0:  f = 0
+%                        1:  
+%
+%           f = 5*sin(3/2*pi*x) - 7*sin(15/2*pi*x) + 10*sin(21/2*pi*x)
+%
+% - u_0:                 Dirichlet boundary condition at x = 0
+% - dudx_1:              Neumann boundary condition at x = 1
+% - dx:                  grid spacing
+% - t_init:              initial time
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
[…]

### 2008-05-12 src/1d/fourth_order_parabolic/demo.m

@@ -21,6 +21,7 @@ t_final = 1e-6;
 % set dx and dt
 N = 100;
 dx = 1/N;
+dt_FE = dx^4*3/40;  % maximum stable time step for forward Euler
 dt_FE = dx^4/16;
 
 % solve diffusion equation using forward Euler with OTS

### 2008-05-13 papers/OTSPDE/optimal_time_steps_for_pdes.tex
OTSPDE:  first completed draft of paper.

@@
 % Manuscript started: 2007/09/01
-% First draft: 2008/05/??, Vitamin D, Inc.
+% First completed draft: 2008/05/12, Vitamin D, Inc.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 equations using only formally low-order finite difference stencils and
-time integration schemes.  By using an optimally chosen the time step for a 
-finite difference scheme (and possibly adding a few correction terms), we 
-are able to boost the order of accuracy of low-order finite difference 
-schemes.  Through simple numerical analysis arguments, we explain the origin 
-of the boost in accuracy and calculate the computational cost of the
-high-order methods obtained via optimal time step (OTS) selection.  We 
-demonstrate the utility of OTS selection on a wide array of classical PDEs
-in one and more space dimensions.   We find that OTS selection is applicable 
-to linear and semilinear PDEs and that it is easily extended to problems 
-on irregular domains. 
+time integration schemes.  We achieve high-order accuracy in a simple
+manner by using an optimally chosen time step (and possibly adding a few 
+correction terms to the finite difference scheme).  Through straightforward 
+numerical analysis arguments, we explain the origin of the boost in accuracy 
+and estimate the computational cost of the high-order methods obtained via 
+optimal time step (OTS) selection.  We demonstrate the utility of OTS 
+selection on a wide array of classical PDEs in one and more space dimensions
+and find that it is applicable to linear and semilinear PDEs on both regular 
+and irregular domains. 
 \end{abstract}
@@
 \markboth{KEVIN T. CHU}
-         {HIGH-ORDER ACCURATE FD SCHEMES VIA OPTIMAL TIME STEP SELECTION}
+         {HIGH-ORDER FINITE DIFF SCHEMES VIA OPTIMAL TIME STEP SELECTION}
 
@@
 
-High-order accurate methods that are simple combinations or modifications of 
-low-order schemes abound.  A few well-known examples include Richardson 
-extrapolation~\cite{richardson_1927,heath_book}, the modified nine-point 
-formula for the Poisson equation~\cite{iserles_book}, and the Lax-Wendroff 
-method for the advection equation~\cite{leveque_book_1992, leveque_book_2002,
-gko_book}.  Cancellation of higher-order errors (whether fortuitous or 
-intentional) is a key to principle underlying the superior accuracy of these 
-each of methods.
+There are many high-order accurate methods that are simple combinations or 
+modifications of low-order schemes.  A few well-known examples include 
+Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
+nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
+Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
+leveque_book_2002, gko_book}.  Cancellation of higher-order errors (whether 
+fortuitous or intentional) is a key principle underlying the superior 
+accuracy of these methods.
 
@@
 constraint is certainly sufficient for high-order accuracy, it is by no means 
-necessary.  This conventional approach for developing numerical methods for 
-PDEs has its roots in the classical numerical analysis of PDEs, which tends to 
-individually analyze the discretization of the terms in the PDE.  Within this 
+necessary.  This conventional approach for constructing numerical methods 
+has its roots in the classical numerical analysis of PDEs, which tends to 
+analyze the discretization of the terms in the PDE individually.  Within this 
 framework, it is natural to seek high-order accuracy by ensuring that the 
-numerical scheme is simultaneously high-order accurate for \emph{all} 
-independent variables.  Unfortunately, this common approach for developing 
-and analyzing numerical methods is often suboptimal because it neglects the 
-potential for achieving high order of accuracy using only simple finite 
-difference schemes based on compact, low-order stencils.
-
-Perhaps the greatest weakness of the conventional method for constructing 
-high-order numerical methods is the use of ``big-O" notation, which hides all 
-of the details about higher-order errors.  The information lost in the ``big-O" 
-notation eliminates the possibility of reducing numerical error by taking 
-advantage of relationships between the terms in the error.  The modified 
-nine-point formula for the Poisson equation (also known under the name 
-Mehrstellen)~\cite{iserles_book} and the Lax-Wendroff scheme for the 
-advection equation~\cite{leveque_book_1992, leveque_book_2002, gko_book} 
-demonstrate the power of explicitly retaining higher-order terms in the 
-discretization error when analyzing numerical schemes.  For both of these 
-methods, examination of higher-order terms allows low-order methods to be 
-transformed into high-order methods with only minor modifications to the 
-numerical scheme.
+numerical scheme is simultaneously high-order accurate for all independent 
+variables.  Unfortunately, this common approach for developing and analyzing 
+numerical methods is often suboptimal because it neglects the potential for 
+achieving high-order accuracy using only simple, low-order schemes.
+
+Perhaps the greatest weakness of the standard method for constructing 
+high-order numerical methods is the use of ``big-O" notation, which hides 
+the details of the discretization error.  The information lost in the 
+``big-O" notation eliminates the possibility of reducing numerical error by 
+taking advantage of relationships between the terms in the error.  The 
+modified nine-point formula for the Poisson equation~\cite{iserles_book} 
+and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
+leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
+higher-order terms in the discretization error when analyzing numerical 
+schemes.  For both of these methods, examination of higher-order terms allows 
+low-order methods to be transformed into higher order methods with only minor 
+modifications to the numerical scheme.
 
 The usual approach for developing numerical methods has another significant
-shortcoming -- it places no value on optimization of numerical parameters 
+shortcoming -- it places no value on the optimization of numerical parameters 
 (\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
@@
 optimizing the numerical parameters can significantly increase the accuracy
-of the numerical solution.  For example, in~\cite{zhao_2006} Zhao showed 
-that appropriately choosing the value of $\theta$ as a function of the time 
-step, $\dt$, in the theta-method\footnote{In~\cite{zhao_2006} refers 
-to $\theta$ as the weight factor and the theta-method as the single-step 
-trapezoidal method.}~\cite{iserles_book, gko_book} can decrease the error 
-of finite element solutions to the diffusion equation.
-
-In this paper, we explore a novel technique based on optimal time step (OTS)
-selection that can be used to construct high-order finite difference schemes 
-for time dependent PDEs using only formally low-order stencils.  The essential 
-idea is that a carefully chosen time step (and possibly the addition of
-a few correction terms) can eliminate low-order terms in the discretization 
-error, allowing a formally low-order method to deliver high-order accuracy.  
-The optimal time step is calculated through a straightforward analysis of the 
-discretization error that uses information from the PDE to relate the 
-leading-order terms.  Optimal time step selection combines the well-known 
-approach of using information provided by the PDE to reduce numerical error 
-with the less-common notion that optimal choice of numerical parameters can 
-boost the effectiveness of algorithmically simple and computationally 
-low-cost methods.
+of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
+that choosing an appropriate value for $\theta$ as a function of the time 
+step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
+the weight factor and the theta-method as the single-step trapezoidal 
+method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
+element solutions to the diffusion equation.
+
+In this paper, we explore a novel technique for constructing high-order 
+finite difference methods for time dependent PDEs using only formally 
+low-order stencils and time integration schemes.  The basic idea is that a 
+carefully chosen time step (and possibly the addition of a few correction 
+terms) can eliminate low-order terms in the discretization error, allowing a 
+formally low-order method to deliver high-order accuracy.  The optimal time 
+step is calculated in a straightforward manner by using knowledge of the 
+PDE to eliminate the leading-order terms in the discretization error.  
+Optimal time step selection combines the well-known approach of using 
+the PDE to reduce numerical error with the less-common notion that optimal 
+choice of numerical parameters can boost the effectiveness of algorithmically 
+simple and computationally low-cost methods.
 
@@
 clever choice of time step can be used to dramatically reduce the error is 
-rarely considered.  In general, little research seems to have been done on 
-optimizing the parameters of numerical methods (\eg time step and grid 
-spacing) to improve accuracy.  To the best of the authors knowledge, Zhao's 
-method for choosing an optimal value of $\theta$ in the theta-method is the 
-only work that attempts to improve accuracy by modifying numerical parameters.
-While he focuses solely on the optimal choice of $\theta$ for a fixed time 
-step, his approach can be viewed as a form of optimal time step selection by 
-reversing the roles of $\theta$ and the time step.
-
-Before moving on, it is worth spending a few moments to draw the distinction
-between optimal time step selection and traditional adaptive time stepping 
+rarely considered.  In general, little research seems to have been done in
+the area of optimizing the parameters of numerical methods (\eg time step and 
+grid spacing) to improve accuracy.  To the best of the author's knowledge, 
+Zhao's method for choosing an optimal value of $\theta$ in the theta-method is 
+the only work that attempts to improve accuracy by modifying numerical 
+parameters.  While he focuses solely on the optimal choice of $\theta$ for a 
+fixed time step, his approach can be viewed as a form of optimal time step 
+selection by reversing the roles of $\theta$ and the time step.
+
+It is worth spending a few moments to draw the distinction between optimal 
+time step selection and traditional adaptive time stepping 
 techniques~\cite{iserles_book,??} that are used in the context of the method 
@@
 uses information from the PDE to choose the time step.  Thus, optimal time 
-step selection has the potential to produce superior results compared with 
+step selection can potentially produce superior results compared with 
 traditional adaptive time stepping approaches for integrating the 
@@
 As a technique for constructing numerical methods, optimal time step selection 
-has several desireable features.  Like any technique that leads to high-order 
+has several desirable features.  Like any technique that leads to high-order 
 methods, optimal time step selection produces schemes that greatly reduce the 
@@
 rectangular domains; irregular domains are straightforward to handle by
-appropriately setting the values in ghost cells~\cite{fedkiw_1999, 
-osher_fedkiw_book, gibou_2005}.  Optimal time step selection is even 
-beneficial when using finite difference schemes to solve some nonlinear PDEs.  
-While there are certainly limitations to optimal time step selection, we will 
-see that there are a many problems where it is useful.
+appropriately setting the values in ghost cells~\cite{gibou_2005, fedkiw_1999, 
+osher_fedkiw_book}.  Optimal time step selection is even beneficial when 
+using finite difference schemes to solve some nonlinear PDEs.  While there 
+are certainly limitations to optimal time step selection, we will see that 
+there are many problems where it is useful.
 
@@
 \section{\label{sec:optimal_time_step_selection} 
-         Optimal Time Step Selection}
+         Optimal Time Step (OTS) Selection}
 Optimal time step selection is not by itself a method for constructing 
@@
 PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
-is well-posed.  Because we will be boosting the accuracy of the FD 
-approximation by optimizing the time step, there is no reason to explicitly
-construct the scheme to be high-order.  In fact, using a high-order FD scheme 
-only complicates the analysis required to calculate the optimal time 
-step.
+is well-posed.  Because we will be boosting the accuracy of the finite 
+difference approximation by optimizing the time step, there is no reason to 
+explicitly construct the scheme to be high-order.  In fact, using a high-order 
+finite difference scheme only complicates the analysis required to calculate 
+the optimal time step.
 
 To compute the optimal time step, we first analyze the discretization errors 
-(both spatial and temporal) for the FD approximation, explicitly keeping track 
-of the leading-order terms.  In contrast to conventional error analysis, we 
-do not sweep all of the errors under ``big-O'' notation.  Rather, we take 
-advantage of the fact that derivation of the leading-order terms in the 
-error is straightforward for FD schemes through the use of Taylor series 
-expansions.  Because each term in the error is proportional to a partial 
-derivative of the solution, we can attempt to use the PDE to relate the 
-leading-order terms to each other.  If we are fortunate (which we are
-for many finite difference schemes, this careful analysis yields a direct 
-relationship between the leading-order spatial and temporal errors that allows 
-them to be combined as a single term of the form:
+(both spatial and temporal) for the finite difference approximation, 
+explicitly keeping track of the leading-order terms.  In contrast to 
+conventional error analysis, we do not sweep all of the errors under ``big-O'' 
+notation.  Rather, we take advantage of the fact that derivation of the 
+leading-order terms in the error is straightforward for finite difference
+schemes through the use of Taylor series expansions.  Because each term in 
+the error is proportional to a partial derivative of the solution, we can 
+attempt to use the PDE to relate the leading-order terms to each other.  If we 
+are fortunate (which we are for many finite difference schemes), this careful 
+analysis yields a direct relationship between the leading-order spatial and 
+temporal errors that allows them to be combined as a single term of the form:
 \beq
@@
 \emph{optimal time step} for the finite difference scheme.  Using the optimal
-time step for time stepping yields a numerical solution with the maximal order 
-of accuracy for a given FD scheme.
-
-An important class of FD schemes for time dependent PDEs are those based
-on the method of lines.   When the method of lines approach is used for 
-(\ref{eq:model_PDE}), $P$ takes a particularly simple 
+time step for time stepping yields a numerical solution with a boosted order 
+of accuracy for a given finite difference scheme.
+
+An important class of finite difference schemes for time dependent PDEs are 
+those based on the method of lines.   When the method of lines approach is 
+used for (\ref{eq:model_PDE}), $P$ takes a particularly simple 
 form:
@@
 where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
-errors, and $\alpha$ and $\beta$ are positive constants that depend on the 
-details of the PDE and FD scheme.  Setting $P = 0$, we find that the optimal 
-time step is given by\footnote{The $\dt = 0$ root of $P$ is discarded 
-because it is not numerically meaningful.}
+discretization errors, and $\alpha$ and $\beta$ are positive constants that 
+depend on the details of the PDE and finite difference scheme.  Setting 
+$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
+root of $P$ is discarded because it is not numerically meaningful.}
 \beq
@@
 satisfies a stability constraint.
-
-Thus far, we have completely ignored the question of whether the FD scheme is 
-stable when the time step is set to $\dto$.  Not surprisingly, OTS selection 
-loses its value, when the optimal time step is greater than the smallest 
-stable time step for an explicit scheme.  Fortunately, we can usually switch 
-to an implicit (or semi-implicit) time integration scheme with the same order 
-of accuracy for which the optimal time step is stable.  Note that the 
-depending on how the original FD scheme is modified, a new optimal time step 
-may need to be calculated.
 
@@
 where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
-$F({\mathbf u})$ is the FD approximation of the spatial derivative 
-operator acting on ${\mathbf u}$.  The second term in the local error arises 
-directly from temporal discretization errors associated with the numerical 
-scheme used to integrate the system of ODEs in time.  The first term arises 
-because $F$ is only an approximation of the spatial derivative operators in 
-the PDE.  As a result, the system of ODEs~(\ref{eq:method_of_lines}) has an 
-error due to the spatial discretization at \emph{every} point in time.  For 
-a spatial discretization error of order $r$, the error in 
-(\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the error 
-accumulated during each time step due to spatial discretization error is 
+$F({\mathbf u})$ is the finite difference approximation of the spatial 
+derivative operator acting on ${\mathbf u}$.  The second term in the local 
+error arises directly from temporal discretization errors associated with the 
+numerical scheme used to integrate the system of ODEs in time.  The first 
+term arises because $F$ is only an approximation of the spatial derivative 
+operators in the PDE.  As a result, the system of 
+ODEs~(\ref{eq:method_of_lines}) has an error due to the spatial discretization 
+at \emph{every} point in time.  For a spatial discretization error of order 
+$r$, the error in (\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the 
+error accumulated during each time step due to spatial discretization error is 
 $O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
 is solved numerically.  Because the method of lines is so frequently used to 
-construct FD approximations for time dependent PDEs, we will focus solely on 
-FD schemes of this type for the remainder of the paper to keep the discussion 
-simple.  It is straightforward to extend the results we obtain to more general 
-FD schemes.
-
-To analyze the accuracy of the FD scheme for~(\ref{eq:model_PDE}) when the 
-optimal time step is used (for a given grid spacing), we examine the 
+construct finite difference approximations for time dependent PDEs, we will 
+focus solely on finite difference schemes of this type for the remainder of 
+the paper to keep the discussion simple.  It is straightforward to extend the 
+results we obtain to more general finite difference schemes.
+
+To analyze the accuracy of the finite difference scheme 
+for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
 higher-order terms in the discretization error following the conventional 
-approach (\ie ``big-O'' notation).  Elimination of the leading term in the 
+approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
 discretization error leaves a local error of the form 
@@
 solution over extended intervals of time by using the heuristic argument
-that the global error is equal to the local discretization 
-error divided by $\dt$.  Using heuristic argument, which is rigorously 
-justified in many texts on numerical methods such as~\cite{gko_book}, we 
-find that the global error of the FD scheme using an optimal time step is 
+that the global error is equal to the local error divided by $\dt$.  Using 
+this heuristic argument, which is rigorously justified in many texts on 
+numerical methods such as~\cite{gko_book}, we find that the global error of 
+the finite difference scheme using an optimal time step is 
 \beq
@@
 \eeq
-Observe that the global error for the FD scheme when an arbitrary time
-step is used has exactly the same form as~(\ref{eq:global_error_ots}) with
-$p$ and $q$ replaced by $r$ and $s$, respectively.  Thus, we see that 
-using an optimal time step boosts the accuracy of the FD scheme.
+Observe that the global error for the finite difference scheme when an 
+arbitrary time step is used has exactly the same form 
+as~(\ref{eq:global_error_ots}) with $p$ and $q$ replaced by $r$ and $s$, 
+respectively.  Thus, because $p > r$ and $q > s$, we see that using an optimal 
+time step boosts the overall accuracy of the finite difference scheme.
 
@@
 in order to derive the error estimate.  In other words, there is really only 
-one parameter, $\dx$, that controls the numerical accuracy of the FD scheme.  
-This important point will be discuss in further detail in the next section.
+one parameter, $\dx$, that controls the numerical accuracy of the finite 
+difference scheme.  This important point will be discuss in further detail in 
+the following section.
 
@@
 interesting and important to understand how the error depends on both the 
-grid spacing $\dx$ and the time step $\dt$, in practice the accuracy 
+grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
 of a numerical scheme is always controlled by only one of the two numerical
@@
 Even though spatial and temporal orders of accuracy for a numerical method
-are formally separate, one will alway dominate for any given choice of 
-$\dx$ and $\dt$.  For example, when solving the diffusion  equation 
+are formally separate, one will always dominate for a given choice of 
+$\dx$ and $\dt$.  For example, when solving the 1D diffusion equation 
 using the backward Euler method for time integration with a second-order 
@@
 
-When there are constraints on the numerical parameters, there is often 
-less freedom in choosing the controlling parameter.  For instance, if we 
-solve the diffusion equation using a forward Euler time integration scheme 
-with a second-order central difference stencil for the Laplacian, stability
+When there are constraints on the numerical parameters, there is less freedom 
+in choosing the controlling parameter.  For instance, if we solve the 
+1D diffusion equation using a forward Euler time integration scheme with a 
+second-order central difference stencil for the Laplacian, stability
 considerations require that we choose $\dt = O(\dx^2)$.  
@@
 discretization scheme and grid spacing so that a desired level of accuracy 
-is achieved and then select a time integration scheme and time step which 
-ensures that the temporal errors do not overcome the spatial errors.  
+is achieved and then select the time integration scheme and time step in 
+such a manner that the temporal errors do not overcome the spatial errors.  
 This methodology explains the tendency to evaluate numerical schemes for
@@
 the error of numerical methods for time dependent PDEs.  Because the optimal 
-time step is calculated explicitly as a function of the grid spacing, the 
-technique of optimal time step selection places an even more restrictive 
-constraint on $\dt$ than stability considerations.  Thus, like numerical
-methods with stability constraints, the accuracy of an FD scheme using 
-the optimal time step is completely controlled by the grid spacing.  The 
-order of accuracy of a FD scheme advanced in time using $\dto$ 
-depends on the functional dependence the optimal time step on $\dx$.
-For the method of lines, we know that $\dto = O(\dx^{r/s})$, 
-so the global error derived in the previous section reduces to 
+time step is a function of the grid spacing, the technique of optimal time 
+step selection places an even more restrictive constraint on $\dt$ than 
+stability considerations.  Thus, like numerical methods with stability 
+constraints, the accuracy of a finite difference scheme using an optimal time 
+step is completely controlled by the grid spacing.  For the method of lines, 
+we know that $\dto = O(\dx^{r/s})$, so the global error derived in the 
+previous section reduces to 
 \beq
@@
 
-\subsubsection*{\label{sec:computational_performance} 
-                Computational Performance}
-For many problems, $r/s > 1$ and yields an optimal time step that is on the 
-order of the time step allowed by the stability constraint for an explicit 
-method.  As a result, using the optimal time step to advance the FD scheme in 
-time may seem to be overly restrictive and lead to poor computational 
-performance.  However, these apparent drawbacks are more than compensated by 
-the boost in the order of accuracy.  For example, when solving the 1D 
-diffusion equation using a forward Euler time integration scheme with a 
-second-order central difference stencil for the Laplacian, we obtain a 
-fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
-Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
-scheme that is computationally cheaper than traditional methods that use 
-the same low-order FD stencils.  The performance gain for problems in higher 
-spatial dimensions is even more impressive.  As we can see in 
-Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
-simple forward Euler scheme with the optimal time step and the Crank-Nicholson
-method grows with the spatial dimension of the problem.
-
-\begin{table}[tbh]
+% TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
+% sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
+\begin{table}[tb]
 \caption{\label{tab:comp_perf_vs_err} 
-   Computational cost of various FD schemes for numerically solving the 1D 
+   Computational cost of various finite difference schemes for the 1D 
    diffusion equation as a function of the global numerical error $e$.
-   For all FD schemes, the standard second-order central difference 
+   For all of the schemes, the standard second-order central difference 
    stencil is used to discretize the Laplacian, and the numerical 
@@
 
-\begin{table}[tbh]
+\subsubsection*{Stability Considerations}
+Thus far, we have not considered the issue of numerical stability of finite
+difference schemes when the time step is set to $\dto$.  For explict time 
+integration schemes, OTS selection loses its value when the optimal time step 
+is greater than the smallest stable time step.  In this situation, we can 
+usually switch to an implicit (or semi-implicit) time integration scheme with 
+the same order of accuracy for which the optimal time step is stable.  Note 
+that a new optimal time step may need to be calculated for the modified finite
+difference sceme.
+
+\subsubsection*{\label{sec:computational_performance} 
+                Computational Performance}
+For many problems, the optimal time step is on the order of the time step 
+allowed by the stability constraint for an explicit method.  
+As a result, using the optimal time step to advance the finite
+difference scheme in time may appear to be overly restrictive and lead to poor 
+computational performance.  However, these apparent drawbacks are more than 
+compensated by the boost in the order of accuracy.  For example, when solving 
+the 1D diffusion equation using a forward Euler time integration scheme with a 
+second-order central difference stencil for the Laplacian, we obtain a 
+fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
+Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
+scheme that is computationally cheaper than traditional methods that use 
+the same low-order finite difference stencils.  The performance gain for 
+problems in higher spatial dimensions is even more impressive.  As we can see 
+in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
+simple forward Euler scheme with the optimal time step and the Crank-Nicholson
+method grows with the spatial dimension of the problem.
+
+\begin{table}[tb]
 \caption{\label{tab:comp_perf_vs_dim}
@@
    of the global numerical error $e$.
-   For both FD schemes, second-order, isotropic stencils are used to discretize 
-   the Laplacian.  For a problem in $d$ spatial dimensions, the memory 
-   requirement for each scheme is estimated to be $O \left(1/\dx^d \right)$, 
-   and the computation time is approximated as 
-   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, which assumes that we 
-   are using an efficient $O(N) = O \left( 1/\dx^d \right)$ algorithm to 
-   invert the matrices that arise for the Crank-Nicholson method.
+   For both schemes, second-order, isotropic stencils are used to discretize 
+   the Laplacian.  For a problem in $d$ space dimensions, the memory 
+   requirement and computation time for each scheme are estimated to be 
+   $O \left(1/\dx^d \right)$ and 
+   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, respectively, which 
+   assumes that we are using an efficient 
+   $O(N) = O \left( 1/\dx^d \right)$ algorithm to invert the matrices that 
+   arise for the Crank-Nicholson method.
 }
@@
 For more general PDEs, the approach of carefully choosing the time step to 
-eliminate leading order terms in the error is still useful, but it needs to 
-be augmented to achieve a boost in the order of accuracy.  In addition to 
-using the optimal time step, correction terms may need to be added to the 
-finite difference scheme.  In this section, we discuss these correction terms 
-in the context of general constant coefficient, linear PDEs and semilinear
-PDEs.
+eliminate leading order terms in the error continues to be useful, but it 
+needs to be augmented to achieve the same improvement in the order of accuracy.
+In addition to using an optimal time step, correction terms may need to be 
+added to the finite difference scheme.  In this section, we discuss these 
+correction terms in the context of general constant coefficient, linear PDEs 
+and semilinear PDEs.
 
@@
   \frac{\partial u}{\partial t} = 
-  \left( \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} \right) + f(x,t)
+  \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} + f(x,t)
   \label{eq:linear_PDE}.
 \eeq
-Because it is unlikely that all of the dominant terms in the error can be
-combined into a single term, the leading-order discretization error is
-a sum of several terms:
+Even if we use the PDE to relate terms in the discretization error, it is
+unlikely that all of the dominant terms in the error can be combined into 
+a single term.   As a result, the leading-order discretization error is
+generally a sum of several terms:
 \bea
@@
 Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
-operators because any that arise can be eliminated by using the PDE.  Also, 
+operators because any that arise can be eliminated using the PDE.  Also, 
 there is no explicit spatial discretization error of order $r$ that originates 
@@
 finite differences applied to the solution $u$ at the current time.
-The last term could also be approximated using finite differences, or,
-alternatively, it could be calculated analytically because the source terms 
-are usually known functions.  With the addition of these correction terms, 
+The last term can be computed using either a finite difference approximation 
+or calculating it analytically.  With the addition of these correction terms, 
 the leading-order discretizaton error is completely eliminated, and the
@@
 approximate the vestigial errors must be chosen carefully.  Special attention 
-is required to ensure that the FD stencils are of sufficiently high-order; 
-otherwise, we will lose the accuracy gained from using an optimal time step 
-for the FD scheme.  Since the specific form of the correction terms are 
-PDE-dependent, we defer further discussion to 
-Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim} where 
-we apply optimal time step selection to several specific PDEs.
+is required to ensure that the finite difference stencils are of sufficiently 
+high-order.  Otherwise, we will lose the accuracy gained from using an optimal 
+time step and adding the correction terms to the finite difference scheme.  
+Since the specific form of the correction terms are PDE-dependent, we defer 
+further discussion to Sections~\ref{sec:applications_1d} 
+and~\ref{sec:applications_multidim} where we apply optimal time step selection 
+to several specific PDEs.
 
@@
 where $F$ is an arbitrary function of the lower-order spatial derivatives
-of $u$ and $f(x,t)$ is a source term.  While, the coefficient on the 
+of $u$ and $f(x,t)$ is a source term.  While the coefficient on the 
 leading-order spatial derivative is generally allowed to be function of 
@@
 term in~(\ref{eq:semilinear_PDE}), and $G$ is a spatio-temporal differential 
-operator that acts on $f$.  To compute the optimal time step and boost
-the accuracy of the finite difference scheme, we proceed in the
-same manner as for general constant coefficient, linear PDEs.
+operator that acts on $f$.  With the discretization error written in the 
+form (\ref{eq:leading_order_error_semilinear_PDE_MOL}), the optimal time 
+step and correction terms can be computed by proceeding in the same manner 
+as for general constant coefficient, linear PDEs.
 
@@
 
-\subsection{OTS Selection in Multiple Space Dimensions
-                \label{sec:ots_higher_spatial_dims}}
-Optimal time selection is applicable to PDEs in any number of spatial 
+\subsection{\label{sec:ots_higher_spatial_dims}
+            OTS Selection in Multiple Space Dimensions}
+Optimal time selection is applicable to PDEs in any number of space
 dimensions.  It is straightforward to generalize the approach outlined in 
@@
 problems in higher dimensions generally require a more careful choice of 
-the finite difference stencil used to approximate spatial derivatives in 
-order to be able to combine leading-order terms in the discretization error, 
+the finite difference stencil used to approximate spatial derivatives,
 there are no fundamental barriers that prevent the use of optimal time step 
@@
 The geometry of the domain is an important feature of problems in more than 
-one spatial dimension because it can significantly impact the way boundary 
+one space dimension because it can significantly impact the way boundary 
 conditions are imposed.  For problems on regular domains, boundary conditions 
@@
 difficulties.  Assuming that a `ghost cell' or `ghost value' 
-approach~\cite{fedkiw_1999, osher_fedkiw_book, gibou_2005} 
-is used to impose boundary conditions, the construction and analysis of finite 
+approach~\cite{gibou_2005, fedkiw_1999, osher_fedkiw_book} is used to impose 
+boundary conditions, the construction and analysis of finite 
 difference schemes and calculation of optimal time steps is identical for 
@@
 spatial directions.  For some problems, the ratio of the grid spacing in
-different coordinate directions may need to be carefully chosen to even 
-be able to calculate an optimal time step.
+different coordinate directions may need to be carefully chosen for an 
+optimal time step to even exist.
 
@@
 systems of PDEs where the optimal time step is the same across the individual 
-PDEs, it is not clear that the method yields any benefit in general.
+PDEs, it is not clear that OTS selection yields any benefit in general.
 
@@
          Application to PDEs in One Space Dimension}
-Because there are no geometry or anistropy considerations, optimal time step 
+Because there are no geometry or anisotropy considerations, optimal time step 
 selection is very useful for boosting the accuracy of many finite difference 
@@
 where $A$ is the flow speed, which we take to be positive for convenience.  
-For an initial condition of the form $u(x,0) = f(x)$, it is easy to use the 
-method of characteristics to show that the analytical solution is $f(x-At)$.  
-While it is unlikely that one would actually use a numerical method to 
-solve~(\ref{eq:advection_eqn_1d}), it provides an interesting example where 
-optimal time step selection dramatically improves the accuracy of the 
-numerical solution.
+For an initial condition of the form $u(x,0) = f(x)$, it is easy to show 
+that the analytical solution is $f(x-At)$.  While it is unlikely that one 
+would actually use a numerical method to solve~(\ref{eq:advection_eqn_1d}), 
+it provides an interesting example where optimal time step selection 
+dramatically improves the accuracy of the numerical solution.
 
@@
 where superscripts and subscripts denote the time and space indices, 
-respectively.  It is straightforward to show that method is formally 
+respectively.  It is straightforward to show that the method is formally 
 first-order in both space and time.
@@
 for the advection equation, it is instructive to retain all of them.  
-Using Taylor series to expand the true solution of the the PDE at time step
+Using Taylor series to expand the true solution of the PDE at time step
 $(n+1)$, we find that
@@
   + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\pt^k} \right|_j
+       \frac{\partial^k \tu^n}{\pt^k} 
   \label{eq:advection_eqn_1d_time_err}
@@
 where $\tu$ denotes the true solution of the PDE and 
-$\tu^n_j \equiv \tilde{u}(j\dx, n\dt)$.  Similarly, Taylor series expansion 
-of the upwind finite difference approximation to the spatial derivative 
-yields
+$\tu^n_j \equiv \tilde{u}(j\dx, n\dt)$.  Similarly, Taylor series expansion of 
+the upwind finite difference approximation to the spatial derivative yields
 \bea
   \frac{\tu^{n}_j - \tu^{n}_{j-1}}{\dx} = 
-  \left. \frac{\partial \tu}{\px} \right|_j
+  \frac{\partial \tu}{\px} 
   - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\px^k} \right|_j.
+       \frac{\partial^k \tu^n}{\px^k}.
   \label{eq:advection_eqn_1d_space_err}
 \eea
+In both of the above equations, note that we have let context implicitly 
+determine the time ($t = t_n = n\dt$) and location ($x = x_j = j\dx$) where 
+partial derivatives of $\tu$ should be evaluated.  For convenience, we shall 
+continue to follow this convention throughout the remainder of the paper.
+
 Next, we use the PDE~(\ref{eq:advection_eqn_1d_FD_scheme}) to replace all of 
@@
   + \sum_{k=1}^\infty \frac{\left( -A \dt \right)^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\px^k} \right|_j.
+       \frac{\partial^k \tu^n}{\px^k}.
   \label{eq:advection_eqn_1d_time_err_modified}
 \eea
-Finally, by combining~(\ref{eq:advection_eqn_1d_time_err_modified}) and
+Finally, by combining~(\ref{eq:advection_eqn_1d_time_err_modified}) 
 and~(\ref{eq:advection_eqn_1d_space_err}) with the finite difference 
@@
     &+& A \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
-        \left. \frac{\partial^k \tu^n}{\px^k} \right|_j
-        \left[ \dx^{k-1} - \left( A \dt \right)^{k-1} \right].
+        \frac{\partial^k \tu^n}{\px^k} 
+        \left( \dx^{k-1} - \left( A \dt \right)^{k-1} \right).
   \label{eq:advection_eqn_1d_err_eqn}
 \eea
-As expected, the error $e$ satifies the same finite difference approximation 
-as the solution with an additional source term which is the discretization 
-error for the scheme.
+As expected, the error $e$ satisfies the same finite difference approximation 
+as the solution with a source term which is the discretization error for the 
+scheme.
 
@@
 \beq
-  \left( \left. \frac{A}{2} \frac{\partial^2 \tu^n}{\px^2} \right|_j \right)
-  \left[ \dx - A \dt \right] \dt,
+  \left( \frac{A}{2} \frac{\partial^2 \tu^n}{\px^2} \right)
+  \left( \dx - A \dt \right) \dt,
 \eeq
@@
 \subsection{Diffusion Equation}
-For an example where optimal time step selection yields a more typical boost 
+As an example where optimal time step selection yields a more typical boost 
 in the order of accuracy, we turn our attention to the diffusion equation:
@@
   \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
-  \left. \frac{\partial^2 \tu^n}{\px^2} \right|_j
-  + \frac{\dx^2}{12} \left. \frac{\partial^4 \tu^n}{\px^4} \right|_j
+  \frac{\partial^2 \tu^n}{\px^2} 
+  + \frac{\dx^2}{12} \frac{\partial^4 \tu^n}{\px^4} 
   + O(\dx^4)
@@
 From~(\ref{eq:diffusion_eqn_1d_err_eqn}), we see that the optimal time step 
-is given by $\dto = \dx^2/6D$.  Using~(\ref{eq:global_error_ots}),
-the order of the scheme with this choice of time step is 
-$O(\dx^4) + O(\dt^2) = O(\dx^4)$; we have boosted the accuracy of original 
+is given by $\dto = \dx^2/6D$.  Using~(\ref{eq:global_error_ots}) or
+(\ref{eq:global_error_ots_simplified}), the order of the scheme with this 
+choice of time step is $O(\dx^4)$; we have boosted the accuracy of original 
 scheme from second to fourth order.  Note that $\dto$ is stable, so there
@@
 
-To demonstrate the improvement in accuracy of the forward Euler scheme when 
-the optimal time step is used, we compare the numerical errors of three finite 
-difference schemes for the diffusion equation: forward Euler with optimal
-time step, forward Euler with suboptimal time step, and Crank-Nicholson.  
-As we can see in Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward 
-Euler with optimal time step is fourth-order accurate while both forward Euler 
-with suboptimal time step and Crank-Nicholson are only second-order accurate.  
-Using the forward Euler scheme with an optimal time step has a clear 
-advantage over both of the two other schemes.  Even with as few as 20 grid 
-points, the error in solution computed by the forward Euler scheme with an 
-optimal time step is orders of magnitude smaller.
-
-\begin{figure}[htb]
+\begin{figure}[tb]
 \begin{center}
@@
 \caption{$L^\infty$ error as a function of number of grid points for three 
-finite difference schemes that solve the diffusion equation: forward Euler 
+finite difference schemes that solve the 1D diffusion equation: forward Euler 
 with optimal time step (circles), forward Euler with suboptimal time step
-(squares), and Crank-Nicholson (diamonds).  All three schemes use the 
-standard second-order, central finite difference approximation for the 
-Laplacian.  Using each method, we solve the diffusion equation with $D = 0.5$ 
-on the domain $0 < x < 1$ subject to the boundary conditions 
-$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
-and the initial condition
+$\dt = 2\dx^2/3D$ (squares), and Crank-Nicholson with time step 
+$\dt = \dx/16 D$ (diamonds).  All three schemes use the standard second-order, 
+central finite difference approximation for the Laplacian.  Using each method, 
+we solve the diffusion equation with $D = 0.5$ to time $t = 0.1$ on the 
+domain $0 < x < 1$ subject to the boundary conditions $u(0,t) = 1$ and 
+$u_x(1,t) = 1/2$ and the initial condition
 $u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
@@
 \end{figure}
+
+To demonstrate the improvement in accuracy of the forward Euler scheme when 
+the optimal time step is used, we compare the numerical errors of three finite 
+difference schemes for the diffusion equation: forward Euler with optimal
+time step, forward Euler with suboptimal time step, and Crank-Nicholson.  
+As we can see in Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward 
+Euler with optimal time step is fourth-order accurate while both forward Euler 
+with suboptimal time step and Crank-Nicholson are only second-order accurate.  
+Using the forward Euler scheme with an optimal time step has a clear 
+advantage over both of the other schemes.  Even with as few as 25 grid 
+points, the error in solution computed by the forward Euler scheme with an 
+optimal time step is orders of magnitude smaller.
 
@@
 with optimal time step will still outperform the Crank-Nicholson scheme 
-because the latter requires the solution of at least one linear system of
-equation for each time step.
+because the latter requires the use of a wider stencil and the solution of 
+at least one linear system of equation for each time step.
 
@@
 where the derivatives of $f$ may be evaluated analytically or using 
-sufficiently high order finite difference stencils.
-
-In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of 
-the OTS forward Euler scheme with and without the correction terms.  As we can 
-see, OTS forward Euler with correction terms is fourth-order accurate while 
-OTS forward Euler without correction terms is only second-order accurate.  
-These results clearly demonstrate the importance of the correction terms for
-achieving high-order accuracy.
-
-\begin{figure}[htb]
+sufficiently high-order finite difference stencils.
+
+\begin{figure}[tb]
 \begin{center}
@@
 $f(x) = 5 \sin(3 \pi x/2) - 7 \sin(15 \pi x/2) + 10 \sin(21 \pi x/2)$
-on the domain $0 < x < 1$ subject to the boundary conditions 
-$u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
+to time $t = 0.5$ on the domain $0 < x < 1$ subject to the boundary 
+conditions $u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
 and the initial condition
@@
 
+In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of 
+the OTS forward Euler scheme with and without the correction terms.  As we can 
+see, OTS forward Euler with correction terms is fourth-order accurate while 
+OTS forward Euler without correction terms is only second-order accurate.  
+These results clearly demonstrate the importance of the correction terms for
+achieving high-order accuracy.
+
 \subsection{Advection-Diffusion Equation}
@@
   + \dt 
-    \left( D \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} 
-         + A \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right).
+    \left( D \left[ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
+         + A \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] \right).
   \label{eq:ADE_1d_FD_scheme}
@@
 $\dx \rightarrow 0$).
+
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.35}{\includegraphics{figures/adv_diff_eqn_1d_error_vs_N}} 
+\caption{$L^\infty$ error as a function of number of grid points for 
+two finite difference schemes that solve the advection-diffusion 
+equation: forward Euler with optimal time step and correction term (circles) 
+and forward Euler with suboptimal time step $\dt = 2\dx^2/3D$ and no correction 
+terms (squares).  Both schemes use second-order, central finite difference 
+approximations for the advection and diffusion terms.
+Using both methods, we solve the advection-diffusion equation with $D = 2$
+and $A = 5$ to time $t = 1$ on the domain $-10 < x < 10$ subject to the 
+initial condition $u(x,0) = \exp(-x^2/4)$.  The boundary conditions are set 
+using the analytical solution 
+$u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
+where $T = Dt+1$.
+}
+\label{fig:ADE_1d_error}
+\end{center}
+\end{figure}
 
@@
   \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
-  \left. \frac{\partial \tu^n}{\px} \right|_j
-  + \frac{\dx^2}{6} \left. \frac{\partial^3 \tu^n}{\px^3} \right|_j
+  \frac{\partial \tu^n}{\px} 
+  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
   + O(\dx^4).
@@
 
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
+\ \ \ \ \
+\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
+\caption{Comparison of numerical solutions for the 1D advection-diffusion
+equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
+optimal time step and correction terms (left) and forward Euler time 
+stepping with suboptimal time step $\dt = 2 \dx^2/3 D$ (right).  
+Both figures show the numerical solution at $t = 1$ computed on the domain 
+$-10 < x < 10$ using 50 grid points.  For details about the initial/boundary 
+conditions and physical parameters used to compute these solutions, see the
+caption for Figure~\ref{fig:ADE_1d_error}.
+}
+\label{fig:ADE_1d_solns}
+\end{center}
+\end{figure}
+
 As we discussed in Section~\ref{sec:ots_linear_pde}, boosting the order
 of accuracy for general linear, constant coefficient PDEs requires both 
-use of an optimal time step and addition of correction terms.  
+the use of an optimal time step and the addition of correction terms.  
 In this case, the optimal time step is $\dto = \dx^2/6D$ and the
@@
 scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
-Figure~\ref{fig:ADE_1d_error} illustrates the improved accuracy of the finite 
-difference scheme~(\ref{eq:ADE_1d_FD_scheme}) when the optimal time step and 
-correction term are used.
-
-\begin{figure}[htb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/adv_diff_eqn_1d_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for 
-two finite difference schemes that solve the advection-diffusion 
-equation: forward Euler with optimal time step and correction term (circles) 
-and forward Euler with suboptimal time step and no correction terms (squares).  
-Both schemes use second-order, central finite difference approximations for 
-the advection and diffusion terms.
-Using both methods, we solve the advection-diffusion equation with $D = 2$
-and $A = 5$ on the domain $-10 < x < 10$ subject to the initial condition
-$u(x,0) = \exp(-x^2/4)$.  The boundary conditions are set using the
-analytical solution 
-$u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
-where $T = Dt+1$.
-}
-\label{fig:ADE_1d_error}
-\end{center}
-\end{figure}
-
-\begin{figure}[htb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D advection-diffusion
-equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step (right).  The latter scheme uses a 
-time step $\Delta t = \Delta x^2 / 3 D$.  Both figures show the numerical
-solution at $t = 1$ computed on the domain $-10 < x < 10$ using 50 grid points.
-The diffusion constant $D$ and the flow speed $A$ are set to $2$ and $5$,
-respectively.
-The initial conditions are taken to be $u(x,0) = \exp \left( -x^2/4 \right)$
-and the boundary conditions are imposed exactly by using the analytical 
-solution $u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
-where $T = Dt+1$.
-}
-\label{fig:burgers_1d_solns}
-\end{center}
-\end{figure}
+Figures~\ref{fig:ADE_1d_error} and \ref{fig:ADE_1d_solns} illustrate the 
+improved accuracy of the finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) 
+when the optimal time step and correction term are used.
 
 \subsubsection*{First-order Upwind Discretization of Advection Term}
-It is common to use upwind discretizations as 
-in~(\ref{eq:advection_eqn_1d_FD_scheme}) for advection terms that arise
+It is common to use upwind discretizations such 
+as~(\ref{eq:advection_eqn_1d_FD_scheme}) for advection terms that arise
 in PDEs.  If instead of a central difference approximation, we use a
@@
 \eea
-Thus, we can see that for a finite difference scheme based a first-order
+Thus, we can see that for a finite difference scheme based on a first-order
 upwind discretization of the advection term, the optimal time step is 
@@
 approximations to all of the derivatives in~(\ref{eq:ADE_1d_upwind_corr_term})
-are at least first-order accurate.
-
-Comparing the accuracies achieved using a second-order versus a 
-first-order discretization of the advection term, it is clear that 
-the power of optimal time step selection is highly dependent on the 
-the original choice of finite difference scheme and underscores the importance 
-of carefully choosing the finite difference scheme even when optimal time
-step selection is used.
+are at least first-order accurate.  Unfortunately, the optimal time step 
+is unstable for the finite difference scheme that uses forward Euler time 
+integration and a first-order upwind stencil for the advection term.  
+In order to use optimal time step selection, we would need to replace the 
+forward Euler time integration scheme with a backward Euler scheme.
+
+A comparison of the finite difference scheme analyzed in this section to the 
+scheme described in the previous section makes it clear that the power 
+of optimal time step selection is highly dependent on the original choice of 
+finite difference scheme.  The first-order upwind scheme not only has a 
+lower overall accuracy when the optimal time step is used, it requires the 
+use of an implicit time integration method and construction of finite 
+difference stencils for higher-order spatial derivatives in order to compute
+the correction term~(\ref{eq:ADE_1d_upwind_corr_term}).  That the first-order
+scheme possesses all of these weaknesses underscores the importance of 
+carefully choosing the finite difference scheme even when optimal time step 
+selection is used.
 
 \subsection{Viscous Burgers Equation}
-For an example of optimal time step selection applied to a semilinear equation,
-we consider the viscous Burgers equation~\cite{whitham_book}:
-\beq
-  \frac{\partial u}{\pt} + u \frac{\partial u}{\px} = 
-     \nu \frac{\partial^2 u}{\px^2} 
-  \label{eq:burgers_1d}
-\eeq
-where $\nu$ is viscosity.  Notice that this problem has the same structure
-to the advection-diffusion equation, except that the advection velocity now 
-depends on the solution $u$.  To solve this equation numerically,
-we will use forward Euler for time integration and second-order central 
-difference approximations for the diffusion and advection terms:
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + \dt 
-    \left( \nu 
-      \left [ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
-         - u^n_j 
-      \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] 
-    \right).
-  \label{eq:burgers_1d_FD_scheme}
-\eeq
-This scheme is formally first-order in time and second-order in space.  
-Using the stability condition for the advection-diffusion equation as a 
-heuristic guide, we expect a stability constraint of the form 
-$\dt \le \dx^2/2 \nu$ in the limit $\dx \rightarrow 0$ (assuming that 
-the solution $u$ remains bounded).  Therefore, the scheme 
-(\ref{eq:burgers_1d_FD_scheme}) is $O(\dx^2)$ accurate overall.
-
-Because the finite difference scheme for the viscous Burgers equation is 
-very similar to the formally second-order advection-diffusion 
-equation~(\ref{eq:ADE_1d_FD_scheme}), the analysis for the error analysis
-is almost the same as in the previous section.  Following the procedure
-described for the advection-diffusion equation, we find that the 
-local truncation error has the form
-\bea
-  \tau^n_j &=&
-      \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
-           - 2 \tu \frac{\partial^3 \tu^n}{\px^3} \right)
-      \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
-      \left( 
-           4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
-         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
-         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
-      \right)
-      + O(\dt \dx^4) + O(\dt^3).
-  \label{eq:burgers_1d_err_eqn}
-\eea
-Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
-for the finite difference scheme is 
-\beq
-  - \frac{\dt^2}{2} 
-      \left( 
-           4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
-         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
-         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
-      \right)
-  \label{eq:burgers_1d_corr_term}
-\eeq 
-As for the advection-diffusion, using the standard second-order central 
-difference stencils to compute the correction term is sufficient to ensure 
-that the overall error for the scheme is $O(\dx^4)$.  
-Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate
-the boost in accuracy obtained by using the optimal time step and 
-adding correction terms to the finite difference 
-scheme~(\ref{eq:burgers_1d_FD_scheme}).
-
-\begin{figure}[htb]
+% FIGURE PLACED HERE TO SPREAD FIGURES OUT
+\begin{figure}[tb]
 \begin{center}
@@
 
-\begin{figure}[htb]
+For an example of optimal time step selection applied to a semilinear equation,
+we consider the viscous Burgers equation~\cite{whitham_book}:
+\beq
+  \frac{\partial u}{\pt} + u \frac{\partial u}{\px} = 
+     \nu \frac{\partial^2 u}{\px^2} 
+  \label{eq:burgers_1d}
+\eeq
+where $\nu$ is viscosity.  Notice that this problem has the same structure
+to the advection-diffusion equation, except that the advection velocity now 
+depends on the solution $u$.  To solve this equation numerically,
+we will use forward Euler for time integration and second-order central 
+difference approximations for the diffusion and advection terms:
+\beq
+  u^{n+1}_j = u^{n}_j 
+  + \dt 
+    \left( \nu 
+      \left [ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
+         - u^n_j 
+      \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] 
+    \right).
+  \label{eq:burgers_1d_FD_scheme}
+\eeq
+This scheme is formally first-order in time and second-order in space.  
+Using the stability condition for the advection-diffusion equation as a 
+heuristic guide, we expect a stability constraint of the form 
+$\dt \le \dx^2/2 \nu$ in the limit $\dx \rightarrow 0$ (assuming that 
+the solution $u$ remains bounded).  Therefore, the scheme 
+(\ref{eq:burgers_1d_FD_scheme}) is $O(\dx^2)$ accurate overall.
+
+Because~(\ref{eq:burgers_1d_FD_scheme}) is very similar to the second-order
+finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) for the 
+advection-diffusion equation, we can reuse many of the results obtained in 
+the previous section to derive the truncation error 
+for~(\ref{eq:burgers_1d_FD_scheme}).  Following a procedure analogous to the
+one we used to analyze~(\ref{eq:ADE_1d_FD_scheme}), we find that the local 
+truncation error for~(\ref{eq:burgers_1d_FD_scheme}) has the form
+\bea
+  \tau^n_j &=&
+      \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
+           - 2 \tu \frac{\partial^3 \tu^n}{\px^3} \right)
+      \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
+  \nonumber \\
+  &+& \frac{\dt^2}{2} 
+      \left( 
+           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
+         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
+         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
+      \right)
+      + O(\dt \dx^4) + O(\dt^3).
+  \label{eq:burgers_1d_err_eqn}
+\eea
+Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
+for the finite difference scheme is 
+\beq
+  - \frac{\dt^2}{2} 
+      \left( 
+           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
+         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
+         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
+      \right)
+  \label{eq:burgers_1d_corr_term}
+\eeq 
+As for the advection-diffusion, using the standard second-order central 
+difference stencils to compute the correction term is sufficient to ensure 
+that the overall error for the scheme is $O(\dx^4)$.  
+Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate
+the boost in accuracy obtained by using the optimal time step and 
+adding correction terms to the finite difference 
+scheme~(\ref{eq:burgers_1d_FD_scheme}).
+
+% FIGURE PLACED HERE TO SPREAD FIGURES OUT
+\begin{figure}[tb]
 \begin{center}
@@
 Next, we substitute $\tu$ into central difference approximation to the 
-bilaplacian and obtain
-\bea
-  B \tu^n = \left. \frac{\partial^4 \tu}{\px^4} \right|_j 
+bilaplacian to obtain
+\bea
+  B \tu^n = \frac{\partial^4 \tu}{\px^4} 
   - \frac{7 \dx^4}{240} 
-    \left. \frac{\partial^8 \tu}{\px^8} \right|_j 
+    \frac{\partial^8 \tu}{\px^8} 
   + O(\dx^6)
@@
 
-\begin{figure}[htb]
+\begin{figure}[tb]
 \begin{center}
@@
 
-As suggested by the discussion in this section, applying optimal time step 
-selection to high-order PDEs requires the use high-order discretizations
+As indicated by the discussion in this section, applying optimal time step 
+selection to high-order PDEs requires the use of high-order discretizations
 of the spatial derivatives.  While the need to derive high-order finite 
@@
 derivatives~\cite{gko_book,greer_2006}, 
-the overall error of finite difference approximations for high-order PDEs
-is usually dominated by the error introduced by the spatial discretization.
-As a result, it is crucial that numerical schemes for high-order PDEs use 
-high-order discretizations for the spatial derivatives.  Using high-order 
-discretizations increases the computational efficiency of numerical schemes 
-and makes it possible to obtain accurate solutions on relatively coarse 
-grids.}. 
+the overall error of finite difference schemes for high-order PDEs is usually 
+dominated by the error introduced by the spatial discretization.  As a result, 
+it is crucial that numerical schemes for high-order PDEs use high-order 
+discretizations for the spatial derivatives.  Using high-order discretizations 
+increases the computational efficiency of numerical schemes and makes it 
+possible to obtain accurate solutions on relatively coarse grids.}. 
 Optimal time step selection merely provides a valuable way to leverage the 
@@
 The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} 
-shows that the theoretical estimates of compute time are in good agreement 
-with timings from actual computational experiments.
-
-Comparing the best forward Euler and Crank-Nicholson schemes shows that
-the optimal forward Euler scheme more efficiently utilizes memory at the
-cost of requiring more computation time.  For the fourth-order parabolic 
-equation, it is important to note that this tradeoff between memory and 
-computation only exists in one space dimension.  In two-dimensions, both 
-schemes require the same amount of compute time; in greater than three space 
-dimensions, the forward Euler scheme requires less compute time than the
-Crank-Nicholson scheme.  In higher dimensions, the origin of the improved 
-performance for the forward Euler scheme with optimal time step is the 
-significantly smaller size of the computational grid required to achieve a 
-desired error level.
-
-\begin{table}[tbh]
+shows that the theoretical estimates for the computation time are in good 
+agreement with timings from actual computational experiments.
+
+\begin{table}[tb]
 \caption{\label{tab:comp_perf_vs_err_4th_order_parabolic} 
    Computational cost as a function of the global numerical error $e$
-   for various FD schemes that numerically solve the 1D fourth-order
-   parabolic equation.  For all FD schemes, the numerical solution is 
-   assumed to be computed over a fixed time interval.
+   for various finite difference schemes that numerically solve the 1D 
+   fourth-order parabolic equation.  For all finite difference schemes, the 
+   numerical solution is assumed to be computed over a fixed time interval.
 }
@@
 
+Comparing the best forward Euler and Crank-Nicholson schemes shows that
+the optimal forward Euler scheme more efficiently utilizes memory at the
+cost of requiring more computation time.  This tradeoff between memory and 
+computation only exists in one space dimension.  In two-dimensions, both 
+schemes require the same amount of compute time; in three or more space 
+dimensions, the forward Euler scheme requires less compute time than the
+Crank-Nicholson scheme.  The origin of the improved performance for the 
+forward Euler scheme with optimal time step in higher dimensions is the 
+significantly smaller size of the computational grid required to achieve a 
+desired error level.
+
 \section{\label{sec:applications_multidim}
@@
             Discretization of Spatial Derivatives}
-A significant distinction between the error for FD schemes in one space 
-dimension and multiple space dimensions is the presence of cross-terms in 
-the discretization error.  When analyzing finite difference schemes using 
-conventional numerical analysis techniques, it is usually safe to ignore the 
-precise form of the cross-terms because they only appear as higher-order 
-corrections in the error.  As a result, the order of accuracy for the FD 
-scheme is not sensitive to the particular choice of discretization for the 
-spatial derivative operators.  In contrast, knowledge of the form of the 
-cross-terms is critical when attempting to use optimal time step selection.  
-Without this knowledge, it is impossible to ensure cancellation of 
-higher-order errors.
-
-Recall that optimal time step selection requires the higher-order temporal 
+A significant distinction between the error for finite difference schemes in 
+one space dimension and multiple space dimensions is the presence of 
+cross-terms in the discretization error.  When analyzing finite difference 
+schemes using conventional numerical analysis techniques, it is usually safe 
+to ignore the precise form of the cross-terms because they only appear as 
+higher-order corrections in the error.  As a result, the order of accuracy for 
+the finite difference scheme is not sensitive to the particular choice of 
+discretization for the spatial derivative operators.  In contrast, knowledge 
+of the form of the cross-terms is critical when attempting to use optimal time 
+step selection.  Without this knowledge, it is impossible to ensure 
+cancellation of higher-order errors.
+
+Recall that optimal time step selection requires higher-order temporal 
 errors to be directly related to higher-order spatial errors via the PDE.  
@@
 \item by explicitly adding correction terms to cancel out any cross-terms 
-      in the original finite difference scheme that are cannot be eliminated 
+      in the original finite difference scheme that cannot be eliminated 
       by using the PDE.
 \end{itemize}
-While these two approaches may appear to be different, they are actually 
-equivalent because the correction terms can be combined with the original 
-FD schemes.  In essense, the correction terms can be viewed as part of an 
-expanded spatial discretization scheme and vice-versa.
+These two approaches are actually equivalent because the correction terms 
+can be combined with the finite difference stencils in the original scheme.  
+In essense, the correction terms can be viewed as part of an expanded spatial 
+discretization scheme and vice-versa.
 
@@
 For our first application of optimal time step selection in multiple
-space dimensions, we consider the two-dimensional transport 
-equation~\cite{evans_book}
+space dimensions, we consider the two-dimensional advection
+equation
 \beq
@@
 a simple forward Euler scheme that uses a first-order upwind discretization
-of the spatial derivative leads to a FD scheme for the advection equation
-with no local truncation error.  The same result can be achieved in 
-two space dimensions.  However, because there are a total of three
+of the spatial derivative leads to a finite difference scheme for the 
+advection equation with no local truncation error.  The same result can be 
+achieved in two space dimensions.  However, because there are a total of three
 numerical parameters (\ie two grid spacings and the time step), it is 
-insufficient to only optimize the time step.  We need to optimally select
-the time step \emph{and} the ratio of the grid spacings.
+insufficient to optimize only the time step; the ratio of the grid 
+spacings also needs to be optimally selected.
 
@@
 \eeq
-where superscripts and subscripts denote the time and space indices, 
-respectively.  Using conventional error analysis, it is straightforward to 
-show that this method is formally first-order in both space and time.
+This method is formally first-order in both space and time.
 
@@
 usual, by analyzing the discretization error 
-for~(\ref{eq:advection_eqn_2d_FD_scheme}).  As in our analysis of the FD
-scheme for the one-dimensional advection equation, we retain all of the terms 
-in the discretization error.  Using Taylor series to expand the true solution 
-of the the PDE at time step $(n+1)$, we find that
+for~(\ref{eq:advection_eqn_2d_FD_scheme}).  As in our analysis of the finite
+difference scheme for the one-dimensional advection equation, we retain all of 
+the terms in the discretization error.  Using Taylor series to expand the true 
+solution of the PDE at time step $(n+1)$, we find that
 \bea
@@
   + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\pt^k} \right|_{i,j}
+       \frac{\partial^k \tu^n}{\pt^k} 
   \label{eq:advection_eqn_2d_time_err}.
@@
   \frac{\tu^{n}_{i,j} - \tu^{n}_{i-1,j}}{\dx} = 
-  \left. \frac{\partial \tu}{\px} \right|_{i,j}
+  \frac{\partial \tu}{\px} 
   - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\px^k} \right|_{i,j} 
+       \frac{\partial^k \tu^n}{\px^k} 
+  \label{eq:advection_eqn_2d_space_err_x}
   \\
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dx} = 
-  \left. \frac{\partial \tu}{\py} \right|_{i,j}
+  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dy} = 
+  \frac{\partial \tu}{\py} 
   - \frac{1}{\dy} \sum_{k=2}^\infty \frac{\left( -\dy \right)^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\py^k} \right|_{i,j} 
-  \label{eq:advection_eqn_2d_space_err}
+       \frac{\partial^k \tu^n}{\py^k} 
+  \label{eq:advection_eqn_2d_space_err_y}
 \eea
@@
   + \sum_{k=1}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left. \left( A_x \partial_x
-                   + A_y \partial_y
-              \right)^k \tu^{n} \right|_{i,j}.
+       \left( A_x \partial_x
+            + A_y \partial_y \right)^k \tu^{n}.
   \label{eq:advection_eqn_2d_time_err_modified}
 \eea
-Combining~(\ref{eq:advection_eqn_2d_time_err_modified}) 
-and~(\ref{eq:advection_eqn_2d_space_err}) with the finite difference 
+Combining~(\ref{eq:advection_eqn_2d_space_err_x}), 
+(\ref{eq:advection_eqn_2d_space_err_y}), 
+and~(\ref{eq:advection_eqn_2d_time_err_modified}) with the finite difference 
 scheme~(\ref{eq:advection_eqn_2d_FD_scheme}), we find that the local
@@
    &+& \frac{A_x\dt}{\dx} \sum_{k=3}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\px^k} \right|_{i,j} 
+       \frac{\partial^k \tu^n}{\px^k} 
    + \frac{A_y\dt}{\dy} \sum_{k=3}^\infty \frac{\left( -\dy \right)^k}{k!} 
-       \left. \frac{\partial^k \tu^n}{\py^k} \right|_{i,j} 
+       \frac{\partial^k \tu^n}{\py^k} 
    \nonumber \\
    &-& \sum_{k=3}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left. \left( A_x \partial_x
-                   + A_y \partial_y
-              \right)^k \tu^{n} \right|_{i,j} 
+       \left( A_x \partial_x + A_y \partial_y
+              \right)^k \tu^{n} 
   \label{eq:advection_eqn_2d_trunc_err}
@@
 With these choices, we are able to boost the order of accuracy for the 
-FD scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from first- to second-order.
+finite difference scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from first- to 
+second-order.
 
@@
        \sum_{k=3}^\infty \frac{\left( -\dx \right)^k}{k!} 
-              \left. \frac{\partial^k \tu^n}{\px^k} \right|_{i,j} 
+              \frac{\partial^k \tu^n}{\px^k} 
    \nonumber \\
@@
        \sum_{k=3}^\infty \frac{\left( -\dy \right)^k}{k!} 
-              \left. \frac{\partial^k \tu^n}{\py^k} \right|_{i,j} 
+              \frac{\partial^k \tu^n}{\py^k} 
    \nonumber \\
@@
               \frac{\dx^l \dy^{k-l}}{k!} 
-              \left. \frac{\partial^k \tu^n}{\px^l \py^{k-l}} \right|_{i,j} 
+              \frac{\partial^k \tu^n}{\px^l \py^{k-l}} 
    \nonumber \\
    &-& \sum_{k=3}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left. \left( A_x \partial_x
-                   + A_y \partial_y
-              \right)^k \tu^{n} \right|_{i,j} 
+       \left( A_x \partial_x + A_y \partial_y
+              \right)^k \tu^{n} 
   \label{eq:advection_eqn_2d_trunc_err_mod},
@@
 which vanishes when the optimal time step and grid spacing ratio are used.
-Thus, when the simple first-order upwind approximation to the correction 
-term based on (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used in
+Thus, when (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used in
 conjunction with the optimal time step and grid spacing ratio, we obtain
-a FD scheme that introduces no truncation error at each time step.
-Figure~\ref{fig:advection_eqn_2d_soln} demonstrates the superior accuracy 
+a finite difference scheme that introduces no truncation error at each time 
+step.  Figure~\ref{fig:advection_eqn_2d_soln} demonstrates the amazing accuracy 
 of this scheme over the first-order forward Euler 
@@
 To see that the correction term can be considered as part of the spatial 
-discretization scheme, we can write out the complete FD scheme including 
-the correction term:
+discretization scheme, we can write out the complete finite difference scheme 
+including the correction term:
 \bea
@@
 \eea
-Thus, we see that the FD scheme (\ref{eq:advection_eqn_2d_FD_scheme_complete})
-is essentially forward Euler time integration with a first-order upwind 
-discretization of the spatial derivatives that is computed using all four 
-upwind neighbors.
+Thus, we see that the finite difference scheme 
+(\ref{eq:advection_eqn_2d_FD_scheme_complete}) is essentially forward Euler 
+time integration with first-order upwind discretizations of the spatial 
+derivatives that are computed using all four upwind neighbors.
 
 In higher space dimensions, it becomes tedious to carry out the analysis 
-to derive the infinite order accurate FD scheme for the advection equation
-because many high-order correction terms must be retained.  However, it is 
-straighforward to generalize 
+to derive the infinite order accurate finite difference scheme for the 
+advection equation because many high-order correction terms must be retained.  
+However, it is straighforward to generalize 
 (\ref{eq:advection_eqn_2d_FD_scheme_simplified}) to $d$ space dimensions by 
-using a first-order discretization for the spatial derivatives that is an 
-average over the $2^{d-1}$ first-order upwind discretizations in each 
-coordinate direction.   As with the two-dimensional problem, the optimal 
-grid spacings satisfy $\dx_k/A_{x_k} = \dt_{opt}$ for $1 \le k \le d$.
+using first-order discretizations for the spatial derivatives that are 
+averages over the $2^{d-1}$ upwind discretizations in each coordinate 
+direction.   As with the two-dimensional problem, the optimal grid spacings 
+satisfy $\dx_k/A_{x_k} = \dt_{opt}$ for $1 \le k \le d$.
 
@@
 In general, optimal parameter selection boosts the accuracy by only a few 
-orders.  In this section, we use the 2D diffusion equation to illustrate the 
-typical accuracy boost gained through optimal time step selection.  We also 
-demonstrate the utility of optimal time step selection for designing simple 
-FD schemes that yield high-order accuracy for problems on irregular domains.
+orders.  To illustrate the typical accuracy boost gained by using optimal 
+time step selection, we apply it to the 2D diffusion equation.  We also 
+use the 2D diffusion equation to demonstrate the utility of optimal time 
+step selection for problems on irregular domains.
 
@@
          +u^{n}_{i+1,j} + u^{n}_{i-1,j} - 4u^{n}_{i,j}}{\dx^2}
-    \right),
+    \right).
   \label{eq:diffusion_eqn_2d_5pt_scheme}
 \eea
-Conventional error analysis shows that this scheme is formally first-order in 
-time and second-order in space.
+This scheme is formally first-order in time and second-order in space.
 
@@
 scheme using the nine-point stencil for the 
-Laplacian~\cite{iserles_book,patra_2005}, which possesses a leading-order 
+Laplacian~\cite{iserles_book,patra_2005}, which does possess a leading-order 
 error directly related to the spatial derivative operator in the PDE 
@@
 \eea
-Like, (\ref{eq:diffusion_eqn_2d_5pt_scheme}), the forward Euler scheme that 
+Like (\ref{eq:diffusion_eqn_2d_5pt_scheme}), the forward Euler scheme that 
 uses the nine-point discretization of the Laplacian 
@@
 naturally arises when directly applying optimal time step selection to the
-five-point scheme through reinterpretion of the correction term as a 
-modification to the spatial discretization.
+five-point scheme if the correction terms are interpreted as a modification 
+to the spatial discretization.
 
@@
   L^{9pt} \tu &=&
-  \left. \nabla^2 \tu \right|_{i,j}
-  + \frac{\dx^2}{12} \left. \nabla^4 \tu \right|_{i,j}
-  + O(\dx^4)
+  \nabla^2 \tu + \frac{\dx^2}{12} \nabla^4 \tu + O(\dx^4)
   \label{eq:diffusion_eqn_2d_space_err}
@@
 \eea
-
 As for the 1D diffusion equation, we see that the optimal time step is given 
-by $\dto = \dx^2/6D$ and that this choice of time step yields a scheme is 
-that is $O(\dx^4) + O(\dt^2) = O(\dx^4)$ overall.  Note that there is no need
-to consider an implicit scheme because $\dto$ is stable for the forward Euler 
-time integration scheme.
-
-\begin{figure}[htb]
+by $\dto = \dx^2/6D$ and that this choice of time step yields a scheme 
+that is fourth-order accurate.  Note that there is no need to consider an 
+implicit scheme because $\dto$ is stable for the forward Euler time 
+integration scheme.
+
+\begin{figure}[tb]
 \begin{center}
@@
 The improvement in the accuracy of the forward Euler scheme when the optimal 
-time step is used is clear in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
-The numerical errors of the following three finite difference schemes for 
-the 2D diffusion equation show the expected orders of accuracy: forward Euler 
-with optimal time step and correction terms, forward Euler with suboptimal 
-time step, and Crank-Nicholson.  As with our other examples, the forward 
-Euler scheme with an optimal time step is much more efficient that the two 
-other schemes.  With as few as 25 grid points in each space direction, the 
-error in solution computed by the forward Euler scheme with an optimal time 
-step is orders of magnitude smaller.
-
-Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also compares the computational
-performance of the three schemes.  For all three schemes, the scaling of the 
-computational time with error is in good agreement with the theoretical 
-expectations presented in Table~\ref{tab:comp_perf_vs_dim}.  Again, the 
-forward Euler with optimal time step demonstrates much greater efficiency 
-than the other schemes.  In addition to having better scaling properties, 
-the line for the forward Euler scheme with optimal time step appears to be 
-significantly shifted to the left (\ie smaller errors).
+time step is used is shown in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
+Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also shows that the 
+scaling of the computational time with the error is in good agreement with 
+theoretical estimates (see Table~\ref{tab:comp_perf_vs_dim}).  As with the
+other examples, the forward Euler scheme with an optimal time step is much 
+more computationally efficient than the two other schemes.
 
@@
 For example, a fourth-order accurate solution to the diffusion equation
-could be obtained by using a pseudospectral spatial
-discretization~\cite{trefethen_spectral_book, boyd_spectral_book} of the 
-Laplacian and a fourth-order time integration scheme (\eg Runge-Kutta).
-When the domain is irregularly shaped, high-order methods are harder to design 
-and implement.  For instance, spectral and pseudospectral methods are not 
-usually appropriate because the domain is not a simple cartesian product of 
-intervals.
-
-As discussed at the end of Section~\ref{sec:ots_higher_spatial_dims}, we can 
-apply optimal time step selection to construct simple finite difference schemes 
-that yield high-order accurate solutions for problems on irregular domains.  
+(in any number of space dimensions) could be obtained by using a 
+pseudospectral spatial discretization~\cite{trefethen_spectral_book, 
+boyd_spectral_book} of the Laplacian and a fourth-order time integration scheme 
+(\eg Runge-Kutta).  When the domain is irregularly shaped, high-order methods 
+are harder to design and implement.  For instance, spectral and pseudospectral 
+methods are not usually appropriate because the domain is not a simple 
+cartesian product of intervals.
+
 In this section, we extend the OTS finite difference scheme for the 2D 
-diffusion equation on regular domains to irregular domains by describing a 
-method for computing the values of the solution in ghost cells that ensures 
-that imposing boundary conditions does not decrease the accuracy of the 
-solution.  For simplicity, we choose to represent the irregular boundary 
-implicitly as the zero level set of a level set function $\phi$ which 
-satisfies:
+diffusion equation on regular domains to irregular domains.  To avoid the 
+need for special stencils at grid points near the irregular boundary, we adopt 
+the `ghost cell' approach for imposing boundary conditions~\cite{fedkiw_1999, 
+osher_fedkiw_book, gibou_2005}.  The basic idea is to set the value of $u$ at 
+any grid point that lies immediately outside of the domain (and thus may be
+required by spatial derivative operators at interior grid points) by 
+extrapolating from values of $u$ at interior grid points and using boundary 
+condition information.  
+The main challenge is to find a method for filling ghost cells that does not 
+decrease the accuracy of the solution in the interior of the domain.  For 
+simplicity, we choose to represent the irregular boundary implicitly as the 
+zero level set of a level set function $\phi$ which satisfies:
 \bea
@@
 
-Devising a strategy for accurately imposing boundary conditions is perhaps 
-the most critical aspect of designing finite difference schemes for problems 
-on irregular domains.  To avoid the need for special stencils at grid points
-near the irregular boundary, we adopt the `ghost cell' or `ghost value'
-approach to imposing boundary conditions~\cite{fedkiw_1999, osher_fedkiw_book, 
-gibou_2005}.  The basic 
-idea is to set the value of grid points that lie outside of the domain 
-and are required by the spatial discretization operator for grid points 
-inside the domain by extrapolating from interior grid points using
-boundary condition information.  Figure~\ref{fig:ghost_cells} shows examples 
-of ghost cells that lie in the vicinity of a irregular boundary.  Notice that
-there are two types of ghost cells:  edge and corner.  Edge and corner
-ghost cells are distinguished by the relative position of the nearest 
-interior grid point.  An edge ghost cell is linked to its nearest interior
-neighbor by an edge of the computational grid whereas a corner ghost cell
-and its nearest interior neighbor lie on the opposite corners of a grid cell.
-
-\begin{figure}[htb]
+Figure~\ref{fig:ghost_cells} shows examples of ghost cells that lie in the 
+vicinity of an irregular boundary.  Notice that there are two types of ghost 
+cells:  edge and corner.  Edge and corner ghost cells are distinguished by 
+the relative position of the nearest interior grid point.  An edge ghost cell 
+is linked to its nearest interior neighbor by an edge of the computational 
+grid whereas a corner ghost cell and its nearest interior neighbor lie on the 
+opposite corners of a grid cell.
+
+\begin{figure}[tb]
 \begin{center}
@@
 the value on the boundary, $u_B$, and values at three interior grid points.  
-The corner ghostcells, $u_C$, is filled by using quadratic extrapolation 
-of the values at interior grid points and neighboring edge ghost cells.
+The corner ghostcell, $u_C$, is filled by using a fourth-order accurate
+Taylor series expansion centered at $u_{0,0}$, which uses the values of $u$ 
+at the specified interior grid points and neighboring edge ghost cells to 
+approximate the partial derivatives in the expansion. 
 }
@@
 
-To set the value in each edge ghost cell, we use 1D, cubic extrapolation of 
-the values of $u$ across the interface.  A cubic Lagrange extrapolant is 
-constructed for each edge ghost cell using a point on the domain boundary 
-and three interior grid points that lie along the line parallel to the edge 
-connecting the ghost cell to its nearest interior neighbor (see 
-Figure~\ref{fig:ghost_cells}).  
+To set the value of $u$ in each edge ghost cell, we use 1D, cubic 
+extrapolation of the values of $u$ across the interface.  A cubic Lagrange 
+extrapolant is constructed for each edge ghost cell using a point on the 
+domain boundary and three interior grid points that lie along the line 
+parallel to the edge connecting the ghost cell to its nearest interior 
+neighbor (see Figure~\ref{fig:ghost_cells}).  
 The location of the point $u_B$ can be calculated by inverting the linear 
@@
 interior neighbor\footnote{We found that using higher-order interpolants to 
-locate boundary points was unnecessary to achieve high-order accuracy in the
-computed solution.}.
+locate boundary points was unnecessary for achieving high-order accuracy in 
+the computed solution.}.
 Because the quality of the extrapolant deteriorates if the boundary point is 
@@
 when the distance between the boundary point and the nearest interior grid 
-point is less than $\dx^2$.  For the numerical scheme presented in this paper, 
-this threshold was too low.  We found that better results were obtained if 
-we used an $O(\dx)$ threshold.
-
-To set the values in corner ghost cells, we use a fourth-order accurate
-Taylor series expansion of $u$ centered at the nearest interior 
-neighbor.  The partial derivatives in the Taylor series expansion are 
-approximated using finite differences constructed from the values of $u$ 
-across the domain boundary and the neighboring edge ghost cells (see 
-Figure~\ref{fig:ghost_cells}).  To obtain a fourth-order accurate value for 
-the corner ghost cell, we need to carefully choose the finite difference 
-stencils used for the partial derivatives.  For the corner ghost cell shown 
-in Figure~\ref{fig:ghost_cells}, the extrapolation stencil is given by
-\bea
-  u_C &=&  -4 u_{0,0} - u_{-1,-1} + 2 u_{1,0} + 2 u_{0,1}
-      - u_{1,-1} - u_{-1,1} + 2 u_{0,-1} + 2 u_{-1,0}
-\eea
-where $u_C$ and $u_{i,j}$ are as indicated in the figure.  The stencils for
-corner ghost cells that are in other positions relative to the interior of 
-the domain are obtained by rotations of Figure~\ref{fig:ghost_cells} and 
-remapping of the indicies in the above stencil.
-
-Using these methods for filling edge and corner ghost cells, it is 
-straightforward to compute high-order accurate solutions of the 2D diffusion 
-equation on irregular domains.  We simply use the OTS forward euler scheme 
-designed for regular domains for grid points in the interior of the domain
-after filling the ghost cells with high-order accurate values.  
-Figure~\ref{fig:diffusion_eqn_2d_starfish_domain} shows an example of 
-a solution computed on a starfish shaped domain.  It also shows that the 
-error in the solution is concentrated near the boundaries of the irregular
-domain.  Thus, we see that it is necessary to use high-order accurate 
-extrapolants for the ghost cells to prevent errors at the boundary from 
-destroying high-order accuracy throughout the domain.  The boosted accuracy 
-of the OTS solution compared to a simple forward Euler scheme with a 
-suboptimal time step is illustrated in 
-Figure~\ref{fig:diffusion_eqn_2d_starfish_error}.
-
-\begin{figure}[htb]
+point is less than $\dx^2$~\cite{gibou_2005}.  For the numerical scheme 
+presented in this paper, this threshold was too low.  We found that better 
+results were obtained if we used an $O(\dx)$ threshold.
+
+% FIGURE SHOWING SOLUTION OF 2D DIFFUSION EQUATION ON STARFISH DOMAIN
+% PLACED HERE TO SPREAD FIGURES OUT
+\begin{figure}[tb]
 \begin{center}
@@
 The numerical solution is calculated using the square computational domain 
-$-1 < x,y < 1$ which is discretized using 800 grid points in each coordinate 
+$-1 < x,y < 1$ which is discretized using 400 grid points in each coordinate 
 direction (the plot of the solution has been downsampled for visualization 
@@
 
-\begin{figure}[htb]
+To set the value of $u$ in a corner ghost cell, we use a fourth-order accurate
+Taylor series expansion of $u$ centered at the nearest interior 
+neighbor.  The partial derivatives in the Taylor series expansion are 
+approximated using finite differences constructed from the values of $u$ 
+at interior grid points and the neighboring edge ghost cells (see 
+Figure~\ref{fig:ghost_cells}).  To obtain a fourth-order accurate value for 
+the corner ghost cell, we need to carefully choose the finite difference 
+stencils used for the partial derivatives.  For the corner ghost cell shown 
+in Figure~\ref{fig:ghost_cells}, the extrapolation stencil is given by
+\bea
+  u_C &=&  -4 u_{0,0} - u_{-1,-1} + 2 u_{1,0} + 2 u_{0,1}
+      - u_{1,-1} - u_{-1,1} + 2 u_{0,-1} + 2 u_{-1,0}
+\eea
+where $u_C$ and $u_{i,j}$ are indicated in the figure.  The stencils for
+corner ghost cells that are in other positions relative to the interior of 
+the domain are obtained by rotations of Figure~\ref{fig:ghost_cells} and 
+remapping of the indicies in the above stencil.
+
+Using these methods for filling edge and corner ghost cells, it is 
+straightforward to compute high-order accurate solutions of the 2D diffusion 
+equation on irregular domains.  We simply use the OTS forward euler scheme 
+designed for regular domains for grid points in the interior of the domain
+after filling the ghost cells with high-order accurate values.  
+Figure~\ref{fig:diffusion_eqn_2d_starfish_domain} shows an example of 
+a solution computed on a starfish shaped domain.  It also shows that the 
+error in the solution is concentrated near the boundaries of the irregular
+domain.  Thus, we see that it is necessary to use high-order accurate 
+extrapolants for the ghost cells to prevent errors at the boundary from 
+destroying high-order accuracy throughout the domain.  The improved accuracy 
+of the OTS solution compared to a simple forward Euler scheme with a 
+suboptimal time step is illustrated in 
+Figure~\ref{fig:diffusion_eqn_2d_starfish_error}.
+
+\begin{figure}[tb]
 \begin{center}
@@
 irregular domain: forward Euler with optimal time step and correction 
-terms (circles) and forward Euler with suboptimal time step (squares).
-Both methods solve the diffusion equation described in the caption for
-Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
+terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/8 D$
+(squares).  Both methods solve the 2D diffusion equation described in the 
+caption for Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
 }
@@
 \subsubsection{Diffusion Equation in Higher Dimensions}
-Using optimal time step selection to boost the accuracy of FD schemes for 
-the diffusion equation in higher space dimensions is straightforward.  
-As for the 2D problem, the key step is identification of a finite difference 
-stencil for the Laplacian that has an \emph{isotropic} discretization error.
-For 3D problems, a catalog of several such stencils is available 
-in~\cite{patra_2005}.  For higher dimensional problems, using symbolic 
-algebra software to derive FD stencils with desired discretization error 
-properties can be helpful~\cite{patra_2005, gupta_1998}.  Once an acceptable
-FD approximation for the Laplacian has been chosen, the usual OTS analysis 
-yields an optimal time step of $\dt_{opt} = \dx^2/6D$ for forward Euler
-time integration and a boost of the order of accuracy from two to four.
+Using optimal time step selection to boost the accuracy of finite difference
+schemes for the diffusion equation in higher space dimensions is 
+straightforward.  As for the 2D problem, the key step is identification of a 
+finite difference stencil for the Laplacian that has an \emph{isotropic} 
+discretization error.  For 3D problems, a catalog of several such stencils is 
+available in~\cite{patra_2005}.  For higher dimensional problems, using 
+symbolic algebra software to derive finite difference stencils with desired 
+discretization error properties can be helpful~\cite{patra_2005, gupta_1998}.  
+Once an acceptable finite difference approximation for the Laplacian has been 
+chosen, the usual OTS analysis yields an optimal time step of 
+$\dt_{opt} = \dx^2/6D$ for forward Euler time integration and a boost of the 
+order of accuracy from two to four.  Above three dimensions, it is necessary 
+to use an implicit time integration scheme because $\dt_{opt}$ is no longer 
+stable for the forward Euler scheme.
 
@@
 and irregular domains.  The examples illustrate the surprisingly high-order
-accuracy that can be achieved using only low-order discretizations for 
+of accuracy that can be achieved using only low-order discretizations for 
 spatial derivatives and simple forward Euler time integration.
@@
 but, as OTS selection demonstrates, optimal selection of numerical parameters 
-can significantly impact on the accuracy of numerical methods.  To realize
+can significantly impact the accuracy of numerical methods.  To realize
 the full potential of numerical methods, both optimal time step selection 
@@
 High-Performance Computing (IHPC) in Singapore and Vitamin D, Inc.
-The author would like to thanks P. Fok and ?? for enlightening discussions 
+The author would like to thank P. Fok and ?? for enlightening discussions 
 and suggestions on the manuscript.

### 2008-05-14 papers/OTSPDE/OTSPDE.tex
OTSPDE: renamed LaTeX files from optimal_time_steps_for_pdes to OTSPDE.
OTSPDE:  added references to work on the use of spectral methods for solving problems on irregular domains.

--- 2008-05-13 papers/OTSPDE/optimal_time_steps_for_pdes.tex
@@
 desired order of accuracy for the solution.  A detailed discussion of 
-appropriate methods for imposing boundary conditions is beyond the scope of 
-this paper.  We refer interested readers to the thorough analysis presented
+methods for imposing boundary conditions is beyond the scope of this paper.  
+We refer interested readers to the thorough analysis presented
 in~\cite{gko_book}.  To avoid complicating the discussion in 
@@
 are harder to design and implement.  For instance, spectral and pseudospectral 
-methods are not usually appropriate because the domain is not a simple 
-cartesian product of intervals.
+methods are not directly applicable because the domain is not a simple 
+cartesian product of intervals.  While it is possible to use spectral methods
+for problems on irregular domains through the clever introduction of a phase 
+field~\cite{bueno_orovio_2006, feng_2006}, this approach does not 
+actually achieve spectral accuracy because the error is dominated by the 
+width of the smoothed approximation to the phase field (as opposed to the 
+accuracy of the discretization for the spatial 
+derivatives)~\cite{bueno_orovio_2006}.
 
@@
 
-Figure~\ref{fig:ghost_cells} shows examples of ghost cells that lie in the 
-vicinity of an irregular boundary.  Notice that there are two types of ghost 
-cells:  edge and corner.  Edge and corner ghost cells are distinguished by 
-the relative position of the nearest interior grid point.  An edge ghost cell 
-is linked to its nearest interior neighbor by an edge of the computational 
-grid whereas a corner ghost cell and its nearest interior neighbor lie on the 
-opposite corners of a grid cell.
-
+% FIGURE SHOWING GHOST CELLS AT IRREGULAR BOUNDARY
+% PLACED HERE TO SPREAD FIGURES OUT
 \begin{figure}[tb]
@@
 \end{figure}
+
+Figure~\ref{fig:ghost_cells} shows examples of ghost cells that lie in the 
+vicinity of an irregular boundary.  Notice that there are two types of ghost 
+cells:  edge and corner.  Edge and corner ghost cells are distinguished by 
+the relative position of the nearest interior grid point.  An edge ghost cell 
+is linked to its nearest interior neighbor by an edge of the computational 
+grid whereas a corner ghost cell and its nearest interior neighbor lie on the 
+opposite corners of a grid cell.
 
@@
 
+% FIGURE SHOWING SOLUTION OF 2D DIFFUSION EQUATION ON STARFISH DOMAIN
+% PLACED HERE TO SPREAD FIGURES OUT
 \begin{figure}[tb]
@@
 
-\bibliography{optimal_time_steps_for_pdes}
+\bibliography{OTSPDE}
 

### 2008-05-14 papers/OTSPDE/OTSPDE.tex
OTSPDE:  minor revisions to captions to make them more consistent.

@@
 at the resolution of the figure.
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }
@@
 - 4 \sin(11 \pi x/2)$.  
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }
@@
 where $T = Dt+1$.
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }
@@
 stepping with suboptimal time step $\dt = 2 \dx^2/3 D$ (right).  
-Both figures show the numerical solution at $t = 1$ computed on the domain 
-$-10 < x < 10$ using 50 grid points.  For details about the initial/boundary 
+Both figures compare the analytical solution (solid line) at $t = 1$ on the 
+domain $-10 < x < 10$ with the numerical solution (circles) computed using 
+50 grid points.  For details about the initial/boundary 
 conditions and physical parameters used to compute these solutions, see the
@@
 the advection and diffusion terms.
-The viscous Burgers equation is solved on the domain $0 < x < 10$ to 
-time $t=2$ with the viscosity $\nu$ set to $0.1$.  The initial conditions are 
+The viscous Burgers equation is solved to time $t=2$ on the domain 
+$0 < x < 10$ with the viscosity $\nu$ set to $0.1$.  The initial conditions are 
 taken to be 
@@
 where $T = t+1$.
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }
@@
 optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step (right).  Both figures show the 
-numerical solution at $t = 2$ computed using $100$ grid points.
+stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
+Both figures compare the analytical solution (solid line) at $t = 2$ on the 
+domain $0 < x < 10$ with the numerical solution (circles) computed using 
+$100$ grid points.  
 For details about the initial/boundary conditions and physical parameters 
@@
 the bilaplacian.  Using each method, we solve the fourth-order parabolic
-equation with $\kappa = 1$ on the domain $0 < x < 1$ subject to the boundary 
-conditions $u(0,t) = u(1,t) = 0$ and the initial condition
-$u(x,0) = 2\sin(2 \pi x) + \sin(5 \pi x)$.  The value of the solution
+equation with $\kappa = 1$ to time $t = 0.001$ on the domain $0 < x < 1
+$ subject to the boundary conditions $u(0,t) = u(1,t) = 0$ and the initial 
+condition $u(x,0) = 2\sin(2 \pi x) + \sin(5 \pi x)$.  The value of the solution
 in the ghost cells (required as a result of the wide stencil for the 
@@
         + \sin(5 \pi x) \left[ 2-\exp(-100\pi^2 t) \right]$.
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }
@@
 forward Euler with optimal time step and correction terms (circles), forward 
-Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
+Euler with suboptimal time step $\dt = \dx^2/4D$ (squares), and 
+Crank-Nicholson with time step $\dt = \dx/16D$ (diamonds).  
 The forward Euler schemes use a nine-point discretization of the Laplacian;
@@
 the Laplacian.  Using each method, we solve the 2D diffusion equation with 
-$D = 0.25$ on the domain $0 < x, y < 1$ subject to the initial condition
+$D = 0.25$ and source term 
+$f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
+            \exp\left(-5 D \pi^2 t\right)
+          - 24 D \pi^2 \sin(5 \pi x) \sin(\pi y) 
+            \exp \left(-2 D \pi^2 t \right)$.
+to time $t = 0.1$ on the domain $0 < x, y < 1$ subject 
+to the initial condition
 $u(x,y,0) =  1 + y/3 + x/2 + xy/4 
-          + \sin(2 \pi x) \sin(3 \pi y) - \sin(5 \pi x) \sin(\pi y)$
-and boundary conditions taken from the analytical solution 
+          + \sin(2 \pi x) \sin(3 \pi y) - \sin(5 \pi x) \sin(\pi y)$.
+The boundary conditions are taken from the analytical solution 
 $u(x,t) = 1 + y/3 + x/2 + xy/4 
@@
 resolution of the figures.  
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }
@@
 
-In this section, we extend the OTS finite difference scheme for the 2D 
-diffusion equation on regular domains to irregular domains.  To avoid the 
-need for special stencils at grid points near the irregular boundary, we adopt 
-the `ghost cell' approach for imposing boundary conditions~\cite{fedkiw_1999, 
-osher_fedkiw_book, gibou_2005}.  The basic idea is to set the value of $u$ at 
-any grid point that lies immediately outside of the domain (and thus may be
-required by spatial derivative operators at interior grid points) by 
-extrapolating from values of $u$ at interior grid points and using boundary 
-condition information.  
-The main challenge is to find a method for filling ghost cells that does not 
-decrease the accuracy of the solution in the interior of the domain.  For 
-simplicity, we choose to represent the irregular boundary implicitly as the 
-zero level set of a level set function $\phi$ which satisfies:
-\bea
-  \begin{array}{ll}
-  \phi < 0 & \mathrm{for} \ x \in \Omega \\
-  \phi > 0 & \mathrm{for} \ x \notin \Omega \\
-  \phi = 0 & \mathrm{for} \ x \in \partial \Omega 
-  \end{array}
-\eea
-where $\Omega$ and $\partial \Omega$ are the interior and boundary of the 
-irregular domain.
-
 % FIGURE SHOWING GHOST CELLS AT IRREGULAR BOUNDARY
@@
 \end{figure}
+
+In this section, we extend the OTS finite difference scheme for the 2D 
+diffusion equation on regular domains to irregular domains.  To avoid the 
+need for special stencils at grid points near the irregular boundary, we adopt 
+the `ghost cell' approach for imposing boundary conditions~\cite{fedkiw_1999, 
+osher_fedkiw_book, gibou_2005}.  The basic idea is to set the value of $u$ at 
+any grid point that lies immediately outside of the domain (and thus may be
+required by spatial derivative operators at interior grid points) by 
+extrapolating from values of $u$ at interior grid points and using boundary 
+condition information.  
+The main challenge is to find a method for filling ghost cells that does not 
+decrease the accuracy of the solution in the interior of the domain.  For 
+simplicity, we choose to represent the irregular boundary implicitly as the 
+zero level set of a level set function $\phi$ which satisfies:
+\bea
+  \begin{array}{ll}
+  \phi < 0 & \mathrm{for} \ x \in \Omega \\
+  \phi > 0 & \mathrm{for} \ x \notin \Omega \\
+  \phi = 0 & \mathrm{for} \ x \in \partial \Omega 
+  \end{array}
+\eea
+where $\Omega$ and $\partial \Omega$ are the interior and boundary of the 
+irregular domain.
 
@@
 where the error in the solution is larger than $25$\% of the $L^\infty$ error 
-of the solution.  For these figures, the diffusion constant $D$ is set to 
-$1/4$ and the source term is taken to be 
-$f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
-            \exp\left(-5 D \pi^2 t\right)
-          - 24 D \pi^2 \sin(5 \pi x) \sin(\pi y) 
-            \exp \left(-2 D \pi^2 t \right).
-$
-The initial conditions for the problem are 
-$u(x,y,0) = 1 + y/3 + x/2 + xy/4 
-          + \sin(2 \pi x) \sin(3 \pi y) 
-          - \sin(5 \pi x) \sin(\pi y)
-$
-and the boundary conditions are obtained from the analytical solution
-$u(x,y,t) = 1 + y/3 + x/2 + xy/4 
-              + \sin(2 \pi x) \sin(3 \pi y) \exp(-5 D \pi^2 t) 
-              - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)
-$.
-The boundary of the starfish domain is represented by the zero-level set of 
-the function
+of the solution.  For these figures, we use the same diffusion constant, 
+source term, and initial conditions as described in the caption for 
+Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.  The boundary of the 
+starfish-shaped domain is represented by the zero-level set of the function
 $\sqrt{x^2 + y^2} - \frac{3}{5} \left(1 + \frac{1}{2}\sin(5 \theta) \right)$, 
 where $\theta = \mathtt{atan}\left( y/x \right)$.
-The numerical solution is calculated using the square computational domain 
-$-1 < x,y < 1$ which is discretized using 400 grid points in each coordinate 
-direction (the plot of the solution has been downsampled for visualization 
-purposes).  Boundary conditions are imposed by using fourth-order accurate
-approximations of the value of the solution in edge and corner ghost cells
-(described in the main text).
+The Dirichlet boundary conditions are taken from the values of the analytical 
+solution (see caption for Figure~\ref{fig:diffusion_eqn_2d_src_analysis})
+on the irregular boundary.  The numerical solution is calculated using the 
+square computational domain $-1 < x,y < 1$ which is discretized using 400 grid 
+points in each coordinate direction (the plot of the solution has been 
+downsampled for visualization purposes).  Boundary conditions are imposed by 
+using fourth-order accurate approximations of the value of the solution in 
+edge and corner ghost cells (described in the main text).
 }
@@
 (squares).  Both methods solve the 2D diffusion equation described in the 
-caption for Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
+caption for Figure~\ref{fig:diffusion_eqn_2d_src_analysis} on the irregular
+domain defined in the caption for 
+Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
+These results were obtained using simple MATLAB implementations of the 
+finite difference schemes.
 }

### 2008-06-08 src/1d/fourth_order_parabolic/solve4thOrderParabolicEqnForwardEulerOTS1d.m
OTSPDE: fixed errors in description of source term in comments for fourth-order parabolic equation examples.

@@ -7,10 +7,10 @@
 %
 % where
 %
-%   f(x,t) = -3*sin(5*pi*x)*sin(3*t)
-%          + 625*pi^4*sin(5*pi*x)*cos(3*t)
-%          + 4*pi^2*sin(2*pi*x)*exp(-4*pi^2*t)
-%          + 16*pi^4*sin(2*pi*x)*(3-exp(-4*pi^2*t))
+%   f(x,t) = -20000*pi*sin(2*pi*x)*sin(10000*pi*t) ...
+%          + 32*pi^4*sin(2*pi*x)*cos(10000*pi*t) ...
+%          + 100*pi^2*sin(5*pi*x)*exp(-100*pi^2*t) ...
+%          + 625*pi^4*sin(5*pi*x)*(2-exp(-100*pi^2*t))
 %
 % on the domain -1 < x < 1 with homogeneous Dirichlet boundary conditions 
 % imposed at all boundaries.  The numerical solution is computed on a 

### 2008-06-09 papers/OTSPDE/OTSPDE.tex
OTSPDE:  final submission revision of OTSPDE paper.

@@
 % First completed draft: 2008/05/12, Vitamin D, Inc.
+% Submission draft: 2008/06/08, Vitamin D, Inc.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\title{Using Optimal Time Step Selection to Boost the Accuracy of
-       Finite Difference Methods for a Class of 
-       Time Dependent Partial Differential Equations}
+\title{High-Order Accurate Finite Difference Schemes \\
+       Via Optimal Time Step Selection}
 
@@
 \begin{abstract}
-In this article, we present a novel technique for obtaining high-order
+In this article, we present a novel technique for computing high-order
 accurate numerical solutions to time dependent partial differential 
 equations using only formally low-order finite difference stencils and
-time integration schemes.  We achieve high-order accuracy in a simple
-manner by using an optimally chosen time step (and possibly adding a few 
-correction terms to the finite difference scheme).  Through straightforward 
-numerical analysis arguments, we explain the origin of the boost in accuracy 
-and estimate the computational cost of the high-order methods obtained via 
-optimal time step (OTS) selection.  We demonstrate the utility of OTS 
-selection on a wide array of classical PDEs in one and more space dimensions
-and find that it is applicable to linear and semilinear PDEs on both regular 
-and irregular domains. 
+time integration schemes.  We achieve this high-order accuracy in a simple 
+manner by optimally choosing the time step for the low-order finite 
+difference scheme (and possibly adding a few correction terms).  Through 
+straightforward numerical analysis arguments, we explain the origin of the 
+boost in accuracy and estimate the computational cost of the high-order 
+methods obtained via optimal time step (OTS) selection.  We demonstrate the 
+utility of OTS selection on a wide array of classical PDEs in one and more 
+space dimensions and find that it is applicable to linear and semilinear PDEs 
+on both regular and irregular domains. 
 \end{abstract}
@@
 \markboth{KEVIN T. CHU}
-         {HIGH-ORDER FINITE DIFF SCHEMES VIA OPTIMAL TIME STEP SELECTION}
+         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEP SELECTION}
 
@@
 Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
-leveque_book_2002, gko_book}.  Cancellation of higher-order errors (whether 
-fortuitous or intentional) is a key principle underlying the superior 
-accuracy of these methods.
+leveque_book_2002, gko_book}.  Cancellation of higher-order errors is a key 
+principle underlying the superior accuracy of these methods.
 
@@
 numerical methods is often suboptimal because it neglects the potential for 
-achieving high-order accuracy using only simple, low-order schemes.
+achieving high-order accuracy using simple low-order schemes.
 
@@
 schemes.  For both of these methods, examination of higher-order terms allows 
-low-order methods to be transformed into higher order methods with only minor 
+low-order methods to be transformed into high-order methods with only minor 
 modifications to the numerical scheme.
@@
 (\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
-somewhat to blame for this situation because it gives the false impression 
+somewhat to blame for this situation because it gives the impression 
 that the error depends on each numerical parameter independently.  Aside
@@
 
-Using the PDE being solved to reduce discretization error has a long 
+Using the PDE to reduce discretization error has a long 
 tradition in numerical analysis.  Indeed, this idea underlies both the 
@@
 simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
-shows that it is only second-order, which might lead to the conclusion that 
-the nine-point stencil has no advantage over the standard five-point stencil.  
-However, a more careful analysis of the discretization error in light of the 
-PDE reveals that the leading-order error, which is proportional to the 
-bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
+shows that it is only second-order accurate, which might lead to the conclusion 
+that the nine-point stencil has no advantage over the standard five-point 
+stencil.  However, a more careful analysis of the discretization error in 
+light of the PDE reveals that the leading-order error, which is proportional 
+to the bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
 related to the Laplacian of the source term in the Poisson equation.  Thus, 
@@
 Typically, the primary concern when choosing a time step is numerical 
-stability with accuracy being a secondary consideration.  The fact that a 
+stability with accuracy being a secondary consideration.  The idea that a 
 clever choice of time step can be used to dramatically reduce the error is 
 rarely considered.  In general, little research seems to have been done in
-the area of optimizing the parameters of numerical methods (\eg time step and 
-grid spacing) to improve accuracy.  To the best of the author's knowledge, 
-Zhao's method for choosing an optimal value of $\theta$ in the theta-method is 
-the only work that attempts to improve accuracy by modifying numerical 
-parameters.  While he focuses solely on the optimal choice of $\theta$ for a 
-fixed time step, his approach can be viewed as a form of optimal time step 
-selection by reversing the roles of $\theta$ and the time step.
+the area of optimizing the parameters of numerical methods for PDEs (\eg time 
+step and grid spacing) to improve accuracy.  To the best of the author's 
+knowledge, Zhao's method for choosing an optimal value of $\theta$ in the 
+theta-method is the only work that attempts to improve accuracy by modifying 
+numerical parameters.  While he focuses solely on the optimal choice of 
+$\theta$ for a fixed time step, his approach can be viewed as a form of 
+optimal time step selection by reversing the roles of $\theta$ and the time 
+step.
 
@@
 time step selection and traditional adaptive time stepping 
-techniques~\cite{iserles_book,??} that are used in the context of the method 
+techniques~\cite{iserles_book, shampine_2005} that are used in the 
+context of the method 
 of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
@@
 In Section~\ref{sec:optimal_time_step_selection}, we begin by 
-discussing the technique of optimal time step selection in detail and 
-presenting the analysis that shows how it boosts the order of accuracy 
+discussing the technique of optimal time step selection and 
+presenting the analysis that explains how it boosts the order of accuracy 
 for formally low-order numerical schemes.  In 
@@
 of time step can be used to eliminate the leading-order (and possibly 
-higher-order) terms in the error.  In this section we present and analyze the 
-technique of optimal time step selection for finite difference schemes 
+higher-order) terms in the error.  In this section, we present and analyze 
+the technique of optimal time step selection for finite difference schemes 
 constructed to solve evolution equations of the form 
@@
 Because this class of time dependent PDEs is so prevalent and important in 
-many problems of applied mathematics, restricting our attention to problems 
-of this form is not a serious limitation.
+applied mathematics, restricting our attention to problems of this form is 
+not a serious limitation.
 
@@
 derivative operator acting on ${\mathbf u}$.  The second term in the local 
-error arises directly from temporal discretization errors associated with the 
-numerical scheme used to integrate the system of ODEs in time.  The first 
-term arises because $F$ is only an approximation of the spatial derivative 
-operators in the PDE.  As a result, the system of 
-ODEs~(\ref{eq:method_of_lines}) has an error due to the spatial discretization 
-at \emph{every} point in time.  For a spatial discretization error of order 
-$r$, the error in (\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the 
-error accumulated during each time step due to spatial discretization error is 
-$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
-is solved numerically.  Because the method of lines is so frequently used to 
-construct finite difference approximations for time dependent PDEs, we will 
-focus solely on finite difference schemes of this type for the remainder of 
-the paper to keep the discussion simple.  It is straightforward to extend the 
-results we obtain to more general finite difference schemes.
+error (\ref{eq:leading_order_error_model_PDE_MOL}) arises directly from 
+temporal discretization errors associated with the numerical scheme used to 
+integrate the system of ODEs in time.  The first term arises because $F$ is 
+only an approximation of the spatial derivative operators in the PDE.  As a 
+result, the system of ODEs~(\ref{eq:method_of_lines}) has an error due to 
+the spatial discretization at \emph{every} point in time.  For a spatial 
+discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
+$O(\dx^r)$.  Therefore, the error accumulated during each time step due to 
+spatial discretization error is $O(\dt \dx^r)$ regardless of how the system of 
+ODEs (\ref{eq:method_of_lines}) is solved numerically.  Because the method of 
+lines is so frequently used to construct finite difference approximations for 
+time dependent PDEs, we will focus solely on finite difference schemes of this 
+type for the remainder of the paper to keep the discussion simple.  It is 
+straightforward to extend the results we obtain to more general finite 
+difference schemes.
 
@@
 of the spatial and temporal discretization errors, respectively, after the 
-leading-order terms in the error have been removed.  As for the leading-order 
-error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
-error is a direct consequence of the method of lines approach to discretizing 
-the PDE.
+leading-order terms in the error have been eliminated.  As for the 
+leading-order error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of 
+the local error is a direct consequence of the method of lines approach to 
+discretizing the PDE.
 
@@
 
+\subsubsection*{Stability Considerations}
+Thus far, we have not considered the issue of numerical stability of finite
+difference schemes when the time step is set to $\dto$.  For explict time 
+integration schemes, OTS selection loses its value when the optimal time step 
+is greater than the smallest stable time step.  In this situation, we can 
+usually switch to an implicit (or semi-implicit) time integration scheme with 
+the same order of accuracy for which the optimal time step is stable.  Note 
+that a new optimal time step may need to be calculated for the modified finite
+difference sceme.
+
 % TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
 % sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
-\begin{table}[tb]
+\begin{table}[!b]
 \caption{\label{tab:comp_perf_vs_err} 
@@
 
-\subsubsection*{Stability Considerations}
-Thus far, we have not considered the issue of numerical stability of finite
-difference schemes when the time step is set to $\dto$.  For explict time 
-integration schemes, OTS selection loses its value when the optimal time step 
-is greater than the smallest stable time step.  In this situation, we can 
-usually switch to an implicit (or semi-implicit) time integration scheme with 
-the same order of accuracy for which the optimal time step is stable.  Note 
-that a new optimal time step may need to be calculated for the modified finite
-difference sceme.
-
 \subsubsection*{\label{sec:computational_performance} 
@@
 
-\begin{table}[tb]
+\begin{table}[!t]
 \caption{\label{tab:comp_perf_vs_dim}
@@
 Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
-operators because any that arise can be eliminated using the PDE.  Also, 
-there is no explicit spatial discretization error of order $r$ that originates 
-from the source term because it is not differentiated with respect to space 
-when the semi-discrete equations~(\ref{eq:method_of_lines}) are derived from 
-the PDE.
+operators because any that arise can be eliminated using the 
+PDE~(\ref{eq:linear_PDE}).  Also, there is no explicit spatial discretization 
+error of order $r$ that originates from the source term because it is not 
+differentiated with respect to space when the semi-discrete 
+equations~(\ref{eq:method_of_lines}) are derived from the PDE.
 
@@
 numerical solution in the domain interior.  For Dirichlet boundaries, it
-is sufficient to impose boundary conditions by using boundary (or ghost cell) 
-values that have the same order of accuracy as the solution.  
-More effort is required to ensure that Neumann boundary conditions do not
-reduce the order of accuracy of the solution.  In general, Neumann boundary
-conditions require that the accuracy of ghost cells slightly exceed the 
-desired order of accuracy for the solution.  A detailed discussion of 
-methods for imposing boundary conditions is beyond the scope of this paper.  
-We refer interested readers to the thorough analysis presented
-in~\cite{gko_book}.  To avoid complicating the discussion in 
-Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim}, 
-Neumann boundary conditions are imposed by taking advantage of knowledge of 
-the analytical solution.
+is sufficient to impose boundary conditions by using boundary values that have 
+the same order of accuracy as the solution.  More effort is required to ensure 
+that Neumann boundary conditions do not reduce the order of accuracy of the 
+solution.  In general, Neumann boundary conditions require that the accuracy 
+of ghost cells slightly exceed the desired order of accuracy for the 
+solution.  A detailed discussion of methods for imposing boundary conditions 
+is beyond the scope of this paper.  We refer interested readers to the 
+thorough analysis presented in~\cite{gko_book}.  To avoid complicating the 
+discussion in Sections~\ref{sec:applications_1d} 
+and~\ref{sec:applications_multidim}, Neumann boundary conditions are imposed 
+by taking advantage of knowledge of the analytical solution.
 
@@
 generally apply to finite difference schemes for systems of PDEs.  The 
-key issue is that each PDE might have its own optimal time step.  While it 
-would be possible to boost the accuracy of finite difference schemes for 
-systems of PDEs where the optimal time step is the same across the individual 
-PDEs, it is not clear that OTS selection yields any benefit in general.
+key issue here is that each PDE might have its own optimal time step.  While 
+it would be possible to boost the accuracy of finite difference schemes for 
+systems of PDEs where the optimal time step is the same for all of the 
+individual PDEs in the system, it is not clear that OTS selection yields any 
+benefit in general.
 
@@
 difference schemes for the diffusion equation: forward Euler with optimal
-time step, forward Euler with suboptimal time step, and Crank-Nicholson.  
+time step, forward Euler with a suboptimal time step, and Crank-Nicholson.  
 As we can see in Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward 
 Euler with optimal time step is fourth-order accurate while both forward Euler 
-with suboptimal time step and Crank-Nicholson are only second-order accurate.  
-Using the forward Euler scheme with an optimal time step has a clear 
+with a suboptimal time step and Crank-Nicholson are only second-order 
+accurate.  Using the forward Euler scheme with an optimal time step has a clear 
 advantage over both of the other schemes.  Even with as few as 25 grid 
@@
   u^{n+1}_j = u^{n}_j 
-  + D\dt 
-    \left( \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} + f^n_j \right).
+  + \dt 
+    \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right] 
+         + f^n_j \right).
   \label{eq:diffusion_eqn_1d_src_FD_scheme}
@@
   \tu^{n+1}_j = \tu^{n}_j 
-  &+& D\dt \left( \frac{\partial^2 \tu^n}{\px^2} + f^n_j \right)
+  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} + f^n_j \right)
   \nonumber \\
@@
 \scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for the 
+\caption{$L^\infty$ error as a function of number of grid points for 
 two finite difference schemes that solve the diffusion equation: 
@@
 $\dx \rightarrow 0$).
+
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
+\ \ \ \ \
+\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
+\caption{Comparison of numerical solutions for the 1D advection-diffusion
+equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
+optimal time step and correction terms (left) and forward Euler time 
+stepping with suboptimal time step $\dt = 2 \dx^2/3 D$ (right).  
+Both figures compare the analytical solution (solid line) at $t = 1$ on the 
+domain $-10 < x < 10$ with the numerical solution (circles) computed using 
+50 grid points.  For details about the initial/boundary 
+conditions and physical parameters used to compute these solutions, see the
+caption for Figure~\ref{fig:ADE_1d_error}.
+}
+\label{fig:ADE_1d_solns}
+\end{center}
+\end{figure}
+
+To analyze the discretization error, we can reuse 
+results~(\ref{eq:diffusion_eqn_1d_time_err}) 
+and~(\ref{eq:diffusion_eqn_1d_space_err})
+from our analysis of the diffusion equation.  In addition, we need the 
+leading-order discretization error for the central difference approximation 
+of $u_x$, which is given by
+\bea
+  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
+  \frac{\partial \tu^n}{\px} 
+  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
+  + O(\dx^4).
+  \label{eq:ADE_1d_ux_err}
+\eea
+Using the advection-diffusion equation~(\ref{eq:ADE_1d})
+to convert time derivatives into spatial derivatives 
+in~(\ref{eq:diffusion_eqn_1d_time_err}), we find that 
+\bea
+  \tu^{n+1}_j = \tu^{n}_j 
+  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} 
+               + A \frac{\partial \tu^n}{\px} \right)
+  \nonumber \\
+  &+& \frac{\dt^2}{2} 
+      \left( 
+        D^2 \frac{\partial^4 \tu^n}{\px^4} 
+      + 2 D A \frac{\partial^3 \tu^n}{\px^3} 
+      + A^2 \frac{\partial^2 \tu^n}{\px^2} 
+      \right) 
+  + O \left( \dt^3 \right).
+  \label{eq:ADE_1d_time_err_modified}
+\eea
+The local truncation error $\tau^n_j$ for the finite difference 
+scheme~(\ref{eq:ADE_1d_FD_scheme}) can be obtained by the usual procedure 
+of deriving the evolution equation for the error and extracting the source 
+term:
+\bea
+  \tau^n_j &=&
+      \left( \frac{\partial^4 \tu^n}{\px^4} 
+           + \frac{2A}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
+      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
+  \nonumber \\
+  &-& \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
+      + O(\dt \dx^4) + O(\dt^3)
+  \label{eq:ADE_1d_err_eqn}
+\eea
+
+As we discussed in Section~\ref{sec:ots_linear_pde}, boosting the order
+of accuracy for general linear, constant coefficient PDEs requires both 
+the use of an optimal time step and the addition of correction terms.  
+In this case, the optimal time step is $\dto = \dx^2/6D$ and the
+correction term is 
+\beq
+  \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
+  \label{eq:ADE_1d_corr_term}
+\eeq 
+Note that using the standard second-order, central difference stencil to 
+compute the Laplacian in the correction term only makes an 
+$O(\dt^2 \dx^2)$ contribution to $\tau_j^n$, which ensures that no 
+errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into the 
+truncation error.
+With the optimal time step and correction term, the finite difference 
+scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
+Figures~\ref{fig:ADE_1d_error} and \ref{fig:ADE_1d_solns} illustrate the 
+improved accuracy of the finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) 
+when the optimal time step and correction term are used.
 
@@
 
-To analyze the discretization error, we can reuse 
-results~(\ref{eq:diffusion_eqn_1d_time_err}) 
-and~(\ref{eq:diffusion_eqn_1d_space_err})
-from our analysis of the diffusion equation.  In addition, we need the 
-leading-order discretization error for the central difference approximation 
-of $u_x$, which is given by
-\bea
-  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
-  \frac{\partial \tu^n}{\px} 
-  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
-  + O(\dx^4).
-  \label{eq:ADE_1d_ux_err}
-\eea
-Using the advection-diffusion equation~(\ref{eq:ADE_1d})
-to convert time derivatives into spatial derivatives 
-in~(\ref{eq:diffusion_eqn_1d_time_err}), we find that 
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} 
-               + A \frac{\partial \tu^n}{\px} \right)
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
-      \left( 
-        D^2 \frac{\partial^4 \tu^n}{\px^4} 
-      + 2 D A \frac{\partial^3 \tu^n}{\px^3} 
-      + A^2 \frac{\partial^2 \tu^n}{\px^2} 
-      \right) 
-  + O \left( \dt^3 \right).
-  \label{eq:ADE_1d_time_err_modified}
-\eea
-The local truncation error $\tau^n_j$ for the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) can be obtained by the usual procedure 
-of deriving the evolution equation for the error and extracting the source 
-term:
-\bea
-  \tau^n_j &=&
-      \left( \frac{\partial^4 \tu^n}{\px^4} 
-           + \frac{2A}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
-      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-  \nonumber \\
-  &-& \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
-      + O(\dt \dx^4) + O(\dt^3)
-  \label{eq:ADE_1d_err_eqn}
-\eea
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D advection-diffusion
-equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step $\dt = 2 \dx^2/3 D$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 1$ on the 
-domain $-10 < x < 10$ with the numerical solution (circles) computed using 
-50 grid points.  For details about the initial/boundary 
-conditions and physical parameters used to compute these solutions, see the
-caption for Figure~\ref{fig:ADE_1d_error}.
-}
-\label{fig:ADE_1d_solns}
-\end{center}
-\end{figure}
-
-As we discussed in Section~\ref{sec:ots_linear_pde}, boosting the order
-of accuracy for general linear, constant coefficient PDEs requires both 
-the use of an optimal time step and the addition of correction terms.  
-In this case, the optimal time step is $\dto = \dx^2/6D$ and the
-correction term is 
-\beq
-  \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
-  \label{eq:ADE_1d_corr_term}
-\eeq 
-Note that using the standard second-order, central difference stencil to 
-compute the Laplacian in the correction term only makes an 
-$O(\dt^2 \dx^2)$ contribution to $\tau_j^n$, which ensures that no 
-errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into the 
-truncation error.
-With the optimal time step and correction term, the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
-Figures~\ref{fig:ADE_1d_error} and \ref{fig:ADE_1d_solns} illustrate the 
-improved accuracy of the finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) 
-when the optimal time step and correction term are used.
-
 \subsubsection*{First-order Upwind Discretization of Advection Term}
@@
 \subsection{Viscous Burgers Equation}
-% FIGURE PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
+\begin{figure}[!b]
+\begin{center}
+\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
+\ \ \ \ \ 
+\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
+\caption{Comparison of numerical solutions for the 1D viscous Burgers 
+equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
+optimal time step and correction terms (left) and forward Euler time 
+stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
+Both figures compare the analytical solution (solid line) at $t = 2$ on the 
+domain $0 < x < 10$ with the numerical solution (circles) computed using 
+$100$ grid points.  
+For details about the initial/boundary conditions and physical parameters 
+used to compute these solutions, see the caption for 
+Figure~\ref{fig:burgers_1d_error}. 
+}
+\label{fig:burgers_1d_solns}
+\end{center}
+\end{figure}
+
+\begin{figure}[!t]
 \begin{center}
@@
 \eeq
-where $\nu$ is viscosity.  Notice that this problem has the same structure
+where $\nu$ is the viscosity.  Notice that this problem has the same structure
 to the advection-diffusion equation, except that the advection velocity now 
@@
 we will use forward Euler for time integration and second-order central 
-difference approximations for the diffusion and advection terms:
+difference approximations for the diffusion and nonlinear terms:
 \beq
@@
 
-% FIGURE PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \ 
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D viscous Burgers 
-equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 2$ on the 
-domain $0 < x < 10$ with the numerical solution (circles) computed using 
-$100$ grid points.  
-For details about the initial/boundary conditions and physical parameters 
-used to compute these solutions, see the caption for 
-Figure~\ref{fig:burgers_1d_error}. 
-}
-\label{fig:burgers_1d_solns}
-\end{center}
-\end{figure}
-
 \subsection{Fourth-Order Parabolic Equation}
@@
 \beq
-  \frac{\partial u}{\pt} = -\kappa \frac{\partial^4 u}{\px^4} + f(x,t).
+  \frac{\partial u}{\pt} = -\kappa \frac{\partial^4 u}{\px^4} + f(x,t), 
   \label{eq:4th_order_parabolic_eqn_1d}
 \eeq
+where $\kappa > 0$.
 From our previous examples, we know that optimal time step selection requires 
@@
 \bea
-  u^{n+1}_j = u^{n}_j - \kappa \dt B u^n, 
+  u^{n+1}_j = u^{n}_j + \dt \left( -\kappa B u^n + f_j^n \right), 
   \label{eq:4th_order_parabolic_eqn_1d_FD_scheme}
@@
 the bilaplacian.  Using each method, we solve the fourth-order parabolic
-equation with $\kappa = 1$ to time $t = 0.001$ on the domain $0 < x < 1
+equation with $\kappa = 1$ and 
+$f(x,t) = \sin(2 \pi x)\left(-20000 \pi \sin(10000\pi t) 
+                           + 32\pi^4\cos(10000\pi t)\right)
+        + \sin(5 \pi x)\left( 100 \pi^2 \exp\left(-100\pi^2 t \right)
+           + 625 \pi^4 \left(2 - \exp\left(-100\pi^2 t \right) \right)
+        \right)$
+to time $t = 0.001$ on the domain $0 < x < 1
 $ subject to the boundary conditions $u(0,t) = u(1,t) = 0$ and the initial 
@@
 \eeq
-This method is formally first-order in both space and time.
+This method is formally first-order in both space and time with a 
+stability constraint of $\dt < \left( A_x/dx + A_y/dy \right)^{-1}$.
 
@@
 To see that the correction term can be considered as part of the spatial 
-discretization scheme, we can write out the complete finite difference scheme 
+discretization stencil, we can write out the complete finite difference scheme 
 including the correction term:
@@
 
-In two space dimensions, the diffusion equation may be written as
-\beq
-  \frac{\partial u}{\pt} = D \nabla^2 u,
+In two space dimensions, the diffusion equation (with source term) may be 
+written as
+\beq
+  \frac{\partial u}{\pt} = D \nabla^2 u + f(x,y,t),
   \label{eq:diffusion_eqn_2d}
 \eeq
-where $\nabla^2$ is the Laplacian operator and $D$ is the diffusion constant. 
+where $\nabla^2$ is the Laplacian operator, $D$ is the diffusion constant,
+and $f$ is the source term.
 While there are many finite difference schemes that are used to solve the 
@@
 which assumes that the grid spacing in the $x$ and $y$ directions are
-equal.  Common choices for the time integration method include forward Euler
-and the Crank-Nicholson method.  Each scheme has its advantages and
-disadvantages.  The forward Euler scheme is very easy to implement but is 
-only first-order accurate and suffers from stability constraints on the time 
-step; the Crank-Nicholson scheme is second-order accurate and unconditionally 
-stable but requires the solution of a linear system of equations at each time 
-step.  The forward Euler scheme using the five-point stencil is given by
+equal.  The forward Euler scheme using the five-point stencil is given by
 \bea
   u^{n+1}_j &=& u^{n}_j 
-  + D \dt 
-    \left( \frac{u^{n}_{i,j+1} + u^{n}_{i,j-1}
-         +u^{n}_{i+1,j} + u^{n}_{i-1,j} - 4u^{n}_{i,j}}{\dx^2}
+  + \dt 
+    \left( D \left[\frac{u^{n}_{i,j+1} + u^{n}_{i,j-1}
+             +u^{n}_{i+1,j} + u^{n}_{i-1,j} - 4u^{n}_{i,j}}{\dx^2} \right]
+         + f_{i,j}^n
     \right).
@@
   u^{n+1}_{i,j} = u^{n}_{i,j}
-  + D \dt L^{9pt} u
+  + \dt \left( D L^{9pt} u + f_{i,j}^n \right)
   \label{eq:diffusion_eqn_2d_9pt_scheme}
@@
 the Laplacian.  Using each method, we solve the 2D diffusion equation with 
-$D = 0.25$ and source term 
+$D = 1/4$ and source term 
 $f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
@@
           - 24 D \pi^2 \sin(5 \pi x) \sin(\pi y) 
-            \exp \left(-2 D \pi^2 t \right)$.
+            \exp \left(-2 D \pi^2 t \right)$
 to time $t = 0.1$ on the domain $0 < x, y < 1$ subject 
@@
 \subsubsection{Irregular Domains}
-All of the previous examples share an important feature: they are all
-defined on simple, rectangular domains.  Thus, any of several methods 
+All of the examples we have considered so far share an important feature: they 
+are all defined on simple, rectangular domains.  Thus, any of several methods 
 could have been used to obtain solutions with similar or even higher accuracy.
@@
 need for special stencils at grid points near the irregular boundary, we adopt 
-the `ghost cell' approach for imposing boundary conditions~\cite{fedkiw_1999, 
-osher_fedkiw_book, gibou_2005}.  The basic idea is to set the value of $u$ at 
+the `ghost cell' approach for imposing boundary conditions~\cite{gibou_2005, 
+fedkiw_1999, osher_fedkiw_book}.  The basic idea is to set the value of $u$ at 
 any grid point that lies immediately outside of the domain (and thus may be
@@
 extrapolation of the values of $u$ across the interface.  A cubic Lagrange 
-extrapolant is constructed for each edge ghost cell using a point on the 
+interpolant is constructed for each edge ghost cell using a point on the 
 domain boundary and three interior grid points that lie along the line 
@@
 locate boundary points was unnecessary for achieving high-order accuracy in 
-the computed solution.}.
-Because the quality of the extrapolant deteriorates if the boundary point is 
-too close to any of the interior grid 
-points used to construct the Lagrange extrapolant, we follow 
-\cite{gibou_2005} and choose the interior grid points so that the nearest
-interior grid point is sufficiently far from the boundary point.  
+the computed solution.}.  Because the quality of the interpolant deteriorates 
+if the boundary point is too close to any of the interior grid points used to 
+construct the Lagrange extrapolant, we follow \cite{gibou_2005} and choose the 
+interior grid points so that the nearest interior grid point is sufficiently 
+far from the boundary point.  Without this procedure, the errors introduced 
+in the values of the ghost cells lead to instability of the numerical method.
 Gibou~\etal~suggest shifting the interpolation points by one grid point 
@@
 presented in this paper, this threshold was too low.  We found that better 
-results were obtained if we used an $O(\dx)$ threshold.
+stability was obtained if we used an $O(\dx)$ threshold.
 
@@
 straightforward to compute high-order accurate solutions of the 2D diffusion 
-equation on irregular domains.  We simply use the OTS forward euler scheme 
+equation on irregular domains.  We simply use the OTS forward Euler scheme 
 designed for regular domains for grid points in the interior of the domain
@@
 irregular domain: forward Euler with optimal time step and correction 
-terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/8 D$
+terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/4 D$
 (squares).  Both methods solve the 2D diffusion equation described in the 
@@
 discretization error.  For 3D problems, a catalog of several such stencils is 
-available in~\cite{patra_2005}.  For higher dimensional problems, using 
-symbolic algebra software to derive finite difference stencils with desired 
-discretization error properties can be helpful~\cite{patra_2005, gupta_1998}.  
+available in~\cite{patra_2005}.   For higher dimensional problems, one simple
+approach for deriving an isotropic finite difference stencil for the Laplacian
+would be to start with the generalization of the standard five point stencil
+for the 2D Laplacian and add finite difference approximations for the missing 
+cross terms that are required to make the leading-order error isotropic 
+(\ie proportional to the bilaplacian of the solution).  Because the missing 
+terms are all of the form 
+\beq
+\frac{\dx^2}{6} \left(\frac{\partial^4 u}{\partial x_i^2 \partial x_j^2}\right),
+\eeq
+this approach is relatively easy to use.  To derive finite difference 
+approximations with more general or complex properties, symbolic algebra 
+software can be helpful~\cite{patra_2005, gupta_1998}.
+
 Once an acceptable finite difference approximation for the Laplacian has been 
@@
 whose nearest interior neighbor is on the opposite corner of a grid cell.
-Symbolic algebra programs can be helpful when deriving methods for 
+Here, too, symbolic algebra programs can be helpful when deriving methods for 
 extrapolating interior and boundary values to ghost cells.
@@
 selection to a wide range of problems.  We showed how OTS selection can be 
-used to very easily to obtain high-order accurate solutions to most linear 
+used to very easily obtain high-order accurate solutions to many linear 
 and semilinear PDEs in any number of space dimensions on both regular
@@
 
-Optimal time step selection is a simple example of the more general problem
+Optimal time step selection is a simple example of the more general technique
 of optimally selecting numerical parameters to boost the accuracy of 
@@
 \section*{Acknowledgments}
-The author gratefully acknowledges the support of the Institute for 
-High-Performance Computing (IHPC) in Singapore and Vitamin D, Inc.
-The author would like to thank P. Fok and ?? for enlightening discussions 
-and suggestions on the manuscript.
+The author gratefully acknowledges the support of Vitamin D, Inc.
+and the Institute for High-Performance Computing (IHPC) in Singapore. 
+The author would like to thank P. Fok, M. Prodanovic, and A. Chiu for 
+helpful suggestions on the manuscript.  The author particularly thanks P. Fok 
+for many enlightening discussions.
 

### 2008-06-09 papers/OTSPDE/OTSPDE.tex
OTSPDE: * changed color scheme in figures so that the graphs show up better in   greyscale * minor changes to keywords and short title in manuscript * added submission information directory

@@
 \begin{keywords}
-optimal time step, finite difference methods, high-order accurate numerical 
+optimal time step, finite difference schemes, high-order accurate numerical 
 methods, time dependent PDEs
@@
 \markboth{KEVIN T. CHU}
-         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEP SELECTION}
+         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEPS}
 

### 2008-06-09 src/1d/advection_diffusion_eqn/genFigs_AdvDiffEqn1d.m

@@ -195,18 +195,18 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(20,2e-7,order_str);
 
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(200,4e-2,order_str);
 

### 2008-06-09 src/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m

@@ -194,17 +194,17 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(45,1e-5,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(250,1e-1,order_str);
 

### 2008-06-09 src/1d/diffusion_eqn/genFigs_DiffusionEqn1d_NoSrc.m

@@ -252,25 +252,25 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(20,1.5e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(100,2e-3,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_CN(1)+P_CN(2)),'k');
 hold on;
-plot(grid_sizes,err_CN, 'rd', ...
+plot(grid_sizes,err_CN, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','r');
+     'MarkerFaceColor','m');
 order_str = sprintf('Crank-Nicholson\nOrder = %1.1f', order_CN);
 text(200,2e-7,order_str);
 
@@ -288,9 +288,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), ...
        'k');
 hold on;
-loglog(err_FE_OTS(3:end), comp_time_FE_OTS(3:end), 'go', ...
+loglog(err_FE_OTS(3:end), comp_time_FE_OTS(3:end), 'bo', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','g');
+       'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(4e-10,4e-3,order_str);
 
@@ -298,9 +298,9 @@ loglog(err_plot, ...
[…]

### 2008-06-09 src/1d/diffusion_eqn/genFigs_DiffusionEqn1d_Src.m

@@ -272,18 +272,18 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler OTS (correction)\nOrder = %1.1f', ...
   order_FE_OTS);
 text(15,4.5e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS_no_corr(1)+P_FE_OTS_no_corr(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS_no_corr, 'm^', ...
+plot(grid_sizes,err_FE_OTS_no_corr, 'g^', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','m');
+     'MarkerFaceColor','g');
 order_str = sprintf('Forward Euler OTS (no correction)\nOrder = %1.1f', ...
   order_FE_OTS_no_corr);
 text(30,3e-2,order_str);
@@ -316,9 +316,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), ...
        'k');
 hold on;
-loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'go', ...
+loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'bo', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','g');
+       'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler OTS (correction)\nSlope = %1.1f', ...
   comp_time_exp_FE_OTS);
 text(3e-10,3e-2,order_str);
@@ -328,9 +328,9 @@ loglog(err_plot, ...
           +P_comp_time_FE_OTS_no_corr(2)), ...
        'k');
 hold on;
-loglog(err_FE_OTS_no_corr(2:end), comp_time_FE_OTS_no_corr(2:end), 'm^', ...
+loglog(err_FE_OTS_no_corr(2:end), comp_time_FE_OTS_no_corr(2:end), 'g^', ...
        'MarkerSize',14, ...
[…]

### 2008-06-09 src/1d/fourth_order_parabolic/genFigs_4thOrderParabolicEqn1d.m

@@ -202,18 +202,18 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(20,2e-8,order_str);
 
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(150,4e-3,order_str);
 
@@ -231,9 +231,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), ...
        'k');
 hold on;
-loglog(err_FE_OTS, comp_time_FE_OTS, 'go', ...
+loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','g');
+       'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(2e-8,5e-2,order_str);
 
@@ -241,9 +241,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), ...
        'k');
 hold on;
-loglog(err_FE, comp_time_FE, 'bs', ...
+loglog(err_FE, comp_time_FE, 'rs', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','b');
+       'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nSlope = %1.1f', comp_time_exp_FE);
 text(5e-4,2e2,order_str);
 

### 2008-06-09 src/2d/advection_eqn/demo.m

@@ -23,7 +23,7 @@ A_y = -3.0;  % flow speed in y-direction
 
 % time integration parameters
 % NOTE: t_init = 0.0
-t_final = 1.0;
+t_final = 2.0;
 
 % set dx and dt
 N = 100;
@@ -31,6 +31,7 @@ x_lo = -10;
 x_hi =  10;
 dx = (x_hi-x_lo)/N;
 dt_FE = 0.5*dx/( abs(A_x) + abs(A_y) );  % suboptimal time step
+dt_FE = 0.99*dx/( abs(A_x) + abs(A_y) );  % suboptimal time step
 
 % set simulation parameters
 debug_on  = 1;

### 2008-06-09 src/2d/advection_eqn/genFigs_AdvectionEqn2d.m

@@ -21,7 +21,7 @@ set(0,'DefaultLineLineWidth',2)
 set(0,'DefaultTextFontSize',18,'DefaultTextFontName','Helvetica')
 
 % set print format
-use_color_figures = 1;
+use_color_figures = 0;
 print_suffix = 'eps';
 if use_color_figures
   print_format = 'epsc';
@@ -103,7 +103,14 @@ u_plot = interp2(reshape(X_FE_OTS,Ny_FE_OTS,N+1), ...
                  reshape(Y_FE_OTS,Ny_FE_OTS,N+1), ...
                  reshape(u_FE_OTS,Ny_FE_OTS,N+1), ...
                  X_plot, Y_plot, '*nearest');
-mesh(X_plot, Y_plot, u_plot);
+if (use_color_figures)
+  mesh(X_plot, Y_plot, u_plot);
+  colormap('default');
+else
+  % black and white
+  mesh(X_plot, Y_plot, u_plot, ones(size(X_plot)));
+  colormap([1 1 1; 0 0 0]);
+end
 axis([-10 10 -10 10 -0.1 1.1])
 xlabel('x'); ylabel('y'); 
 %title('Forward Euler OTS Solution');
@@ -118,7 +125,14 @@ u_plot = interp2(reshape(X_FE,N+1,N+1), ...
                  reshape(Y_FE,N+1,N+1), ...
                  reshape(u_FE,N+1,N+1), ...
                  X_plot, Y_plot, '*nearest');
-mesh(X_plot, Y_plot, u_plot);
+if (use_color_figures)
+  mesh(X_plot, Y_plot, u_plot);
+  colormap('default');
+else
+  % black and white
+  mesh(X_plot, Y_plot, u_plot, ones(size(X_plot)));
+  colormap([1 1 1; 0 0 0]);
+end
 axis([-10 10 -10 10 -0.1 1.1])
 xlabel('x'); ylabel('y'); 
 %title('Forward Euler Solution');

### 2008-06-09 src/2d/diffusion_eqn/irregular_domain/circle_domain/genFigs_DiffusionEqn2d_CircleDomain.m

@@ -26,6 +26,7 @@ set(0,'DefaultTextFontSize',18,'DefaultTextFontName','Helvetica')
 
 % set print format
 use_color_figures = 1;
+use_color_for_meshplot = 0;
 print_suffix = 'eps';
 if use_color_figures 
   print_format = 'epsc';
@@ -228,17 +229,17 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot, exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes, err_FE_OTS, 'go', ...
+plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
-     'MarkerFaceColor', 'g');
+     'MarkerFaceColor', 'b');
 order_str = sprintf('Forward Euler OTS\nOrder = %1.1f', order_FE_OTS);
 text(12,5e-4,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
 hold on;
-plot(grid_sizes, err_FE, 'bs', ...
+plot(grid_sizes, err_FE, 'rs', ...
      'MarkerSize', 14, ...
-     'MarkerFaceColor', 'b');
+     'MarkerFaceColor', 'r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(250,1e-2,order_str);
 
@@ -258,17 +259,17 @@ err_plot = [1e-10 1e0];
 loglog(err_plot, ...
   exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), 'k');
 hold on;
-loglog(err_FE_OTS, comp_time_FE_OTS, 'go', ...
+loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
-       'MarkerFaceColor', 'g');
+       'MarkerFaceColor', 'b');
 order_str = sprintf('Forward Euler OTS\nOrder = %1.1f', comp_time_exp_FE_OTS);
 text(5e-8,5,order_str);
 
 loglog(err_plot, exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), 'k');
 hold on;
-loglog(err_FE, comp_time_FE, 'bs', ...
+loglog(err_FE, comp_time_FE, 'rs', ...
        'MarkerSize', 14, ...
[…]

### 2008-06-09 src/2d/diffusion_eqn/irregular_domain/starfish_domain/genFigs_DiffusionEqn2d_StarfishDomain.m

@@ -26,6 +26,7 @@ set(0,'DefaultTextFontSize',18,'DefaultTextFontName','Helvetica')
 
 % set print format
 use_color_figures = 1;
+use_color_for_meshplot = 0;
 print_suffix = 'eps';
 if use_color_figures 
   print_format = 'epsc';
@@ -227,17 +228,17 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot, exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)), 'k');
 hold on;
-plot(grid_sizes, err_FE_OTS, 'go', ...
+plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
-     'MarkerFaceColor', 'g');
+     'MarkerFaceColor', 'b');
 order_str = sprintf('Forward Euler OTS\nOrder = %1.1f', order_FE_OTS);
 text(30,1e-5,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
 hold on;
-plot(grid_sizes, err_FE, 'bs', ...
+plot(grid_sizes, err_FE, 'rs', ...
      'MarkerSize', 14, ...
-     'MarkerFaceColor', 'b');
+     'MarkerFaceColor', 'r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(250,1e-2,order_str);
 
@@ -257,17 +258,17 @@ err_plot = [1e-10 1e0];
 loglog(err_plot, ...
   exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), 'k');
 hold on;
-loglog(err_FE_OTS, comp_time_FE_OTS, 'go', ...
+loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
-       'MarkerFaceColor', 'g');
+       'MarkerFaceColor', 'b');
 order_str = sprintf('Forward Euler OTS\nOrder = %1.1f', comp_time_exp_FE_OTS);
 text(5e-8,5,order_str);
 
 loglog(err_plot, exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), 'k');
 hold on;
-loglog(err_FE, comp_time_FE, 'bs', ...
+loglog(err_FE, comp_time_FE, 'rs', ...
        'MarkerSize', 14, ...
[…]

### 2008-06-09 src/2d/diffusion_eqn/irregular_domain/starfish_domain/genPresoFigs_DiffusionEqn2d_StarfishDomain.m

@@ -215,19 +215,19 @@ comp_time_exp_FE = P_comp_time_FE(1);
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 figure(1); clf;
 N_plot = [10 1000];
-loglog(N_plot, exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)), 'g');
+loglog(N_plot, exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)), 'b');
 hold on;
-plot(grid_sizes, err_FE_OTS, 'go', ...
+plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
-     'MarkerFaceColor', 'g');
+     'MarkerFaceColor', 'b');
 order_str = sprintf('Forward Euler OTS\nOrder = %1.1f', order_FE_OTS);
 text(30,1e-5,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
 hold on;
-plot(grid_sizes, err_FE, 'bs', ...
+plot(grid_sizes, err_FE, 'rs', ...
      'MarkerSize', 14, ...
-     'MarkerFaceColor', 'b');
+     'MarkerFaceColor', 'r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(250,1e-2,order_str);
 
@@ -245,20 +245,20 @@ print([fig_dir, '/', filename], format_str);
 figure(2); clf;
 err_plot = [1e-10 1e0];
 loglog(err_plot, ...
-  exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), 'g');
+  exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), 'b');
 hold on;
-loglog(err_FE_OTS, comp_time_FE_OTS, 'go', ...
+loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
-       'MarkerFaceColor', 'g');
+       'MarkerFaceColor', 'b');
 order_str = sprintf('Forward Euler OTS\nOrder = %1.1f', ...
                      comp_time_exp_FE_OTS);
 text(5e-8,5,order_str);
 
[…]

### 2008-06-09 src/2d/diffusion_eqn/square_domain/genFigs_DiffusionEqn2d_NoSrc.m

@@ -213,25 +213,25 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(13,1.5e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(100,7e-4,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_CN(1)+P_CN(2)),'k');
 hold on;
-plot(grid_sizes,err_CN, 'rd', ...
+plot(grid_sizes,err_CN, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','r');
+     'MarkerFaceColor','m');
 order_str = sprintf('Crank-Nicholson\nOrder = %1.1f', order_CN);
 text(140,9e-8,order_str);
 
@@ -249,9 +249,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), ...
        'k');
 hold on;
-loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'go', ...
+loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'bo', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','g');
+       'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(5e-11,4e-3,order_str);
 
@@ -259,9 +259,9 @@ loglog(err_plot, ...
[…]

### 2008-06-09 src/2d/diffusion_eqn/square_domain/genFigs_DiffusionEqn2d_Src.m

@@ -223,25 +223,25 @@ figure(1); clf;
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'k');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(18,4e-8,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(100,3e-2,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_CN(1)+P_CN(2)),'k');
 hold on;
-plot(grid_sizes,err_CN, 'rd', ...
+plot(grid_sizes,err_CN, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','r');
+     'MarkerFaceColor','m');
 order_str = sprintf('Crank-Nicholson\nOrder = %1.1f', order_CN);
 text(150,7e-6,order_str);
 
@@ -259,9 +259,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), ...
        'k');
 hold on;
-loglog(err_FE_OTS, comp_time_FE_OTS, 'go', ...
+loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','g');
+       'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(5e-8,5e-2,order_str);
 
@@ -269,9 +269,9 @@ loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), ...
        'k');
[…]

### 2008-06-09 src/2d/diffusion_eqn/square_domain/genPresoFigs_DiffusionEqn2d_Src.m

@@ -215,27 +215,27 @@ comp_time_exp_CN = P_comp_time_CN(1);
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 figure(1); clf;
 N_plot = [10 1000];
-loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'g');
+loglog(N_plot,exp(log(N_plot)*P_FE_OTS(1)+P_FE_OTS(2)),'b');
 hold on;
-plot(grid_sizes,err_FE_OTS, 'go', ...
+plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','g');
+     'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS)\nOrder = %1.1f', order_FE_OTS);
 text(18,4e-8,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
-plot(grid_sizes,err_FE, 'bs', ...
+plot(grid_sizes,err_FE, 'rs', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','b');
+     'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nOrder = %1.1f', order_FE);
 text(100,3e-2,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_CN(1)+P_CN(2)),'k');
 hold on;
-plot(grid_sizes,err_CN, 'rd', ...
+plot(grid_sizes,err_CN, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','r');
+     'MarkerFaceColor','m');
 order_str = sprintf('Crank-Nicholson\nOrder = %1.1f', order_CN);
 text(150,7e-6,order_str);
 
@@ -251,11 +251,11 @@ figure(2); clf;
 err_plot = [1e-10 1e0];
 loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS(1)+P_comp_time_FE_OTS(2)), ...
-       'g');
+       'b');
 hold on;
-loglog(err_FE_OTS, comp_time_FE_OTS, 'go', ...
+loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
-       'MarkerFaceColor','g');
+       'MarkerFaceColor','b');
[…]

### 2008-08-10 papers/OTSPDE-Systems/OTSPDE-Systems.tex
OTSPDE: * Checked in place holders for upcoming papers. * Updated information on potential journals to submit to.

--- 2008-05-14 papers/OTSPDE/OTSPDE.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Manuscript started: 2007/09/01
-% First completed draft: 2008/05/12, Vitamin D, Inc.
+% Manuscript started: 2008/09/0?
+% First completed draft: 2008/??/??, Vitamin D, Inc.
+% Submission draft: 2008/??/??, Vitamin D, Inc.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\title{Using Optimal Time Step Selection to Boost the Accuracy of
-       Finite Difference Methods for a Class of 
-       Time Dependent Partial Differential Equations}
+\title{High-Order Accurate Finite Difference Schemes \\
+       for Systems of PDEs via Optimal Time Step Selection}
 
@@
 \begin{abstract}
-In this article, we present a novel technique for obtaining high-order
-accurate numerical solutions to time dependent partial differential 
-equations using only formally low-order finite difference stencils and
-time integration schemes.  We achieve high-order accuracy in a simple
-manner by using an optimally chosen time step (and possibly adding a few 
-correction terms to the finite difference scheme).  Through straightforward 
-numerical analysis arguments, we explain the origin of the boost in accuracy 
-and estimate the computational cost of the high-order methods obtained via 
-optimal time step (OTS) selection.  We demonstrate the utility of OTS 
-selection on a wide array of classical PDEs in one and more space dimensions
-and find that it is applicable to linear and semilinear PDEs on both regular 
-and irregular domains. 
+place holder...
 \end{abstract}
@@
 \begin{keywords}
-optimal time step, finite difference methods, high-order accurate numerical 
-methods, time dependent PDEs
+optimal time step, finite difference schemes, high-order accurate numerical 
+methods, time dependent PDEs, systems of PDEs
 \end{keywords}
@@
 \begin{AMS}
-65-02, 65M06, 65M12, 65M20
+65-02, 65M06, 65M12, 65M20, ??
 \end{AMS}
@@
 \markboth{KEVIN T. CHU}
-         {HIGH-ORDER FINITE DIFF SCHEMES VIA OPTIMAL TIME STEP SELECTION}
+         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEPS}
 
@@
 
-There are many high-order accurate methods that are simple combinations or 
-modifications of low-order schemes.  A few well-known examples include 
-Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
-nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
-Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
-leveque_book_2002, gko_book}.  Cancellation of higher-order errors (whether 
-fortuitous or intentional) is a key principle underlying the superior 
-accuracy of these methods.
+\section{\label{sec:summary} Summary}
 
-More commonly, high-order accurate numerical methods are constructed by 
-requiring that the discretization error possesses the desired formal order of 
-accuracy for \emph{all} of the independent variables in the PDE.  While this 
-constraint is certainly sufficient for high-order accuracy, it is by no means 
-necessary.  This conventional approach for constructing numerical methods 
-has its roots in the classical numerical analysis of PDEs, which tends to 
-analyze the discretization of the terms in the PDE individually.  Within this 
-framework, it is natural to seek high-order accuracy by ensuring that the 
-numerical scheme is simultaneously high-order accurate for all independent 
-variables.  Unfortunately, this common approach for developing and analyzing 
-numerical methods is often suboptimal because it neglects the potential for 
-achieving high-order accuracy using only simple, low-order schemes.
+\section*{Acknowledgments}
+The author gratefully acknowledges the support of Vitamin D, Inc.
+and the Institute for High-Performance Computing (IHPC) in Singapore. 
+The author would like to thank K.-H. Chiam, ?? 
+for helpful suggestions on the manuscript.  
+The author particularly thanks K.-H. Chiam for pointing him to the 
+mathematical biology literature for interesting examples of reaction-diffusion 
+systems.
 
-Perhaps the greatest weakness of the standard method for constructing 
-high-order numerical methods is the use of ``big-O" notation, which hides 
-the details of the discretization error.  The information lost in the 
-``big-O" notation eliminates the possibility of reducing numerical error by 
-taking advantage of relationships between the terms in the error.  The 
-modified nine-point formula for the Poisson equation~\cite{iserles_book} 
-and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
-leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
-higher-order terms in the discretization error when analyzing numerical 
-schemes.  For both of these methods, examination of higher-order terms allows 
-low-order methods to be transformed into higher order methods with only minor 
-modifications to the numerical scheme.
-
-The usual approach for developing numerical methods has another significant
-shortcoming -- it places no value on the optimization of numerical parameters 
-(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
-somewhat to blame for this situation because it gives the false impression 
-that the error depends on each numerical parameter independently.  Aside
-from constraints needed for numerical stability, ``big-O" notation makes 
-it appear as though any combination of numerical parameters is equally good 
-as long as they are all ``small enough.''  For some problems, however, 
-optimizing the numerical parameters can significantly increase the accuracy
-of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
-that choosing an appropriate value for $\theta$ as a function of the time 
-step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
-the weight factor and the theta-method as the single-step trapezoidal 
-method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
-element solutions to the diffusion equation.
-
-In this paper, we explore a novel technique for constructing high-order 
-finite difference methods for time dependent PDEs using only formally 
-low-order stencils and time integration schemes.  The basic idea is that a 
-carefully chosen time step (and possibly the addition of a few correction 
-terms) can eliminate low-order terms in the discretization error, allowing a 
-formally low-order method to deliver high-order accuracy.  The optimal time 
-step is calculated in a straightforward manner by using knowledge of the 
-PDE to eliminate the leading-order terms in the discretization error.  
-Optimal time step selection combines the well-known approach of using 
-the PDE to reduce numerical error with the less-common notion that optimal 
-choice of numerical parameters can boost the effectiveness of algorithmically 
-simple and computationally low-cost methods.
-
-Using the PDE being solved to reduce discretization error has a long 
-tradition in numerical analysis.  Indeed, this idea underlies both the 
-modified nine-point formula for the Poisson equation~\cite{iserles_book} and 
-the Lax-Wendroff method for the linear advection 
-equation~\cite{leveque_book_1992, leveque_book_2002, gko_book}.  
-In the case of the modified nine-point formula for the Poisson equation, a 
-simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
-shows that it is only second-order, which might lead to the conclusion that 
-the nine-point stencil has no advantage over the standard five-point stencil.  
-However, a more careful analysis of the discretization error in light of the 
-PDE reveals that the leading-order error, which is proportional to the 
-bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
-related to the Laplacian of the source term in the Poisson equation.  Thus, 
-the numerical scheme is easily boosted to fourth-order by simply adding a 
-correction term that cancels out the leading-order error.   
-Similarly, use of the PDE to eliminate higher-order discretization errors also 
-plays a critical role in the Lax-Wendroff method because it allows the second 
-derivative of the solution in time to be written in terms of the second 
-derivative in space.  This observation transforms an unstable, first-order 
-scheme based on a central difference approximation of 
-$\partial/\partial x$ into a stable, second-order 
-scheme~\cite{leveque_book_1992, leveque_book_2002}.
-
-Optimal time step selection as a technique for improving the overall accuracy 
-of numerical schemes for time dependent PDEs appears to be a novel idea.
-Typically, the primary concern when choosing a time step is numerical 
-stability with accuracy being a secondary consideration.  The fact that a 
-clever choice of time step can be used to dramatically reduce the error is 
-rarely considered.  In general, little research seems to have been done in
-the area of optimizing the parameters of numerical methods (\eg time step and 
-grid spacing) to improve accuracy.  To the best of the author's knowledge, 
-Zhao's method for choosing an optimal value of $\theta$ in the theta-method is 
-the only work that attempts to improve accuracy by modifying numerical 
-parameters.  While he focuses solely on the optimal choice of $\theta$ for a 
-fixed time step, his approach can be viewed as a form of optimal time step 
-selection by reversing the roles of $\theta$ and the time step.
-
-It is worth spending a few moments to draw the distinction between optimal 
-time step selection and traditional adaptive time stepping 
-techniques~\cite{iserles_book,??} that are used in the context of the method 
-of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
-selection may appear to be just a crude version of adaptive time step control.
-However, the two methods are actually very different.  While both methods seek 
-to reduce numerical error by controlling the time step, the numerical errors 
-they control are fundamentally different.  Adaptive time stepping can only 
-be used to reduce temporal discretization errors because it controls the 
-errors that arise when solving the coupled system of ODEs for the 
-semi-discrete approximation to the PDE; the spatial discretization errors 
-are completely controlled by the finite difference scheme selected to 
-approximate the spatial derivatives.  
-In contrast, optimal time step selection simultaneously reduces 
-\emph{both} the spatial and temporal discretization errors because it
-uses information from the PDE to choose the time step.  Thus, optimal time 
-step selection can potentially produce superior results compared with 
-traditional adaptive time stepping approaches for integrating the 
-semi-discrete equations associated with time dependent PDEs.  Another 
-important distinction between the two techniques is that optimal time step 
-selection uses a \emph{fixed} time step which makes it significantly less 
-computationally complex than adaptive time stepping methods.
-
-As a technique for constructing numerical methods, optimal time step selection 
-has several desirable features.  Like any technique that leads to high-order 
-methods, optimal time step selection produces schemes that greatly reduce the 
-computational cost required to obtain a numerical solution of a desired 
-accuracy.  However, its real power comes from the fact that it achieves
-high-order accuracy while being based solely on very simple finite difference 
-schemes.  For instance, we will see how optimal time step selection makes 
-it possible to achieve 4th-order accuracy for the diffusion equation using only 
-a second-order stencil for the Laplacian and forward Euler time integration.  
-Moreover, high-order accuracy is not restricted to problems on simple, 
-rectangular domains; irregular domains are straightforward to handle by
-appropriately setting the values in ghost cells~\cite{gibou_2005, fedkiw_1999, 
-osher_fedkiw_book}.  Optimal time step selection is even beneficial when 
-using finite difference schemes to solve some nonlinear PDEs.  While there 
-are certainly limitations to optimal time step selection, we will see that 
-there are many problems where it is useful.
-
-The remainder of this paper is organized as follows.  
-In Section~\ref{sec:optimal_time_step_selection}, we begin by 
-discussing the technique of optimal time step selection in detail and 
-presenting the analysis that shows how it boosts the order of accuracy 
-for formally low-order numerical schemes.  In 
-Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
-optimal time selection to problems in one space dimension by applying it to 
-finite difference schemes for several classical linear and nonlinear PDEs. 
-In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
-to improve the accuracy of simple finite difference schemes for the linear 
-advection equation and diffusion equation in multiple space dimensions.  
-To illustrate the ease with which high-order accuracy can be achieved for
-problems on arbitrary domains, the 2D diffusion equation is solved on both 
-the regular and irregular domains.  
-For each PDE in Sections~\ref{sec:applications_1d} and 
-\ref{sec:applications_multidim}, we provide the numerical analysis required 
-to calculate the optimal time step size and order of accuracy.  
-Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.
-
-\section{\label{sec:optimal_time_step_selection} 
-         Optimal Time Step (OTS) Selection}
-Optimal time step selection is not by itself a method for constructing 
-finite difference schemes.  Rather it is a way to enhance to the performance
-of existing finite difference schemes by carefully choosing the time step to 
-eliminate low-order numerical errors.  The two basic ideas underlying 
-optimal time step selection are (1) the PDE provides valuable insight into the 
-discretization errors for finite difference schemes and (2) a judicious choice 
-of time step can be used to eliminate the leading-order (and possibly 
-higher-order) terms in the error.  In this section we present and analyze the 
-technique of optimal time step selection for finite difference schemes 
-constructed to solve evolution equations of the form 
-\beq
-  \frac{\partial u}{\partial t} = 
-    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
-\eeq
-where the right hand side is an arbitrary function of the solution $u$, its 
-spatial derivatives, and the independent variable $x$.
-Because this class of time dependent PDEs is so prevalent and important in 
-many problems of applied mathematics, restricting our attention to problems 
-of this form is not a serious limitation.
-
-\subsection{\label{sec:ots_model_pde} 
-            Optimal Time Step for a Model PDE}
-Let us begin by considering a finite difference approximation for the
-following time dependent PDE in one spatial dimension: 
-\beq
-  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
-  \label{eq:model_PDE}
-\eeq
-where $u$ is the solution to the PDE, $n$ is the spatial order of the 
-PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
-is well-posed.  Because we will be boosting the accuracy of the finite 
-difference approximation by optimizing the time step, there is no reason to 
-explicitly construct the scheme to be high-order.  In fact, using a high-order 
-finite difference scheme only complicates the analysis required to calculate 
-the optimal time step.
-
-To compute the optimal time step, we first analyze the discretization errors 
-(both spatial and temporal) for the finite difference approximation, 
-explicitly keeping track of the leading-order terms.  In contrast to 
-conventional error analysis, we do not sweep all of the errors under ``big-O'' 
-notation.  Rather, we take advantage of the fact that derivation of the 
-leading-order terms in the error is straightforward for finite difference
-schemes through the use of Taylor series expansions.  Because each term in 
-the error is proportional to a partial derivative of the solution, we can 
-attempt to use the PDE to relate the leading-order terms to each other.  If we 
-are fortunate (which we are for many finite difference schemes), this careful 
-analysis yields a direct relationship between the leading-order spatial and 
-temporal errors that allows them to be combined as a single term of the form:
-\beq
-  (L u) P(\dx, \dt) ,
-  \label{eq:leading_order_error_model_PDE_general}
-\eeq
-where $L$ is a differential operator and $P$ is a polynomial in the numerical 
-parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
-$P = 0$, we can completely eliminate the leading-order term in the 
-discretization error.  Because time dependent PDEs are often solved by 
-stepping in time, it is natural to choose the time step to be a function of 
-the grid spacing.  The time step selected in this way is the
-\emph{optimal time step} for the finite difference scheme.  Using the optimal
-time step for time stepping yields a numerical solution with a boosted order 
-of accuracy for a given finite difference scheme.
-
-An important class of finite difference schemes for time dependent PDEs are 
-those based on the method of lines.   When the method of lines approach is 
-used for (\ref{eq:model_PDE}), $P$ takes a particularly simple 
-form:
-\beq
-  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
-  \label{eq:leading_order_error_model_PDE_MOL}
-\eeq
-where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
-discretization errors, and $\alpha$ and $\beta$ are positive constants that 
-depend on the details of the PDE and finite difference scheme.  Setting 
-$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
-root of $P$ is discarded because it is not numerically meaningful.}
-\beq
-  \dto = \gamma \dx^{r/s},
-  \label{eq:optimal_time_step}
-\eeq
-where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
-has a similar functional dependence on the grid spacing as a time step that 
-satisfies a stability constraint.
-
-\subsubsection*{\label{sec:error_analysis} 
-            Error Analysis}
-The form of the local error in 
-(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
-that the method of lines first approximates the PDE as a system of ODEs in 
-time by only discretizing the spatial derivatives: 
-\beq
-{\mathbf u}_t = F({\mathbf u}),
-\label{eq:method_of_lines}
-\eeq
-where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
-$F({\mathbf u})$ is the finite difference approximation of the spatial 
-derivative operator acting on ${\mathbf u}$.  The second term in the local 
-error arises directly from temporal discretization errors associated with the 
-numerical scheme used to integrate the system of ODEs in time.  The first 
-term arises because $F$ is only an approximation of the spatial derivative 
-operators in the PDE.  As a result, the system of 
-ODEs~(\ref{eq:method_of_lines}) has an error due to the spatial discretization 
-at \emph{every} point in time.  For a spatial discretization error of order 
-$r$, the error in (\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the 
-error accumulated during each time step due to spatial discretization error is 
-$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
-is solved numerically.  Because the method of lines is so frequently used to 
-construct finite difference approximations for time dependent PDEs, we will 
-focus solely on finite difference schemes of this type for the remainder of 
-the paper to keep the discussion simple.  It is straightforward to extend the 
-results we obtain to more general finite difference schemes.
-
-To analyze the accuracy of the finite difference scheme 
-for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
-higher-order terms in the discretization error following the conventional 
-approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
-discretization error leaves a local error of the form 
-$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
-of the spatial and temporal discretization errors, respectively, after the 
-leading-order terms in the error have been removed.  As for the leading-order 
-error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
-error is a direct consequence of the method of lines approach to discretizing 
-the PDE.
-
-From the local error, we can compute the global error of the numerical 
-solution over extended intervals of time by using the heuristic argument
-that the global error is equal to the local error divided by $\dt$.  Using 
-this heuristic argument, which is rigorously justified in many texts on 
-numerical methods such as~\cite{gko_book}, we find that the global error of 
-the finite difference scheme using an optimal time step is 
-\beq
-O(\dx^p) + O(\dt^q).
-\label{eq:global_error_ots}
-\eeq
-Observe that the global error for the finite difference scheme when an 
-arbitrary time step is used has exactly the same form 
-as~(\ref{eq:global_error_ots}) with $p$ and $q$ replaced by $r$ and $s$, 
-respectively.  Thus, because $p > r$ and $q > s$, we see that using an optimal 
-time step boosts the overall accuracy of the finite difference scheme.
-
-It is important to emphasize that $\dx$ and $\dt$ are \emph{not} independent 
-in~(\ref{eq:global_error_ots}) because the time step must be set to $\dto$ 
-in order to derive the error estimate.  In other words, there is really only 
-one parameter, $\dx$, that controls the numerical accuracy of the finite 
-difference scheme.  This important point will be discuss in further detail in 
-the following section.
-
-\subsubsection*{Formal vs. Practical Accuracy}
-For time dependent PDEs, evaluating the order of accuracy for a numerical
-method is subtle because there are formally two separate orders of accuracy 
-to consider\footnote{The situation is more straightforward for 
-time-independent PDEs because there is usually only one order of accuracy to 
-be concerned with.}:  temporal and spatial.  While it is theoretically
-interesting and important to understand how the error depends on both the 
-grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
-of a numerical scheme is always controlled by only one of the two numerical
-parameters.
-
-Even though spatial and temporal orders of accuracy for a numerical method
-are formally separate, one will always dominate for a given choice of 
-$\dx$ and $\dt$.  For example, when solving the 1D diffusion equation 
-using the backward Euler method for time integration with a second-order 
-central difference stencil for the Laplacian, formal analysis shows that 
-the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
-stability constraints on the numerical parameters, the time
-step and grid spacing are free to vary independently.  In this situation, the 
-practical error for the method depends on the relative sizes of the time step 
-and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
-$O(\dt)$ which means that the error in the numerical solution is 
-primarily controlled by the time step.  Similarly, when 
-$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
-the error is controlled by the grid spacing.  Finally, when 
-$\dt  = O(\dx^2)$, the practical error is 
-$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
-primarily controlled by one of the two numerical parameters, and varying
-the subdominant parameter while holding the dominant parameter fixed does 
-not significantly affect the error.
-
-When there are constraints on the numerical parameters, there is less freedom 
-in choosing the controlling parameter.  For instance, if we solve the 
-1D diffusion equation using a forward Euler time integration scheme with a 
-second-order central difference stencil for the Laplacian, stability
-considerations require that we choose $\dt = O(\dx^2)$.  
-Combining this stability constraint with the formal error for the scheme
-shows that the practical error is 
-$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
-method is completely controlled by the grid spacing; the temporal error can 
-never dominate the spatial error.
-
-Numerical methods for time dependent PDEs are commonly designed so that 
-the global error is primarily controlled by the grid spacing.  The numerical
-errors introduced by the time integration scheme are typically \emph{forced}
-to be subdominant to the spatial discretization errors manually or as a result
-of stability constraints.  In other words, the operational procedure used 
-when developing numerical methods is usually to first choose a spatial 
-discretization scheme and grid spacing so that a desired level of accuracy 
-is achieved and then select the time integration scheme and time step in 
-such a manner that the temporal errors do not overcome the spatial errors.  
-This methodology explains the tendency to evaluate numerical schemes for
-time dependent PDEs predominantly on their order of accuracy with respect 
-to the grid spacing.
-
-Optimal time step selection fits into this general framework for controlling
-the error of numerical methods for time dependent PDEs.  Because the optimal 
-time step is a function of the grid spacing, the technique of optimal time 
-step selection places an even more restrictive constraint on $\dt$ than 
-stability considerations.  Thus, like numerical methods with stability 
-constraints, the accuracy of a finite difference scheme using an optimal time 
-step is completely controlled by the grid spacing.  For the method of lines, 
-we know that $\dto = O(\dx^{r/s})$, so the global error derived in the 
-previous section reduces to 
-\beq
-O \left( \dx^{\min(p,rq/s)} \right).
-\label{eq:global_error_ots_simplified}
-\eeq
-
-% TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
-% sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
-\begin{table}[tb]
-\caption{\label{tab:comp_perf_vs_err} 
-   Computational cost of various finite difference schemes for the 1D 
-   diffusion equation as a function of the global numerical error $e$.
-   For all of the schemes, the standard second-order central difference 
-   stencil is used to discretize the Laplacian, and the numerical 
-   solution is assumed to be computed over a fixed time interval.  
-}
-\begin{minipage}{\textwidth}
-\begin{center} \footnotesize
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|l|c|c|c|c|}
-  \hline
-  {\bf Numerical Scheme} & $\dx$ 
-  & $\dt$\footnote{The 
-   time step for the forward/backward Euler and Crank-Nicholson schemes 
-   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
-   time step for the backward Euler scheme is chosen so that the temporal 
-   error is subdominant to the spatial error.}
-  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
-    to be $O(1/\dx)$.} 
-  & {\bf Compute Time}\footnote{The computation time is approximated as 
-    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
-    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
-    arise for the implicit methods.}  \\
-  \hline 
-  Forward Euler    & $O\left( e^{1/2} \right)$ 
-                   & $O\left( e \right)$ 
-                   & $O\left( e^{-1/2} \right)$ 
-                   & $O\left( e^{-3/2} \right)$ \\
-  Backward Euler   & $O\left( e^{1/2} \right)$ 
-                   & $O\left( e \right)$ 
-                   & $O\left( e^{-1/2} \right)$ 
-                   & $O\left( e^{-3/2} \right)$ \\
-  Crank-Nicholson  & $O\left( e^{1/2} \right)$ 
-                   & $O\left( e^{1/2} \right)$ 
-                   & $O\left( e^{-1/2} \right)$ 
-                   & $O\left( e^{-1} \right)$ \\
-  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
-                   & $O\left( e^{1/2} \right)$ 
-                   & $O\left( e^{-1/4} \right)$ 
-                   & $O\left( e^{-3/4} \right)$ \\ 
-  \hline
-\end{tabular}
-\end{center}
-\end{minipage}
-\end{table}
-
-\subsubsection*{Stability Considerations}
-Thus far, we have not considered the issue of numerical stability of finite
-difference schemes when the time step is set to $\dto$.  For explict time 
-integration schemes, OTS selection loses its value when the optimal time step 
-is greater than the smallest stable time step.  In this situation, we can 
-usually switch to an implicit (or semi-implicit) time integration scheme with 
-the same order of accuracy for which the optimal time step is stable.  Note 
-that a new optimal time step may need to be calculated for the modified finite
-difference sceme.
-
-\subsubsection*{\label{sec:computational_performance} 
-                Computational Performance}
-For many problems, the optimal time step is on the order of the time step 
-allowed by the stability constraint for an explicit method.  
-As a result, using the optimal time step to advance the finite
-difference scheme in time may appear to be overly restrictive and lead to poor 
-computational performance.  However, these apparent drawbacks are more than 
-compensated by the boost in the order of accuracy.  For example, when solving 
-the 1D diffusion equation using a forward Euler time integration scheme with a 
-second-order central difference stencil for the Laplacian, we obtain a 
-fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
-Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
-scheme that is computationally cheaper than traditional methods that use 
-the same low-order finite difference stencils.  The performance gain for 
-problems in higher spatial dimensions is even more impressive.  As we can see 
-in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
-simple forward Euler scheme with the optimal time step and the Crank-Nicholson
-method grows with the spatial dimension of the problem.
-
-\begin{table}[tb]
-\caption{\label{tab:comp_perf_vs_dim}
-   Computational cost of the forward Euler scheme with optimal time step and 
-   the Crank-Nicholson method for solving the diffusion equation as a function 
-   of the global numerical error $e$.
-   For both schemes, second-order, isotropic stencils are used to discretize 
-   the Laplacian.  For a problem in $d$ space dimensions, the memory 
-   requirement and computation time for each scheme are estimated to be 
-   $O \left(1/\dx^d \right)$ and 
-   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, respectively, which 
-   assumes that we are using an efficient 
-   $O(N) = O \left( 1/\dx^d \right)$ algorithm to invert the matrices that 
-   arise for the Crank-Nicholson method.
-}
-\begin{minipage}{\textwidth}
-\begin{center} \footnotesize
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|c|c|c|c|c|}
-  \hline
-  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
-    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
-  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
-  \cline{2-3} \cline{4-5} 
-    {\bf Dimensions} & {\bf Memory} & {\bf Compute Time} 
-  & {\bf Memory} & {\bf Compute Time} \\
-  \hline 
-  $1$ & $O\left( e^{-1/4} \right)$ 
-      & $O\left( e^{-3/4} \right)$ 
-      & $O\left( e^{-1/2} \right)$ 
-      & $O\left( e^{-1} \right)$ \\ 
-  $2$ & $O\left( e^{-1/2} \right)$ 
-      & $O\left( e^{-1} \right)$ 
-      & $O\left( e^{-1} \right)$ 
-      & $O\left( e^{-3/2} \right)$ \\ 
-  $3$ & $O\left( e^{-3/4} \right)$ 
-      & $O\left( e^{-5/4} \right)$ 
-      & $O\left( e^{-3/2} \right)$ 
-      & $O\left( e^{-2} \right)$ \\
-  $d$ & $O\left( e^{-d/4} \right)$ 
-      & $O\left( e^{-(d+2)/4} \right)$ 
-      & $O\left( e^{-d/2} \right)$ 
-      & $O\left( e^{-(d+1)/2} \right)$ \\ 
-  \hline 
-\end{tabular}
-\end{center}
-\end{minipage}
-\end{table}
-
-\subsection{\label{sec:ots_general_1d_pdes} 
-            OTS Selection for General PDEs in One Space Dimension} 
-In the previous section, we analyzed the technique of optimal time step 
-selection in the context of a simple model PDE in one space dimension.  
-For more general PDEs, the approach of carefully choosing the time step to 
-eliminate leading order terms in the error continues to be useful, but it 
-needs to be augmented to achieve the same improvement in the order of accuracy.
-In addition to using an optimal time step, correction terms may need to be 
-added to the finite difference scheme.  In this section, we discuss these 
-correction terms in the context of general constant coefficient, linear PDEs 
-and semilinear PDEs.
-
-\subsubsection*{\label{sec:ots_linear_pde} 
-            General Constant Coefficient, Linear PDEs} 
-Let us consider a general constant coefficient, linear PDE of the form
-\beq
-  \frac{\partial u}{\partial t} = 
-  \sum_{k=1}^n A_k \frac{\partial^k u}{\partial x^k} + f(x,t)
-  \label{eq:linear_PDE}.
-\eeq
-Even if we use the PDE to relate terms in the discretization error, it is
-unlikely that all of the dominant terms in the error can be combined into 
-a single term.   As a result, the leading-order discretization error is
-generally a sum of several terms:
-\bea
-  \dt\left[ (L u) (\alpha \dx^r - \beta \dt^s) 
-  + \sum_k (L_k u) \dx^r 
-  + \sum_k (L'_k u) \dt^s 
-  + (G f) \dt^s \right]
-  \label{eq:leading_order_error_linear_PDE_MOL}
-\eea
-where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
-(\ref{eq:leading_order_error_model_PDE_general}) and
-(\ref{eq:leading_order_error_model_PDE_MOL}), 
-$L_k$ and $L'_k$ are the linear spatial differential operators associated 
-with the leading-order terms in the discretization error that are not involved 
-in the choice of the optimal time step, and $G$ is a spatio-temporal 
-differential operator that acts on source term $f$.  
-Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
-operators because any that arise can be eliminated using the PDE.  Also, 
-there is no explicit spatial discretization error of order $r$ that originates 
-from the source term because it is not differentiated with respect to space 
-when the semi-discrete equations~(\ref{eq:method_of_lines}) are derived from 
-the PDE.
-
-From~(\ref{eq:leading_order_error_linear_PDE_MOL}), it is clear that setting 
-$\dt = \dto$ is not enough to completely eliminate the 
-leading-order error.  To eliminate the vestigial errors, we need to add a few 
-correction terms to the numerical scheme in exactly the same spirit used in 
-the design of the modified nine-point formula for the Poisson 
-equation~\cite{iserles_book}.  The required correction terms are precisely 
-those that cancel out the second, third, and fourth terms in 
-(\ref{eq:leading_order_error_linear_PDE_MOL}).  Fortunately, it is 
-straightforward to compute and incorporate these terms into the numerical 
-scheme.  At each time step, the second and third terms in 
-(\ref{eq:leading_order_error_linear_PDE_MOL}) can be approximated using
-finite differences applied to the solution $u$ at the current time.
-The last term can be computed using either a finite difference approximation 
-or calculating it analytically.  With the addition of these correction terms, 
-the leading-order discretizaton error is completely eliminated, and the
-global error can be determined using the analysis in 
-Section~\ref{sec:error_analysis}.
-
-It is important to point out that the finite difference stencils used to 
-approximate the vestigial errors must be chosen carefully.  Special attention 
-is required to ensure that the finite difference stencils are of sufficiently 
-high-order.  Otherwise, we will lose the accuracy gained from using an optimal 
-time step and adding the correction terms to the finite difference scheme.  
-Since the specific form of the correction terms are PDE-dependent, we defer 
-further discussion to Sections~\ref{sec:applications_1d} 
-and~\ref{sec:applications_multidim} where we apply optimal time step selection 
-to several specific PDEs.
-
-\subsubsection*{Semilinear PDEs}
-Semilinear PDEs are an important class of PDEs that have the form
-\beq
-  \frac{\partial u}{\partial t} = A(x,t) \frac{\partial^n u}{\partial x^n}
-  + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}},
-      \frac{\partial^{n-2} u}{\partial x^{n-2}}, \ldots,
-      \frac{\partial u}{\partial x}, u \right)
-  + f(x,t)
-  \label{eq:semilinear_PDE}
-\eeq
-where $F$ is an arbitrary function of the lower-order spatial derivatives
-of $u$ and $f(x,t)$ is a source term.  While the coefficient on the 
-leading-order spatial derivative is generally allowed to be function of 
-space and time, optimal time step selection can only be applied to PDEs where
-$A(x,t)$ is a constant.  Because the PDE is linear in the leading-order 
-spatial derivative, it is natural to construct a finite difference scheme that 
-treats $F$ in an explicit manner and lets stability considerations determine 
-whether the $A \frac{\partial^n u}{\partial x^n}$ term is handled explicitly 
-or implicitly.  Assuming that sufficiently high-order stencils are used to 
-compute $F$, the leading-order discretization error for such a finite 
-difference scheme is given by
-\bea
-  \dt \left[ (L u) (\alpha \dx^r - \beta \dt^s) 
-  + \left( H u \right) \dx^r 
-  + \left( H' u \right) \dt^s
-  + (G f) \dt^s \right]
-  \label{eq:leading_order_error_semilinear_PDE_MOL}
-\eea
-where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
-(\ref{eq:leading_order_error_model_PDE_general}) and
-(\ref{eq:leading_order_error_model_PDE_MOL}), $H$ and $H'$ are spatial 
-differential operators (potentially nonlinear) associated with the 
-leading-order terms in the discretization error that arise from the nonlinear 
-term in~(\ref{eq:semilinear_PDE}), and $G$ is a spatio-temporal differential 
-operator that acts on $f$.  With the discretization error written in the 
-form (\ref{eq:leading_order_error_semilinear_PDE_MOL}), the optimal time 
-step and correction terms can be computed by proceeding in the same manner 
-as for general constant coefficient, linear PDEs.
-
-\subsection{Boundary Conditions}
-In our analysis and discussion thus far, we have completely ignored the issue 
-of boundary conditions.  The choice of discretization for boundary conditions
-can have a significant impact on the accuracy of the numerical solution.  In 
-general, high-order discretizations of boundary conditions are required to 
-ensure that errors introduced at the domain boundary do not pollute the 
-numerical solution in the domain interior.  For Dirichlet boundaries, it
-is sufficient to impose boundary conditions by using boundary (or ghost cell) 
-values that have the same order of accuracy as the solution.  
-More effort is required to ensure that Neumann boundary conditions do not
-reduce the order of accuracy of the solution.  In general, Neumann boundary
-conditions require that the accuracy of ghost cells slightly exceed the 
-desired order of accuracy for the solution.  A detailed discussion of 
-methods for imposing boundary conditions is beyond the scope of this paper.  
-We refer interested readers to the thorough analysis presented
-in~\cite{gko_book}.  To avoid complicating the discussion in 
-Sections~\ref{sec:applications_1d} and~\ref{sec:applications_multidim}, 
-Neumann boundary conditions are imposed by taking advantage of knowledge of 
-the analytical solution.
-
-\subsection{\label{sec:ots_higher_spatial_dims}
-            OTS Selection in Multiple Space Dimensions}
-Optimal time selection is applicable to PDEs in any number of space
-dimensions.  It is straightforward to generalize the approach outlined in 
-the previous sections to problems in multiple space dimensions.  While 
-problems in higher dimensions generally require a more careful choice of 
-the finite difference stencil used to approximate spatial derivatives,
-there are no fundamental barriers that prevent the use of optimal time step 
-selection.
-
-The geometry of the domain is an important feature of problems in more than 
-one space dimension because it can significantly impact the way boundary 
-conditions are imposed.  For problems on regular domains, boundary conditions 
-can usually be handled by simply using 1D boundary condition discretizations 
-in each coordinate direction.  Problems on irregular domains pose a greater 
-challenge because the location of the domain boundary may not be explicitly 
-specified and grid points near the boundary require special consideration.  
-Fortunately, because optimal time step selection is based on simple, localized 
-finite difference stencils, irregular domains do not create any serious 
-difficulties.  Assuming that a `ghost cell' or `ghost value' 
-approach~\cite{gibou_2005, fedkiw_1999, osher_fedkiw_book} is used to impose 
-boundary conditions, the construction and analysis of finite 
-difference schemes and calculation of optimal time steps is identical for 
-problems on both regular and irregular domains.  For problems on 
-irregular domains, the key issue is how values at ghost cell points should 
-be calculated.  Once this issue has been addressed, devising high-order 
-accurate methods for PDEs on irregular domains is simply a matter of applying 
-optimal time step selection to the finite difference scheme used for problems 
-on regular domains.
-
-Another important issue that arises in multiple space dimensions is the choice
-of grid spacing.  In general, the grid spacing need not be equal in all 
-spatial directions.  For some problems, the ratio of the grid spacing in
-different coordinate directions may need to be carefully chosen for an 
-optimal time step to even exist.
-
-\subsection{Limitations of Optimal Time Step Selection}
-As with any numerical method, optimal time step selection has its limitations.
-The most serious limitation is the requirement that the coefficient on the 
-leading-order spatial derivative in the PDE be constant in time and space.  
-Without this critical assumption, the first term in the leading-order 
-discretization error would be of the form 
-$(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
-where $\alpha$ and $\beta$ are now functions of space and time.  This
-small change in the form of the error makes it impossible to derive a single 
-optimal time step.  In a sense, each grid point would possess a different 
-optimal time step that could also depend on the time.  For a similar reason, 
-optimal time step selection also fails for quasilinear and fully nonlinear 
-PDEs.  The issue for these types of PDEs is that the coefficients in 
-$P(\dx, \dt)$ become functions of the solution $u$.
-
-Another limitation of optimal time step selection is that it does not 
-generally apply to finite difference schemes for systems of PDEs.  The 
-key issue is that each PDE might have its own optimal time step.  While it 
-would be possible to boost the accuracy of finite difference schemes for 
-systems of PDEs where the optimal time step is the same across the individual 
-PDEs, it is not clear that OTS selection yields any benefit in general.
-
-\section{\label{sec:applications_1d} 
-         Application to PDEs in One Space Dimension}
-Because there are no geometry or anisotropy considerations, optimal time step 
-selection is very useful for boosting the accuracy of many finite difference 
-schemes in one spatial dimension.  In this section, we demonstrate its 
-utility in designing finite difference schemes to solve several classical 
-linear and semilinear PDEs where analytical solutions are available for 
-comparison with numerical solutions.
-
-\subsection{Linear Advection Equation}
-For our first application of optimal time step selection, we consider 
-finite difference schemes for the linear advection equation
-~\cite{leveque_book_1992, leveque_book_2002, gko_book}:
-\beq
-  \frac{\partial u}{\partial t} + A \frac{\partial u}{\partial x} = 0,
-  \label{eq:advection_eqn_1d}
-\eeq
-where $A$ is the flow speed, which we take to be positive for convenience.  
-For an initial condition of the form $u(x,0) = f(x)$, it is easy to show 
-that the analytical solution is $f(x-At)$.  While it is unlikely that one 
-would actually use a numerical method to solve~(\ref{eq:advection_eqn_1d}), 
-it provides an interesting example where optimal time step selection 
-dramatically improves the accuracy of the numerical solution.
-
-There are many finite difference schemes that can be used to 
-solve~(\ref{eq:advection_eqn_1d}).  Several examples are given 
-in~\cite{leveque_book_1992}.  Perhaps the simplest, stable method is a 
-forward Euler for time integration scheme that uses a first-order upwind 
-discretization for the spatial derivative:
-\beq
-  u^{n+1}_j = u^{n}_j 
-  - A \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right),
-  \label{eq:advection_eqn_1d_FD_scheme}
-\eeq
-where superscripts and subscripts denote the time and space indices, 
-respectively.  It is straightforward to show that the method is formally 
-first-order in both space and time.
-
-To compute the optimal time step, we begin by analyzing the discretization
-error for~(\ref{eq:advection_eqn_1d_FD_scheme}).  In general, we only need to 
-keep track of the leading-order terms in the discretization error.  However,
-for the advection equation, it is instructive to retain all of them.  
-Using Taylor series to expand the true solution of the PDE at time step
-$(n+1)$, we find that
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
-       \frac{\partial^k \tu^n}{\pt^k} 
-  \label{eq:advection_eqn_1d_time_err}
-\eea
-where $\tu$ denotes the true solution of the PDE and 
-$\tu^n_j \equiv \tilde{u}(j\dx, n\dt)$.  Similarly, Taylor series expansion of 
-the upwind finite difference approximation to the spatial derivative yields
-\bea
-  \frac{\tu^{n}_j - \tu^{n}_{j-1}}{\dx} = 
-  \frac{\partial \tu}{\px} 
-  - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k}.
-  \label{eq:advection_eqn_1d_space_err}
-\eea
-In both of the above equations, note that we have let context implicitly 
-determine the time ($t = t_n = n\dt$) and location ($x = x_j = j\dx$) where 
-partial derivatives of $\tu$ should be evaluated.  For convenience, we shall 
-continue to follow this convention throughout the remainder of the paper.
-
-Next, we use the PDE~(\ref{eq:advection_eqn_1d_FD_scheme}) to replace all of 
-the time derivatives in~(\ref{eq:advection_eqn_1d_time_err}) with spatial
-derivatives to obtain
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + \sum_{k=1}^\infty \frac{\left( -A \dt \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k}.
-  \label{eq:advection_eqn_1d_time_err_modified}
-\eea
-Finally, by combining~(\ref{eq:advection_eqn_1d_time_err_modified}) 
-and~(\ref{eq:advection_eqn_1d_space_err}) with the finite difference 
-scheme~(\ref{eq:advection_eqn_1d_FD_scheme}), we obtain the evolution
-equation for the error $e \equiv u - \tu$:
-\bea
-  e^{n+1}_j = e^{n}_j 
-    &-& A \dt \left( \frac{e^{n}_j - e^{n}_{j-1}}{\dx} \right) \nonumber \\
-    &+& A \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
-        \frac{\partial^k \tu^n}{\px^k} 
-        \left( \dx^{k-1} - \left( A \dt \right)^{k-1} \right).
-  \label{eq:advection_eqn_1d_err_eqn}
-\eea
-As expected, the error $e$ satisfies the same finite difference approximation 
-as the solution with a source term which is the discretization error for the 
-scheme.
-
-Extracting the leading-order term of the discretization error
-\beq
-  \left( \frac{A}{2} \frac{\partial^2 \tu^n}{\px^2} \right)
-  \left( \dx - A \dt \right) \dt,
-\eeq
-we find that the optimal time step is $\Delta t_{opt} = \Delta x / A$.
-For the linear advection equation, setting the time step equal to the optimal 
-time step has an amazing consequence -- not only does the leading-order term
-in the discretization error vanish, the \emph{entire} local truncation 
-error at each time step is completely eliminated!  In other words, the error 
-in the numerical solution is due solely to errors in the initial conditions.
-Assuming that the initial conditions are set exactly, there is absolutely no 
-error in the numerical solution for all times!  
-This surprising result, which is unique to the advection equation, is 
-well-known in the numerical methods community and is referred to as the unit 
-CFL condition~\cite{leveque_book_2002}.  It is usually proved by examining 
-how characteristic lines of the advection equation pass through grid points.  
-It is interesting to see that it also naturally arises through an analysis of 
-the discretization errors and optimal time step selection.
-
-\subsection{Diffusion Equation}
-As an example where optimal time step selection yields a more typical boost 
-in the order of accuracy, we turn our attention to the diffusion equation:
-\beq
-  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2},
-  \label{eq:diffusion_eqn_1d}
-\eeq
-where $D$ is the diffusion constant.  As with the advection equation,
-there are many finite difference schemes that can be used to solve the 
-diffusion equation.  Keeping with our philosophy of using simple, low-order 
-methods to construct finite difference schemes, we apply optimal time step 
-selection to a scheme that uses forward Euler for time integration and the 
-standard second-order central difference approximation for the Laplacian:
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + D \dt 
-    \left( \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right),
-  \label{eq:diffusion_eqn_1d_FD_scheme}
-\eeq
-Conventional error analysis shows that scheme is formally first-order in time 
-and second-order in space.  Due to the stability constraint 
-$\dt \le \dx^2/2D$, the scheme is $O(\dx^2)$ accurate overall.
-
-As before, we begin by analyzing the leading-order terms of the 
-discretization error for the finite difference scheme.  Employing Taylor
-series expansions, we find that the true solution satisfies
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + \dt \frac{\partial \tu^n}{\pt} 
-  + \frac{\dt^2}{2} \frac{\partial^2 \tu^n}{\pt^2} + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_time_err}
-\eea
-and the central difference for the Laplacian satisfies
-\bea
-  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
-  \frac{\partial^2 \tu^n}{\px^2} 
-  + \frac{\dx^2}{12} \frac{\partial^4 \tu^n}{\px^4} 
-  + O(\dx^4)
-  \label{eq:diffusion_eqn_1d_space_err}
-\eea
-Making use of the PDE~(\ref{eq:diffusion_eqn_1d}), 
-(\ref{eq:diffusion_eqn_1d_time_err}) becomes
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + D \dt \frac{\partial^2 \tu^n}{\px^2} 
-  + \frac{(D \dt)^2}{2} \frac{\partial^4 \tu^n}{\px^4} 
-  + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_time_err_modified}
-\eea
-Combining these results with the finite difference 
-scheme~(\ref{eq:diffusion_eqn_1d_FD_scheme}), we arrive at the evolution 
-equation for the error $e \equiv u - \tu$:
-\bea
-  e^{n+1}_j = e^{n}_j 
-  &+& D \dt 
-    \left( \frac{e^{n}_{j+1} -2 e^{n}_j + e^{n}_{j-1}}{\dx^2} \right)
-  \nonumber \\
-  &+& \frac{\partial^4 \tu^n}{\px^4} 
-      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-      + O(\dt \dx^4) + O(\dt^3)
-  \label{eq:diffusion_eqn_1d_err_eqn}
-\eea
-Again, we see that the error satisfies the same evolution equation
-as the numerical solution and that the discretization error acts as a 
-source term.
-
-From~(\ref{eq:diffusion_eqn_1d_err_eqn}), we see that the optimal time step 
-is given by $\dto = \dx^2/6D$.  Using~(\ref{eq:global_error_ots}) or
-(\ref{eq:global_error_ots_simplified}), the order of the scheme with this 
-choice of time step is $O(\dx^4)$; we have boosted the accuracy of original 
-scheme from second to fourth order.  Note that $\dto$ is stable, so there
-is no need to switch to an implicit time integration scheme.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_no_src_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for three 
-finite difference schemes that solve the 1D diffusion equation: forward Euler 
-with optimal time step (circles), forward Euler with suboptimal time step
-$\dt = 2\dx^2/3D$ (squares), and Crank-Nicholson with time step 
-$\dt = \dx/16 D$ (diamonds).  All three schemes use the standard second-order, 
-central finite difference approximation for the Laplacian.  Using each method, 
-we solve the diffusion equation with $D = 0.5$ to time $t = 0.1$ on the 
-domain $0 < x < 1$ subject to the boundary conditions $u(0,t) = 1$ and 
-$u_x(1,t) = 1/2$ and the initial condition
-$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
-- 4 \sin(11 \pi x/2)$.  
-Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figure.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:diffusion_eqn_1d_no_src_error}
-\end{center}
-\end{figure}
-
-To demonstrate the improvement in accuracy of the forward Euler scheme when 
-the optimal time step is used, we compare the numerical errors of three finite 
-difference schemes for the diffusion equation: forward Euler with optimal
-time step, forward Euler with suboptimal time step, and Crank-Nicholson.  
-As we can see in Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward 
-Euler with optimal time step is fourth-order accurate while both forward Euler 
-with suboptimal time step and Crank-Nicholson are only second-order accurate.  
-Using the forward Euler scheme with an optimal time step has a clear 
-advantage over both of the other schemes.  Even with as few as 25 grid 
-points, the error in solution computed by the forward Euler scheme with an 
-optimal time step is orders of magnitude smaller.
-
-As mentioned in Section~\ref{sec:computational_performance}, one of the 
-advantages of using optimal time step selection is the superior scaling of 
-the computational cost with respect to the numerical error.  It is worth 
-mentioning that another way to achieve the same computational performance
-would be to use a Crank-Nicholson time integration scheme with a fourth-order 
-discretization of the Laplacian~\cite{gibou_2005}.  By using an 
-$O\left(\dx^2 \right)$ time step and a linear solver that scales linearly 
-with the size of the matrix, we achieve an overall fourth-order accurate 
-method that has the same computational complexity as the forward Euler scheme
-described above with an optimal time step.  However, the forward Euler scheme 
-with optimal time step will still outperform the Crank-Nicholson scheme 
-because the latter requires the use of a wider stencil and the solution of 
-at least one linear system of equation for each time step.
-
-\subsubsection*{Diffusion Equation with Source Term} 
-As discussed in Section~(\ref{sec:ots_linear_pde}), the procedure for optimal 
-time step selection needs to be slightly modified when source terms are 
-present.  If we add a source term, $f(x,t)$, to the right-hand side 
-of~(\ref{eq:diffusion_eqn_1d}), the finite difference 
-scheme~(\ref{eq:diffusion_eqn_1d_FD_scheme}) picks up an additional term
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + D\dt 
-    \left( \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} + f^n_j \right).
-  \label{eq:diffusion_eqn_1d_src_FD_scheme}
-\eeq
-and~(\ref{eq:diffusion_eqn_1d_time_err_modified}) becomes
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  &+& D\dt \left( \frac{\partial^2 \tu^n}{\px^2} + f^n_j \right)
-  \nonumber \\
-  &+& \frac{(D\dt)^2}{2} \frac{\partial^4 \tu^n}{\px^4} 
-  + \frac{\dt^2}{2} \left( D\frac{\partial^2 f}{\px^2}
-                         + \frac{\partial f}{\pt} \right)
-  + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_src_time_err_modified}
-\eea
-Therefore, the leading-order discretization error is given by
-\beq
-  \frac{\partial^4 \tu^n}{\px^4} 
-    \left[ \frac{\dx^2}{12} - \frac{D \dt}{2} \right] (D \dt)
-    - \frac{\dt^2}{2} \left( D \frac{\partial^2 f}{\px^2} 
-                           + \frac{\partial f}{\pt} \right).
-  \label{eq:diffusion_eqn_1d_src_err}
-\eeq
-To eliminate the low-order errors introduced by the source term, we 
-simply add the following correction term to the right-hand side of 
-the finite difference scheme~(\ref{eq:diffusion_eqn_1d_src_FD_scheme}):
-\beq
-\frac{\dt^2}{2} \left( D \left( f_{xx} \right)^n_j 
-  + \left( f_t \right)^n_j \right),
-\eeq
-where the derivatives of $f$ may be evaluated analytically or using 
-sufficiently high-order finite difference stencils.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for the 
-two finite difference schemes that solve the diffusion equation: 
-forward Euler with optimal time step and correction terms (circles) and 
-forward Euler with optimal time step but \emph{without} correction terms 
-(triangles).  Both schemes use the standard second-order, central finite 
-difference approximation for the Laplacian.
-Using each method, we solve the diffusion equation with $D = 0.5$ and
-source term 
-$f(x) = 5 \sin(3 \pi x/2) - 7 \sin(15 \pi x/2) + 10 \sin(21 \pi x/2)$
-to time $t = 0.5$ on the domain $0 < x < 1$ subject to the boundary 
-conditions $u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
-and the initial condition
-$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
-- 4 \sin(11 \pi x/2)$.  
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:diffusion_eqn_1d_src_error}
-\end{center}
-\end{figure}
-
-In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of 
-the OTS forward Euler scheme with and without the correction terms.  As we can 
-see, OTS forward Euler with correction terms is fourth-order accurate while 
-OTS forward Euler without correction terms is only second-order accurate.  
-These results clearly demonstrate the importance of the correction terms for
-achieving high-order accuracy.
-
-\subsection{Advection-Diffusion Equation}
-We now demonstrate the use of optimal time step selection in the context of 
-a general linear, constant coefficient PDE -- the advection-diffusion 
-equation:
-\beq
-  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2} 
-  + A \frac{\partial u}{\px},
-  \label{eq:ADE_1d}
-\eeq
-where $D$ is the diffusivity and $A$ is the flow speed.  For this problem,
-we use forward Euler for time integration and second-order central 
-difference approximations for the diffusion and advection terms:
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + \dt 
-    \left( D \left[ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
-         + A \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] \right).
-  \label{eq:ADE_1d_FD_scheme}
-\eeq
-This scheme is formally first-order in time and second-order in space.  
-Due to the stability constraint $\dt < \min \{2D/A^2, \dx^2/2D \}$ 
-\cite{chan_1984}, the scheme is $O(\dx^2)$ accurate overall (in the limit 
-$\dx \rightarrow 0$).
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/adv_diff_eqn_1d_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for 
-two finite difference schemes that solve the advection-diffusion 
-equation: forward Euler with optimal time step and correction term (circles) 
-and forward Euler with suboptimal time step $\dt = 2\dx^2/3D$ and no correction 
-terms (squares).  Both schemes use second-order, central finite difference 
-approximations for the advection and diffusion terms.
-Using both methods, we solve the advection-diffusion equation with $D = 2$
-and $A = 5$ to time $t = 1$ on the domain $-10 < x < 10$ subject to the 
-initial condition $u(x,0) = \exp(-x^2/4)$.  The boundary conditions are set 
-using the analytical solution 
-$u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
-where $T = Dt+1$.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:ADE_1d_error}
-\end{center}
-\end{figure}
-
-To analyze the discretization error, we can reuse 
-results~(\ref{eq:diffusion_eqn_1d_time_err}) 
-and~(\ref{eq:diffusion_eqn_1d_space_err})
-from our analysis of the diffusion equation.  In addition, we need the 
-leading-order discretization error for the central difference approximation 
-of $u_x$, which is given by
-\bea
-  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
-  \frac{\partial \tu^n}{\px} 
-  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
-  + O(\dx^4).
-  \label{eq:ADE_1d_ux_err}
-\eea
-Using the advection-diffusion equation~(\ref{eq:ADE_1d})
-to convert time derivatives into spatial derivatives 
-in~(\ref{eq:diffusion_eqn_1d_time_err}), we find that 
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} 
-               + A \frac{\partial \tu^n}{\px} \right)
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
-      \left( 
-        D^2 \frac{\partial^4 \tu^n}{\px^4} 
-      + 2 D A \frac{\partial^3 \tu^n}{\px^3} 
-      + A^2 \frac{\partial^2 \tu^n}{\px^2} 
-      \right) 
-  + O \left( \dt^3 \right).
-  \label{eq:ADE_1d_time_err_modified}
-\eea
-The local truncation error $\tau^n_j$ for the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) can be obtained by the usual procedure 
-of deriving the evolution equation for the error and extracting the source 
-term:
-\bea
-  \tau^n_j &=&
-      \left( \frac{\partial^4 \tu^n}{\px^4} 
-           + \frac{2A}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
-      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-  \nonumber \\
-  &-& \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
-      + O(\dt \dx^4) + O(\dt^3)
-  \label{eq:ADE_1d_err_eqn}
-\eea
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D advection-diffusion
-equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step $\dt = 2 \dx^2/3 D$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 1$ on the 
-domain $-10 < x < 10$ with the numerical solution (circles) computed using 
-50 grid points.  For details about the initial/boundary 
-conditions and physical parameters used to compute these solutions, see the
-caption for Figure~\ref{fig:ADE_1d_error}.
-}
-\label{fig:ADE_1d_solns}
-\end{center}
-\end{figure}
-
-As we discussed in Section~\ref{sec:ots_linear_pde}, boosting the order
-of accuracy for general linear, constant coefficient PDEs requires both 
-the use of an optimal time step and the addition of correction terms.  
-In this case, the optimal time step is $\dto = \dx^2/6D$ and the
-correction term is 
-\beq
-  \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
-  \label{eq:ADE_1d_corr_term}
-\eeq 
-Note that using the standard second-order, central difference stencil to 
-compute the Laplacian in the correction term only makes an 
-$O(\dt^2 \dx^2)$ contribution to $\tau_j^n$, which ensures that no 
-errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into the 
-truncation error.
-With the optimal time step and correction term, the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
-Figures~\ref{fig:ADE_1d_error} and \ref{fig:ADE_1d_solns} illustrate the 
-improved accuracy of the finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) 
-when the optimal time step and correction term are used.
-
-\subsubsection*{First-order Upwind Discretization of Advection Term}
-It is common to use upwind discretizations such 
-as~(\ref{eq:advection_eqn_1d_FD_scheme}) for advection terms that arise
-in PDEs.  If instead of a central difference approximation, we use a
-first-order upwind stencil\footnote{The appropriate upwind stencil depends
-on the sign of $A$.} for the advection term in~(\ref{eq:ADE_1d}), the 
-discretization error for the finite difference scheme is dominated by the 
-first-order error introduced by the upwind stencil for the advection term:
-\bea
-  \tau^n_j &=&
-      \frac{\partial^2 \tu^n}{\px^2} 
-      \left[ \frac{\dx}{2} - \frac{|A| \dt}{2}  \right] (|A| \dt)
-  \nonumber \\
-   & & 
-    - \frac{\dt^2}{2} 
-      \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
-           + 2 A D \frac{\partial^3 \tu^n}{\px^3}
-      \right)
-      + O(\dt \dx^2) + O(\dt^3)
-  \label{eq:ADE_1d_upwind_err_eqn}.
-\eea
-Thus, we can see that for a finite difference scheme based on a first-order
-upwind discretization of the advection term, the optimal time step is 
-$\dto = \dx/|A|$ (as for the pure advection equation) and the required
-correction term is
-\beq
-    \frac{\dt^2}{2} 
-      \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
-           + 2 A D \frac{\partial^3 \tu^n}{\px^3}
-      \right)
-  \label{eq:ADE_1d_upwind_corr_term}.
-\eeq
-This scheme is second-order overall as long as the finite difference 
-approximations to all of the derivatives in~(\ref{eq:ADE_1d_upwind_corr_term})
-are at least first-order accurate.  Unfortunately, the optimal time step 
-is unstable for the finite difference scheme that uses forward Euler time 
-integration and a first-order upwind stencil for the advection term.  
-In order to use optimal time step selection, we would need to replace the 
-forward Euler time integration scheme with a backward Euler scheme.
-
-A comparison of the finite difference scheme analyzed in this section to the 
-scheme described in the previous section makes it clear that the power 
-of optimal time step selection is highly dependent on the original choice of 
-finite difference scheme.  The first-order upwind scheme not only has a 
-lower overall accuracy when the optimal time step is used, it requires the 
-use of an implicit time integration method and construction of finite 
-difference stencils for higher-order spatial derivatives in order to compute
-the correction term~(\ref{eq:ADE_1d_upwind_corr_term}).  That the first-order
-scheme possesses all of these weaknesses underscores the importance of 
-carefully choosing the finite difference scheme even when optimal time step 
-selection is used.
-
-\subsection{Viscous Burgers Equation}
-% FIGURE PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/burgers_eqn_1d_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for
-two finite difference schemes that solve the viscous Burgers equation:
-forward Euler with optimal time step and correction term (circles)
-and forward Euler with suboptimal time step $\dt = \dx^2/4\nu$ and no 
-correction terms (squares).
-Both schemes use second-order, central finite difference approximations for
-the advection and diffusion terms.
-The viscous Burgers equation is solved to time $t=2$ on the domain 
-$0 < x < 10$ with the viscosity $\nu$ set to $0.1$.  The initial conditions are 
-taken to be 
-$u(x,0) = 1 + \gamma \sqrt{\frac{\nu}{\pi}} \exp\left( -(x-1)^2/4 \nu \right) 
-            / \left[ 1 + \frac{\gamma}{2} \ 
-                         \mathtt{erfc}\left((x-1)/\sqrt{4\nu} \right) 
-    \right]$,
-where $\ln(1+\gamma)$ is the effective Reynolds number (set equal to 10) and 
-the wave speed is set equal to $1$.  The boundary conditions are imposed 
-exactly by using
-the analytical solution
-$u(x,t) = 1 + \gamma \sqrt{\frac{\nu}{\pi T}} 
-              \exp\left( -(x-T)^2/4 \nu T \right) 
-            / \left[ 1 + \frac{\gamma}{2} \ 
-                         \mathtt{erfc}\left((x-T)/\sqrt{4\nu T} \right) 
-    \right]$,
-where $T = t+1$.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:burgers_1d_error}
-\end{center}
-\end{figure}
-
-For an example of optimal time step selection applied to a semilinear equation,
-we consider the viscous Burgers equation~\cite{whitham_book}:
-\beq
-  \frac{\partial u}{\pt} + u \frac{\partial u}{\px} = 
-     \nu \frac{\partial^2 u}{\px^2} 
-  \label{eq:burgers_1d}
-\eeq
-where $\nu$ is viscosity.  Notice that this problem has the same structure
-to the advection-diffusion equation, except that the advection velocity now 
-depends on the solution $u$.  To solve this equation numerically,
-we will use forward Euler for time integration and second-order central 
-difference approximations for the diffusion and advection terms:
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + \dt 
-    \left( \nu 
-      \left [ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
-         - u^n_j 
-      \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] 
-    \right).
-  \label{eq:burgers_1d_FD_scheme}
-\eeq
-This scheme is formally first-order in time and second-order in space.  
-Using the stability condition for the advection-diffusion equation as a 
-heuristic guide, we expect a stability constraint of the form 
-$\dt \le \dx^2/2 \nu$ in the limit $\dx \rightarrow 0$ (assuming that 
-the solution $u$ remains bounded).  Therefore, the scheme 
-(\ref{eq:burgers_1d_FD_scheme}) is $O(\dx^2)$ accurate overall.
-
-Because~(\ref{eq:burgers_1d_FD_scheme}) is very similar to the second-order
-finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) for the 
-advection-diffusion equation, we can reuse many of the results obtained in 
-the previous section to derive the truncation error 
-for~(\ref{eq:burgers_1d_FD_scheme}).  Following a procedure analogous to the
-one we used to analyze~(\ref{eq:ADE_1d_FD_scheme}), we find that the local 
-truncation error for~(\ref{eq:burgers_1d_FD_scheme}) has the form
-\bea
-  \tau^n_j &=&
-      \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
-           - 2 \tu \frac{\partial^3 \tu^n}{\px^3} \right)
-      \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
-      \left( 
-           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
-         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
-         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
-      \right)
-      + O(\dt \dx^4) + O(\dt^3).
-  \label{eq:burgers_1d_err_eqn}
-\eea
-Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
-for the finite difference scheme is 
-\beq
-  - \frac{\dt^2}{2} 
-      \left( 
-           4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
-         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
-         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
-      \right)
-  \label{eq:burgers_1d_corr_term}
-\eeq 
-As for the advection-diffusion, using the standard second-order central 
-difference stencils to compute the correction term is sufficient to ensure 
-that the overall error for the scheme is $O(\dx^4)$.  
-Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate
-the boost in accuracy obtained by using the optimal time step and 
-adding correction terms to the finite difference 
-scheme~(\ref{eq:burgers_1d_FD_scheme}).
-
-% FIGURE PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \ 
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D viscous Burgers 
-equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 2$ on the 
-domain $0 < x < 10$ with the numerical solution (circles) computed using 
-$100$ grid points.  
-For details about the initial/boundary conditions and physical parameters 
-used to compute these solutions, see the caption for 
-Figure~\ref{fig:burgers_1d_error}. 
-}
-\label{fig:burgers_1d_solns}
-\end{center}
-\end{figure}
-
-\subsection{Fourth-Order Parabolic Equation}
-As a final example in one space dimension, we consider a PDE with high-order
-spatial derivatives -- a fourth-order parabolic equation: 
-\beq
-  \frac{\partial u}{\pt} = -\kappa \frac{\partial^4 u}{\px^4} + f(x,t).
-  \label{eq:4th_order_parabolic_eqn_1d}
-\eeq
-From our previous examples, we know that optimal time step selection requires 
-the coefficient on leading-order spatial discretization error to be directly 
-related to the second derivative of the solution $u$ with respect to time.  
-From~(\ref{eq:4th_order_parabolic_eqn_1d}), we see that 
-\beq
-  \frac{\partial^2 u}{\pt^2} = 
-    \kappa^2 \frac{\partial^8 u}{\px^8} 
-  - \kappa \frac{\partial^4 f}{\px^4} 
-  + \frac{\partial f}{\pt}
-  \label{eq:4th_order_parabolic_eqn_1d_second_time_derivative},
-\eeq
-which indicates that the leading-order term in the spatial discretization 
-error should involve the eighth-derivative of $u$.  In other words, the
-finite difference approximation to the bilaplacian needs to be at least 
-fourth-order accurate.  With this insight, let us consider a finite 
-difference scheme for~(\ref{eq:4th_order_parabolic_eqn_1d}) that 
-uses forward Euler for time integration and the a fourth-order central 
-difference approximation for the bilaplacian:
-\bea
-  u^{n+1}_j = u^{n}_j - \kappa \dt B u^n, 
-  \label{eq:4th_order_parabolic_eqn_1d_FD_scheme}
-\eea
-where the fourth-order discrete bilaplacian is given by
-\beq
-  B u^n = \frac{ -u^{n}_{j+3} + 12 u^{n}_{j+2} - 39 u^{n}_{j+1}
-               + 56 u^{n}_j
-               - 39 u^{n}_{j-1} + 12 u^{n}_{j-2} -u^{n}_{j-3} }
-               {6 \dx^4}.
-\eeq
-This scheme is formally first-order in time and fourth-order in space.  
-Due to the stability constraint $\dt \le 3\dx^4/40 \kappa$, the scheme is 
-$O(\dx^4)$ accurate overall.
-
-Proceeding in the usual way, we use Taylor series expansions and the PDE to 
-express the true solution, $\tu$, at time step $(n+1)$ in terms of the true
-solution and its spatial derivatives at time step $n$: 
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  &+& \dt \left( -\kappa \frac{\partial^4 \tu^n}{\px^4}  
-             + f^n_j
-        \right)
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
-    \left(
-      \kappa^2 \frac{\partial^8 \tu^n}{\px^8} 
-    - \kappa \frac{\partial^4 f}{\px^4} 
-    + \frac{\partial f}{\pt}
-    \right)
-  + O \left( \dt^3 \right)
-  \label{eq:4th_order_parabolic_eqn_1d_time_err_modified}.
-\eea
-Next, we substitute $\tu$ into central difference approximation to the 
-bilaplacian to obtain
-\bea
-  B \tu^n = \frac{\partial^4 \tu}{\px^4} 
-  - \frac{7 \dx^4}{240} 
-    \frac{\partial^8 \tu}{\px^8} 
-  + O(\dx^6)
-  \label{eq:4th_order_parabolic_eqn_1d_space_err}
-\eea
-Combining these results with the finite difference 
-scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}), we find that the
-local truncation error is given by
-\bea
-  \tau^n_j &=& 
-     \frac{\partial^8 \tu^n}{\px^8} 
-    \left[ \frac{7 \dx^4}{240} - \frac{\kappa \dt}{2}  \right] (\kappa \dt)
-    \nonumber \\
-    &+& \frac{\dt^2}{2} 
-        \left( \kappa \frac{\partial^4 f}{\px^4} 
-             - \frac{\partial f}{\pt}
-        \right)
-    + O(\dt \dx^6) + O(\dt^3)
-  \label{eq:4th_order_parabolic_eqn_1d_err_eqn}
-\eea
-Thus, the optimal time step is $\dto = 7 \dx^4/120\kappa$ and the correction
-term is 
-\beq
-  - \frac{\dt^2}{2} 
-      \left( \kappa \frac{\partial^4 f}{\px^4} 
-           - \frac{\partial f}{\pt}
-      \right).
-\eeq
-Together, these modifications lead to a finite difference scheme that is 
-sixth-order accurate.  As with the previous examples, the correction term 
-can be computed analytically or numerically.  The improved accuracy of 
-(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) when the optimal time step 
-and correction terms are used is illustrated in 
-Figure~\ref{fig:4th_order_parabolic_eqn_1d_error}.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
-\ \ 
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_comp_time}} 
-\caption{$L^\infty$ error as a function of number of grid points (left)
-and computation time as a function of $L^\infty$ error (right) for two 
-finite difference schemes that solve the fourth-order parabolic equation 
-(\ref{eq:4th_order_parabolic_eqn_1d_space_err}):
-forward Euler with optimal time step and correction term (circles)
-and forward Euler with suboptimal time step $\dt = \dx^2/16$ and no 
-correction terms (squares).
-Both schemes use a fourth-order, central finite difference approximation for
-the bilaplacian.  Using each method, we solve the fourth-order parabolic
-equation with $\kappa = 1$ to time $t = 0.001$ on the domain $0 < x < 1
-$ subject to the boundary conditions $u(0,t) = u(1,t) = 0$ and the initial 
-condition $u(x,0) = 2\sin(2 \pi x) + \sin(5 \pi x)$.  The value of the solution
-in the ghost cells (required as a result of the wide stencil for the 
-discrete bilaplacian) are filled by taking values from the exact solution: 
-$u(x,t) = 2\sin(2 \pi x)\cos(10000 \pi t) 
-        + \sin(5 \pi x) \left[ 2-\exp(-100\pi^2 t) \right]$.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:4th_order_parabolic_eqn_1d_error}
-\end{center}
-\end{figure}
-
-As indicated by the discussion in this section, applying optimal time step 
-selection to high-order PDEs requires the use of high-order discretizations
-of the spatial derivatives.  While the need to derive high-order finite 
-difference stencils may seem like an unnecessary burden, it is important
-to remember that high-order finite difference schemes for spatial derivatives
-are almost always necessary to maximize the efficiency of explicit time
-integration schemes for high-order PDEs\footnote{Because restrictive time 
-step constraints are always a major limitation of explicit schemes for 
-time dependent PDEs with high-order spatial 
-derivatives~\cite{gko_book,greer_2006}, 
-the overall error of finite difference schemes for high-order PDEs is usually 
-dominated by the error introduced by the spatial discretization.  As a result, 
-it is crucial that numerical schemes for high-order PDEs use high-order 
-discretizations for the spatial derivatives.  Using high-order discretizations 
-increases the computational efficiency of numerical schemes and makes it 
-possible to obtain accurate solutions on relatively coarse grids.}. 
-Optimal time step selection merely provides a valuable way to leverage the 
-effort spent deriving high-order finite difference schemes in order to further 
-boost the overall accuracy of the numerical method.
-
-Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} shows the computational 
-performance of various finite difference methods for solving 
-equation~(\ref{eq:4th_order_parabolic_eqn_1d_second_time_derivative}).
-Observe the dramatic decrease in computational cost as a function of the
-error by simply replacing a second-order discretization for the bilaplacian 
-operator with a fourth-order discretization.  For the forward Euler scheme,
-using an optimal time step drives the computational cost down even more.
-The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} 
-shows that the theoretical estimates for the computation time are in good 
-agreement with timings from actual computational experiments.
-
-\begin{table}[tb]
-\caption{\label{tab:comp_perf_vs_err_4th_order_parabolic} 
-   Computational cost as a function of the global numerical error $e$
-   for various finite difference schemes that numerically solve the 1D 
-   fourth-order parabolic equation.  For all finite difference schemes, the 
-   numerical solution is assumed to be computed over a fixed time interval.
-}
-\begin{minipage}{\textwidth}
-\begin{center} \footnotesize
-\renewcommand{\arraystretch}{1.3}
-\begin{tabular}{|l|c|c|c|c|}
-  \hline
-  {\bf Numerical Scheme} & $\dx$ 
-  & $\dt$\footnote{The time step for the forward Euler scheme is $O(\dx^4)$
-   as a result of stability considerations.  The time step for the 
-   Crank-Nicholson schemes with second- and fourth-order spatial 
-   discretizations are taken to be $O(\dx)$ and $O(\dx^2)$, respectively, so 
-   that the temporal error is subdominant to the spatial error.}
-  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
-    to be $O(1/\dx)$.} 
-  & {\bf Compute Time}\footnote{The computation time is approximated as 
-    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
-    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
-    arise for the Crank-Nicholson method.}  \\
-  \hline 
-  Forward Euler with 
-    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{2} \right)$}
-    & \multirow{2}{*}{$O\left( e^{-1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-5/2} \right)$} \\
-  \ \ 2nd-Order Bilaplacian & & & & \\
-  \hline 
-  Forward Euler with 
-    & \multirow{2}{*}{$O\left( e^{1/4} \right)$}
-    & \multirow{2}{*}{$O\left( e \right)$}
-    & \multirow{2}{*}{$O\left( e^{-1/4} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-{5/4}} \right)$} \\
-  \ \ 4th-Order Bilaplacian & & & & \\
-  \hline 
-  Forward Euler with 
-    & \multirow{2}{*}{$O\left( e^{1/6} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{2/3} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-1/6} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-5/6} \right)$} \\ 
-  \ \ 4th-Order Bilaplacian and OTS & & & & \\
-  \hline 
-  Crank-Nicholson with 
-    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-1} \right)$} \\
-  \ \ 2nd-Order Bilaplacian & & & & \\
-  \hline 
-  Crank-Nicholson with 
-    & \multirow{2}{*}{$O\left( e^{1/4} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{1/2} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-1/4} \right)$} 
-    & \multirow{2}{*}{$O\left( e^{-3/4} \right)$} \\
-  \ \ 4th-Order Bilaplacian & & & & \\
-  \hline
-\end{tabular}
-\end{center}
-\end{minipage}
-\end{table}
-
-Comparing the best forward Euler and Crank-Nicholson schemes shows that
-the optimal forward Euler scheme more efficiently utilizes memory at the
-cost of requiring more computation time.  This tradeoff between memory and 
-computation only exists in one space dimension.  In two-dimensions, both 
-schemes require the same amount of compute time; in three or more space 
-dimensions, the forward Euler scheme requires less compute time than the
-Crank-Nicholson scheme.  The origin of the improved performance for the 
-forward Euler scheme with optimal time step in higher dimensions is the 
-significantly smaller size of the computational grid required to achieve a 
-desired error level.
-
-\section{\label{sec:applications_multidim}
-         Application to PDEs in Multiple Space Dimensions}
-Optimal time step selection can be used to boost the order of accuracy for 
-finite difference schemes in any number of space dimensions.  However, 
-several important issues arise for problems in more than one space dimension.  
-These issues are primarily related to the presence of cross-terms in the 
-discretization error, multiple grid spacing parameters,  and the domain 
-geometry.  In this section, we address these issues and demonstrate the use 
-of optimal time step selection in the construction of finite difference 
-schemes for two linear, time dependent PDEs in two space dimensions on both 
-regular and irregular domains.  It is straightforward to generalize the 
-approach presented in this section to more space dimensions and to construct 
-methods for semilinear PDEs by incorporating ideas discussed in 
-Sections~\ref{sec:optimal_time_step_selection} and~\ref{sec:applications_1d}.
-
-\subsection{\label{sec:ots_multidim_crossterms}
-            Discretization of Spatial Derivatives}
-A significant distinction between the error for finite difference schemes in 
-one space dimension and multiple space dimensions is the presence of 
-cross-terms in the discretization error.  When analyzing finite difference 
-schemes using conventional numerical analysis techniques, it is usually safe 
-to ignore the precise form of the cross-terms because they only appear as 
-higher-order corrections in the error.  As a result, the order of accuracy for 
-the finite difference scheme is not sensitive to the particular choice of 
-discretization for the spatial derivative operators.  In contrast, knowledge 
-of the form of the cross-terms is critical when attempting to use optimal time 
-step selection.  Without this knowledge, it is impossible to ensure 
-cancellation of higher-order errors.
-
-Recall that optimal time step selection requires higher-order temporal 
-errors to be directly related to higher-order spatial errors via the PDE.  
-This requirement can be satisfied in one of two ways:
-\begin{itemize}
-\item by carefully designing the spatial discretization so that 
-      higher-order spatial errors are directly related to the spatial 
-      derivative operators in the PDE; or
-\item by explicitly adding correction terms to cancel out any cross-terms 
-      in the original finite difference scheme that cannot be eliminated 
-      by using the PDE.
-\end{itemize}
-These two approaches are actually equivalent because the correction terms 
-can be combined with the finite difference stencils in the original scheme.  
-In essense, the correction terms can be viewed as part of an expanded spatial 
-discretization scheme and vice-versa.
-
-\subsection{Linear Advection Equation}
-For our first application of optimal time step selection in multiple
-space dimensions, we consider the two-dimensional advection
-equation
-\beq
-  \frac{\partial u}{\partial t} + \vec{A} \cdot \nabla u = 0,
-  \label{eq:advection_eqn_2d}
-\eeq
-where $\vec{A} = (A_x, A_y)$ is the advection velocity.  For convenience, we 
-take $A_x, A_y > 0$.  This choice allows us to avoid unnecessary complexity 
-associated with the dependence of the upwind direction on the sign of $A_x$ 
-and $A_y$.  It is straightforward to generalize our discussion to $A_x$ and 
-$A_y$ of arbitrary sign.
-
-Recall that in one space dimension, using an optimal time step for
-a simple forward Euler scheme that uses a first-order upwind discretization
-of the spatial derivative leads to a finite difference scheme for the 
-advection equation with no local truncation error.  The same result can be 
-achieved in two space dimensions.  However, because there are a total of three
-numerical parameters (\ie two grid spacings and the time step), it is 
-insufficient to optimize only the time step; the ratio of the grid 
-spacings also needs to be optimally selected.
-
-For the purposes of illustration, we consider a simple forward Euler scheme 
-with first-order upwind discretizations of the advection terms:
-\beq
-  u^{n+1}_{i,j} = u^{n}_{i,j}
-  - A_x \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} \right)
-  - A_y \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} \right)
-  \label{eq:advection_eqn_2d_FD_scheme}
-\eeq
-This method is formally first-order in both space and time.
-
-To compute the optimal time step and grid spacing ratio, we begin, as
-usual, by analyzing the discretization error 
-for~(\ref{eq:advection_eqn_2d_FD_scheme}).  As in our analysis of the finite
-difference scheme for the one-dimensional advection equation, we retain all of 
-the terms in the discretization error.  Using Taylor series to expand the true 
-solution of the PDE at time step $(n+1)$, we find that
-\bea
-  \tu^{n+1}_{i,j} = \tu^{n}_{i,j }
-  + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
-       \frac{\partial^k \tu^n}{\pt^k} 
-  \label{eq:advection_eqn_2d_time_err}.
-\eea
-where $\tu$ denotes the true solution of the PDE. 
-Similarly, Taylor series expansion of the upwind finite difference 
-approximation to the spatial derivative yields
-\bea
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i-1,j}}{\dx} = 
-  \frac{\partial \tu}{\px} 
-  - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k} 
-  \label{eq:advection_eqn_2d_space_err_x}
-  \\
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dy} = 
-  \frac{\partial \tu}{\py} 
-  - \frac{1}{\dy} \sum_{k=2}^\infty \frac{\left( -\dy \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\py^k} 
-  \label{eq:advection_eqn_2d_space_err_y}
-\eea
-Next, we use the PDE~(\ref{eq:advection_eqn_2d}) to replace all of 
-the time derivatives in~(\ref{eq:advection_eqn_2d_time_err}) with spatial
-derivatives to obtain
-\bea
-  \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
-  + \sum_{k=1}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left( A_x \partial_x
-            + A_y \partial_y \right)^k \tu^{n}.
-  \label{eq:advection_eqn_2d_time_err_modified}
-\eea
-Combining~(\ref{eq:advection_eqn_2d_space_err_x}), 
-(\ref{eq:advection_eqn_2d_space_err_y}), 
-and~(\ref{eq:advection_eqn_2d_time_err_modified}) with the finite difference 
-scheme~(\ref{eq:advection_eqn_2d_FD_scheme}), we find that the local
-truncation error is given by:
-\bea
-  \tau^{n}_{i,j} &=&
-      \frac{A_x \dt}{2} \frac{\partial^2 \tu}{\px^2}
-      \left( \dx - A_x \dt \right)
-    + \frac{A_y \dt}{2} \frac{\partial^2 \tu}{\py^2}
-      \left( \dy - A_y \dt \right)
-    - A_x A_y \dt^2 \frac{\partial^2 \tu}{\px \py}
-   \nonumber \\
-   &+& \frac{A_x\dt}{\dx} \sum_{k=3}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k} 
-   + \frac{A_y\dt}{\dy} \sum_{k=3}^\infty \frac{\left( -\dy \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\py^k} 
-   \nonumber \\
-   &-& \sum_{k=3}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left( A_x \partial_x + A_y \partial_y
-              \right)^k \tu^{n} 
-  \label{eq:advection_eqn_2d_trunc_err}
-\eea
-From this expression, we see that choosing the ratio of the grid spacings to
-be $\dx/\dy = A_x/A_y$ and setting the time step to be 
-$\dt_{opt} = \dx/A_x = \dy/A_y$ eliminates two of the three leading-order 
-terms in the discretization error.   The last leading-order term can be 
-eliminated by adding the correction term
-\bea
-   A_x A_y \dt^2 \frac{\partial^2 \tu}{\px \py}.
-  \label{eq:advection_eqn_2d_corr_term}
-\eea
-With these choices, we are able to boost the order of accuracy for the 
-finite difference scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from first- to 
-second-order.
-
-With a careful choice of discretization for the correction term, we can
-boost the order of accuracy even further.  Using the first-order upwind 
-discretization for $\frac{\partial^2 \tu}{\px \py}$
-\beq
-   \frac{u_{i,j} - u_{i-1,j} - u_{i,j-1} + u_{i-1,j-1}}
-        {\dx \dy} 
-  \label{eq:advection_eqn_2d_upwind_corr_approx}
-\eeq
-the truncation error~(\ref{eq:advection_eqn_2d_trunc_err}) becomes
-\bea
-  \tau^{n}_{i,j} &=&
-      \frac{A_x \dt}{2} \frac{\partial^2 \tu}{\px^2}
-      \left( \dx - A_x \dt \right)
-    + \frac{A_y \dt}{2} \frac{\partial^2 \tu}{\py^2}
-      \left( \dy - A_y \dt \right)
-   \nonumber \\
-   &+& \frac{A_x\dt}{\dx \dy} ( \dy - A_y \dt )
-       \sum_{k=3}^\infty \frac{\left( -\dx \right)^k}{k!} 
-              \frac{\partial^k \tu^n}{\px^k} 
-   \nonumber \\
-   &+& \frac{A_y\dt}{\dx \dy} ( \dx - A_x \dt )
-       \sum_{k=3}^\infty \frac{\left( -\dy \right)^k}{k!} 
-              \frac{\partial^k \tu^n}{\py^k} 
-   \nonumber \\
-   &+& \frac{A_x A_y\dt^2}{\dx \dy} 
-       \sum_{k=3}^\infty \sum_{l=0}^{k}
-              (-1)^k {k \choose l}
-              \frac{\dx^l \dy^{k-l}}{k!} 
-              \frac{\partial^k \tu^n}{\px^l \py^{k-l}} 
-   \nonumber \\
-   &-& \sum_{k=3}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left( A_x \partial_x + A_y \partial_y
-              \right)^k \tu^{n} 
-  \label{eq:advection_eqn_2d_trunc_err_mod},
-\eea
-which vanishes when the optimal time step and grid spacing ratio are used.
-Thus, when (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used in
-conjunction with the optimal time step and grid spacing ratio, we obtain
-a finite difference scheme that introduces no truncation error at each time 
-step.  Figure~\ref{fig:advection_eqn_2d_soln} demonstrates the amazing accuracy 
-of this scheme over the first-order forward Euler 
-scheme~(\ref{eq:advection_eqn_2d_FD_scheme}) without any modifications.
-Notice how the sharp corners of the initial conditions are perfectly 
-preserved when the optimal time step, optimal grid spacing ratio, and 
-correction term are used.  In contrast, the sharp corners are
-completely smoothed out in the solution computed using the standard 
-first-order upwind scheme.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/adv_eqn_2d_FE_OTS_soln}} 
-\scalebox{0.35}{\includegraphics{figures/adv_eqn_2d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 2D advection
-equation~(\ref{eq:advection_eqn_2d}) using forward Euler time stepping 
-with optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step (right).  The latter scheme uses a 
-time step $\Delta t = \Delta x / 2 (|A_x|+|A_y|)$.  
-Both figures show the numerical solution at $t = 3$ computed on the domain 
-$-10 < x,y < 10$ using 100 grid points in each direction.
-The advection velocity $(A_x, A_y)$ is set to $(-1, -2)$.  The initial 
-conditions are taken to be a 2D square function centered at the
-origin: $u(x,y,0) = 1$ if $|x| + |y| \le 2$ and 
-$u(x,y,0) = 0$ if $|x| + |y| > 2$. 
-For the time interval considered, $u$ is identically zero 
-on the boundary.
-}
-\label{fig:advection_eqn_2d_soln}
-\end{center}
-\end{figure}
-
-To see that the correction term can be considered as part of the spatial 
-discretization scheme, we can write out the complete finite difference scheme 
-including the correction term:
-\bea
-  u^{n+1}_{i,j} &=& u^{n}_{i,j}
-  - A_x \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} \right)
-  - A_y \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} \right)
-  \nonumber \\
-  & & + A_x A_y \dt^2 
-        \left( \frac{u_{i,j} - u_{i-1,j} - u_{i,j-1} + u_{i-1,j-1}}
-                    {\dx \dy} 
-        \right).
-  \label{eq:advection_eqn_2d_FD_scheme_complete}
-\eea
-Using the relationship between the optimal time step and the grid spacings, 
-this scheme can be rewritten as
-\bea
-  u^{n+1}_{i,j} = u^{n}_{i,j}
-  &-& \frac{A_x \dt}{2}
-    \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} 
-         + \frac{u^{n}_{i,j-1} - u^{n}_{i-1,j-1}}{\dx} 
-    \right)
-  \nonumber \\
-  &-& \frac{A_y \dt}{2}
-    \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} 
-         + \frac{u^{n}_{i-1,j} - u^{n}_{i-1,j-1}}{\dy} 
-    \right).
-  \label{eq:advection_eqn_2d_FD_scheme_simplified}
-\eea
-Thus, we see that the finite difference scheme 
-(\ref{eq:advection_eqn_2d_FD_scheme_complete}) is essentially forward Euler 
-time integration with first-order upwind discretizations of the spatial 
-derivatives that are computed using all four upwind neighbors.
-
-In higher space dimensions, it becomes tedious to carry out the analysis 
-to derive the infinite order accurate finite difference scheme for the 
-advection equation because many high-order correction terms must be retained.  
-However, it is straighforward to generalize 
-(\ref{eq:advection_eqn_2d_FD_scheme_simplified}) to $d$ space dimensions by 
-using first-order discretizations for the spatial derivatives that are 
-averages over the $2^{d-1}$ upwind discretizations in each coordinate 
-direction.   As with the two-dimensional problem, the optimal grid spacings 
-satisfy $\dx_k/A_{x_k} = \dt_{opt}$ for $1 \le k \le d$.
-
-Like the unit CFL condition for the one-dimensional advection equation, the 
-analogue of (\ref{eq:advection_eqn_2d_FD_scheme_simplified}) in any number
-of space dimensions could also have been derived by examining how 
-characteristic lines of the $d$-dimensional advection equation pass 
-through grid points.  For the two-dimensional problem, this perspective 
-becomes apparent when we fully simplify 
-(\ref{eq:advection_eqn_2d_FD_scheme_simplified}) using the 
-optimal time step and grid spacing ratio to obtain
-$u^{n+1}_{i,j} = u^{n}_{i-1,j-1}$.
-
-\subsection{Diffusion Equation}
-As for PDEs in one space dimension, achieving infinite order accuracy 
-through optimal parameter selection is unique to the advection equation.
-In general, optimal parameter selection boosts the accuracy by only a few 
-orders.  To illustrate the typical accuracy boost gained by using optimal 
-time step selection, we apply it to the 2D diffusion equation.  We also 
-use the 2D diffusion equation to demonstrate the utility of optimal time 
-step selection for problems on irregular domains.
-
-In two space dimensions, the diffusion equation may be written as
-\beq
-  \frac{\partial u}{\pt} = D \nabla^2 u,
-  \label{eq:diffusion_eqn_2d}
-\eeq
-where $\nabla^2$ is the Laplacian operator and $D$ is the diffusion constant. 
-While there are many finite difference schemes that are used to solve the 
-2D diffusion equation, the most common ones employ a five-point, second-order 
-central difference stencil for the Laplacian~\cite{iserles_book}:
-\beq
-  L^{5pt} u = \frac{u_{i,j+1} + u_{i,j-1}
-                   +u_{i+1,j} + u_{i-1,j} - 4u_{i,j}}{\dx^2},
-  \label{eq:laplacian_2d_5pt_stencil}
-\eeq
-which assumes that the grid spacing in the $x$ and $y$ directions are
-equal.  Common choices for the time integration method include forward Euler
-and the Crank-Nicholson method.  Each scheme has its advantages and
-disadvantages.  The forward Euler scheme is very easy to implement but is 
-only first-order accurate and suffers from stability constraints on the time 
-step; the Crank-Nicholson scheme is second-order accurate and unconditionally 
-stable but requires the solution of a linear system of equations at each time 
-step.  The forward Euler scheme using the five-point stencil is given by
-\bea
-  u^{n+1}_j &=& u^{n}_j 
-  + D \dt 
-    \left( \frac{u^{n}_{i,j+1} + u^{n}_{i,j-1}
-         +u^{n}_{i+1,j} + u^{n}_{i-1,j} - 4u^{n}_{i,j}}{\dx^2}
-    \right).
-  \label{eq:diffusion_eqn_2d_5pt_scheme}
-\eea
-This scheme is formally first-order in time and second-order in space.
-
-While it is certainly possible to directly apply optimal time step selection 
-to (\ref{eq:diffusion_eqn_2d_5pt_scheme}), we would be forced to use additional
-correction terms to deal with the fact that the leading-order error for 
-(\ref{eq:laplacian_2d_5pt_stencil}), 
-$\dx^2 \left(u_{xxxx} + u_{yyyy}\right)/12$, cannot be directly related 
-to the spatial derivative operator in the diffusion equation.  
-The analysis is more straightforward if we construct our finite difference 
-scheme using the nine-point stencil for the 
-Laplacian~\cite{iserles_book,patra_2005}, which does possess a leading-order 
-error directly related to the spatial derivative operator in the PDE 
-(\ref{eq:diffusion_eqn_2d}):
-\bea
-  L^{9pt} u = \frac{1}{6 \dx^2} \left( 
-      \begin{array}{l}
-         \ \ \ u_{i+1,j+1} + u_{i+1,j-1}
-            + u_{i-1,j+1} + u_{i-1,j-1} \\
-         +\ 4 u_{i+1,j} + 4 u_{i-1,j}
-           + 4 u_{i,j+1} + 4 u_{i,j-1}
-           -20 u_{i,j} 
-      \end{array}
-    \right). 
-  \label{eq:laplacian_2d_9pt_stencil}
-\eea
-Observe that the nine-point stencil is simply the standard five-point 
-stencil plus $\dx^2/6$ times a second-order, central difference 
-approximation to $u_{xxyy}$.  Therefore, we can obtain the leading-order 
-error in the nine-point stencil by combining the leading-order error for the 
-five-point stencil with the extra contribution of 
-$\dx^2 u_{xxyy}/6$
-\bea
-L^{9pt} u &=& \nabla^2 u 
-              + \frac{\dx^2}{12} \left(u_{xxxx} + 2 u_{xxyy} + u_{yyyy}\right)
-              + O(\dx^4)
-          \nonumber \\
-          &=& \nabla^2 u + \frac{\dx^2}{12} \nabla^4 u + O(\dx^4).
-  \label{eq:laplacian_2d_9pt_stencil_error}
-\eea
-Like (\ref{eq:diffusion_eqn_2d_5pt_scheme}), the forward Euler scheme that 
-uses the nine-point discretization of the Laplacian 
-\beq
-  u^{n+1}_{i,j} = u^{n}_{i,j}
-  + D \dt L^{9pt} u
-  \label{eq:diffusion_eqn_2d_9pt_scheme}
-\eeq
-is $O(\dx^2)$ accurate overall because it has a stability constraint 
-$\dt \le 3\dx^2/8D = O(\dx^2)$.
-It is interesting to note that use of a nine-point stencil for the Laplacian 
-naturally arises when directly applying optimal time step selection to the
-five-point scheme if the correction terms are interpreted as a modification 
-to the spatial discretization.
-
-To apply optimal time step selection to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we 
-proceed in the usual way by first analyzing the leading-order terms of the 
-discretization error for the finite difference scheme.  Taylor series 
-expansions show that the true solution satisfies the following equations:
-\bea
-  \tu^{n+1}_{i,j} &=& \tu^{n}_{i,j}
-  + \dt \frac{\partial \tu^n}{\pt} 
-  + \frac{\dt^2}{2} \frac{\partial^2 \tu^n}{\pt^2} + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_2d_time_err} 
-  \\
-  L^{9pt} \tu &=&
-  \nabla^2 \tu + \frac{\dx^2}{12} \nabla^4 \tu + O(\dx^4)
-  \label{eq:diffusion_eqn_2d_space_err}
-\eea
-Using the PDE~(\ref{eq:diffusion_eqn_2d}) to replace time derivatives
-with space derivatives, (\ref{eq:diffusion_eqn_2d_time_err}) becomes
-\bea
-  \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
-  + D \dt \nabla^2 \tu^{n}_{i,j}
-  + \frac{(D \dt)^2}{2} \nabla^4 \tu^{n}_{i,j}
-  + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_2d_time_err_modified}
-\eea
-Combining these results, we find that the local truncation error for the
-finite difference scheme (\ref{eq:diffusion_eqn_2d_9pt_scheme}) is given
-by 
-\bea
-  \tau^{n}_j = 
-  \nabla^4 \tu
-      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-      + O(\dt \dx^4) + O(\dt^3).
-  \label{eq:diffusion_eqn_2d_trunc_err}
-\eea
-As for the 1D diffusion equation, we see that the optimal time step is given 
-by $\dto = \dx^2/6D$ and that this choice of time step yields a scheme 
-that is fourth-order accurate.  Note that there is no need to consider an 
-implicit scheme because $\dto$ is stable for the forward Euler time 
-integration scheme.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_error_vs_N}} 
-\ \ 
-\scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_comp_time}} 
-\caption{$L^\infty$ error as a function of number of grid points (left)
-and computation time as a function of $L^\infty$ error (right) for three
-finite difference schemes that solve the 2D diffusion equation:
-forward Euler with optimal time step and correction terms (circles), forward 
-Euler with suboptimal time step $\dt = \dx^2/4D$ (squares), and 
-Crank-Nicholson with time step $\dt = \dx/16D$ (diamonds).  
-The forward Euler schemes use a nine-point discretization of the Laplacian;
-the Crank-Nicholson scheme uses the standard five-point discretization of
-the Laplacian.  Using each method, we solve the 2D diffusion equation with 
-$D = 0.25$ and source term 
-$f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
-            \exp\left(-5 D \pi^2 t\right)
-          - 24 D \pi^2 \sin(5 \pi x) \sin(\pi y) 
-            \exp \left(-2 D \pi^2 t \right)$.
-to time $t = 0.1$ on the domain $0 < x, y < 1$ subject 
-to the initial condition
-$u(x,y,0) =  1 + y/3 + x/2 + xy/4 
-          + \sin(2 \pi x) \sin(3 \pi y) - \sin(5 \pi x) \sin(\pi y)$.
-The boundary conditions are taken from the analytical solution 
-$u(x,t) = 1 + y/3 + x/2 + xy/4 
-          + \sin(2 \pi x) \sin(3 \pi y) \exp(-5 D \pi^2 t) 
-          - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)
-$.
-Note that the errors for the forward Euler with suboptimal time step and 
-Crank-Nicholson schemes lie almost directly on top of each other at the 
-resolution of the figures.  
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:diffusion_eqn_2d_src_analysis}
-\end{center}
-\end{figure}
-
-The improvement in the accuracy of the forward Euler scheme when the optimal 
-time step is used is shown in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
-Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also shows that the 
-scaling of the computational time with the error is in good agreement with 
-theoretical estimates (see Table~\ref{tab:comp_perf_vs_dim}).  As with the
-other examples, the forward Euler scheme with an optimal time step is much 
-more computationally efficient than the two other schemes.
-
-\subsubsection{Irregular Domains}
-All of the previous examples share an important feature: they are all
-defined on simple, rectangular domains.  Thus, any of several methods 
-could have been used to obtain solutions with similar or even higher accuracy.
-For example, a fourth-order accurate solution to the diffusion equation
-(in any number of space dimensions) could be obtained by using a 
-pseudospectral spatial discretization~\cite{trefethen_spectral_book, 
-boyd_spectral_book} of the Laplacian and a fourth-order time integration scheme 
-(\eg Runge-Kutta).  When the domain is irregularly shaped, high-order methods 
-are harder to design and implement.  For instance, spectral and pseudospectral 
-methods are not directly applicable because the domain is not a simple 
-cartesian product of intervals.  While it is possible to use spectral methods
-for problems on irregular domains through the clever introduction of a phase 
-field~\cite{bueno_orovio_2006, feng_2006}, this approach does not 
-actually achieve spectral accuracy because the error is dominated by the 
-width of the smoothed approximation to the phase field (as opposed to the 
-accuracy of the discretization for the spatial 
-derivatives)~\cite{bueno_orovio_2006}.
-
-% FIGURE SHOWING GHOST CELLS AT IRREGULAR BOUNDARY
-% PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.25}{\includegraphics{figures/ghostcell_edge}} 
-\ \ \ \ \ \ \ \ \ \ \ \ \ 
-\scalebox{0.25}{\includegraphics{figures/ghostcell_corner}}
-\caption{Illustrations of edge (left) and corner (right) ghostcells.
-The edge ghostcell, $u_E$, is filled by using cubic extrapolation of 
-the value on the boundary, $u_B$, and values at three interior grid points.  
-The corner ghostcell, $u_C$, is filled by using a fourth-order accurate
-Taylor series expansion centered at $u_{0,0}$, which uses the values of $u$ 
-at the specified interior grid points and neighboring edge ghost cells to 
-approximate the partial derivatives in the expansion. 
-}
-\label{fig:ghost_cells}
-\end{center}
-\end{figure}
-
-In this section, we extend the OTS finite difference scheme for the 2D 
-diffusion equation on regular domains to irregular domains.  To avoid the 
-need for special stencils at grid points near the irregular boundary, we adopt 
-the `ghost cell' approach for imposing boundary conditions~\cite{fedkiw_1999, 
-osher_fedkiw_book, gibou_2005}.  The basic idea is to set the value of $u$ at 
-any grid point that lies immediately outside of the domain (and thus may be
-required by spatial derivative operators at interior grid points) by 
-extrapolating from values of $u$ at interior grid points and using boundary 
-condition information.  
-The main challenge is to find a method for filling ghost cells that does not 
-decrease the accuracy of the solution in the interior of the domain.  For 
-simplicity, we choose to represent the irregular boundary implicitly as the 
-zero level set of a level set function $\phi$ which satisfies:
-\bea
-  \begin{array}{ll}
-  \phi < 0 & \mathrm{for} \ x \in \Omega \\
-  \phi > 0 & \mathrm{for} \ x \notin \Omega \\
-  \phi = 0 & \mathrm{for} \ x \in \partial \Omega 
-  \end{array}
-\eea
-where $\Omega$ and $\partial \Omega$ are the interior and boundary of the 
-irregular domain.
-
-Figure~\ref{fig:ghost_cells} shows examples of ghost cells that lie in the 
-vicinity of an irregular boundary.  Notice that there are two types of ghost 
-cells:  edge and corner.  Edge and corner ghost cells are distinguished by 
-the relative position of the nearest interior grid point.  An edge ghost cell 
-is linked to its nearest interior neighbor by an edge of the computational 
-grid whereas a corner ghost cell and its nearest interior neighbor lie on the 
-opposite corners of a grid cell.
-
-To set the value of $u$ in each edge ghost cell, we use 1D, cubic 
-extrapolation of the values of $u$ across the interface.  A cubic Lagrange 
-extrapolant is constructed for each edge ghost cell using a point on the 
-domain boundary and three interior grid points that lie along the line 
-parallel to the edge connecting the ghost cell to its nearest interior 
-neighbor (see Figure~\ref{fig:ghost_cells}).  
-The location of the point $u_B$ can be calculated by inverting the linear 
-interpolant of $\phi$ that passes through the ghost cell and its nearest
-interior neighbor\footnote{We found that using higher-order interpolants to 
-locate boundary points was unnecessary for achieving high-order accuracy in 
-the computed solution.}.
-Because the quality of the extrapolant deteriorates if the boundary point is 
-too close to any of the interior grid 
-points used to construct the Lagrange extrapolant, we follow 
-\cite{gibou_2005} and choose the interior grid points so that the nearest
-interior grid point is sufficiently far from the boundary point.  
-Gibou~\etal~suggest shifting the interpolation points by one grid point 
-when the distance between the boundary point and the nearest interior grid 
-point is less than $\dx^2$~\cite{gibou_2005}.  For the numerical scheme 
-presented in this paper, this threshold was too low.  We found that better 
-results were obtained if we used an $O(\dx)$ threshold.
-
-% FIGURE SHOWING SOLUTION OF 2D DIFFUSION EQUATION ON STARFISH DOMAIN
-% PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_2d_starfish_domain_soln}} 
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_2d_starfish_domain_error}} 
-\caption{Numerical solutions (left) and dominant error (right) for the 2D 
-diffusion equation on a starfish-shaped domain.  The solution is computed at 
-time $t = 0.1$ using forward Euler time stepping with an optimal time step 
-and correction terms.  In the error plot, the dark regions represent points 
-where the error in the solution is larger than $25$\% of the $L^\infty$ error 
-of the solution.  For these figures, we use the same diffusion constant, 
-source term, and initial conditions as described in the caption for 
-Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.  The boundary of the 
-starfish-shaped domain is represented by the zero-level set of the function
-$\sqrt{x^2 + y^2} - \frac{3}{5} \left(1 + \frac{1}{2}\sin(5 \theta) \right)$, 
-where $\theta = \mathtt{atan}\left( y/x \right)$.
-The Dirichlet boundary conditions are taken from the values of the analytical 
-solution (see caption for Figure~\ref{fig:diffusion_eqn_2d_src_analysis})
-on the irregular boundary.  The numerical solution is calculated using the 
-square computational domain $-1 < x,y < 1$ which is discretized using 400 grid 
-points in each coordinate direction (the plot of the solution has been 
-downsampled for visualization purposes).  Boundary conditions are imposed by 
-using fourth-order accurate approximations of the value of the solution in 
-edge and corner ghost cells (described in the main text).
-}
-\label{fig:diffusion_eqn_2d_starfish_domain}
-\end{center}
-\end{figure}
-
-To set the value of $u$ in a corner ghost cell, we use a fourth-order accurate
-Taylor series expansion of $u$ centered at the nearest interior 
-neighbor.  The partial derivatives in the Taylor series expansion are 
-approximated using finite differences constructed from the values of $u$ 
-at interior grid points and the neighboring edge ghost cells (see 
-Figure~\ref{fig:ghost_cells}).  To obtain a fourth-order accurate value for 
-the corner ghost cell, we need to carefully choose the finite difference 
-stencils used for the partial derivatives.  For the corner ghost cell shown 
-in Figure~\ref{fig:ghost_cells}, the extrapolation stencil is given by
-\bea
-  u_C &=&  -4 u_{0,0} - u_{-1,-1} + 2 u_{1,0} + 2 u_{0,1}
-      - u_{1,-1} - u_{-1,1} + 2 u_{0,-1} + 2 u_{-1,0}
-\eea
-where $u_C$ and $u_{i,j}$ are indicated in the figure.  The stencils for
-corner ghost cells that are in other positions relative to the interior of 
-the domain are obtained by rotations of Figure~\ref{fig:ghost_cells} and 
-remapping of the indicies in the above stencil.
-
-Using these methods for filling edge and corner ghost cells, it is 
-straightforward to compute high-order accurate solutions of the 2D diffusion 
-equation on irregular domains.  We simply use the OTS forward euler scheme 
-designed for regular domains for grid points in the interior of the domain
-after filling the ghost cells with high-order accurate values.  
-Figure~\ref{fig:diffusion_eqn_2d_starfish_domain} shows an example of 
-a solution computed on a starfish shaped domain.  It also shows that the 
-error in the solution is concentrated near the boundaries of the irregular
-domain.  Thus, we see that it is necessary to use high-order accurate 
-extrapolants for the ghost cells to prevent errors at the boundary from 
-destroying high-order accuracy throughout the domain.  The improved accuracy 
-of the OTS solution compared to a simple forward Euler scheme with a 
-suboptimal time step is illustrated in 
-Figure~\ref{fig:diffusion_eqn_2d_starfish_error}.
-
-% FIGURE SHOWING SOLUTION OF 2D DIFFUSION EQUATION ON STARFISH DOMAIN
-% PLACED HERE TO SPREAD FIGURES OUT
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{
-  \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for two
-finite difference schemes that solve the 2D diffusion equation on an 
-irregular domain: forward Euler with optimal time step and correction 
-terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/8 D$
-(squares).  Both methods solve the 2D diffusion equation described in the 
-caption for Figure~\ref{fig:diffusion_eqn_2d_src_analysis} on the irregular
-domain defined in the caption for 
-Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:diffusion_eqn_2d_starfish_error}
-\end{center}
-\end{figure}
-
-\subsubsection{Diffusion Equation in Higher Dimensions}
-Using optimal time step selection to boost the accuracy of finite difference
-schemes for the diffusion equation in higher space dimensions is 
-straightforward.  As for the 2D problem, the key step is identification of a 
-finite difference stencil for the Laplacian that has an \emph{isotropic} 
-discretization error.  For 3D problems, a catalog of several such stencils is 
-available in~\cite{patra_2005}.  For higher dimensional problems, using 
-symbolic algebra software to derive finite difference stencils with desired 
-discretization error properties can be helpful~\cite{patra_2005, gupta_1998}.  
-Once an acceptable finite difference approximation for the Laplacian has been 
-chosen, the usual OTS analysis yields an optimal time step of 
-$\dt_{opt} = \dx^2/6D$ for forward Euler time integration and a boost of the 
-order of accuracy from two to four.  Above three dimensions, it is necessary 
-to use an implicit time integration scheme because $\dt_{opt}$ is no longer 
-stable for the forward Euler scheme.
-
-Problems on irregular domains can be handled using the ghost cell approach. 
-While the derivation of high accuracy stencils for ghost cells becomes more
-tedious, the general approach is the same as for the 2D problem.  The ghost 
-cells can first be grouped into separate classes based on the number of 
-grid cell edges that must be traversed to reach the nearest interior neighbor.
-The values in the ghost cells can then be set in order starting with those 
-directly connected to their nearest interior neighbor and ending with those 
-whose nearest interior neighbor is on the opposite corner of a grid cell.
-Symbolic algebra programs can be helpful when deriving methods for 
-extrapolating interior and boundary values to ghost cells.
-
-\section{\label{sec:summary} Summary} 
-In this paper, we have presented a novel approach for constructing high-order
-finite difference methods for time dependent PDEs \emph{without} requiring
-the use of high-order stencils or high-order time integration schemes.  
-The basic idea is that the leading-order terms in the discretization error
-can be eliminated by carefully choosing the time step and possibly adding
-a few simple correction terms.  It is straightforward to derive both the 
-optimal time step and the correction terms by (1) taking advantage of the 
-regular structure of finite difference schemes and (2) making use of the
-PDE to relate terms in the discretization error.  The boost in the order
-of accuracy achieved when the optimal time step and correction terms is well
-worth the extra effort required to explicitly retain and analyze the 
-leading-order terms in the discretization error.
-
-Through many examples, we have demonstrated the utility of optimal time step
-selection to a wide range of problems.  We showed how OTS selection can be 
-used to very easily to obtain high-order accurate solutions to most linear 
-and semilinear PDEs in any number of space dimensions on both regular
-and irregular domains.  The examples illustrate the surprisingly high-order
-of accuracy that can be achieved using only low-order discretizations for 
-spatial derivatives and simple forward Euler time integration.
-
-Optimal time step selection is a simple example of the more general problem
-of optimally selecting numerical parameters to boost the accuracy of 
-numerical methods.  Little research appears to have been done in this area,
-but, as OTS selection demonstrates, optimal selection of numerical parameters 
-can significantly impact the accuracy of numerical methods.  To realize
-the full potential of numerical methods, both optimal time step selection 
-and optimal numerical parameter selection deserve further investigation.
-
-% ACKNOWLEDGMENTS
-\section*{Acknowledgments}
-The author gratefully acknowledges the support of the Institute for 
-High-Performance Computing (IHPC) in Singapore and Vitamin D, Inc.
-The author would like to thank P. Fok and ?? for enlightening discussions 
-and suggestions on the manuscript.
-
-\bibliography{OTSPDE}
+\bibliography{OTSPDE-Systems}
 

### 2008-08-10 papers/OTSPDE-VarCoefs/OTSPDE-VarCoefs.tex
OTSPDE: * Checked in place holders for upcoming papers. * Updated information on potential journals to submit to.

--- 2008-08-10 papers/OTSPDE-Systems/OTSPDE-Systems.tex
@@
 \title{High-Order Accurate Finite Difference Schemes \\
-       for Systems of PDEs via Optimal Time Step Selection}
+       for Variable-Coefficient PDEs via Optimal Time Step Selection}
 
@@
 Kevin T. Chu\footnotemark[2] \footnotemark[3]
+and 
+James V. Lambers\footnotemark[4]
 }
@@
 
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\footnotetext[4]{Department of Energy Resources Engineering, Stanford University, Stanford, CA 94305}
+
 \renewcommand{\thefootnote}{\arabic{footnote}}
@@
 \begin{abstract}
-place holder...
+place holder
 \end{abstract}
@@
 optimal time step, finite difference schemes, high-order accurate numerical 
-methods, time dependent PDEs, systems of PDEs
+methods, time dependent PDEs, variable-coefficient PDEs
 \end{keywords}
@@
 \begin{AMS}
-65-02, 65M06, 65M12, 65M20, ??
+65-02, 65M06, 65M12, 65M20 ??
 \end{AMS}
@@
 \thispagestyle{plain}
-\markboth{KEVIN T. CHU}
-         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEPS}
+\markboth{KEVIN T. CHU AND JAMES V. LAMBERS}
+         {??}
 
 \section*{Introduction}
-High-order numerical methods for partial differential equations (PDEs) will 
-always be valuable for increasing the computational efficiency of numerical 
-simulations.  Thus, it is not at all surprising that a great deal of effort in 
-numerical PDEs continues to be focused on the development of high-order 
-numerical schemes~\cite{bruger_2005, gibou_2005, ito_2005, shukla_2005, 
-shukla_2007}.  
-Typically, high-order accuracy is achieved by constructing
-schemes that have high formal orders of accuracy.  However, high-order 
-accurate numerical solutions can also be obtained by using formally low-order 
-schemes in clever ways.  When possible, the latter approach can be a powerful 
-way to boost the accuracy of a numerical method \emph{without} introducing too 
-much additional algorithmic (and programming) complexity.
+\cite{zoppou1999, kreiss2002, guidotti2006}
 
-\section{\label{sec:summary} Summary}
-
+% ACKNOWLEDGMENTS
 \section*{Acknowledgments}
@@
 and the Institute for High-Performance Computing (IHPC) in Singapore. 
-The author would like to thank K.-H. Chiam, ?? 
+The author would like to thank ??
 for helpful suggestions on the manuscript.
-The author particularly thanks K.-H. Chiam for pointing him to the 
-mathematical biology literature for interesting examples of reaction-diffusion 
-systems.
 
-\bibliography{OTSPDE-Systems}
+\bibliography{OTSPDE-VarCoefs}
 

### 2008-11-15 papers/OTSPDE/OTSPDE.tex
OTSPDE: - revised draft of OTSPDE paper for submission to JCP

@@
 % Submission draft: 2008/06/08, Vitamin D, Inc.
+% Resubmission draft: 2008/11/15, Vitamin D, Inc.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\documentclass[oneeqnum,onefignum,onetabnum,onethmnum]{siamltex}
-\usepackage{graphics}
+%\documentclass[fleqn,12pt,twoside,draft]{article}
+\documentclass[fleqn,12pt,twoside]{article}
+\usepackage[headings]{espcrc1}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\PassOptionsToPackage{ctagsplt,righttag}{amsmath}
 \usepackage{multirow}
@@
 
-\title{High-Order Accurate Finite Difference Schemes \\
-       Via Optimal Time Step Selection}
-
-\author{
-Kevin T. Chu\footnotemark[2] \footnotemark[3]
-}
+%******************************SIZE DECLARATIONS********************************
+\topmargin -0.7in
+
+%*******************************DOCUMENT BEGIN**********************************
 
@@
 \bibliographystyle{unsrt}
+
+\title{Boosting the Accuracy of Finite Difference Schemes 
+       via Optimal Time Step Selection}
+
+\author{K. T. Chu\address{Vitamin D, Inc., Menlo Park, CA 94025}$^,$\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore} 
+}
+
+\runtitle{Optimal Time Step Selection for FD Schemes}
+\runauthor{K. T. Chu}
+
 \maketitle
 
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
-\footnotetext[2]{Vitamin D, Inc., Menlo Park, CA 94025}
-
-\renewcommand{\thefootnote}{\fnsymbol{footnote}}
-\footnotetext[3]{Institute of High Performance Computing, A*STAR, Singapore, Singapore}
-
-\renewcommand{\thefootnote}{\arabic{footnote}}
+\noindent \rule{6.3in}{1pt}
 
 \begin{abstract}
-In this article, we present a novel technique for computing high-order
-accurate numerical solutions to time dependent partial differential 
-equations using only formally low-order finite difference stencils and
-time integration schemes.  We achieve this high-order accuracy in a simple 
-manner by optimally choosing the time step for the low-order finite 
-difference scheme (and possibly adding a few correction terms).  Through 
-straightforward numerical analysis arguments, we explain the origin of the 
-boost in accuracy and estimate the computational cost of the high-order 
-methods obtained via optimal time step (OTS) selection.  We demonstrate the 
-utility of OTS selection on a wide array of classical PDEs in one and more 
-space dimensions and find that it is applicable to linear and semilinear PDEs 
-on both regular and irregular domains. 
+In this article, we present a novel technique for boosting the order of 
+accuracy of formally low-order finite difference schemes for time dependent 
+partial differential equations by optimally selecting the time step used
+to advance the numerical solution.  This technique allows us to extract as
+much accuracy as possible from existing finite difference schemes with 
+minimal additional effort.  Through straightforward numerical 
+analysis arguments, we explain the origin of the boost in accuracy and 
+estimate the computational cost of the resulting numerical 
+method.  We demonstrate the utility of optimal time step (OTS) selection on a 
+wide array of classical PDEs in one and more space dimensions to illustrate 
+its applicability to linear and semilinear PDEs on both regular and irregular 
+domains.  
 \end{abstract}
 
-\begin{keywords}
-optimal time step, finite difference schemes, high-order accurate numerical 
-methods, time dependent PDEs
-\end{keywords}
-
-\begin{AMS}
-65-02, 65M06, 65M12, 65M20
-\end{AMS}
-
-\pagestyle{myheadings}
-\thispagestyle{plain}
-\markboth{KEVIN T. CHU}
-         {HIGH-ORDER FD SCHEMES VIA OPTIMAL TIME STEPS}
-
-\section*{Introduction}
+\noindent \rule{6.3in}{1pt}
+
+\section{Introduction}
 High-order numerical methods for partial differential equations (PDEs) will 
@@
 numerical PDEs continues to be focused on the development of high-order 
-numerical schemes~\cite{bruger_2005, gibou_2005, ito_2005, shukla_2005, 
-shukla_2007}.  
+numerical 
+schemes~\cite{bruger_2005,gibou_2005,ito_2005,shukla_2005,shukla_2007}.  
 Typically, high-order accuracy is achieved by constructing
-schemes that have high formal orders of accuracy.  However, high-order 
-accurate numerical solutions can also be obtained by using formally low-order 
-schemes in clever ways.  When possible, the latter approach can be a powerful 
-way to boost the accuracy of a numerical method \emph{without} introducing too 
-much additional algorithmic (and programming) complexity.
-
-There are many high-order accurate methods that are simple combinations or 
-modifications of low-order schemes.  A few well-known examples include 
-Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
-nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
-Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
-leveque_book_2002, gko_book}.  Cancellation of higher-order errors is a key 
-principle underlying the superior accuracy of these methods.
-
-More commonly, high-order accurate numerical methods are constructed by 
-requiring that the discretization error possesses the desired formal order of 
-accuracy for \emph{all} of the independent variables in the PDE.  While this 
-constraint is certainly sufficient for high-order accuracy, it is by no means 
-necessary.  This conventional approach for constructing numerical methods 
-has its roots in the classical numerical analysis of PDEs, which tends to 
-analyze the discretization of the terms in the PDE individually.  Within this 
-framework, it is natural to seek high-order accuracy by ensuring that the 
-numerical scheme is simultaneously high-order accurate for all independent 
-variables.  Unfortunately, this common approach for developing and analyzing 
-numerical methods is often suboptimal because it neglects the potential for 
-achieving high-order accuracy using simple low-order schemes.
-
-Perhaps the greatest weakness of the standard method for constructing 
-high-order numerical methods is the use of ``big-O" notation, which hides 
-the details of the discretization error.  The information lost in the 
-``big-O" notation eliminates the possibility of reducing numerical error by 
-taking advantage of relationships between the terms in the error.  The 
-modified nine-point formula for the Poisson equation~\cite{iserles_book} 
-and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
-leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
-higher-order terms in the discretization error when analyzing numerical 
-schemes.  For both of these methods, examination of higher-order terms allows 
-low-order methods to be transformed into high-order methods with only minor 
-modifications to the numerical scheme.
-
-The usual approach for developing numerical methods has another significant
-shortcoming -- it places no value on the optimization of numerical parameters 
-(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
-somewhat to blame for this situation because it gives the impression 
-that the error depends on each numerical parameter independently.  Aside
-from constraints needed for numerical stability, ``big-O" notation makes 
-it appear as though any combination of numerical parameters is equally good 
-as long as they are all ``small enough.''  For some problems, however, 
-optimizing the numerical parameters can significantly increase the accuracy
-of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
-that choosing an appropriate value for $\theta$ as a function of the time 
-step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
-the weight factor and the theta-method as the single-step trapezoidal 
-method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
-element solutions to the diffusion equation.
-
-In this paper, we explore a novel technique for constructing high-order 
-finite difference methods for time dependent PDEs using only formally 
-low-order stencils and time integration schemes.  The basic idea is that a 
-carefully chosen time step (and possibly the addition of a few correction 
-terms) can eliminate low-order terms in the discretization error, allowing a 
-formally low-order method to deliver high-order accuracy.  The optimal time 
-step is calculated in a straightforward manner by using knowledge of the 
-PDE to eliminate the leading-order terms in the discretization error.  
-Optimal time step selection combines the well-known approach of using 
-the PDE to reduce numerical error with the less-common notion that optimal 
-choice of numerical parameters can boost the effectiveness of algorithmically 
-simple and computationally low-cost methods.
-
-Using the PDE to reduce discretization error has a long 
-tradition in numerical analysis.  Indeed, this idea underlies both the 
-modified nine-point formula for the Poisson equation~\cite{iserles_book} and 
-the Lax-Wendroff method for the linear advection 
-equation~\cite{leveque_book_1992, leveque_book_2002, gko_book}.  
-In the case of the modified nine-point formula for the Poisson equation, a 
-simple analysis of the nine-point stencil for the Laplacian on a uniform grid 
-shows that it is only second-order accurate, which might lead to the conclusion 
-that the nine-point stencil has no advantage over the standard five-point 
-stencil.  However, a more careful analysis of the discretization error in 
-light of the PDE reveals that the leading-order error, which is proportional 
-to the bilaplacian of the solution~\cite{iserles_book, patra_2005}, is directly 
-related to the Laplacian of the source term in the Poisson equation.  Thus, 
-the numerical scheme is easily boosted to fourth-order by simply adding a 
-correction term that cancels out the leading-order error.   
-Similarly, use of the PDE to eliminate higher-order discretization errors also 
-plays a critical role in the Lax-Wendroff method because it allows the second 
-derivative of the solution in time to be written in terms of the second 
-derivative in space.  This observation transforms an unstable, first-order 
-scheme based on a central difference approximation of 
-$\partial/\partial x$ into a stable, second-order 
-scheme~\cite{leveque_book_1992, leveque_book_2002}.
-
-Optimal time step selection as a technique for improving the overall accuracy 
-of numerical schemes for time dependent PDEs appears to be a novel idea.
-Typically, the primary concern when choosing a time step is numerical 
-stability with accuracy being a secondary consideration.  The idea that a 
-clever choice of time step can be used to dramatically reduce the error is 
-rarely considered.  In general, little research seems to have been done in
-the area of optimizing the parameters of numerical methods for PDEs (\eg time 
-step and grid spacing) to improve accuracy.  To the best of the author's 
-knowledge, Zhao's method for choosing an optimal value of $\theta$ in the 
-theta-method is the only work that attempts to improve accuracy by modifying 
-numerical parameters.  While he focuses solely on the optimal choice of 
-$\theta$ for a fixed time step, his approach can be viewed as a form of 
-optimal time step selection by reversing the roles of $\theta$ and the time 
-step.
+schemes that are formally high-order accurate.  However, high-order 
+accuracy can also be obtained by using formally low-order schemes in clever 
+ways (\eg Richardson extrapolation and compact finite difference schemes).  
+When possible, the latter approach can be a powerful way to boost the accuracy 
+of a numerical method \emph{without} introducing too much additional 
+algorithmic (and programming) complexity.
+
+Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
+but relatively uncommon, approach for maximizing the accuracy of numerical 
+schemes.  Often, all combinations of numerical parameters are considered 
+equally good.  That is, the simplest model of the error as a function 
+of the numerical parameters is assumed:  each numerical parameter contributes
+independently to the error.  For some problems, however, optimizing the 
+numerical parameters can significantly increase the accuracy of the 
+numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed that 
+choosing an appropriate value for the weight factor $\theta$ in the 
+theta-method can decrease the error in finite element solutions to the 
+diffusion equation.
+
+In this paper, we present a novel technique for boosting the order of 
+accuracy of finite difference schemes for time dependent PDEs by optimally 
+choosing the time step (and possibly adding a few correction terms).  
+Optimal time step (OTS) selection is a simple method based on the observation 
+that a carefully chosen time step can eliminate low-order terms in the 
+discretization error, allowing a formally low-order method to deliver 
+high-order accuracy.  It combines the well-known approach of using knowledge 
+of the structure of the PDE to reduce numerical error with the less-common 
+notion that an optimal choice of numerical parameters can boost the 
+effectiveness of algorithmically simple and computationally low-cost numerical 
+schemes.
+
+As a technique for enhancing numerical methods, optimal time step selection 
+has several desirable features.  Like any technique that leads to high-order 
+methods, optimal time step selection produces schemes that greatly reduce the 
+computational cost required to obtain a numerical solution of a desired 
+accuracy.  However, its real power comes from the fact that it achieves
+high-order accuracy while being based solely on very simple, formally 
+low-order finite difference schemes.  In other words, OTS selection allows us
+to extract as much accuracy as possible from an existing numerical scheme
+with minimal additional work.  For instance, we will see how optimal time step 
+selection makes it possible to achieve 4th-order accuracy for the diffusion 
+equation using only a second-order stencil for the Laplacian and forward 
+Euler time integration.  Moreover, high-order accuracy is not restricted to 
+problems on simple, rectangular domains; irregular domains are straightforward 
+to handle by appropriately setting the values in ghost 
+cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
+Optimal time step selection is even beneficial when 
+using finite difference schemes to solve some nonlinear PDEs.  While there 
+are certainly limitations to optimal time step selection, we will see that 
+there are many problems where it is useful.
 
@@
 time step selection and traditional adaptive time stepping 
-techniques~\cite{iserles_book, shampine_2005} that are used in the 
-context of the method 
-of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
-selection may appear to be just a crude version of adaptive time step control.
-However, the two methods are actually very different.  While both methods seek 
-to reduce numerical error by controlling the time step, the numerical errors 
-they control are fundamentally different.  Adaptive time stepping can only 
-be used to reduce temporal discretization errors because it controls the 
-errors that arise when solving the coupled system of ODEs for the 
+techniques~\cite{iserles_book,shampine_2005} that are used in the 
+context of the method of lines~\cite{iserles_book,gko_book}.  
+While both methods reduce numerical error by controlling the time step, the 
+numerical errors they control are fundamentally different.  Adaptive time 
+stepping can only be used to reduce temporal discretization errors because it 
+controls the errors that arise when solving the coupled system of ODEs for the 
 semi-discrete approximation to the PDE; the spatial discretization errors 
@@
 \emph{both} the spatial and temporal discretization errors because it
-uses information from the PDE to choose the time step.  Thus, optimal time 
-step selection can potentially produce superior results compared with 
-traditional adaptive time stepping approaches for integrating the 
-semi-discrete equations associated with time dependent PDEs.  Another 
-important distinction between the two techniques is that optimal time step 
-selection uses a \emph{fixed} time step which makes it significantly less 
+uses information from the PDE to choose the time step.  Another important 
+distinction between the two techniques is that optimal time step selection 
+uses a \emph{fixed} time step which makes it significantly less 
 computationally complex than adaptive time stepping methods.
-
-As a technique for constructing numerical methods, optimal time step selection 
-has several desirable features.  Like any technique that leads to high-order 
-methods, optimal time step selection produces schemes that greatly reduce the 
-computational cost required to obtain a numerical solution of a desired 
-accuracy.  However, its real power comes from the fact that it achieves
-high-order accuracy while being based solely on very simple finite difference 
-schemes.  For instance, we will see how optimal time step selection makes 
-it possible to achieve 4th-order accuracy for the diffusion equation using only 
-a second-order stencil for the Laplacian and forward Euler time integration.  
-Moreover, high-order accuracy is not restricted to problems on simple, 
-rectangular domains; irregular domains are straightforward to handle by
-appropriately setting the values in ghost cells~\cite{gibou_2005, fedkiw_1999, 
-osher_fedkiw_book}.  Optimal time step selection is even beneficial when 
-using finite difference schemes to solve some nonlinear PDEs.  While there 
-are certainly limitations to optimal time step selection, we will see that 
-there are many problems where it is useful.
 
@@
 for formally low-order numerical schemes.  In 
-Section~\ref{sec:applications_1d}, we demonstrate the general utility of 
+Section~\ref{sec:applications_1d}, we demonstrate the utility of 
 optimal time selection to problems in one space dimension by applying it to 
 finite difference schemes for several classical linear and nonlinear PDEs. 
-In Section~\ref{sec:applications_multidim}, we use optimal time step selection 
-to improve the accuracy of simple finite difference schemes for the linear 
-advection equation and diffusion equation in multiple space dimensions.  
-To illustrate the ease with which high-order accuracy can be achieved for
-problems on arbitrary domains, the 2D diffusion equation is solved on both 
-the regular and irregular domains.  
+In Section~\ref{sec:applications_multidim}, we show that optimal time step
+selection is useful for problems in multiple space dimensions by applying it
+the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
+the ease with which high-order accuracy can be achieved for problems on 
+arbitrary domains, the 2D diffusion equation is solved on both regular 
+and irregular domains.  
 For each PDE in Sections~\ref{sec:applications_1d} and 
@@
 to calculate the optimal time step size and order of accuracy.  
-Finally, in Section~\ref{sec:summary}, we provide some concluding remarks.
+Finally, in Section~\ref{sec:summary}, we end some concluding remarks.
 
@@
 eliminate low-order numerical errors.  The two basic ideas underlying 
-optimal time step selection are (1) the PDE provides valuable insight into the 
-discretization errors for finite difference schemes and (2) a judicious choice 
-of time step can be used to eliminate the leading-order (and possibly 
-higher-order) terms in the error.  In this section, we present and analyze 
-the technique of optimal time step selection for finite difference schemes 
-constructed to solve evolution equations of the form 
+optimal time step selection are 
+(1) a judicious choice of time step can be used to eliminate the leading-order 
+(and possibly higher-order) terms in the error
+and
+(2) the PDE provides valuable insight into the discretization errors for 
+finite difference schemes.
+In this section, we present and analyze the technique of optimal time step 
+selection for finite difference schemes constructed to solve scalar evolution 
+equations of the form 
 \beq
@@
 spatial derivatives, and the independent variable $x$.
-Because this class of time dependent PDEs is so prevalent and important in 
-applied mathematics, restricting our attention to problems of this form is 
-not a serious limitation.
+Because this class of time dependent PDEs is so prevalent and important, 
+restricting our attention to problems of this form is not a serious 
+limitation.
 
@@
 \eeq
-where $u$ is the solution to the PDE, $n$ is the spatial order of the 
-PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
+where $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
 is well-posed.  Because we will be boosting the accuracy of the finite 
@@
 explicitly keeping track of the leading-order terms.  In contrast to 
-conventional error analysis, we do not sweep all of the errors under ``big-O'' 
+conventional error analysis, we do not sweep all of the errors under big O
 notation.  Rather, we take advantage of the fact that derivation of the 
@@
 the error is proportional to a partial derivative of the solution, we can 
-attempt to use the PDE to relate the leading-order terms to each other.  If we 
-are fortunate (which we are for many finite difference schemes), this careful 
+use the PDE to relate the leading-order terms to each other.  
+If we have chosen an appropriate finite difference approximation for 
+(\ref{eq:model_PDE}), which is relatively straightforward to do, this careful 
 analysis yields a direct relationship between the leading-order spatial and 
@@
 \emph{optimal time step} for the finite difference scheme.  Using the optimal
-time step for time stepping yields a numerical solution with a boosted order 
-of accuracy for a given finite difference scheme.
+time step for time stepping yields a numerical solution with an order 
+of accuracy that is \emph{higher} than the formal order of the original finite 
+difference scheme.
 
@@
 discretization error of order $r$, the error in (\ref{eq:method_of_lines}) is 
-$O(\dx^r)$.  Therefore, the error accumulated during each time step due to 
-spatial discretization error is $O(\dt \dx^r)$ regardless of how the system of 
-ODEs (\ref{eq:method_of_lines}) is solved numerically.  Because the method of 
+$O(\dx^r)$, so the error accumulated during each time step due to 
+spatial discretization error is $O(\dt \dx^r)$.  Because the method of 
 lines is so frequently used to construct finite difference approximations for 
 time dependent PDEs, we will focus solely on finite difference schemes of this 
-type for the remainder of the paper to keep the discussion simple.  It is 
-straightforward to extend the results we obtain to more general finite 
-difference schemes.
+type for the remainder of the paper.  It is straightforward to extend the 
+results we obtain to more general finite difference schemes.
 
@@
 for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
-higher-order terms in the discretization error following the conventional 
-approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
-discretization error leaves a local error of the form 
+higher-order terms in the discretization error.  Elimination of the leading 
+terms in the discretization error leaves a local error of the form 
 $O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
 of the spatial and temporal discretization errors, respectively, after the 
-leading-order terms in the error have been eliminated.  As for the 
-leading-order error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of 
-the local error is a direct consequence of the method of lines approach to 
-discretizing the PDE.
-
+leading-order terms in the error have been eliminated.  
 From the local error, we can compute the global error of the numerical 
-solution over extended intervals of time by using the heuristic argument
-that the global error is equal to the local error divided by $\dt$.  Using 
-this heuristic argument, which is rigorously justified in many texts on 
-numerical methods such as~\cite{gko_book}, we find that the global error of 
-the finite difference scheme using an optimal time step is 
-\beq
-O(\dx^p) + O(\dt^q).
+solution over extended intervals of time by using the heuristic 
+argument that the global error is equal to the local error divided by 
+$\dt$~\cite{gko_book}:
+\beq
+e = O(\dx^p) + O(\dt^q).
 \label{eq:global_error_ots}
 \eeq
-Observe that the global error for the finite difference scheme when an 
-arbitrary time step is used has exactly the same form 
-as~(\ref{eq:global_error_ots}) with $p$ and $q$ replaced by $r$ and $s$, 
-respectively.  Thus, because $p > r$ and $q > s$, we see that using an optimal 
-time step boosts the overall accuracy of the finite difference scheme.
+Thus, because $p > r$ and $q > s$, we see that using an optimal time step 
+boosts the overall accuracy of the finite difference scheme.
 
@@
 one parameter, $\dx$, that controls the numerical accuracy of the finite 
-difference scheme.  This important point will be discuss in further detail in 
-the following section.
-
-\subsubsection*{Formal vs. Practical Accuracy}
-For time dependent PDEs, evaluating the order of accuracy for a numerical
-method is subtle because there are formally two separate orders of accuracy 
-to consider\footnote{The situation is more straightforward for 
-time-independent PDEs because there is usually only one order of accuracy to 
-be concerned with.}:  temporal and spatial.  While it is theoretically
-interesting and important to understand how the error depends on both the 
-grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
-of a numerical scheme is always controlled by only one of the two numerical
-parameters.
-
-Even though spatial and temporal orders of accuracy for a numerical method
-are formally separate, one will always dominate for a given choice of 
-$\dx$ and $\dt$.  For example, when solving the 1D diffusion equation 
-using the backward Euler method for time integration with a second-order 
-central difference stencil for the Laplacian, formal analysis shows that 
-the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
-stability constraints on the numerical parameters, the time
-step and grid spacing are free to vary independently.  In this situation, the 
-practical error for the method depends on the relative sizes of the time step 
-and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
-$O(\dt)$ which means that the error in the numerical solution is 
-primarily controlled by the time step.  Similarly, when 
-$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
-the error is controlled by the grid spacing.  Finally, when 
-$\dt  = O(\dx^2)$, the practical error is 
-$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
-primarily controlled by one of the two numerical parameters, and varying
-the subdominant parameter while holding the dominant parameter fixed does 
-not significantly affect the error.
-
-When there are constraints on the numerical parameters, there is less freedom 
-in choosing the controlling parameter.  For instance, if we solve the 
-1D diffusion equation using a forward Euler time integration scheme with a 
-second-order central difference stencil for the Laplacian, stability
-considerations require that we choose $\dt = O(\dx^2)$.  
-Combining this stability constraint with the formal error for the scheme
-shows that the practical error is 
-$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
-method is completely controlled by the grid spacing; the temporal error can 
-never dominate the spatial error.
-
-Numerical methods for time dependent PDEs are commonly designed so that 
-the global error is primarily controlled by the grid spacing.  The numerical
-errors introduced by the time integration scheme are typically \emph{forced}
-to be subdominant to the spatial discretization errors manually or as a result
-of stability constraints.  In other words, the operational procedure used 
-when developing numerical methods is usually to first choose a spatial 
-discretization scheme and grid spacing so that a desired level of accuracy 
-is achieved and then select the time integration scheme and time step in 
-such a manner that the temporal errors do not overcome the spatial errors.  
-This methodology explains the tendency to evaluate numerical schemes for
-time dependent PDEs predominantly on their order of accuracy with respect 
-to the grid spacing.
-
-Optimal time step selection fits into this general framework for controlling
-the error of numerical methods for time dependent PDEs.  Because the optimal 
-time step is a function of the grid spacing, the technique of optimal time 
-step selection places an even more restrictive constraint on $\dt$ than 
-stability considerations.  Thus, like numerical methods with stability 
-constraints, the accuracy of a finite difference scheme using an optimal time 
-step is completely controlled by the grid spacing.  For the method of lines, 
-we know that $\dto = O(\dx^{r/s})$, so the global error derived in the 
-previous section reduces to 
-\beq
-O \left( \dx^{\min(p,rq/s)} \right).
+difference scheme (see Appendix~\ref{app:formal_vs_practical_accuracy} for 
+further discussion about this important point).  Since $\dto = O(\dx^{r/s})$, 
+the global error (\ref{eq:global_error_ots}) reduces to
+\beq
+e = O \left( \dx^{\min(p,rq/s)} \right).
 \label{eq:global_error_ots_simplified}
@@
 Thus far, we have not considered the issue of numerical stability of finite
-difference schemes when the time step is set to $\dto$.  For explict time 
-integration schemes, OTS selection loses its value when the optimal time step 
+difference schemes when the time step is set to $\dto$.  For explicit time 
+integration schemes, OTS selection is ineffective when the optimal time step 
 is greater than the smallest stable time step.  In this situation, we can 
-usually switch to an implicit (or semi-implicit) time integration scheme with 
-the same order of accuracy for which the optimal time step is stable.  Note 
-that a new optimal time step may need to be calculated for the modified finite
-difference sceme.
-
-% TABLE SHOWING COMPUTATIONAL COSTS DISCUSSED IN SECTION
-% sec:computational_performance.  PLACED HERE TO SPREAD TABLES OUT
-\begin{table}[!b]
-\caption{\label{tab:comp_perf_vs_err} 
+switch to an implicit or semi-implicit time integration scheme with the 
+same order of accuracy for which the optimal time step is stable.  
+While the need to shift to implicit or semi-implicit time integration results 
+in a more complicated finite difference scheme, OTS selection is still 
+beneficial because it allows us to extract more accuracy from the simplest, 
+low-order implicit or semi-implicit methods.
+Note that a new optimal time step may need to be calculated for the modified 
+finite difference scheme.
+
+\subsubsection*{\label{sec:computational_performance} 
+                Computational Performance}
+For many problems, the optimal time step is on the order of the maximum 
+time step allowed by the stability constraint for an explicit method.  
+As a result, using the optimal time step to advance the finite
+difference scheme in time may appear to be overly restrictive and lead to poor 
+computational performance.  However, this apparent drawback is more than 
+compensated by the boost in the order of accuracy.  For example, when solving 
+the 1D diffusion equation using a forward Euler time integration scheme with a 
+second-order central difference stencil for the Laplacian, we obtain a 
+fourth-order scheme when $\dt$ is set equal to the optimal time step 
+(discussed further in Sections~\ref{sec:diffusion_eqn_1d} and
+\ref{sec:diffusion_eqn_2d}).  
+As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
+scheme that is computationally cheaper than traditional methods that use 
+the same low-order finite difference stencils.  The performance gain for 
+problems in higher spatial dimensions is even more impressive.  As we can see 
+in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
+simple forward Euler scheme with the optimal time step and the Crank-Nicholson
+method grows with the spatial dimension of the problem.
+
+\begin{table}[tb]
+\caption{
 Computational cost of various finite difference schemes for the 1D 
@@
 For all of the schemes, the standard second-order central difference 
-   stencil is used to discretize the Laplacian, and the numerical 
-   solution is assumed to be computed over a fixed time interval.  
+stencil is used to discretize the Laplacian.  
+Note that the time step for the backward Euler scheme is chosen so that the
+temporal error is subdominant to the spatial error and that the computation 
+time for the Crank-Nicholson method assumes an $O(N) = O \left( 1/\dx\right)$ 
+matrix inversion algorithm. 
 }
-\begin{minipage}{\textwidth}
-\begin{center} \footnotesize
+\label{tab:comp_perf_vs_err} 
 \renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|l|c|c|c|c|}
+\begin{tabular}{lcccc}
   \hline
   {\bf Numerical Scheme} & $\dx$ 
-  & $\dt$\footnote{The 
-   time step for the forward/backward Euler and Crank-Nicholson schemes 
-   are $O(\dx^2)$ and $O(\dx)$, respectively.  Note that the 
-   time step for the backward Euler scheme is chosen so that the temporal 
-   error is subdominant to the spatial error.}
-  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
-    to be $O(1/\dx)$.} 
-  & {\bf Compute Time}\footnote{The computation time is approximated as 
-    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
-    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
-    arise for the implicit methods.}  \\
+  & $\dt$
+  & {\bf Memory}
+  & {\bf Compute Time}
+  \\
   \hline 
@@
 \end{tabular}
-\end{center}
-\end{minipage}
 \end{table}
 
-\subsubsection*{\label{sec:computational_performance} 
-                Computational Performance}
-For many problems, the optimal time step is on the order of the time step 
-allowed by the stability constraint for an explicit method.  
-As a result, using the optimal time step to advance the finite
-difference scheme in time may appear to be overly restrictive and lead to poor 
-computational performance.  However, these apparent drawbacks are more than 
-compensated by the boost in the order of accuracy.  For example, when solving 
-the 1D diffusion equation using a forward Euler time integration scheme with a 
-second-order central difference stencil for the Laplacian, we obtain a 
-fourth-order scheme when $\dt$ is set equal to the optimal time step.  As
-Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
-scheme that is computationally cheaper than traditional methods that use 
-the same low-order finite difference stencils.  The performance gain for 
-problems in higher spatial dimensions is even more impressive.  As we can see 
-in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
-simple forward Euler scheme with the optimal time step and the Crank-Nicholson
-method grows with the spatial dimension of the problem.
-
-\begin{table}[!t]
-\caption{\label{tab:comp_perf_vs_dim}
+\begin{table}[tb]
+\caption{
 Computational cost of the forward Euler scheme with optimal time step and 
@@
 For both schemes, second-order, isotropic stencils are used to discretize 
-   the Laplacian.  For a problem in $d$ space dimensions, the memory 
-   requirement and computation time for each scheme are estimated to be 
-   $O \left(1/\dx^d \right)$ and 
-   $O\left( \left(1/\dx^d \right) (1/\dt) \right)$, respectively, which 
-   assumes that we are using an efficient 
-   $O(N) = O \left( 1/\dx^d \right)$ algorithm to invert the matrices that 
-   arise for the Crank-Nicholson method.
+the Laplacian.  
+Note that the computation time for the Crank-Nicholson method assumes an 
+$O(N) = O \left( 1/\dx^d \right)$ matrix inversion algorithm 
+and that the optimal time step for the forward Euler scheme is $\dto = \dx^2/6$.
 }
-\begin{minipage}{\textwidth}
-\begin{center} \footnotesize
+\label{tab:comp_perf_vs_dim}
 \renewcommand{\arraystretch}{1.5}
-\begin{tabular}{|c|c|c|c|c|}
+\begin{tabular}{ccccc}
   \hline
-  & \multicolumn{2}{|c|}{{\bf Forward Euler with OTS}\footnote{For the forward 
-    Euler scheme, the optimal time step is $\dto = \dx^2/6$.}}
-  & \multicolumn{2}{|c|}{\bf Crank-Nicholson} \\
+  & \multicolumn{2}{c}{\bf Forward Euler with OTS} 
+  & \multicolumn{2}{c}{\bf Crank-Nicholson} \\
   \cline{2-3} \cline{4-5} 
@@
 \end{tabular}
-\end{center}
-\end{minipage}
 \end{table}
@@
             OTS Selection for General PDEs in One Space Dimension} 
-In the previous section, we analyzed the technique of optimal time step 
-selection in the context of a simple model PDE in one space dimension.  
-For more general PDEs, the approach of carefully choosing the time step to 
-eliminate leading order terms in the error continues to be useful, but it 
-needs to be augmented to achieve the same improvement in the order of accuracy.
-In addition to using an optimal time step, correction terms may need to be 
-added to the finite difference scheme.  In this section, we discuss these 
-correction terms in the context of general constant coefficient, linear PDEs 
-and semilinear PDEs.
+In the previous section, we analyzed optimal time step selection in the 
+context of a simple model PDE in one space dimension.  For more general PDEs, 
+OTS selection continues to be useful, but it needs to be augmented with 
+correction terms to achieve the same improvement in the order of accuracy.  
+In this section, we discuss these correction terms in the context of general 
+constant coefficient, linear PDEs and semilinear PDEs.
 
@@
 generally a sum of several terms:
-\bea
+\beq
   \dt\left[ (L u) (\alpha \dx^r - \beta \dt^s) 
@@
   \label{eq:leading_order_error_linear_PDE_MOL}
-\eea
+\eeq
 where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
@@
 in the choice of the optimal time step, and $G$ is a spatio-temporal 
-differential operator that acts on source term $f$.  
+differential operator that acts on the source term $f$.  
 Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
 operators because any that arise can be eliminated using the 
-PDE~(\ref{eq:linear_PDE}).  Also, there is no explicit spatial discretization 
-error of order $r$ that originates from the source term because it is not 
-differentiated with respect to space when the semi-discrete 
-equations~(\ref{eq:method_of_lines}) are derived from the PDE.
+PDE~(\ref{eq:linear_PDE}).
 
@@
 $\dt = \dto$ is not enough to completely eliminate the 
-leading-order error.  To eliminate the vestigial errors, we need to add a few 
-correction terms to the numerical scheme in exactly the same spirit used in 
-the design of the modified nine-point formula for the Poisson 
-equation~\cite{iserles_book}.  The required correction terms are precisely 
-those that cancel out the second, third, and fourth terms in 
-(\ref{eq:leading_order_error_linear_PDE_MOL}).  Fortunately, it is 
-straightforward to compute and incorporate these terms into the numerical 
-scheme.  At each time step, the second and third terms in 
-(\ref{eq:leading_order_error_linear_PDE_MOL}) can be approximated using
-finite differences applied to the solution $u$ at the current time.
+leading-order error.  To eliminate the vestigial errors, we need to add 
+correction terms to cancel out the second, third, and fourth terms in 
+(\ref{eq:leading_order_error_linear_PDE_MOL}).  At each time step, the second 
+and third terms in (\ref{eq:leading_order_error_linear_PDE_MOL}) can be 
+approximated using finite differences applied to the solution $u$ at the 
+current time.
 The last term can be computed using either a finite difference approximation 
-or calculating it analytically.  With the addition of these correction terms, 
-the leading-order discretizaton error is completely eliminated, and the
-global error can be determined using the analysis in 
+or a direct analytic calculation.  With the addition of these correction 
+terms, the global error can be determined using the analysis in 
 Section~\ref{sec:error_analysis}.
 
-It is important to point out that the finite difference stencils used to 
-approximate the vestigial errors must be chosen carefully.  Special attention 
-is required to ensure that the finite difference stencils are of sufficiently 
-high-order.  Otherwise, we will lose the accuracy gained from using an optimal 
-time step and adding the correction terms to the finite difference scheme.  
-Since the specific form of the correction terms are PDE-dependent, we defer 
-further discussion to Sections~\ref{sec:applications_1d} 
-and~\ref{sec:applications_multidim} where we apply optimal time step selection 
-to several specific PDEs.
+Some care must be exercised when choosing the finite difference 
+stencils for the correction terms.  While the order of accuracy will always
+be boosted, the finite difference stencils used for the correction terms
+can introduce errors into the numerical solution that will cause the order
+of accuracy to fall below the estimate given by 
+(\ref{eq:global_error_ots_simplified}).
 
 \subsubsection*{Semilinear PDEs}
-Semilinear PDEs are an important class of PDEs that have the form
+Optimal time step selection can also be applied to semilinear PDEs of the
+form
 \beq
@@
 of $u$ and $f(x,t)$ is a source term.  While the coefficient on the 
-leading-order spatial derivative is generally allowed to be function of 
+leading-order spatial derivative is generally allowed to be a function of 
 space and time, optimal time step selection can only be applied to PDEs where
@@
 difference scheme is given by
-\bea
+\beq
   \dt \left[ (L u) (\alpha \dx^r - \beta \dt^s) 
@@
   \label{eq:leading_order_error_semilinear_PDE_MOL}
-\eea
+\eeq
 where $L$, $u$, $\alpha$, $\beta,$ $r$, and $s$ are defined as in
@@
 form (\ref{eq:leading_order_error_semilinear_PDE_MOL}), the optimal time 
-step and correction terms can be computed by proceeding in the same manner 
-as for general constant coefficient, linear PDEs.
+step and correction terms can be derived by proceeding in the same manner 
+as in the previous section.
 
 \subsection{Boundary Conditions}
-In our analysis and discussion thus far, we have completely ignored the issue 
-of boundary conditions.  The choice of discretization for boundary conditions
-can have a significant impact on the accuracy of the numerical solution.  In 
-general, high-order discretizations of boundary conditions are required to 
-ensure that errors introduced at the domain boundary do not pollute the 
-numerical solution in the domain interior.  For Dirichlet boundaries, it
-is sufficient to impose boundary conditions by using boundary values that have 
-the same order of accuracy as the solution.  More effort is required to ensure 
-that Neumann boundary conditions do not reduce the order of accuracy of the 
-solution.  In general, Neumann boundary conditions require that the accuracy 
-of ghost cells slightly exceed the desired order of accuracy for the 
-solution.  A detailed discussion of methods for imposing boundary conditions 
-is beyond the scope of this paper.  We refer interested readers to the 
-thorough analysis presented in~\cite{gko_book}.  To avoid complicating the 
-discussion in Sections~\ref{sec:applications_1d} 
-and~\ref{sec:applications_multidim}, Neumann boundary conditions are imposed 
-by taking advantage of knowledge of the analytical solution.
+In general, high-order discretizations of boundary conditions are required 
+to ensure that errors introduced at the domain boundary do not pollute 
+the numerical solution in the domain interior.  For Dirichlet boundaries, 
+it is sufficient to impose boundary conditions by using boundary values 
+that have the same order of accuracy as the solution.  More effort is 
+required to ensure that Neumann boundary conditions do not reduce the order 
+of accuracy of the solution.  In general, Neumann boundary conditions require 
+that the accuracy of ghost cells slightly exceeds the desired order of 
+accuracy for the solution.  A detailed discussion of methods for imposing 
+boundary conditions is beyond the scope of this paper.  We refer interested 
+readers to the thorough analysis presented in~\cite{gko_book}.  To avoid 
+complicating the discussion in Sections~\ref{sec:applications_1d} 
+and~\ref{sec:applications_multidim}, Neumann boundary conditions for
+all examples are imposed by taking advantage of knowledge of the analytical 
+solution.
 
@@
             OTS Selection in Multiple Space Dimensions}
-Optimal time selection is applicable to PDEs in any number of space
-dimensions.  It is straightforward to generalize the approach outlined in 
-the previous sections to problems in multiple space dimensions.  While 
-problems in higher dimensions generally require a more careful choice of 
-the finite difference stencil used to approximate spatial derivatives,
-there are no fundamental barriers that prevent the use of optimal time step 
-selection.
-
-The geometry of the domain is an important feature of problems in more than 
-one space dimension because it can significantly impact the way boundary 
-conditions are imposed.  For problems on regular domains, boundary conditions 
-can usually be handled by simply using 1D boundary condition discretizations 
-in each coordinate direction.  Problems on irregular domains pose a greater 
-challenge because the location of the domain boundary may not be explicitly 
-specified and grid points near the boundary require special consideration.  
-Fortunately, because optimal time step selection is based on simple, localized 
-finite difference stencils, irregular domains do not create any serious 
-difficulties.  Assuming that a `ghost cell' or `ghost value' 
-approach~\cite{gibou_2005, fedkiw_1999, osher_fedkiw_book} is used to impose 
-boundary conditions, the construction and analysis of finite 
-difference schemes and calculation of optimal time steps is identical for 
-problems on both regular and irregular domains.  For problems on 
-irregular domains, the key issue is how values at ghost cell points should 
-be calculated.  Once this issue has been addressed, devising high-order 
-accurate methods for PDEs on irregular domains is simply a matter of applying 
-optimal time step selection to the finite difference scheme used for problems 
-on regular domains.
-
-Another important issue that arises in multiple space dimensions is the choice
+Optimal time step selection is applicable to PDEs in any number of space
+dimensions.  While problems in higher dimensions generally require a more 
+careful choice of the finite difference stencil used to approximate spatial 
+derivatives, there are no fundamental barriers that prevent the use of optimal 
+time step selection.  Moreover, by using a `ghost cell' or `ghost value' 
+method for imposing boundary 
+conditions~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book},
+optimal time step selection can even be directly applied to problems on 
+irregular domains.
+
+One important issue that arises in multiple space dimensions is the choice
 of grid spacing.  In general, the grid spacing need not be equal in all 
@@
 different coordinate directions may need to be carefully chosen for an 
-optimal time step to even exist.
+optimal time step to exist.  In other words, it may be necessary to optimally 
+choose both the time step \emph{and} the mesh in order to boost the order 
+of accuracy.
 
@@
 leading-order spatial derivative in the PDE be constant in time and space.  
-Without this critical assumption, the first term in the leading-order 
-discretization error would be of the form 
+Without this assumption, the first term in the leading-order discretization 
+error would be of the form 
 $(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
@@
 small change in the form of the error makes it impossible to derive a single 
-optimal time step.  In a sense, each grid point would possess a different 
-optimal time step that could also depend on the time.  For a similar reason, 
-optimal time step selection also fails for quasilinear and fully nonlinear 
-PDEs.  The issue for these types of PDEs is that the coefficients in 
-$P(\dx, \dt)$ become functions of the solution $u$.
-
-Another limitation of optimal time step selection is that it does not 
-generally apply to finite difference schemes for systems of PDEs.  The 
-key issue here is that each PDE might have its own optimal time step.  While 
-it would be possible to boost the accuracy of finite difference schemes for 
-systems of PDEs where the optimal time step is the same for all of the 
-individual PDEs in the system, it is not clear that OTS selection yields any 
-benefit in general.
+optimal time step.  Instead, each grid point would possess its own individual 
+optimal time step.  For a similar reason, optimal time step selection also 
+fails for quasilinear and fully nonlinear PDEs.  The issue for these types of 
+PDEs is that the coefficients in $P(\dx, \dt)$ become functions of the 
+solution $u$.
+
+Another limitation of optimal time step selection is that it does not,
+in its current incarnation, apply to finite difference schemes for general
+systems of PDEs.  The main issue here is that each PDE might have its 
+own optimal time step.  OTS is, however, certainly applicable for problems 
+where the optimal time step is the same for all of the individual PDEs in the 
+system.
 
@@
 utility in designing finite difference schemes to solve several classical 
-linear and semilinear PDEs where analytical solutions are available for 
-comparison with numerical solutions.
+linear and semilinear PDEs.  The examples in this section were selected 
+because they have analytical solutions that are useful to compare numerical 
+solutions against.  However, the utility of OTS selection is by no means 
+limited to these simple PDEs.
 
@@
 For our first application of optimal time step selection, we consider 
-finite difference schemes for the linear advection equation
-~\cite{leveque_book_1992, leveque_book_2002, gko_book}:
+a finite difference scheme for the linear advection equation
+~\cite{gko_book,leveque_book_1992,leveque_book_2002}:
 \beq
@@
 where $A$ is the flow speed, which we take to be positive for convenience.  
-For an initial condition of the form $u(x,0) = f(x)$, it is easy to show 
-that the analytical solution is $f(x-At)$.  While it is unlikely that one 
-would actually use a numerical method to solve~(\ref{eq:advection_eqn_1d}), 
-it provides an interesting example where optimal time step selection 
-dramatically improves the accuracy of the numerical solution.
-
-There are many finite difference schemes that can be used to 
-solve~(\ref{eq:advection_eqn_1d}).  Several examples are given 
-in~\cite{leveque_book_1992}.  Perhaps the simplest, stable method is a 
-forward Euler for time integration scheme that uses a first-order upwind 
-discretization for the spatial derivative:
+For an initial condition of the form $u(x,0) = f(x)$, the analytical solution 
+is $f(x-At)$.  While it is unlikely that one would actually use a numerical 
+method to solve~(\ref{eq:advection_eqn_1d}), it is an interesting example 
+where optimal time step selection dramatically improves the accuracy of the 
+numerical solution.
+
+Perhaps the simplest, stable finite difference scheme 
+for (\ref{eq:advection_eqn_1d}) uses forward Euler for time integration 
+scheme and a first-order upwind discretization for the spatial derivative:
 \beq
   u^{n+1}_j = u^{n}_j 
-  - A \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right),
+  - A \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right).
   \label{eq:advection_eqn_1d_FD_scheme}
 \eeq
-where superscripts and subscripts denote the time and space indices, 
-respectively.  It is straightforward to show that the method is formally 
-first-order in both space and time.
+This scheme is formally first-order in both space and time.
 
@@
 keep track of the leading-order terms in the discretization error.  However,
-for the advection equation, it is instructive to retain all of them.  
-Using Taylor series to expand the true solution of the PDE at time step
-$(n+1)$, we find that
+for the advection equation, it is valuable to retain all of them.  
+Using Taylor series expansions of the true solution of the PDE, we find that 
 \bea
-  \tu^{n+1}_j = \tu^{n}_j 
+  \tu^{n+1}_j &=& \tu^{n}_j 
   + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
        \frac{\partial^k \tu^n}{\pt^k} 
+  = \tu^{n}_j + \sum_{k=1}^\infty \frac{\left( -A \dt \right)^k}{k!} 
+       \frac{\partial^k \tu^n}{\px^k},
   \label{eq:advection_eqn_1d_time_err} 
-\eea
-where $\tu$ denotes the true solution of the PDE and 
-$\tu^n_j \equiv \tilde{u}(j\dx, n\dt)$.  Similarly, Taylor series expansion of 
-the upwind finite difference approximation to the spatial derivative yields
-\bea
-  \frac{\tu^{n}_j - \tu^{n}_{j-1}}{\dx} = 
-  \frac{\partial \tu}{\px} 
+  \\
+  \frac{\tu^{n}_j - \tu^{n}_{j-1}}{\dx} &=& 
+  \frac{\partial \tu^n}{\px} 
   - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
@@
 \eea
-In both of the above equations, note that we have let context implicitly 
-determine the time ($t = t_n = n\dt$) and location ($x = x_j = j\dx$) where 
-partial derivatives of $\tu$ should be evaluated.  For convenience, we shall 
-continue to follow this convention throughout the remainder of the paper.
-
-Next, we use the PDE~(\ref{eq:advection_eqn_1d_FD_scheme}) to replace all of 
-the time derivatives in~(\ref{eq:advection_eqn_1d_time_err}) with spatial
-derivatives to obtain
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + \sum_{k=1}^\infty \frac{\left( -A \dt \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k}.
-  \label{eq:advection_eqn_1d_time_err_modified}
-\eea
-Finally, by combining~(\ref{eq:advection_eqn_1d_time_err_modified}) 
+where $\tu$ denotes the true solution of the PDE, all derivatives are 
+evaluated at $x_j$, and we have used the 
+PDE~(\ref{eq:advection_eqn_1d_FD_scheme}) to replace all of the time 
+derivatives with spatial derivatives in (\ref{eq:advection_eqn_1d_time_err}).
+For convenience, we will continue to suppress the subscripts on spatial
+derivatives throughout the paper; the location where they should be 
+evaluated will be apparent from the context.
+
+Combining~(\ref{eq:advection_eqn_1d_time_err}) 
 and~(\ref{eq:advection_eqn_1d_space_err}) with the finite difference 
-scheme~(\ref{eq:advection_eqn_1d_FD_scheme}), we obtain the evolution
+scheme~(\ref{eq:advection_eqn_1d_FD_scheme}) yields the evolution
 equation for the error $e \equiv u - \tu$:
-\bea
+\beq
   e^{n+1}_j = e^{n}_j 
-    &-& A \dt \left( \frac{e^{n}_j - e^{n}_{j-1}}{\dx} \right) \nonumber \\
-    &+& A \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
+    - A \dt \left( \frac{e^{n}_j - e^{n}_{j-1}}{\dx} \right) 
+    + A \dt \sum_{k=2}^\infty \frac{\left( -1 \right)^k}{k!} 
         \frac{\partial^k \tu^n}{\px^k} 
@@
   \label{eq:advection_eqn_1d_err_eqn}
-\eea
-As expected, the error $e$ satisfies the same finite difference approximation 
-as the solution with a source term which is the discretization error for the 
-scheme.
-
-Extracting the leading-order term of the discretization error
+\eeq
+Therefore, the leading-order term of the discretization error is given
+by
 \beq
@@
 \eeq
-we find that the optimal time step is $\Delta t_{opt} = \Delta x / A$.
+which leads to an optimal time step of $\Delta t_{opt} = \Delta x / A$.
+
 For the linear advection equation, setting the time step equal to the optimal 
@@
 Assuming that the initial conditions are set exactly, there is absolutely no 
-error in the numerical solution for all times!  
-This surprising result, which is unique to the advection equation, is 
-well-known in the numerical methods community and is referred to as the unit 
-CFL condition~\cite{leveque_book_2002}.  It is usually proved by examining 
+error in the numerical solution for all times.
+This result, which is unique to the advection equation, is well-known in the 
+numerical analysis and scientific computing community as the unit CFL 
+condition~\cite{leveque_book_2002} and is usually proved by examining 
 how characteristic lines of the advection equation pass through grid points.  
-It is interesting to see that it also naturally arises through an analysis of 
-the discretization errors and optimal time step selection.
-
-\subsection{Diffusion Equation}
+It is interesting to see that it also naturally arises through optimal time 
+step selection.
+
+\subsection{Diffusion Equation \label{sec:diffusion_eqn_1d}}
 As an example where optimal time step selection yields a more typical boost 
-in the order of accuracy, we turn our attention to the diffusion equation:
-\beq
-  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2},
+in the order of accuracy, let us consider the diffusion equation:
+\beq
+  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2} + f(x,t),
   \label{eq:diffusion_eqn_1d}
 \eeq
-where $D$ is the diffusion constant.  As with the advection equation,
-there are many finite difference schemes that can be used to solve the 
-diffusion equation.  Keeping with our philosophy of using simple, low-order 
-methods to construct finite difference schemes, we apply optimal time step 
-selection to a scheme that uses forward Euler for time integration and the 
-standard second-order central difference approximation for the Laplacian:
+where $D$ is the diffusion constant and $f(x,t)$ is a source term.  
+We apply optimal time step selection to the simplest finite difference 
+scheme for (\ref{eq:diffusion_eqn_1d}):
 \beq
   u^{n+1}_j = u^{n}_j 
-  + D \dt 
-    \left( \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right),
+  + \dt 
+    \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right] 
+         + f^n_j \right),
   \label{eq:diffusion_eqn_1d_FD_scheme}
 \eeq
-Conventional error analysis shows that scheme is formally first-order in time 
-and second-order in space.  Due to the stability constraint 
-$\dt \le \dx^2/2D$, the scheme is $O(\dx^2)$ accurate overall.
-
-As before, we begin by analyzing the leading-order terms of the 
-discretization error for the finite difference scheme.  Employing Taylor
-series expansions, we find that the true solution satisfies
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + \dt \frac{\partial \tu^n}{\pt} 
-  + \frac{\dt^2}{2} \frac{\partial^2 \tu^n}{\pt^2} + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_time_err}
-\eea
-and the central difference for the Laplacian satisfies
-\bea
-  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
-  \frac{\partial^2 \tu^n}{\px^2} 
-  + \frac{\dx^2}{12} \frac{\partial^4 \tu^n}{\px^4} 
-  + O(\dx^4)
-  \label{eq:diffusion_eqn_1d_space_err}
-\eea
-Making use of the PDE~(\ref{eq:diffusion_eqn_1d}), 
-(\ref{eq:diffusion_eqn_1d_time_err}) becomes
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  + D \dt \frac{\partial^2 \tu^n}{\px^2} 
-  + \frac{(D \dt)^2}{2} \frac{\partial^4 \tu^n}{\px^4} 
-  + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_time_err_modified}
-\eea
-Combining these results with the finite difference 
-scheme~(\ref{eq:diffusion_eqn_1d_FD_scheme}), we arrive at the evolution 
-equation for the error $e \equiv u - \tu$:
-\bea
-  e^{n+1}_j = e^{n}_j 
-  &+& D \dt 
-    \left( \frac{e^{n}_{j+1} -2 e^{n}_j + e^{n}_{j-1}}{\dx^2} \right)
-  \nonumber \\
-  &+& \frac{\partial^4 \tu^n}{\px^4} 
-      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-      + O(\dt \dx^4) + O(\dt^3)
-  \label{eq:diffusion_eqn_1d_err_eqn}
-\eea
-Again, we see that the error satisfies the same evolution equation
-as the numerical solution and that the discretization error acts as a 
-source term.
-
-From~(\ref{eq:diffusion_eqn_1d_err_eqn}), we see that the optimal time step 
-is given by $\dto = \dx^2/6D$.  Using~(\ref{eq:global_error_ots}) or
-(\ref{eq:global_error_ots_simplified}), the order of the scheme with this 
-choice of time step is $O(\dx^4)$; we have boosted the accuracy of original 
-scheme from second to fourth order.  Note that $\dto$ is stable, so there
-is no need to switch to an implicit time integration scheme.
+which uses forward Euler for time integration and the standard second-order 
+central difference approximation for the Laplacian.
+This scheme is formally first-order in time and second-order in space.  
+The stability constraint $\dt \le \dx^2/2D$ implies that the scheme is 
+$O(\dx^2)$ accurate overall.
 
@@
 with optimal time step (circles), forward Euler with suboptimal time step
-$\dt = 2\dx^2/3D$ (squares), and Crank-Nicholson with time step 
-$\dt = \dx/16 D$ (diamonds).  All three schemes use the standard second-order, 
-central finite difference approximation for the Laplacian.  Using each method, 
-we solve the diffusion equation with $D = 0.5$ to time $t = 0.1$ on the 
-domain $0 < x < 1$ subject to the boundary conditions $u(0,t) = 1$ and 
-$u_x(1,t) = 1/2$ and the initial condition
-$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
-- 4 \sin(11 \pi x/2)$.  
-Note that the errors for the forward Euler with suboptimal time 
-step and Crank-Nicholson schemes lie almost directly on top of each other
-at the resolution of the figure.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
+(squares), and Crank-Nicholson (diamonds).  
+All three schemes use the standard second-order, central finite difference 
+approximation for the Laplacian.  
 }
@@
 
-To demonstrate the improvement in accuracy of the forward Euler scheme when 
-the optimal time step is used, we compare the numerical errors of three finite 
-difference schemes for the diffusion equation: forward Euler with optimal
-time step, forward Euler with a suboptimal time step, and Crank-Nicholson.  
-As we can see in Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward 
-Euler with optimal time step is fourth-order accurate while both forward Euler 
-with a suboptimal time step and Crank-Nicholson are only second-order 
-accurate.  Using the forward Euler scheme with an optimal time step has a clear 
-advantage over both of the other schemes.  Even with as few as 25 grid 
-points, the error in solution computed by the forward Euler scheme with an 
-optimal time step is orders of magnitude smaller.
-
-As mentioned in Section~\ref{sec:computational_performance}, one of the 
-advantages of using optimal time step selection is the superior scaling of 
-the computational cost with respect to the numerical error.  It is worth 
-mentioning that another way to achieve the same computational performance
-would be to use a Crank-Nicholson time integration scheme with a fourth-order 
-discretization of the Laplacian~\cite{gibou_2005}.  By using an 
-$O\left(\dx^2 \right)$ time step and a linear solver that scales linearly 
-with the size of the matrix, we achieve an overall fourth-order accurate 
-method that has the same computational complexity as the forward Euler scheme
-described above with an optimal time step.  However, the forward Euler scheme 
-with optimal time step will still outperform the Crank-Nicholson scheme 
-because the latter requires the use of a wider stencil and the solution of 
-at least one linear system of equation for each time step.
-
-\subsubsection*{Diffusion Equation with Source Term} 
-As discussed in Section~(\ref{sec:ots_linear_pde}), the procedure for optimal 
-time step selection needs to be slightly modified when source terms are 
-present.  If we add a source term, $f(x,t)$, to the right-hand side 
-of~(\ref{eq:diffusion_eqn_1d}), the finite difference 
-scheme~(\ref{eq:diffusion_eqn_1d_FD_scheme}) picks up an additional term
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + \dt 
-    \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right] 
-         + f^n_j \right).
-  \label{eq:diffusion_eqn_1d_src_FD_scheme}
-\eeq
-and~(\ref{eq:diffusion_eqn_1d_time_err_modified}) becomes
+As before, we begin by analyzing the leading-order terms of the 
+discretization error for the finite difference scheme.  Employing Taylor
+series expansions and the PDE (\ref{eq:diffusion_eqn_1d}),
+we find that the true solution satisfies
 \bea
@@
   + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_src_time_err_modified}
+  \label{eq:diffusion_eqn_1d_time_err} 
 \eea
-Therefore, the leading-order discretization error is given by
+and that the central difference approximation for the Laplacian satisfies
+\beq
+  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
+  \frac{\partial^2 \tu^n}{\px^2} 
+  + \frac{\dx^2}{12} \frac{\partial^4 \tu^n}{\px^4} 
+  + O(\dx^4)
+  \label{eq:diffusion_eqn_1d_space_err}
+\eeq
+Therefore, the truncation error for (\ref{eq:diffusion_eqn_1d_FD_scheme})
+is given by
 \beq
@@
     - \frac{\dt^2}{2} \left( D \frac{\partial^2 f}{\px^2} 
+                           + \frac{\partial f}{\pt} \right)
+      + O(\dt \dx^4) + O(\dt^3).
+  \label{eq:diffusion_eqn_1d_trunc_err}
+\eeq
+From this expression, it is easy to see that the optimal time step is 
+$\dto = \dx^2/6D$ and that the correction term is
+\beq
+\frac{\dt^2}{2} \left( D \frac{\partial^2 f}{\px^2} 
                      + \frac{\partial f}{\pt} \right).
-  \label{eq:diffusion_eqn_1d_src_err}
-\eeq
-To eliminate the low-order errors introduced by the source term, we 
-simply add the following correction term to the right-hand side of 
-the finite difference scheme~(\ref{eq:diffusion_eqn_1d_src_FD_scheme}):
-\beq
-\frac{\dt^2}{2} \left( D \left( f_{xx} \right)^n_j 
-  + \left( f_t \right)^n_j \right),
-\eeq
-where the derivatives of $f$ may be evaluated analytically or using 
-sufficiently high-order finite difference stencils.
+\eeq
+From (\ref{eq:global_error_ots_simplified}), we see that using the optimal 
+time step and adding this simple correction term to 
+(\ref{eq:diffusion_eqn_1d_FD_scheme}) boosts the accuracy of original scheme 
+from second to fourth order.  Note that to achieve fourth order accuracy, 
+the spatial derivatives in the correction term must be calculated using a
+finite difference approximation that is at least second-order accurate; 
+otherwise, the truncation error will be $O(\dt^2 \dx)$ which implies an 
+$O(\dt \dx) = O(\dx^3)$ global error.
+
+To demonstrate the improvement in accuracy of the forward Euler scheme when 
+the optimal time step is used, we compare the numerical errors of three finite 
+difference schemes for the diffusion equation: forward Euler with optimal
+time step and correction term, forward Euler with a suboptimal time step, and 
+Crank-Nicholson.  As we can see in 
+Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward Euler with optimal 
+time step is fourth-order accurate while both forward Euler with a suboptimal 
+time step and Crank-Nicholson are only second-order accurate.  
+Using the forward Euler scheme with an optimal time step and correction term 
+has a clear advantage over the other finite difference schemes.  Even with as 
+few as 25 grid points, the error in solution computed by the forward Euler 
+scheme with an optimal time step and correction term is orders of magnitude 
+smaller.
+
+The importance of the correction term is illustrated in 
+Figure~\ref{fig:diffusion_eqn_1d_src_error}, which compares the accuracy of 
+the OTS forward Euler scheme with and without the correction term.  As we 
+can see, OTS forward Euler with correction term is fourth-order accurate 
+while OTS forward Euler without correction term is only second-order 
+accurate.
 
@@
 two finite difference schemes that solve the diffusion equation: 
-forward Euler with optimal time step and correction terms (circles) and 
-forward Euler with optimal time step but \emph{without} correction terms 
+forward Euler with optimal time step and correction term (circles) and 
+forward Euler with optimal time step \emph{without} correction term 
 (triangles).  Both schemes use the standard second-order, central finite 
 difference approximation for the Laplacian.
-Using each method, we solve the diffusion equation with $D = 0.5$ and
-source term 
-$f(x) = 5 \sin(3 \pi x/2) - 7 \sin(15 \pi x/2) + 10 \sin(21 \pi x/2)$
-to time $t = 0.5$ on the domain $0 < x < 1$ subject to the boundary 
-conditions $u(0,t) = 1$ and $u_x(1,t) = 1/2$ 
-and the initial condition
-$u(x,0) = 1 + x/2 + 2 \sin(5 \pi x/2) + 3 \sin(7 \pi x/2) 
-- 4 \sin(11 \pi x/2)$.  
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
 }
@@
 
-In Figure~\ref{fig:diffusion_eqn_1d_src_error}, we compare the accuracy of 
-the OTS forward Euler scheme with and without the correction terms.  As we can 
-see, OTS forward Euler with correction terms is fourth-order accurate while 
-OTS forward Euler without correction terms is only second-order accurate.  
-These results clearly demonstrate the importance of the correction terms for
-achieving high-order accuracy.
-
-\subsection{Advection-Diffusion Equation}
+As mentioned in Section~\ref{sec:computational_performance}, one of the 
+advantages of using optimal time step selection is the superior scaling of 
+the computational cost with respect to the numerical error.  
+Another way to achieve computational performance with the same scaling would 
+be to use a Crank-Nicholson time integration scheme with a fourth-order 
+discretization of the Laplacian~\cite{gibou_2005}.  By using an 
+$O\left(\dx^2 \right)$ time step and a linear solver that scales linearly 
+with the size of the matrix, we achieve an overall fourth-order accurate 
+method that has the same computational complexity as the forward Euler scheme
+described above with an optimal time step.  However, the forward Euler scheme 
+with optimal time step will still outperform the Crank-Nicholson scheme 
+because the latter requires the use of a wider stencil and the solution of 
+at least one linear system of equation for each time step.  Round-off error
+associated with matrix inversion may also be an issue for the Crank-Nicholson 
+method.
+
+\subsection{Viscous Burgers Equation}
 We now demonstrate the use of optimal time step selection in the context of 
-a general linear, constant coefficient PDE -- the advection-diffusion 
-equation:
-\beq
-  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2} 
-  + A \frac{\partial u}{\px},
-  \label{eq:ADE_1d}
-\eeq
-where $D$ is the diffusivity and $A$ is the flow speed.  For this problem,
-we use forward Euler for time integration and second-order central 
-difference approximations for the diffusion and advection terms:
+a well-known semilinear PDE -- the viscous Burgers 
+equation~\cite{whitham_book}:
+\beq
+  \frac{\partial u}{\pt} + u \frac{\partial u}{\px} = 
+     \nu \frac{\partial^2 u}{\px^2} 
+  \label{eq:burgers_1d}
+\eeq
+where $\nu$ is the viscosity.  For this equation, let us consider the finite
+difference scheme constructed using a forward Euler time integration scheme
+and second-order, central difference approximations for the diffusion and 
+nonlinear advection terms:
 \beq
@@
   + \dt 
-    \left( D \left[ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
-         + A \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] \right).
-  \label{eq:ADE_1d_FD_scheme}
+    \left( \nu 
+      \left [ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
+         - u^n_j 
+      \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] 
+    \right).
+  \label{eq:burgers_1d_FD_scheme}
 \eeq
 This scheme is formally first-order in time and second-order in space.  
-Due to the stability constraint $\dt < \min \{2D/A^2, \dx^2/2D \}$ 
-\cite{chan_1984}, the scheme is $O(\dx^2)$ accurate overall (in the limit 
-$\dx \rightarrow 0$).
+Using the stability condition for the advection-diffusion as a heuristic 
+guide\footnote{The stability constraint for the advection-diffusion equation
+is $\dt < \min \{2D/A^2, \dx^2/2D \}$~\cite{chan_1984}.}, 
+we expect a stability constraint of the form $\dt \le \dx^2/2 \nu$ in the 
+limit $\dx \rightarrow 0$ (assuming that the solution $u$ remains bounded).  
+Therefore, the scheme (\ref{eq:burgers_1d_FD_scheme}) is $O(\dx^2)$ accurate 
+overall.
 
 \begin{figure}[tb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \
-\scalebox{0.32}{\includegraphics{figures/adv_diff_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D advection-diffusion
-equation~(\ref{eq:ADE_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step $\dt = 2 \dx^2/3 D$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 1$ on the 
-domain $-10 < x < 10$ with the numerical solution (circles) computed using 
-50 grid points.  For details about the initial/boundary 
-conditions and physical parameters used to compute these solutions, see the
-caption for Figure~\ref{fig:ADE_1d_error}.
-}
-\label{fig:ADE_1d_solns}
-\end{center}
-\end{figure}
-
-To analyze the discretization error, we can reuse 
-results~(\ref{eq:diffusion_eqn_1d_time_err}) 
-and~(\ref{eq:diffusion_eqn_1d_space_err})
-from our analysis of the diffusion equation.  In addition, we need the 
-leading-order discretization error for the central difference approximation 
-of $u_x$, which is given by
-\bea
-  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
-  \frac{\partial \tu^n}{\px} 
-  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
-  + O(\dx^4).
-  \label{eq:ADE_1d_ux_err}
-\eea
-Using the advection-diffusion equation~(\ref{eq:ADE_1d})
-to convert time derivatives into spatial derivatives 
-in~(\ref{eq:diffusion_eqn_1d_time_err}), we find that 
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} 
-               + A \frac{\partial \tu^n}{\px} \right)
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
-      \left( 
-        D^2 \frac{\partial^4 \tu^n}{\px^4} 
-      + 2 D A \frac{\partial^3 \tu^n}{\px^3} 
-      + A^2 \frac{\partial^2 \tu^n}{\px^2} 
-      \right) 
-  + O \left( \dt^3 \right).
-  \label{eq:ADE_1d_time_err_modified}
-\eea
-The local truncation error $\tau^n_j$ for the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) can be obtained by the usual procedure 
-of deriving the evolution equation for the error and extracting the source 
-term:
-\bea
-  \tau^n_j &=&
-      \left( \frac{\partial^4 \tu^n}{\px^4} 
-           + \frac{2A}{D} \frac{\partial^3 \tu^n}{\px^3} \right)
-      \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-  \nonumber \\
-  &-& \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}
-      + O(\dt \dx^4) + O(\dt^3)
-  \label{eq:ADE_1d_err_eqn}
-\eea
-
-As we discussed in Section~\ref{sec:ots_linear_pde}, boosting the order
-of accuracy for general linear, constant coefficient PDEs requires both 
-the use of an optimal time step and the addition of correction terms.  
-In this case, the optimal time step is $\dto = \dx^2/6D$ and the
-correction term is 
-\beq
-  \frac{A^2 \dt^2}{2} \frac{\partial^2 \tu^n}{\px^2}.
-  \label{eq:ADE_1d_corr_term}
-\eeq 
-Note that using the standard second-order, central difference stencil to 
-compute the Laplacian in the correction term only makes an 
-$O(\dt^2 \dx^2)$ contribution to $\tau_j^n$, which ensures that no 
-errors larger than $O(\dt \dx^4) + O(\dt^3)$ are introduced into the 
-truncation error.
-With the optimal time step and correction term, the finite difference 
-scheme~(\ref{eq:ADE_1d_FD_scheme}) is now fourth-order accurate.  
-Figures~\ref{fig:ADE_1d_error} and \ref{fig:ADE_1d_solns} illustrate the 
-improved accuracy of the finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) 
-when the optimal time step and correction term are used.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/adv_diff_eqn_1d_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for 
-two finite difference schemes that solve the advection-diffusion 
-equation: forward Euler with optimal time step and correction term (circles) 
-and forward Euler with suboptimal time step $\dt = 2\dx^2/3D$ and no correction 
-terms (squares).  Both schemes use second-order, central finite difference 
-approximations for the advection and diffusion terms.
-Using both methods, we solve the advection-diffusion equation with $D = 2$
-and $A = 5$ to time $t = 1$ on the domain $-10 < x < 10$ subject to the 
-initial condition $u(x,0) = \exp(-x^2/4)$.  The boundary conditions are set 
-using the analytical solution 
-$u(x,t) = \frac{1}{\sqrt{T}} \exp \left( -(x+At)^2/4T \right)$
-where $T = Dt+1$.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:ADE_1d_error}
-\end{center}
-\end{figure}
-
-\subsubsection*{First-order Upwind Discretization of Advection Term}
-It is common to use upwind discretizations such 
-as~(\ref{eq:advection_eqn_1d_FD_scheme}) for advection terms that arise
-in PDEs.  If instead of a central difference approximation, we use a
-first-order upwind stencil\footnote{The appropriate upwind stencil depends
-on the sign of $A$.} for the advection term in~(\ref{eq:ADE_1d}), the 
-discretization error for the finite difference scheme is dominated by the 
-first-order error introduced by the upwind stencil for the advection term:
-\bea
-  \tau^n_j &=&
-      \frac{\partial^2 \tu^n}{\px^2} 
-      \left[ \frac{\dx}{2} - \frac{|A| \dt}{2}  \right] (|A| \dt)
-  \nonumber \\
-   & & 
-    - \frac{\dt^2}{2} 
-      \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
-           + 2 A D \frac{\partial^3 \tu^n}{\px^3}
-      \right)
-      + O(\dt \dx^2) + O(\dt^3)
-  \label{eq:ADE_1d_upwind_err_eqn}.
-\eea
-Thus, we can see that for a finite difference scheme based on a first-order
-upwind discretization of the advection term, the optimal time step is 
-$\dto = \dx/|A|$ (as for the pure advection equation) and the required
-correction term is
-\beq
-    \frac{\dt^2}{2} 
-      \left( D^2 \frac{\partial^4 \tu^n}{\px^4}  
-           + 2 A D \frac{\partial^3 \tu^n}{\px^3}
-      \right)
-  \label{eq:ADE_1d_upwind_corr_term}.
-\eeq
-This scheme is second-order overall as long as the finite difference 
-approximations to all of the derivatives in~(\ref{eq:ADE_1d_upwind_corr_term})
-are at least first-order accurate.  Unfortunately, the optimal time step 
-is unstable for the finite difference scheme that uses forward Euler time 
-integration and a first-order upwind stencil for the advection term.  
-In order to use optimal time step selection, we would need to replace the 
-forward Euler time integration scheme with a backward Euler scheme.
-
-A comparison of the finite difference scheme analyzed in this section to the 
-scheme described in the previous section makes it clear that the power 
-of optimal time step selection is highly dependent on the original choice of 
-finite difference scheme.  The first-order upwind scheme not only has a 
-lower overall accuracy when the optimal time step is used, it requires the 
-use of an implicit time integration method and construction of finite 
-difference stencils for higher-order spatial derivatives in order to compute
-the correction term~(\ref{eq:ADE_1d_upwind_corr_term}).  That the first-order
-scheme possesses all of these weaknesses underscores the importance of 
-carefully choosing the finite difference scheme even when optimal time step 
-selection is used.
-
-\subsection{Viscous Burgers Equation}
-\begin{figure}[!b]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \ 
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D viscous Burgers 
-equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
-optimal time step and correction terms (left) and forward Euler time 
-stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 2$ on the 
-domain $0 < x < 10$ with the numerical solution (circles) computed using 
-$100$ grid points.  
-For details about the initial/boundary conditions and physical parameters 
-used to compute these solutions, see the caption for 
-Figure~\ref{fig:burgers_1d_error}. 
-}
-\label{fig:burgers_1d_solns}
-\end{center}
-\end{figure}
-
-\begin{figure}[!t]
 \begin{center}
@@
 and forward Euler with suboptimal time step $\dt = \dx^2/4\nu$ and no 
-correction terms (squares).
+correction term (squares).
 Both schemes use second-order, central finite difference approximations for
@@
 where $T = t+1$.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
 }
@@
 
-For an example of optimal time step selection applied to a semilinear equation,
-we consider the viscous Burgers equation~\cite{whitham_book}:
-\beq
-  \frac{\partial u}{\pt} + u \frac{\partial u}{\px} = 
-     \nu \frac{\partial^2 u}{\px^2} 
-  \label{eq:burgers_1d}
-\eeq
-where $\nu$ is the viscosity.  Notice that this problem has the same structure
-to the advection-diffusion equation, except that the advection velocity now 
-depends on the solution $u$.  To solve this equation numerically,
-we will use forward Euler for time integration and second-order central 
-difference approximations for the diffusion and nonlinear terms:
-\beq
-  u^{n+1}_j = u^{n}_j 
-  + \dt 
-    \left( \nu 
-      \left [ \frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2} \right]
-         - u^n_j 
-      \left[ \frac{u^{n}_{j+1} - u^{n}_{j-1}}{2 \dx} \right] 
-    \right).
-  \label{eq:burgers_1d_FD_scheme}
-\eeq
-This scheme is formally first-order in time and second-order in space.  
-Using the stability condition for the advection-diffusion equation as a 
-heuristic guide, we expect a stability constraint of the form 
-$\dt \le \dx^2/2 \nu$ in the limit $\dx \rightarrow 0$ (assuming that 
-the solution $u$ remains bounded).  Therefore, the scheme 
-(\ref{eq:burgers_1d_FD_scheme}) is $O(\dx^2)$ accurate overall.
-
-Because~(\ref{eq:burgers_1d_FD_scheme}) is very similar to the second-order
-finite difference scheme~(\ref{eq:ADE_1d_FD_scheme}) for the 
-advection-diffusion equation, we can reuse many of the results obtained in 
-the previous section to derive the truncation error 
-for~(\ref{eq:burgers_1d_FD_scheme}).  Following a procedure analogous to the
-one we used to analyze~(\ref{eq:ADE_1d_FD_scheme}), we find that the local 
-truncation error for~(\ref{eq:burgers_1d_FD_scheme}) has the form
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
+\ \ \ \ \ 
+\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
+\caption{Comparison of numerical solutions for the 1D viscous Burgers 
+equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
+optimal time step and correction term (left) and forward Euler time 
+stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
+Both figures compare the analytical solution (solid line) at $t = 2$ on the 
+domain $0 < x < 10$ with the numerical solution (circles) computed using 
+$100$ grid points.  
+For information about the initial/boundary conditions and physical parameters 
+used to compute these solutions, see the caption for 
+Figure~\ref{fig:burgers_1d_error}. 
+}
+\label{fig:burgers_1d_solns}
+\end{center}
+\end{figure}
+
+To analyze the discretization error for (\ref{eq:burgers_1d_FD_scheme}), 
+we combine~(\ref{eq:diffusion_eqn_1d_space_err}) from our analysis of 
+the diffusion equation with the leading-order discretization error for 
+the central difference approximation of $u_x$,
+\beq
+  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
+  \frac{\partial \tu^n}{\px} 
+  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
+  + O(\dx^4),
+  \label{eq:burgers_1d_ux_err}
+\eeq
+and the Burgers equation analogue of (\ref{eq:diffusion_eqn_1d_time_err}),
+\bea
+  \tu^{n+1}_j = \tu^{n}_j 
+    &+& \dt \left( \nu \frac{\partial^2 \tu^n}{\px^2} 
+                 - \tu^{n}_j \frac{\partial \tu^n}{\px} \right)
+    \nonumber \\
+    &+& \frac{\dt^2}{2} 
+        \left( \nu^2 \frac{\partial^4 \tu^n}{\px^4} 
+             - 2 \nu \tu^n_j\frac{\partial^3 \tu^n}{\px^3} 
+             - 4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2} 
+             + 2 \tu^n_j \left(\frac{\partial \tu^n}{\px} \right)^2
+             + \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2} 
+        \right) 
+    \nonumber \\
+    &+& O \left( \dt^3 \right).
+    \label{eq:burgers_1d_time_err}
+\eea
+Together these imply that the leading order term in the local truncation error 
+for~(\ref{eq:burgers_1d_FD_scheme}) is given by
 \bea
@@
       \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
-           - 2 \tu \frac{\partial^3 \tu^n}{\px^3} \right)
+           - 2 \tu^n_j \frac{\partial^3 \tu^n}{\px^3} \right)
       \left[ \frac{\dx^2}{12} - \frac{\nu \dt}{2}  \right] \dt
@@
            4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
-         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
-         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
+         - 2 \tu^n_j \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
+         - \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2}
       \right)
-      + O(\dt \dx^4) + O(\dt^3).
   \label{eq:burgers_1d_err_eqn}
@@
 Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
-for the finite difference scheme is 
+is
 \beq
@@
            4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2}
-         - 2 \tu \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
-         - \tu^2 \frac{\partial^2 \tu^n}{\px^2}
+         - 2 \tu^n_j \left( \frac{\partial^2 \tu^n}{\px^2} \right)^2
+         - \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2}
       \right)
@@
 \eeq 
-As for the advection-diffusion, using the standard second-order central 
-difference stencils to compute the correction term is sufficient to ensure 
-that the overall error for the scheme is $O(\dx^4)$.  
+By using the optimal time step and the correction term, the overall error
+for the finite difference scheme~(\ref{eq:burgers_1d_FD_scheme}) is reduced
+from $O(\dx^2)$ to $O(\dx^4)$.  As for the diffusion equation, we need
+to use a second-order accurate discretization for the spatial derivatives
+in the correction term.  
 Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate 
-the boost in accuracy obtained by using the optimal time step and 
-adding correction terms to the finite difference 
+the boost in accuracy obtained by using the optimal time step and adding the
+correction term to the finite difference 
 scheme~(\ref{eq:burgers_1d_FD_scheme}).
+
+The Burgers equation example illustrates an important aspect of optimal time 
+step selection: the ease with which OTS selection can be used to boost the 
+order of accuracy depends on the choice of the finite difference scheme.  
+Had we chosen an upwind approximation for the gradient term in the Burgers 
+equation, the correction term would have been more complicated and included 
+terms involving higher-order spatial derivatives than those that appear in the
+original equation.  While it should be possible to handle this situation with 
+OTS selection, it is clearly less desirable.
 
@@
 \eeq
-where $\kappa > 0$.
+where $\kappa > 0$.  This example illustrates the importance of the choice of
+discretization for the spatial derivative on the applicability of OTS selection.
+
 From our previous examples, we know that optimal time step selection requires 
@@
 difference scheme for~(\ref{eq:4th_order_parabolic_eqn_1d}) that 
-uses forward Euler for time integration and the a fourth-order central 
+uses forward Euler for time integration and a fourth-order central 
 difference approximation for the bilaplacian:
-\bea
+\beq
   u^{n+1}_j = u^{n}_j + \dt \left( -\kappa B u^n + f_j^n \right), 
   \label{eq:4th_order_parabolic_eqn_1d_FD_scheme}
-\eea
+\eeq
 where the fourth-order discrete bilaplacian is given by
@@
 
-Proceeding in the usual way, we use Taylor series expansions and the PDE to 
-express the true solution, $\tu$, at time step $(n+1)$ in terms of the true
-solution and its spatial derivatives at time step $n$: 
-\bea
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.33}{
+  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
+\ \ 
+\scalebox{0.33}{
+  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N_roundoff}} 
+\caption{$L^\infty$ error as a function of number of grid points for four
+finite difference schemes that solve the fourth-order parabolic equation
+(\ref{eq:4th_order_parabolic_eqn_1d}):
+forward Euler with optimal time step and correction term (circles),
+forward Euler with suboptimal time step and no correction term (squares),
+Crank-Nicholson with a second-order stencil for the bilaplacian (diamonds), 
+and Crank-Nicholson with a fourth-order stencil for the bilaplacian and
+an $O(\dx^2)$ time step (triangles).
+Both forward Euler schemes use a fourth-order, central finite difference 
+approximation for the bilaplacian.  The boundary and ghost cells values
+are set by taking values from the known analytical solution.
+Note that the errors for the forward Euler with suboptimal time step and 
+fourth-order Crank-Nicholson schemes lie almost directly on top of each 
+other at the resolution of the figures.  
+The figure on the left highlights the impact of round-off error associated 
+with matrix inversion on the accuracy of the solution for the two 
+Crank-Nicholson schemes.
+}
+\label{fig:4th_order_parabolic_eqn_1d_error}
+\end{center}
+\end{figure}
+
+Proceeding in the usual way, we find that the true solution satisfies
+\beq
   \tu^{n+1}_j = \tu^{n}_j 
-  &+& \dt \left( -\kappa \frac{\partial^4 \tu^n}{\px^4}  
+  + \dt \left( -\kappa \frac{\partial^4 \tu^n}{\px^4}  
              + f^n_j
         \right)
-  \nonumber \\
-  &+& \frac{\dt^2}{2} 
+  + \frac{\dt^2}{2} 
     \left(
@@
   + O \left( \dt^3 \right)
-  \label{eq:4th_order_parabolic_eqn_1d_time_err_modified}.
-\eea
-Next, we substitute $\tu$ into central difference approximation to the 
-bilaplacian to obtain
-\bea
+  \label{eq:4th_order_parabolic_eqn_1d_time_err}
+\eeq
+and that the central difference approximation to the bilaplacian satisfies
+\beq
   B \tu^n = \frac{\partial^4 \tu}{\px^4} 
@@
   + O(\dx^6)
-  \label{eq:4th_order_parabolic_eqn_1d_space_err}
-\eea
+  \label{eq:4th_order_parabolic_eqn_1d_space_err}.
+\eeq
 Combining these results with the finite difference 
 scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}), we find that the
-local truncation error is given by
-\bea
-  \tau^n_j &=& 
+leading-order local truncation error is given by
+\beq
+  \tau^n_j = 
      \frac{\partial^8 \tu^n}{\px^8} 
     \left[ \frac{7 \dx^4}{240} - \frac{\kappa \dt}{2}  \right] (\kappa \dt)
-    \nonumber \\
-    &+& \frac{\dt^2}{2} 
+    + \frac{\dt^2}{2} 
         \left( \kappa \frac{\partial^4 f}{\px^4} 
@@
         \right)
-    + O(\dt \dx^6) + O(\dt^3)
   \label{eq:4th_order_parabolic_eqn_1d_err_eqn}
-\eea
+\eeq
 Thus, the optimal time step is $\dto = 7 \dx^4/120\kappa$ and the correction
@@
 \eeq
-Together, these modifications lead to a finite difference scheme that is 
-sixth-order accurate.  As with the previous examples, the correction term 
-can be computed analytically or numerically.  The improved accuracy of 
-(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) when the optimal time step 
-and correction terms are used is illustrated in 
+Together, these modifications boost the accuracy of the original
+finite difference scheme from fourth- to sixth-order accurate 
+overall.  As with the previous examples, the spatial derivatives in the 
+correction term should be computed using a finite difference approximation 
+that is at least second-order accurate.  
+The improved accuracy of (\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) 
+when the optimal time step and correction term are used is illustrated in 
 Figure~\ref{fig:4th_order_parabolic_eqn_1d_error}.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
-\ \ 
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_comp_time}} 
-\caption{$L^\infty$ error as a function of number of grid points (left)
-and computation time as a function of $L^\infty$ error (right) for two 
-finite difference schemes that solve the fourth-order parabolic equation 
-(\ref{eq:4th_order_parabolic_eqn_1d_space_err}):
-forward Euler with optimal time step and correction term (circles)
-and forward Euler with suboptimal time step $\dt = \dx^2/16$ and no 
-correction terms (squares).
-Both schemes use a fourth-order, central finite difference approximation for
-the bilaplacian.  Using each method, we solve the fourth-order parabolic
-equation with $\kappa = 1$ and 
-$f(x,t) = \sin(2 \pi x)\left(-20000 \pi \sin(10000\pi t) 
-                           + 32\pi^4\cos(10000\pi t)\right)
-        + \sin(5 \pi x)\left( 100 \pi^2 \exp\left(-100\pi^2 t \right)
-           + 625 \pi^4 \left(2 - \exp\left(-100\pi^2 t \right) \right)
-        \right)$
-to time $t = 0.001$ on the domain $0 < x < 1
-$ subject to the boundary conditions $u(0,t) = u(1,t) = 0$ and the initial 
-condition $u(x,0) = 2\sin(2 \pi x) + \sin(5 \pi x)$.  The value of the solution
-in the ghost cells (required as a result of the wide stencil for the 
-discrete bilaplacian) are filled by taking values from the exact solution: 
-$u(x,t) = 2\sin(2 \pi x)\cos(10000 \pi t) 
-        + \sin(5 \pi x) \left[ 2-\exp(-100\pi^2 t) \right]$.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
-}
-\label{fig:4th_order_parabolic_eqn_1d_error}
-\end{center}
-\end{figure}
 
@@
 it is crucial that numerical schemes for high-order PDEs use high-order 
-discretizations for the spatial derivatives.  Using high-order discretizations 
-increases the computational efficiency of numerical schemes and makes it 
-possible to obtain accurate solutions on relatively coarse grids.}. 
+discretizations for the spatial derivatives.}
 Optimal time step selection merely provides a valuable way to leverage the 
@@
 
-Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} shows the computational 
-performance of various finite difference methods for solving 
-equation~(\ref{eq:4th_order_parabolic_eqn_1d_second_time_derivative}).
-Observe the dramatic decrease in computational cost as a function of the
-error by simply replacing a second-order discretization for the bilaplacian 
-operator with a fourth-order discretization.  For the forward Euler scheme,
-using an optimal time step drives the computational cost down even more.
-The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} 
-shows that the theoretical estimates for the computation time are in good 
-agreement with timings from actual computational experiments.
-
 \begin{table}[tb]
-\caption{\label{tab:comp_perf_vs_err_4th_order_parabolic} 
+\caption{
 Computational cost as a function of the global numerical error $e$
@@
 fourth-order parabolic equation.  For all finite difference schemes, the 
-   numerical solution is assumed to be computed over a fixed time interval.
+time step is chosen so that the temporal error is subdominant to the spatial
+error.
 }
-\begin{minipage}{\textwidth}
-\begin{center} \footnotesize
+\label{tab:comp_perf_vs_err_4th_order_parabolic} 
 \renewcommand{\arraystretch}{1.3}
-\begin{tabular}{|l|c|c|c|c|}
+\begin{tabular}{lcccc}
   \hline
   {\bf Numerical Scheme} & $\dx$ 
-  & $\dt$\footnote{The time step for the forward Euler scheme is $O(\dx^4)$
-   as a result of stability considerations.  The time step for the 
-   Crank-Nicholson schemes with second- and fourth-order spatial 
-   discretizations are taken to be $O(\dx)$ and $O(\dx^2)$, respectively, so 
-   that the temporal error is subdominant to the spatial error.}
-  & {\bf Memory}\footnote{The memory requirement for each scheme is estimated 
-    to be $O(1/\dx)$.} 
-  & {\bf Compute Time}\footnote{The computation time is approximated as 
-    $O\left( (1/\dx) (1/\dt) \right)$, which assumes that we are using an 
-    efficient $O(N) = O(1/\dx)$ algorithm to invert the matrices that 
-    arise for the Crank-Nicholson method.}  \\
+  & $\dt$
+  & {\bf Memory}
+  & {\bf Compute Time}
+  \\
   \hline 
@@
   \ \ 2nd-Order Bilaplacian & & & & \\
-  \hline 
   Forward Euler with 
@@
   \ \ 4th-Order Bilaplacian & & & & \\
-  \hline 
   Forward Euler with 
@@
   \ \ 4th-Order Bilaplacian and OTS & & & & \\
-  \hline 
   Crank-Nicholson with 
@@
   \ \ 2nd-Order Bilaplacian & & & & \\
-  \hline 
   Crank-Nicholson with 
@@
 \end{tabular}
+\end{table}
+
+Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} shows the computational 
+performance of various finite difference methods for solving 
+equation~(\ref{eq:4th_order_parabolic_eqn_1d_second_time_derivative}).
+Observe the dramatic decrease in computational cost as a function of the
+error by simply replacing a second-order discretization for the bilaplacian 
+operator with a fourth-order discretization.  For the forward Euler scheme,
+using an optimal time step drives the computational cost down even more.
+The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf} 
+shows that the theoretical estimates are in good agreement with timings 
+from computational experiments.  Note that the empirical performance for the 
+forward Euler schemes is better than theoretically expected because the 
+computational work per time step is not dominated by the 
+$O(1/\dx)$ update calculations until $\dx$ gets very small.
+
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.33}{
+  \includegraphics{figures/4th_order_parabolic_eqn_1d_comp_time}} 
+\caption{Computation time as a function of $L^\infty$ error for four
+finite difference schemes that solve the fourth-order parabolic equation
+(\ref{eq:4th_order_parabolic_eqn_1d}).  
+These results were obtained using MATLAB implementations of the 
+finite difference schemes run on a 2.4 GHz MacBook Pro.
+Note the impact of round-off error associated with matrix inversion on 
+the accuracy of the solution for the two Crank-Nicholson schemes.
+See the caption of Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} for 
+descriptions of the four finite difference schemes.
+}
+\label{fig:4th_order_parabolic_eqn_1d_perf}
 \end{center}
-\end{minipage}
-\end{table}
-
-Comparing the best forward Euler and Crank-Nicholson schemes shows that
+\end{figure}
+
+In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
+2nd-order Crank-Nicholson scheme actually outperforms the OTS scheme over 
+the high (but still practically useful) range of errors even though 
+Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS
+scheme should do better.  The origin of this discrepancy is the constant 
+hidden by the big O notation.  In particular, the factor of $7/120$ in 
+the optimal time step for the forward Euler scheme with a fourth-order 
+approximation of the bilaplacian leads to a relatively large hidden constant 
+for the compute time. 
+However, as Figures~\ref{fig:4th_order_parabolic_eqn_1d_error} and
+\ref{fig:4th_order_parabolic_eqn_1d_perf} show, there is an important 
+limitation of the 2nd-order Crank-Nicholson scheme even if we ignore the 
+cross-over which will occur at sufficiently small values of the error -- 
+round-off error associated with matrix inversion limits the accuracy that is 
+achievable by the scheme.  This limitation also exists for the 4th-order 
+Crank-Nicholson scheme.
+
+Comparing the best forward Euler and Crank-Nicholson schemes, we see that
 the optimal forward Euler scheme more efficiently utilizes memory at the
@@
 dimensions, the forward Euler scheme requires less compute time than the
-Crank-Nicholson scheme.  The origin of the improved performance for the 
-forward Euler scheme with optimal time step in higher dimensions is the 
-significantly smaller size of the computational grid required to achieve a 
-desired error level.
+Crank-Nicholson scheme.
 
@@
 discretization error, multiple grid spacing parameters,  and the domain 
-geometry.  In this section, we address these issues and demonstrate the use 
-of optimal time step selection in the construction of finite difference 
-schemes for two linear, time dependent PDEs in two space dimensions on both 
-regular and irregular domains.  It is straightforward to generalize the 
-approach presented in this section to more space dimensions and to construct 
-methods for semilinear PDEs by incorporating ideas discussed in 
+geometry.  In this section, we show how optimal time step selection easily 
+overcomes these issues by using it to construct high-order finite difference
+schemes for the 2D advection equation and the 2D diffusion equation.  
+It is straightforward to generalize the approach presented in this section to 
+more space dimensions and to construct methods for semilinear PDEs by 
+incorporating ideas discussed in 
 Sections~\ref{sec:optimal_time_step_selection} and~\ref{sec:applications_1d}.
@@
 one space dimension and multiple space dimensions is the presence of 
-cross-terms in the discretization error.  When analyzing finite difference 
-schemes using conventional numerical analysis techniques, it is usually safe 
-to ignore the precise form of the cross-terms because they only appear as 
-higher-order corrections in the error.  As a result, the order of accuracy for 
-the finite difference scheme is not sensitive to the particular choice of 
-discretization for the spatial derivative operators.  In contrast, knowledge 
-of the form of the cross-terms is critical when attempting to use optimal time 
-step selection.  Without this knowledge, it is impossible to ensure 
-cancellation of higher-order errors.
+cross-terms in the discretization error.  When formally analyzing the accuracy
+of a finite difference scheme, it is usually safe to ignore the details of 
+these terms because they only appear as higher-order corrections in the error.
+However, precise knowledge of the cross-terms is critical when attempting to 
+use optimal time step selection.  Without this knowledge, it is impossible to 
+ensure cancellation of higher-order errors.
 
@@
 These two approaches are actually equivalent because the correction terms 
-can be combined with the finite difference stencils in the original scheme.  
-In essense, the correction terms can be viewed as part of an expanded spatial 
-discretization scheme and vice-versa.
+can be combined with the finite difference stencils in the original scheme 
+and be viewed as part of an expanded spatial discretization scheme.
 
@@
 \eeq
-where $\vec{A} = (A_x, A_y)$ is the advection velocity.  For convenience, we 
-take $A_x, A_y > 0$.  This choice allows us to avoid unnecessary complexity 
-associated with the dependence of the upwind direction on the sign of $A_x$ 
-and $A_y$.  It is straightforward to generalize our discussion to $A_x$ and 
-$A_y$ of arbitrary sign.
-
-Recall that in one space dimension, using an optimal time step for
-a simple forward Euler scheme that uses a first-order upwind discretization
-of the spatial derivative leads to a finite difference scheme for the 
-advection equation with no local truncation error.  The same result can be 
-achieved in two space dimensions.  However, because there are a total of three
-numerical parameters (\ie two grid spacings and the time step), it is 
-insufficient to optimize only the time step; the ratio of the grid 
-spacings also needs to be optimally selected.
-
-For the purposes of illustration, we consider a simple forward Euler scheme 
-with first-order upwind discretizations of the advection terms:
+where $\vec{A} = (A_x, A_y)$ is the advection velocity.  
+The most interesting and important aspect of this problem is that elimination 
+of the truncation error requires optimal selection of the grid spacing ratio 
+as well as the time step.  That is, the more general idea of numerical 
+parameter optimization is needed in order to boost the accuracy of the scheme.
+
+For convenience, let us assume that $A_x, A_y > 0$.  The simple forward Euler 
+scheme for (\ref{eq:advection_eqn_2d}) with first-order upwind discretizations 
+of the advection terms is 
 \beq
@@
 This method is formally first-order in both space and time with a 
-stability constraint of $\dt < \left( A_x/dx + A_y/dy \right)^{-1}$.
-
-To compute the optimal time step and grid spacing ratio, we begin, as
-usual, by analyzing the discretization error 
+stability constraint of $\dt < \left( A_x/\dx + A_y/\dy \right)^{-1}$.
+
+To compute the optimal time step and grid spacing ratio, we begin by analyzing 
+the discretization error 
 for~(\ref{eq:advection_eqn_2d_FD_scheme}).  As in our analysis of the finite
 difference scheme for the one-dimensional advection equation, we retain all of 
-the terms in the discretization error.  Using Taylor series to expand the true 
-solution of the PDE at time step $(n+1)$, we find that
+the terms in the discretization error.  Using Taylor series expansions and
+the PDE (\ref{eq:advection_eqn_2d}), we find that the true solution $\tu$ 
+satisfies
 \bea
-  \tu^{n+1}_{i,j} = \tu^{n}_{i,j }
-  + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
-       \frac{\partial^k \tu^n}{\pt^k} 
-  \label{eq:advection_eqn_2d_time_err}.
-\eea
-where $\tu$ denotes the true solution of the PDE. 
-Similarly, Taylor series expansion of the upwind finite difference 
-approximation to the spatial derivative yields
-\bea
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i-1,j}}{\dx} = 
+  \tu^{n+1}_{i,j} &=& \tu^{n}_{i,j }
+    + \sum_{k=1}^\infty \frac{\left( -\dt \right)^k}{k!} 
+         \left( A_x \partial_x
+              + A_y \partial_y \right)^k \tu^{n}.
+  \label{eq:advection_eqn_2d_time_err}
+  \\
+  \frac{\tu^{n}_{i,j} - \tu^{n}_{i-1,j}}{\dx} &=& 
   \frac{\partial \tu}{\px} 
@@
   \\
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dy} = 
+  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dy} &=&
   \frac{\partial \tu}{\py} 
@@
        \frac{\partial^k \tu^n}{\py^k} 
-  \label{eq:advection_eqn_2d_space_err_y}
+  \label{eq:advection_eqn_2d_space_err_y}.
 \eea
-Next, we use the PDE~(\ref{eq:advection_eqn_2d}) to replace all of 
-the time derivatives in~(\ref{eq:advection_eqn_2d_time_err}) with spatial
-derivatives to obtain
-\bea
-  \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
-  + \sum_{k=1}^\infty \frac{\left( -\dt \right)^k}{k!} 
-       \left( A_x \partial_x
-            + A_y \partial_y \right)^k \tu^{n}.
-  \label{eq:advection_eqn_2d_time_err_modified}
-\eea
-Combining~(\ref{eq:advection_eqn_2d_space_err_x}), 
-(\ref{eq:advection_eqn_2d_space_err_y}), 
-and~(\ref{eq:advection_eqn_2d_time_err_modified}) with the finite difference 
-scheme~(\ref{eq:advection_eqn_2d_FD_scheme}), we find that the local
-truncation error is given by:
+Therefore, the local truncation error is given by:
 \bea
@@
 \eea
-With these choices, we are able to boost the order of accuracy for the 
-finite difference scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from first- to 
+These choices enable us to boost the order of accuracy for the finite 
+difference scheme (\ref{eq:advection_eqn_2d_FD_scheme}) from first- to 
 second-order.
@@
 \beq
-   \frac{u_{i,j} - u_{i-1,j} - u_{i,j-1} + u_{i-1,j-1}}
+   \frac{ \left(u_{i,j} - u_{i-1,j}\right) 
+        - \left(u_{i,j-1} - u_{i-1,j-1}\right)}
         {\dx \dy} 
@@
 equation~(\ref{eq:advection_eqn_2d}) using forward Euler time stepping 
-with optimal time step and correction terms (left) and forward Euler time 
+with optimal time step and correction term (left) and forward Euler time 
 stepping with suboptimal time step (right).  The latter scheme uses a 
@@
 
-To see that the correction term can be considered as part of the spatial 
-discretization stencil, we can write out the complete finite difference scheme 
-including the correction term:
+It is interesting to push the analysis of the optimal finite difference 
+scheme a little further and examine the complete finite difference scheme
+(including correction term):
 \bea
-  u^{n+1}_{i,j} &=& u^{n}_{i,j}
-  - A_x \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} \right)
+  u^{n+1}_{i,j} = u^{n}_{i,j}
+  &-& A_x \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i-1,j}}{\dx} \right)
   - A_y \dt \left( \frac{u^{n}_{i,j} - u^{n}_{i,j-1}}{\dy} \right)
   \nonumber \\
-  & & + A_x A_y \dt^2 
+  &+& A_x A_y \dt^2 
         \left( \frac{u_{i,j} - u_{i-1,j} - u_{i,j-1} + u_{i-1,j-1}}
@@
 \eea
-Using the relationship between the optimal time step and the grid spacings, 
-this scheme can be rewritten as
+Using the relationship between the optimal time step and the grid spacing
+ratio, this scheme can be rewritten as
 \bea
@@
          + \frac{u^{n}_{i-1,j} - u^{n}_{i-1,j-1}}{\dy} 
-    \right).
+    \right),
   \label{eq:advection_eqn_2d_FD_scheme_simplified}
 \eea
-Thus, we see that the finite difference scheme 
-(\ref{eq:advection_eqn_2d_FD_scheme_complete}) is essentially forward Euler 
-time integration with first-order upwind discretizations of the spatial 
-derivatives that are computed using all four upwind neighbors.
+which is essentially a forward Euler time integration scheme with 
+first-order upwind discretizations of the spatial derivatives that are 
+computed using all four upwind neighbors.  This formulation of 
+(\ref{eq:advection_eqn_2d_FD_scheme_complete}) suggests that the highest
+accuracy finite difference scheme for the 2D linear advection equation arises
+by using a fully upwind discretization of the advection terms with the 
+optimal time step and optimal ratio of grid spacings.
 
@@
 advection equation because many high-order correction terms must be retained.  
-However, it is straighforward to generalize 
+However, it is straightforward to generalize 
 (\ref{eq:advection_eqn_2d_FD_scheme_simplified}) to $d$ space dimensions by 
@@
 
-\subsection{Diffusion Equation}
+\subsection{Diffusion Equation \label{sec:diffusion_eqn_2d}}
 As for PDEs in one space dimension, achieving infinite order accuracy 
@@
 In general, optimal parameter selection boosts the accuracy by only a few 
-orders.  To illustrate the typical accuracy boost gained by using optimal 
-time step selection, we apply it to the 2D diffusion equation.  We also 
-use the 2D diffusion equation to demonstrate the utility of optimal time 
-step selection for problems on irregular domains.
-
-In two space dimensions, the diffusion equation (with source term) may be 
-written as
-\beq
-  \frac{\partial u}{\pt} = D \nabla^2 u + f(x,y,t),
+orders.  To illustrate the typical accuracy boost gained by using OTS
+selection, we apply it to the 2D diffusion equation.  This example underscores 
+the need to carefully choose the discretization for the spatial derivatives 
+for problems in more than one space dimension and demonstrates the ease with 
+which OTS selection can be applied to problems on irregular domains.
+
+In two space dimensions, the diffusion equation may be written as
+\beq
+  \frac{\partial u}{\pt} = D \nabla^2 u + f(x,y,t).
   \label{eq:diffusion_eqn_2d}
 \eeq
-where $\nabla^2$ is the Laplacian operator, $D$ is the diffusion constant,
-and $f$ is the source term.
-While there are many finite difference schemes that are used to solve the 
-2D diffusion equation, the most common ones employ a five-point, second-order 
-central difference stencil for the Laplacian~\cite{iserles_book}:
+The most common finite difference schemes for (\ref{eq:diffusion_eqn_2d}) 
+employ a five-point, second-order central difference stencil for the 
+Laplacian:
 \beq
@@
 equal.  The forward Euler scheme using the five-point stencil is given by
-\bea
-  u^{n+1}_j &=& u^{n}_j 
+\beq
+  u^{n+1}_j = u^{n}_j 
   + \dt 
@@
   \label{eq:diffusion_eqn_2d_5pt_scheme}
-\eea
+\eeq
 This scheme is formally first-order in time and second-order in space.  
+The stability constraint $\dt \le \dx^2/4D$ implies that this scheme is
+$O(\dx^2)$ accurate overall.
 
@@
 (\ref{eq:laplacian_2d_5pt_stencil}), 
-$\dx^2 \left(u_{xxxx} + u_{yyyy}\right)/12$, cannot be directly related 
+$\dx^2 \left(\tu_{xxxx} + \tu_{yyyy}\right)/12$, cannot be directly related 
 to the spatial derivative operator in the diffusion equation.  
@@
 (\ref{eq:diffusion_eqn_2d}):
-\bea
+\beq
   L^{9pt} u = \frac{1}{6 \dx^2} \left( 
@@
   \label{eq:laplacian_2d_9pt_stencil}
-\eea
+\eeq
 Observe that the nine-point stencil is simply the standard five-point 
@@
 five-point stencil with the extra contribution of 
-$\dx^2 u_{xxyy}/6$
+$\dx^2 \tu_{xxyy}/6$
 \bea
-L^{9pt} u &=& \nabla^2 u 
-              + \frac{\dx^2}{12} \left(u_{xxxx} + 2 u_{xxyy} + u_{yyyy}\right)
+L^{9pt} \tu &=& \nabla^2 \tu 
+              + \frac{\dx^2}{12} 
+                \left(\tu_{xxxx} + 2 \tu_{xxyy} + \tu_{yyyy}\right)
               + O(\dx^4)
           \nonumber \\
-          &=& \nabla^2 u + \frac{\dx^2}{12} \nabla^4 u + O(\dx^4).
+          &=& \nabla^2 \tu + \frac{\dx^2}{12} \nabla^4 \tu + O(\dx^4).
   \label{eq:laplacian_2d_9pt_stencil_error}
@@
 naturally arises when directly applying optimal time step selection to the
-five-point scheme if the correction terms are interpreted as a modification 
-to the spatial discretization.
-
-To apply optimal time step selection to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we 
-proceed in the usual way by first analyzing the leading-order terms of the 
+five-point scheme if the correction term is interpreted as a modification 
+to the original spatial discretization.
+
+To apply optimal time step selection to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we proceed in the usual way by first analyzing the leading-order terms of the 
 discretization error for the finite difference scheme.  Taylor series 
-expansions show that the true solution satisfies the following equations:
-\bea
-  \tu^{n+1}_{i,j} &=& \tu^{n}_{i,j}
-  + \dt \frac{\partial \tu^n}{\pt} 
-  + \frac{\dt^2}{2} \frac{\partial^2 \tu^n}{\pt^2} + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_2d_time_err} 
-  \\
-  L^{9pt} \tu &=&
-  \nabla^2 \tu + \frac{\dx^2}{12} \nabla^4 \tu + O(\dx^4)
-  \label{eq:diffusion_eqn_2d_space_err}
-\eea
-Using the PDE~(\ref{eq:diffusion_eqn_2d}) to replace time derivatives
-with space derivatives, (\ref{eq:diffusion_eqn_2d_time_err}) becomes
+expansions and the PDE show that the true solution satisfies 
 \bea
   \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
-  + D \dt \nabla^2 \tu^{n}_{i,j}
-  + \frac{(D \dt)^2}{2} \nabla^4 \tu^{n}_{i,j}
+  + \dt \left(\nabla^2 \tu^{n} +f^n_{i,j} \right)
+  + \frac{(D \dt)^2}{2} \nabla^4 \tu^{n}
+  + \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right)
   + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_2d_time_err_modified}
+  \label{eq:diffusion_eqn_2d_time_err} 
 \eea
-Combining these results, we find that the local truncation error for the
-finite difference scheme (\ref{eq:diffusion_eqn_2d_9pt_scheme}) is given
-by 
+Combining this result with~(\ref{eq:laplacian_2d_9pt_stencil_error}), 
+we find that the leading-order terms in the 
+local truncation error for the finite difference 
+scheme (\ref{eq:diffusion_eqn_2d_9pt_scheme}) are
 \bea
@@
       \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-      + O(\dt \dx^4) + O(\dt^3).
+  - \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right).
   \label{eq:diffusion_eqn_2d_trunc_err}
@@
 As for the 1D diffusion equation, we see that the optimal time step is given 
-by $\dto = \dx^2/6D$ and that this choice of time step yields a scheme 
-that is fourth-order accurate.  Note that there is no need to consider an 
-implicit scheme because $\dto$ is stable for the forward Euler time 
-integration scheme.
+by $\dto = \dx^2/6D$ and the correction term is 
+\beq
+  \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right).
+\eeq
+This choice of time step and correction term yields a scheme that is 
+fourth-order accurate.
 
@@
 finite difference schemes that solve the 2D diffusion equation:
-forward Euler with optimal time step and correction terms (circles), forward 
-Euler with suboptimal time step $\dt = \dx^2/4D$ (squares), and 
-Crank-Nicholson with time step $\dt = \dx/16D$ (diamonds).  
+forward Euler with optimal time step and correction term (circles), forward 
+Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
 The forward Euler schemes use a nine-point discretization of the Laplacian;
 the Crank-Nicholson scheme uses the standard five-point discretization of
-the Laplacian.  Using each method, we solve the 2D diffusion equation with 
-$D = 1/4$ and source term 
-$f(x,y,t) = 8 D \pi^2 \sin(2 \pi x) \sin(3 \pi y) 
-            \exp\left(-5 D \pi^2 t\right)
-          - 24 D \pi^2 \sin(5 \pi x) \sin(\pi y) 
-            \exp \left(-2 D \pi^2 t \right)$
-to time $t = 0.1$ on the domain $0 < x, y < 1$ subject 
-to the initial condition
-$u(x,y,0) =  1 + y/3 + x/2 + xy/4 
-          + \sin(2 \pi x) \sin(3 \pi y) - \sin(5 \pi x) \sin(\pi y)$.
-The boundary conditions are taken from the analytical solution 
-$u(x,t) = 1 + y/3 + x/2 + xy/4 
-          + \sin(2 \pi x) \sin(3 \pi y) \exp(-5 D \pi^2 t) 
-          - \sin(5 \pi x) \sin(\pi y) \exp(-2 D \pi^2 t)
-$.
+the Laplacian.  
 Note that the errors for the forward Euler with suboptimal time step and 
@@
 resolution of the figures.  
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
+These results were obtained using MATLAB implementations of the 
+finite difference schemes run on a 2.4 GHz MacBook Pro.
 }
@@
 \end{figure}
-
 The improvement in the accuracy of the forward Euler scheme when the optimal 
@@
 (in any number of space dimensions) could be obtained by using a 
-pseudospectral spatial discretization~\cite{trefethen_spectral_book, 
-boyd_spectral_book} of the Laplacian and a fourth-order time integration scheme 
+pseudospectral spatial 
+discretization~\cite{trefethen_spectral_book,boyd_spectral_book} 
+of the Laplacian and a fourth-order time integration scheme 
 (\eg Runge-Kutta).  When the domain is irregularly shaped, high-order methods 
-are harder to design and implement.  For instance, spectral and pseudospectral 
-methods are not directly applicable because the domain is not a simple 
-cartesian product of intervals.  While it is possible to use spectral methods
-for problems on irregular domains through the clever introduction of a phase 
-field~\cite{bueno_orovio_2006, feng_2006}, this approach does not 
-actually achieve spectral accuracy because the error is dominated by the 
-width of the smoothed approximation to the phase field (as opposed to the 
-accuracy of the discretization for the spatial 
-derivatives)~\cite{bueno_orovio_2006}.
+are harder to design and implement.  
+In this section, we extend the OTS finite difference scheme for the 2D 
+diffusion equation on regular domains to irregular domains.  Other 
+researchers have developed fourth-order accurate methods for the diffusion 
+equation on irregular domains~\cite{gibou_2005,ito_2005}.  However, these 
+earlier methods require a wider ghost cell region~\cite{gibou_2005}, use
+implicit time integration~\cite{gibou_2005,ito_2005}, or involve a more 
+complex calculation for the boundary conditions~\cite{ito_2005}.  As we shall 
+see, it is possible to achieve fourth-order accuracy for the diffusion 
+equation on irregular domains using only forward Euler integration and a 
+compact, second-order stencil for the Laplacian.
 
@@
 \scalebox{0.25}{\includegraphics{figures/ghostcell_corner}}
-\caption{Illustrations of edge (left) and corner (right) ghostcells.
-The edge ghostcell, $u_E$, is filled by using cubic extrapolation of 
+\caption{Illustrations of edge (left) and corner (right) ghost cells.
+The edge ghost cell, $u_E$, is filled by using cubic extrapolation of 
 the value on the boundary, $u_B$, and values at three interior grid points.  
-The corner ghostcell, $u_C$, is filled by using a fourth-order accurate
+The corner ghost cell, $u_C$, is filled by using a fourth-order accurate
 Taylor series expansion centered at $u_{0,0}$, which uses the values of $u$ 
@@
 
-In this section, we extend the OTS finite difference scheme for the 2D 
-diffusion equation on regular domains to irregular domains.  To avoid the 
-need for special stencils at grid points near the irregular boundary, we adopt 
-the `ghost cell' approach for imposing boundary conditions~\cite{gibou_2005, 
-fedkiw_1999, osher_fedkiw_book}.  The basic idea is to set the value of $u$ at 
-any grid point that lies immediately outside of the domain (and thus may be
-required by spatial derivative operators at interior grid points) by 
-extrapolating from values of $u$ at interior grid points and using boundary 
-condition information.  
-The main challenge is to find a method for filling ghost cells that does not 
-decrease the accuracy of the solution in the interior of the domain.  For 
-simplicity, we choose to represent the irregular boundary implicitly as the 
-zero level set of a level set function $\phi$ which satisfies:
-\bea
-  \begin{array}{ll}
-  \phi < 0 & \mathrm{for} \ x \in \Omega \\
-  \phi > 0 & \mathrm{for} \ x \notin \Omega \\
-  \phi = 0 & \mathrm{for} \ x \in \partial \Omega 
-  \end{array}
-\eea
-where $\Omega$ and $\partial \Omega$ are the interior and boundary of the 
-irregular domain.
-
+To avoid the need for special stencils at grid points near the irregular 
+boundary, we adopt the `ghost cell' approach for imposing boundary 
+conditions~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.
+The main challenge in using the ghost cell method is ensuring ghost cells 
+are filled using a sufficiently high-order extrapolation scheme. 
 Figure~\ref{fig:ghost_cells} shows examples of ghost cells that lie in the 
@@
 
-To set the value of $u$ in each edge ghost cell, we use 1D, cubic 
-extrapolation of the values of $u$ across the interface.  A cubic Lagrange 
-interpolant is constructed for each edge ghost cell using a point on the 
-domain boundary and three interior grid points that lie along the line 
-parallel to the edge connecting the ghost cell to its nearest interior 
-neighbor (see Figure~\ref{fig:ghost_cells}).  
-The location of the point $u_B$ can be calculated by inverting the linear 
+To set the value of $u$ in each edge ghost cell, we use 1D cubic 
+(\ie fourth-order accurate) extrapolation of the values of $u$ across the 
+interface.  A cubic Lagrange interpolant is constructed for each edge ghost 
+cell using a point on the domain boundary and three interior grid points that 
+lie along the line parallel to the edge connecting the ghost cell to its 
+nearest interior neighbor (see Figure~\ref{fig:ghost_cells})\footnote{If
+a level set representation of the boundary~\cite{osher_fedkiw_book} is used, 
+the location of the point $u_B$ can be calculated by inverting the linear 
 interpolant of $\phi$ that passes through the ghost cell and its nearest
-interior neighbor\footnote{We found that using higher-order interpolants to 
+interior neighbor.  We found that using higher-order interpolants to 
 locate boundary points was unnecessary for achieving high-order accuracy in 
 the computed solution.}.  Because the quality of the interpolant deteriorates 
-if the boundary point is too close to any of the interior grid points used to 
-construct the Lagrange extrapolant, we follow \cite{gibou_2005} and choose the 
-interior grid points so that the nearest interior grid point is sufficiently 
-far from the boundary point.  Without this procedure, the errors introduced 
-in the values of the ghost cells lead to instability of the numerical method.
+if the boundary point $u_B$ is too close to any of the interior grid points 
+used to construct the Lagrange extrapolant, we follow \cite{gibou_2005} and 
+choose the interior grid points so that the nearest interior grid point is 
+sufficiently far from the boundary point.  Without this procedure, the errors 
+introduced in the values of the ghost cells lead to instability of the 
+numerical method.
 Gibou~\etal~suggest shifting the interpolation points by one grid point 
@@
 point is less than $\dx^2$~\cite{gibou_2005}.  For the numerical scheme 
-presented in this paper, this threshold was too low.  We found that better 
-stability was obtained if we used an $O(\dx)$ threshold.
+presented in this paper, this threshold was too low.  Better stability was 
+obtained if an $O(\dx)$ threshold was used.
 
@@
 \caption{Numerical solutions (left) and dominant error (right) for the 2D 
-diffusion equation on a starfish-shaped domain.  The solution is computed at 
-time $t = 0.1$ using forward Euler time stepping with an optimal time step 
-and correction terms.  In the error plot, the dark regions represent points 
+diffusion equation on a starfish-shaped domain.  The solution is computed 
+using forward Euler time stepping with an optimal time step and correction 
+term.  In the error plot, the dark regions represent points 
 where the error in the solution is larger than $25$\% of the $L^\infty$ error 
-of the solution.  For these figures, we use the same diffusion constant, 
-source term, and initial conditions as described in the caption for 
-Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.  The boundary of the 
-starfish-shaped domain is represented by the zero-level set of the function
-$\sqrt{x^2 + y^2} - \frac{3}{5} \left(1 + \frac{1}{2}\sin(5 \theta) \right)$, 
-where $\theta = \mathtt{atan}\left( y/x \right)$.
-The Dirichlet boundary conditions are taken from the values of the analytical 
-solution (see caption for Figure~\ref{fig:diffusion_eqn_2d_src_analysis})
-on the irregular boundary.  The numerical solution is calculated using the 
-square computational domain $-1 < x,y < 1$ which is discretized using 400 grid 
-points in each coordinate direction (the plot of the solution has been 
-downsampled for visualization purposes).  Boundary conditions are imposed by 
-using fourth-order accurate approximations of the value of the solution in 
-edge and corner ghost cells (described in the main text).
+of the solution.  
 }
@@
 at interior grid points and the neighboring edge ghost cells (see 
-Figure~\ref{fig:ghost_cells}).  To obtain a fourth-order accurate value for 
-the corner ghost cell, we need to carefully choose the finite difference 
-stencils used for the partial derivatives.  For the corner ghost cell shown 
+Figure~\ref{fig:ghost_cells}).  For the corner ghost cell shown 
 in Figure~\ref{fig:ghost_cells}, the extrapolation stencil is given by
@@
 the domain are obtained by rotations of Figure~\ref{fig:ghost_cells} and 
-remapping of the indicies in the above stencil.
+remapping of the indices in the above stencil.
 
@@
 error in the solution is concentrated near the boundaries of the irregular
-domain.  Thus, we see that it is necessary to use high-order accurate 
-extrapolants for the ghost cells to prevent errors at the boundary from 
-destroying high-order accuracy throughout the domain.  The improved accuracy 
-of the OTS solution compared to a simple forward Euler scheme with a 
-suboptimal time step is illustrated in 
+domain.  The improved accuracy of the OTS solution compared to a simple 
+forward Euler scheme with a suboptimal time step is illustrated in 
 Figure~\ref{fig:diffusion_eqn_2d_starfish_error}.
@@
 irregular domain: forward Euler with optimal time step and correction 
-terms (circles) and forward Euler with suboptimal time step $\dt = \dx^2/4 D$
-(squares).  Both methods solve the 2D diffusion equation described in the 
-caption for Figure~\ref{fig:diffusion_eqn_2d_src_analysis} on the irregular
-domain defined in the caption for 
-Figure~\ref{fig:diffusion_eqn_2d_starfish_domain}.
-These results were obtained using simple MATLAB implementations of the 
-finite difference schemes.
+terms (circles) and forward Euler with suboptimal time step (squares).  
 }
@@
 approximations with more general or complex properties, symbolic algebra 
-software can be helpful~\cite{patra_2005, gupta_1998}.
+software can be helpful~\cite{patra_2005,gupta_1998}.
 
@@
 order of accuracy from two to four.  Above three dimensions, it is necessary 
-to use an implicit time integration scheme because $\dt_{opt}$ is no longer 
+to use an implicit time integration scheme if $\dt_{opt}$ is no longer 
 stable for the forward Euler scheme.
@@
 PDE to relate terms in the discretization error.  The boost in the order
-of accuracy achieved when the optimal time step and correction terms is well
-worth the extra effort required to explicitly retain and analyze the 
-leading-order terms in the discretization error.
+of accuracy achieved when the optimal time step and correction terms are
+used is well worth the extra effort required to explicitly retain and analyze 
+the leading-order terms in the discretization error.
 
@@
 and the Institute for High-Performance Computing (IHPC) in Singapore. 
-The author would like to thank P. Fok, M. Prodanovic, and A. Chiu for 
+The author would like to thank P. Fok, M. Prodanovi\'c, and A. Chiu for 
 helpful suggestions on the manuscript.  The author particularly thanks P. Fok 
@@
 
+\appendix
+\section{Formal vs. Practical Accuracy
+         \label{app:formal_vs_practical_accuracy} }
+For time dependent PDEs, evaluating the order of accuracy for a numerical
+method is subtle because there are formally two separate orders of accuracy 
+to consider:  temporal and spatial.  While it is theoretically
+interesting and important to understand how the error depends on both the 
+grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
+of a numerical scheme is always controlled by only one of the two numerical
+parameters.
+
+Even though spatial and temporal orders of accuracy for a numerical method
+are formally separate, one will always dominate for a given choice of 
+$\dx$ and $\dt$.  For example, when solving the diffusion equation 
+using the backward Euler method for time integration with a second-order 
+central difference stencil for the Laplacian, formal analysis shows that 
+the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
+stability constraints on the numerical parameters, the time
+step and grid spacing are free to vary independently.  In this situation, the 
+practical error for the method depends on the relative sizes of the time step 
+and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
+$O(\dt)$ which means that the error in the numerical solution is 
+primarily controlled by the time step.  Similarly, when 
+$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
+the error is controlled by the grid spacing.  Finally, when 
+$\dt  = O(\dx^2)$, the practical error is 
+$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
+primarily controlled by one of the two numerical parameters, and varying
+the subdominant parameter while holding the dominant parameter fixed does 
+not significantly affect the error.
+
+When there are constraints on the numerical parameters, there is less freedom 
+in choosing the controlling parameter.  For instance, if we solve the 
+diffusion equation using a forward Euler time integration scheme with a 
+second-order central difference stencil for the Laplacian, stability
+considerations require that we choose $\dt = O(\dx^2)$.  
+Combining this stability constraint with the formal error for the scheme
+shows that the practical error is 
+$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
+method is completely controlled by the grid spacing; the temporal error can 
+never dominate the spatial error.
+
 \bibliography{OTSPDE}

### 2008-11-15 papers/OTSPDE/getFigures.sh
OTSPDE: - updated getFigures.sh script.

@@ -10,14 +10,12 @@
 FIGURES_DIR="figures"
 
 SRC_DIR="../../src"
-EXAMPLE_DIR_LIST="1d/advection_diffusion_eqn                            \
-                  1d/burgers_eqn/single_hump                            \
-                  1d/diffusion_eqn                                      \
-                  1d/fourth_order_parabolic                             \
-                  2d/advection_eqn                                      \
-                  2d/diffusion_eqn/square_domain                        \
-                  2d/diffusion_eqn/irregular_domain/circle_domain       \
-                  2d/diffusion_eqn/irregular_domain/starfish_domain     \
+EXAMPLE_DIR_LIST="scalar_PDEs/1d/burgers_eqn/single_hump                       \
+                  scalar_PDEs/1d/diffusion_eqn                                 \
+                  scalar_PDEs/1d/fourth_order_parabolic                        \
+                  scalar_PDEs/2d/advection_eqn                                 \
+                  scalar_PDEs/2d/diffusion_eqn/square_domain                   \
+                  scalar_PDEs/2d/diffusion_eqn/irregular_domain/starfish_domain\
                  "
 
 # create directory for figures if it doesn't already exist

### 2008-11-15 papers/OTSPDE/OTSPDE.tex
OTSPDE: - updated OTSPDE paper so that the macros and bibliography are in the main   LaTEX file.

@@
 
-\include{macros}
+%******* MACROS *******
+
+% MATH MACROS
+\newcommand{\beq}{\begin{equation}}
+\newcommand{\eeq}{\end{equation}}
+\newcommand{\bea}{\begin{eqnarray}}
+\newcommand{\eea}{\end{eqnarray}}
+
+\def\pt{\partial t}
+\def\px{\partial x}
+\def\py{\partial y}
+
+% NUMERICS MACROS
+\def\tu{\tilde{u}}
+\def\dt{\Delta t}
+\def\dx{\Delta x}
+\def\dy{\Delta y}
+\def\dto{\dt_{opt}}
+
+% MISC MACROS
+\def\eg{\emph{e.g., }}
+\def\ie{\emph{i.e., }}
+\def\etal{\emph{et al.}}
 
@@
 
-\bibliography{OTSPDE}
+%*********************************BIBLIOGRAPHY**********************************
+%\bibliography{OTSPDE}
+
+\begin{thebibliography}{10}
+
+\bibitem{bruger_2005}
+A.~Br\"uger, B.~Gustafsson, P.~L\"otstedt, and J.~Nilsson.
+\newblock High order accurate solution of the incompressible {N}avier-{S}tokes
+  equations.
+\newblock {\em J. Comput. Phys.}, 203:49--71, 2005.
+
+\bibitem{gibou_2005}
+F.~Gibou and R.~Fedkiw.
+\newblock A fourth order accurate discretization for the {L}aplace and heat
+  equations on arbitrary domains, with applications to the {S}tefan problem.
+\newblock {\em J. Comput. Phys.}, 202:577--601, 2005.
+
+\bibitem{ito_2005}
+K.~Ito, Z.~Li, and Y.~Kyei.
+\newblock {H}igher-order, {C}artesian {G}rid {B}ased {F}inite {D}ifference
+  {S}chemes for {E}lliptic {E}quations on {I}rregular {D}omains.
+\newblock {\em SIAM J. SCI. COMPUT.}, 27:346--367, 2005.
+
+\bibitem{shukla_2005}
+R.~K. Shukla and X.~Zhong.
+\newblock Derivation of high-order compact finite difference schemes for
+  non-uniform grid using polynomial interpolation.
+\newblock {\em J. Comput. Phys.}, 204:404--429, 2005.
+
+\bibitem{shukla_2007}
+R.~K. Shukla, M.~Tatineni, and X.~Zhong.
+\newblock Very high-order compact finite difference schemes on non-uniform
+  grids for incompressible {N}avier-{S}tokes equations.
+\newblock {\em J. Comput. Phys.}, 224:1064--1094, 2007.
+
+\bibitem{zhao_2006}
+F.~Zhao.
+\newblock Weighting {F}actors for {S}ingle-{S}tep {T}rapezoidal {M}ethod.
+\newblock {\em J. Heat Transf.-Trans. ASME}, 128:409--412, 2006.
+
+\bibitem{fedkiw_1999}
+R.~Fedkiw, T.~Aslam, B.~Merriman, and S.~Osher.
+\newblock A {N}on-{O}scillatory {E}ulerian {A}pproach to {I}nterfaces in
+  {M}ultimaterial {F}lows ({T}he {G}host {F}luid {M}ethod).
+\newblock {\em J. Comput. Phys.}, 152:457--492, 1999.
+
+\bibitem{osher_fedkiw_book}
+S.~Osher and R.~Fedkiw.
+\newblock {\em Level Set Methods and Dynamic Implicit Surfaces}.
+\newblock Springer, New York, NY, 2003.
+
+\bibitem{iserles_book}
+A.~Iserles.
+\newblock {\em A First Course in the Numerical Analysis of Differential
+  Equations}.
+\newblock Cambridge University Press, Cambridge, UK, 1996.
+
+\bibitem{shampine_2005}
+L.F. Shampine.
+\newblock {E}rror {E}stimation and {C}ontrol for {ODE}s.
+\newblock {\em J. Sci. Comput.}, 25:3--16, 2005.
+
+\bibitem{gko_book}
+B.~Gustafsson, H.-O. Kreiss, and J.~Oliger.
+\newblock {\em Time Dependent Problems and Difference Methods}.
+\newblock John Wiley \& Sons, Inc., New York, NY, 1995.
+
+\bibitem{leveque_book_1992}
+R.~J. LeVeque.
+\newblock {\em Numerical Methods for Conservation Laws}.
+\newblock Birkh\"auser Verlag, Berlin, 1992.
+
+\bibitem{leveque_book_2002}
+R.~J. LeVeque.
+\newblock {\em Finite Volume Methods for Hyperbolic Problems}.
+\newblock Cambridge University Press, Cambridge, UK, 2002.
+
+\bibitem{whitham_book}
+G.~B. Whitham.
+\newblock {\em Linear and Nonlinear Waves}.
+\newblock John Wiley \& Sons, Inc., New York, NY, 1999.
+
+\bibitem{chan_1984}
+T.~Chan.
+\newblock Stability {A}nalysis of {F}inite {D}ifference {S}chemes for the
+  {A}vection-{D}iffusion {E}quation.
+\newblock {\em SIAM J. NUMER. ANAL.}, 21:272--284, 1984.
+
+\bibitem{greer_2006}
+J.~B. Greer, A.~L. Bertozzi, and G.~Sapiro.
+\newblock Fourth order partial differential equations on general geometries.
+\newblock {\em J. Comput. Phys.}, 216:216--246, 2006.
+
+\bibitem{patra_2005}
+M.~Patra and M.~Karttunen.
+\newblock {S}tencils with {I}sotropic {D}iscretization {E}rror for
+  {D}ifferential {O}perators.
+\newblock {\em Numer. Meth. for Part. Diff. Eq.}, 22:936--953, 2005.
+
+\bibitem{trefethen_spectral_book}
+L.~N. Trefethen.
+\newblock {\em Spectral Methods in MATLAB}.
+\newblock SIAM, Philadelphia, PA, 2000.
+
+\bibitem{boyd_spectral_book}
+J.~P. Boyd.
+\newblock {\em Chebyshev and Fourier Spectral Methods}.
+\newblock Dover Publications, Inc., Mineola, NY, 2nd edition, 2001.
+
+\bibitem{gupta_1998}
+M.~M. Gupta and J.~Kouatchou.
+\newblock {S}ymbolic {D}erivation of {F}inite {D}ifference {A}pproximations for
+  the {T}hree-{D}imensional {P}oisson {E}quation.
+\newblock {\em Numer. Meth. for Part. Diff. Eq.}, 14:593--606, 1998.
+
+\end{thebibliography}
 

### 2008-11-16 src/scalar_PDEs/1d/fourth_order_parabolic/genFigs_4thOrderParabolicEqn1d.m
OTSPDE: - changed figures for fourth-order parabolic equation   - changed data points used to fit the OTS performance graph   - changed colors to make them darker when printed in grayscale

@@ -268,8 +268,8 @@ order_CN_4thOrder = -P_CN_4thOrder(1);
 P_CN_2ndOrder = polyfit(log(grid_sizes(1:9)),log(err_CN_2ndOrder(1:9)),1);
 order_CN_2ndOrder = -P_CN_2ndOrder(1);
 
-P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(1:4)), ...
-                             log(comp_time_FE_OTS(1:4)),1);
+P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(1:5)), ...
+                             log(comp_time_FE_OTS(1:5)),1);
 comp_time_exp_FE_OTS = P_comp_time_FE_OTS(1);
 P_comp_time_FE = polyfit(log(err_FE(1:5)),log(comp_time_FE(1:5)),1);
 comp_time_exp_FE = P_comp_time_FE(1);
@@ -310,9 +310,9 @@ annotation(f, 'arrow', [.25 .37], [.55 .7], ...
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_CN_4thOrder(1)+P_CN_4thOrder(2)),'k');
 hold on;
-plot(grid_sizes,err_CN_4thOrder, 'cd', ...
+plot(grid_sizes,err_CN_4thOrder, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','c');
+     'MarkerFaceColor','m');
 order_str = sprintf('4th-Order Crank-Nicholson \nOrder = %1.1f', ...
                     order_CN_4thOrder);
 text(12,7e-8,order_str);
@@ -344,9 +344,9 @@ figure(2); clf;
 N_plot = [10 100000];
 loglog(N_plot,exp(log(N_plot)*P_CN_4thOrder(1)+P_CN_4thOrder(2)),'k');
 hold on;
-plot(grid_sizes,err_CN_4thOrder, 'cd', ...
+plot(grid_sizes,err_CN_4thOrder, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','c');
+     'MarkerFaceColor','m');
 order_str = sprintf('4th-Order\nCrank-Nicholson');
 text(30,1e-7,order_str);
 
@@ -379,7 +379,7 @@ loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerFaceColor','b');
[…]

### 2008-11-16 src/scalar_PDEs/1d/fourth_order_parabolic/genPresoFigs_4thOrderParabolicEqn1d.m

@@ -253,7 +253,7 @@ end
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Analyze Results
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-P_FE_OTS = polyfit(log(grid_sizes(1:5)),log(err_FE_OTS(1:5)),1);
+P_FE_OTS = polyfit(log(grid_sizes(1:4)),log(err_FE_OTS(1:4)),1);
 order_FE_OTS = -P_FE_OTS(1);
 P_FE = polyfit(log(grid_sizes(1:5)),log(err_FE(1:5)),1);
 order_FE = -P_FE(1);
@@ -267,8 +267,8 @@ P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(1:5)), ...
 comp_time_exp_FE_OTS = P_comp_time_FE_OTS(1);
 P_comp_time_FE = polyfit(log(err_FE(1:5)),log(comp_time_FE(1:5)),1);
 comp_time_exp_FE = P_comp_time_FE(1);
-P_comp_time_CN_4thOrder = polyfit(log(err_CN_4thOrder(4:5)), ...
-                                  log(comp_time_CN_4thOrder(4:5)),1);
+P_comp_time_CN_4thOrder = polyfit(log(err_CN_4thOrder(4:6)), ...
+                                  log(comp_time_CN_4thOrder(4:6)),1);
 comp_time_exp_CN_4thOrder = P_comp_time_CN_4thOrder(1);
 P_comp_time_CN_2ndOrder = polyfit(log(err_CN_2ndOrder(7:9)), ...
                                   log(comp_time_CN_2ndOrder(7:9)),1);
@@ -304,9 +304,9 @@ annotation(f, 'arrow', [.25 .37], [.55 .7], ...
 N_plot = [10 1000];
 loglog(N_plot,exp(log(N_plot)*P_CN_4thOrder(1)+P_CN_4thOrder(2)),'k');
 hold on;
-plot(grid_sizes,err_CN_4thOrder, 'cd', ...
+plot(grid_sizes,err_CN_4thOrder, 'md', ...
      'MarkerSize',14, ...
-     'MarkerFaceColor','c');
+     'MarkerFaceColor','m');
 order_str = sprintf('4th-Order Crank-Nicholson \nOrder = %1.1f', ...
                     order_CN_4thOrder);
 text(12,7e-8,order_str);
[…]

### 2008-11-16 papers/OTSPDE/OTSPDE.tex
OTSPDE: - minor revisions to OTSPDE paper.

@@
 % Submission draft: 2008/06/08, Vitamin D, Inc.
-% Resubmission draft: 2008/11/15, Vitamin D, Inc.
+% Resubmission draft: 2008/11/16, Vitamin D, Inc.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 method.  We demonstrate the utility of optimal time step (OTS) selection on a 
-wide array of classical PDEs in one and more space dimensions to illustrate 
-its applicability to linear and semilinear PDEs on both regular and irregular 
-domains.  
+wide array of classical linear and semilinear PDEs in one and more space 
+dimensions on both regular and irregular domains.  
 \end{abstract}
@@
 \eeq
-Thus, because $p > r$ and $q > s$, we see that using an optimal time step 
+Because $p > r$ and $q > s$, we see that using an optimal time step 
 boosts the overall accuracy of the finite difference scheme.
@@
 \subsubsection*{Stability Considerations}
-Thus far, we have not considered the issue of numerical stability of finite
+We have not yet considered the issue of numerical stability of finite
 difference schemes when the time step is set to $\dto$.  For explicit time 
@@
 $O(N) = O \left( 1/\dx^d \right)$ matrix inversion algorithm 
-and that the optimal time step for the forward Euler scheme is $\dto = \dx^2/6$.
+and that the optimal time step for the forward Euler scheme is 
+$\dto = \dx^2/6D$.
 }
@@
 \eeq
-Thus, the optimal time step is $\dto = 7 \dx^4/120\kappa$ and the correction
-term is 
+Therefore, the optimal time step is $\dto = 7 \dx^4/120\kappa$ and the 
+correction term is 
 \beq
@@
 which vanishes when the optimal time step and grid spacing ratio are used.
-Thus, when (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used in
-conjunction with the optimal time step and grid spacing ratio, we obtain
+In other words, when (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used 
+in conjunction with the optimal time step and grid spacing ratio, we obtain
 a finite difference scheme that introduces no truncation error at each time 
@@
 Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also shows that the 
-scaling of the computational time with the error is in good agreement with 
-theoretical estimates (see Table~\ref{tab:comp_perf_vs_dim}).  As with the
+scaling of the computational time with the error is in good agreement with the
+theoretical estimates given in Table~\ref{tab:comp_perf_vs_dim}.  As with the
 other examples, the forward Euler scheme with an optimal time step is much 
@@
 All of the examples we have considered so far share an important feature: they 
-are all defined on simple, rectangular domains.  Thus, any of several methods 
+are all defined on simple, rectangular domains.  Any of several methods 
 could have been used to obtain solutions with similar or even higher accuracy.
@@
 shows that the practical error is 
-$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
+$O(\dx^2) + O(\dt) = O(\dx^2)$.  Therefore, the accuracy of the 
 method is completely controlled by the grid spacing; the temporal error can 

### 2008-11-18 papers/OTSPDE/submissions/200811-JCP/cover_letter.tex
OTSPDE: - Checked in submission information for OTSPDE paper.

\documentclass[12pt]{letter}

% page settings
\textwidth 6.5in
\textheight 9in
\oddsidemargin -0.05in
\evensidemargin -0.05in
\topmargin -0.85in

\begin{document}

\address{}
\signature{Kevin T. Chu}

\begin{letter}
{
% NORMALLY ADDRESS GOES HERE
}

\opening{Dear J. Computational Physics Editor/Referee,}

My article titled \emph{Boosting the Accuracy of Finite Difference Schemes 
via Optimal Time Step Selection} discusses a novel technique for boosting 
the order of accuracy of finite difference schemes for time dependent 
PDEs by choosing an optimal, \emph{fixed} time step.  As discussed in the 
paper, a careful choice of time step can dramatically increase the order of 
accuracy for simple finite difference schemes and make it possible to achieve 
high-order accuracy using only formally low-order finite difference stencils 
and time integration schemes.  For example, optimal time step selection can be 
used to 
\begin{itemize}
\item obtain infinite-order accuracy for the advection equation using 
      a first-order upwind discretization of the gradient operator and
      forward Euler time integration;
\item achieve fourth-order accuracy for the diffusion equation using only 
      formally second-order finite difference stencils and forward Euler 
      time integration. 
\end{itemize}
This paper describes the method for calculating the optimal time step and 
explains the observed boost in accuracy using straightforward numerical 
analysis arguments.  The technique of optimal time step selection is 
applicable to linear and semilinear PDEs in any number of space dimensions 
and both regular and irregular domains.  Examples of each of these types of 
problems are discussed and analyzed in the paper.

Thank you for your review of the paper.  I look forward to hearing your 
comments and suggestions for improvement of the manuscript.

\closing{Sincerely,}

\end{letter}

\end{document}

### 2008-12-15 papers/OTSPDE/OTSPDE.tex
OTSPDE: - added keywords to manuscript
OTSPDE - fixed typo in OTSPDE paper
OTSPDE: - fixed references for shapes of markers in figure.

@@
 \noindent \rule{6.3in}{1pt}
+
+% KEYWORDS
+% optimal time step, finite difference schemes, 
+% high-order accurate numerical methods, time dependent PDEs
 
@@
              - 2 \nu \tu^n_j\frac{\partial^3 \tu^n}{\px^3} 
-             - 4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2} 
+             - 4 \nu \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2} 
              + 2 \tu^n_j \left(\frac{\partial \tu^n}{\px} \right)^2
@@
 forward Euler with suboptimal time step and no correction term (squares),
-Crank-Nicholson with a second-order stencil for the bilaplacian (diamonds), 
 and Crank-Nicholson with a fourth-order stencil for the bilaplacian and
-an $O(\dx^2)$ time step (triangles).
+an $O(\dx^2)$ time step (diamonds), and Crank-Nicholson with a second-order 
+stencil for the bilaplacian (triangles).
 Both forward Euler schemes use a fourth-order, central finite difference 

### 2008-12-15 papers/OTS-MaxwellsEqns/genPDFFigures.sh
OTSPDE: - checked in outline of OTS paper for Maxwell's equation - updated potential journals file

@@ -1,0 +1,17 @@
+#!/bin/sh
+#
+# This script generates PDF figures from the eps versions of figures.
+#
+# Kevin T. Chu
+#
+
+FIGURES_DIR="figures"
+
+# convert figures
+cd ${FIGURES_DIR}
+for FIG in `ls *.eps`; do
+  echo -n "Converting ${FIG} ..."
+  epstopdf ${FIG}
+  echo "done"
+done
+

### 2008-12-15 papers/OTS-MaxwellsEqns/getFigures.sh

@@ -1,0 +1,40 @@
+#!/bin/sh
+#
+# This script collects the figures from the src directory and places them
+# in the figures directory.
+#
+# Kevin T. Chu
+# February 2008
+#
+
+FIGURES_DIR="figures"
+
+SRC_DIR="../../src"
+EXAMPLE_DIR_LIST="1d/advection_diffusion_eqn                            \
+                  1d/burgers_eqn/single_hump                            \
+                  1d/diffusion_eqn                                      \
+                  1d/fourth_order_parabolic                             \
+                  2d/advection_eqn                                      \
+                  2d/diffusion_eqn/square_domain                        \
+                  2d/diffusion_eqn/irregular_domain/circle_domain       \
+                  2d/diffusion_eqn/irregular_domain/starfish_domain     \
+                 "
+
+# create directory for figures if it doesn't already exist
+if [ ! -d "${FIGURES_DIR}" ]; then
+  mkdir -p ${FIGURES_DIR}
+fi
+
+# move figures 
+for DIR in ${EXAMPLE_DIR_LIST}; do
+  FIGURES=`ls ${SRC_DIR}/${DIR}/figures/*.eps 2> /dev/null` 
+  if [ ! -z "${FIGURES}" ]; then
+    echo -n "Moving figures from "
+    echo "${SRC_DIR}/${DIR}/figures/ to ${FIGURES_DIR} ..."
+    mv ${SRC_DIR}/${DIR}/figures/*.eps ${FIGURES_DIR}/
+  else
+    echo -n "Figures missing from "
+    echo "${SRC_DIR}/${DIR}/figures/"
+  fi
+done
+

### 2009-01-25 src/scalar_PDEs/1d/wave_eqn/genFigs_WaveEqn1d_Src.m
OTSPDE: - replaced quadratic interpolation with cubic interpolation to get fourth-order   accuracy - replaced fourth-order accurate first step with fifth-order first step - minor changes to the figures generated for genFigs_WaveEqn1d_Src.m

@@ -227,9 +227,9 @@ plot(grid_sizes,err_KPY_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
 order_str = sprintf('KPY-OTS\nOrder = %1.1f', order_KPY_OTS);
-text(200,1e-9,order_str);
+text(200,4e-12,order_str);
 
-N_plot = [100 10000];
+N_plot = [95 10000];
 loglog(N_plot,exp(log(N_plot)*P_KPY(1)+P_KPY(2)),'k');
 hold on;
 plot(grid_sizes,err_KPY, 'rs', ...
@@ -238,8 +238,8 @@ plot(grid_sizes,err_KPY, 'rs', ...
 order_str = sprintf('KPY\nOrder = %1.1f', order_KPY);
 text(2000,2e-4,order_str);
 
-axis([100 10000 1e-12 1e-2]);
-set(gca, 'ytick', 10.^[-12:2:-2]);
+axis([100 10000 1e-14 1e-2]);
+set(gca, 'ytick', 10.^[-14:2:-2]);
 xlabel('N');
 ylabel('L^\infty Error');
 filename = sprintf('wave_eqn_1d_src_error_vs_N.%s', print_suffix);

### 2009-01-25 src/scalar_PDEs/1d/wave_eqn/solveWaveEqn1dKPY_OTS.m

@@ -31,16 +31,6 @@
 % time step of dt = dx/c (unit CFL condition) is used and a correction term
 % is added when a source term is present.
 %
-% NOTES:
-% - Because quadratic interpolation is used when on the final time step, 
-%   the numerical solution is only guaranteed to be third-order accurate. 
-%   This situation occurs when t_final is not close to a multiple of the 
-%   optimal time step dt = dx/c.
-% - When the final time step is close to a multiple of the optimal time
-%   step dt = dx/c, then the numerical solution is fourth-order accurate 
-%   when there is no source term and third-order accurate when a source
-%   term is present.
-%
 % USAGE:
 %   function [u, u_exact, x] = solveWaveEqn1dKPY_OTS(c, ...
 %                                                    use_source_term, ...
@@ -128,17 +118,32 @@ else
   u_t = -0.5*pi*sin(pi*x);
 end
 
-% use second-order forward Euler for first time step
+% use fifth-order Taylor-series expansion for first time step
+% NOTE: the numerical approximation to u_tt provides part of 
+%       contribution to the O(dt^4) term
 if (use_source_term > 0)
   f = 0.125/c*pi^2*sin(pi*x);
   f_t = -0.03125/c*pi^3*cos(pi*x);
+  f_tt = -0.0078125/c*pi^4*sin(pi*x);
 else
   f = 0;
   f_t = 0;
 end
-u_tt = c^2*L*u + f;
-u_next = u + dt*u_t + 0.5*dt^2*u_tt ...
-       + 1/6*dt^3*(c^2*L*u_t+f_t);
+
+u_tt = c^2*L*u + f;  % second-order accurate approx to u_tt.
+                     % the leading-order spatial error partially cancels
+                     % out the O(dt^4) term in the Taylor series expansion
+
+if (use_source_term > 0)
+  % NOTE: 1/24*dt^4*u_tttt term is accounted for by the leading-order
+  %       error in the numerical approximation for u_tt
+  u_next = u + dt*u_t + 0.5*dt^2*u_tt + 1/6*dt^3*(c^2*L*u_t+f_t) ...
+         + 1/24*dt^4*(c^2*L*f + f_tt); 
+else
+  % NOTE: 1/24*dt^4*u_tttt term is accounted for by the leading-order
+  %       error in the numerical approximation for u_tt
[…]

### 2009-01-25 src/variable_coefs/wave_eqn/genFigs_WaveEqn1d.m
OTSPDE: - replaced fourth-order Taylor series expansions with fifth-order expansion   for first time step in OTS code - replaced quadratic interpolation with cubic interpolation in OTS code - updated figures in genFigs_WaveEqn1d.m - minor updates to comments throughout code

@@ -35,7 +35,7 @@ if ~exist(fig_dir, 'dir')
 end
 
 % set flag for loading data from saved files (instead of recomputing solution)
-use_saved_data = 1;
+use_saved_data = 0;
 data_dir = 'data-var_coef_wave_eqn_1d';
 if ~exist(data_dir, 'dir')
   mkdir(data_dir);
@@ -271,7 +271,7 @@ plot(grid_sizes,err_KPY_Transformed_OTS, 'cd', ...
      'MarkerFaceColor','c');
 order_str = sprintf('KYP-OptMesh-OTS,\nKPY-Transformed-OTS\nOrder = %1.1f', ...
   order_KPY_Transformed_OTS);
-text(200,2e-8,order_str);
+text(150,1e-10,order_str);
 
 N_plot = [100 10000];
 loglog(N_plot, ...
@@ -297,10 +297,10 @@ plot(grid_sizes,err_KPY, 'rs', ...
      'MarkerFaceColor','r');
 order_str = sprintf('KPY, KPY-OptMesh,\nKPY-Transformed\nOrder = %1.1f', ...
                     order_KPY);
-text(800,5e-3,order_str);
+text(1200,3e-3,order_str);
 
-axis([100 10000 1e-9 1e-1]);
-set(gca, 'ytick', 10.^[-9:2:-1]);
+axis([100 10000 1e-12 1e-1]);
+set(gca, 'ytick', 10.^[-12:2:-1]);
 xlabel('N');
 ylabel('L^\infty Error');
 filename = sprintf('var_coef_wave_eqn_1d_error_vs_N.%s', print_suffix);

### 2009-01-25 src/variable_coefs/wave_eqn/solveWaveEqn1dKPY.m

@@ -101,7 +101,7 @@ u_t = -pi*(2*cos(2*pi*x) + 3*sin(3*pi*x));
 % compute wave speed
 c = 1 + 0.5*sin(0.5*pi*x);
 
-% use second-order forward Euler for first time step
+% use second-order Taylor series expansion for first time step
 f =  0.25*pi^2*sin(0.5*pi*x) ...
   .* ( 16*sin(2*pi*x) + 36*cos(3*pi*x) ...
      + 4*sin(0.5*pi*x).*sin(2*pi*x)   ...

### 2009-01-25 src/variable_coefs/wave_eqn/solveWaveEqn1dKPY_Transformed.m

@@ -135,7 +135,7 @@ t = 0.0;
 u   = sin(2*pi*x) + cos(3*pi*x);
 u_t = -pi*(2*cos(2*pi*x) + 3*sin(3*pi*x));
 
-% use second-order forward Euler for first time step
+% use second-order Taylor series expansion for first time step
 f =  0.25*pi^2*sin(0.5*pi*x) ...
   .* ( 16*sin(2*pi*x) + 36*cos(3*pi*x) ...
      + 4*sin(0.5*pi*x).*sin(2*pi*x)   ...

### 2009-01-25 src/variable_coefs/wave_eqn/solveWaveEqn1dKPY_Transformed_OTS.m

@@ -138,7 +138,7 @@ t = 0.0;
 u   = sin(2*pi*x) + cos(3*pi*x);
 u_t = -pi*(2*cos(2*pi*x) + 3*sin(3*pi*x));
 
-% use second-order forward Euler for first time step
+% use fifth-order Taylor series expansion for first time step
 f =  0.25*pi^2*sin(0.5*pi*x) ...
   .* ( 16*sin(2*pi*x) + 36*cos(3*pi*x) ...
      + 4*sin(0.5*pi*x).*sin(2*pi*x)   ...
@@ -147,9 +147,16 @@ f_t = -0.25*pi^3*sin(0.5*pi*x) ...
     .* ( 32*cos(2*pi*x) + 108*sin(3*pi*x) ...
        + 8*sin(0.5*pi*x).*cos(2*pi*x) ...
        + 27*sin(0.5*pi*x).*sin(3*pi*x) );
-u_tt = 4*c_bar^2*(L*u) - 2*c_bar*c_prime.*(G*u) + f;
+f_tt = -0.25*pi^4*sin(0.5*pi*x) ...
+     .*( 64*sin(2*pi*x) + 324*cos(3*pi*x) ...
+       + 16*sin(0.5*pi*x).*sin(2*pi*x) ...
+       + 81*sin(0.5*pi*x).*cos(3*pi*x) );
+
+c = 1 + 0.5*sin(0.5*pi*x); % variable wave speed in original domain
+u_tt = -c.^2.*(4*pi^2*sin(2*pi*x) + 9*pi^2*cos(3*pi*x)) + f;
 u_ttt = 4*c_bar^2*(L*u_t) -2*c_bar*c_prime.*(G*u_t) + f_t;
-u_next = u + dt*u_t + 0.5*dt^2*u_tt + 1/6*dt^3*u_ttt;
+u_tttt = 4*c_bar^2*(L*u_tt) - 2*c_bar*c_prime.*(G*u_tt) + f_tt;
+u_next = u + dt*u_t + 0.5*dt^2*u_tt + 1/6*dt^3*u_ttt + 1/24*dt^4*u_tttt;
 
 % update u_prev and u
 u_prev = u;
@@ -219,18 +226,20 @@ while (t < t_final)
 
   % update u_prev and u 
   if (t < t_final)
+    u_minus_two = u_prev;
     u_prev = u;
     u = u_next;
 
   else
-    % if we have overstepped t_final, use quadratic interpolation to obtain
[…]

### 2009-01-25 src/variable_coefs/wave_eqn/solveWaveEqn1dKPY_VarMesh.m

@@ -127,7 +127,7 @@ u_t = -pi*(2*cos(2*pi*x) + 3*sin(3*pi*x));
 % compute wave speed
 c = 1 + 0.5*sin(0.5*pi*x);
 
-% use second-order forward Euler for first time step
+% use second-order Taylor series expansion for first time step
 f =  0.25*pi^2*sin(0.5*pi*x) ...
   .* ( 16*sin(2*pi*x) + 36*cos(3*pi*x) ...
      + 4*sin(0.5*pi*x).*sin(2*pi*x)   ...

### 2009-01-25 src/variable_coefs/wave_eqn/solveWaveEqn1dKPY_VarMesh_OTS.m

@@ -142,7 +142,7 @@ u_t = -pi*(2*cos(2*pi*x) + 3*sin(3*pi*x));
 % compute wave speed
 c = 1 + 0.5*sin(0.5*pi*x);
 
-% use second-order forward Euler for first time step
+% use fifth-order Taylor series expansion for first time step
 f =  0.25*pi^2*sin(0.5*pi*x) ...
   .* ( 16*sin(2*pi*x) + 36*cos(3*pi*x) ...
      + 4*sin(0.5*pi*x).*sin(2*pi*x)   ...
@@ -151,9 +151,15 @@ f_t = -0.25*pi^3*sin(0.5*pi*x) ...
     .* ( 32*cos(2*pi*x) + 108*sin(3*pi*x) ...
        + 8*sin(0.5*pi*x).*cos(2*pi*x) ...
        + 27*sin(0.5*pi*x).*sin(3*pi*x) );
-u_tt = c.^2.*(L*u) + f;
+f_tt = -0.25*pi^4*sin(0.5*pi*x) ...
+     .*( 64*sin(2*pi*x) + 324*cos(3*pi*x) ...
+       + 16*sin(0.5*pi*x).*sin(2*pi*x) ...
+       + 81*sin(0.5*pi*x).*cos(3*pi*x) );
+
+u_tt = -c.^2.*(4*pi^2*sin(2*pi*x) + 9*pi^2*cos(3*pi*x)) + f;
 u_ttt = c.^2.*(L*u_t) + f_t;
-u_next = u + dt*u_t + 0.5*dt^2*u_tt + 1/6*dt^3*u_ttt;
+u_tttt = c.^2.*(L*u_tt) + f_tt;
+u_next = u + dt*u_t + 0.5*dt^2*u_tt + 1/6*dt^3*u_ttt + 1/24*dt^4*u_tttt;
 
 % update u_prev and u
 u_prev = u;
@@ -218,20 +224,22 @@ while (t < t_final)
   % update time
   t = t + dt;
 
-  % update u_prev and u 
+  % update u_minus_two, u_prev and u 
   if (t < t_final)
+    u_minus_two = u_prev;
     u_prev = u;
     u = u_next;
 
   else
-    % if we have overstepped t_final, use quadratic interpolation to obtain
+    % if we have overstepped t_final, use cubic interpolation to obtain
     % u(t_final).
-    % NOTE:  solution is third-order accurate because error in quadratic 
-    %        interpolation is O(dt^3).
[…]

### 2009-01-25 papers/OTSPDE-VarCoefs/OTSPDE-VarCoefs.tex
OTSPDE: - updated noted for variable-coefficient case (includes new results showing   how to achieve fourth-order accuracy when there is a source term present)

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% Manuscript started: 2008/09/0?
+% Manuscript started: 2008/10/12
 % First completed draft: 2008/??/??, Vitamin D, Inc.
@@
 \section*{Introduction}
+Place holder...
 \cite{zoppou1999, kreiss2002, guidotti2006}
+
+Key points
+\begin{itemize}
+\item OTS applicable to problems with variable-coefficient equations by
+      (1) optimally selecting the computational mesh or 
+      (2) solving the equations on a transformed domain.  
+      These two approaches are essentially the same.
+\item Demonstration of OTS on discretization on a non-single step time 
+      integration schemes.
+\item Demonstration of boost in order of accuracy for variable-coefficient 
+      wave equation. 
+\item Demonstration of boost in order of accuracy for variable-coefficient 
+      diffusion equation. 
+\item Demonstration of boost in order of accuracy for 2D problem. Schrodinger
+      equation?
+\end{itemize}
+
+\section{Optimal Time Step Selection for Variable-Coefficient PDEs}
+Place holder...
+
+\section{Wave Equation}
+Second-order wave equation with source term:
+\bea
+    u_{tt} - c^2 u_{xx} = f.
+    \label{eq:wave_eqn}
+\eea
+on the domain $-1 < x < 1$ subject to periodic boundary conditions and
+initial conditions
+\bea
+  u(x,0)   &=& 2 \exp \left( -10 (\sin(0.5 \pi x))^2 \right) \nonumber \\
+  u_t(x,0) &=& -0.5 \pi \sin(\pi x) + g(x)
+\eea
+
+\subsection{OTS Selection for Kriess-Petersson-Ystr\"om Discretization}
+In~\cite{kreiss2002}, Kriess, Petersson, and Ystr\"om analyzed the following
+direct discretization of the second-order wave equation without conversion 
+to a first-order system of equations:
+\bea
+    \frac{u^{n+1}_i - 2 u^n_i + u^{n-1}_i}{\dt^2}
+  = c^2 \left( \frac{u^{n}_{i+1} - 2 u^n_i + u^n_{i-1}}{\dx^2} \right)
+  + f
+  \label{eq:wave_eqn_KPY}
+\eea
+This scheme is second-order accurate in space and time and stable when 
+$\dt < ??$.
+
+To compute the optimal time step and correction terms for this scheme, we 
+begin by deriving the evolution equation for the error, 
+$e^n_i = u^n_i - \tu^n_i$, where $\tu^n_i \equiv \tu(i\dx, n\dt)$.  Employing
+Taylor series expansions, it is straightforward to show that the true solution
+satisfies:
+\bea
+  \frac{\tu^{n+1}_i - 2 \tu^n_i + \tu^{n-1}_i}{\dt^2}
+    &-& \frac{\dt^2}{12} \tu_{tttt} + O(\dt^4)
+  = \nonumber \\
+  & & c^2 \left( \frac{\tu^{n}_{i+1} - 2 \tu^n_i + \tu^n_{i-1}}{\dx^2} 
+             -\frac{\dx^2}{12} \tu_{xxxx} + O(\dx^4)
+        \right)
+  + f
+\eea
+Therefore, the error evolution equation is 
+\bea
+  \frac{e^{n+1}_i - 2 e^n_i + e^{n-1}_i}{\dt^2}
+    &+& \frac{\dt^2}{12} \tu_{tttt} + O(\dt^4)
+  = \nonumber \\
+  & & c^2 \left( \frac{e^{n}_{i+1} - 2 e^n_i + e^n_{i-1}}{\dx^2} 
+             +\frac{\dx^2}{12} \tu_{xxxx} + O(\dx^4)
+        \right)
+  \label{eq:wave_eqn_error_eqn}
+\eea
+Using the wave equation (\ref{eq:wave_eqn}) to express $\tu_{tttt}$ in terms
+of spatial derivatives, we find that
+\beq
+  \tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt}
+\eeq
+Substituting this result into (\ref{eq:wave_eqn_error_eqn}), we find that
+\bea
+  \frac{e^{n+1}_i - 2 e^n_i + e^{n-1}_i}{\dt^2}
+    &+& \frac{\dt^2}{12} \left(c^4\tu_{xxxx} + c^2 f_{xx} + f_{tt} \right) 
+    + O(\dt^4)
+  = \nonumber \\
+  & & c^2 \left( \frac{e^{n}_{i+1} - 2 e^n_i + e^n_{i-1}}{\dx^2} 
+             +\frac{\dx^2}{12} \tu_{xxxx} + O(\dx^4)
+        \right)
+\eea
+Thus, we see that we can eliminate the leading-order term in the discretization
+error by choosing $\dt = \dx/c$ and adding the correction term 
+\beq
+  \frac{\dt^2}{12} \left(c^2 f_{xx} + f_{tt} \right) 
+\eeq
+to the right-hand side of (\ref{eq:wave_eqn_KPY}).
+
+With this choice of time step and the correction term, the local truncation
+error is $O(\dt^6) + O(\dx^4 \dt^2)$.  Using the heuristic for single-step 
+calculations leads to the conclusion that the global error should be 
+approximately $1/\dt^2$ times the local error, which leads to a global error
+of $O(\dx^4) = O(\dt^4)$.  
+Figures~\ref{fig:wave_eqn_no_src_error} and~\ref{fig:wave_eqn_src_error} 
+demonstrate that the expected accuracy is achieved.
+
+Note that to get higher-order accuracy, we also need to use a fifth-order
+accurate first step for the scheme.  For the wave equation, this is 
+straightforward to achieve by using a fifth-order Taylor series expansion
+in time to obtain $u^1_i$.  The reason that the first time step needs to be 
+fifth-order accurate can be seen by observing that the solution of the 
+difference equation for the normal modes (von Neumann analysis):
+\bea
+  \frac{e^{n+1} - 2 e^{n} + e^{n-1}}{\dt^2} = \dt^2 \lambda e^{n}
+  \label{eq:error_eqn_normal_mode}
+\eea
+is~\cite{carrier_pearson_book}
+\bea
+  e^n = e^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-} 
+      + e^0 \kappa_+ \kappa_- 
+        \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-} 
+  \label{eq:error_eqn_normal_mode_soln}
+\eea
+where $\kappa_\pm$, the roots of the characteristic equation for 
+(\ref{eq:error_eqn_normal_mode}), are given by~\cite{kreiss2002}
+\beq
+  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2 
+             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
+\eeq
+From the solution to the error equation for the normal modes 
+(\ref{eq:error_eqn_normal_mode_soln}), we see that the solution loses
+one temporal order of accuracy compared to the initial conditions because
+the denominator of both terms is $O(\dt)$.
+In other words, the initial errors $e^0$ and $e^1$ must be $O(\dt^5)$ in 
+order to for the solution to be $O(\dt^4)$.
+
+\begin{figure}[thb]
+\begin{center}
+\scalebox{0.35}{\includegraphics{figures/wave_eqn_1d_no_src_error_vs_N}}
+\caption{$L^\infty$ error as a function of number of grid points for the
+KPY discretization of the second-order wave equation without source term.
+}
+\label{fig:wave_eqn_no_src_error}
+\end{center}
+\end{figure}
+
+\begin{figure}[thb]
+\begin{center}
+\scalebox{0.35}{\includegraphics{figures/wave_eqn_1d_src_error_vs_N}}
+\caption{$L^\infty$ error as a function of number of grid points for the
+KPY discretization of the second-order wave equation with source term.
+}
+\label{fig:wave_eqn_src_error}
+\end{center}
+\end{figure}
+
+\subsection{OTS Selection for Variable Coefficient Wave Equation}
+We consider the variable-coefficient second-order wave equation:
+\beq
+    u_{tt} - c(x)^2 u_{xx} = f,
+    \label{eq:var_coef_wave_eqn}
+\eeq
+on the domain $-1 < x < 1$.
+
+The formulation of optimal time step selection presented in~\cite{chu2009}
+is restricted to problems which are constant coefficient in the leading-order 
+temporal and spatial derivatives.  To extend apply the technique to the
+variable-coefficient wave equation, we can take one of two approaches:
+(1) transformation of the domain so that the leading-order equations are 
+constant coefficient or (2) use of an optimal computational grid with
+variable grid spacing when solving the equation on the original domain.
+In fact, these two approaches are closely related because the optimal 
+computational grid is the mapping of a uniform grid in the transformed domain
+onto the original domain.
+
+Following~\cite{guidotti_2006}, we use the following simple change of variables
+to transform (\ref{eq:var_coef_wave_eqn}) into a constant coefficient 
+wave equation:
+\beq
+  y = \Phi(x) = \cbar \int_{-1}^x \frac{1}{c(\xi)} d\xi,
+\eeq
+where 
+\beq
+  \cbar = \left( \int_{-1}^1 \frac{1}{c(\xi)} d\xi \right)^{-1}.
+\eeq
+On the transformed domain, (\ref{eq:var_coef_wave_eqn}) becomes
+\beq
+  u_{tt} = 4\cbar^2 u_{yy} - 2 \cbar c' u_y + f,
+\eeq
+where $c' = dc/dx$.
+
+\subsection{OTS on Transformed Domain}
+If we use second-order central differences for both the Laplacian and 
+gradient terms, the optimal time step remains the same as for the constant
+coefficient wave equation $\dt = \dx/2 \cbar$.  The correction term is also
+straightforward (but tedious) to calculate:
+\bea
+  \frac{\dt^2}{12} 
+  \left(
+    \begin{array}{l}
+        -16 \cbar^3 (c')_y u_{yy}
+        -8 \cbar^3 (c')_{yy} u_{y}
+        +4 \cbar^2 (c')^2 u_{yy} 
+        \\
+        +\ 4 \cbar^2 c' (c')_y u_{y}
+        +4 \cbar^2 f_{yy}
+        -2 \cbar c' f_y
+        + f_{tt}
+    \end{array}
+  \right)
+\eea
+With the optimal time step and this correction term, we obtain a third-order
+accurate scheme for the variable-coefficient wave equation.  
+Figure~\ref{fig:var_coef_wave_eqn_src_error} shows the results.
+
+\begin{figure}[thb]
+\begin{center}
+\scalebox{0.35}{\includegraphics{figures/var_coef_wave_eqn_1d_error_vs_N}}
+\caption{$L^\infty$ error as a function of number of grid points for the
+KPY discretization of the variable-coefficient second-order wave equation 
+with source term.
+}
+\label{fig:var_coef_wave_eqn_src_error}
+\end{center}
+\end{figure}
+
+\subsection{OTS with Optimal Mesh}
+While applying OTS to the equation on the transformed domain yields a
+boost in order of accuracy, the lower-order spatial derivative and the
+correction term are tedious to deal with.  A convenient alternative is to 
+work on the original domain but optimally choose the mesh so that use of
+an optimal time step still leads to fortuitous cancellation of the 
+leading-order error.
+
+We calculate the location of the grid points in the optimal mesh by mapping
+a uniform mesh in the $y$-domain back to the $x$-domain.  The finite 
+difference stencil for the Laplacian on this mesh is natural generalization 
+of Laplaacian to irregularly spaced grid points:
+\bea
+  (u_{xx})_i &\approx&
+    \frac{1}{(x_{i+1}+x_i)/2 - (x_i+x_{i-1})/2}
+    \left( \frac{u_{i+1}-u_i}{x_{i+1} - x_i}
+         - \frac{u_{i}-u_{i-1}}{x_{i} - x_{i-1}}
+    \right)
+  \nonumber \\
+  &=&
+    \frac{2}{x_{i+1}-x_{i-1}}
+    \left( \frac{u_{i+1}-u_i}{x_{i+1} - x_i}
+         - \frac{u_{i}-u_{i-1}}{x_{i} - x_{i-1}}
+    \right)
+\eea
+
+Using our analysis of the problem in the transformed domain, we find that the
+optimal time step for the scheme on the variable mesh is given by
+$\dt_{opt} = \dy/2\cbar$.  The correction term can be derived by first
+computing $u_{tttt}$ on the original domain
+\bea
+  u_{tttt} = 
+    c^4 u_{xxxx}
+  + c^2 \left( (c^2)_xx u_{xx} + 2 (c^2)_x u_{xxx} + f_{xx} \right)
+  + f_{tt}
+\eea
+Because the odd derivatives do not show up in the correction term on
+the transformed domain (they cancelled themselves out), the odd derivative
+is not part of the correction term for the scheme on the variable mesh.  
+Thus, the correction term is
+\beq
+  \frac{\dt^2}{12} \left[ c^2 (c^2)_xx u_{xx} + c^2 f_{xx} + f_{tt} \right]
+\eeq
+
+Using the optimal time step and this correction term yields a third-order 
+accurate scheme for the variable-coefficient wave equation.  The error on
+the variable mesh is almost identical to the error obtained when solving
+the equation on the transformed domain (see 
+Figure~\ref{fig:var_coef_wave_eqn_src_error}).
+
+*** KTC: 
+I haven't thought of a nice elegant way to derive the optimal time
+step and correction term yet... 
+***
+
+\section{Diffusion Equation}
+Not done yet...
 
@@
 \section*{Acknowledgments}
-The author gratefully acknowledges the support of Vitamin D, Inc.
+K.T. Chu gratefully acknowledges the support of Vitamin D, Inc.
 and the Institute for High-Performance Computing (IHPC) in Singapore. 
-The author would like to thank ??
+The authors would like to thank ??
 for helpful suggestions on the manuscript.

### 2009-03-02 papers/OTSPDE-WCE2009/OTSPDE-WCE2009.tex
OTSPDE: - added draft of paper for World Congress on Engineering 2009

\documentclass[twocolumn]{article} % default is 10 pt
\usepackage{graphicx} % needed for including graphics e.g. EPS, PS

\long\def\comment#1{}

% uncomment if don't want page numbers
\pagestyle{empty}

%set dimensions of columns, gap between columns, and paragraph indent

\setlength{\textheight}{9.5in} \setlength{\columnsep}{1pc}
\setlength{\textwidth}{7in}

\setlength{\topmargin}{0in}

\setlength{\headheight}{0.0in}
\setlength{\headsep}{0.0in}
\setlength{\oddsidemargin}{-0.35in}
\setlength{\evensidemargin}{-0.35in}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.12in}
\makeatletter
\def\@normalsize{\@setsize\normalsize{10pt}\xpt\@xpt
\abovedisplayskip 10pt plus2pt minus5pt\belowdisplayskip
\abovedisplayskip \abovedisplayshortskip \z@
plus3pt\belowdisplayshortskip 6pt plus3pt
minus3pt\let\@listi\@listI}

%need an 11 pt font size for subsection and abstract headings
\def\subsize{\@setsize\subsize{12pt}\xipt\@xipt}
%make section titles bold and 12 point, 2 blank lines before, 1 after
\def\section{\@startsection {section}{1}{\z@}{1.0ex plus
1ex minus .2ex}{.2ex plus .2ex}{\large\bf}}
%make subsection titles bold and 11 point, 1 blank line before, 1 after
\def\subsection{\@startsection
   {subsection}{2}{\z@}{.2ex plus 1ex} {.2ex plus .2ex}{\subsize\bf}}
\makeatother

% GENERAL MACROS
\newcommand{\bc}{\begin{center}}
\newcommand{\ec}{\end{center}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}

\def\ie{{\it i.e. }}
\def\eg{{\it e.g. }}
\def\etal{{\it et. al. }}

% PDE MACROS
\def\tu{\tilde{u}}
\def\abar{\bar{a}}
\def\cbar{\bar{c}}
\def\dbar{\bar{d}}

% NUMERICS MACROS
\def\dt{\Delta t}
\def\dx{\Delta x}
\def\dy{\Delta y}
\def\pt{\partial t}
\def\px{\partial x}
\def\dto{\Delta t_{opt}}
\def\he{\hat{e}}

\begin{document}
\bibliographystyle{unsrt}

% don't want date printed
\date{}

% >>>>>>>>>>>>>>>>>>>>>>>  Put your title here <<<<<<<<<<<<<<<<<<<<<<<<
% make title bold and 14 pt font (Latex default is non-bold, 16pt)
\title{\huge \bf {Using Optimal Time Step Selection to \\ 
       Boost the Accuracy of FD Schemes for \\
       Variable Coefficient PDEs }}

% >>>>>>>>>>>>>>>>>>>>>>> Author's Name, Thanks or Affliation <<<<<<<<
\author{Kevin T. Chu\thanks{
  K.T. Chu gratefully acknowledges the support of Vitamin D, Inc.
  and the Institute for High-Performance Computing (IHPC) in Singapore. 
  Vitamin D, Inc., Menlo Park, CA 94025 \& 
  Institute of High Performance Computing, A*STAR, Singapore, Singapore.
  Email: ktchu@serendipityresearch.org.
  Manuscript submitted 2008 March 2. 
 } 
 \ and
 James V. Lambers\thanks{Department of Energy Resources Engineering, 
 Stanford University, Stanford, CA 94305.  Email: lambers@stanford.edu.}
}

\maketitle
\thispagestyle{empty}

% >>>>>>>>>>>>>>>>>>>>>>>>> Keywords and Abstract <<<<<<<<<<<<<<<<<<<<<
{\hspace{1pc} {\it{\small Abstract}}{\bf{\small---We extend the technique 
of optimal time step (OTS) selection for finite difference (FD) schemes of time 
dependent PDEs to PDEs where the leading-order spatial derivative term has a
spatially varying coefficient.  The basic approach involves identifying a 
transformation of the domain that eliminates the spatial dependence of the 
coefficient for the leading-order term.  This change of variables is then used 
to define an optimal computational grid for the FD scheme on the 
\emph{original} domain.   By using both the optimal grid and OTS, we are
able to boost the order of accuracy above what would be expected from a formal
analysis of the FD scheme.  We illustrate the utility of our method by 
applying it to variable coefficient wave and diffusion equations.  In 
addition, we demonstrate the viability of OTS selection to the two-step 
Kreiss-Petersson-Ystr\"om discretization of the wave equation.

\em Keywords: optimal time step selection, optimal grid selection, finite 
    difference schemes, time dependent PDEs, variable coefficient PDEs}}
 }

% >>>>>>>>>>>>>>>>>>>>>> START OF PAPER <<<<<<<<<<<<<<<<<<<<<<<<<<<<<<
\section{Introduction}
\label{Introduction}
Optimal time step (OTS) selection is a surprisingly simple and effective way to 
boost the order of accuracy of finite difference (FD) schemes for 
time-dependent partial differential equations~\cite{chu_otspde}.  The basic
principle underlying OTS selection is that a careful choice of time step (and 
the addition of a few correction terms) can boost the \emph{order} of 
accuracy for formally low-order finite difference schemes.  For instance, it 
is possible to obtain a fourth-order accurate solution of the 1D diffusion 
equation using only the standard second-order central difference approximation 
and forward Euler time integration.  As demonstrated in~\cite{chu_otspde}, OTS 
selection works for linear and semilinear PDEs in any number of space 
dimensions on both regular and irregular domains.

Unfortunately, a limitation of the existing formulation of OTS selection is 
the requirement that the leading-order spatial derivative of the PDE be 
constant coefficient.  In this paper, we remove this limitation and extend OTS 
selection to be applicable to general variable coefficient semilinear PDEs.  
Our approach is to identify a transformation of the domain which 
eliminates the spatial dependence of the coefficient on the leading-order 
spatial derivative.  Using this transformation, we define an optimal 
computational grid which can be used in conjunction with OTS selection to 
boost the order of accuracy of formally low-order FD schemes for the PDE.  
Optimal grid selection is an important extension of the philosophy that the 
accuracy of numerical methods can be boosted via optimization of the 
parameters used to compute the solution.

This paper is organized as follows.  First, we briefly review OTS selection 
for PDEs with a constant coefficient leading-order spatial derivative term.  
Next, we show how to combine optimal grid and optimal time step selection to 
boost the accuracy of finite difference schemes for variable coefficient PDEs.  
We then demonstrate the use of variable coefficient OTS selection on the 
variable coefficient wave and heat equations.  Our analysis of the wave is 
equation is of particular note because it is \emph{not} based on a single step 
time integration scheme.  We conclude with a summary of our main results
and thoughts on future directions for research.

\section{Review of OTS Selection for Constant Coefficient PDEs
  \label{sec:ots_const_coef_pdes}}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather, it is a technique for enhancing the 
performance of existing FD schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  There are two fundamental ideas 
underlying OTS selection.  First, a judicious choice of time step can be 
used to eliminate the leading-order terms in the error.  Second, the PDE 
provides valuable insight into the discretization errors for FD schemes.  
Combining these two simple ideas often yields an optimal time step which 
can be used to \emph{boost} the order of accuracy of a given FD scheme 
above what would be expected from a formal analysis of the scheme.

To illustrate the basic theory behind OTS selection, let us consider the 
constant coefficient diffusion equation:
\beq
  u_{t} = D u_{xx} + f(x,t),
 \label{eq:diffusion_eqn_1d}
\eeq
where $D$ is the diffusion constant and $f(x,t)$ is a source term.  Perhaps
the simplest FD scheme for this equation uses forward Euler time integration 
and the standard second-order central difference approximation for the 
Laplacian:
\beq
  u^{n+1}_j = u^{n}_j
    + \dt
      \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right]
           + f^n_j \right).
 \label{eq:diffusion_eqn_1d_FD_scheme}
\eeq
This scheme is formally first-order in time and second-order in 
space.  The stability constraint $\dt \le \dx^2/2D$ implies that the scheme is
$O(\dx^2)$ accurate overall.

OTS selection is based on a detailed analysis of the leading-order errors in 
FD schemes.  So, we begin by deriving the truncation equation for the scheme 
(\ref{eq:diffusion_eqn_1d_FD_scheme}).  Employing Taylor series expansions and 
the PDE (\ref{eq:diffusion_eqn_1d}), we find that the true solution satisfies
\bea
  \tu^{n+1}_j &=& \tu^{n}_j
  + \dt \left( D \tu_{xx} + f \right)
  \nonumber \\
  &+& \frac{\dt^2}{2} \left( D^2 \tu_{xxxx} + D f_{xx} + f_t \right)
  + O \left( \dt^3 \right) \ \ 
  \label{eq:diffusion_eqn_1d_time_err}
\eea
and that the central difference approximation for the Laplacian satisfies
\beq
  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
  \tu_{xx} + \frac{\dx^2}{12} \tu_{xxxx}
  + O(\dx^4)
  \label{eq:diffusion_eqn_1d_space_err}
\eeq
Therefore, the truncation error for (\ref{eq:diffusion_eqn_1d_FD_scheme})
is given by
\bea
  & & \tu_{xxxx}
  \left[ \frac{\dx^2}{12} - \frac{D \dt}{2} \right] (D \dt)
  - \frac{\dt^2}{2} \left( D f_{xx}
  + f_t \right)
  \nonumber \\
  & & + \ O(\dt \dx^4) + O(\dt^3).
  \label{eq:diffusion_eqn_1d_trunc_err}
\eea
From this expression, we see that choosing the time step to be
$\dt = \dx^2/6D$ and adding the correction term 
\beq
  \frac{\dt^2}{2} \left( D f_{xx} + f_t \right)
\eeq
eliminates the leading-order truncation error.  Using the heuristic
argument that the global error is equal to the local error divided by 
$\dt$~~\cite{gko_book}, we find that the numerical solution is 
$O(\dx^4) + O(\dt^2) = O(\dx^4)$ accurate -- the FD scheme has been
boosted from second- to fourth-order accuracy.

As shown in~\cite{chu_otspde}, this general procedure can be used to boost 
the accuracy of FD schemes for any semilinear PDE that has a constant 
coefficient leading-order spatial derivative.  It is important to emphasize 
that variable coefficients and nonlinearities in the lower-order spatial 
derivative terms do not pose a problem for OTS selection.

\section{OTS Selection for Variable Coefficient PDEs
  \label{sec:ots_var_coef_pdes}}
Transformation of the spatial domain is a natural way to extend OTS selection 
to semilinear PDEs that have variable coefficient leading-order spatial 
derivatives.  Consider a semilinear PDE of the form:
\bea
  \frac{\partial u}{\partial t} = a^n(x) \frac{\partial^n u}{\partial x^n}
 + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}}, \ldots,
           \frac{\partial u}{\partial x}, u \right)
 + f(x,t)
\label{eq:semilinear_PDE}
\eea
on the domain $0 < x < 1$.  Using the change of variables\footnote{This
transformation is a generalization of the change of variables proposed
in~\cite{guidotti2006} for the wave equation.}
\beq
  y = \Phi(x) = \abar \int_0^x \frac{1}{a(\xi)} d\xi
  \label{eq:change_of_variables}
\eeq
with 
\beq
  \abar \equiv \left( \int_0^1 \frac{1}{a(\xi)} d\xi \right)^{-1},
  \label{eq:abar}
\eeq
we can completely eliminate the spatial dependence of the coefficient on the 
leading-order spatial derivative (at the expense of adding 
variable coefficient lower-order spatial derivative terms).  The PDE 
(\ref{eq:semilinear_PDE}) on the transformed domain has the form
\bea
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
 + \tilde{F} \left( \frac{\partial^{n-1} u}{\partial y^{n-1}},
           \ldots,
           \frac{\partial u}{\partial y}, u \right)
 + f(y,t),
\label{eq:semilinear_PDE_transformed}
\eea
where $\tilde{F}$ includes any lower-order terms introduced by the process of 
changing variables.

The change of variables (\ref{eq:change_of_variables}) allows us to apply 
OTS selection in one of two ways.  First, we could simply apply OTS selection
to the transformed PDE on the transformed domain.  An important alternative,
however, is to solve the PDE on the \emph{original} domain using a variable 
spaced grid generated by mapping a uniform grid from the transformed to the 
original domain.  Amazingly, OTS selection can be used to boost the accuracy 
of natural choices of finite difference schemes defined on this 
\emph{optimal computational grid}.

Mathematically, these two approaches are equivalent.  The former approach is 
numerically simpler but requires the solution of a more complicated PDE.
The latter approach places more complexity on the construction of the FD
scheme but requires no modification to the PDE.  When using OTS selection,
the latter approach is generally superior because it dramatically simplifies 
the derivation of the correction terms.

\subsection{Construction of FD Schemes for Optimal Grid}
When constructing FD schemes for the optimal grid on the original domain, it 
is important to define them as divided differences that use grid points 
corresponding to the ones used on the transformed domain.  Doing so ensures 
that the FD scheme for the PDE on the original domain is compatible with the 
FD scheme for the transformed PDE on the transformed domain (\ie the FD scheme 
on the variable spaced grid automatically encapsulates the required 
transformation of the PDE that result from the change of variables).  This 
compatibility preserves the fortuitous cancellation of errors on the optimal 
grid even though it is not uniform.

For instance, the generalization of the second-order central difference 
operator for the 1D Laplacian on a variable spaced grid is
\bea
  (u_{xx})_i &\approx&
    \frac{1}{(x_{i+1}+x_i)/2 - (x_i+x_{i-1})/2}
  \nonumber \\
    & & \times \left( \frac{u_{i+1}-u_i}{x_{i+1} - x_i}
         - \frac{u_{i}-u_{i-1}}{x_{i} - x_{i-1}}
    \right)
  \nonumber \\
  &=&
    \frac{2}{x_{i+1}-x_{i-1}}
    \left( \frac{u_{i+1}-u_i}{x_{i+1} - x_i}
         - \frac{u_{i}-u_{i-1}}{x_{i} - x_{i-1}}
    \right), \ \ \ \ \ 
  \label{eq:laplacian_variable_grid}
\eea
which is essentially a difference of fluxes computed at 
$(x_{i+1}+x_i)/2$ and $(x_i+x_{i-1})/2$.

\subsection{Derivation of OTS and Correction Terms for Optimal Grid
  \label{sec:derivation_ots_corr_terms} }
To derive the optimal time step, we apply the procedure presented 
in~\cite{chu_otspde} to the PDE in the transformed domain 
(\ref{eq:semilinear_PDE_transformed}).  However, rather than analyzing the 
full PDE, we focus only on the simplified PDE containing only the time 
derivative term and leading-order spatial derivative:
\beq
  \frac{\partial u}{\partial t} = {\abar}^n \frac{\partial^n u}{\partial y^n}
  \label{eq:simplified_PDE_transformed}.
\eeq
When a uniform grid and standard finite difference schemes are used to solve
this PDE on the transformed domain, we find that the optimal time step is given
by $\dt_{opt} = \alpha (\dy / \abar)^n$, where $\alpha$ depends on the 
particular FD scheme used approximate (\ref{eq:simplified_PDE_transformed}).
Because the FD schemes on the original and transformed domains are chosen 
to be compatible, $\dt_{opt}$ is also the optimal time step to use when 
solving the PDE on the optimal grid.

A more formal way to arrive at the same conclusion is to observe that 
$\dy \approx \abar/a \dx$, which follows from (\ref{eq:change_of_variables}).  
Using this approximation to convert $\dx$'s to $\dy$'s in the FD scheme 
defined on the optimal grid yields an approximation to the leading-order 
truncation error for the spatial derivative of the form:
\beq
  \frac{\alpha \abar^n \dt \dy^n}{2} \frac{\partial^{2n} \tu}{\partial y^{2n}}
  = \frac{\alpha a(x)^n \dt \dx^n}{2} \frac{\partial^{2n} \tu}{\partial x^{2n}} 
  + \ldots,
\eeq
where the factor of $1/2$ is included for convenience.  This observation makes 
it possible to cancel out the leading-order temporal error 
\beq
 \frac{\dt^2}{2} \frac{\partial^2 \tu}{\partial t^2} = 
  \frac{a(x)^{2n} \dt^2}{2} \frac{\partial^{2n} \tu}{\partial x^{2n}}.
\eeq
by choosing $\dt = \alpha \dx^n/a(x)^n = \alpha (\dy/\abar)^n$.

To boost the accuracy, we also need to derive the appropriate correction terms.
When working on the transformed domain, we simply follow the procedure 
described in~\cite{chu_otspde}.  However, when working on the original domain 
using the optimal mesh, the analysis is more complicated and generally depends
on the details of the FD scheme and PDE.  The main challenge is deciding which 
terms of the leading-order temporal error are not canceled out by the choice 
of time step.

Fortunately, a simple heuristic that can be used to derive the correction 
terms.  When deriving $u_{tt}$, each spatial derivative term in the PDE (on 
the original domain) gives rise to several potential correction terms.  We 
conjecture that of these candidates, we need only retain those terms involving 
spatial derivatives of $u$ with order not exceeding the order the originating 
spatial derivative term in the PDE.  As a concrete example, consider a term in 
the PDE involving $u_x$.  The corresponding correction terms should only 
involve $u_x$ and $u$ but not $u_{xx}$ or higher derivatives.  While this 
assertion has yet to be proven, it seems to hold for the PDEs we have examined.

\section{Application to Model Problems}

\subsection{Variable Coefficient Wave Equation}
In this section, we apply OTS and optimal grid selection to the
variable coefficient wave equation.  Because we choose a FD scheme based on 
the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
equation~\cite{kreiss2002}, we begin by analyzing OTS selection for the of 
the constant coefficient wave equation and demonstrate its applicability for 
this multistep method.

\subsubsection*{OTS Selection for the KPY Discretization of the 
  Constant-Coefficient Wave Equation}
In~\cite{kreiss2002}, Kreiss, Petersson, and Ystr\"om analyzed the following
direct discretization of the second-order wave equation, 
$u_{tt} - c^2 u_{xx} = f$,
without conversion to a first-order system of equations:
\bea
    \frac{u^{n+1}_i - 2 u^n_i + u^{n-1}_i}{\dt^2}
  = c^2 \left( \frac{u^{n}_{i+1} - 2 u^n_i + u^n_{i-1}}{\dx^2} \right)
  + f
  \label{eq:wave_eqn_KPY}
\eea
and showed that it is second-order accurate in space and time with a 
stability constraint of the form $\dt = O(\dx)$.

Since the stability constraint implies that $\dt$ and $\dx$ are not truly
independent numerical parameters, it is beneficial to use OTS selection to 
optimally choose the time step as a function of the grid 
spacing\footnote{We choose to optimize $\dt$ as a function of $\dx$ (as 
opposed to $\dx$ as a function of $\dt$) because time stepping is the most 
commonly used approach for numerically solving time-dependent PDEs.}.

To compute the optimal time step and correction terms for this scheme, we 
follow the procedure outlined in~\cite{chu_otspde} and 
Section~\ref{sec:ots_const_coef_pdes}.  We begin by deriving the truncation 
error for the scheme.  Employing Taylor series expansions, it is 
straightforward to show that the true solution satisfies:
\bea
  & &\frac{\tu^{n+1}_i - 2 \tu^n_i + \tu^{n-1}_i}{\dt^2}
    - \frac{\dt^2}{12} \tu_{tttt} + O(\dt^4)
  = \nonumber \\
  & & c^2 \left( \frac{\tu^{n}_{i+1} - 2 \tu^n_i + \tu^n_{i-1}}{\dx^2} 
  -\frac{\dx^2}{12} \tu_{xxxx} + O(\dx^4) \right)
  \nonumber \\
  & & +f
\eea
Combining this result with the observation that 
\beq
  \tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt}
\eeq
we find that the truncation error is given by
\bea
    & & c^2 \frac{\dx^2}{12} \tu_{xxxx} 
    - \frac{\dt^2}{12} \left(c^4\tu_{xxxx} + c^2 f_{xx} + f_{tt} \right) 
    \nonumber \\
    & & + \ O(\dx^4) + O(\dt^4) 
\eea
Thus, we can eliminate the leading-order term in the discretization
error by choosing $\dt = \dx/c$ and adding the correction term 
\beq
  \frac{\dt^2}{12} \left(c^2 f_{xx} + f_{tt} \right) 
\eeq
to the right-hand side of (\ref{eq:wave_eqn_KPY}).

With this choice of time step and correction term, the local truncation
error is $O(\dt^6) + O(\dx^4 \dt^2)$.  Using the heuristic for two-step
methods that the global error should be approximately $1/\dt^2$ times the 
local error leads to a global error of $O(\dx^4) = O(\dt^4)$.  That is, using
OTS selection boosts the order of accuracy of the KPY scheme from second-
to fourth-order.  Figure~\ref{fig:wave_eqn_1d_error} demonstrates that the 
expected accuracy is indeed achieved by applying OTS to the KPY scheme.

\begin{figure}[thb]
\begin{center}
\scalebox{0.30}{\includegraphics{figures/wave_eqn_1d_src_error_vs_N}}
\caption{$L^\infty$ error as a function of number of grid points for the
KPY discretization of the second-order wave equation on with OTS selection 
(blue circles) and without OTS selection (red squares).
}
\label{fig:wave_eqn_1d_error}
\end{center}
\end{figure}

\paragraph{High Accuracy Required for First Time Step}
For the KPY discretization of the wave equation, it is critical that a 
high-order accurate method is used to compute the first time step because 
the error introduced during this step affects the global error at all times.
Specifically, for the error of the KPY scheme to be $O(\dx^p)$, the numerical
solution at the first time step must be accurate to $O(\dx^{p+1})$.  In other 
words the error introduced by the first time step must be at least one order
of accuracy higher than desired for the overall solution.

We can understand this need for higher-order accuracy of the first time step 
by solving the difference equation for the normal modes of the error:
\bea
  \frac{\he^{n+1} - 2 \he^{n} + \he^{n-1}}
       {\dt^2} = \dt^2 \lambda \he^{n}
  \label{eq:error_eqn_normal_mode},
\eea
where $\he$ is the coefficient of an arbitrary normal mode of the spatial
operator for the error $e \equiv u - \tu$.  The solution of this equation 
is~\cite{carrier_pearson_book}
\bea
  \he^n = \he^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-} 
      + \he^0 \kappa_+ \kappa_- 
        \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-} 
  \label{eq:error_eqn_normal_mode_soln}
\eea
where $\kappa_\pm$, the roots of the characteristic equation for 
(\ref{eq:error_eqn_normal_mode}), are given by~\cite{kreiss2002}
\beq
  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2 
             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
\eeq
Notice that the denominator of both terms in 
(\ref{eq:error_eqn_normal_mode_soln}) is $O(\dt)$.  As a result, the 
global error is always at least one temporal order of accuracy less than the 
error in the initial conditions.  Therefore, the initial errors $e^0$ and 
$e^1$ must be $O(\dt^{p+1})$ in order to for the solution to be $O(\dt^p)$. 
For KPY without OTS selection, this observation indicates that the first
time step should be at least third-order accurate, which is exactly the way
that Kreiss \etal chose to compute the first time step in~\cite{kreiss2002}.
Because the OTS selection transforms KPY into a fourth-order accurate scheme, 
we require that the first time step is at least fifth-order accurate.

It is straightforward to construct a fifth-order approximation for the first 
time step by using a fifth-order Taylor series expansion in time:
\bea
  u^1 &=& u^0 + \dt u_t^0 + \frac{\dt^2}{2} u_{tt}^0
      + \frac{\dt^3}{6} u_{ttt}^0 
   + \frac{\dt^4}{24} u_{tttt}^0
  \nonumber \\
  &=& u^0 + \dt u_t^0 + \frac{\dt^2}{2} \left(c^2 u_{xx}^0 + f\right)
  \nonumber \\
  & & + \frac{\dt^3}{6} \left(c^2 \left(u_t\right)_{xx}^0 + f_t\right)
  \nonumber \\
  & & + \ \frac{\dt^4}{24} \left(c^4 u_{xxxx}^0 + c^2 f_{xx} + f_{tt}\right)
  \label{eq:KPY_fifth_order_first_step}
\eea
Note that sufficiently high-order accurate finite difference stencils may be 
used to compute some of the higher order terms in this expression.  However,
when advancing the solution using the optimal time step, special care must be 
exercised if finite differences are used to compute the derivatives in 
(\ref{eq:KPY_fifth_order_first_step}) because higher-order terms may be 
automatically (implicitly) included by lower-order terms in the expansion.

\subsubsection*{OTS Selection for Variable Coefficient Wave Equation}
We now consider the variable coefficient second-order wave equation:
\beq
    u_{tt} - c(x)^2 u_{xx} = f,
    \label{eq:var_coef_wave_eqn}
\eeq
on the domain $0 < x < 1$.  Using the change of variables defined in 
(\ref{eq:change_of_variables}) and (\ref{eq:abar}), we can transform 
(\ref{eq:var_coef_wave_eqn}) into the constant coefficient wave equation:
\beq
  u_{tt} = \cbar^2 u_{yy} - \cbar c' u_y + f,
\eeq
where $c' = dc/dx$.

\paragraph*{OTS on Transformed Domain}
On the transformed domain, it is straightforward to apply OTS selection 
because the leading-order spatial derivative has a constant coefficient.  
If we use second-order central differences for both the Laplacian and gradient 
terms (on a uniform grid in the transformed domain), the optimal time step is 
$\dt_{opt} = \dx/\cbar$ and the correction term, which is calculated by 
identifying the terms in $\left(\dt^2 \tu_{tttt}/12\right)$ that are not 
eliminated by the use of $\dt_{opt}$, is given by:
\bea
  \frac{\dt^2}{12} 
  \left(
    \begin{array}{l}
        -2 \cbar^3 (c')_y u_{yy}
        -\cbar^3 (c')_{yy} u_{y}
        + \cbar^2 (c')^2 u_{yy} 
        \\
        +\ \cbar^2 c' (c')_y u_{y}
        + \cbar^2 f_{yy}
        - \cbar c' f_y
        + f_{tt}
    \end{array}
  \right)
\eea
Using the optimal time step and the correction term, we obtain a fourth-order
accurate KPY scheme for the variable coefficient wave equation 
(see Figure~\ref{fig:var_coef_wave_eqn_1d_error}).  As mentioned earlier,
a fifth-order accurate first time step is required to achieve this level of
accuracy.

\begin{figure}[thb]
\begin{center}
\scalebox{0.30}{\includegraphics{figures/var_coef_wave_eqn_1d_error_vs_N}}
\caption{$L^\infty$ error as a function of number of grid points for various
KPY discretizations of the variable coefficient second-order wave equation:
KPY using a uniform grid on the original domain (red squares),
KPY using a uniform grid on the transformed domain (green triangles),
KPY using the optimal grid without OTS selection the original domain 
(magenta triangles),
KPY using a uniform grid with OTS selection on the transformed domain 
(cyan diamonds),
KPY using the optimal grid with OTS selection on the original domain 
(blue circles).  Note that the green triangles overlap the magenta triangles 
and the cyan diamonds overlap the blue circles. 
}
\label{fig:var_coef_wave_eqn_1d_error}
\end{center}
\end{figure}

\paragraph*{OTS with Optimal Grid}
While applying OTS to the equation on the transformed domain yields the 
desired boost in order of accuracy, the lower-order spatial derivative and 
the correction term are tedious to deal with.  It is more convenient to 
work on the original domain but optimally choose the grid so that use of
an optimal time step still leads to fortuitous cancellation of the 
leading-order error.

Following the procedure outlined in Section~\ref{sec:ots_var_coef_pdes}, we 
define the location of the grid points in the optimal grid by mapping
a uniform grid in the $y$-domain back to the $x$-domain and use 
the generalized finite difference approximation for the Laplacian 
(\ref{eq:laplacian_variable_grid}).  The optimal time step for the optimal
grid is given by $\dt_{opt} = \dy/\cbar$, where $\dy$ is the grid spacing 
in the $y$-domain.

The required correction term can be derived by first computing $u_{tttt}$ on 
the original domain
\bea
  u_{tttt} =
    c^4 u_{xxxx} + 2 c^2 (c^2)_x u_{xxx} 
    + c^2 (c^2)_{xx} u_{xx} + c^2 f_{xx} + f_{tt}
\eea
Using the heuristic mentioned in Section~\ref{sec:derivation_ots_corr_terms},
the terms involving $u_{xxxx}$ and $u_{xxx}$ are eliminated through the use 
of the optimal time step.  Thus, the correction term is
\beq
  \frac{\dt^2}{12} \left[ c^2 (c^2)_{xx} u_{xx} + c^2 f_{xx} + f_{tt} \right]
\eeq

As shown in Figure~\ref{fig:var_coef_wave_eqn_1d_error}, by using the optimal 
grid, optimal time step, and the correction terms, we are able to obtain a 
fourth-order accurate scheme for the variable coefficient wave equation on
the original domain.  The error on the variable-spaced grid is almost 
identical to the error obtained when solving the equation on the transformed 
domain.

% DIFFUSION EQUATION FIGURE PLACED HERE TO SPREAD OUT FIGURES
\begin{figure}[thb]
\begin{center}
\scalebox{0.30}{\includegraphics{figures/var_coef_diffusion_eqn_1d_error_vs_N}}
\caption{$L^\infty$ error as a function of number of grid points with 
OTS/Optimal Grid selection (blue circles) and on a uniform grid without 
OTS selection (red squares).
}
\label{fig:var_coef_diffusion_eqn_1d_error}
\end{center}
\end{figure}

\subsection{Application to Variable Coefficient Diffusion Equation}
In this section, we apply OTS and optimal grid selection to the variable
coefficient diffusion equation:
\beq
  u_t = (D(x) u)_{xx} + f(x,t),
  \label{eq:var_coef_diffusion_eqn}
\eeq
on the domain $0 < x < 1$.  Note that we have expressed diffusion in a 
spatially inhomogeneous medium using the Fokker-Planck diffusivity 
law~\cite{vanmilligen2005}, which has been shown to often be more physically 
correct than Fick's law.

Following the procedure outlined in Section~\ref{sec:ots_var_coef_pdes}, the
change of variables used to define the optimal grid is given by
$y = \dbar\int_0^x D(\xi)^{-1/2} d\xi$ where
$\dbar = \left(\int_0^1 D(\xi)^{-1/2} d\xi \right) ^ {-1}$.
Since the optimal time step for the constant coefficient diffusion equation 
is given by $\dto = \dx^2/6D$, the optimal time step for the variable
coefficient diffusion equation is given by $\dto = \left(\dy/\dbar\right)^2/6$.
The correction term for this scheme is 
\bea
  \frac{\dt^2}{2} \left[
  \begin{array}{c}
  6 D D_{xx} u_{xx} + 4 D D_{xxx} u_x D D_{xxxx} u + D f_{xx} \\
  + \ 6 D_x D_{xx} u_x + 2 D_x D_{xxx} u + 2 D_x f_x \\
  + \ D_{xx} u_t + f_t
  \end{array}
  \right]
\eea
which is derived by computing $u_{tt}$ and using the heuristic described in 
Section~\ref{sec:derivation_ots_corr_terms} to exclude the following terms:
\beq
  D^2u_{xxxx} \ , \ 4 D D_x u_{xxx} \ , \ 
  2 D_x D u_{xxx} \ , \ 6 \left(D_x\right)^2 u_{xx}.
\eeq
Combining the optimal grid, optimal time step, and the correction term, we are 
able to obtain fourth-order solutions for (\ref{eq:var_coef_diffusion_eqn})
using only a formally second-order FD scheme (see 
Figure~\ref{fig:var_coef_diffusion_eqn_1d_error}).  
Figure~\ref{fig:var_coef_diffusion_eqn_1d_solns} shows a comparison of 
solutions computed with and without the optimal grid and OTS.

\begin{figure}[tbh]
\begin{center}
\scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_OTS_OptGrid_soln}} \ 
\scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_soln}}
\caption{Comparison of numerical solutions for the variable coefficient
diffusion equation~(\ref{eq:var_coef_diffusion_eqn}) computed using 
the optimal grid and OTS selection (left) and a uniform grid without OTS
selection (right).  Both figures compare the analytical solution (solid line)
with the numerical solution (circles) computed using $25$ grid points.
}
\label{fig:var_coef_diffusion_eqn_1d_solns}
\end{center}
\end{figure}

\section{Summary and Conclusions}
In this paper, we have extended the philosophy of optimizing the parameters
of FD schemes (\eg the time step size) for accuracy to the selection of the 
computational grid.  In particular, we have presented a method to boost the 
order of accuracy of FD schemes for PDEs where the leading-order has a 
variable coefficient by using an optimal time step and an optimal grid.  Using 
straightforward numerical analysis arguments, we have described the basic 
procedure for constructing the optimal grid, computing the optimal time step, 
and deriving the required correction terms.  To demonstrate the power of these 
techniques, we have applied them to the variable coefficient wave equation and 
the Fokker-Planck form of the diffusion equation in an inhomogeneous medium.  
Our analysis of the KPY scheme (used to solve the wave equation) also serves 
as an example of OTS selection applied to multistep time integration schemes.

\subsection{Future Work}
An important direction for future work is the generalization of the current 
work to PDEs in higher spatial dimensions.  We believe that the use of a 
change of variables may still be viable.  However, we anticipate several
challenges.  For instance, we may need to allow for a change in the shape 
of domain and the possibility the optimal grid may no longer be orthogonal.  
Both of these issues complicate the construction of the finite difference
scheme on the original domain.

\bibliography{OTSPDE-WCE2009}

\end{document}

% >>>>>>>>>>>>>>>>>>>>>> Bibliography <<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<<
% A label is give for each bibitem.  Your paper references this label

%\begin{thebibliography}{99}
%
%% >>>>>>>>> Book examples <<<<<<<<<
%\bibitem{CarpenterBOOK} Carpenter, R.H.S., {\it Movements of the Eyes},
% 2nd Edition, Pion Publishing, 1988.
%
%\bibitem{FranklinBOOK} Franklin, G.F., Powel, J.D., Workman, M.L.,
%{\it Digital Control of Dynamic Systems}, Second Edition,
%Addison-Wesley, 1990.
%
%% >>>>>>>>> Conference Proceedings Example <<<<<<<<<
%\bibitem{OhICRA1998} Oh, P.Y., Allen, P.K., ``Design a Partitioned
% Visual Feedback Controller,'' {\it IEEE Int Conf Robotics
% \& Automation}, Leuven, Belgium, pp. 1360-1365 5/98
%
%% >>>>>>>>> Journal Example <<<<<<<<<<<<<<<<<<<<<<<<
%\bibitem{OhTRA2001} Oh, P.Y., Allen, P.K., ``Visual Servoing
% by Partitioning Degrees of Freedom,'' {\it IEEE Trans on
% Robotics \& Automation}, V17, N1, pp. 1-17, 2/01
%
%\end{thebibliography}
\end{document}

### 2009-03-02 papers/OTSPDE-WCE2009/getFigures.sh

@@ -1,0 +1,34 @@
+#!/bin/bash
+#
+# This script collects the figures from the src directory and places them
+# in the figures directory.
+#
+# Kevin T. Chu
+# February 2008
+#
+
+FIGURES_DIR="figures"
+
+SRC_DIR="../../src"
+EXAMPLE_DIR_LIST="scalar_PDEs/1d/wave_eqn                     \
+                  variable_coefs/wave_eqn                     \
+                  variable_coefs/diffusion_eqn"
+
+# create directory for figures if it doesn't already exist
+if [ ! -d "${FIGURES_DIR}" ]; then
+  mkdir -p ${FIGURES_DIR}
+fi
+
+# move figures 
+for DIR in ${EXAMPLE_DIR_LIST}; do
+  FIGURES=`ls ${SRC_DIR}/${DIR}/figures/*.eps 2> /dev/null` 
+  if [ ! -z "${FIGURES}" ]; then
+    echo -n "Moving figures from "
+    echo "${SRC_DIR}/${DIR}/figures/ to ${FIGURES_DIR} ..."
+    mv ${SRC_DIR}/${DIR}/figures/*.eps ${FIGURES_DIR}/
+  else
+    echo -n "Figures missing from "
+    echo "${SRC_DIR}/${DIR}/figures/"
+  fi
+done
+

### 2009-03-20 papers/OTSPDE-WCE2009/OTSPDE-WCE2009.tex
OTSPDE: - Checking in Jim's revisions to WCE2009 manuscript.

@@
   Email: ktchu@serendipityresearch.org.
-  Manuscript submitted 2008 March 2. 
+  Manuscript submitted 2009 March 20. 
  } 
@@
 to define an optimal computational grid for the FD scheme on the 
-\emph{original} domain.   By using both the optimal grid and OTS, we are
+\emph{original} domain.   By using both the optimal grid and OTS selection, we are
 able to boost the order of accuracy above what would be expected from a formal
@@
 applying it to variable coefficient wave and diffusion equations.  In 
-addition, we demonstrate the viability of OTS selection to the two-step 
+addition, we demonstrate the viability of OTS selection for the two-step 
 Kreiss-Petersson-Ystr\"om discretization of the wave equation.
@@
 Optimal time step (OTS) selection is a surprisingly simple and effective way to 
-boost the order of accuracy of finite difference (FD) schemes for 
+boost the order of accuracy of finite-difference (FD) schemes for 
 time-dependent partial differential equations~\cite{chu_otspde}.  The basic
@@
 Unfortunately, a limitation of the existing formulation of OTS selection is 
-the requirement that the leading-order spatial derivative of the PDE be 
+the requirement that the leading-order spatial derivative of the PDE have a 
 constant coefficient.  In this paper, we remove this limitation and extend OTS 
-selection to be applicable to general variable coefficient semilinear PDEs.  
+selection to general variable-coefficient semilinear PDEs.  
 Our approach is to identify a transformation of the domain which 
@@
 This paper is organized as follows.  First, we briefly review OTS selection 
-for PDEs with a constant coefficient leading-order spatial derivative term.  
+for PDEs with a constant-coefficient leading-order spatial derivative term.  
 Next, we show how to combine optimal grid and optimal time step selection to 
-boost the accuracy of finite difference schemes for variable coefficient PDEs.  
-We then demonstrate the use of variable coefficient OTS selection on the 
-variable coefficient wave and heat equations.  Our analysis of the wave is 
-equation is of particular note because it is \emph{not} based on a single step 
+boost the accuracy of finite difference schemes for variable-coefficient PDEs.  
+We then demonstrate the use of variable-coefficient OTS selection on the 
+variable-coefficient wave and heat equations.  Our analysis of the wave is 
+equation is particularly noteworthy because it is \emph{not} based on a one-step 
 time integration scheme.  We conclude with a summary of our main results
@@
 Optimal time step selection is not by itself a method for constructing 
-finite difference schemes.  Rather, it is a technique for enhancing the 
+finite-difference schemes.  Rather, it is a technique for enhancing the 
 performance of existing FD schemes by carefully choosing the time step to 
@@
 To illustrate the basic theory behind OTS selection, let us consider the 
-constant coefficient diffusion equation:
+constant-coefficient diffusion equation
 \beq
@@
 OTS selection is based on a detailed analysis of the leading-order errors in 
-FD schemes.  So, we begin by deriving the truncation equation for the scheme 
+FD schemes.  We begin by deriving the truncation error for the scheme 
 (\ref{eq:diffusion_eqn_1d_FD_scheme}).  Employing Taylor series expansions and 
@@
   \tu_{xx} + \frac{\dx^2}{12} \tu_{xxxx}
-  + O(\dx^4)
+  + O(\dx^4).
   \label{eq:diffusion_eqn_1d_space_err}
@@
 leading-order spatial derivative (at the expense of adding 
-variable coefficient lower-order spatial derivative terms).  The PDE 
+lower-order variable-coefficient spatial derivative terms).  The PDE 
 (\ref{eq:semilinear_PDE}) on the transformed domain has the form
@@
 
-\subsection{Derivation of OTS and Correction Terms for Optimal Grid
+\subsection{Derivation of OTS Selection and Correction Terms for Optimal Grid
   \label{sec:derivation_ots_corr_terms} }
@@
 on the details of the FD scheme and PDE.  The main challenge is deciding which 
-terms of the leading-order temporal error are not canceled out by the choice 
+terms of the leading-order temporal error are not cancelled out by the choice 
 of time step.
 
-Fortunately, a simple heuristic that can be used to derive the correction 
+Fortunately, a simple heuristic can be used to derive the correction 
 terms.  When deriving $u_{tt}$, each spatial derivative term in the PDE (on 
@@
 the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
-equation~\cite{kreiss2002}, we begin by analyzing OTS selection for the of 
-the constant coefficient wave equation and demonstrate its applicability for 
+equation~\cite{kreiss2002}, we begin by analyzing OTS selection for 
+the constant-coefficient wave equation and demonstrate its applicability to
 this multistep method.
@@
   = c^2 \left( \frac{u^{n}_{i+1} - 2 u^n_i + u^n_{i-1}}{\dx^2} \right)
-  + f
+  + f.
   \label{eq:wave_eqn_KPY}
 \eea
-and showed that it is second-order accurate in space and time with a 
+They showed that it is second-order accurate in space and time with a 
 stability constraint of the form $\dt = O(\dx)$.
@@
 error for the scheme.  Employing Taylor series expansions, it is 
-straightforward to show that the true solution satisfies:
+straightforward to show that the true solution satisfies
 \bea
@@
   \nonumber \\
-  & & +f
+  & & +f.
 \eea
@@
 \beq
-  \tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt}
+  \tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt},
 \eeq
@@
     \nonumber \\
-    & & + \ O(\dx^4) + O(\dt^4) 
+    & & + \ O(\dx^4) + O(\dt^4).
 \eea
@@
 For the KPY discretization of the wave equation, it is critical that a 
-high-order accurate method is used to compute the first time step because 
+high-order accurate method is used to perform the first time step because 
 the error introduced during this step affects the global error at all times.
@@
 solution at the first time step must be accurate to $O(\dx^{p+1})$.  In other 
-words the error introduced by the first time step must be at least one order
+words, the error introduced by the first time step must be at least one order
 of accuracy higher than desired for the overall solution.
@@
 We can understand this need for higher-order accuracy of the first time step 
-by solving the difference equation for the normal modes of the error:
+by solving the difference equation
 \bea
@@
 \eea
+for the normal modes of the error, 
 where $\he$ is the coefficient of an arbitrary normal mode of the spatial
@@
       + \he^0 \kappa_+ \kappa_- 
-        \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-} 
+        \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-},
   \label{eq:error_eqn_normal_mode_soln}
@@
   \nonumber \\
-  & & + \ \frac{\dt^4}{24} \left(c^4 u_{xxxx}^0 + c^2 f_{xx} + f_{tt}\right)
+  & & + \ \frac{\dt^4}{24} \left(c^4 u_{xxxx}^0 + c^2 f_{xx} + f_{tt}\right).
   \label{eq:KPY_fifth_order_first_step}
@@
 \subsubsection*{OTS Selection for Variable Coefficient Wave Equation}
-We now consider the variable coefficient second-order wave equation:
+We now consider the variable coefficient second-order wave equation
 \beq
@@
 (\ref{eq:change_of_variables}) and (\ref{eq:abar}), we can transform 
-(\ref{eq:var_coef_wave_eqn}) into the constant coefficient wave equation:
+(\ref{eq:var_coef_wave_eqn}) into the constant-coefficient wave equation
 \beq
@@
 
-\paragraph*{OTS on Transformed Domain}
+\paragraph*{OTS Selection on Transformed Domain}
 On the transformed domain, it is straightforward to apply OTS selection 
@@
 identifying the terms in $\left(\dt^2 \tu_{tttt}/12\right)$ that are not 
-eliminated by the use of $\dt_{opt}$, is given by:
+eliminated by the use of $\dt_{opt}$, is given by
 \bea
@@
     \end{array}
-  \right)
+  \right).
 \eea
@@
 
-\paragraph*{OTS with Optimal Grid}
-While applying OTS to the equation on the transformed domain yields the 
+\paragraph*{OTS Selection with Optimal Grid}
+While applying OTS selection to the equation on the transformed domain yields the 
 desired boost in order of accuracy, the lower-order spatial derivative and 
@@
 \bea
-  u_{tttt} =
+  u_{tttt} & = & 
     c^4 u_{xxxx} + 2 c^2 (c^2)_x u_{xxx} 
-    + c^2 (c^2)_{xx} u_{xx} + c^2 f_{xx} + f_{tt}
+    + c^2 (c^2)_{xx} u_{xx} + \nonumber \\ 
+& & c^2 f_{xx} + f_{tt}.
 \eea
@@
 \beq
-  \frac{\dt^2}{12} \left[ c^2 (c^2)_{xx} u_{xx} + c^2 f_{xx} + f_{tt} \right]
+  \frac{\dt^2}{12} \left[ c^2 (c^2)_{xx} u_{xx} + c^2 f_{xx} + f_{tt} \right].
 \eeq
@@
 \subsection{Application to Variable Coefficient Diffusion Equation}
-In this section, we apply OTS and optimal grid selection to the variable
-coefficient diffusion equation:
+In this section, we apply OTS and optimal grid selection to the 
+variable-coefficient diffusion equation
 \beq
@@
 Figure~\ref{fig:var_coef_diffusion_eqn_1d_solns} shows a comparison of 
-solutions computed with and without the optimal grid and OTS.
+solutions computed with and without the optimal grid and OTS selection.
 
@@
 computational grid.  In particular, we have presented a method to boost the 
-order of accuracy of FD schemes for PDEs where the leading-order has a 
-variable coefficient by using an optimal time step and an optimal grid.  Using 
-straightforward numerical analysis arguments, we have described the basic 
-procedure for constructing the optimal grid, computing the optimal time step, 
+order of accuracy of FD schemes for PDEs where the leading-order spatial derivative has a 
+variable coefficient by using an optimal time step and an optimal grid.  We have 
+described the basic procedure for constructing the optimal grid, computing the optimal time step, 
 and deriving the required correction terms.  To demonstrate the power of these 
@@
 work to PDEs in higher spatial dimensions.  We believe that the use of a 
-change of variables may still be viable.  However, we anticipate several
-challenges.  For instance, we may need to allow for a change in the shape 
-of domain and the possibility the optimal grid may no longer be orthogonal.  
+change of variables may still be viable.  However, we may need to allow for a change in the shape 
+of domain and the possibility that the optimal grid may no longer be orthogonal.  
 Both of these issues complicate the construction of the finite difference

### 2009-05-25 papers/OTSPDE/OTSPDE.tex
OTSPDE: - checked in version of OTSPDE paper submitted to Applied Numerical Mathematics

--- 2008-11-16 papers/OTSPDE/OTSPDE.tex
@@
 % Resubmission draft: 2008/11/16, Vitamin D, Inc.
+% Resubmission draft: 2009/05/25, Serendipity Research
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \def\tu{\tilde{u}}
+\def\he{\hat{e}}
 \def\dt{\Delta t}
@@
 % MISC MACROS
-\def\eg{\emph{e.g., }}
-\def\ie{\emph{i.e., }}
+\def\eg{\emph{e.g. }}
+\def\ie{\emph{i.e. }}
 \def\etal{\emph{et al.}}
@@
 
-\title{Boosting the Accuracy of Finite Difference Schemes 
-       via Optimal Time Step Selection}
-
-\author{K. T. Chu\address{Vitamin D, Inc., Menlo Park, CA 94025}$^,$\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore} 
+\title{Boosting the Accuracy of Finite Difference Schemes via
+       Optimal Time Step Selection and Non-Iterative Defect Correction}
+
+\author{Kevin T. Chu\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore}$^,$\address{Vitamin D, Inc., Menlo Park, CA 94025, United States}
 }
 
-\runtitle{Optimal Time Step Selection for FD Schemes}
-\runauthor{K. T. Chu}
+\runtitle{Optimal Time Step Selection and Non-Iterative Defect Correction for 
+          FD Schemes}
+\runauthor{K.T. Chu}
 
@@
 \begin{abstract}
-In this article, we present a novel technique for boosting the order of 
-accuracy of formally low-order finite difference schemes for time dependent 
+In this article, we present a simple technique for boosting the order of 
+accuracy of finite difference schemes for time dependent 
 partial differential equations by optimally selecting the time step used
-to advance the numerical solution.  This technique allows us to extract as
-much accuracy as possible from existing finite difference schemes with 
-minimal additional effort.  Through straightforward numerical 
-analysis arguments, we explain the origin of the boost in accuracy and 
-estimate the computational cost of the resulting numerical 
-method.  We demonstrate the utility of optimal time step (OTS) selection on a 
-wide array of classical linear and semilinear PDEs in one and more space 
-dimensions on both regular and irregular domains.  
+to advance the numerical solution and adding defect correction terms in a
+non-iterative manner.  The power of the technique is its ability to extract 
+as much accuracy as possible from existing finite difference schemes with 
+minimal additional effort.  Through straightforward numerical analysis 
+arguments, we explain the origin of the boost in accuracy and estimate the 
+computational cost of the resulting numerical method.  We demonstrate the 
+utility of optimal time step (OTS) selection combined with non-iterative 
+defect correction (NIDC) on several different types of finite difference
+schemes for a wide array of classical linear and semilinear PDEs in one and 
+more space dimensions on both regular and irregular domains.  
 \end{abstract}
@@
 \noindent \rule{6.3in}{1pt}
+
+% KEYWORDS
+% optimal time step; non-iterative defect correction; finite difference 
+% schemes; high-order accurate numerical methods; time dependent PDEs
 
@@
 numerical 
-schemes~\cite{bruger_2005,gibou_2005,ito_2005,shukla_2005,shukla_2007}.  
+schemes~\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
 Typically, high-order accuracy is achieved by constructing
@@
 accuracy can also be obtained by using formally low-order schemes in clever 
-ways (\eg Richardson extrapolation and compact finite difference schemes).  
-When possible, the latter approach can be a powerful way to boost the accuracy 
-of a numerical method \emph{without} introducing too much additional 
-algorithmic (and programming) complexity.
+ways (\eg Richardson extrapolation and compact finite difference 
+schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
+powerful way to boost the accuracy of a numerical method \emph{without} 
+introducing too much additional algorithmic (and programming) complexity.
 
@@
 schemes.  Often, all combinations of numerical parameters are considered 
-equally good.  That is, the simplest model of the error as a function 
-of the numerical parameters is assumed:  each numerical parameter contributes
-independently to the error.  For some problems, however, optimizing the 
-numerical parameters can significantly increase the accuracy of the 
-numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed that 
-choosing an appropriate value for the weight factor $\theta$ in the 
-theta-method can decrease the error in finite element solutions to the 
-diffusion equation.
-
-In this paper, we present a novel technique for boosting the order of 
+equally good.  For some problems, however, optimizing the numerical parameters 
+can significantly increase the accuracy of the numerical solution.  For 
+example, the well-known unit CFL condition, $\dt = \dx/a$, for the first-order 
+upwind forward Euler discretization of the linear advection equation 
+completely eliminates the numerical error introduced by the 
+scheme~\cite{leveque_book_2002}.  Another well-known example is the boost in 
+the order of accuracy that results from setting the time step to 
+$\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
+time integration with the standard second-order central difference 
+approximation for the Laplacian.  As a final example, Zhao showed that an 
+appropriate choice of the weight factor $\theta$ in the theta-method can 
+significantly decrease the error in finite element solutions to the diffusion 
+equation~\cite{zhao_2006}.
+
+In this article, we present a simple technique for boosting the order of 
 accuracy of finite difference schemes for time dependent PDEs by optimally 
-choosing the time step (and possibly adding a few correction terms).  
-Optimal time step (OTS) selection is a simple method based on the observation 
-that a carefully chosen time step can eliminate low-order terms in the 
-discretization error, allowing a formally low-order method to deliver 
-high-order accuracy.  It combines the well-known approach of using knowledge 
-of the structure of the PDE to reduce numerical error with the less-common 
-notion that an optimal choice of numerical parameters can boost the 
-effectiveness of algorithmically simple and computationally low-cost numerical 
-schemes.
-
-As a technique for enhancing numerical methods, optimal time step selection 
-has several desirable features.  Like any technique that leads to high-order 
-methods, optimal time step selection produces schemes that greatly reduce the 
-computational cost required to obtain a numerical solution of a desired 
-accuracy.  However, its real power comes from the fact that it achieves
-high-order accuracy while being based solely on very simple, formally 
-low-order finite difference schemes.  In other words, OTS selection allows us
-to extract as much accuracy as possible from an existing numerical scheme
-with minimal additional work.  For instance, we will see how optimal time step 
-selection makes it possible to achieve 4th-order accuracy for the diffusion 
-equation using only a second-order stencil for the Laplacian and forward 
-Euler time integration.  Moreover, high-order accuracy is not restricted to 
-problems on simple, rectangular domains; irregular domains are straightforward 
-to handle by appropriately setting the values in ghost 
+choosing the time step and, when necessary, adding defect correction terms
+in a \emph{non-iterative} manner.  This technique is a systematic 
+generalization of the ideas underlying some of the examples mentioned above.
+Optimal time step (OTS) selection is based on the observation that a carefully 
+chosen time step can simultaneously eliminate low-order terms in both the
+spatial and temporal the discretization errors.  It combines the well-known 
+approach of using knowledge of the structure of the PDE to reduce numerical 
+error with the less-common notion that an optimal choice of numerical 
+parameters can boost the effectiveness of algorithmically simple and 
+computationally low-cost numerical schemes.  For some problems, OTS alone is 
+sufficient to allow a formally low-order method to deliver high-order 
+accuracy.  However, addition of defect correction terms is generally required 
+to eliminate residual terms in the leading-order truncation error.  It is 
+important to emphasize that, unlike traditional defect/deferred correction 
+methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006} 
+which use defect correction terms to iteratively refine the solution at 
+each time step, we take a non-iterative defect correction (NIDC) approach
+that incorporates defect correction terms directly into the original finite 
+difference scheme.
+
+As a technique for enhancing numerical methods, OTS-NIDC has several desirable 
+features.  Like any technique that leads to high-order methods, OTS-NIDC
+produces schemes that greatly reduce the computational cost required to obtain 
+a numerical solution of a desired accuracy.  However, its real power comes 
+from the fact that it achieves high-order accuracy while being based solely on 
+very simple, formally low-order finite difference schemes.  In other words, 
+OTS-NIDC allows us to extract as much accuracy as possible from an existing 
+numerical scheme with minimal additional work.  For instance, we will see how 
+OTS-NIDC makes it possible to achieve 4th-order accuracy for the diffusion 
+equation in any number of space dimensions using only a second-order stencil 
+for the Laplacian and explicit time integration.  
+Moreover, high-order accuracy is not restricted to problems 
+on simple, rectangular domains; irregular domains are straightforward to 
+handle by appropriately setting the values in ghost 
 cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
-Optimal time step selection is even beneficial when 
-using finite difference schemes to solve some nonlinear PDEs.  While there 
-are certainly limitations to optimal time step selection, we will see that 
-there are many problems where it is useful.
-
-It is worth spending a few moments to draw the distinction between optimal 
-time step selection and traditional adaptive time stepping 
+OTS-NIDC is even beneficial when using finite difference schemes to solve some 
+nonlinear PDEs.  While there are certainly limitations to OTS-NIDC, we will 
+see that there are many problems where it is useful.
+
+This article is organized as follows.  In the remainder of this section,
+we compare OTS-NIDC to a few related numerical techniques.
+In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
+OTS-NIDC and show how it boosts the order of accuracy for formally low-order 
+numerical schemes.  
+In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
+OTS-NIDC to problems in one space dimension by applying it to finite 
+difference schemes for several classical linear and nonlinear PDEs.  
+In Section~\ref{sec:applications_multidim}, we show that 
+OTS-NIDC is useful for problems in multiple space dimensions by applying it
+the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
+the ease with which high-order accuracy can be achieved for problems on 
+arbitrary domains, the 2D diffusion equation is solved on both regular 
+and irregular domains.  
+For each PDE in Sections~\ref{sec:applications_1d} and 
+\ref{sec:applications_multidim}, we provide the numerical analysis required 
+to calculate the optimal time step size, the defect correction terms, and the 
+order of accuracy.  Finally, we offer some concluding remarks in 
+Section~\ref{sec:summary}.
+
+\subsection{Comparison with Traditional Defect/Deferred Correction Methods}
+Defect/deferred correction is a well-known strategy for boosting the order of 
+accuracy of finite difference schemes for time dependent PDEs.  It usually 
+takes one of two forms: 
+(1) iterative refinement of the numerical solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
+or 
+(2) derivation of high-order compact schemes for the spatial derivative 
+operator treating time derivatives as source 
+terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.
+
+For iterated defect/deferred correction, the numerical solution at each 
+time step is computed by iteratively adding correction terms that are 
+calculated from lower accuracy iterates.  When used to improve the 
+temporal order of accuracy, the iterative approach can help to ensure 
+stability of the time integration scheme~\cite{kress_2006}.  Although there 
+are instances where iterative defect correction has been used to eliminate 
+both spatial and temporal errors~\cite{gustafsson_2002}, it is usually used to 
+improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
+Note that unlike iterated defect correction, OTS-NIDC requires no iteration 
+during each time step because it computes correction terms using only the 
+solution at the previous time step(s) (as opposed to requiring an estimate 
+of the solution at the current time).
+
+High-order compact schemes focus on eliminating spatial discretization errors 
+when deriving the semi-discrete equations.  When applied to time dependent 
+PDEs, a high-order compact scheme for the spatial derivative operator is 
+derived treating time derivatives as source 
+terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.  
+After deriving the high-order spatial discretization, any time differencing 
+scheme may be applied to discretize the time derivatives.  While this 
+approach is very general, it really only provides a mechanism for reducing 
+spatial discretization error.  Temporal discretization errors are completely 
+controlled by the choice of time differencing scheme.  Note that because 
+high-order compact schemes for the spatial derivative operators involve spatial 
+derivatives of the source term, the resulting schemes are always effectively 
+implicit (even if all of the spatial derivative terms are treated explicitly).
+
+OTS-NIDC has several advantages over both of these defect correction 
+techniques.  First, OTS-NIDC is generally more computationally efficient than 
+both iterated defect correction and the high-order compact scheme approach
+because it requires less memory and, for many problems, admits simple 
+explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
+spatial and temporal numerical errors at the same time.  Finally, OTS-NIDC 
+does not rely solely on the addition of finite differences of solution values 
+to cancel out discretization errors.  Instead, it takes advantage of 
+relationships between spatial derivatives, temporal derivatives and source 
+terms to eliminate the need for finite differences approximations of 
+high-order derivatives of the solution.  Compared with iterated defect 
+correction, OTS-NIDC makes it easier to avoid instabilities that arise from 
+the presence of finite difference discretizations of high-order temporal 
+derivatives and may eliminate some of the correction terms.  Comparing to the
+high-order compact scheme approach, the optimization of numerical parameters 
+in OTS-NIDC means that mixed partial derivatives do not need to appear in the 
+fully discrete equations.
+
+\subsection{Comparison with Adaptive Time Stepping Methods}
+OTS selection is distinct and separate from traditional adaptive time stepping 
 techniques~\cite{iserles_book,shampine_2005} that are used in the 
@@
 
-The remainder of this paper is organized as follows.  
-In Section~\ref{sec:optimal_time_step_selection}, we begin by 
-discussing the technique of optimal time step selection and 
-presenting the analysis that explains how it boosts the order of accuracy 
-for formally low-order numerical schemes.  In 
-Section~\ref{sec:applications_1d}, we demonstrate the utility of 
-optimal time selection to problems in one space dimension by applying it to 
-finite difference schemes for several classical linear and nonlinear PDEs. 
-In Section~\ref{sec:applications_multidim}, we show that optimal time step
-selection is useful for problems in multiple space dimensions by applying it
-the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
-the ease with which high-order accuracy can be achieved for problems on 
-arbitrary domains, the 2D diffusion equation is solved on both regular 
-and irregular domains.  
-For each PDE in Sections~\ref{sec:applications_1d} and 
-\ref{sec:applications_multidim}, we provide the numerical analysis required 
-to calculate the optimal time step size and order of accuracy.  
-Finally, in Section~\ref{sec:summary}, we end some concluding remarks.
-
-\section{\label{sec:optimal_time_step_selection} 
-         Optimal Time Step (OTS) Selection}
-Optimal time step selection is not by itself a method for constructing 
-finite difference schemes.  Rather it is a way to enhance to the performance
-of existing finite difference schemes by carefully choosing the time step to 
-eliminate low-order numerical errors.  The two basic ideas underlying 
-optimal time step selection are 
+\section{\label{sec:OTS-NIDC}
+         Theoretical Foundations of OTS Selection and NIDC}
+OTS-NIDC is not by itself a method for constructing finite difference 
+schemes.  Rather, it is a way to enhance to the performance of existing 
+finite difference schemes by carefully choosing the time step and adding
+defect correction terms to eliminate low-order numerical errors.  
+The two key ideas underlying OTS-NIDC are 
 (1) a judicious choice of time step can be used to eliminate the leading-order 
@@
 and
-(2) the PDE provides valuable insight into the discretization errors for 
-finite difference schemes.
-In this section, we present and analyze the technique of optimal time step 
-selection for finite difference schemes constructed to solve scalar evolution 
-equations of the form 
+(2) the PDE provides valuable insight into the relationships between terms
+in the discretization error. 
+In this section, we present and analyze the technique of OTS-NIDC for finite 
+difference schemes constructed to solve scalar evolution equations of the form 
 \beq
@@
 restricting our attention to problems of this form is not a serious 
-limitation.
+limitation.  The same principles apply to finite difference schemes for 
+general PDEs.
 
@@
 explicitly keeping track of the leading-order terms.  In contrast to 
-conventional error analysis, we do not sweep all of the errors under big O
+conventional error analysis, we do not sweep all of the errors under big-$O$
 notation.  Rather, we take advantage of the fact that derivation of the 
@@
 use the PDE to relate the leading-order terms to each other.  
-If we have chosen an appropriate finite difference approximation for 
-(\ref{eq:model_PDE}), which is relatively straightforward to do, this careful 
-analysis yields a direct relationship between the leading-order spatial and 
-temporal errors that allows them to be combined as a single term of the form:
+For many finite difference schemes, this careful analysis yields a direct 
+relationship between the leading-order spatial and temporal errors that allows 
+them to be combined as a single term of the form:
 \beq
@@
 \eeq
-where $L$ is a differential operator and $P$ is a polynomial in the numerical 
-parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
-$P = 0$, we can completely eliminate the leading-order term in the 
+where $L$ is a differential operator and $P$ is a simple function of the 
+numerical parameters $\dx$ and $\dt$.  By selecting the numerical parameters 
+so that $P = 0$, we can completely eliminate the leading-order term in the 
 discretization error.  Because time dependent PDEs are often solved by 
@@
 those based on the method of lines.   When the method of lines approach is 
-used for (\ref{eq:model_PDE}), $P$ takes a particularly simple 
+used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
 form:
@@
 where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
-discretization errors, and $\alpha$ and $\beta$ are positive constants that 
+discretization errors, and $\alpha$ and $\beta$ are constants that 
 depend on the details of the PDE and finite difference scheme.  Setting 
@@
 \beq
-  \dto = \gamma \dx^{r/s},
+  \dto = \left(\alpha/\beta \right)^{1/s} \dx^{r/s}.
   \label{eq:optimal_time_step}
 \eeq
-where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
-has a similar functional dependence on the grid spacing as a time step that 
-satisfies a stability constraint.
+Notice that the optimal time step has a similar functional dependence on the 
+grid spacing as a time step that satisfies a stability constraint.
+
+An important assumption in the above analysis is that $\alpha$ and $\beta$ 
+have the same sign.  If this condition is not satisfied, (\eg finite
+difference scheme for the diffusion equation based on the standard central
+difference stencil for the Laplacian and backward Euler time integration),
+then OTS selection is not applicable because none of the zeros of 
+(\ref{eq:leading_order_error_model_PDE_MOL}) is positive.  Fortunately, 
+in this situation, it is usually possible to find a closely related finite 
+difference scheme which \emph{does} have a positive optimal time step.
 
@@
 lines is so frequently used to construct finite difference approximations for 
-time dependent PDEs, we will focus solely on finite difference schemes of this 
-type for the remainder of the paper.  It is straightforward to extend the 
-results we obtain to more general finite difference schemes.
+time dependent PDEs, we will focus primarily on finite difference schemes of 
+this type for the remainder of the article.  It is straightforward to extend 
+the results we obtain to more general finite difference schemes (\eg the
+Lax-Wendroff scheme for the linear advection equation).
 
@@
 integration schemes, OTS selection is ineffective when the optimal time step 
-is greater than the smallest stable time step.  In this situation, we can 
-switch to an implicit or semi-implicit time integration scheme with the 
-same order of accuracy for which the optimal time step is stable.  
-While the need to shift to implicit or semi-implicit time integration results 
-in a more complicated finite difference scheme, OTS selection is still 
-beneficial because it allows us to extract more accuracy from the simplest, 
-low-order implicit or semi-implicit methods.
-Note that a new optimal time step may need to be calculated for the modified 
-finite difference scheme.
+is greater than the smallest stable time step.  In this situation, it is 
+necessary to replace the finite difference approximation of the PDE.  
+For example, if the diffusion equation is solved using a 
+finite difference scheme based on forward Euler time integration and a 
+discretization of the Laplacian with an isotropic leading-order truncation 
+error (see Section~\ref{sec:diffusion_eqn_2d}), the optimal time step becomes 
+unstable when the dimension of space is greater than 3.  To make OTS selection 
+useful for diffusion problems in 4 and more space dimensions, we can switch 
+to the DuFort-Frankel scheme, which is explicit and unconditionally 
+stable~\cite{gko_book}.
 
@@
 As a result, using the optimal time step to advance the finite
-difference scheme in time may appear to be overly restrictive and lead to poor 
-computational performance.  However, this apparent drawback is more than 
+difference scheme in time may appear to be overly restrictive.
+However, this apparent drawback is more than 
 compensated by the boost in the order of accuracy.  For example, when solving 
-the 1D diffusion equation using a forward Euler time integration scheme with a 
+the diffusion equation using a forward Euler time integration scheme with a 
 second-order central difference stencil for the Laplacian, we obtain a 
@@
 \renewcommand{\arraystretch}{1.5}
+\centering
 \begin{tabular}{lcccc}
@@
                    & $O\left( e^{-1} \right)$ \\
-  Forward Euler with OTS  & $O\left( e^{1/4} \right)$ 
+  Forward Euler with OTS-NIDC  & $O\left( e^{1/4} \right)$ 
                    & $O\left( e^{1/2} \right)$ 
@@
 \caption{
-Computational cost of the forward Euler scheme with optimal time step and 
+Computational cost of the forward Euler scheme with OTS-NIDC and
 the Crank-Nicholson method for solving the diffusion equation as a function 
 of the global numerical error $e$.
-For both schemes, second-order, isotropic stencils are used to discretize 
-the Laplacian.  
 Note that the computation time for the Crank-Nicholson method assumes an 
-$O(N) = O \left( 1/\dx^d \right)$ matrix inversion algorithm 
-and that the optimal time step for the forward Euler scheme is 
-$\dto = \dx^2/6D$.
+$O(N) = O \left( 1/\dx^d \right)$ matrix inversion algorithm. 
+A comparison of computational costs in higher spatial dimensions continues 
+along the same trends but requires replacement of the forward Euler time 
+integration with an alternative for which the optimal time step is stable 
+(\eg the DuFort-Frankel scheme). 
 }
@@
 \renewcommand{\arraystretch}{1.5}
+\centering
 \begin{tabular}{ccccc}
   \hline
-  & \multicolumn{2}{c}{\bf Forward Euler with OTS} 
+  & \multicolumn{2}{c}{\bf Forward Euler with OTS-NIDC} 
   & \multicolumn{2}{c}{\bf Crank-Nicholson} \\
@@
       & $O\left( e^{-2} \right)$ \\
-  $d$ & $O\left( e^{-d/4} \right)$ 
-      & $O\left( e^{-(d+2)/4} \right)$ 
-      & $O\left( e^{-d/2} \right)$ 
-      & $O\left( e^{-(d+1)/2} \right)$ \\ 
   \hline 
@@
 
-\subsection{\label{sec:ots_general_1d_pdes} 
-            OTS Selection for General PDEs in One Space Dimension} 
+\subsection{\label{sec:ots_nidc_general_1d_pdes} 
+            OTS-NIDC for General PDEs in One Space Dimension} 
 In the previous section, we analyzed optimal time step selection in the 
@@
 OTS selection continues to be useful, but it needs to be augmented with 
-correction terms to achieve the same improvement in the order of accuracy.  
-In this section, we discuss these correction terms in the context of general 
-constant coefficient, linear PDEs and semilinear PDEs.
+defect correction (incorporated in a non-iterative manner) to achieve 
+the same improvement in the order of accuracy.  In this section, we discuss 
+these correction terms in the context of general constant coefficient, linear 
+PDEs and semilinear PDEs.
 
@@
 leading-order error.  To eliminate the vestigial errors, we need to add 
-correction terms to cancel out the second, third, and fourth terms in 
+defect correction terms to cancel out the second, third, and fourth terms in 
 (\ref{eq:leading_order_error_linear_PDE_MOL}).  At each time step, the second 
@@
 \subsubsection*{Semilinear PDEs}
-Optimal time step selection can also be applied to semilinear PDEs of the
-form
-\beq
-  \frac{\partial u}{\partial t} = A(x,t) \frac{\partial^n u}{\partial x^n}
+OTS-NIDC can also be applied to semilinear PDEs of the form
+\beq
+  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n}
   + F \left( \frac{\partial^{n-1} u}{\partial x^{n-1}},
@@
 of $u$ and $f(x,t)$ is a source term.  While the coefficient on the 
-leading-order spatial derivative is generally allowed to be a function of 
-space and time, optimal time step selection can only be applied to PDEs where
-$A(x,t)$ is a constant.  Because the PDE is linear in the leading-order 
+leading-order spatial derivative for a semilinear PDE is generally allowed 
+to be a function of space and time, optimal time step selection can only be 
+applied when it is a constant.  Because the PDE is linear in the leading-order 
 spatial derivative, it is natural to construct a finite difference scheme that 
 treats $F$ in an explicit manner and lets stability considerations determine 
-whether the $A \frac{\partial^n u}{\partial x^n}$ term is handled explicitly 
-or implicitly.  Assuming that sufficiently high-order stencils are used to 
+whether the $A \frac{\partial^n u}{\partial x^n}$ term is handled explicitly or 
+implicitly.  Assuming that sufficiently high-order stencils are used to 
 compute $F$, the leading-order discretization error for such a finite 
@@
 form (\ref{eq:leading_order_error_semilinear_PDE_MOL}), the optimal time 
-step and correction terms can be derived by proceeding in the same manner 
-as in the previous section.
+step and defect correction terms can be derived by proceeding in the same 
+manner as in the previous section.
 
@@
 accuracy for the solution.  A detailed discussion of methods for imposing 
-boundary conditions is beyond the scope of this paper.  We refer interested 
+boundary conditions is beyond the scope of this article.  We refer interested 
 readers to the thorough analysis presented in~\cite{gko_book}.  To avoid 
@@
 
-\subsection{\label{sec:ots_higher_spatial_dims}
-            OTS Selection in Multiple Space Dimensions}
-Optimal time step selection is applicable to PDEs in any number of space
-dimensions.  While problems in higher dimensions generally require a more 
-careful choice of the finite difference stencil used to approximate spatial 
-derivatives, there are no fundamental barriers that prevent the use of optimal 
-time step selection.  Moreover, by using a `ghost cell' or `ghost value' 
-method for imposing boundary 
+\subsection{\label{sec:ots_nidc_higher_spatial_dims}
+            OTS-NIDC Selection in Multiple Space Dimensions}
+OTS-NIDC is applicable to PDEs in any number of space dimensions.  While 
+problems in higher dimensions generally require a more careful choice of the 
+finite difference stencil used to approximate spatial derivatives, there are 
+no fundamental barriers that prevent the use of OTS-NIDC.  Moreover, by using 
+a `ghost cell' or `ghost value' method for imposing boundary 
 conditions~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book},
@@
 Without this assumption, the first term in the leading-order discretization 
-error would be of the form 
-$(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
+error would be of the form $(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
 where $\alpha$ and $\beta$ are now functions of space and time.  This
@@
 
+While these limitations certainly place restrictions on the range of problems 
+that we can apply OTS-NIDC to, there are many relevant and important PDEs 
+arising in practice that satisfy the requirements for OTS-NIDC.
+
 \section{\label{sec:applications_1d} 
          Application to PDEs in One Space Dimension}
-Because there are no geometry or anisotropy considerations, optimal time step 
-selection is very useful for boosting the accuracy of many finite difference 
-schemes in one spatial dimension.  In this section, we demonstrate its 
-utility in designing finite difference schemes to solve several classical 
-linear and semilinear PDEs.  The examples in this section were selected 
-because they have analytical solutions that are useful to compare numerical 
-solutions against.  However, the utility of OTS selection is by no means 
-limited to these simple PDEs.
-
-\subsection{Linear Advection Equation}
-For our first application of optimal time step selection, we consider 
-a finite difference scheme for the linear advection equation
-~\cite{gko_book,leveque_book_1992,leveque_book_2002}:
-\beq
-  \frac{\partial u}{\partial t} + A \frac{\partial u}{\partial x} = 0,
-  \label{eq:advection_eqn_1d}
-\eeq
-where $A$ is the flow speed, which we take to be positive for convenience.  
-For an initial condition of the form $u(x,0) = f(x)$, the analytical solution 
-is $f(x-At)$.  While it is unlikely that one would actually use a numerical 
-method to solve~(\ref{eq:advection_eqn_1d}), it is an interesting example 
-where optimal time step selection dramatically improves the accuracy of the 
-numerical solution.
-
-Perhaps the simplest, stable finite difference scheme 
-for (\ref{eq:advection_eqn_1d}) uses forward Euler for time integration 
-scheme and a first-order upwind discretization for the spatial derivative:
+Because there are no geometry or anisotropy considerations, OTS-NIDC is very 
+useful for boosting the accuracy of many finite difference schemes in one 
+spatial dimension.  In this section, we demonstrate its utility in designing 
+finite difference schemes that solve several classical linear and semilinear 
+PDEs.  To emphasize the importance of including numerical parameter 
+optimization as part of the design of any finite difference scheme, we 
+include applications of OTS-NIDC to several different types of finite 
+difference schemes.  The examples in this section were selected because they 
+have analytical solutions that are useful to compare numerical solutions 
+against.  However, the utility of OTS-NIDC is by no means limited to 
+these simple PDEs.
+
+\subsection{Unit CFL Conditions for Wave Equations}
+In this section, we show how OTS selection naturally leads to unit CFL 
+conditions for the linear advection equation and second-order wave equation 
+in one space dimension.
+
+\subsubsection{Linear Advection Equation}
+When the linear advection equation, $u_t + A u_x = 0$,
+is solved using a first-order upwind scheme with forward Euler time
+integration, it is well-known that choosing $\Delta t = \Delta x / |A|$ 
+completely eliminates all numerical error in the solution (except for 
+errors in the initial conditions).  This result is known as the unit CFL 
+condition~\cite{leveque_book_2002} and is usually proved by examining how 
+characteristic lines of the advection equation pass through grid points.  
+Interestingly, the same optimal choice of time step arises by applying OTS 
+selection.
+
+Using OTS selection, we begin by analyzing the discretization error for the 
+first-order upwind, forward Euler scheme:
 \beq
   u^{n+1}_j = u^{n}_j 
-  - A \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right).
+  - A \dt \left( \frac{u^{n}_j - u^{n}_{j-1}}{\dx} \right),
   \label{eq:advection_eqn_1d_FD_scheme}
 \eeq
-This scheme is formally first-order in both space and time.
-
-To compute the optimal time step, we begin by analyzing the discretization
-error for~(\ref{eq:advection_eqn_1d_FD_scheme}).  In general, we only need to 
-keep track of the leading-order terms in the discretization error.  However,
-for the advection equation, it is valuable to retain all of them.  
+where we have taken $A$ to be positive for convenience.  In general, we only 
+need to keep track of the leading-order terms in the discretization error.  
+However, for the advection equation, it is valuable to retain all of them.  
 Using Taylor series expansions of the true solution of the PDE, we find that 
@@
 where $\tu$ denotes the true solution of the PDE, all derivatives are 
-evaluated at $x_j$, and we have used the 
-PDE~(\ref{eq:advection_eqn_1d_FD_scheme}) to replace all of the time 
+evaluated at $x_j$, and we have used the PDE to replace all of the time 
 derivatives with spatial derivatives in (\ref{eq:advection_eqn_1d_time_err}).
 For convenience, we will continue to suppress the subscripts on spatial
-derivatives throughout the paper; the location where they should be 
+derivatives throughout the article; the location where they should be 
 evaluated will be apparent from the context.
@@
 \eeq
-Therefore, the leading-order term of the discretization error is given
-by
-\beq
-  \left( \frac{A}{2} \frac{\partial^2 \tu^n}{\px^2} \right)
-  \left( \dx - A \dt \right) \dt,
-\eeq
-which leads to an optimal time step of $\Delta t_{opt} = \Delta x / A$.
-
-For the linear advection equation, setting the time step equal to the optimal 
-time step has an amazing consequence -- not only does the leading-order term
-in the discretization error vanish, the \emph{entire} local truncation 
-error at each time step is completely eliminated!  In other words, the error 
-in the numerical solution is due solely to errors in the initial conditions.
-Assuming that the initial conditions are set exactly, there is absolutely no 
-error in the numerical solution for all times.
-This result, which is unique to the advection equation, is well-known in the 
-numerical analysis and scientific computing community as the unit CFL 
-condition~\cite{leveque_book_2002} and is usually proved by examining 
-how characteristic lines of the advection equation pass through grid points.  
-It is interesting to see that it also naturally arises through optimal time 
-step selection.
+Therefore, choosing the time step to eliminate the leading-order term 
+of the discretization error leads to the unit CFL condition and 
+\emph{complete elimination} of the local truncation error at each time step.
+It is interesting to note that OTS selection applied to the Lax-Wendroff 
+scheme~\cite{leveque_book_2002,leveque_book_1992} yields the same optimal 
+time step and total elimination of the truncation error.
+
+\subsection{Second-Order Wave Equation\label{sec:wave_eqn_1d}}
+Numerical methods for the wave equation $u_{tt} - c^2 u_{xx} = f$ are typically 
+derived by transforming the second-order equation into an equivalent
+first-order system of equations.  However, as Kreiss, Petersson, and Ystr\"om 
+showed in~\cite{kreiss2002}, direct discretization of the second-order wave 
+equation is a viable option and leads to the following finite difference 
+scheme that is second-order accurate in both space and time:
+\bea
+  \frac{u^{n+1}_i - 2 u^n_i + u^{n-1}_i}{\dt^2}
+  = c^2 \left( \frac{u^{n}_{i+1} - 2 u^n_i + u^n_{i-1}}{\dx^2} \right)
+  + f.
+  \label{eq:wave_eqn_KPY}
+\eea
+The stability constraint for this scheme of the form $\dt = K \dx$.
+
+To compute the optimal time step and correction terms for this scheme, we
+begin by deriving the truncation error for the scheme.  Employing Taylor 
+series expansions, it is straightforward to show that the true solution 
+satisfies
+\bea
+  \frac{\tu^{n+1}_i - 2 \tu^n_i + \tu^{n-1}_i}{\dt^2}
+    - \frac{\dt^2}{12} \frac{\partial^4 \tu}{\pt^4}
+  &=& c^2 \left( \frac{\tu^{n}_{i+1} - 2 \tu^n_i + \tu^n_{i-1}}{\dx^2}
+  -\frac{\dx^2}{12} \frac{\partial^4 \tu}{\px^4} \right) + f
+  \nonumber \\
+  &+& O(\dt^4) + O(\dx^4) 
+\eea
+Combining this result with the observation that
+$\tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt}$ and rearranging a bit, 
+we find that the error equation for (\ref{eq:wave_eqn_KPY}) is given by
+\bea
+  e^{n+1}_i &=& 2 e^n_i - e^{n-1}_i
+  + \dt^2 c^2 \left( \frac{e^{n}_{i+1} - 2 e^n_i + e^n_{i-1}}{\dx^2} \right)
+  + \frac{\dt^2}{12} c^2 \frac{\partial^4 \tu}{\px^4} 
+    \left( \dx^2 - \dt^2 c^2 \right)
+  \nonumber \\
+  &-& \frac{\dt^4}{12} \left( c^2 \frac{\partial^2 f}{\px^2} 
+                            + \frac{\partial^2 f}{\pt^2} \right)
+      + O(\dt^2 \dx^4) + O(\dt^6).
+\eea
+Choosing $\dt = \dx/c$ and adding the correction term 
+$\dt^2 \left(c^2 f_{xx} + f_{tt} \right)/12$ to the right-hand side of 
+(\ref{eq:wave_eqn_KPY}) eliminates the leading-order term in the local 
+truncation error.  Using the heuristic for two-step methods that the global 
+error should be approximately $1/\dt^2$ times the local error leads to a 
+global error of $O(\dx^4) = O(\dt^4)$ -- a boost in the order of accuracy 
+from second- to fourth-order.  Notice that as for first-order upwind, forward 
+Euler and Lax-Wendroff schemes for the linear advection equation, applying OTS 
+selection to the KPY scheme leads to an optimal time step that satisfies the 
+unit CFL condition.  However, unlike the schemes for the linear advection 
+equation, the unit CFL condition for the KPY scheme eliminates only the 
+leading-order term in the truncation error.  Figure~\ref{fig:wave_eqn_1d_error} 
+demonstrates that the expected accuracy is indeed achieved by applying 
+OTS-NIDC to the KPY scheme.  
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.35}{\includegraphics{figures/wave_eqn_1d_src_error_vs_N}}
+\caption{$L^\infty$ error as a function of number of grid points for the
+  KPY discretization of the second-order wave equation on with OTS-NIDC 
+  (blue circles) and without OTS-NIDC (red squares).
+}
+\label{fig:wave_eqn_1d_error}
+\end{center}
+\end{figure}
+
+An important property of the KPY scheme is that the error introduced during 
+the first time step affects the global error at all times.  More specifically, 
+a $(p+1)$-order accurate first time introduces an $O(\dx^p)$ error into the 
+solution at all future times.
+Thus, we must ensure that the error introduced by the first time step is
+at least one order of accuracy higher than desired accuracy for the numerical 
+solution.  See Appendix~\ref{appendix:KPY_analysis} for a derivation of this
+condition.
+
+Constructing a fifth-order approximation for the first time step is 
+straightforward using a fifth-order Taylor series expansion in time:
+\bea
+  u^1 &=& u^0 + \dt \frac{\partial u^0}{\pt} 
+              + \frac{\dt^2}{2} \frac{\partial^2 u^0}{\pt^2}
+              + \frac{\dt^3}{6} \frac{\partial^3 u^0}{\pt^3} 
+           + \frac{\dt^4}{24} \frac{\partial^4 u^0}{\pt^4}
+  \nonumber \\
+  &=& u^0 + \dt \frac{\partial u^0}{\pt} 
+          + \frac{\dt^2}{2} \left(c^2 \frac{\partial^2 u^0}{\px^2} + f\right)
+  + \frac{\dt^3}{6} \left(c^2 
+    \frac{\partial^2}{\px^2} \left(\frac{\partial u^0}{\pt} \right) 
+    + \frac{\partial f}{\pt}\right)
+  \nonumber \\
+  & & +\  \frac{\dt^4}{24} \left(c^4 \frac{\partial^4 u^0}{\px^4} 
+  + c^2 \frac{\partial ^2 f}{\px^2} + \frac{\partial^2 f}{\pt^2} \right).
+  \label{eq:KPY_fifth_order_first_step}
+\eea
+Note that special care must be exercised if finite differences are used to 
+compute the derivatives in (\ref{eq:KPY_fifth_order_first_step}) because
+higher-order terms may be implicitly included by lower-order terms in the 
+expansion if the initial time step is taken using $\dto$.
 
 \subsection{Diffusion Equation \label{sec:diffusion_eqn_1d}}
-As an example where optimal time step selection yields a more typical boost 
-in the order of accuracy, let us consider the diffusion equation:
-\beq
-  \frac{\partial u}{\pt} = D \frac{\partial^2 u}{\px^2} + f(x,t),
-  \label{eq:diffusion_eqn_1d}
-\eeq
-where $D$ is the diffusion constant and $f(x,t)$ is a source term.  
-We apply optimal time step selection to the simplest finite difference 
-scheme for (\ref{eq:diffusion_eqn_1d}):
+Solution of the diffusion equation, $u_t = D u_{xx}$, using forward Euler 
+time integration with a second-order central difference approximation for 
+the Laplacian provides a familiar example where careful choice of the time 
+step leads to cancellation of the leading order spatial and temporal 
+truncation errors.  It is well-known that choosing the time step to be 
+$\dto = \dx^2/6D$ boosts the order of accuracy for this simple scheme 
+from $O(\dx^2)$ to $O(\dx^4)$.
+
+When a source term is present, boosting the order of accuracy requires
+the addition of defect correction terms.  These correction terms are 
+straightforward to calculate by examining the discretization error for the
+scheme
 \beq
@@
 \eeq
-which uses forward Euler for time integration and the standard second-order 
-central difference approximation for the Laplacian.
-This scheme is formally first-order in time and second-order in space.  
-The stability constraint $\dt \le \dx^2/2D$ implies that the scheme is 
-$O(\dx^2)$ accurate overall.
-
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_no_src_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for three 
-finite difference schemes that solve the 1D diffusion equation: forward Euler 
-with optimal time step (circles), forward Euler with suboptimal time step
-(squares), and Crank-Nicholson (diamonds).  
-All three schemes use the standard second-order, central finite difference 
-approximation for the Laplacian.  
-}
-\label{fig:diffusion_eqn_1d_no_src_error}
-\end{center}
-\end{figure}
-
-As before, we begin by analyzing the leading-order terms of the 
-discretization error for the finite difference scheme.  Employing Taylor
-series expansions and the PDE (\ref{eq:diffusion_eqn_1d}),
-we find that the true solution satisfies
+where $f$ is the source term.  Note that even though this scheme is formally 
+first-order in time and second-order in space, the stability constraint 
+$\dt \le \dx^2/2D$ implies that the scheme is $O(\dx^2)$ accurate 
+overall (\ie inclusion of the temporal order of accuracy is irrelevant).
+Using Taylor series expansions and the PDE $u_t = D u_{xx} + f$,
+we see that the true solution satisfies
 \bea
@@
   + \frac{\dx^2}{12} \frac{\partial^4 \tu^n}{\px^4} 
-  + O(\dx^4)
+  + O(\dx^4).
   \label{eq:diffusion_eqn_1d_space_err}
@@
 \eeq
-From this expression, it is easy to see that the optimal time step is 
-$\dto = \dx^2/6D$ and that the correction term is
-\beq
-\frac{\dt^2}{2} \left( D \frac{\partial^2 f}{\px^2} 
-                     + \frac{\partial f}{\pt} \right).
-\eeq
+From this expression, we easily recover the well-known optimal time step for
+the diffusion equation, $\dto = \dx^2/6D$, and identify the correction term 
+$\dt^2 \left( D f_{xx} + f_t \right)  /2$.
+
 From (\ref{eq:global_error_ots_simplified}), we see that using the optimal 
@@
 $O(\dt \dx) = O(\dx^3)$ global error. 
-
-To demonstrate the improvement in accuracy of the forward Euler scheme when 
-the optimal time step is used, we compare the numerical errors of three finite 
-difference schemes for the diffusion equation: forward Euler with optimal
-time step and correction term, forward Euler with a suboptimal time step, and 
-Crank-Nicholson.  As we can see in 
-Figure~\ref{fig:diffusion_eqn_1d_no_src_error}, forward Euler with optimal 
-time step is fourth-order accurate while both forward Euler with a suboptimal 
-time step and Crank-Nicholson are only second-order accurate.  
-Using the forward Euler scheme with an optimal time step and correction term 
-has a clear advantage over the other finite difference schemes.  Even with as 
-few as 25 grid points, the error in solution computed by the forward Euler 
-scheme with an optimal time step and correction term is orders of magnitude 
-smaller.
-
 The importance of the correction term is illustrated in 
@@
 two finite difference schemes that solve the diffusion equation: 
-forward Euler with optimal time step and correction term (circles) and 
-forward Euler with optimal time step \emph{without} correction term 
-(triangles).  Both schemes use the standard second-order, central finite 
-difference approximation for the Laplacian.
+forward Euler with optimal time step and defect correction term (circles) and 
+forward Euler with optimal time step \emph{without} defect correction term 
+(triangles).  
 }
@@
 
-As mentioned in Section~\ref{sec:computational_performance}, one of the 
-advantages of using optimal time step selection is the superior scaling of 
-the computational cost with respect to the numerical error.  
-Another way to achieve computational performance with the same scaling would 
-be to use a Crank-Nicholson time integration scheme with a fourth-order 
-discretization of the Laplacian~\cite{gibou_2005}.  By using an 
-$O\left(\dx^2 \right)$ time step and a linear solver that scales linearly 
-with the size of the matrix, we achieve an overall fourth-order accurate 
-method that has the same computational complexity as the forward Euler scheme
-described above with an optimal time step.  However, the forward Euler scheme 
-with optimal time step will still outperform the Crank-Nicholson scheme 
-because the latter requires the use of a wider stencil and the solution of 
-at least one linear system of equation for each time step.  Round-off error
-associated with matrix inversion may also be an issue for the Crank-Nicholson 
-method.
+\subsubsection{DuFort-Frankel Scheme \label{sec:dufort_frankel}}
+As mentioned earlier, the forward Euler scheme for the diffusion equation is
+unstable for problems in more than 3 space dimensions.  Fortunately, the 
+OTS-NIDC makes it possible to construct a fourth-order accurate 
+DuFort-Frankel scheme.
+
+We begin by writing the DuFort-Frankel scheme in form that is easily 
+generalizable to higher space dimensions:
+\beq
+  u^{n+1}_j = u^{n-1}_j 
+  + 2 \dt 
+    \left( D \left[\frac{u^{n}_{j+1} -2 u^{n}_j + u^{n}_{j-1}}{\dx^2}\right] 
+         + f^n_j \right)
+  - 2 D \frac{\dt^3}{\dx^2}
+    \left( \frac{u^{n+1}_j -2 u^{n}_j + u^{n-1}_j}{\dt^2}\right).
+  \label{eq:diffusion_eqn_1d_DF_scheme}
+\eeq
+Note that the DuFort-Frankel scheme is essentially a leap-frog time integration 
+scheme that is stabilized by the addition of a term proportional to $u_{tt}$
+that vanishes in the limit $\dt \rightarrow 0$.  This scheme is 
+unconditionally stable and has a global error that is formally 
+$O(\dt^2/\dx^2) + O(\dt^2) + O(\dx^2) + O(\dt^4/\dx^2)$~\cite{gko_book}.  
+This expression for the formal error is ``minimized'' when $\dt = O(\dx^2)$, 
+which yields an $O(\dx^2)$ global error.
+
+Using Taylor series expansions centered at $(x_j, t_n)$, 
+(\ref{eq:diffusion_eqn_1d_space_err}) and the PDE to derive the local 
+truncation error for the DuFort-Frankel scheme, we find that 
+\bea
+  \tau &=& 2D \dt \frac{\partial^4 \tu}{\px^4}
+  \left(\frac{\dx^2}{12} - \frac{D^2 \dt^2}{\dx^2} \right)
+  - 2 D \frac{\dt^3}{\dx^2} \left(D \frac{\partial^2 f}{\px^2}  
+                                 + \frac{\partial f}{\pt} \right)
+  \nonumber \\
+  &+& O(\dt^3) + O(\dt \dx^4) + O\left(\frac{\dt^5}{\dx^2} \right).
+\eea
+Therefore the optimal time step is $\dto = \dx^2/D\sqrt{12}$ and the defect
+correction term is $2 D \dt^3 \left(D f_{xx} + f_{t} \right)/\dx^2$.  As with 
+the forward Euler scheme for the diffusion equation, OTS-NIDC boosts the
+order of accuracy for the DuFort-Frankel scheme from two to four.  
+Figure~\ref{fig:diffusion_eqn_1d_dufort_frankel_error} shows that OTS-NIDC
+applied to the DuFort-Frankel scheme achieves the expected boost in accuracy.
+
+\begin{figure}[tb]
+\begin{center}
+\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_DF_error_vs_N}} 
+\caption{$L^\infty$ error in the numerical solution of the 1D diffusion 
+equation as a function of number of grid points for the DuFort-Frankel scheme 
+with (circles) and without (squares) OTS-NIDC.
+}
+\label{fig:diffusion_eqn_1d_dufort_frankel_error} 
+\end{center}
+\end{figure}
 
 \subsection{Viscous Burgers Equation}
-We now demonstrate the use of optimal time step selection in the context of 
-a well-known semilinear PDE -- the viscous Burgers 
-equation~\cite{whitham_book}:
+We now demonstrate the use of OTS-NIDC in the context of a well-studied
+semilinear PDE -- the viscous Burgers equation~\cite{whitham_book}:
 \beq
@@
 two finite difference schemes that solve the viscous Burgers equation:
-forward Euler with optimal time step and correction term (circles)
+forward Euler with OTS-NIDC (circles)
 and forward Euler with suboptimal time step $\dt = \dx^2/4\nu$ and no 
 correction term (squares).
-Both schemes use second-order, central finite difference approximations for
-the advection and diffusion terms.
-The viscous Burgers equation is solved to time $t=2$ on the domain 
-$0 < x < 10$ with the viscosity $\nu$ set to $0.1$.  The initial conditions are 
-taken to be 
+The viscous Burgers equation is solved with initial conditions taken to be 
 $u(x,0) = 1 + \gamma \sqrt{\frac{\nu}{\pi}} \exp\left( -(x-1)^2/4 \nu \right) 
@@
 
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_OTS_soln}} 
-\ \ \ \ \ 
-\scalebox{0.32}{\includegraphics{figures/burgers_eqn_1d_FE_soln}} 
-\caption{Comparison of numerical solutions for the 1D viscous Burgers 
-equation~(\ref{eq:burgers_1d}) using forward Euler time stepping with 
-optimal time step and correction term (left) and forward Euler time 
-stepping with suboptimal time step $\dt = \dx^2/4 \nu$ (right).  
-Both figures compare the analytical solution (solid line) at $t = 2$ on the 
-domain $0 < x < 10$ with the numerical solution (circles) computed using 
-$100$ grid points.  
-For information about the initial/boundary conditions and physical parameters 
-used to compute these solutions, see the caption for 
-Figure~\ref{fig:burgers_1d_error}. 
-}
-\label{fig:burgers_1d_solns}
-\end{center}
-\end{figure}
-
 To analyze the discretization error for (\ref{eq:burgers_1d_FD_scheme}), 
@@
 \eeq
-and the Burgers equation analogue of (\ref{eq:diffusion_eqn_1d_time_err}),
+to obtain the local truncation error for~(\ref{eq:burgers_1d_FD_scheme}): 
 \bea
-  \tu^{n+1}_j = \tu^{n}_j 
-    &+& \dt \left( \nu \frac{\partial^2 \tu^n}{\px^2} 
-                 - \tu^{n}_j \frac{\partial \tu^n}{\px} \right)
-    \nonumber \\
-    &+& \frac{\dt^2}{2} 
-        \left( \nu^2 \frac{\partial^4 \tu^n}{\px^4} 
-             - 2 \nu \tu^n_j\frac{\partial^3 \tu^n}{\px^3} 
-             - 4 \frac{\partial \tu^n}{\px} \frac{\partial^2 \tu^n}{\px^2} 
-             + 2 \tu^n_j \left(\frac{\partial \tu^n}{\px} \right)^2
-             + \left(\tu^n_j\right)^2 \frac{\partial^2 \tu^n}{\px^2} 
-        \right) 
-    \nonumber \\
-    &+& O \left( \dt^3 \right).
-    \label{eq:burgers_1d_time_err}
-\eea
-Together these imply that the leading order term in the local truncation error 
-for~(\ref{eq:burgers_1d_FD_scheme}) is given by
-\bea
-  \tau^n_j &=&
+  & &
       \left( \nu \frac{\partial^4 \tu^n}{\px^4} 
@@
       \right) 
+  \nonumber \\
+  &+& O(\dt^3) + O(\dt \dx^4).
   \label{eq:burgers_1d_err_eqn}
 \eea
-Therefore, the optimal time step is $\dto = \dx^2/6D$ and the correction term 
-is
+Therefore, the optimal time step is $\dto = \dx^2/6\nu$ and the defect 
+correction term is
 \beq
@@
 \eeq 
+Note that the optimal time step eliminates errors introduced by 
+both the diffusion term \emph{and} the nonlinear advection term.
+
 By using the optimal time step and the correction term, the overall error
@@
 in the correction term.  
-Figures~\ref{fig:burgers_1d_error} and \ref{fig:burgers_1d_solns} demonstrate 
-the boost in accuracy obtained by using the optimal time step and adding the
-correction term to the finite difference 
+Figure~\ref{fig:burgers_1d_error} demonstrates the boost in accuracy obtained 
+by applying OTS-NIDC to the finite difference 
 scheme~(\ref{eq:burgers_1d_FD_scheme}).
 
-The Burgers equation example illustrates an important aspect of optimal time 
-step selection: the ease with which OTS selection can be used to boost the 
-order of accuracy depends on the choice of the finite difference scheme.  
-Had we chosen an upwind approximation for the gradient term in the Burgers 
-equation, the correction term would have been more complicated and included 
-terms involving higher-order spatial derivatives than those that appear in the
-original equation.  While it should be possible to handle this situation with 
-OTS selection, it is clearly less desirable.
+The Burgers equation example illustrates an important aspect of OTS-NIDC:
+the ease with which OTS-NIDC can be used to boost the order of accuracy 
+depends on the choice of the finite difference scheme.  Had we chosen an 
+upwind approximation for the gradient term in the Burgers equation, the 
+correction term would have been more complicated and included terms involving 
+higher-order spatial derivatives than those that appear in the original 
+equation.  While it should be possible to handle this situation with OTS-NIDC 
+selection, it is clearly less desirable.
 
@@
 where $\kappa > 0$.  This example illustrates the importance of the choice of
-discretization for the spatial derivative on the applicability of OTS selection.
+discretization for the spatial derivative in order for OTS-NIDC to be
+applicable.
 
@@
 the coefficient on leading-order spatial discretization error to be directly 
-related to the second derivative of the solution $u$ with respect to time.  
-From~(\ref{eq:4th_order_parabolic_eqn_1d}), we see that 
+related to the second derivative of the solution $u$ with respect to time:
 \beq
@@
   + \frac{\partial f}{\pt}
-  \label{eq:4th_order_parabolic_eqn_1d_second_time_derivative},
-\eeq
-which indicates that the leading-order term in the spatial discretization 
-error should involve the eighth-derivative of $u$.  In other words, the
+  \label{eq:4th_order_parabolic_eqn_1d_second_time_derivative}.
+\eeq
+Therefore, the leading-order term in the spatial discretization 
+error should involve the eighth-derivative of $u$, which implies that the
 finite difference approximation to the bilaplacian needs to be at least 
@@
 (\ref{eq:4th_order_parabolic_eqn_1d}):
-forward Euler with optimal time step and correction term (circles),
+forward Euler with OTS-NIDC (circles),
 forward Euler with suboptimal time step and no correction term (squares),
-Crank-Nicholson with a second-order stencil for the bilaplacian (diamonds), 
-and Crank-Nicholson with a fourth-order stencil for the bilaplacian and
-an $O(\dx^2)$ time step (triangles).
+Crank-Nicholson with a fourth-order stencil for the bilaplacian and
+an $O(\dx^2)$ time step (diamonds), and Crank-Nicholson with a second-order 
+stencil for the bilaplacian (triangles).
 Both forward Euler schemes use a fourth-order, central finite difference 
@@
 other at the resolution of the figures.  
-The figure on the left highlights the impact of round-off error associated 
-with matrix inversion on the accuracy of the solution for the two 
-Crank-Nicholson schemes.
+The figure on the left highlights the impact (when $N$ is large) of round-off 
+error associated with matrix inversion on the accuracy of the solution for the 
+two Crank-Nicholson schemes.
 }
@@
 scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}), we find that the
-leading-order local truncation error is given by
-\beq
-  \tau^n_j = 
+local truncation error is given by
+\beq
      \frac{\partial^8 \tu^n}{\px^8} 
@@
         \right)
+    + O(\dt^3) + O(\dt \dx^6)
   \label{eq:4th_order_parabolic_eqn_1d_err_eqn}
@@
 Together, these modifications boost the accuracy of the original
-finite difference scheme from fourth- to sixth-order accurate 
-overall.  As with the previous examples, the spatial derivatives in the 
+finite difference scheme from fourth- to sixth-order.
+As with the previous examples, the spatial derivatives in the 
 correction term should be computed using a finite difference approximation 
@@
 The improved accuracy of (\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) 
-when the optimal time step and correction term are used is illustrated in 
-Figure~\ref{fig:4th_order_parabolic_eqn_1d_error}.
-
-As indicated by the discussion in this section, applying optimal time step 
-selection to high-order PDEs requires the use of high-order discretizations
-of the spatial derivatives.  While the need to derive high-order finite 
-difference stencils may seem like an unnecessary burden, it is important
-to remember that high-order finite difference schemes for spatial derivatives
-are almost always necessary to maximize the efficiency of explicit time
-integration schemes for high-order PDEs\footnote{Because restrictive time 
-step constraints are always a major limitation of explicit schemes for 
-time dependent PDEs with high-order spatial 
+when the optimal time step and defect correction terms are used is 
+illustrated in Figure~\ref{fig:4th_order_parabolic_eqn_1d_error}.
+
+Applying optimal time step selection to high-order PDEs requires the use of 
+high-order discretizations of the spatial derivatives.  While the need to 
+derive high-order finite difference stencils may seem like an unnecessary 
+burden, it is important to remember that high-order finite difference schemes 
+for spatial derivatives are almost always necessary to maximize the efficiency 
+of explicit time integration schemes for high-order 
+PDEs\footnote{Because restrictive time step constraints are always a major 
+limitation of explicit schemes for time dependent PDEs with high-order spatial 
 derivatives~\cite{gko_book,greer_2006}, 
@@
 it is crucial that numerical schemes for high-order PDEs use high-order 
-discretizations for the spatial derivatives.}
+discretizations for the spatial derivatives.}.
 Optimal time step selection merely provides a valuable way to leverage the 
@@
 \renewcommand{\arraystretch}{1.3}
+\centering
 \begin{tabular}{lcccc}
@@
     & \multirow{2}{*}{$O\left( e^{-5/6} \right)$} \\ 
-  \ \ 4th-Order Bilaplacian and OTS & & & & \\
+  \ \ 4th-Order Bilaplacian and OTS-NIDC & & & & \\
   Crank-Nicholson with 
@@
 operator with a fourth-order discretization.  For the forward Euler scheme,
-using an optimal time step drives the computational cost down even more.
+using an OTS-NIDC drives the computational cost down even more.
 The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf} 
@@
 In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
-2nd-order Crank-Nicholson scheme actually outperforms the OTS scheme over 
+2nd-order Crank-Nicholson scheme actually outperforms the OTS-NIDC scheme over 
 the high (but still practically useful) range of errors even though 
-Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS
+Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-NIDC
 scheme should do better.  The origin of this discrepancy is the constant 
-hidden by the big O notation.  In particular, the factor of $7/120$ in 
+hidden by the big-$O$ notation.  In particular, the factor of $7/120$ in 
 the optimal time step for the forward Euler scheme with a fourth-order 
@@
 
-Comparing the best forward Euler and Crank-Nicholson schemes, we see that
-the optimal forward Euler scheme more efficiently utilizes memory at the
-cost of requiring more computation time.  This tradeoff between memory and 
-computation only exists in one space dimension.  In two-dimensions, both 
-schemes require the same amount of compute time; in three or more space 
-dimensions, the forward Euler scheme requires less compute time than the
-Crank-Nicholson scheme.
+Comparing the computational cost of the best forward Euler and Crank-Nicholson 
+schemes, we see that the OTS-NIDC forward Euler scheme more efficiently 
+utilizes memory at the cost of requiring more computation time.  This tradeoff 
+between memory and computation only exists in one space dimension.  In 
+two-dimensions, both schemes require the same amount of compute time; in three 
+or more space dimensions, the forward Euler scheme requires less compute time 
+than the Crank-Nicholson scheme.
 
@@
          Application to PDEs in Multiple Space Dimensions}
-Optimal time step selection can be used to boost the order of accuracy for 
-finite difference schemes in any number of space dimensions.  However, 
-several important issues arise for problems in more than one space dimension.  
-These issues are primarily related to the presence of cross-terms in the 
-discretization error, multiple grid spacing parameters,  and the domain 
-geometry.  In this section, we show how optimal time step selection easily 
-overcomes these issues by using it to construct high-order finite difference
-schemes for the 2D advection equation and the 2D diffusion equation.  
+OTS-NIDC can be used to boost the order of accuracy for finite difference 
+schemes in any number of space dimensions.  However, several important issues 
+arise for problems in more than one space dimension.  These issues are 
+primarily related to the presence of cross-terms in the discretization error, 
+multiple grid spacing parameters,  and the domain geometry.  In this section, 
+we show how OTS-NIDC selection easily overcomes these issues by using it to 
+construct high-order finite difference schemes for the 2D advection equation 
+and the 2D diffusion equation.  
 It is straightforward to generalize the approach presented in this section to 
@@
 incorporating ideas discussed in 
-Sections~\ref{sec:optimal_time_step_selection} and~\ref{sec:applications_1d}.
+Sections~\ref{sec:OTS-NIDC} and~\ref{sec:applications_1d}.
 
@@
 of a finite difference scheme, it is usually safe to ignore the details of 
-these terms because they only appear as higher-order corrections in the error.
+these terms because they can be absorbed within big-$O$ notation.
 However, precise knowledge of the cross-terms is critical when attempting to 
-use optimal time step selection.  Without this knowledge, it is impossible to 
-ensure cancellation of higher-order errors.
-
-Recall that optimal time step selection requires higher-order temporal 
-errors to be directly related to higher-order spatial errors via the PDE.  
-This requirement can be satisfied in one of two ways:
+use OTS-NIDC.  Without this knowledge, it is impossible to 
+ensure cancellation of leading-order errors.
+
+Recall that OTS-NIDC requires the leading-order temporal errors to be directly 
+related to the leading-order spatial errors via the PDE.  This requirement can 
+be satisfied in one of two ways:
 \begin{itemize}
-\item by carefully designing the spatial discretization so that 
-      higher-order spatial errors are directly related to the spatial 
+\item by carefully designing the spatial discretization so that the
+      leading-order spatial errors are directly related to the spatial 
       derivative operators in the PDE; or
-\item by explicitly adding correction terms to cancel out any cross-terms 
-      in the original finite difference scheme that cannot be eliminated 
+\item by explicitly adding defect correction terms to cancel out any 
+      cross-terms in the discretization error that cannot be eliminated 
       by using the PDE.
@@
 can be combined with the finite difference stencils in the original scheme 
-and be viewed as part of an expanded spatial discretization scheme.
+and be viewed as part of an alternate spatial discretization scheme.
 
 \subsection{Linear Advection Equation}
-For our first application of optimal time step selection in multiple
-space dimensions, we consider the two-dimensional advection
-equation
+For our first application of OTS-NIDC in multiple space dimensions, we 
+consider the two-dimensional advection equation
 \beq
@@
 $\dt_{opt} = \dx/A_x = \dy/A_y$ eliminates two of the three leading-order 
-terms in the discretization error.   The last leading-order term can be 
-eliminated by adding the correction term
+terms in the discretization error\footnote{Note that the use of different 
+values for $\dx$ and $\dy$ is consistent with physical intuition.  When 
+computing the upwind derivative in a given coordinate direction, the upstream 
+value we use should be at a distance that is proportional to the flow speed in 
+that direction.  Otherwise, we will be using upstream data that is too far or 
+too near in one of the coordinate directions.}.
+The last leading-order term can be eliminated by adding the defect correction 
+term
 \bea
@@
 boost the order of accuracy even further.  Using the first-order upwind 
-discretization for $\frac{\partial^2 \tu}{\px \py}$
+discretization for $\tu_{xy}$
 \beq
@@
         - \left(u_{i,j-1} - u_{i-1,j-1}\right)}
-        {\dx \dy} 
+        {\dx \dy},
   \label{eq:advection_eqn_2d_upwind_corr_approx}
@@
 which vanishes when the optimal time step and grid spacing ratio are used.
-In other words, when (\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used 
+In other words, when the defect correction 
+(\ref{eq:advection_eqn_2d_upwind_corr_approx}) is used 
 in conjunction with the optimal time step and grid spacing ratio, we obtain
@@
 equation~(\ref{eq:advection_eqn_2d}) using forward Euler time stepping 
-with optimal time step and correction term (left) and forward Euler time 
+with OTS-NIDC (left) and forward Euler time 
 stepping with suboptimal time step (right).  The latter scheme uses a 
@@
 scheme a little further and examine the complete finite difference scheme
-(including correction term):
+(including defect correction term):
 \bea
@@
 As for PDEs in one space dimension, achieving infinite order accuracy 
-through optimal parameter selection is unique to the advection equation.
-In general, optimal parameter selection boosts the accuracy by only a few 
-orders.  To illustrate the typical accuracy boost gained by using OTS
-selection, we apply it to the 2D diffusion equation.  This example underscores 
+through OTS-NIDC is unique to the advection equation.
+To illustrate the typical accuracy boost gained by using OTS-NIDC,
+we apply it to the 2D diffusion equation.  This example underscores 
 the need to carefully choose the discretization for the spatial derivatives 
 for problems in more than one space dimension and demonstrates the ease with 
-which OTS selection can be applied to problems on irregular domains.
+which OTS-NIDC selection can be applied to problems on irregular domains.
 
@@
 
-While it is certainly possible to directly apply optimal time step selection 
+While it is certainly possible to directly apply OTS-NIDC
 to (\ref{eq:diffusion_eqn_2d_5pt_scheme}), we would be forced to use additional
-correction terms to deal with the fact that the leading-order error for 
+defect correction terms to deal with the fact that the leading-order error for 
 (\ref{eq:laplacian_2d_5pt_stencil}), 
@@
 scheme using the nine-point stencil for the 
-Laplacian~\cite{iserles_book,patra_2005}, which does possess a leading-order 
+Laplacian~\cite{iserles_book,patra_2005}, which possesses a leading-order 
 error directly related to the spatial derivative operator in the PDE 
@@
 It is interesting to note that use of a nine-point stencil for the Laplacian 
-naturally arises when directly applying optimal time step selection to the
-five-point scheme if the correction term is interpreted as a modification 
-to the original spatial discretization.
-
-To apply optimal time step selection to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we proceed in the usual way by first analyzing the leading-order terms of the 
-discretization error for the finite difference scheme.  Taylor series 
-expansions and the PDE show that the true solution satisfies 
+naturally arises when directly applying OTS-NIDC to the five-point scheme if 
+the defect correction term is interpreted as a modification to the original 
+spatial discretization.
+
+To apply OTS-NIDC to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we proceed in the 
+usual way by first analyzing the leading-order terms of the local truncation
+error for the finite difference scheme:
 \bea
-  \tu^{n+1}_{i,j} = \tu^{n}_{i,j}
-  + \dt \left(\nabla^2 \tu^{n} +f^n_{i,j} \right)
-  + \frac{(D \dt)^2}{2} \nabla^4 \tu^{n}
-  + \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right)
-  + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_2d_time_err} 
-\eea
-Combining this result with~(\ref{eq:laplacian_2d_9pt_stencil_error}), 
-we find that the leading-order terms in the 
-local truncation error for the finite difference 
-scheme (\ref{eq:diffusion_eqn_2d_9pt_scheme}) are
-\bea
-  \tau^{n}_j = 
   \nabla^4 \tu
       \left[ \frac{\dx^2}{12} - \frac{D \dt}{2}  \right] (D \dt)
-  - \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right).
+  - \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right)
+  + O(\dt^3) + O(\dt \dx^4).
   \label{eq:diffusion_eqn_2d_trunc_err}
 \eea
-As for the 1D diffusion equation, we see that the optimal time step is given 
-by $\dto = \dx^2/6D$ and the correction term is 
-\beq
-  \frac{\dt^2}{2} \left( D\nabla^2 f + \frac{\partial f}{\pt} \right).
-\eeq
+Therefore, the optimal time step is given by $\dto = \dx^2/6D$ and the 
+defect correction term is $\dt^2 \left( D\nabla^2 f + f_t \right)$.
 This choice of time step and correction term yields a scheme that is 
-fourth-order accurate.
+fourth-order accurate.  It is interesting to note that the optimal time step
+for the forward Euler scheme for the diffusion equation is \emph{independent} 
+of the number of spatial dimensions.
 
@@
 \scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_comp_time}} 
-\caption{$L^\infty$ error as a function of number of grid points (left)
-and computation time as a function of $L^\infty$ error (right) for three
-finite difference schemes that solve the 2D diffusion equation:
-forward Euler with optimal time step and correction term (circles), forward 
+\caption{$L^\infty$ error as a function of number of grid points in
+each coordinate direction (left) and computation time as a function of 
+$L^\infty$ error (right) for three finite difference schemes that solve the 
+2D diffusion equation:
+forward Euler with OTS-NIDC (circles), forward 
 Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
@@
 \end{figure}
-The improvement in the accuracy of the forward Euler scheme when the optimal 
-time step is used is shown in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
+The improvement in the accuracy of the forward Euler scheme with OTS-NIDC
+is shown in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
 Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also shows that the 
@@
 theoretical estimates given in Table~\ref{tab:comp_perf_vs_dim}.  As with the
-other examples, the forward Euler scheme with an optimal time step is much 
+other examples, the forward Euler scheme with OTS-NIDC is much 
 more computationally efficient than the two other schemes.
@@
 are harder to design and implement.  
-In this section, we extend the OTS finite difference scheme for the 2D 
+In this section, we extend the OTS-NIDC finite difference scheme for the 2D 
 diffusion equation on regular domains to irregular domains.  Other 
@@
 point is less than $\dx^2$~\cite{gibou_2005}.  For the numerical scheme 
-presented in this paper, this threshold was too low.  Better stability was 
+presented in this article, this threshold was too low.  Better stability was 
 obtained if an $O(\dx)$ threshold was used.
@@
 diffusion equation on a starfish-shaped domain.  The solution is computed 
-using forward Euler time stepping with an optimal time step and correction 
-term.  In the error plot, the dark regions represent points 
-where the error in the solution is larger than $25$\% of the $L^\infty$ error 
-of the solution.  
+using forward Euler time stepping with OTS-NIDC.  In the error plot, the dark 
+regions represent points where the error in the solution is larger than 
+$25$\% of the $L^\infty$ error of the solution.  
 }
@@
 straightforward to compute high-order accurate solutions of the 2D diffusion 
-equation on irregular domains.  We simply use the OTS forward Euler scheme 
-designed for regular domains for grid points in the interior of the domain
-after filling the ghost cells with high-order accurate values.  
+equation on irregular domains.  We simply use the OTS-NIDC forward Euler 
+scheme designed for regular domains for grid points in the interior of the 
+domain after filling the ghost cells with high-order accurate values.  
 Figure~\ref{fig:diffusion_eqn_2d_starfish_domain} shows an example of 
@@
 error in the solution is concentrated near the boundaries of the irregular
-domain.  The improved accuracy of the OTS solution compared to a simple 
+domain.  The improved accuracy of the OTS-NIDC solution compared to a simple 
 forward Euler scheme with a suboptimal time step is illustrated in 
@@
   \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for two
-finite difference schemes that solve the 2D diffusion equation on an 
-irregular domain: forward Euler with optimal time step and correction 
-terms (circles) and forward Euler with suboptimal time step (squares).  
+\caption{$L^\infty$ error as a function of number of grid points in each
+coordinate direction for two finite difference schemes that solve the 2D 
+diffusion equation on an irregular domain: forward Euler with OTS-NIDC 
+(circles) and forward Euler with suboptimal time step (squares).  
 }
@@
 \subsubsection{Diffusion Equation in Higher Dimensions}
-Using optimal time step selection to boost the accuracy of finite difference
+Using OTS-NIDC to boost the accuracy of finite difference
 schemes for the diffusion equation in higher space dimensions is 
@@
 $\dt_{opt} = \dx^2/6D$ for forward Euler time integration and a boost of the 
-order of accuracy from two to four.  Above three dimensions, it is necessary 
-to use an implicit time integration scheme if $\dt_{opt}$ is no longer 
-stable for the forward Euler scheme.
+order of accuracy from two to four.  Above three dimensions, the optimal
+time step is no longer stable, and the forward Euler scheme should be replaced
+with the DuFort-Frankel scheme (see Section~\ref{sec:dufort_frankel}).
 
@@
 \section{\label{sec:summary} Summary} 
-In this paper, we have presented a novel approach for constructing high-order
-finite difference methods for time dependent PDEs \emph{without} requiring
-the use of high-order stencils or high-order time integration schemes.  
-The basic idea is that the leading-order terms in the discretization error
-can be eliminated by carefully choosing the time step and possibly adding
-a few simple correction terms.  It is straightforward to derive both the 
-optimal time step and the correction terms by (1) taking advantage of the 
-regular structure of finite difference schemes and (2) making use of the
-PDE to relate terms in the discretization error.  The boost in the order
-of accuracy achieved when the optimal time step and correction terms are
-used is well worth the extra effort required to explicitly retain and analyze 
-the leading-order terms in the discretization error.
-
-Through many examples, we have demonstrated the utility of optimal time step
-selection to a wide range of problems.  We showed how OTS selection can be 
-used to very easily obtain high-order accurate solutions to many linear 
-and semilinear PDEs in any number of space dimensions on both regular
-and irregular domains.  The examples illustrate the surprisingly high-order
-of accuracy that can be achieved using only low-order discretizations for 
-spatial derivatives and simple forward Euler time integration.
-
-Optimal time step selection is a simple example of the more general technique
+In this article, we have presented OTS-NIDC, a simple approach for 
+constructing high-order finite difference methods for time dependent PDEs 
+\emph{without} requiring the use of high-order stencils or high-order time 
+integration schemes.  The major idea underlying OTS-NIDC is that 
+spatial and temporal discretization errors can be simultaneously eliminated
+by (1) using the PDE to relate terms in the discretization error, (2)
+optimally selecting numerical parameters, and (3) adding a few defect 
+correction terms in a non-iterative manner.  The boost in the order of 
+accuracy achieved by using OTS-NIDC is well worth the extra effort required to 
+derive the optimal time step and defect correction terms.
+
+Through many examples, we have demonstrated the utility of OTS-NIDC
+to several types of finite difference schemes for a wide range of problems.  
+We showed how OTS-NIDC can be used to very easily obtain high-order accurate 
+solutions to many linear and semilinear PDEs in any number of space dimensions 
+on both regular and irregular domains.  The examples illustrate the 
+high-order of accuracy that can be achieved using only low-order 
+discretizations for spatial derivatives and simple forward Euler time 
+integration.  They also demonstrate the applicability of OTS-NIDC to more 
+exotic finite difference schemes, such as the Kreiss-Petersson-Ystr\"om 
+discretization of the second-order wave equation and the DuFort-Frankel scheme 
+for the diffusion equation.
+
+Optimal time step selection is an example of the more general technique
 of optimally selecting numerical parameters to boost the accuracy of 
 numerical methods.  Little research appears to have been done in this area,
-but, as OTS selection demonstrates, optimal selection of numerical parameters 
-can significantly impact the accuracy of numerical methods.  To realize
-the full potential of numerical methods, both optimal time step selection 
-and optimal numerical parameter selection deserve further investigation.
+but as OTS-NIDC selection demonstrates, optimal selection of numerical 
+parameters can significantly impact the accuracy of numerical methods.  To 
+realize the full potential of existing and novel numerical schemes, 
+both optimal time step selection and optimal numerical parameter selection 
+deserve further investigation and attention.
 
@@
 and the Institute for High-Performance Computing (IHPC) in Singapore. 
-The author would like to thank P. Fok, M. Prodanovi\'c, and A. Chiu for 
-helpful suggestions on the manuscript.  The author particularly thanks P. Fok 
-for many enlightening discussions.
+The would like to thank P. Fok and J. Lambers for many enlightening 
+discussions and P. Fok, M. Prodanovi\'c and A. Chiu for helpful suggestions 
+on the manuscript.
 
@@
 
+\section{Importance of High-Accuracy for First Time Step for the KPY Scheme
+         \label{appendix:KPY_analysis} }
+The need for higher-order accuracy when taking the first time step of the KPY 
+scheme for the second-order wave equation can be understood by solving the 
+difference equation 
+\bea
+  \he^{n+1} - 2 \he^{n} + \he^{n-1} = \dt^2 \lambda \he^{n}
+  \label{eq:error_eqn_normal_mode},
+\eea
+for the normal modes of the error, where $\he$ is the coefficient of an 
+arbitrary normal mode of the spatial operator for the error 
+$e \equiv u - \tu$.  Using standard methods for solving linear difference
+equations~\cite{carrier_pearson_book} yields the solution
+\bea
+  \he^n = \he^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-}
+        + \he^0 \kappa_+ \kappa_-
+          \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-},
+  \label{eq:error_eqn_normal_mode_soln}
+\eea
+where $\kappa_\pm$ are the roots of the characteristic equation for
+(\ref{eq:error_eqn_normal_mode})~\cite{kreiss2002}
+\beq
+  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2
+             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
+\eeq
+Notice that the denominator of both terms in
+(\ref{eq:error_eqn_normal_mode_soln}) is $O(\dt)$, which implies that the
+error in the initial conditions is degraded by one temporal order of accuracy
+by the KPY scheme.  Therefore, KPY without OTS-NIDC requires the first time 
+step must be at least third-order accurate; KPY with OTS-NIDC requires the 
+first time step must be at least fifth-order accurate.
+
 %*********************************BIBLIOGRAPHY**********************************
@@
 \begin{thebibliography}{10}
+
+\bibitem{spotz_2001}
+W.~F. Spotz and G.~F. Carey.
+\newblock Extension of high-order compact schemes to time-dependent problems.
+\newblock {\em Numer. Meth. Part. D. E.}, 17:657--672, 2001.
 
@@
   {S}chemes for {E}lliptic {E}quations on {I}rregular {D}omains.
-\newblock {\em SIAM J. SCI. COMPUT.}, 27:346--367, 2005.
+\newblock {\em SIAM J. Sci. Comput.}, 27:346--367, 2005.
 
@@
 
+\bibitem{heidenreich_2007}
+E.~A. Heidenreich, J.~F. Rodr\'iguez, F.~J. Gaspar, and M.~Doblar\'e.
+\newblock Fourth-order compact schemes with adaptive time step for monodomain
+  reaction-diffusion equations.
+\newblock {\em J. Comput. Appl. Math.}, 216:39--55, 2007.
+
 \bibitem{shukla_2007}
@@
 
+\bibitem{leveque_book_2002}
+R.~J. LeVeque.
+\newblock {\em Finite Volume Methods for Hyperbolic Problems}.
+\newblock Cambridge University Press, Cambridge, UK, 2002.
+
 \bibitem{zhao_2006}
@@
 \newblock {\em J. Heat Transf.-Trans. ASME}, 128:409--412, 2006.
+
+\bibitem{pereyra_1968}
+V.~Pereyra.
+\newblock Iterated {D}eferred {C}orrections for {N}onlinear {B}oundary {V}alue
+  {P}roblems.
+\newblock {\em Numer. Math.}, 11:111--125, 1968.
+
+\bibitem{stetter_1978}
+H.~J. Stetter.
+\newblock The {D}efect {C}orrection {P}rinciple and {D}iscretization {M}ethods.
+\newblock {\em Numer. Math.}, 29:425--443, 1978.
+
+\bibitem{gustafsson_2002}
+B.~Gustafsson and L.~Hemmingsson-Fr\"nd\'en.
+\newblock Deferred correction in space and time.
+\newblock {\em J. Sci. Comput.}, 17:541--550, 2002.
+
+\bibitem{kress_2002}
+W.~Kress and B.~Gustafsson.
+\newblock Deferred correction methods for initial boundary value problems.
+\newblock {\em J. Sci. Comput.}, 17:241--251, 2002.
+
+\bibitem{kress_2006}
+W.~Kress.
+\newblock Error estimates for deferred correction methods in time.
+\newblock {\em Apll. Numer. Math}, 57:335--353, 2006.
 
@@
 
-\bibitem{leveque_book_2002}
-R.~J. LeVeque.
-\newblock {\em Finite Volume Methods for Hyperbolic Problems}.
-\newblock Cambridge University Press, Cambridge, UK, 2002.
+\bibitem{kreiss2002}
+H.-O. Kreiss, N.~A. Petersson, and J.~Ystr\"om.
+\newblock {D}ifference {A}pproximations for the {S}econd {O}rder {W}ave
+  {E}quation.
+\newblock {\em SIAM. J. Numer. Anal}, 40:1940--1967, 2002.
 
@@
   {A}vection-{D}iffusion {E}quation.
-\newblock {\em SIAM J. NUMER. ANAL.}, 21:272--284, 1984.
+\newblock {\em SIAM J. Numer. Anal.}, 21:272--284, 1984.
 
@@
 
+\bibitem{carrier_pearson_book}
+G.~F. Carrier and C.~E. Pearson.
+\newblock {\em Ordinary Differential Equations}.
+\newblock Society for Industrial and Applied Mathematics, Philadelphia, 1991.
+
 \end{thebibliography}

### 2009-05-25 papers/OTSPDE/submissions/200905-Applied_Numerical_Math/cover_letter.tex
OTSPDE: - checked in version of OTSPDE paper submitted to Applied Numerical Mathematics

\documentclass[12pt]{letter}

% page settings
\textwidth 6.5in
\textheight 9in
\oddsidemargin -0.05in
\evensidemargin -0.05in
\topmargin -0.85in

\begin{document}

\address{}
\signature{Kevin T. Chu}

\begin{letter}
{
% NORMALLY ADDRESS GOES HERE
}

\opening{Dear Applied Numerical Mathematics Editor/Referee,}

My article titled \emph{Boosting the Accuracy of Finite Difference Schemes via
Optimal Time Step Selection and Non-Iterative Defect Correction} discusses a 
simple technique for boosting the order of accuracy of finite difference 
schemes for time dependent PDEs by choosing an optimal, \emph{fixed} time step
and adding defect correction terms in a \emph{non-iterative} manner.  As 
discussed in the paper, a careful choice of time step can dramatically 
reduce the numerical error of simple finite difference schemes and make 
it possible to achieve high-order accuracy using only formally low-order 
finite difference stencils and time integration schemes.

This paper describes the method for calculating the optimal time step and
defect correcction terms and explains the observed boost in accuracy using 
straightforward numerical analysis arguments.  The technique of optimal time 
step selection is applicable to several types of finite difference schemes
for linear and semilinear PDEs in any number of space dimensions and both 
regular and irregular domains.  Examples of each of these types of problems 
are discussed and analyzed in the paper.

Thank you for your review of the paper.  I look forward to hearing your 
comments and suggestions for improvement of the manuscript.

\closing{Sincerely,}

\end{letter}

\end{document}

### 2009-05-25 papers/OTSPDE/getFigures.sh

@@ -1,4 +1,4 @@
-#!/bin/sh
+#!/bin/bash
 #
 # This script collects the figures from the src directory and places them
 # in the figures directory.
@@ -13,6 +13,7 @@ SRC_DIR="../../src"
 EXAMPLE_DIR_LIST="scalar_PDEs/1d/burgers_eqn/single_hump                       \
                   scalar_PDEs/1d/diffusion_eqn                                 \
                   scalar_PDEs/1d/fourth_order_parabolic                        \
+                  scalar_PDEs/1d/wave_eqn                                      \
                   scalar_PDEs/2d/advection_eqn                                 \
                   scalar_PDEs/2d/diffusion_eqn/square_domain                   \
                   scalar_PDEs/2d/diffusion_eqn/irregular_domain/starfish_domain\

### 2010-01-18 src/scalar_PDEs/1d/diffusion_eqn/nonperiodic_src_term/demo.m
Added nonperiodic source term example to 1d diffusion equation.

@@ -1,0 +1,92 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script demonstrates the use of 
+% solveDiffusionEqnForwardEuler1d(),
+% solveDiffusionEqnForwardEulerOTS1d(), and
+% solveDiffusionEqnCrankNicholson1d().
+%  
+% Kevin T. Chu
+% 2008 April
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+
+% physical parameters
+D = 1.0;  % diffusion coefficient 
+
+% time integration parameters
+t_init  = 0.0;
+t_final = 0.05;
+
+% set dx and dt
+N = 100;
+N = 200;
+N = 400;
+N = 800;
+dx = 1/N;
+dt_FE = dx^2/4/D;
+dt_CN = dx/8/D;
+
+% solve diffusion equation using forward Euler with OTS
+debug_on = 0;
+[u_FE_OTS, u_exact, x] = solveDiffusionEqnForwardEulerOTS1d( ...
+                             D, ...
+                             dx, ...
+                             t_final, ...
+                             debug_on);
+
+% solve diffusion equation using forward Euler
+%[u_FE, u_exact, x] = solveDiffusionEqnForwardEuler1d( ...
+%                             D, ...
+%                             dx, dt_FE, ...
+%                             t_final, ...
+%                             debug_on);
+%
+%% solve diffusion equation using Crank-Nicholson
+%[u_CN, u_exact, x] = solveDiffusionEqnCrankNicholson1d( ...
+%                             D, ...
+%                             dx, dt_CN, ...
+%                             t_final, ...
+%                             debug_on);
+
+% compute error
+err_FE_OTS = u_FE_OTS-u_exact;
+err_L_inf_FE_OTS = norm(err_FE_OTS, 'inf')
+%err_FE = u_FE-u_exact;
+%err_L_inf_FE = norm(err_FE,'inf')
+%err_CN = u_CN-u_exact;
+%err_L_inf_CN = norm(err_CN,'inf')
+
+% plot results
[…]

### 2010-01-18 src/scalar_PDEs/1d/diffusion_eqn/nonperiodic_src_term/solveDiffusionEqnCrankNicholson1d.m

@@ -1,0 +1,195 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveDiffusionEqnCrankNicholson1d() computes the solutions of the 1d 
+% diffusion equation 
+%
+%   u_t = D u_xx + f(x)
+%
+% on the domain 0 < x < 1 with f(x) = 2*t - 42*x^5 - 6*x and boundary
+% conditions given by u(0) = 0 and u'(1) = 10.  The initial condition is
+% taken to be u(x,0) = x^3 + x^3.  The numerical solution is computed on 
+% a node-centered grid using Crank-Nicholson time integration with a 
+% second-order central difference approximation to the Laplacian.  A 
+% suboptimal time step is used.
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveDiffusionEqnCrankNicholson1d(D, ...
+%                                       dx, dt, ...
+%                                       t_final, ...
+%                                       debug_on, timing_on)
+%
+% Arguments:
+% - D:                   diffusion coefficient
+% - dx:                  grid spacing
+% - dt:                  time step
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
+% - timing_data:         time used to compute the solution (set up time
[…]

### 2010-01-18 src/scalar_PDEs/1d/diffusion_eqn/nonperiodic_src_term/solveDiffusionEqnForwardEuler1d.m

@@ -1,0 +1,183 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveDiffusionEqnForwardEuler1d() computes the solutions of the 1d 
+% diffusion equation 
+%
+%   u_t = D u_xx + f(x)
+%
+% on the domain 0 < x < 1 with f(x) = 2*t - 42*x^5 - 6*x and boundary
+% conditions given by u(0) = 0 and u'(1) = 10.  The initial condition is
+% taken to be u(x,0) = x^3 + x^3.  The numerical solution is computed on 
+% a node-centered grid using forward Euler time integration with a 
+% second-order central difference approximation to the Laplacian.  
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveDiffusionEqnForwardEuler1d(D, ...
+%                                     dx, dt, ...
+%                                     t_final, ...
+%                                     debug_on, timing_on)
+%
+% Arguments:
+% - D:                   diffusion coefficient 
+% - dx:                  grid spacing
+% - dt:                  time step
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
+% - timing_data:         time used to compute the solution (set up time
+%                        excluded).  If timing is not activated, timing_data 
+%                        is set to -1.
+%
+%
[…]

### 2010-01-18 src/scalar_PDEs/1d/diffusion_eqn/nonperiodic_src_term/solveDiffusionEqnForwardEulerOTS1d.m

@@ -1,0 +1,223 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveDiffusionEqnForwardEulerOTS1d() computes the solutions of the 1d 
+% diffusion equation 
+%
+%   u_t = D u_xx + f(x)
+%
+% on the domain 0 < x < 1 with f(x) = 2*t - 42*x^5 + 6*x and boundary
+% conditions given by u(0) = 0 and u'(1) = 4.  The initial condition
+% is taken to be u(x,0) = x^7 - x^3.  The numerical solution is computed on 
+% a node-centered grid using forward Euler time integration with a 
+% second-order central difference approximation to the Laplacian.  The 
+% optimal time step of dx^2/(6D) and correction terms are used.
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveDiffusionEqnForwardEulerOTS1d(D, ...
+%                                        dx, ...
+%                                        t_final, ...
+%                                        debug_on, timing_on)
+%
+% Arguments:
+% - D:                   diffusion coefficient 
+% - dx:                  grid spacing
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
[…]

### 2010-01-18 src/scalar_PDEs/1d/reaction_diffusion_eqn/demo.m
Added reaction diffusion equation example.
* demonstrates time interpolation for systems of PDEs.

@@ -1,0 +1,80 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script demonstrates the use of 
+% solveRxnDiffusionEqn1dForwardEuler(),
+% solveRxnDiffusionEqn1dForwardEulerOTS(), and
+% solveRxnDiffusionEqn1dNSFD().
+%  
+% Kevin T. Chu
+% 2009 April
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+
+% time integration parameters
+t_final = 5.0;
+
+% set dx and dt
+x_lo = -5;
+x_hi =  5;
+N = 50;
+dx = (x_hi-x_lo)/N;
+dt_FE = 0.25*dx^2;
+dt_NSFD = 0.5*dx^2;
+
+% debug flag
+debug_on = 0;
+
+% solve diffusion equation using forward Euler with OTS
+[u_FE_OTS, u_exact, x] = solveRxnDiffusionEqn1dForwardEulerOTS( ...
+                             dx, t_final, debug_on);
+
+% solve reaction-diffusion equation using forward Euler
+[u_FE, u_exact, x] = solveRxnDiffusionEqn1dForwardEuler(dx, dt_FE, t_final, ...
+                                                        debug_on);
+
+% solve reaction-diffusion equation using non-standard finite difference scheme
+%[u_NSFD, u_exact, x] = solveRxnDiffusionEqn1dNSFD(dx, dt_NSFD, t_final, ...
+%                                                  debug_on);
+
+% compute error
+err_FE_OTS = u_FE_OTS-u_exact;
+err_L_inf_FE_OTS = norm(err_FE_OTS,'inf')
+err_FE = u_FE-u_exact;
+err_L_inf_FE = norm(err_FE,'inf')
+%err_NSFD = u_NSFD-u_exact;
+%err_L_inf_NSFD = norm(err_NSFD,'inf')
+
+% plot results
+figure(1); clf;
+plot(x,u_FE_OTS,'bo')
+hold on;
+plot(x,u_exact,'r')
+title('Forward Euler OTS Solution')
+
+figure(2); clf;
[…]

### 2010-01-18 src/scalar_PDEs/1d/reaction_diffusion_eqn/genFigs_RxnDiffusionEqn1d.m

@@ -1,0 +1,286 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script generates figures of the results for the 
+% reaction-diffusion equation. 
+%  
+% Kevin T. Chu
+% 2009 April
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Set parameters for computation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+set(0,'DefaultAxesFontSize',18,'DefaultAxesFontName','Helvetica')
+set(0,'DefaultLineLineWidth',2)
+set(0,'DefaultTextFontSize',18,'DefaultTextFontName','Helvetica')
+
+% set print format
+use_color_figures = 1;
+print_suffix = 'eps';
+if use_color_figures
+  print_format = 'epsc';
+else
+  print_format = 'eps';
+end
+fig_dir = 'figures';
+if ~exist(fig_dir, 'dir')
+  mkdir(fig_dir);
+end
+
+% set flag for loading data from saved files (instead of recomputing solution)
+use_saved_data = 1;
+data_dir = 'data-rxn_diffusion_eqn_1d';
+if ~exist(data_dir, 'dir')
+  mkdir(data_dir);
+end
+
+% set simulation parameters
+debug_on = 0;
+timing_on = 1;
+
+% time integration parameters
+t_final = 5.0;
+
+% grid parameters
+x_lo = -5;
+x_hi =  5;
+
+% grid sizes to collect data on
+grid_sizes = [25 50 100 200 400 800];
+
+% allocate memory for errors
+err_FE_OTS         = zeros(size(grid_sizes));
+err_FE             = zeros(size(grid_sizes));
+err_NSFD           = zeros(size(grid_sizes));
+
+% start clock for timing plot generation time
+t_start = cputime;
+
+
[…]

### 2010-01-18 src/scalar_PDEs/1d/reaction_diffusion_eqn/genPresoFigs_RxnDiffusionEqn1d.m

@@ -1,0 +1,399 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script generates figures of the results for the 
+% reaction-diffusion equation for presentation purposes.
+%  
+% Kevin T. Chu
+% 2009 April
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Set parameters for computation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+set(0,'DefaultAxesFontSize',18,'DefaultAxesFontName','Helvetica')
+set(0,'DefaultLineLineWidth',2)
+set(0,'DefaultTextFontSize',18,'DefaultTextFontName','Helvetica')
+
+% set print format
+print_format = 'png';
+fig_dir = 'figures';
+if ~exist(fig_dir, 'dir')
+  mkdir(fig_dir);
+end
+
+% set flag for loading data from saved files (instead of recomputing solution)
+use_saved_data = 1;
+data_dir = 'data-diffusion_eqn_1d_src';
+if ~exist(data_dir, 'dir')
+  mkdir(data_dir);
+end
+
+% set simulation parameters
+debug_on = 0;
+timing_on = 1;
+
+% source term type
+source_term_type = 1;
+
+% physical parameters
+D = 0.5;  % diffusion coefficient 
+
+% boundary conditions 
+u_0 = 1;
+dudx_1 = 0.5;
+
+% time integration parameters
+t_init  = 0.0;
+t_final = 0.5;
+
+% grid sizes to collect data on
+grid_sizes = [25 50 100 200 400 800];
+
+% allocate memory for errors
+err_FE_OTS         = zeros(size(grid_sizes));
+err_FE_OTS_no_corr = zeros(size(grid_sizes));
+err_FE             = zeros(size(grid_sizes));
+err_CN             = zeros(size(grid_sizes));
+
[…]

### 2010-01-18 src/scalar_PDEs/1d/reaction_diffusion_eqn/solveRxnDiffusionEqn1dForwardEuler.m

@@ -1,0 +1,169 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveRxnDiffusionEqn1dForwardEuler() computes the solutions of the 
+% following 1d reaction-diffusion equation 
+%
+%   u_t = u_xx - u*(u-2)*(u+1)
+%
+% on the domain -5 < x < 5 with boundary conditions imposed from the 
+% exact travelling-wave solution:  
+%
+%   u(x,t) = 2*(exp(eta) - 1)/(exp(eta) + 2),
+%
+% where eta = (k*x + w*t) with k = 3/sqrt(2) and w = 3/2, which implies a 
+% wave speed of -1/sqrt(2).  The initial conditions at t = 0 is given by 
+%
+%   u(x,0) = 2*(exp(k*x) - 1)/(exp(k*x) + 2).
+%
+% The numerical solution is computed on a node-centered grid using 
+% forward Euler time integration with a second-order central difference 
+% approximation to the Laplacian.  
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveRxnDiffusionEqn1dForwardEuler(dx, dt, ...
+%                                        t_final, ...
+%                                        debug_on, timing_on)
+%
+% Arguments:
+% - dx:                  grid spacing
+% - dt:                  time step
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
+% - timing_data:         time used to compute the solution (set up time
+%                        excluded).  If timing is not activated, timing_data 
[…]

### 2010-01-18 src/scalar_PDEs/1d/reaction_diffusion_eqn/solveRxnDiffusionEqn1dForwardEulerOTS.m

@@ -1,0 +1,179 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveRxnDiffusionEqn1dForwardEulerOTS() computes the solutions of the 
+% following 1d reaction-diffusion equation 
+%
+%   u_t = u_xx - u*(u-2)*(u+1)
+%
+% on the domain -5 < x < 5 with boundary conditions imposed from the 
+% exact travelling-wave solution:  
+%
+%   u(x,t) = 2*(exp(eta) - 1)/(exp(eta) + 2),
+%
+% where eta = (k*x + w*t) with k = 3/sqrt(2) and w = 3/2, which implies a 
+% wave speed of -1/sqrt(2).  The initial conditions at t = 0 is given by 
+%
+%   u(x,0) = 2*(exp(k*x) - 1)/(exp(k*x) + 2).
+%
+% The numerical solution is computed on a node-centered grid using 
+% forward Euler time integration with a second-order central difference 
+% approximation to the Laplacian.  The optimal time step of dt = dx^2/6
+% and correction terms are used to boost the order of accuracy of the 
+% solution.
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveRxnDiffusionEqn1dForwardEulerOTS(dx, ...
+%                                           t_final, ...
+%                                           debug_on, timing_on)
+%
+% Arguments:
+% - dx:                  grid spacing
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
[…]

### 2010-01-18 src/scalar_PDEs/1d/reaction_diffusion_eqn/solveRxnDiffusionEqn1dNSFD.m

@@ -1,0 +1,173 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveRxnDiffusionEqn1dNSFD() computes the solutions of the following 1d 
+% reaction-diffusion equation 
+%
+%   u_t = u_xx - u*(u-2)*(u+1)
+%
+% on the domain -5 < x < 5 with boundary conditions imposed from the 
+% exact travelling-wave solution:  
+%
+%   u(x,t) = 2*(exp(eta) - 1)/(exp(eta) + 2),
+%
+% where eta = (k*x + w*t) with k = 3/sqrt(2) and w = 3/2, which implies a 
+% wave speed of -1/sqrt(2).  The initial conditions at t = 0 is given by 
+%
+%   u(x,0) = 2*(exp(k*x) - 1)/(exp(k*x) + 2).
+%
+% The numerical solution is computed on a node-centered grid using 
+% a non-standard finite difference scheme.  
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solveRxnDiffusionEqn1dNSFD(dx, dt, t_final, ...
+%                                debug_on, timing_on)
+%
+% Arguments:
+% - dx:                  grid spacing
+% - dt:                  time step
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
+% - timing_data:         time used to compute the solution (set up time
+%                        excluded).  If timing is not activated, timing_data 
+%                        is set to -1.
+%
+%
[…]

### 2010-01-18 src/scalar_PDEs/1d/fourth_order_parabolic/solve4thOrderParabolicEqn1dCrankNicholson2ndOrder.m
Added examples for fourth-order parabolic equation using Crank-Nicholson method.

@@ -1,0 +1,273 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solve4thOrderParabolicEqn1dCrankNicholson2ndOrder() computes the solutions 
+% of the 4th-order parabolic equation in one spatial dimensions
+%
+%   u_t = -u_xxxx + f
+%
+% where
+%
+%   f(x,t) = -20*pi*sin(2*pi*x)*sin(10*pi*t) ...
+%          + 32*pi^4*sin(2*pi*x)*cos(10*pi*t) ...
+%          + 10*pi^2*sin(5*pi*x)*exp(-10*pi^2*t) ...
+%          + 625*pi^4*sin(5*pi*x)*(2-exp(-10*pi^2*t))
+%
+% on the domain -1 < x < 1 with homogeneous Dirichlet boundary conditions 
+% imposed at all boundaries.  The numerical solution is computed on a 
+% node-centered grid using Crank-Nicholson time integration with a 
+% second-order central difference approximation to the bilaplacian.  
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solve4thOrderParabolicEqn1dCrankNicholson2ndOrder(dx, dt, ...
+%                                                       t_init, t_final, ...
+%                                                       debug_on, timing_on)
+%
+% Arguments:
+% - dx:                  grid spacing
+% - dt:                  time step
+% - t_init:              initial time
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
[…]

### 2010-01-18 src/scalar_PDEs/1d/fourth_order_parabolic/solve4thOrderParabolicEqn1dCrankNicholson4thOrder.m

@@ -1,0 +1,275 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solve4thOrderParabolicEqn1dCrankNicholson4thOrder() computes the solutions 
+% of the 4th-order parabolic equation in one spatial dimensions
+%
+%   u_t = -u_xxxx + f
+%
+% where
+%
+%   f(x,t) = -20*pi*sin(2*pi*x)*sin(10*pi*t) ...
+%          + 32*pi^4*sin(2*pi*x)*cos(10*pi*t) ...
+%          + 10*pi^2*sin(5*pi*x)*exp(-10*pi^2*t) ...
+%          + 625*pi^4*sin(5*pi*x)*(2-exp(-10*pi^2*t))
+%
+% on the domain -1 < x < 1 with homogeneous Dirichlet boundary conditions 
+% imposed at all boundaries.  The numerical solution is computed on a 
+% node-centered grid using Crank-Nicholson time integration with a 
+% fourth-order central difference approximation to the bilaplacian.  
+%
+% USAGE:
+%   function [u, u_exact, x, timing_data] = ...
+%     solve4thOrderParabolicEqn1dCrankNicholson4thOrder(dx, dt, ...
+%                                                       t_init, t_final, ...
+%                                                       debug_on, timing_on)
+%
+% Arguments:
+% - dx:                  grid spacing
+% - dt:                  time step
+% - t_init:              initial time
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+% - timing_on:           flag indicating whether timing information
+%                        should be collected.  To activate timing,
+%                        set timing_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
[…]

### 2010-01-18 src/scalar_PDEs/2d/fourth_order_parabolic/forward_euler_2d.m
Added initial version of 2D fourth-order parabolic example.

@@ -1,0 +1,311 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script computes the solutions of the 4th-order parabolic
+% equation in two spatial dimensions
+%
+%   u_t = -(u_xxxx + 2 u_xxyy + u_yyyy) + f
+%
+% where
+%
+%   f(x,y,t) = -3*sin(2*pi*x)*sin(5*pi*y)*sin(3*t)
+%            + 841*pi^4*sin(2*pi*x)*sin(5*pi*y)*cos(3*t)
+%            + 10*pi^2*sin(pi*x)*sin(3*pi*y)*exp(-10*pi^2*t)
+%            + 100*pi^4*sin(pi*x)*sin(3*pi*y)*(1-exp(-10*pi^2*t)
+%
+% on the domain -1 < x,y < 1 with homogeneous Dirichlet boundary conditions 
+% imposed at all boundaries.  The numerical solution is computed on a 
+% node-centered grid using forward Euler time integration with a 
+% fourth-order central difference approximation to the bilaplacian.  The 
+% optimal time step for this problem is dt = 7/120*dx^4.
+%
+% NOTE:
+% - For the purposes of illustration, the boundary conditions for the 
+%   example are imposed by filling the ghostcells from the exact 
+%   solution.
+%
+% Kevin T. Chu
+% 2007 October
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Set parameters for computation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+
+% Set timing_run to 1 to disable error calculation and plotting
+% during time integration loop.
+timing_run = 0;
+
+% time integration parameters
+t_init  = 0.0;
+t_final = 0.01;
+
+% number of grid cells to use in computation
+% NOTE: number of grid points in computational domain is (N_cell+1)
+N_cell = 50;
+
[…]

### 2010-01-18 src/variable_coefs/diffusion_eqn_non_div_form/demo.m
Added variable-coefficient example with diffusion equation in non-divergence form.

@@ -1,0 +1,66 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script demonstrates the use of solveDiffusionEqn1dFE(), 
+% solveDiffusionEqn1dFE_OTS_OptGrid().
+%  
+% Kevin T. Chu
+% 2009 February
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+
+% simulation parameters
+debug_on = 0;
+
+% time integration parameters
+% NOTE: t_init = 0.0
+t_final = 0.1;
+
+% set dx and dt
+x_lo = 0.0;
+x_hi = 1.0;
+N = 50;
+dx = (x_hi-x_lo)/N;
+dt_FE = 0.5*dx^2/9;
+
+% solve 1d diffusion equation using FE time integration with OTS 
+% and optimal grid
+[u_FE_OTS_OptGrid, u_exact_OTS_OptGrid, x_OTS_OptGrid] = ...
+  solveDiffusionEqn1dFE_OTS_OptGrid(N, t_final, debug_on);
+
+% solve 1d diffusion equation using FE time integration without OTS
+[u_FE, u_exact, x] = solveDiffusionEqn1dFE(N, dt_FE, t_final, debug_on);
+
+
+% compute error
+err_FE = u_FE-u_exact;
+err_L_inf_FE = norm(err_FE,'inf')
+err_FE_OTS_OptGrid = u_FE_OTS_OptGrid-u_exact_OTS_OptGrid;
+err_L_inf_FE_OTS_OptGrid = norm(err_FE_OTS_OptGrid,'inf')
+
+% plot results
+figure(1); clf;
+plot(x,u_FE,'bo')
+hold on;
+plot(x,u_exact,'r')
+title('FE Solution')
+axis([x_lo x_hi 0 4]);
+
+figure(2); clf;
+plot(x,err_FE);
+title('Error in FE Solution')
+
+figure(3); clf;
+plot(x_OTS_OptGrid,u_FE_OTS_OptGrid,'bo')
+hold on;
[…]

### 2010-01-18 src/variable_coefs/diffusion_eqn_non_div_form/genFigs_DiffusionEqn1d.m

@@ -1,0 +1,265 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% This MATLAB script generates figures of the results for a
+% 1d variable-coefficient diffusion equation with source term. 
+%  
+% Kevin T. Chu
+% 2009 February
+% Serendipity Research
+% 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Set parameters for computation
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+clear 
+format long
+format compact
+set(0,'DefaultAxesFontSize',18,'DefaultAxesFontName','Helvetica')
+set(0,'DefaultLineLineWidth',2)
+set(0,'DefaultTextFontSize',18,'DefaultTextFontName','Helvetica')
+
+% set print format
+use_color_figures = 1;
+print_suffix = 'eps';
+if use_color_figures
+  print_format = 'epsc';
+else
+  print_format = 'eps';
+end
+fig_dir = 'figures';
+if ~exist(fig_dir, 'dir')
+  mkdir(fig_dir);
+end
+
+% set flag for loading data from saved files (instead of recomputing solution)
+use_saved_data = 0;
+data_dir = 'data-var_coef_diffusion_eqn_1d';
+if ~exist(data_dir, 'dir')
+  mkdir(data_dir);
+end
+
+% debug switch
+debug_on = 0;
+
+% grid parameters
+x_lo = 0;
+x_hi = 1;
+
+% time integration parameters
+% NOTE: t_init = 0.0
+t_final = 0.1;
+
+% grid sizes to collect data on
+grid_sizes = [25 50 100 200 400 800]; 
+
+% allocate memory for errors
+err_FE                 = zeros(size(grid_sizes));
+
+% start clock for timing plot generation time
+t_start = cputime;
+
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Compute errors
[…]

### 2010-01-18 src/variable_coefs/diffusion_eqn_non_div_form/solveDiffusionEqn1dFE.m

@@ -1,0 +1,157 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveDiffusionEqn1dFE() computes the solutions of the 1d diffusion 
+% equation with variable diffusivity
+%
+%   u_t = D*u_xx + f(x,t)
+%
+% on the domain 0 < x < 1 subject to the initial conditions
+%
+%   u(x,0)   = 2 + sin(3*pi*x)
+%
+% and homogeneous Dirichlet boundary conditions.  The diffusivity and 
+% source term are given by
+%
+%   D(x) = (2+sin(2*pi*x))^2
+%
+%   f(x,t) = -3*pi*cos(3*pi*(x-t))
+%          + 9*pi^2*(2+sin(2*pi*x))^2*sin(3*pi*(x-t))
+%
+% The analytical solution to this problem is 
+%
+%   u(x,t) = 2+sin(3*pi*(x-t))
+%
+% The numerical solution is computed on a node-centered grid using 
+% central difference operators for the Laplacian and gradient term of
+% the expanded equation and forward Euler time integration.
+%
+% USAGE:
+%   function [u, u_exact, x] = solveDiffusionEqn1dFE(N, dt, ...
+%                                                    t_final, ...
+%                                                    debug_on)
+%
+% Arguments:
+% - N:                   number of grid cells to use for computation.
+%                        NOTE:  num grid points = (num grid cells + 1)
+% - dt:                  time step
+% - t_final:             final time
+% - debug_on:            flag indicating whether debugging information
+%                        should be displayed.  To turn on debugging,
+%                        set debug_on to 1.
+%                        (default = 0)
+%
+% Return values:
+% - u:                   numerical solution
+% - u_exact:             analytical solution
+% - x:                   grid points
+%
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% Kevin T. Chu
+%
[…]

### 2010-01-18 src/variable_coefs/diffusion_eqn_non_div_form/solveDiffusionEqn1dFE_OTS_OptGrid.m

@@ -1,0 +1,204 @@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+% solveDiffusionEqn1dFE_OTS_OptGrid() computes the solutions of the 1d 
+% diffusion equation with variable diffusivity
+%
+%   u_t = D*u_xx + f(x,t)
+%
+% on the domain 0 < x < 1 subject to the initial conditions
+%
+%   u(x,0)   = 2 + sin(3*pi*x)
+%
+% and homogeneous Dirichlet boundary conditions.  The diffusivity and 
+% source term are given by
+%
+%   D(x) = (2+sin(2*pi*x))^2
+%
+%   f(x,t) = -3*pi*cos(3*pi*(x-t))
+%          + 9*pi^2*(2+sin(2*pi*x))^2*sin(3*pi*(x-t))
+%
+% The analytical solution to this problem is 
+%
+%   u(x,t) = 2+sin(3*pi*(x-t))
+%
+% The numerical solution is computed on the variable grid size mesh induced 
+% by taking a uniform mesh in a transformed domain where the diffusion 
+% coefficient is constant.  Forward Euler time integration is used to 
+% advance the solution in time.  The optimal time step of dt = (dy/D_bar)^2/6
+% is used.  In this expression, dy is the grid spacing of the uniform mesh
+% on the transformed domain and D_bar is the harmonic mean of the square-root
+% of the diffusion coefficient.  The expression for D_bar is given below.
+%
+% On the transformed domain 
+%
+%   y = 1/pi * ( atan( (2*tan(pi*x)+1)/sqrt(3) ) - pi/6 )
+%
+% the variable coefficient diffusion equation above becomes a diffusion 
+% equation that has a constant coefficient leading-order spatial derivative 
+% term:
+%
+%   u_tt = D_bar^2 u_yy + ...
+%
+% where D_bar has a value of sqrt(3).  Note that in the change of variables,
+% the appropriate branch of atan() must be used to avoid a discontinuity in 
+% y.
+%
+%
+% USAGE:
+%   function [u, u_exact, x] = solveDiffusionEqn1dFE_OTS_OptGrid(N, ...
[…]

### 2010-01-18 papers/OTS-NavierStokes/OTS-NavierStokes.tex
Added OTS-NavierStokes paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2008/11/24
% First completed draft: 2008/??/??
% Submission draft: 2008/??/??
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[fleqn,12pt,twoside]{article}
\usepackage[headings]{espcrc1}
\usepackage{amsmath}
\usepackage{graphicx}
\PassOptionsToPackage{ctagsplt,righttag}{amsmath}
\usepackage{multirow}

%********************************** MACROS ************************************

% COMMON EXPRESSIONS
\def\eg{\emph{e.g., }}
\def\ie{\emph{i.e., }}
\def\etal{\emph{et al.}}
\def\etc{\emph{etc.}}

% UTILITY COMMANDS
\newcommand{\bc}{\begin{center}}
\newcommand{\ec}{\end{center}}
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}
\newcommand{\ba}{\begin{array}}
\newcommand{\ea}{\end{array}}

% NOTATION
\def\div{\ensuremath{\nabla \cdot}}
\def\grad{\ensuremath{\nabla}}
\def\lapl{\ensuremath{\nabla^2}}
\def\bilapl{\ensuremath{\nabla^4}}

% PDE MACROS
\def\pt{\partial t}
\def\px{\partial x}
\def\py{\partial y}
\def\pz{\partial z}
\def\tU{\tilde{U}}
\def\tP{\tilde{P}}
\def\tu{\tilde{u}}
\def\tv{\tilde{v}}
\def\tw{\tilde{w}}

% NUMERICS MACROS
\def\dt{\Delta t}
\def\dx{\Delta x}
\def\dy{\Delta y}
\def\dz{\Delta z}
\def\dto{\dt_{opt}}

%****************************** SIZE DECLARATIONS *****************************
\topmargin -0.7in

%****************************** DOCUMENT BEGIN ********************************

\begin{document}
\bibliographystyle{unsrt}

\title{A Fourth-Order Finite Difference Scheme for the 
Incompressible Navier-Stokes Equations via Optimal Time Step Selection}

\author{
K. T. Chu\address{Vitamin D, Inc., Menlo Park, CA 94025}$^,$\address[IHPC]{Institute of High Performance Computing, A*STAR, Singapore, Singapore}
%D. V. Le\addressmark[IHPC],
%and K. H. Chiam\addressmark[IHPC]
}

\runtitle{Fourth-Order Navier Stokes Scheme via Optimal Time Step Selection}
\runauthor{??}

\maketitle

\noindent \rule{6.3in}{1pt}

\begin{abstract}
place holder...
\end{abstract}

\noindent \rule{6.3in}{1pt}

% KEYWORDS
% optimal time step, Navier-Stokes equations, computational fluid dynamics

\section{Introduction}
place holder...
Prior work~\cite{bruger_2005,shukla_2007}

Optimal time step selection~\cite{chu_2009}

\section{Method}

\subsection{Highlights}
The propsed fourth-order scheme for the incompressible Navier-Stokes equations
only requires forward Euler time integration and a fourth-order accurate
solution of the the pressure correction equation.  Does not require 
correction terms that involve spatial or temporal derivatives higher than
those that are present in the original equations.

There are two main complication in deriving the scheme:
(1) choosing an appropriate discretization for the nonlinear convection
term and 
(2) calculating the correction terms in the forward Euler scheme.  These
two derivations are closely interrelated.

We can use a standard staggered grid configuration with $\dx = \dy = \dz $ 
to avoid unphysical oscillations in the pressure.

\subsection{Method}
Incompressible Navier-Stokes equations:
\bea
  \frac{\partial U}{\pt} + \left(U \cdot \grad \right) U 
    &=& -\grad P + \frac{1}{Re} \lapl U + f
    \\
  \div U &=& 0
\eea
Pressure equation:
\bea
  \lapl P = \div f - \div \left( (U \cdot \grad) U \right)
\eea

Let variables with and without tildes denote the true and numerical solution,
respectively.

\subsubsection{Time discretization}
Forward Euler time integration for $U$:
\beq
  \frac{\partial U^n}{\pt} = \frac{U^{n+1}-U^{n}}{\dt} 
\eeq
True solution satisfies
\beq
  \frac{\partial \tU^n}{\pt} = 
    \frac{\tU^{n+1}-\tU^{n}}{\dt} - \frac{\dt}{2} \tU^n_{tt} + O(\dt^2)
  \label{eq:forward_Euler_error}
\eeq

For the Laplacian term, we use any stencil which has an isotropic error:
\beq
  \lapl \tU = L_{iso} \tU - \frac{\dx^2}{12} \bilapl \tU + O(\dx^4).
\eeq
In 2D, the standard nine-point stencil works fine.  In 3D, any of the
isotropic stencils in Patra and Karttunen~\cite{patra_2005} would work.

As with the diffusion equation, the optimal time step $\dt_{opt}$ will
come from a balance between the bilaplacian terms that appear in the 
leading-order terms in the spatial and temporal errors.  This leads to 
an optimal time step $\dt_{opt} = Re \dx^2/6$.

The discretization of the other spatial derivatives will be specially
derived so that they exactly cancel out terms in the leading-order temporal
error when the optimal time step is used.

From (\ref{eq:forward_Euler_error}), we see that the leading-order temporal
error involves $\tU_{tt}$:
\bea
  \tU_{tt} &=&
    -\left(\tU_t \cdot \grad \right) \tU 
    -\left(\tU \cdot \grad \right) \tU_t 
    -\grad P_t
    + \frac{1}{Re} \lapl \tU_t
    + f_t
    \nonumber \\
    &=& \frac{1}{Re^2} \bilapl \tU - \frac{1}{Re} \lapl \grad P  
    \nonumber \\
    & & - \frac{1}{Re} 
    \left[ \lapl \left( (\tU \cdot \grad) \tU \right)
         + \left(\tU \cdot \grad \right) \lapl \tU 
         + \left(\lapl \tU \cdot \grad \right) \tU 
    \right]
    \nonumber \\
    & & + \left(\tU \cdot \grad \right) \left(\tU \cdot \grad \right) \tU
    + \left( (\tU \cdot \grad) \tU \cdot \grad \right) \tU
    \nonumber \\
    & & - \left(\tU \cdot \grad \right) \left(-\grad P + f \right)
        - \left((-\grad P + f) \cdot \grad \right) \tU
    \nonumber \\
    & & -\grad P_t + f_t + \frac{1}{Re} \lapl f
\eea

This expression looks messy, but it turns out that the first two lines
can be partially cancelled out by careful choice of the spatial 
discretizations.  The last three lines are correction terms that need
to be added to the RHS of the forward Euler scheme.  A few of the terms 
are already computed (\eg $-\grad P + f$) and can be reused to reduce the
computational cost of evaluating the correction terms.  For a heuristic
way to see whether a term is can be cancelled out or must be explicitly 
included as a correction term, it is helpful to look at the analysis of 
the viscous Burgers equation in~\cite{chu_2009}.

To make it easier to derive the spatial discretizations, it is convenient
to expand the cancellation terms in component form.  The $x$-component 
of 
\bea
  \lapl \left( (\tU \cdot \grad) \tU \right)
  + \left(\tU \cdot \grad \right) \lapl \tU 
  + \left(\lapl \tU \cdot \grad \right) \tU 
\eea
is given by 
\bea
  \lapl \left( (\tU \cdot \grad) \tu \right)
  &+& \left(\tU \cdot \grad \right) \lapl \tu 
  \ + \ \left(\lapl \tU \cdot \grad \right) \tu =  
  \nonumber \\
  &2& \left[ 
  \begin{array}{c}
    \tu \lapl \tu_x + \tv \lapl \tu_y + \tw \lapl \tu_z \\
    + \ \grad \tu_x \cdot \grad \tu 
    + \grad \tu_y \cdot \grad \tv
    + \grad \tu_z \cdot \grad \tw \\
    + \ \tu_x \lapl \tu + \tu_y \lapl \tv + \tu_z \lapl \tw 
  \end{array}
  \right]
%  \nonumber \\
%  &2& \left[ 
%  \begin{array}{c}
%    \tu \lapl \tu_x + \tv \lapl \tu_y + \tw \lapl \tu_z \\
%    + \div \left( \tu_x \cdot \grad \tu \right)
%    + \div \left( \tu_y \cdot \grad \tv \right)
%    + \div \left( \tu_z \cdot \grad \tw \right)
%  \end{array}
%  \right]
  \label{eq:u_temporal_error}
\eea
Similarly, the $x$-component of the first pressure term in $\tU_{tt}$ is
\beq
  -\frac{1}{Re} 
  \left( P_{xxx} + P_{xyy} + P_{xzz} \right)
  \label{eq:P_temporal_error}
\eeq
The $y$- and $z$-components have analogous forms.

The leading-order term in the local truncation error for the forward
Euler scheme is
\beq
 e_{spatial} \dt + \frac{\dt^2}{2} \tU_{tt},
\eeq
so we seek finite difference discretizations of the spatial derivatives that
will exactly balance some of the terms in the temporal error.

For the pressure gradient term, use a second-order central-difference 
approximation.  In the $x$-direction:
\bea
  \frac{\partial P_{i+1/2,j,k}}{\px} &=& 
  \alpha \left(\frac{P_{i+1,j,k}-P_{i,j,k}}{\dx}\right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{P_{i+1,j+1,k}-P_{i,j+1,k}}{\dx} 
                           + \frac{P_{i+1,j-1,k}-P_{i,j-1,k}}{\dx} 
                      \right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{P_{i+1,j,k+1}-P_{i,j,k+1}}{\dx} 
                           + \frac{P_{i+1,j,k-1}-P_{i,j,k-1}}{\dx} 
                      \right),
\eea
with $\alpha + 2 \beta = 1$.
The discretization error for this scheme is 
\beq
  - \left(\alpha + 2 \beta \right ) \frac{\dx^2}{12} \tP_{xxx}
  - \beta \frac{\dx^2}{4} + \tP_{xyy}
  - \beta \frac{\dx^2}{4} + \tP_{xzz}
  + O(\dx^4),
\eeq
where we have use the fact that the grid spacing is the same in all
directions.  Therefore, to cancel out the contribution to the truncation
error from (\ref{eq:P_temporal_error}) when $\dt = \dt_{opt}$, we want to 
choose $\alpha = 1/3 = \beta$.  For a 2D problem, we would choose 
$\alpha = 2/3$ and $\beta = 1/3$.

For the nonlinear convection term, we can eliminate the first three terms in
(\ref{eq:u_temporal_error}) by computing gradients using an analogous scheme 
as for the pressure gradient.  Next, we can carefully choose the approximation 
of $\hat{u}_{i+1/2,j,k}$, $\hat{v}_{i+1/2,j,k}$, and $\hat{w}_{i+1/2,j,k}$ 
so that the fourth term is completely eliminated and the error when computing
$v u_y + w u_z$ are automatically eliminated by adding a multiple of the
$u_y \lapl v + u_z \lapl w$ terms in (\ref{eq:u_temporal_error}).
The remaining three terms need to be directly added as a correction term.  
The can be computed using things that have already been computed anyways (I 
think).  For the $x$-component of the fluid velocity, we use the following 
finite difference approximations for the gradient:
\bea
  \left. \frac{\partial \tu}{\px}\right|_{i+1/2,j,k} &=& 
  \alpha \left(\frac{u_{i+3/2,j,k}-u_{i-1/2,j,k}}{2\dx}\right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{u_{i+3/2,j+1,k}-u_{i-1/2,j+1,k}}{2\dx} 
                           + \frac{u_{i+3/2,j-1,k}-u_{i-1/2,j-1,k}}{2\dx} 
                      \right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{u_{i+3/2,j,k+1}-u_{i-1/2,j,k+1}}{2\dx} 
                           + \frac{u_{i+3/2,j,k-1}-u_{i-1/2,j,k-1}}{2\dx} 
                      \right).
\eea
\bea
  \left. \frac{\partial \tu}{\py}\right|_{i+1/2,j,k} &=& 
  \alpha \left(\frac{u_{i+1/2,j+1,k}-u_{i+1/2,j-1,k}}{2\dy}\right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{u_{i+3/2,j+1,k}-u_{i+3/2,j-1,k}}{2\dy} 
                           + \frac{u_{i-1/2,j+1,k}-u_{i-1/2,j-1,k}}{2\dy} 
                      \right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{u_{i+1/2,j+1,k+1}-u_{i+1/2,j-1,k+1}}{2\dy} 
                           + \frac{u_{i+1/2,j+1,k-1}-u_{i+1/2,j-1,k-1}}{2\dy} 
                      \right).
\eea
\bea
  \left. \frac{\partial \tu}{\pz}\right|_{i+1/2,j,k} &=& 
  \alpha \left(\frac{u_{i+1/2,j,k+1}-u_{i+1/2,j,k-1}}{2\dz}\right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{u_{i+3/2,j,k+1}-u_{i-1/2,j,k-1}}{2\dz} 
                           + \frac{u_{i+3/2,j,k+1}-u_{i-1/2,j,k-1}}{2\dz} 
                      \right)
  \nonumber \\
  &+& \frac{\beta}{2} \left( \frac{u_{i+1/2,j+1,k+1}-u_{i+1/2,j+1,k-1}}{2\dz} 
                           + \frac{u_{i+1/2,j-1,k-1}-u_{i+1/2,j-1,k-1}}{2\dz} 
                      \right),
\eea
with $\alpha = 1/3 = \beta$.
For a 2D problem, we use analogous formulae with 
$\alpha = 2/3, \beta = 1/3$.

In 3D, we use the following for
$\hat{u}_{i+1/2,j,k}$, $\hat{v}_{i+1/2,j,k}$, and $\hat{w}_{i+1/2,j,k}$, 
\bea
  \hat{u}_{i+1/2,j,k} = \frac{1}{6}
  \left[
      u_{i+3/2,j,k} + u_{i-1/2,j,k} 
    + u_{i+1/2,j+1,k} + u_{i+1/2,j-1,k} 
    + u_{i+1/2,j,k+1} + u_{i+1/2,j,k-1}
  \right]
\eea
\bea
  \hat{v}_{i+1/2,j,k} &=& 
  \frac{3}{16}
  \left[
    \begin{array}{c}
    \ \ v_{i+1,j+1/2,k} + v_{i,j+1/2,k} \\
    + v_{i+1,j-1/2,k} + v_{i,j-1/2,k}
    \end{array}
  \right] \nonumber \\
  &+& \frac{1}{32}
  \left[
    \begin{array}{c}
    \ \ v_{i+1,j+1/2,k+1} + v_{i,j+1/2,k+1} \\
    + v_{i+1,j-1/2,k+1} + v_{i,j-1/2,k+1} \\
    + v_{i+1,j+1/2,k-1} + v_{i,j+1/2,k-1} \\
    + v_{i+1,j-1/2,k-1} + v_{i,j-1/2,k-1}
    \end{array}
  \right]
\eea
\bea
  \hat{w}_{i+1/2,j,k} &=& 
  \frac{3}{16}
  \left[
    \begin{array}{c}
    \ \ w_{i+1,j,k+1/2} + w_{i,j,k-1/2} \\
    + w_{i+1,j,k+1/2} + w_{i,j,k-1/2}
    \end{array}
  \right] \nonumber \\
  &+& \frac{1}{32}
  \left[
    \begin{array}{c}
    \ \ w_{i+1,j+1,k+1/2} + w_{i,j+1,k+1/2} \\
    + w_{i+1,j+1,k-1/2} + w_{i,j+1,k-1/2} \\
    + w_{i+1,j-1,k+1/2} + w_{i,j-1,k+1/2} \\
    + w_{i+1,j-1,k-1/2} + w_{i,j-1,k-1/2} \\
    \end{array}
  \right]
\eea
In 2D, the analogous expressions are:
\bea
  \hat{u}_{i+1/2,j} = \frac{1}{3} u_{i+1/2,j}
  + \frac{1}{6}
  \left[
      u_{i+3/2,j} + u_{i-1/2,j} 
    + u_{i+1/2,j+1} + u_{i+1/2,j-1} 
  \right]
\eea
\bea
  \hat{v}_{i+1/2,j} &=& 
  \frac{1}{4}
  \left[
      v_{i+1,j+1/2} + v_{i,j+1/2} 
    + v_{i+1,j-1/2} + v_{i,j-1/2}
  \right] 
\eea

In the 3D case, the discretization error for $-(U \cdot \grad) u$ is
\bea
  \frac{\dx^2}{6}
  \left[\tu \lapl \tu_x + \tv \lapl \tu_y + \tw \lapl \tu_z \right]
  + \frac{\dx^2}{6} \tu_x \lapl \tu 
  + \frac{\dx^2}{8} \left[ \tu_y \lapl \tv + \tu_z \lapl \tw \right]
  \label{eq:convection_err_3d}
\eea
Therefore, when the optimal time step is used the local truncation error
associated with the convection terms is:
\bea
  -\frac{\dto \dx^2}{24} \left[ \tu_y \lapl \tv + \tu_z \lapl \tw \right]
  -\frac{\dto^2}{Re}
    \left[ \grad \tu_x \cdot \grad \tu
         + \grad \tu_y \cdot \grad \tv
         + \grad \tu_z \cdot \grad \tw
    \right] = 
    \nonumber \\
  -\frac{\dto^2}{Re} 
  \left[ \frac{1}{4} \left( \tu_y \lapl \tv + \tu_z \lapl \tw \right)
  +  \grad \tu_x \cdot \grad \tu
  + \grad \tu_y \cdot \grad \tv
  + \grad \tu_z \cdot \grad \tw
  \right]
\eea
This is the correction term that is required for the convection term's
contribution to the local truncation error.

In the 2D case, the discretization error for $-(U \cdot \grad) u$ is
\bea
  \frac{\dx^2}{6}
  \left[\tu \lapl \tu_x + \tv \lapl \tu_y \right]
  + \frac{\dx^2}{6} \tu_x \lapl \tu 
  + \frac{\dx^2}{8} \tu_y \lapl \tv 
\eea
and an analogous result is obtained for the convection term's contribution
to the correction term.

The discretizations of the $y$- and $z$- components of the Navier-Stokes
equations are analogous.

\subsubsection{Pressure discretization}
Use a compact fourth-order scheme for the pressure Poisson equation (or 
Poisson equation for the pressure correction).  To ensure that the RHS 
is fourth-order accurate, compute the $\div f$ term analytically or (use
a fourth-order discretization ... this will probably require wider stencils).
For the $\div \left( \left(U \cdot \grad \right) U \right)$ term, we can
use the same stencil as for the time advance and explicitly subtract
the discretization error (\ref{eq:convection_err_3d}) to get fourth-order
accuracy.

\section{Concerns/Thoughts}

\begin{itemize}
\item I have some concerns that while the the above discretization is
fourth-order accurate, if we only have a fourth-order accurate solution for 
the pressure, the accuracy of the gradient will only be third-order accurate
leading to an overall third-order accurate scheme.  It may work out that this
is not the case, but it's something to look out for.  A more careful 
analysis of the pressure scheme may be necessary to determine if this will 
affect the overall scheme.

\item It would be nice to think of an interpretation of the fancy spatial
stencils as cancelling out artificial dissipation or something like that.

\item If it's acceptable to have pressure oscillations (because we only
case about the fluid velocity), then using an unstaggered grid may be nice
because we can use the same simple formula for 
$\hat{u}$, $\hat{v}$, and $\hat{w}$ which leads to a discretization error
for the convection term that has errors that are the same in all directions:
\beq
  \frac{\dx^2}{6}
  \left[\tu \lapl \tu_x + \tv \lapl \tu_y + \tw \lapl \tu_z \right]
  + \frac{\dx^2}{6} 
  \left[ \tu_x \lapl \tu + \tu_y \lapl \tv + \tu_z \lapl \tw \right]
\eeq
so that six of the terms in (\ref{eq:u_temporal_error}) can be eliminated
by using the optimal time step approach.  The convection contribution to 
the correction term in this case is:
\beq
  -\frac{\dto^2}{Re} 
  \left[ \grad \tu_x \cdot \grad \tu
       + \grad \tu_y \cdot \grad \tv
       + \grad \tu_z \cdot \grad \tw
  \right]
\eeq
The formulas for the pressure gradient discretization would also have be 
adjusted a little bit, but it would end up being basically the same as the
gradient expression for the convection term.

\end{itemize}

\section{\label{sec:summary} Summary}

\section*{Acknowledgments}
The author gratefully acknowledges the support of Vitamin D, Inc.
and the Institute for High-Performance Computing (IHPC) in Singapore. 
The author would like to thank ??  for helpful suggestions on the manuscript.

\bibliography{OTS-NavierStokes}

\end{document}

### 2010-01-18 papers/OTS-NavierStokes/genPDFFigures.sh

@@ -1,0 +1,17 @@
+#!/bin/bash
+#
+# This script generates PDF figures from the eps versions of figures.
+#
+# Kevin T. Chu
+#
+
+FIGURES_DIR="figures"
+
+# convert figures
+cd ${FIGURES_DIR}
+for FIG in `ls *.eps`; do
+  echo -n "Converting ${FIG} ..."
+  epstopdf ${FIG}
+  echo "done"
+done
+

### 2010-01-18 papers/OTSPDE-VarCoefs/getFigures.sh
Added getFigures.sh and genPDFFigures.sh to OTSPDE-VarCoefs paper.

@@ -1,0 +1,33 @@
+#!/bin/bash
+#
+# This script collects the figures from the src directory and places them
+# in the figures directory.
+#
+# Kevin T. Chu
+# February 2008
+#
+
+FIGURES_DIR="figures"
+
+SRC_DIR="../../src"
+EXAMPLE_DIR_LIST="scalar_PDEs/1d/wave_eqn                     \
+                  variable_coefs/wave_eqn"
+
+# create directory for figures if it doesn't already exist
+if [ ! -d "${FIGURES_DIR}" ]; then
+  mkdir -p ${FIGURES_DIR}
+fi
+
+# move figures 
+for DIR in ${EXAMPLE_DIR_LIST}; do
+  FIGURES=`ls ${SRC_DIR}/${DIR}/figures/*.eps 2> /dev/null` 
+  if [ ! -z "${FIGURES}" ]; then
+    echo -n "Moving figures from "
+    echo "${SRC_DIR}/${DIR}/figures/ to ${FIGURES_DIR} ..."
+    mv ${SRC_DIR}/${DIR}/figures/*.eps ${FIGURES_DIR}/
+  else
+    echo -n "Figures missing from "
+    echo "${SRC_DIR}/${DIR}/figures/"
+  fi
+done
+

### 2010-05-02 papers/OTSPDE/OTSPDE.tex
Revised OTSPDE paper.

@@
 % Resubmission draft: 2009/05/25, Serendipity Research
+% Resubmission draft: 2010/03/25, Serendipity Research
+% Resubmission draft: 2010/05/02, Serendipity Research
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\title{Boosting the Accuracy of Finite Difference Schemes via
-       Optimal Time Step Selection and Non-Iterative Defect Correction}
-
-\author{Kevin T. Chu\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore}$^,$\address{Vitamin D, Inc., Menlo Park, CA 94025, United States}
+\title{Using Optimal Time Step Selection to Boost the Accuracy of
+       Finite Difference Schemes for Time Dependent PDEs}
+
+\author{Kevin T. Chu\address{Serendipity Research, Mountain View, CA 94041, United States}
 }
 
-\runtitle{Optimal Time Step Selection and Non-Iterative Defect Correction for 
-          FD Schemes}
+\runtitle{OTS Selection for FD Approximations of Time Dependent PDEs}
 \runauthor{K.T. Chu}
@@
 partial differential equations by optimally selecting the time step used
-to advance the numerical solution and adding defect correction terms in a
-non-iterative manner.  The power of the technique is its ability to extract 
-as much accuracy as possible from existing finite difference schemes with 
-minimal additional effort.  Through straightforward numerical analysis 
-arguments, we explain the origin of the boost in accuracy and estimate the 
-computational cost of the resulting numerical method.  We demonstrate the 
-utility of optimal time step (OTS) selection combined with non-iterative 
-defect correction (NIDC) on several different types of finite difference
-schemes for a wide array of classical linear and semilinear PDEs in one and 
-more space dimensions on both regular and irregular domains.  
+to advance the numerical solution. The power of the technique is its ability
+to increase the accuracy of existing finite difference schemes with 
+minimal modification to the original numerical algorithm.  Through
+straightforward numerical analysis arguments, we explain the origin of the
+boost in accuracy and estimate the computational cost of the resulting
+numerical method.  We demonstrate the utility of optimal time step (OTS)
+selection on several different types of finite difference schemes
+for a wide array of classical linear and semilinear PDEs in one and more space
+dimensions on both regular and irregular domains.  
 \end{abstract}
@@
 schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
-powerful way to boost the accuracy of a numerical method \emph{without} 
+powerful way to boost the accuracy of a numerical method without
 introducing too much additional algorithmic (and programming) complexity.
 
-Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
-but relatively uncommon, approach for maximizing the accuracy of numerical 
-schemes.  Often, all combinations of numerical parameters are considered 
-equally good.  For some problems, however, optimizing the numerical parameters 
-can significantly increase the accuracy of the numerical solution.  For 
-example, the well-known unit CFL condition, $\dt = \dx/a$, for the first-order 
-upwind forward Euler discretization of the linear advection equation 
-completely eliminates the numerical error introduced by the 
-scheme~\cite{leveque_book_2002}.  Another well-known example is the boost in 
-the order of accuracy that results from setting the time step to 
+One important, but relatively uncommon, approach for increasing the accuracy
+of a numerical algorithm is optimization of its parameters
+(\eg $\dx$ and $\dt$).  Often, all combinations of numerical parameters are
+considered equally good.  For some problems, however, optimizing the numerical
+parameters can dramatically increase the accuracy of the numerical solution.
+For example, the well-known unit CFL condition, $\dt = \dx/a$, for the
+first-order upwind forward Euler discretization of the linear advection
+equation completely eliminates the numerical error introduced by the 
+scheme~\cite{leveque_book_2002}.  Another relatively well-known example is the
+boost in the order of accuracy that results from setting the time step to 
 $\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
 time integration with the standard second-order central difference 
-approximation for the Laplacian.  As a final example, Zhao showed that an 
-appropriate choice of the weight factor $\theta$ in the theta-method can 
-significantly decrease the error in finite element solutions to the diffusion 
-equation~\cite{zhao_2006}.
-
-In this article, we present a simple technique for boosting the order of 
+approximation for the Laplacian.  As a final example, Zhao recently showed
+that an appropriate choice of the weight factor $\theta$ in the theta-method
+can significantly decrease the error in finite element solutions to the
+diffusion equation~\cite{zhao_2006}.
+
+In this article, we present a simple technique for boosting the \emph{order} of 
 accuracy of finite difference schemes for time dependent PDEs by optimally 
-choosing the time step and, when necessary, adding defect correction terms
-in a \emph{non-iterative} manner.  This technique is a systematic 
-generalization of the ideas underlying some of the examples mentioned above.
-Optimal time step (OTS) selection is based on the observation that a carefully 
-chosen time step can simultaneously eliminate low-order terms in both the
-spatial and temporal the discretization errors.  It combines the well-known 
-approach of using knowledge of the structure of the PDE to reduce numerical 
-error with the less-common notion that an optimal choice of numerical 
-parameters can boost the effectiveness of algorithmically simple and 
-computationally low-cost numerical schemes.  For some problems, OTS alone is 
-sufficient to allow a formally low-order method to deliver high-order 
-accuracy.  However, addition of defect correction terms is generally required 
-to eliminate residual terms in the leading-order truncation error.  It is 
-important to emphasize that, unlike traditional defect/deferred correction 
-methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006} 
-which use defect correction terms to iteratively refine the solution at 
-each time step, we take a non-iterative defect correction (NIDC) approach
-that incorporates defect correction terms directly into the original finite 
-difference scheme.
-
-As a technique for enhancing numerical methods, OTS-NIDC has several desirable 
-features.  Like any technique that leads to high-order methods, OTS-NIDC
-produces schemes that greatly reduce the computational cost required to obtain 
-a numerical solution of a desired accuracy.  However, its real power comes 
-from the fact that it achieves high-order accuracy while being based solely on 
-very simple, formally low-order finite difference schemes.  In other words, 
-OTS-NIDC allows us to extract as much accuracy as possible from an existing 
-numerical scheme with minimal additional work.  For instance, we will see how 
-OTS-NIDC makes it possible to achieve 4th-order accuracy for the diffusion 
-equation in any number of space dimensions using only a second-order stencil 
-for the Laplacian and explicit time integration.  
-Moreover, high-order accuracy is not restricted to problems 
-on simple, rectangular domains; irregular domains are straightforward to 
-handle by appropriately setting the values in ghost 
-cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
-OTS-NIDC is even beneficial when using finite difference schemes to solve some 
-nonlinear PDEs.  While there are certainly limitations to OTS-NIDC, we will 
-see that there are many problems where it is useful.
+choosing the time step.  This technique is a systematic generalization of
+the ideas underlying some of the examples mentioned above.  Optimal time step
+(OTS) selection is based on the observation that a carefully chosen time step
+can simultaneously eliminate low-order terms in both the spatial and temporal
+the discretization errors.  At its essence, OTS selection combines the 
+powerful but uncommon approach of optimally choosing numerical parameters 
+with the well-known approach of using knowledge of the structure of the
+PDE to reduce numerical error.  The end result is a strategy for boosting
+the order of accuracy of algorithmically simple and computationally low-cost
+numerical schemes.
+
+While OTS selection alone is sufficient to allow a formally low-order method to
+deliver high-order accuracy for some problems, defect correction terms are
+generally required to eliminate residual terms in the leading-order truncation
+error.  However, unlike traditional defect correction 
+methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}, 
+using OTS selection eliminates the need to incorporate correction terms that
+involve derivatives of higher-order than those that appear in the original PDE.
+Another significant feature of OTS selection with defect correction (OTS-DC)
+is that the defect terms are incorporated in a \emph{non-iterative} fashion.
+In contrast, traditional defect/deferred correction approaches typically
+require several refinement iterations at each time step to boost the order of
+accuracy of the numerical solution.  OTS-DC also differs from the method of
+modified equations~\cite{griffiths_1986} for eliminating higher-order
+corrections to the truncation error because it uses an optimal time
+step, not explicit inclusion of correction terms, to regularize the
+singularly perturbed modified equations - the correction terms that remain
+when using OTS-DC are all regular perturbations to the original equation.
+
+As a technique for enhancing numerical methods, OTS selection has several
+desirable features.  Like any technique that leads to high-order methods,
+OTS selection produces schemes that greatly reduce the computational cost
+required to obtain a numerical solution of a desired accuracy.  However, its
+real power comes from the fact that it achieves high-order accuracy while
+being based solely on very simple, formally low-order finite difference
+schemes \emph{without} requiring special algorithmic steps.  In other words,
+OTS selection allows us to extract as much accuracy as possible from an
+existing numerical scheme with minimal additional work.  For instance, we will
+see how OTS selection makes it possible to achieve 4th-order accuracy for the
+diffusion equation in any number of space dimensions using only a second-order
+stencil for the Laplacian and explicit time integration.  Moreover, high-order
+accuracy is not restricted to problems on simple, rectangular domains;
+irregular domains are straightforward to handle by appropriately setting the
+values in ghost cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
+OTS selection is even beneficial when using finite difference schemes to solve
+some nonlinear PDEs.
 
 This article is organized as follows.  In the remainder of this section,
-we compare OTS-NIDC to a few related numerical techniques.
-In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
-OTS-NIDC and show how it boosts the order of accuracy for formally low-order 
-numerical schemes.  
+we compare OTS selection to a few related numerical techniques.
+In Section~\ref{sec:OTS}, we present the theoretical foundations of 
+OTS selection and show how it boosts the order of accuracy for formally
+low-order numerical schemes.  
 In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
-OTS-NIDC to problems in one space dimension by applying it to finite 
+OTS selection to problems in one space dimension by applying it to finite 
 difference schemes for several classical linear and nonlinear PDEs.  
 In Section~\ref{sec:applications_multidim}, we show that 
-OTS-NIDC is useful for problems in multiple space dimensions by applying it
-the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
-the ease with which high-order accuracy can be achieved for problems on 
-arbitrary domains, the 2D diffusion equation is solved on both regular 
-and irregular domains.  
-For each PDE in Sections~\ref{sec:applications_1d} and 
-\ref{sec:applications_multidim}, we provide the numerical analysis required 
-to calculate the optimal time step size, the defect correction terms, and the 
-order of accuracy.  Finally, we offer some concluding remarks in 
-Section~\ref{sec:summary}.
-
-\subsection{Comparison with Traditional Defect/Deferred Correction Methods}
+OTS selection is useful for problems in multiple space dimensions by applying
+it the 2D linear advection equation and the 2D diffusion equation.  To
+illustrate the ease with which high-order accuracy can be achieved for
+problems on arbitrary domains, the 2D diffusion equation is solved on both
+regular and irregular domains.  For each PDE in
+Sections~\ref{sec:applications_1d} and \ref{sec:applications_multidim}, we
+provide the numerical analysis required to calculate the optimal time step
+size, the defect correction terms, and the order of accuracy.  Note that in
+order to emphasize the method, we have chosen to apply it to relatively simple
+problems rather than use it to tackle ``real-life'' applications.  The method,
+however, is not restricted to these simple examples.  Finally, in
+Section~\ref{sec:summary}, we summarize the paper with some concluding
+remarks.
+
+\subsection{Comparison with Defect/Deferred Correction Methods}
 Defect/deferred correction is a well-known strategy for boosting the order of 
@@
 takes one of two forms: 
-(1) iterative refinement of the numerical solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
+(1) iterative refinement of the numerical
+solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
 or 
@@
 calculated from lower accuracy iterates.  When used to improve the 
-temporal order of accuracy, the iterative approach can help to ensure 
+temporal order of accuracy, the iterative approach helps to ensure 
 stability of the time integration scheme~\cite{kress_2006}.  Although there 
@@
 improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
-Note that unlike iterated defect correction, OTS-NIDC requires no iteration 
-during each time step because it computes correction terms using only the 
-solution at the previous time step(s) (as opposed to requiring an estimate 
-of the solution at the current time).
+Unlike iterated defect correction, OTS-DC requires no iteration during each
+time step because it computes correction terms using only the solution at the
+previous time step(s) (as opposed to requiring an estimate of the solution at
+the current time).
 
@@
 
-OTS-NIDC has several advantages over both of these defect correction 
-techniques.  First, OTS-NIDC is generally more computationally efficient than 
+OTS-DC has several advantages over both of these defect correction 
+techniques.  First, OTS-DC is generally more computationally efficient than 
 both iterated defect correction and the high-order compact scheme approach
 because it requires less memory and, for many problems, admits simple 
-explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
-spatial and temporal numerical errors at the same time.  Finally, OTS-NIDC 
-does not rely solely on the addition of finite differences of solution values 
-to cancel out discretization errors.  Instead, it takes advantage of 
-relationships between spatial derivatives, temporal derivatives and source 
-terms to eliminate the need for finite differences approximations of 
-high-order derivatives of the solution.  Compared with iterated defect 
-correction, OTS-NIDC makes it easier to avoid instabilities that arise from 
-the presence of finite difference discretizations of high-order temporal 
-derivatives and may eliminate some of the correction terms.  Comparing to the
-high-order compact scheme approach, the optimization of numerical parameters 
-in OTS-NIDC means that mixed partial derivatives do not need to appear in the 
-fully discrete equations.
+explicit time integration schemes.  Second, OTS-DC inherently controls both 
+spatial and temporal numerical errors at the same time.  Finally, taking
+advantage of relationships between spatial derivatives, temporal derivatives
+and source terms, OTS-DC avoids the need to explicitly incorporate finite
+differences of high-order derivatives of intermediate solutions in order to
+eliminate discretization errors. 
+Compared with iterated defect correction, OTS-DC makes it easier to avoid
+instabilities that arise from the presence of finite difference discretizations
+of high-order temporal derivatives by eliminating the most problematic
+correction terms.  Comparing to the high-order compact scheme approach, the
+optimization of numerical parameters in OTS-DC means that mixed partial
+derivatives do not need to appear in the fully discrete equations.
+
+\subsection{Comparison with the Modified Equation Method}
+Both defect and deferred correction can be viewed as special cases of the
+method of modified equations~\cite{griffiths_1986}.  When using any modified
+equation method, the basic approach is to derive a modified version of the 
+original PDE through the introduction of the leading-order terms of the
+discretization error into the PDE.  Unfortunately, the modified PDE is a
+singular perturbation to the original PDE due to the presence of the
+discretization parameters in the correction terms.  Because the the singular
+perturbation can lead to numerical instabilities, finding a way regularize
+the modified equation is an important step for developing any numerical method
+based on the modified equations method.
+
+Whereas deferred correction methods deal with the singular perturbation by
+using an iterative procedure to avoid numerical instabilities, the traditional
+method of modified equations~\cite{griffiths_1986} handles the
+singularity by ``pushing'' the singularity to a higher-order term in the
+discretization error.  The basic procedure involves substituting derivatives
+that appear in the singular term with expressions obtained by taking
+derivatives of the original PDE.  This procedure yields a new modified
+equation where the new singular term is of sufficiently high-order in the
+discretization parameter that it can be neglected in the construction of the
+numerical scheme.
+
+Like defect/deferred correction, OTS selection (even without defect correction)
+falls within the general framework of the method of modified equations.
+In OTS selection, regularization of the singular perturbation in the modified
+equations is achieved by choosing the time step so that the spatial and
+temporal singular perturbation terms exactly cancel each other.  More
+specifically, by constraining the relationship between the time step and
+grid spacing, the modified equation becomes a regular perturbation to the
+original PDE.  In the special case when OTS selection does not require
+defect correction, the modified equation is actually exactly the same as the
+original equation -- no regular perturbation terms remain after the singular
+perturbation terms have been eliminated.
+
+There are several advantage of OTS selection over traditional modified
+equation methods.  First, because no finite difference approximations of
+higher-order derivatives are required, numerical schemes based on OTS
+selection do not require additional boundary conditions.  Second, derivation
+of the optimal time step and correction terms for OTS-DC requires a comparable
+amount of analysis as the traditional modified equation method.  In all cases,
+we need explicit knowledge of the leading-order terms in the discretization
+error.  Finally, because the leading-order terms in the discretization error
+are eliminated implicitly rather than explicitly, the computational
+performance of OTS-DC is typically superior to the performance of traditional
+modified equation methods.
 
@@
 
-\section{\label{sec:OTS-NIDC}
-         Theoretical Foundations of OTS Selection and NIDC}
-OTS-NIDC is not by itself a method for constructing finite difference 
+\section{\label{sec:OTS} Theoretical Foundations of OTS Selection}
+OTS selection is not by itself a method for constructing finite difference 
 schemes.  Rather, it is a way to enhance to the performance of existing 
-finite difference schemes by carefully choosing the time step and adding
-defect correction terms to eliminate low-order numerical errors.  
-The two key ideas underlying OTS-NIDC are 
+finite difference schemes by carefully choosing the time step to eliminate
+leading-order terms in the discretization error.  The two key ideas
+underlying OTS selection are 
 (1) a judicious choice of time step can be used to eliminate the leading-order 
@@
 in the discretization error. 
-In this section, we present and analyze the technique of OTS-NIDC for finite 
-difference schemes constructed to solve scalar evolution equations of the form 
+In this section, we present and analyze the technique of OTS selection for
+finite difference schemes constructed to solve scalar evolution equations of
+the form 
 \beq
@@
 \subsection{\label{sec:ots_model_pde} 
-            Optimal Time Step for a Model PDE}
+            Optimal Time Step Selection for a Model PDE}
 Let us begin by considering a finite difference approximation for the
@@
 Notice that the optimal time step has a similar functional dependence on the 
-grid spacing as a time step that satisfies a stability constraint.
+grid spacing as a time step constraint required for stability of an explicit
+time-stepping scheme.
 
@@
 As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
-scheme that is computationally cheaper than traditional methods that use 
-the same low-order finite difference stencils.  The performance gain for 
+scheme that is computationally cheaper than traditional implicit schemes that
+use the same low-order finite difference stencils.  The performance gain for 
 problems in higher spatial dimensions is even more impressive.  As we can see 
@@
                    & $O\left( e^{-1} \right)$ \\
-  Forward Euler with OTS-NIDC  & $O\left( e^{1/4} \right)$ 
+  Forward Euler with OTS-DC & $O\left( e^{1/4} \right)$ 
                    & $O\left( e^{1/2} \right)$ 
@@
 \caption{
-Computational cost of the forward Euler scheme with OTS-NIDC and
+Computational cost of the forward Euler scheme with OTS-DC and
 the Crank-Nicholson method for solving the diffusion equation as a function 
@@
   \hline
-  & \multicolumn{2}{c}{\bf Forward Euler with OTS-NIDC} 
+  & \multicolumn{2}{c}{\bf Forward Euler with OTS-DC} 
   & \multicolumn{2}{c}{\bf Crank-Nicholson} \\
@@
 
-\subsection{\label{sec:ots_nidc_general_1d_pdes} 
-            OTS-NIDC for General PDEs in One Space Dimension} 
+\subsection{\label{sec:ots_dc_general_1d_pdes} 
+            OTS Selection for General PDEs in One Space Dimension} 
 In the previous section, we analyzed optimal time step selection in the 
@@
 the same improvement in the order of accuracy.  In this section, we discuss 
-these correction terms in the context of general constant coefficient, linear 
-PDEs and semilinear PDEs.
+OTS selection with defect correction (OTS-DC) in the context of general
+constant coefficient, linear PDEs and semilinear PDEs.
 
@@
   \dt\left[ (L u) (\alpha \dx^r - \beta \dt^s) 
-  + \sum_k (L_k u) \dx^r 
-  + \sum_k (L'_k u) \dt^s 
+  + (L' u) \dx^r 
+  + (L'' u) \dt^s 
   + (G f) \dt^s \right]
@@
 (\ref{eq:leading_order_error_model_PDE_MOL}), 
-$L_k$ and $L'_k$ are the linear spatial differential operators associated 
+$L'$ and $L''$ are the linear spatial differential operators associated 
 with the leading-order terms in the discretization error that are not involved 
@@
 differential operator that acts on the source term $f$.  
-Note that neither $L_k$ nor $L'_k$ involve any temporal differential 
+Note that neither $L'$ nor $L''$ involve any temporal differential 
 operators because any that arise can be eliminated using the 
@@
 defect correction terms to cancel out the second, third, and fourth terms in 
-(\ref{eq:leading_order_error_linear_PDE_MOL}).  At each time step, the second 
-and third terms in (\ref{eq:leading_order_error_linear_PDE_MOL}) can be 
-approximated using finite differences applied to the solution $u$ at the 
-current time.
+(\ref{eq:leading_order_error_linear_PDE_MOL}).  To advance the solution to
+the next time step, the second and third terms in 
+(\ref{eq:leading_order_error_linear_PDE_MOL}) can be approximated using finite
+differences applied to the solution $u$ at the current time.
 The last term can be computed using either a finite difference approximation 
@@
 \subsubsection*{Semilinear PDEs}
-OTS-NIDC can also be applied to semilinear PDEs of the form
+OTS-DC can also be applied to semilinear PDEs of the form
 \beq
@@
 
-\subsection{\label{sec:ots_nidc_higher_spatial_dims}
-            OTS-NIDC Selection in Multiple Space Dimensions}
-OTS-NIDC is applicable to PDEs in any number of space dimensions.  While 
+\subsection{\label{sec:ots_dc_higher_spatial_dims}
+            OTS-DC Selection in Multiple Space Dimensions}
+OTS-DC is applicable to PDEs in any number of space dimensions.  While 
 problems in higher dimensions generally require a more careful choice of the 
 finite difference stencil used to approximate spatial derivatives, there are 
-no fundamental barriers that prevent the use of OTS-NIDC.  Moreover, by using 
+no fundamental barriers that prevent the use of OTS-DC.  Moreover, by using 
 a `ghost cell' or `ghost value' method for imposing boundary 
@@
 
-\subsection{Limitations of Optimal Time Step Selection}
-As with any numerical method, optimal time step selection has its limitations.
-The most serious limitation is the requirement that the coefficient on the 
-leading-order spatial derivative in the PDE be constant in time and space.  
-Without this assumption, the first term in the leading-order discretization 
-error would be of the form $(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
-where $\alpha$ and $\beta$ are now functions of space and time.  This
-small change in the form of the error makes it impossible to derive a single 
-optimal time step.  Instead, each grid point would possess its own individual 
-optimal time step.  For a similar reason, optimal time step selection also 
-fails for quasilinear and fully nonlinear PDEs.  The issue for these types of 
-PDEs is that the coefficients in $P(\dx, \dt)$ become functions of the 
-solution $u$.
-
-Another limitation of optimal time step selection is that it does not,
-in its current incarnation, apply to finite difference schemes for general
-systems of PDEs.  The main issue here is that each PDE might have its 
-own optimal time step.  OTS is, however, certainly applicable for problems 
-where the optimal time step is the same for all of the individual PDEs in the 
-system.
-
-While these limitations certainly place restrictions on the range of problems 
-that we can apply OTS-NIDC to, there are many relevant and important PDEs 
-arising in practice that satisfy the requirements for OTS-NIDC.
-
 \section{\label{sec:applications_1d} 
          Application to PDEs in One Space Dimension}
-Because there are no geometry or anisotropy considerations, OTS-NIDC is very 
+Because there are no geometry or anisotropy considerations, OTS-DC is very 
 useful for boosting the accuracy of many finite difference schemes in one 
@@
 optimization as part of the design of any finite difference scheme, we 
-include applications of OTS-NIDC to several different types of finite 
+include applications of OTS-DC to several different types of finite 
 difference schemes.  The examples in this section were selected because they 
 have analytical solutions that are useful to compare numerical solutions 
-against.  However, the utility of OTS-NIDC is by no means limited to 
-these simple PDEs.
+against.  However, the utility of OTS-DC is by no means limited to these
+simple PDEs.
 
@@
 demonstrates that the expected accuracy is indeed achieved by applying 
-OTS-NIDC to the KPY scheme.  
+OTS-DC to the KPY scheme.  
 \begin{figure}[tb]
@@
 \caption{$L^\infty$ error as a function of number of grid points for the
-  KPY discretization of the second-order wave equation on with OTS-NIDC 
-  (blue circles) and without OTS-NIDC (red squares).
+  KPY discretization of the second-order wave equation on with OTS-DC 
+  (blue circles) and without OTS-DC (red squares).
 }
@@
 unstable for problems in more than 3 space dimensions.  Fortunately, the 
-OTS-NIDC makes it possible to construct a fourth-order accurate 
+OTS-DC makes it possible to construct a fourth-order accurate 
 DuFort-Frankel scheme.
@@
 correction term is $2 D \dt^3 \left(D f_{xx} + f_{t} \right)/\dx^2$.  As with 
-the forward Euler scheme for the diffusion equation, OTS-NIDC boosts the
+the forward Euler scheme for the diffusion equation, OTS-DC boosts the
 order of accuracy for the DuFort-Frankel scheme from two to four.  
-Figure~\ref{fig:diffusion_eqn_1d_dufort_frankel_error} shows that OTS-NIDC
+Figure~\ref{fig:diffusion_eqn_1d_dufort_frankel_error} shows that OTS-DC
 applied to the DuFort-Frankel scheme achieves the expected boost in accuracy.
@@
 equation as a function of number of grid points for the DuFort-Frankel scheme 
-with (circles) and without (squares) OTS-NIDC.
+with (circles) and without (squares) OTS-DC.
 }
@@
 \subsection{Viscous Burgers Equation}
-We now demonstrate the use of OTS-NIDC in the context of a well-studied
+We now demonstrate the use of OTS-DC in the context of a well-studied
 semilinear PDE -- the viscous Burgers equation~\cite{whitham_book}:
@@
 two finite difference schemes that solve the viscous Burgers equation:
-forward Euler with OTS-NIDC (circles)
+forward Euler with OTS-DC (circles)
 and forward Euler with suboptimal time step $\dt = \dx^2/4\nu$ and no 
@@
 Note that the optimal time step eliminates errors introduced by 
-both the diffusion term \emph{and} the nonlinear advection term.
+both the diffusion term \emph{and} the nonlinear advection term so that the
+defect correction terms only involve spatial derivatives of order two and
+lower.
 
@@
 Figure~\ref{fig:burgers_1d_error} demonstrates the boost in accuracy obtained 
-by applying OTS-NIDC to the finite difference 
+by applying OTS-DC to the finite difference 
 scheme~(\ref{eq:burgers_1d_FD_scheme}).
 
-The Burgers equation example illustrates an important aspect of OTS-NIDC:
-the ease with which OTS-NIDC can be used to boost the order of accuracy 
+The Burgers equation example illustrates an important aspect of OTS-DC:
+the ease with which OTS-DC can be used to boost the order of accuracy 
 depends on the choice of the finite difference scheme.  Had we chosen an 
@@
 higher-order spatial derivatives than those that appear in the original 
-equation.  While it should be possible to handle this situation with OTS-NIDC 
+equation.  While it should be possible to handle this situation with OTS-DC 
 selection, it is clearly less desirable.
@@
 where $\kappa > 0$.  This example illustrates the importance of the choice of
-discretization for the spatial derivative in order for OTS-NIDC to be
+discretization for the spatial derivative in order for OTS-DC to be
 applicable.
@@
 (\ref{eq:4th_order_parabolic_eqn_1d}):
-forward Euler with OTS-NIDC (circles),
+forward Euler with OTS-DC (circles),
 forward Euler with suboptimal time step and no correction term (squares),
@@
     & \multirow{2}{*}{$O\left( e^{-5/6} \right)$} \\ 
-  \ \ 4th-Order Bilaplacian and OTS-NIDC & & & & \\
+  \ \ 4th-Order Bilaplacian and OTS-DC & & & & \\
   Crank-Nicholson with 
@@
 operator with a fourth-order discretization.  For the forward Euler scheme,
-using an OTS-NIDC drives the computational cost down even more.
+using an OTS-DC drives the computational cost down even more.
 The performance plot in Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf} 
@@
 In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
-2nd-order Crank-Nicholson scheme actually outperforms the OTS-NIDC scheme over 
+2nd-order Crank-Nicholson scheme actually outperforms the OTS-DC scheme over 
 the high (but still practically useful) range of errors even though 
-Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-NIDC
+Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-DC
 scheme should do better.  The origin of this discrepancy is the constant 
@@
 Comparing the computational cost of the best forward Euler and Crank-Nicholson 
-schemes, we see that the OTS-NIDC forward Euler scheme more efficiently 
-utilizes memory at the cost of requiring more computation time.  This tradeoff 
+schemes, we see that the OTS-DC forward Euler scheme more efficiently 
+utilizes memory at the cost of requiring more computation time.  This trade off 
 between memory and computation only exists in one space dimension.  In 
@@
          Application to PDEs in Multiple Space Dimensions}
-OTS-NIDC can be used to boost the order of accuracy for finite difference 
+OTS-DC can be used to boost the order of accuracy for finite difference 
 schemes in any number of space dimensions.  However, several important issues 
@@
 multiple grid spacing parameters,  and the domain geometry.  In this section, 
-we show how OTS-NIDC selection easily overcomes these issues by using it to 
+we show how OTS-DC selection easily overcomes these issues by using it to 
 construct high-order finite difference schemes for the 2D advection equation 
@@
 incorporating ideas discussed in 
-Sections~\ref{sec:OTS-NIDC} and~\ref{sec:applications_1d}.
+Sections~\ref{sec:OTS} and~\ref{sec:applications_1d}.
 
@@
 However, precise knowledge of the cross-terms is critical when attempting to 
-use OTS-NIDC.  Without this knowledge, it is impossible to 
+use OTS-DC.  Without this knowledge, it is impossible to 
 ensure cancellation of leading-order errors.
 
-Recall that OTS-NIDC requires the leading-order temporal errors to be directly 
+Recall that OTS-DC requires the leading-order temporal errors to be directly 
 related to the leading-order spatial errors via the PDE.  This requirement can 
@@
 \subsection{Linear Advection Equation}
-For our first application of OTS-NIDC in multiple space dimensions, we 
+For our first application of OTS-DC in multiple space dimensions, we 
 consider the two-dimensional advection equation
@@
 equation~(\ref{eq:advection_eqn_2d}) using forward Euler time stepping 
-with OTS-NIDC (left) and forward Euler time 
+with OTS-DC (left) and forward Euler time 
 stepping with suboptimal time step (right).  The latter scheme uses a 
@@
 As for PDEs in one space dimension, achieving infinite order accuracy 
-through OTS-NIDC is unique to the advection equation.
-To illustrate the typical accuracy boost gained by using OTS-NIDC,
+through OTS-DC is unique to the advection equation.
+To illustrate the typical accuracy boost gained by using OTS-DC,
 we apply it to the 2D diffusion equation.  This example underscores 
@@
 for problems in more than one space dimension and demonstrates the ease with 
-which OTS-NIDC selection can be applied to problems on irregular domains.
+which OTS-DC selection can be applied to problems on irregular domains.
 
@@
 
-While it is certainly possible to directly apply OTS-NIDC
+While it is certainly possible to directly apply OTS-DC
 to (\ref{eq:diffusion_eqn_2d_5pt_scheme}), we would be forced to use additional
@@
 It is interesting to note that use of a nine-point stencil for the Laplacian 
-naturally arises when directly applying OTS-NIDC to the five-point scheme if 
+naturally arises when directly applying OTS-DC to the five-point scheme if 
 the defect correction term is interpreted as a modification to the original 
@@
 
-To apply OTS-NIDC to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we proceed in the 
+To apply OTS-DC to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we proceed in the 
 usual way by first analyzing the leading-order terms of the local truncation
@@
 2D diffusion equation:
-forward Euler with OTS-NIDC (circles), forward 
+forward Euler with OTS-DC (circles), forward 
 Euler with suboptimal time step (squares), and Crank-Nicholson (diamonds).  
@@
 \end{figure}
-The improvement in the accuracy of the forward Euler scheme with OTS-NIDC
+The improvement in the accuracy of the forward Euler scheme with OTS-DC
 is shown in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
@@
 theoretical estimates given in Table~\ref{tab:comp_perf_vs_dim}.  As with the
-other examples, the forward Euler scheme with OTS-NIDC is much 
+other examples, the forward Euler scheme with OTS-DC is much 
 more computationally efficient than the two other schemes.
@@
 are harder to design and implement.  
-In this section, we extend the OTS-NIDC finite difference scheme for the 2D 
+In this section, we extend the OTS-DC finite difference scheme for the 2D 
 diffusion equation on regular domains to irregular domains.  Other 
@@
 diffusion equation on a starfish-shaped domain.  The solution is computed 
-using forward Euler time stepping with OTS-NIDC.  In the error plot, the dark 
+using forward Euler time stepping with OTS-DC.  In the error plot, the dark 
 regions represent points where the error in the solution is larger than 
@@
 straightforward to compute high-order accurate solutions of the 2D diffusion 
-equation on irregular domains.  We simply use the OTS-NIDC forward Euler 
+equation on irregular domains.  We simply use the OTS-DC forward Euler 
 scheme designed for regular domains for grid points in the interior of the 
@@
 error in the solution is concentrated near the boundaries of the irregular
-domain.  The improved accuracy of the OTS-NIDC solution compared to a simple 
+domain.  The improved accuracy of the OTS-DC solution compared to a simple 
 forward Euler scheme with a suboptimal time step is illustrated in 
@@
 coordinate direction for two finite difference schemes that solve the 2D 
-diffusion equation on an irregular domain: forward Euler with OTS-NIDC 
+diffusion equation on an irregular domain: forward Euler with OTS-DC 
 (circles) and forward Euler with suboptimal time step (squares).  
@@
 \subsubsection{Diffusion Equation in Higher Dimensions}
-Using OTS-NIDC to boost the accuracy of finite difference
+Using OTS-DC to boost the accuracy of finite difference
 schemes for the diffusion equation in higher space dimensions is 
@@
 \section{\label{sec:summary} Summary} 
-In this article, we have presented OTS-NIDC, a simple approach for 
+In this article, we have presented OTS selection, a simple approach for
 constructing high-order finite difference methods for time dependent PDEs
 \emph{without} requiring the use of high-order stencils or high-order time
-integration schemes.  The major idea underlying OTS-NIDC is that 
-spatial and temporal discretization errors can be simultaneously eliminated
-by (1) using the PDE to relate terms in the discretization error, (2)
-optimally selecting numerical parameters, and (3) adding a few defect 
-correction terms in a non-iterative manner.  The boost in the order of 
-accuracy achieved by using OTS-NIDC is well worth the extra effort required to 
-derive the optimal time step and defect correction terms.
-
-Through many examples, we have demonstrated the utility of OTS-NIDC
-to several types of finite difference schemes for a wide range of problems.  
-We showed how OTS-NIDC can be used to very easily obtain high-order accurate 
-solutions to many linear and semilinear PDEs in any number of space dimensions 
-on both regular and irregular domains.  The examples illustrate the 
-high-order of accuracy that can be achieved using only low-order 
-discretizations for spatial derivatives and simple forward Euler time 
-integration.  They also demonstrate the applicability of OTS-NIDC to more 
-exotic finite difference schemes, such as the Kreiss-Petersson-Ystr\"om 
-discretization of the second-order wave equation and the DuFort-Frankel scheme 
-for the diffusion equation.
+integration schemes.  The essential idea underlying OTS selection is that 
+leading order spatial and temporal discretization errors can be simultaneously
+eliminated by optimally selecting numerical discretization parameters.  Any
+terms in the truncation error that remain are handled in a straightforward
+manner using non-iterative defect correction.  The boost in the order of
+accuracy achieved by using OTS-DC is well worth the extra effort required to
+derive the optimal time step and defect correction terms.  In fact, the amount
+of analysis required to compute the OTS and correction terms is comparable to
+traditional defect/deferred correction and modified equation methods.
+
+Through many examples, we have demonstrated the utility of OTS-DC to several
+types of finite difference schemes for a wide range of problems.  We showed
+how OTS-DC can be used to very easily obtain high-order accurate solutions to
+many linear and semilinear PDEs in any number of space dimensions on both
+regular and irregular domains.  The examples illustrate the high-order of
+accuracy that can be achieved using only low-order discretizations for spatial
+derivatives and simple forward Euler time integration.  They also demonstrate
+the applicability of OTS-DC to more exotic finite difference schemes, such as
+the Kreiss-Petersson-Ystr\"om discretization of the second-order wave equation
+and the DuFort-Frankel scheme for the diffusion equation.
 
@@
 numerical methods.  Little research appears to have been done in this area,
-but as OTS-NIDC selection demonstrates, optimal selection of numerical 
+but as OTS-DC selection demonstrates, optimal selection of numerical 
 parameters can significantly impact the accuracy of numerical methods.  To 
@@
 error in the initial conditions is degraded by one temporal order of accuracy
-by the KPY scheme.  Therefore, KPY without OTS-NIDC requires the first time 
-step must be at least third-order accurate; KPY with OTS-NIDC requires the 
+by the KPY scheme.  Therefore, KPY without OTS-DC requires the first time 
+step must be at least third-order accurate; KPY with OTS-DC requires the 
 first time step must be at least fifth-order accurate.
@@
 
+\bibitem{griffiths_1986}
+D.~F. Griffiths and J.~M. Sanz-Serna.
+\newblock On the {S}cope of the {M}ethod of {M}odified {E}quations.
+\newblock {\em SIAM J. Sci. Stat. Comput.}, 7:994--1008, 1986.
+
 \bibitem{fedkiw_1999}
@@
 \end{thebibliography}
-
 \end{document}

### 2010-05-21 papers/OTSPDE/OTSPDE.tex
Added brief discussion of benefits of OTS selection for GPU computing.

@@
 % Resubmission draft: 2010/03/25, Serendipity Research
-% Resubmission draft: 2010/05/02, Serendipity Research
+% Resubmission draft: 2010/05/21, Serendipity Research
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 for a wide array of classical linear and semilinear PDEs in one and more space
-dimensions on both regular and irregular domains.  
+dimensions on both regular and irregular domains.  OTS selection is 
+particularly valuable in light of recent advances in GPGPU computing because
+it significantly increases the usefulness of simple explicit time stepping
+schemes, which are especially well-suited to GPU hardware.
 \end{abstract}
@@
 
+From a scientific computing perspective, recent advances in GPGPU computing
+make optimal time step selection especially relevant.  Because explicit time
+integration schemes are better suited to the GPU hardware architectures than
+implicit time integration schemes, the ability of optimal time step selection
+to dramatically boost the accuracy of simple explicit schemes makes it a
+powerful tool for achieving high-performance on GPUs.
+
 % ACKNOWLEDGMENTS

### 2010-05-21 src/scalar_PDEs/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m
Updated MATLAB scripts for generating figures.

@@ -197,8 +197,8 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(20,1e-5,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(25,1e-5,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;

### 2010-05-21 src/scalar_PDEs/1d/burgers_eqn/single_hump/genPresoFigs_BurgersEqn1d_SingleHump.m

@@ -191,8 +191,8 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(20,1e-5,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(25,1e-5,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;

### 2010-05-21 src/scalar_PDEs/1d/diffusion_eqn/genFigs_DiffusionEqn1d_NoSrc.m

@@ -255,7 +255,7 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
 text(12,1e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
@@ -291,7 +291,7 @@ hold on;
 loglog(err_FE_OTS(3:end), comp_time_FE_OTS(3:end), 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(3e-10,7e-4,order_str);
 
 loglog(err_plot, ...

### 2010-05-21 src/scalar_PDEs/1d/diffusion_eqn/genFigs_DiffusionEqn1d_NoSrc_DuFortFrankel.m

@@ -225,7 +225,7 @@ hold on;
 plot(grid_sizes,err_DF_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('DuFort-Frankel (OTS-NIDC)\nOrder = %1.1f', order_DF_OTS);
+order_str = sprintf('DuFort-Frankel (OTS-DC)\nOrder = %1.1f', order_DF_OTS);
 text(13,1e-10,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_DF(1)+P_DF(2)),'k');
@@ -254,7 +254,7 @@ hold on;
 loglog(err_DF_OTS, comp_time_DF_OTS, 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('DuFort-Frankel (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_DF_OTS);
+order_str = sprintf('DuFort-Frankel (OTS-DC)\nSlope = %1.1f', comp_time_exp_DF_OTS);
 text(2e-10,1e-3,order_str);
 
 loglog(err_plot, ...

### 2010-05-21 src/scalar_PDEs/1d/diffusion_eqn/genFigs_DiffusionEqn1d_Src.m

@@ -275,9 +275,9 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler OTS-NIDC\nOrder = %1.1f', ...
+order_str = sprintf('Forward Euler OTS-DC\nOrder = %1.1f', ...
   order_FE_OTS);
-text(15,4.5e-9,order_str);
+text(20,4.5e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS_no_corr(1)+P_FE_OTS_no_corr(2)),'k');
 hold on;
@@ -319,7 +319,7 @@ hold on;
 loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler OTS-NIDC\nSlope = %1.1f', ...
+order_str = sprintf('Forward Euler OTS-DC\nSlope = %1.1f', ...
   comp_time_exp_FE_OTS);
 text(1e-9,3e-2,order_str);
 
@@ -368,7 +368,7 @@ plot(x_lo_res,u_FE_OTS_lo_res,'bo')
 hold on;
 plot(x,u_exact,'r')
 xlabel('x');
-%title('Forward Euler OTS-NIDC Solution')
+%title('Forward Euler OTS-DC Solution')
 filename = sprintf('diffusion_eqn_1d_src_FE_OTS_soln.%s', print_suffix);
 format_str = sprintf('-d%s', print_format);
 print([fig_dir, '/', filename], format_str);

### 2010-05-21 src/scalar_PDEs/1d/diffusion_eqn/genFigs_DiffusionEqn1d_Src_DuFortFrankel.m

@@ -225,8 +225,8 @@ hold on;
 plot(grid_sizes,err_DF_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('DuFort-Frankel (OTS-NIDC)\nOrder = %1.1f', order_DF_OTS);
-text(13,1e-9,order_str);
+order_str = sprintf('DuFort-Frankel (OTS-DC)\nOrder = %1.1f', order_DF_OTS);
+text(20,1e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_DF(1)+P_DF(2)),'k');
 hold on;
@@ -254,8 +254,8 @@ hold on;
 loglog(err_DF_OTS, comp_time_DF_OTS, 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('DuFort-Frankel (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_DF_OTS);
-text(2e-10,5e-3,order_str);
+order_str = sprintf('DuFort-Frankel (OTS-DC)\nSlope = %1.1f', comp_time_exp_DF_OTS);
+text(9e-10,5e-3,order_str);
 
 loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_DF(1)+P_comp_time_DF(2)), ...

### 2010-05-21 src/scalar_PDEs/1d/diffusion_eqn/genPresoFigs_DiffusionEqn1d_NoSrc.m

@@ -249,7 +249,7 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
 text(12,1e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
@@ -285,7 +285,7 @@ hold on;
 loglog(err_FE_OTS(3:end), comp_time_FE_OTS(3:end), 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(3e-10,7e-4,order_str);
 
 loglog(err_plot, ...

### 2010-05-21 src/scalar_PDEs/1d/diffusion_eqn/genPresoFigs_DiffusionEqn1d_Src.m

@@ -269,9 +269,9 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler OTS-NIDC\nOrder = %1.1f', ...
+order_str = sprintf('Forward Euler OTS-DC\nOrder = %1.1f', ...
   order_FE_OTS);
-text(15,4.5e-9,order_str);
+text(20,4.5e-9,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE_OTS_no_corr(1)+P_FE_OTS_no_corr(2)),'k');
 hold on;
@@ -313,9 +313,9 @@ hold on;
 loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler OTS-NIDC\nSlope = %1.1f', ...
+order_str = sprintf('Forward Euler OTS-DC\nSlope = %1.1f', ...
   comp_time_exp_FE_OTS);
-  text(1e-9,3e-2,order_str);
+text(1e-9,3e-2,order_str);
 
 loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE_OTS_no_corr(1) ...

### 2010-05-21 src/scalar_PDEs/1d/fourth_order_parabolic/genFigs_4thOrderParabolicEqn1d.m

@@ -291,8 +291,8 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(15,1e-9,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(20,1e-9,order_str);
 annotation(f, 'arrow', [.63 .7], [.2 .2], ...
            'linewidth', 4, 'headstyle', 'plain');
 
@@ -377,8 +377,8 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.2f', comp_time_exp_FE_OTS);
-text(2e-6,9e4,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.2f', comp_time_exp_FE_OTS);
+text(3e-6,9e4,order_str);
 annotation(f, 'arrow', [.48 .32], [.86 .77], ...
            'linewidth', 4, 'headstyle', 'plain');
 

### 2010-05-21 src/scalar_PDEs/1d/fourth_order_parabolic/genPresoFigs_4thOrderParabolicEqn1d.m

@@ -285,8 +285,8 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(15,1e-9,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(20,1e-9,order_str);
 annotation(f, 'arrow', [.63 .7], [.2 .2], ...
            'linewidth', 4, 'headstyle', 'plain');
 
@@ -371,8 +371,8 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.2f', comp_time_exp_FE_OTS);
-text(2e-6,9e4,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.2f', comp_time_exp_FE_OTS);
+text(3e-6,9e4,order_str);
 annotation(f, 'arrow', [.48 .32], [.86 .77], ...
            'linewidth', 4, 'headstyle', 'plain');
 

### 2010-05-21 src/scalar_PDEs/1d/wave_eqn/genFigs_WaveEqn1d_NoSrc.m

@@ -101,7 +101,7 @@ for i = 1:length(grid_sizes)
     dt_KPY = 0.5*dx/c;
 
     % solve wave equation using KPY with OTS
-    disp('KPY-OTS-NIDC');
+    disp('KPY-OTS-DC');
     [u_KPY_OTS, u_exact, x] = solveWaveEqn1dKPY_OTS(c, ...
                                                     use_source_term, ...
                                                     N, ...
@@ -162,7 +162,7 @@ else
   dt_KPY = 0.5*dx/c;
 
   % solve wave equation using KPY with OTS
-  disp('KPY-OTS-NIDC');
+  disp('KPY-OTS-DC');
   [u_KPY_OTS_lo_res, u_exact_lo_res, x_lo_res] = ...
     solveWaveEqn1dKPY_OTS(c, ...
                           use_source_term, ...
@@ -190,7 +190,7 @@ else
   dt_KPY = 0.5*dx/c;
 
   % solve wave equation using KPY with OTS
-  disp('KPY-OTS-NIDC');
+  disp('KPY-OTS-DC');
   [u_KPY_OTS_hi_res, u_exact_hi_res, x_hi_res] = ...
     solveWaveEqn1dKPY_OTS(c, ...
                           use_source_term, ...
@@ -226,7 +226,7 @@ hold on;
 plot(grid_sizes,err_KPY_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('KPY (OTS-NIDC)\nOrder = %1.1f', order_KPY_OTS);
+order_str = sprintf('KPY (OTS-DC)\nOrder = %1.1f', order_KPY_OTS);
 text(150,1e-13,order_str);
 
 N_plot = [100 10000];
@@ -251,7 +251,7 @@ hold on;
 plot(x_hi_res,u_exact_hi_res,'r')
 axis([x_lo x_hi -5 5]);
 xlabel('x');
-%title('KPY (OTS-NIDC) Solution')
+%title('KPY (OTS-DC) Solution')
 filename = sprintf('wave_eqn_1d_no_src_KPY_OTS_soln.%s', print_suffix);
 format_str = sprintf('-d%s',print_format);
 print([fig_dir, '/', filename], format_str);

### 2010-05-21 src/scalar_PDEs/1d/wave_eqn/genFigs_WaveEqn1d_Src.m

@@ -101,7 +101,7 @@ for i = 1:length(grid_sizes)
     dt_KPY = 0.5*dx/c;
 
     % solve wave equation using KPY with OTS
-    disp('KPY-OTS-NIDC');
+    disp('KPY-OTS-DC');
     [u_KPY_OTS, u_exact, x] = solveWaveEqn1dKPY_OTS(c, ...
                                                     use_source_term, ...
                                                     N, ...
@@ -162,7 +162,7 @@ else
   dt_KPY = 0.5*dx/c;
 
   % solve wave equation using KPY with OTS
-  disp('KPY-OTS-NIDC');
+  disp('KPY-OTS-DC');
   [u_KPY_OTS_lo_res, u_exact_lo_res, x_lo_res] = ...
     solveWaveEqn1dKPY_OTS(c, ...
                           use_source_term, ...
@@ -190,7 +190,7 @@ else
   dt_KPY = 0.5*dx/c;
 
   % solve wave equation using KPY with OTS
-  disp('KPY-OTS-NIDC');
+  disp('KPY-OTS-DC');
   [u_KPY_OTS_hi_res, u_exact_hi_res, x_hi_res] = ...
     solveWaveEqn1dKPY_OTS(c, ...
                           use_source_term, ...
@@ -226,8 +226,8 @@ hold on;
 plot(grid_sizes,err_KPY_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('KPY (OTS-NIDC)\nOrder = %1.1f', order_KPY_OTS);
-text(180,2e-13,order_str);
+order_str = sprintf('KPY (OTS-DC)\nOrder = %1.1f', order_KPY_OTS);
+text(200,2e-13,order_str);
 
 N_plot = [95 10000];
 loglog(N_plot,exp(log(N_plot)*P_KPY(1)+P_KPY(2)),'k');
@@ -252,7 +252,7 @@ hold on;
 plot(x_hi_res,u_exact_hi_res,'r')
 axis([x_lo x_hi -5 5]);
 xlabel('x');
-%title('KPY (OTS-NIDC) Solution')
+%title('KPY (OTS-DC) Solution')
 filename = sprintf('wave_eqn_1d_src_KPY_OTS_soln.%s', print_suffix);
 format_str = sprintf('-d%s',print_format);
 print([fig_dir, '/', filename], format_str);

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/irregular_domain/circle_domain/genFigs_DiffusionEqn2d_CircleDomain.m

@@ -130,9 +130,9 @@ for i = 1:length(grid_sizes)
                        N, bc_order);
     disp(disp_str);
 
-    % solve diffusion equation using forward Euler with OTS-NIDC
+    % solve diffusion equation using forward Euler with OTS-DC
     disp('---------------------------');
-    disp('  Forward Euler OTS-NIDC');
+    disp('  Forward Euler OTS-DC');
     disp('---------------------------');
     [u_FE_OTS, u_exact, X, Y, timing_data_FE_OTS] = ...
        solveDiffusionEqnForwardEulerOTS2d(D, ...
@@ -232,7 +232,7 @@ hold on;
 plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
      'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
 text(12,3e-5,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
@@ -262,7 +262,7 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
        'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', comp_time_exp_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', comp_time_exp_FE_OTS);
 text(3e-8,0.5,order_str);
 
 loglog(err_plot, exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), 'k');

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/irregular_domain/circle_domain/genPresoFigs_DiffusionEqn2d_CircleDomain.m

@@ -123,9 +123,9 @@ for i = 1:length(grid_sizes)
                        N, bc_order);
     disp(disp_str);
 
-    % solve diffusion equation using forward Euler with OTS-NIDC
+    % solve diffusion equation using forward Euler with OTS-DC
     disp('---------------------------');
-    disp('  Forward Euler OTS-NIDC');
+    disp('  Forward Euler OTS-DC');
     disp('---------------------------');
     [u_FE_OTS, u_exact, X, Y, timing_data_FE_OTS] = ...
        solveDiffusionEqnForwardEulerOTS2d(D, ...
@@ -225,7 +225,7 @@ hold on;
 plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
      'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
 text(12,3e-5,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
@@ -255,7 +255,7 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
        'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', comp_time_exp_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', comp_time_exp_FE_OTS);
 text(3e-8,0.5,order_str);
 
 loglog(err_plot, exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), 'k');

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/irregular_domain/starfish_domain/genFigs_DiffusionEqn2d_StarfishDomain.m

@@ -130,9 +130,9 @@ for i = 1:length(grid_sizes)
                        N, bc_order);
     disp(disp_str);
 
-    % solve diffusion equation using forward Euler with OTS-NIDC
+    % solve diffusion equation using forward Euler with OTS-DC
     disp('---------------------------');
-    disp('  Forward Euler OTS-NIDC');
+    disp('  Forward Euler OTS-DC');
     disp('---------------------------');
     [u_FE_OTS, u_exact, X, Y, timing_data_FE_OTS] = ...
        solveDiffusionEqnForwardEulerOTS2d(D, ...
@@ -231,8 +231,8 @@ hold on;
 plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
      'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(20,3e-6,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(25,3e-6,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
 hold on;
@@ -261,8 +261,8 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
        'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', comp_time_exp_FE_OTS);
-text(3e-8,0.5,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', comp_time_exp_FE_OTS);
+text(5e-8,0.5,order_str);
 
 loglog(err_plot, exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), 'k');
 hold on;

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/irregular_domain/starfish_domain/genPresoFigs_DiffusionEqn2d_StarfishDomain.m

@@ -123,9 +123,9 @@ for i = 1:length(grid_sizes)
                        N, bc_order);
     disp(disp_str);
 
-    % solve diffusion equation using forward Euler with OTS-NIDC
+    % solve diffusion equation using forward Euler with OTS-DC
     disp('---------------------------');
-    disp('  Forward Euler OTS-NIDC');
+    disp('  Forward Euler OTS-DC');
     disp('---------------------------');
     [u_FE_OTS, u_exact, X, Y, timing_data_FE_OTS] = ...
        solveDiffusionEqnForwardEulerOTS2d(D, ...
@@ -220,8 +220,8 @@ hold on;
 plot(grid_sizes, err_FE_OTS, 'bo', ...
      'MarkerSize', 14, ...
      'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(20,3e-6,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(25,3e-6,order_str);
 
 loglog(N_plot, exp(log(N_plot)*P_FE(1)+P_FE(2)), 'k');
 hold on;
@@ -250,9 +250,9 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize', 14, ...
        'MarkerFaceColor', 'b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', ...
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', ...
                      comp_time_exp_FE_OTS);
-text(3e-8,0.5,order_str);
+text(5e-8,0.5,order_str);
 
 loglog(err_plot, exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), 'k');
 hold on;

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/square_domain/genFigs_DiffusionEqn2d_NoSrc.m

@@ -216,7 +216,7 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
 text(13,2e-10,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
@@ -252,7 +252,7 @@ hold on;
 loglog(err_FE_OTS(2:end), comp_time_FE_OTS(2:end), 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
 text(5e-12,1e-3,order_str);
 
 loglog(err_plot, ...

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/square_domain/genFigs_DiffusionEqn2d_Src.m

@@ -226,8 +226,8 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(13,1e-8,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(15,1e-8,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
@@ -262,8 +262,8 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
-text(3e-8,3e-3,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
+text(8e-8,3e-3,order_str);
 
 loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), ...

### 2010-05-21 src/scalar_PDEs/2d/diffusion_eqn/square_domain/genPresoFigs_DiffusionEqn2d_Src.m

@@ -220,8 +220,8 @@ hold on;
 plot(grid_sizes,err_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nOrder = %1.1f', order_FE_OTS);
-text(13,1e-8,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nOrder = %1.1f', order_FE_OTS);
+text(15,1e-8,order_str);
 
 loglog(N_plot,exp(log(N_plot)*P_FE(1)+P_FE(2)),'k');
 hold on;
@@ -256,8 +256,8 @@ hold on;
 loglog(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
        'MarkerSize',14, ...
        'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
-text(3e-8,3e-3,order_str);
+order_str = sprintf('Forward Euler (OTS-DC)\nSlope = %1.1f', comp_time_exp_FE_OTS);
+text(8e-8,3e-3,order_str);
 
 loglog(err_plot, ...
        exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)), ...

### 2010-05-21 papers/OTSPDE/submissions/201005-Applied_Mathematics_and_Computation/cover_letter.tex
Added Applied Mathematics and Computations submission information.

--- 2009-05-25 papers/OTSPDE/submissions/200905-Applied_Numerical_Math/cover_letter.tex
@@
 
-\opening{Dear Applied Numerical Mathematics Editor/Referee,}
+\opening{Dear Applied Mathematics and Computations Editor/Reviewer,}
 
-My article titled \emph{Boosting the Accuracy of Finite Difference Schemes via
-Optimal Time Step Selection and Non-Iterative Defect Correction} discusses a 
-simple technique for boosting the order of accuracy of finite difference 
-schemes for time dependent PDEs by choosing an optimal, \emph{fixed} time step
-and adding defect correction terms in a \emph{non-iterative} manner.  As 
-discussed in the paper, a careful choice of time step can dramatically 
-reduce the numerical error of simple finite difference schemes and make 
-it possible to achieve high-order accuracy using only formally low-order 
-finite difference stencils and time integration schemes.
+My article titled \emph{Using Optimal Time Step Selection to Boost the
+Accuracy of Finite Difference Schemes for Time Dependent PDEs} discusses a 
+simple, systematic technique for boosting the order of accuracy of finite
+difference schemes for time dependent PDEs by choosing an optimal, 
+\emph{fixed} time step (and possibly adding defect correction terms in a 
+\emph{non-iterative} manner).  As discussed in the paper, a careful choice of
+time step can reduce the order of the numerical error of simple finite 
+difference schemes and make it possible to achieve high-order accuracy using 
+only formally low-order finite difference stencils and time integration 
+schemes.
 
@@
 
+In addition to being an interesting method, I believe that this method has
+value in light of recent advances in GPGPU computing.  Because modern GPU
+hardware is more compatible with explicit time stepping schemes, the ability
+of optimal time step selection to dramaticaly boost the accuracy of simple
+explicit schemes makes it a powerful tool for achieving high-performance
+on GPUs.
+
 Thank you for your review of the paper.  I look forward to hearing your 

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m
Minor variable name change.

@@ -61,9 +61,9 @@ err_FE_OTS = zeros(size(grid_sizes));
 err_FE     = zeros(size(grid_sizes));
 
 % allocate memory for computation times
-comp_time_FE_OTS           = zeros(size(grid_sizes));
-comp_time_FE_OTS_corr_term = zeros(size(grid_sizes));
-comp_time_FE               = zeros(size(grid_sizes));
+comp_time_FE_OTS      = zeros(size(grid_sizes));
+corr_term_time_FE_OTS = zeros(size(grid_sizes));
+comp_time_FE          = zeros(size(grid_sizes));
 
 % start clock for timing plot generation time
 tic;
@@ -137,7 +137,7 @@ for i = 1:length(grid_sizes)
 
 % collect timing data
   comp_time_FE_OTS(i) = timing_data_FE_OTS(1);
-  comp_time_FE_OTS_corr_term(i) = timing_data_FE_OTS(2);
+  corr_term_time_FE_OTS(i) = timing_data_FE_OTS(2);
   comp_time_FE(i) = timing_data_FE(1);
 
 end % end loop over grid sizes

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genPresoFigs_BurgersEqn1d_SingleHump.m

@@ -55,9 +55,9 @@ err_FE_OTS = zeros(size(grid_sizes));
 err_FE     = zeros(size(grid_sizes));
 
 % allocate memory for computation times
-comp_time_FE_OTS           = zeros(size(grid_sizes));
-comp_time_FE_OTS_corr_term = zeros(size(grid_sizes));
-comp_time_FE               = zeros(size(grid_sizes));
+comp_time_FE_OTS      = zeros(size(grid_sizes));
+corr_term_time_FE_OTS = zeros(size(grid_sizes));
+comp_time_FE          = zeros(size(grid_sizes));
 
 % start clock for timing plot generation time
 tic;
@@ -131,7 +131,7 @@ for i = 1:length(grid_sizes)
 
 % collect timing data
   comp_time_FE_OTS(i) = timing_data_FE_OTS(1);
-  comp_time_FE_OTS_corr_term(i) = timing_data_FE_OTS(2);
+  corr_term_time_FE_OTS(i) = timing_data_FE_OTS(2);
   comp_time_FE(i) = timing_data_FE(1);
 
 end % end loop over grid sizes

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m
Updated calculation of slope for performance plot for Burgers equation.

@@ -34,7 +34,7 @@ if ~exist(fig_dir, 'dir')
 end
 
 % set flag for loading data from saved files (instead of recomputing solution)
-use_saved_data = 0;
+use_saved_data = 1;
 data_dir = 'data-burgers_eqn_1d-single_hump';
 if ~exist(data_dir, 'dir')
   mkdir(data_dir);
@@ -201,10 +201,10 @@ order_FE_OTS = -P_FE_OTS(1);
 P_FE = polyfit(log(grid_sizes(2:end)),log(err_FE(2:end)),1);
 order_FE = -P_FE(1);
 
-P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(2:end)), ...
-                             log(comp_time_FE_OTS(2:end)),1);
-P_comp_time_FE = polyfit(log(err_FE(3:end)), ...
-                         log(comp_time_FE(3:end)),1);
+P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(5:end)), ...
+                             log(comp_time_FE_OTS(5:end)),1);
+P_comp_time_FE = polyfit(log(err_FE(5:end)), ...
+                         log(comp_time_FE(5:end)),1);
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Plot Results
@@ -243,7 +243,7 @@ hold on;
 plot(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', P_comp_time_FE_OTS(1));
+order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.2f', P_comp_time_FE_OTS(1));
 text(1e-9,2e-2,order_str);
 
 loglog(err_plot,exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)),'k');

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genPresoFigs_BurgersEqn1d_SingleHump.m

@@ -28,7 +28,7 @@ if ~exist(fig_dir, 'dir')
 end
 
 % set flag for loading data from saved files (instead of recomputing solution)
-use_saved_data = 0;
+use_saved_data = 1;
 data_dir = 'data-burgers_eqn_1d-single_hump';
 if ~exist(data_dir, 'dir')
   mkdir(data_dir);
@@ -195,10 +195,10 @@ order_FE_OTS = -P_FE_OTS(1);
 P_FE = polyfit(log(grid_sizes(2:end)),log(err_FE(2:end)),1);
 order_FE = -P_FE(1);
 
-P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(2:end)), ...
-                             log(comp_time_FE_OTS(2:end)),1);
-P_comp_time_FE = polyfit(log(err_FE(3:end)), ...
-                         log(comp_time_FE(3:end)),1);
+P_comp_time_FE_OTS = polyfit(log(err_FE_OTS(5:end)), ...
+                             log(comp_time_FE_OTS(5:end)),1);
+P_comp_time_FE = polyfit(log(err_FE(5:end)), ...
+                         log(comp_time_FE(5:end)),1);
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Plot Results
@@ -237,7 +237,7 @@ hold on;
 plot(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
-order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.1f', P_comp_time_FE_OTS(1));
+order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.2f', P_comp_time_FE_OTS(1));
 text(1e-9,2e-2,order_str);
 
 loglog(err_plot,exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)),'k');

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/solveBurgersEqn1dForwardEuler.m
Removed collection of correction term computation time because it affects collection of total solution time.

@@ -59,8 +59,8 @@
 % - u_exact:             analytical solution
 % - x:                   grid points
 % - timing_data:         array of the following form containing timing data
-%                        [total solution time.  If timing is not activated,
-%                        timing_data is set to be an array of -1's.
+%                        [total solution time].  If timing is not activated,
+%                        timing_data is set to [-1].
 %
 %
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/solveBurgersEqn1dForwardEulerOTS.m

@@ -59,9 +59,8 @@
 % - u_exact:             analytical solution
 % - x:                   grid points
 % - timing_data:         array of the following form containing timing data
-%                        [total solution time, time to compute correction term].
-%                        If timing is not activated, timing_data is
-%                        set to [-1, -1].
+%                        [total solution time].  If timing is not activated,
+%                        timing_data is [-1].
 %
 %
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@ -163,17 +162,9 @@ while (t < t_final)
     dt = t_final - t; 
   end
 
-  % compute correction term
-  if (timing_on == 1)
-    t_start_correction_term = cputime;
-  end
-  corr_term = - 0.5*dt^2*( 4*nu*(G*u).*(L*u) - 2*u.*(G*u).^2 - (u.^2).*(L*u) ); 
-  if (timing_on == 1)
-    t_corr_term = t_corr_term + cputime - t_start_correction_term; 
-  end
-
   % update solution
-  u = u + dt*(nu*(L*u) - u.*(G*u)) + corr_term;
+  u = u + dt*(nu*(L*u) - u.*(G*u)) ...
+    - 0.5*dt^2*( 4*nu*(G*u).*(L*u) - 2*u.*(G*u).^2 - (u.^2).*(L*u) ); 
 
   % update time
   t = t + dt;
@@ -199,7 +190,7 @@ u_exact = U +  sqrt(nu/T)*gamma*exp(-(x-U*T).^2/(4*nu*T)) ...
 
 % set timing data
 if (timing_on == 1)
-  timing_data = [t_solve, t_corr_term];
+  timing_data = [t_solve];
 else
-  timing_data = [-1, -1];
+  timing_data = [-1];
 end

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m
Removed unused variable.

@@ -62,7 +62,6 @@ err_FE     = zeros(size(grid_sizes));
 
 % allocate memory for computation times
 comp_time_FE_OTS      = zeros(size(grid_sizes));
-corr_term_time_FE_OTS = zeros(size(grid_sizes));
 comp_time_FE          = zeros(size(grid_sizes));
 
 % start clock for timing plot generation time
@@ -137,7 +136,6 @@ for i = 1:length(grid_sizes)
 
 % collect timing data
   comp_time_FE_OTS(i) = timing_data_FE_OTS(1);
-  corr_term_time_FE_OTS(i) = timing_data_FE_OTS(2);
   comp_time_FE(i) = timing_data_FE(1);
 
 end % end loop over grid sizes

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genPresoFigs_BurgersEqn1d_SingleHump.m

@@ -56,7 +56,6 @@ err_FE     = zeros(size(grid_sizes));
 
 % allocate memory for computation times
 comp_time_FE_OTS      = zeros(size(grid_sizes));
-corr_term_time_FE_OTS = zeros(size(grid_sizes));
 comp_time_FE          = zeros(size(grid_sizes));
 
 % start clock for timing plot generation time
@@ -131,7 +130,6 @@ for i = 1:length(grid_sizes)
 
 % collect timing data
   comp_time_FE_OTS(i) = timing_data_FE_OTS(1);
-  corr_term_time_FE_OTS(i) = timing_data_FE_OTS(2);
   comp_time_FE(i) = timing_data_FE(1);
 
 end % end loop over grid sizes

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m
Regenerated data for 1D Burgers equation so that the timing data does not include overhead of timing the correction term calculation.

@@ -34,7 +34,7 @@ if ~exist(fig_dir, 'dir')
 end
 
 % set flag for loading data from saved files (instead of recomputing solution)
-use_saved_data = 1;
+use_saved_data = 0;
 data_dir = 'data-burgers_eqn_1d-single_hump';
 if ~exist(data_dir, 'dir')
   mkdir(data_dir);

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genFigs_BurgersEqn1d_SingleHump.m
Fixed output filename and adjusted text placement.

@@ -34,7 +34,7 @@ if ~exist(fig_dir, 'dir')
 end
 
 % set flag for loading data from saved files (instead of recomputing solution)
-use_saved_data = 0;
+use_saved_data = 1;
 data_dir = 'data-burgers_eqn_1d-single_hump';
 if ~exist(data_dir, 'dir')
   mkdir(data_dir);
@@ -242,7 +242,7 @@ plot(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.2f', P_comp_time_FE_OTS(1));
-text(1e-9,2e-2,order_str);
+text(1e-9,1e-2,order_str);
 
 loglog(err_plot,exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)),'k');
 hold on;
@@ -250,14 +250,14 @@ plot(err_FE, comp_time_FE, 'rs', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nSlope = %1.1f', P_comp_time_FE(1));
-text(1e-4,1e2,order_str);
+text(4e-4,1e2,order_str);
 
 axis([err_plot(1) err_plot(2) 1e-4 1e4]);
 xlabel('L^\infty Error');
 ylabel('Compute Time (s)');
 set(gca, 'XTick', 10.^[-10:2:0]);
 set(gca, 'YTick', 10.^[-4:2:4]);
-filename = sprintf('burgers_eqn_1d_error_vs_N_comp_time.%s', print_suffix);
+filename = sprintf('burgers_eqn_1d_comp_time.%s', print_suffix);
 format_str = sprintf('-d%s',print_format);
 print([fig_dir, '/', filename], format_str);
 

### 2011-07-27 src/scalar_PDEs/1d/burgers_eqn/single_hump/genPresoFigs_BurgersEqn1d_SingleHump.m

@@ -236,7 +236,7 @@ plot(err_FE_OTS, comp_time_FE_OTS, 'bo', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','b');
 order_str = sprintf('Forward Euler (OTS-NIDC)\nSlope = %1.2f', P_comp_time_FE_OTS(1));
-text(1e-9,2e-2,order_str);
+text(1e-9,1e-2,order_str);
 
 loglog(err_plot,exp(log(err_plot)*P_comp_time_FE(1)+P_comp_time_FE(2)),'k');
 hold on;
@@ -244,14 +244,14 @@ plot(err_FE, comp_time_FE, 'rs', ...
      'MarkerSize',14, ...
      'MarkerFaceColor','r');
 order_str = sprintf('Forward Euler\nSlope = %1.1f', P_comp_time_FE(1));
-text(1e-4,1e2,order_str);
+text(4e-4,1e2,order_str);
 
 axis([err_plot(1) err_plot(2) 1e-4 1e4]);
 xlabel('L^\infty Error');
 ylabel('Compute Time (s)');
 set(gca, 'XTick', 10.^[-10:2:0]);
 set(gca, 'YTick', 10.^[-4:2:4]);
-filename = sprintf('burgers_eqn_1d_error_vs_N_comp_time.%s', print_format);
+filename = sprintf('burgers_eqn_1d_comp_time.%s', print_format);
 format_str = sprintf('-d%s',print_format);
 print([fig_dir, '/', filename], format_str);
 

### 2011-07-27 papers/OTSPDE/OTSPDE.tex
Added keywords in OTSPDE.tex
Revised OTSPDE paper for resubmission to AMC.

--- 2009-05-25 papers/OTSPDE/OTSPDE.tex
@@
 % Resubmission draft: 2009/05/25, Serendipity Research
+% Resubmission draft: 2010/03/25, Serendipity Research
+% Resubmission draft: 2010/05/21, Serendipity Research
+% Resubmission draft: 2011/07/30, Serendipity Research
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \usepackage{graphicx}
+\usepackage{subfig}
 \PassOptionsToPackage{ctagsplt,righttag}{amsmath}
@@
 
-\title{Boosting the Accuracy of Finite Difference Schemes via
-       Optimal Time Step Selection and Non-Iterative Defect Correction}
-
-\author{Kevin T. Chu\address{Institute of High Performance Computing, A*STAR, Singapore, Singapore}$^,$\address{Vitamin D, Inc., Menlo Park, CA 94025, United States}
+\title{Boosting the Accuracy of Finite Difference Schemes via Optimal Time
+Step Selection and Non-Iterative Defect Correction}
+
+\author{Kevin T. Chu\address{Serendipity Research, Mountain View, CA 94041, United States}
 }
@@
 \begin{abstract}
-In this article, we present a simple technique for boosting the order of 
-accuracy of finite difference schemes for time dependent 
-partial differential equations by optimally selecting the time step used
-to advance the numerical solution and adding defect correction terms in a
-non-iterative manner.  The power of the technique is its ability to extract 
-as much accuracy as possible from existing finite difference schemes with 
-minimal additional effort.  Through straightforward numerical analysis 
-arguments, we explain the origin of the boost in accuracy and estimate the 
-computational cost of the resulting numerical method.  We demonstrate the 
-utility of optimal time step (OTS) selection combined with non-iterative 
-defect correction (NIDC) on several different types of finite difference
-schemes for a wide array of classical linear and semilinear PDEs in one and 
-more space dimensions on both regular and irregular domains.  
+In this article, we present a unified analysis of the simple technique for
+boosting the order of accuracy of finite difference schemes for time dependent
+partial differential equations by optimally selecting the time step used to
+advance the numerical solution and adding defect correction terms in a
+non-iterative manner.
+The power of the technique is its ability to increase the accuracy of formally
+low-order finite difference schemes without major modification to the basic 
+numerical algorithm.  Through straightforward numerical analysis arguments, we
+explain the origin of the boost in accuracy and estimate the computational cost
+of the resulting numerical method.  We demonstrate the utility of optimal time
+step (OTS) selection combined with non-iterative defect correction (NIDC) on
+several different types of finite difference schemes for a wide array of
+classical linear and semilinear PDEs in one and more space dimensions on both
+regular and irregular domains.
+
+% KEYWORDS
+\ \newline
+Keywords: optimal time step; non-iterative defect correction; finite difference
+schemes; high-order accurate numerical methods; time dependent PDEs
+
 \end{abstract}
@@
 \noindent \rule{6.3in}{1pt}
-
-% KEYWORDS
-% optimal time step; non-iterative defect correction; finite difference 
-% schemes; high-order accurate numerical methods; time dependent PDEs
 
@@
 numerical PDEs continues to be focused on the development of high-order 
-numerical 
-schemes~\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
+numerical schemes
+\cite{spotz_2001,bruger_2005,gibou_2005,ito_2005,shukla_2005,heidenreich_2007,shukla_2007}.  
 Typically, high-order accuracy is achieved by constructing
@@
 ways (\eg Richardson extrapolation and compact finite difference 
-schemes~\cite{spotz_2001}).  When possible, the latter approach can be a 
-powerful way to boost the accuracy of a numerical method \emph{without} 
+schemes \cite{spotz_2001}).  When possible, the latter approach can be a 
+powerful way to boost the accuracy of a numerical method without
 introducing too much additional algorithmic (and programming) complexity.
 
-Optimization of numerical parameters (\eg $\dx$ and $\dt$) is an important, 
-but relatively uncommon, approach for maximizing the accuracy of numerical 
-schemes.  Often, all combinations of numerical parameters are considered 
-equally good.  For some problems, however, optimizing the numerical parameters 
-can significantly increase the accuracy of the numerical solution.  For 
-example, the well-known unit CFL condition, $\dt = \dx/a$, for the first-order 
-upwind forward Euler discretization of the linear advection equation 
-completely eliminates the numerical error introduced by the 
-scheme~\cite{leveque_book_2002}.  Another well-known example is the boost in 
-the order of accuracy that results from setting the time step to 
-$\dt = \dx^2/6D$ when the 1D diffusion equation is solved using forward Euler 
+One increasingly important approach for improving the accuracy of a numerical
+algorithm is optimization of its numerical parameters (\eg $\dx$ and $\dt$).
+For some problems, this optimization can dramatically increase the accuracy of
+the numerical solution. Two well-known examples are: (1) the unit 
+Courant-Friedrichs-Lewy (CFL) condition,
+$\dt = \dx/a$, for the first-order upwind forward Euler discretization of the
+linear advection equation, which completely eliminates the numerical error
+introduced by the scheme \cite{leveque_book_2002} and (2) the boost from
+second- to fourth-order accuracy that results if the time step is set to
+$\dt = \dx^2/6D$ when solving the 1D diffusion equation using forward Euler
 time integration with the standard second-order central difference
-approximation for the Laplacian.  As a final example, Zhao showed that an 
-appropriate choice of the weight factor $\theta$ in the theta-method can 
-significantly decrease the error in finite element solutions to the diffusion 
-equation~\cite{zhao_2006}.
-
-In this article, we present a simple technique for boosting the order of 
-accuracy of finite difference schemes for time dependent PDEs by optimally 
-choosing the time step and, when necessary, adding defect correction terms
-in a \emph{non-iterative} manner.  This technique is a systematic 
-generalization of the ideas underlying some of the examples mentioned above.
+approximation for the Laplacian.
+
+In this article, we present a unified analysis of the simple technique for
+boosting the \emph{order} of accuracy of finite difference schemes for time
+dependent PDEs by optimally choosing the time step and, when necessary, adding
+defect correction terms in a \emph{non-iterative} manner.  This technique is a
+systematic generalization of the ideas underlying the previously mentioned
+examples and specific examples that are scattered throughout the literature.
 Optimal time step (OTS) selection is based on the observation that a carefully
-chosen time step can simultaneously eliminate low-order terms in both the
-spatial and temporal the discretization errors.  It combines the well-known 
-approach of using knowledge of the structure of the PDE to reduce numerical 
-error with the less-common notion that an optimal choice of numerical 
-parameters can boost the effectiveness of algorithmically simple and 
-computationally low-cost numerical schemes.  For some problems, OTS alone is 
-sufficient to allow a formally low-order method to deliver high-order 
-accuracy.  However, addition of defect correction terms is generally required 
-to eliminate residual terms in the leading-order truncation error.  It is 
-important to emphasize that, unlike traditional defect/deferred correction 
-methods~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006} 
-which use defect correction terms to iteratively refine the solution at 
-each time step, we take a non-iterative defect correction (NIDC) approach
-that incorporates defect correction terms directly into the original finite 
-difference scheme.
+chosen time step can \emph{simultaneously} eliminate low-order terms in both
+the spatial and temporal the discretization errors.  At its essence, OTS
+selection combines the powerful idea of optimally choosing numerical parameters
+with the approach of using knowledge of the structure of the PDE to reduce
+numerical error.  The end result is a strategy for boosting the order of
+accuracy of algorithmically simple and computationally low-cost numerical
+schemes.
+
+While OTS selection alone is sufficient to allow a formally low-order method to
+deliver high-order accuracy for some problems, defect correction terms are
+generally required to eliminate residual terms in the leading-order truncation
+error.  However, unlike traditional defect correction 
+methods \cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}, 
+using OTS selection eliminates the need to incorporate correction terms that
+involve derivatives of higher-order than those that appear in the original PDE.
+Another significant feature of OTS selection with defect correction
+is that the defect terms are incorporated in a \emph{non-iterative} fashion.
+In contrast, traditional defect/deferred correction approaches typically
+require several refinement iterations at each time step to boost the order of
+accuracy of the numerical solution.  OTS with non-iterative defect correction
+(OTS-NIDC) also differs from the method of modified
+equations \cite{griffiths_1986} for eliminating higher-order
+corrections to the truncation error because it uses an optimal time
+step, not explicit inclusion of correction terms, to regularize the
+singularly perturbed modified equations - the correction terms that remain
+when using OTS-NIDC are all \emph{regular} perturbations to the original
+equation.
 
@@
 from the fact that it achieves high-order accuracy while being based solely on
-very simple, formally low-order finite difference schemes.  In other words, 
-OTS-NIDC allows us to extract as much accuracy as possible from an existing 
-numerical scheme with minimal additional work.  For instance, we will see how 
-OTS-NIDC makes it possible to achieve 4th-order accuracy for the diffusion 
-equation in any number of space dimensions using only a second-order stencil 
-for the Laplacian and explicit time integration.  
-Moreover, high-order accuracy is not restricted to problems 
-on simple, rectangular domains; irregular domains are straightforward to 
-handle by appropriately setting the values in ghost 
-cells~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
-OTS-NIDC is even beneficial when using finite difference schemes to solve some 
-nonlinear PDEs.  While there are certainly limitations to OTS-NIDC, we will 
-see that there are many problems where it is useful.
-
-This article is organized as follows.  In the remainder of this section,
-we compare OTS-NIDC to a few related numerical techniques.
+very simple, formally low-order finite difference schemes \emph{without}
+requiring special algorithmic steps.  In other words, OTS-NIDC allows us to
+extract as much accuracy as possible from simple numerical schemes with
+minimal additional work.  For instance, we will see how OTS-NIDC makes it
+possible to achieve 4th-order accuracy for the diffusion equation in any
+number of space dimensions using only a second-order stencil for the Laplacian
+and explicit time integration.  Moreover, high-order accuracy is not
+restricted to problems on simple, rectangular domains; irregular domains are
+straightforward to handle by appropriately setting the values in ghost
+cells \cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.  
+OTS-NIDC is even beneficial when using finite difference schemes to solve
+some nonlinear PDEs.
+
+This article is organized as follows.
+In Section~\ref{sec:theoretical_comparisons}, we compare OTS-NIDC to 
+related numerical techniques.
 In Section~\ref{sec:OTS-NIDC}, we present the theoretical foundations of 
-OTS-NIDC and show how it boosts the order of accuracy for formally low-order 
-numerical schemes.  
+OTS-NIDC and show how it boosts the order of accuracy for formally
+low-order numerical schemes.  
 In Section~\ref{sec:applications_1d}, we demonstrate the utility of 
@@
 In Section~\ref{sec:applications_multidim}, we show that 
-OTS-NIDC is useful for problems in multiple space dimensions by applying it
-the 2D linear advection equation and the 2D diffusion equation.  To illustrate 
-the ease with which high-order accuracy can be achieved for problems on 
-arbitrary domains, the 2D diffusion equation is solved on both regular 
-and irregular domains.  
-For each PDE in Sections~\ref{sec:applications_1d} and 
-\ref{sec:applications_multidim}, we provide the numerical analysis required 
-to calculate the optimal time step size, the defect correction terms, and the 
-order of accuracy.  Finally, we offer some concluding remarks in 
-Section~\ref{sec:summary}.
-
-\subsection{Comparison with Traditional Defect/Deferred Correction Methods}
+OTS-NIDC is useful for problems in multiple space dimensions by applying
+it to the 2D linear advection equation and the 2D diffusion equation.  To
+illustrate the ease with which high-order accuracy can be achieved for
+problems on arbitrary domains, the 2D diffusion equation is solved on both
+regular and irregular domains.  For each PDE in
+Sections~\ref{sec:applications_1d} and \ref{sec:applications_multidim}, we
+provide the numerical analysis required to calculate the optimal time step
+size, the defect correction terms, and the order of accuracy.  Note that in
+order to emphasize the method, we have chosen to apply it to relatively simple
+problems rather than use it to tackle ``real-life'' applications.  The method,
+however, is not restricted to these simple examples.  Finally, 
+Section~\ref{sec:conclusions} offers some concluding remarks.
+
+\section{Theoretical Comparisons with Related Numerical Techniques
+         \label{sec:theoretical_comparisons}}
+
+\subsection{Comparison with Defect/Deferred Correction Methods}
 Defect/deferred correction is a well-known strategy for boosting the order of 
@@
 takes one of two forms: 
-(1) iterative refinement of the numerical solution~\cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
+(1) iterative refinement of the numerical
+solution \cite{pereyra_1968,stetter_1978,gustafsson_2002,kress_2002,kress_2006}
 or 
@@
 operator treating time derivatives as source 
-terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.
+terms \cite{spotz_2001,ito_2005,heidenreich_2007}.
 
@@
 calculated from lower accuracy iterates.  When used to improve the 
-temporal order of accuracy, the iterative approach can help to ensure 
-stability of the time integration scheme~\cite{kress_2006}.  Although there 
+temporal order of accuracy, the iterative approach helps to ensure 
+stability of the time integration scheme \cite{kress_2006}.  Although there 
 are instances where iterative defect correction has been used to eliminate 
-both spatial and temporal errors~\cite{gustafsson_2002}, it is usually used to 
-improve only one of these errors~\cite{pereyra_1968,kress_2002,kress_2006}.
-Note that unlike iterated defect correction, OTS-NIDC requires no iteration 
-during each time step because it computes correction terms using only the 
-solution at the previous time step(s) (as opposed to requiring an estimate 
-of the solution at the current time).
+both spatial and temporal errors \cite{gustafsson_2002}, it is usually used to 
+improve only one of these errors \cite{pereyra_1968,kress_2002,kress_2006}.
+Unlike iterated defect correction, OTS-NIDC requires no iteration during each
+time step because it computes correction terms using only the solution at the
+previous time step(s) (as opposed to requiring an estimate of the solution at
+the current time).
 
@@
 derived treating time derivatives as source 
-terms~\cite{spotz_2001,ito_2005,heidenreich_2007}.  
+terms \cite{spotz_2001,ito_2005,heidenreich_2007}.  
 After deriving the high-order spatial discretization, any time differencing 
@@
 explicit time integration schemes.  Second, OTS-NIDC inherently controls both 
-spatial and temporal numerical errors at the same time.  Finally, OTS-NIDC 
-does not rely solely on the addition of finite differences of solution values 
-to cancel out discretization errors.  Instead, it takes advantage of 
-relationships between spatial derivatives, temporal derivatives and source 
-terms to eliminate the need for finite differences approximations of 
-high-order derivatives of the solution.  Compared with iterated defect 
-correction, OTS-NIDC makes it easier to avoid instabilities that arise from 
-the presence of finite difference discretizations of high-order temporal 
-derivatives and may eliminate some of the correction terms.  Comparing to the
-high-order compact scheme approach, the optimization of numerical parameters 
-in OTS-NIDC means that mixed partial derivatives do not need to appear in the 
-fully discrete equations.
+spatial and temporal numerical errors at the same time.  Finally, taking
+advantage of relationships between spatial derivatives, temporal derivatives
+and source terms, OTS-NIDC avoids the need to explicitly incorporate finite
+differences of high-order derivatives of intermediate solutions in order to
+eliminate discretization errors. 
+Compared with iterated defect correction, OTS-NIDC makes it easier to avoid
+instabilities that arise from the presence of finite difference discretizations
+of high-order temporal derivatives by automatically eliminating the most
+problematic correction terms.  Comparing to the high-order compact scheme
+approach, the use of an optimal time step in OTS-NIDC means that mixed 
+spatio-temporal partial derivatives do not need to appear in the fully
+discrete equations.
+
+\subsection{Comparison with the Modified Equation Method}
+Both defect and deferred correction can be viewed as special cases of the
+method of modified equations \cite{griffiths_1986}.  When using any modified
+equation method, the basic approach is to derive a modified version of the 
+original PDE through the introduction of the leading-order terms of the
+discretization error into the PDE.  Unfortunately, the modified PDE is a
+singular perturbation to the original PDE due to the presence of the
+discretization parameters in the correction terms.  Because the the singular
+perturbation can lead to numerical instabilities, finding a way regularize
+the modified equation is an important step for developing any numerical method
+based on the modified equations method.
+
+Whereas deferred correction methods deal with the singular perturbation by
+using an iterative procedure to avoid numerical instabilities, the traditional
+method of modified equations \cite{griffiths_1986} handles the
+singularity by ``pushing'' the singularity to a higher-order term in the
+discretization error.  The basic procedure involves substituting derivatives
+that appear in the singular term with expressions obtained by taking
+derivatives of the original PDE.  This procedure yields a new modified
+equation where high-order derivatives of the solution are replaced by
+derivatives of source terms and the new singular term is of sufficiently
+high-order in the discretization parameter that it can be neglected in the
+construction of the numerical scheme.
+
+While OTS-NIDC, like defect/deferred correction, falls within the general
+framework of the method of modified equations, it is an important special
+case because regularization of the singular perturbation in the modified
+equations is \emph{automatic}.  Regularization results from selecting the time
+step so that the spatial and temporal singular perturbation terms exactly
+cancel each other rather than through an iterative refinement procedure.
+By constraining the relationship between the time step and grid spacing, the
+modified equation becomes a regular perturbation to the original PDE.  In the
+special case when OTS-NIDC does not require defect correction, the modified
+equation is actually exactly the same as the original equation -- no regular
+perturbation terms remain after the singular perturbation terms have been
+eliminated.
+
+There are several advantages of OTS-NIDC over traditional modified
+equation methods.  First, because no finite difference approximations of
+higher-order derivatives are required, numerical schemes based on OTS-NIDC
+do not require additional boundary conditions.  Second, derivation
+of the optimal time step and correction terms for OTS-NIDC requires a comparable
+amount of analysis as the traditional modified equation method.  In both cases,
+we need explicit knowledge of the leading-order terms in the discretization
+error.  Finally, because the leading-order terms in the discretization error
+are eliminated implicitly rather than explicitly as additional terms in the 
+numerical scheme, the computational performance of OTS-NIDC is often
+superior to the performance of traditional modified equation methods.
 
 \subsection{Comparison with Adaptive Time Stepping Methods}
-OTS selection is distinct and separate from traditional adaptive time stepping 
-techniques~\cite{iserles_book,shampine_2005} that are used in the 
-context of the method of lines~\cite{iserles_book,gko_book}.  
+OTS-NIDC is distinct and separate from traditional adaptive time stepping 
+techniques \cite{iserles_book,shampine_2005} that are used in the 
+context of the method of lines \cite{iserles_book,gko_book}.  
 While both methods reduce numerical error by controlling the time step, the 
@@
 
-\section{\label{sec:OTS-NIDC}
-         Theoretical Foundations of OTS Selection and NIDC}
-OTS-NIDC is not by itself a method for constructing finite difference 
-schemes.  Rather, it is a way to enhance to the performance of existing 
-finite difference schemes by carefully choosing the time step and adding
-defect correction terms to eliminate low-order numerical errors.  
-The two key ideas underlying OTS-NIDC are 
+\section{\label{sec:OTS-NIDC} Theoretical Foundations of OTS-NIDC}
+Technically, OTS-NIDC is not by itself a method for constructing finite
+difference schemes.  Rather, it is a way to enhance to the performance of
+existing finite difference schemes by carefully choosing the time step to
+eliminate leading-order terms in the discretization error.  The two key ideas
+underlying OTS-NIDC are 
 (1) a judicious choice of time step can be used to eliminate the leading-order 
@@
 in the discretization error. 
-In this section, we present and analyze the technique of OTS-NIDC for finite 
-difference schemes constructed to solve scalar evolution equations of the form 
+In this section, we present and analyze the technique of OTS-NIDC for
+finite difference schemes constructed to solve scalar evolution equations of
+the form 
 \beq
@@
 To compute the optimal time step, we first analyze the discretization errors 
-(both spatial and temporal) for the finite difference approximation, 
-explicitly keeping track of the leading-order terms.  In contrast to 
-conventional error analysis, we do not sweep all of the errors under big-$O$
-notation.  Rather, we take advantage of the fact that derivation of the 
-leading-order terms in the error is straightforward for finite difference
-schemes through the use of Taylor series expansions.  Because each term in 
-the error is proportional to a partial derivative of the solution, we can 
-use the PDE to relate the leading-order terms to each other.  
-For many finite difference schemes, this careful analysis yields a direct 
-relationship between the leading-order spatial and temporal errors that allows 
-them to be combined as a single term of the form:
+for the finite difference approximation, explicitly keeping track of the
+leading-order terms.  Because each term in the error is proportional to a
+partial derivative of the solution, we can use the PDE to relate the
+leading-order terms to each other.  For many finite difference schemes, this
+careful analysis yields a direct relationship between the leading-order spatial
+and temporal errors that allows them to be combined as a single term of the
+form:
 \beq
@@
 where $L$ is a differential operator and $P$ is a simple function of the 
-numerical parameters $\dx$ and $\dt$.  By selecting the numerical parameters 
+numerical parameters $\dx$ and $\dt$.  By selecting the time step
 so that $P = 0$, we can completely eliminate the leading-order term in the 
-discretization error.  Because time dependent PDEs are often solved by 
-stepping in time, it is natural to choose the time step to be a function of 
-the grid spacing.  The time step selected in this way is the
+discretization error.  The time step selected in this way is the
 \emph{optimal time step} for the finite difference scheme.  Using the optimal
-time step for time stepping yields a numerical solution with an order 
-of accuracy that is \emph{higher} than the formal order of the original finite 
-difference scheme.
+time step for time stepping yields a numerical solution with an
+\emph{effective} order of accuracy that is higher than the formal order of the
+original finite difference scheme.
 
 An important class of finite difference schemes for time dependent PDEs are 
-those based on the method of lines.   When the method of lines approach is 
+those based on the method of lines.   When the method of 
+lines \cite{iserles_book,gko_book,cash_1996,psihoyios_2007}.  
+approach is 
 used for (\ref{eq:model_PDE}), $P$ often takes a particularly simple 
@@
 \eeq
-Notice that the optimal time step has a similar functional dependence on the 
-grid spacing as a time step that satisfies a stability constraint.
 
@@
 in this situation, it is usually possible to find a closely related finite 
-difference scheme which \emph{does} have a positive optimal time step.
-
-\subsubsection*{\label{sec:error_analysis} 
-            Error Analysis}
+difference scheme that has a positive optimal time step.
+
+\subsubsection*{Error Analysis \label{sec:error_analysis}}
 The form of the local error in 
@@
 argument that the global error is equal to the local error divided by 
-$\dt$~\cite{gko_book}:
+$\dt$ \cite{gko_book}:
 \beq
@@
 one parameter, $\dx$, that controls the numerical accuracy of the finite 
-difference scheme (see Appendix~\ref{app:formal_vs_practical_accuracy} for 
-further discussion about this important point).  Since $\dto = O(\dx^{r/s})$, 
-the global error (\ref{eq:global_error_ots}) reduces to
+difference scheme.  Since $\dto = O(\dx^{r/s})$, the global error
+(\ref{eq:global_error_ots}) reduces to
 \beq
@@
 difference schemes when the time step is set to $\dto$.  For explicit time 
-integration schemes, OTS selection is ineffective when the optimal time step 
+integration schemes, OTS-NIDC is ineffective when the optimal time step 
 is greater than the smallest stable time step.  In this situation, it is 
@@
 error (see Section~\ref{sec:diffusion_eqn_2d}), the optimal time step becomes 
-unstable when the dimension of space is greater than 3.  To make OTS selection 
+unstable when the dimension of space is greater than 3.  To make OTS-NIDC 
 useful for diffusion problems in 4 and more space dimensions, we can switch 
 to the DuFort-Frankel scheme, which is explicit and unconditionally 
-stable~\cite{gko_book}.
-
-\subsubsection*{\label{sec:computational_performance} 
-                Computational Performance}
+stable \cite{gko_book}.
+
+\subsubsection*{Computational Performance \label{sec:computational_performance}}
 For many problems, the optimal time step is on the order of the maximum 
@@
 As Table~\ref{tab:comp_perf_vs_err} shows, this boost in accuracy leads to a
-scheme that is computationally cheaper than traditional methods that use 
-the same low-order finite difference stencils.  The performance gain for 
-problems in higher spatial dimensions is even more impressive.  As we can see 
-in Table~\ref{tab:comp_perf_vs_dim}, the gap between the performance of a
-simple forward Euler scheme with the optimal time step and the Crank-Nicholson
-method grows with the spatial dimension of the problem.
+scheme that is asymptotically computationally cheaper than traditional implicit
+schemes that use the same low-order finite difference stencils.  The
+performance gain for problems in higher spatial dimensions is even more
+impressive.  As we can see in Table~\ref{tab:comp_perf_vs_dim}, the gap
+between the performance of a simple forward Euler scheme with the optimal time
+step and the Crank-Nicholson method grows with the spatial dimension of the
+problem.
 
@@
 A comparison of computational costs in higher spatial dimensions continues 
-along the same trends but requires replacement of the forward Euler time 
-integration with an alternative for which the optimal time step is stable 
-(\eg the DuFort-Frankel scheme). 
+along the same trends but requires replacement of the conventional forward Euler
+time integration scheme with an alternative for which the optimal time step is
+stable (\eg the DuFort-Frankel scheme). 
 }
@@
 the same improvement in the order of accuracy.  In this section, we discuss 
-these correction terms in the context of general constant coefficient, linear 
-PDEs and semilinear PDEs.
+OTS-NIDC in the context of general constant coefficient, linear PDEs and
+semilinear PDEs.
 
@@
 it is sufficient to impose boundary conditions by using boundary values 
-that have the same order of accuracy as the solution.  More effort is 
+that have the same order of accuracy as the numerical scheme.  More effort is 
 required to ensure that Neumann boundary conditions do not reduce the order 
@@
 boundary conditions is beyond the scope of this article.  We refer interested 
-readers to the thorough analysis presented in~\cite{gko_book}.  To avoid 
+readers to the thorough analysis presented in \cite{gko_book}.  To avoid 
 complicating the discussion in Sections~\ref{sec:applications_1d} 
@@
 a `ghost cell' or `ghost value' method for imposing boundary 
-conditions~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book},
+conditions \cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book},
 optimal time step selection can even be directly applied to problems on 
@@
 
-\subsection{Limitations of Optimal Time Step Selection}
-As with any numerical method, optimal time step selection has its limitations.
-The most serious limitation is the requirement that the coefficient on the 
-leading-order spatial derivative in the PDE be constant in time and space.  
-Without this assumption, the first term in the leading-order discretization 
-error would be of the form $(L u) (\alpha(x,t) \dx^r - \beta(x,t) \dt^s) \dt$,
-where $\alpha$ and $\beta$ are now functions of space and time.  This
-small change in the form of the error makes it impossible to derive a single 
-optimal time step.  Instead, each grid point would possess its own individual 
-optimal time step.  For a similar reason, optimal time step selection also 
-fails for quasilinear and fully nonlinear PDEs.  The issue for these types of 
-PDEs is that the coefficients in $P(\dx, \dt)$ become functions of the 
-solution $u$.
-
-Another limitation of optimal time step selection is that it does not,
-in its current incarnation, apply to finite difference schemes for general
-systems of PDEs.  The main issue here is that each PDE might have its 
-own optimal time step.  OTS is, however, certainly applicable for problems 
-where the optimal time step is the same for all of the individual PDEs in the 
-system.
-
-While these limitations certainly place restrictions on the range of problems 
-that we can apply OTS-NIDC to, there are many relevant and important PDEs 
-arising in practice that satisfy the requirements for OTS-NIDC.
-
 \section{\label{sec:applications_1d} 
@@
 finite difference schemes that solve several classical linear and semilinear 
-PDEs.  To emphasize the importance of including numerical parameter 
+PDEs.  To emphasize the value of including numerical parameter 
 optimization as part of the design of any finite difference scheme, we 
@@
 have analytical solutions that are useful to compare numerical solutions 
-against.  However, the utility of OTS-NIDC is by no means limited to 
-these simple PDEs.
+against.  However, the utility of OTS-NIDC is by no means limited to these
+simple PDEs.
 
@@
 is solved using a first-order upwind scheme with forward Euler time
-integration, it is well-known that choosing $\Delta t = \Delta x / |A|$ 
+integration, it is well-known that choosing $\dt = \dx / |A|$ 
 completely eliminates all numerical error in the solution (except for 
 errors in the initial conditions).  This result is known as the unit CFL 
-condition~\cite{leveque_book_2002} and is usually proved by examining how 
+condition \cite{leveque_book_2002} and is usually proved by examining how 
 characteristic lines of the advection equation pass through grid points.  
@@
 \eeq
-where we have taken $A$ to be positive for convenience.  In general, we only 
-need to keep track of the leading-order terms in the discretization error.  
-However, for the advection equation, it is valuable to retain all of them.  
-Using Taylor series expansions of the true solution of the PDE, we find that 
-\bea
-  \tu^{n+1}_j &=& \tu^{n}_j 
-  + \sum_{k=1}^\infty \frac{\dt^k}{k!} 
-       \frac{\partial^k \tu^n}{\pt^k} 
-  = \tu^{n}_j + \sum_{k=1}^\infty \frac{\left( -A \dt \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k},
-  \label{eq:advection_eqn_1d_time_err} 
-  \\
-  \frac{\tu^{n}_j - \tu^{n}_{j-1}}{\dx} &=& 
-  \frac{\partial \tu^n}{\px} 
-  - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k}.
-  \label{eq:advection_eqn_1d_space_err}
-\eea
-where $\tu$ denotes the true solution of the PDE, all derivatives are 
-evaluated at $x_j$, and we have used the PDE to replace all of the time 
-derivatives with spatial derivatives in (\ref{eq:advection_eqn_1d_time_err}).
-For convenience, we will continue to suppress the subscripts on spatial
-derivatives throughout the article; the location where they should be 
-evaluated will be apparent from the context.
-
-Combining~(\ref{eq:advection_eqn_1d_time_err}) 
-and~(\ref{eq:advection_eqn_1d_space_err}) with the finite difference 
-scheme~(\ref{eq:advection_eqn_1d_FD_scheme}) yields the evolution
-equation for the error $e \equiv u - \tu$:
+where we have taken $A$ to be positive for convenience.  When applying OTS 
+selection, we generally only need to keep track of the leading-order terms in
+the discretization error.  However, for the advection equation, it is valuable
+to retain all of them.  Using Taylor series expansions of the true solution of
+the PDE, it is straightforward to show that the evolution equation for the error
+$e \equiv u - \tu$ is
 \beq
@@
 \eeq
-Therefore, choosing the time step to eliminate the leading-order term 
-of the discretization error leads to the unit CFL condition and 
-\emph{complete elimination} of the local truncation error at each time step.
-It is interesting to note that OTS selection applied to the Lax-Wendroff 
-scheme~\cite{leveque_book_2002,leveque_book_1992} yields the same optimal 
-time step and total elimination of the truncation error.
+where $\tu$ denotes the true solution of the PDE, all derivatives are 
+evaluated at $(t_n, x_j)$, and we have used the PDE to replace all of time 
+derivatives with spatial derivatives.
+
+From (\ref{eq:advection_eqn_1d_err_eqn}), we see that choosing the
+time step to eliminate the leading-order term of the discretization error
+leads to the unit CFL condition and \emph{complete elimination} of the local
+truncation error at each time step.  It is interesting to note that OTS
+selection applied to the Lax-Wendroff 
+scheme \cite{leveque_book_2002,leveque_book_1992} yields the same optimal 
+time step and the same total elimination of the truncation error.
 
@@
 first-order system of equations.  However, as Kreiss, Petersson, and Ystr\"om 
-showed in~\cite{kreiss2002}, direct discretization of the second-order wave 
+showed in \cite{kreiss2002}, direct discretization of the second-order wave 
 equation is a viable option and leads to the following finite difference 
@@
 \eea
-The stability constraint for this scheme of the form $\dt = K \dx$.
+The stability constraint for this scheme of the form $\dt = K \dx$, where 
+$K$ is a constant.
 
@@
 begin by deriving the truncation error for the scheme.  Employing Taylor 
-series expansions, it is straightforward to show that the true solution 
-satisfies
-\bea
-  \frac{\tu^{n+1}_i - 2 \tu^n_i + \tu^{n-1}_i}{\dt^2}
-    - \frac{\dt^2}{12} \frac{\partial^4 \tu}{\pt^4}
-  &=& c^2 \left( \frac{\tu^{n}_{i+1} - 2 \tu^n_i + \tu^n_{i-1}}{\dx^2}
-  -\frac{\dx^2}{12} \frac{\partial^4 \tu}{\px^4} \right) + f
-  \nonumber \\
-  &+& O(\dt^4) + O(\dx^4) 
-\eea
-Combining this result with the observation that
-$\tu_{tttt} = c^4 \tu_{xxxx} + c^2 f_{xx} + f_{tt}$ and rearranging a bit, 
-we find that the error equation for (\ref{eq:wave_eqn_KPY}) is given by
+series expansions and the observation that the true solution satisfies
+$u_{tttt} = c^4 u_{xxxx} + c^2 f_{xx} + f_{tt}$, we find that the error
+equation for (\ref{eq:wave_eqn_KPY}) is given by
 \bea
@@
 \begin{center}
-\scalebox{0.35}{\includegraphics{figures/wave_eqn_1d_src_error_vs_N}}
+\subfloat{\includegraphics[width=0.40\textwidth]{figures/wave_eqn_1d_src_error_vs_N}}
 \caption{$L^\infty$ error as a function of number of grid points for the
-  KPY discretization of the second-order wave equation on with OTS-NIDC 
+  KPY discretization of the second-order wave equation with OTS-NIDC 
   (blue circles) and without OTS-NIDC (red squares).
@@
 a $(p+1)$-order accurate first time introduces an $O(\dx^p)$ error into the 
-solution at all future times.
-Thus, we must ensure that the error introduced by the first time step is
-at least one order of accuracy higher than desired accuracy for the numerical 
-solution.  See Appendix~\ref{appendix:KPY_analysis} for a derivation of this
-condition.
+solution at all future times.  Thus, we must ensure that the error introduced
+by the first time step is at least one order of accuracy higher than desired
+accuracy for the numerical solution\footnote{This requirement follows from a
+straightforward analysis of the second-order difference equation for the
+normal modes of the error.}.
 
@@
 overall (\ie inclusion of the temporal order of accuracy is irrelevant).
-Using Taylor series expansions and the PDE $u_t = D u_{xx} + f$,
-we see that the true solution satisfies
-\bea
-  \tu^{n+1}_j = \tu^{n}_j 
-  &+& \dt \left( D \frac{\partial^2 \tu^n}{\px^2} + f^n_j \right)
-  \nonumber \\
-  &+& \frac{(D\dt)^2}{2} \frac{\partial^4 \tu^n}{\px^4} 
-  + \frac{\dt^2}{2} \left( D\frac{\partial^2 f}{\px^2}
-                         + \frac{\partial f}{\pt} \right)
-  + O \left( \dt^3 \right)
-  \label{eq:diffusion_eqn_1d_time_err} 
-\eea
-and that the central difference approximation for the Laplacian satisfies
-\beq
-  \frac{\tu^{n}_{j+1} -2 \tu^{n}_j + \tu^{n}_{j-1}}{\dx^2}  =
-  \frac{\partial^2 \tu^n}{\px^2} 
-  + \frac{\dx^2}{12} \frac{\partial^4 \tu^n}{\px^4} 
-  + O(\dx^4).
-  \label{eq:diffusion_eqn_1d_space_err}
-\eeq
-Therefore, the truncation error for (\ref{eq:diffusion_eqn_1d_FD_scheme})
-is given by
+Using Taylor series expansions and the PDE, $u_t = D u_{xx} + f$,
+we find that the truncation error is given by
 \beq
@@
 \begin{center}
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for 
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/diffusion_eqn_1d_src_error_vs_N} \label{fig:diffusion_eqn_1d_src_error}}
+\ \ \ \ \ \ \ 
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/diffusion_eqn_1d_src_DF_error_vs_N} \label{fig:diffusion_eqn_1d_dufort_frankel_error}}
+\caption{(a) $L^\infty$ error as a function of number of grid points for 
 two finite difference schemes that solve the diffusion equation: 
@@
 (triangles).  
+(b) $L^\infty$ error in the numerical solution of the 1D diffusion 
+equation as a function of number of grid points for the DuFort-Frankel scheme 
+with (circles) and without (squares) OTS-NIDC.
 }
-\label{fig:diffusion_eqn_1d_src_error}
 \end{center}
@@
 
-We begin by writing the DuFort-Frankel scheme in form that is easily 
+We begin by writing the DuFort-Frankel scheme in a form that is easily 
 generalizable to higher space dimensions:
@@
 unconditionally stable and has a global error that is formally 
-$O(\dt^2/\dx^2) + O(\dt^2) + O(\dx^2) + O(\dt^4/\dx^2)$~\cite{gko_book}.  
+$O(\dt^2/\dx^2) + O(\dt^2) + O(\dx^2) + O(\dt^4/\dx^2)$ \cite{gko_book}.  
 This expression for the formal error is ``minimized'' when $\dt = O(\dx^2)$, 
@@
 
-Using Taylor series expansions centered at $(x_j, t_n)$, 
-(\ref{eq:diffusion_eqn_1d_space_err}) and the PDE to derive the local 
-truncation error for the DuFort-Frankel scheme, we find that 
+Using Taylor series expansions centered at $(t_n, x_j)$ and the PDE, we derive
+the local truncation error for the DuFort-Frankel scheme:
 \bea
-  \tau &=& 2D \dt \frac{\partial^4 \tu}{\px^4}
+  & & 2D \dt \frac{\partial^4 \tu}{\px^4}
   \left(\frac{\dx^2}{12} - \frac{D^2 \dt^2}{\dx^2} \right)
@@
 
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.35}{\includegraphics{figures/diffusion_eqn_1d_src_DF_error_vs_N}} 
-\caption{$L^\infty$ error in the numerical solution of the 1D diffusion 
-equation as a function of number of grid points for the DuFort-Frankel scheme 
-with (circles) and without (squares) OTS-NIDC.
-}
-\label{fig:diffusion_eqn_1d_dufort_frankel_error} 
-\end{center}
-\end{figure}
-
 \subsection{Viscous Burgers Equation}
 We now demonstrate the use of OTS-NIDC in the context of a well-studied
-semilinear PDE -- the viscous Burgers equation~\cite{whitham_book}:
+semilinear PDE -- the viscous Burgers equation \cite{whitham_book}:
 \beq
@@
 guide\footnote{The stability constraint for the advection-diffusion equation
-is $\dt < \min \{2D/A^2, \dx^2/2D \}$~\cite{chan_1984}.}, 
+is $\dt < \min \{2D/A^2, \dx^2/2D \}$ \cite{chan_1984}.}, 
 we expect a stability constraint of the form $\dt \le \dx^2/2 \nu$ in the 
@@
 \begin{center}
-\scalebox{0.35}{\includegraphics{figures/burgers_eqn_1d_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points for
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/burgers_eqn_1d_error_vs_N} \label{fig:burgers_1d_error}}
+\ \ \ \ \ \ \ 
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/burgers_eqn_1d_comp_time} \label{fig:burgers_1d_perf}}
+\caption{(a) $L^\infty$ error as a function of number of grid points for
 two finite difference schemes that solve the viscous Burgers equation:
@@
 correction term (squares).
-The viscous Burgers equation is solved with initial conditions taken to be 
-$u(x,0) = 1 + \gamma \sqrt{\frac{\nu}{\pi}} \exp\left( -(x-1)^2/4 \nu \right) 
-            / \left[ 1 + \frac{\gamma}{2} \ 
-                         \mathtt{erfc}\left((x-1)/\sqrt{4\nu} \right) 
-    \right]$,
-where $\ln(1+\gamma)$ is the effective Reynolds number (set equal to 10) and 
-the wave speed is set equal to $1$.  The boundary conditions are imposed 
-exactly by using
-the analytical solution
-$u(x,t) = 1 + \gamma \sqrt{\frac{\nu}{\pi T}} 
-              \exp\left( -(x-T)^2/4 \nu T \right) 
-            / \left[ 1 + \frac{\gamma}{2} \ 
-                         \mathtt{erfc}\left((x-T)/\sqrt{4\nu T} \right) 
-    \right]$,
-where $T = t+1$.
+(b) Computation time as a function of $L^\infty$ error.  These results were
+obtained using MATLAB implementations of the finite difference schemes run on
+a dual-core 2.4 GHz MacBook Pro.
 }
-\label{fig:burgers_1d_error}
 \end{center}
@@
 
-To analyze the discretization error for (\ref{eq:burgers_1d_FD_scheme}), 
-we combine~(\ref{eq:diffusion_eqn_1d_space_err}) from our analysis of 
-the diffusion equation with the leading-order discretization error for 
-the central difference approximation of $u_x$,
-\beq
-  \frac{\tu^{n}_{j+1} - \tu^{n}_{j-1}}{2 \dx}  =
-  \frac{\partial \tu^n}{\px} 
-  + \frac{\dx^2}{6} \frac{\partial^3 \tu^n}{\px^3} 
-  + O(\dx^4),
-  \label{eq:burgers_1d_ux_err}
-\eeq
-to obtain the local truncation error for~(\ref{eq:burgers_1d_FD_scheme}): 
+Using the usual OTS-NIDC analysis procedure, we obtain the local truncation
+error for~(\ref{eq:burgers_1d_FD_scheme}): 
 \bea
@@
 Note that the optimal time step eliminates errors introduced by 
-both the diffusion term \emph{and} the nonlinear advection term.
+both the diffusion term \emph{and} the nonlinear advection term so that 
+the highest order spatial derivative in the defect correction terms only
+is no higher than the highest order spatial derivative in the original PDE.
 
@@
 
+The asymptotic computational performance of numerical schemes for the
+1D viscous Burgers equation is the same as for the 1D diffusion equation.
+Figure~\ref{fig:burgers_1d_perf} shows that the theoretical estimates
+from Table~\ref{tab:comp_perf_vs_err} for the conventional forward Euler and
+OTS-NIDC forward Euler algorithms are in good agreement with timing data
+collected from computational experiments.  Note that the data for the OTS-NIDC
+scheme bends upwards below an $L^\infty$ error of about $10^{-5}$, which
+indicates that the $O(1/\dx)$ work per time step does not dominate the
+computational work per time step until $\dx$ is sufficiently small.
+
 The Burgers equation example illustrates an important aspect of OTS-NIDC:
@@
 \begin{center}
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N}} 
+\subfloat[]{\includegraphics[width=0.31\textwidth]{figures/4th_order_parabolic_eqn_1d_error_vs_N} \label{fig:4th_order_parabolic_eqn_1d_error}}
 \ \  
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_error_vs_N_roundoff}} 
-\caption{$L^\infty$ error as a function of number of grid points for four
+\subfloat[]{\includegraphics[width=0.31\textwidth]{figures/4th_order_parabolic_eqn_1d_comp_time} \label{fig:4th_order_parabolic_eqn_1d_perf}}
+\ \ 
+\subfloat[]{\includegraphics[width=0.31\textwidth]{figures/4th_order_parabolic_eqn_1d_error_vs_N_roundoff} \label{fig:4th_order_parabolic_eqn_1d_error_roundoff}}
+\caption{(a) $L^\infty$ error as a function of number of grid points for four
 finite difference schemes that solve the fourth-order parabolic equation
@@
 Both forward Euler schemes use a fourth-order, central finite difference 
-approximation for the bilaplacian.  The boundary and ghost cells values
-are set by taking values from the known analytical solution.
+approximation for the bilaplacian.
 Note that the errors for the forward Euler with suboptimal time step and 
@@
 other at the resolution of the figures.  
-The figure on the left highlights the impact (when $N$ is large) of round-off 
-error associated with matrix inversion on the accuracy of the solution for the 
-two Crank-Nicholson schemes.
+(b) Computation time as a function of $L^\infty$ error.  These results were
+obtained using MATLAB implementations of the finite difference schemes run on
+a dual-core 2.4 GHz MacBook Pro.  Note the impact of round-off error
+associated with matrix inversion on the accuracy of the solution for the two
+Crank-Nicholson schemes.
+(c) Illustration of impact of round-off error associated with matrix inversion
+on the accuracy of the solution for the two Crank-Nicholson schemes when
+$N$ is large.
 }
-\label{fig:4th_order_parabolic_eqn_1d_error}
 \end{center}
@@
 
-Proceeding in the usual way, we find that the true solution satisfies
-\beq
-  \tu^{n+1}_j = \tu^{n}_j 
-  + \dt \left( -\kappa \frac{\partial^4 \tu^n}{\px^4}  
-             + f^n_j
-        \right)
-  + \frac{\dt^2}{2} 
-    \left(
-      \kappa^2 \frac{\partial^8 \tu^n}{\px^8} 
-    - \kappa \frac{\partial^4 f}{\px^4} 
-    + \frac{\partial f}{\pt}
-    \right)
-  + O \left( \dt^3 \right)
-  \label{eq:4th_order_parabolic_eqn_1d_time_err}
-\eeq
-and that the central difference approximation to the bilaplacian satisfies
+Proceeding in the usual way and noting that the central difference
+approximation to the bilaplacian satisfies
 \beq
@@
   + O(\dx^6)
-  \label{eq:4th_order_parabolic_eqn_1d_space_err}.
-\eeq
-Combining these results with the finite difference 
-scheme~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}), we find that the
-local truncation error is given by
+  \label{eq:4th_order_parabolic_eqn_1d_space_err},
+\eeq
+we find that the local truncation error
+for~(\ref{eq:4th_order_parabolic_eqn_1d_FD_scheme}) is given by
 \beq
@@
 limitation of explicit schemes for time dependent PDEs with high-order spatial 
-derivatives~\cite{gko_book,greer_2006}, 
+derivatives \cite{gko_book,greer_2006}, 
 the overall error of finite difference schemes for high-order PDEs is usually 
@@
 
-Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} shows the computational 
-performance of various finite difference methods for solving 
+Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} shows the asymptotic
+computational performance of various finite difference methods for solving 
 equation~(\ref{eq:4th_order_parabolic_eqn_1d_second_time_derivative}).
@@
 
-\begin{figure}[tb]
-\begin{center}
-\scalebox{0.33}{
-  \includegraphics{figures/4th_order_parabolic_eqn_1d_comp_time}} 
-\caption{Computation time as a function of $L^\infty$ error for four
-finite difference schemes that solve the fourth-order parabolic equation
-(\ref{eq:4th_order_parabolic_eqn_1d}).  
-These results were obtained using MATLAB implementations of the 
-finite difference schemes run on a 2.4 GHz MacBook Pro.
-Note the impact of round-off error associated with matrix inversion on 
-the accuracy of the solution for the two Crank-Nicholson schemes.
-See the caption of Figure~\ref{fig:4th_order_parabolic_eqn_1d_error} for 
-descriptions of the four finite difference schemes.
-}
-\label{fig:4th_order_parabolic_eqn_1d_perf}
-\end{center}
-\end{figure}
-
 In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
@@
 Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-NIDC
-scheme should do better.  The origin of this discrepancy is the constant 
-hidden by the big-$O$ notation.  In particular, the factor of $7/120$ in 
-the optimal time step for the forward Euler scheme with a fourth-order 
-approximation of the bilaplacian leads to a relatively large hidden constant 
+scheme should do better asymptotically.  The origin of this discrepancy is the
+constant hidden by the big-$O$ notation.  In particular, the factor of $7/120$
+in the optimal time step for the forward Euler scheme with a fourth-order 
+approximation of the bilaplacian leads to a large hidden constant 
 for the compute time. 
-However, as Figures~\ref{fig:4th_order_parabolic_eqn_1d_error} and
-\ref{fig:4th_order_parabolic_eqn_1d_perf} show, there is an important 
+However, as Figures~\ref{fig:4th_order_parabolic_eqn_1d_perf} and
+\ref{fig:4th_order_parabolic_eqn_1d_error_roundoff} show, there is an important 
 limitation of the 2nd-order Crank-Nicholson scheme even if we ignore the 
@@
 schemes, we see that the OTS-NIDC forward Euler scheme more efficiently 
-utilizes memory at the cost of requiring more computation time.  This tradeoff 
+utilizes memory at the cost of requiring more computation time.  This trade off 
 between memory and computation only exists in one space dimension.  In 
@@
 one space dimension and multiple space dimensions is the presence of 
-cross-terms in the discretization error.  When formally analyzing the accuracy
-of a finite difference scheme, it is usually safe to ignore the details of 
-these terms because they can be absorbed within big-$O$ notation.
-However, precise knowledge of the cross-terms is critical when attempting to 
-use OTS-NIDC.  Without this knowledge, it is impossible to 
-ensure cancellation of leading-order errors.
-
-Recall that OTS-NIDC requires the leading-order temporal errors to be directly 
-related to the leading-order spatial errors via the PDE.  This requirement can 
-be satisfied in one of two ways:
+cross-terms in the discretization error.  Recall that OTS-NIDC requires the
+leading-order temporal errors to be directly related to the leading-order
+spatial errors via the PDE.  This requirement can be satisfied in one of two
+ways:
 \begin{itemize}
@@
 of the truncation error requires optimal selection of the grid spacing ratio 
-as well as the time step.  That is, the more general idea of numerical 
-parameter optimization is needed in order to boost the accuracy of the scheme.
+as well as the time step.
 
@@
 the terms in the discretization error.  Using Taylor series expansions and
-the PDE (\ref{eq:advection_eqn_2d}), we find that the true solution $\tu$ 
+the PDE (\ref{eq:advection_eqn_2d}), we find that the truncation error 
 satisfies
 \bea
-  \tu^{n+1}_{i,j} &=& \tu^{n}_{i,j }
-    + \sum_{k=1}^\infty \frac{\left( -\dt \right)^k}{k!} 
-         \left( A_x \partial_x
-              + A_y \partial_y \right)^k \tu^{n}.
-  \label{eq:advection_eqn_2d_time_err}
-  \\
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i-1,j}}{\dx} &=& 
-  \frac{\partial \tu}{\px} 
-  - \frac{1}{\dx} \sum_{k=2}^\infty \frac{\left( -\dx \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\px^k} 
-  \label{eq:advection_eqn_2d_space_err_x}
-  \\
-  \frac{\tu^{n}_{i,j} - \tu^{n}_{i,j-1}}{\dy} &=&
-  \frac{\partial \tu}{\py} 
-  - \frac{1}{\dy} \sum_{k=2}^\infty \frac{\left( -\dy \right)^k}{k!} 
-       \frac{\partial^k \tu^n}{\py^k} 
-  \label{eq:advection_eqn_2d_space_err_y}.
-\eea
-Therefore, the local truncation error is given by:
-\bea
-  \tau^{n}_{i,j} &=&
+  & &
       \frac{A_x \dt}{2} \frac{\partial^2 \tu}{\px^2}
@@
 \bea
-  \tau^{n}_{i,j} &=&
+  & &
       \frac{A_x \dt}{2} \frac{\partial^2 \tu}{\px^2}
@@
 step.  Figure~\ref{fig:advection_eqn_2d_soln} demonstrates the amazing accuracy 
-of this scheme over the first-order forward Euler 
-scheme~(\ref{eq:advection_eqn_2d_FD_scheme}) without any modifications.
+of the defect corrected scheme over the first-order forward Euler 
+scheme~(\ref{eq:advection_eqn_2d_FD_scheme}) without defect correction.
 Notice how the sharp corners of the initial conditions are perfectly 
@@
 \begin{center}
-\scalebox{0.35}{\includegraphics{figures/adv_eqn_2d_FE_OTS_soln}} 
-\scalebox{0.35}{\includegraphics{figures/adv_eqn_2d_FE_soln}} 
+\subfloat{\includegraphics[width=0.40\textwidth]{figures/adv_eqn_2d_FE_OTS_soln}} 
+\ \ \ \ \ \ \ 
+\subfloat{\includegraphics[width=0.40\textwidth]{figures/adv_eqn_2d_FE_soln}} 
 \caption{Comparison of numerical solutions for the 2D advection
@@
 stepping with suboptimal time step (right).  The latter scheme uses a 
-time step $\Delta t = \Delta x / 2 (|A_x|+|A_y|)$.  
-Both figures show the numerical solution at $t = 3$ computed on the domain 
-$-10 < x,y < 10$ using 100 grid points in each direction.
-The advection velocity $(A_x, A_y)$ is set to $(-1, -2)$.  The initial 
-conditions are taken to be a 2D square function centered at the
+time step $\dt = \dx / 2 (|A_x|+|A_y|)$.  
+Both figures show the numerical solution at the same time when the 
+initial conditions are taken to be a 2D square function centered at the
 origin: $u(x,y,0) = 1$ if $|x| + |y| \le 2$ and 
 $u(x,y,0) = 0$ if $|x| + |y| > 2$. 
-For the time interval considered, $u$ is identically zero 
-on the boundary.
 }
@@
 It is interesting to push the analysis of the optimal finite difference 
-scheme a little further and examine the complete finite difference scheme
-(including defect correction term):
+scheme a little further and examine the defect-corrected finite difference
+scheme:
 \bea
@@
 \beq
-  u^{n+1}_j = u^{n}_j 
-  + \dt 
-    \left( D \left[\frac{u^{n}_{i,j+1} + u^{n}_{i,j-1}
-             +u^{n}_{i+1,j} + u^{n}_{i-1,j} - 4u^{n}_{i,j}}{\dx^2} \right]
-         + f_{i,j}^n
-    \right).
+  u^{n+1}_{i,j} = u^{n}_{i,j}
+  + \dt \left( D L^{5pt} u + f_{i,j}^n \right)
   \label{eq:diffusion_eqn_2d_5pt_scheme}
@@
 scheme using the nine-point stencil for the 
-Laplacian~\cite{iserles_book,patra_2005}, which possesses a leading-order 
+Laplacian \cite{iserles_book,patra_2005}, which possesses a leading-order 
 error directly related to the spatial derivative operator in the PDE 
@@
 stencil plus $\dx^2/6$ times a second-order, central difference 
-approximation to $u_{xxyy}$.  Therefore, we can obtain the leading-order 
-error in the nine-point stencil by combining the leading-order error for the 
-five-point stencil with the extra contribution of 
-$\dx^2 \tu_{xxyy}/6$
-\bea
-L^{9pt} \tu &=& \nabla^2 \tu 
-              + \frac{\dx^2}{12} 
-                \left(\tu_{xxxx} + 2 \tu_{xxyy} + \tu_{yyyy}\right)
-              + O(\dx^4)
-          \nonumber \\
-          &=& \nabla^2 \tu + \frac{\dx^2}{12} \nabla^4 \tu + O(\dx^4).
-  \label{eq:laplacian_2d_9pt_stencil_error}
-\eea
-Like (\ref{eq:diffusion_eqn_2d_5pt_scheme}), the forward Euler scheme that 
-uses the nine-point discretization of the Laplacian 
-\beq
-  u^{n+1}_{i,j} = u^{n}_{i,j}
-  + \dt \left( D L^{9pt} u + f_{i,j}^n \right)
-  \label{eq:diffusion_eqn_2d_9pt_scheme}
-\eeq
-is $O(\dx^2)$ accurate overall because it has a stability constraint 
-$\dt \le 3\dx^2/8D = O(\dx^2)$.
-It is interesting to note that use of a nine-point stencil for the Laplacian 
+approximation to $u_{xxyy}$.  Therefore, the error in the nine-point stencil
+is $\frac{\dx^2}{12} \nabla^4 u + O(\dx^4)$.  The forward Euler scheme that 
+uses the nine-point discretization of the Laplacian is also $O(\dx^2)$ accurate
+overall because it has a stability constraint $\dt \le 3\dx^2/8D = O(\dx^2)$.
+It is worth noting that use of a nine-point stencil for the Laplacian 
 naturally arises when directly applying OTS-NIDC to the five-point scheme if 
@@
 
-To apply OTS-NIDC to (\ref{eq:diffusion_eqn_2d_9pt_scheme}), we proceed in the 
-usual way by first analyzing the leading-order terms of the local truncation
-error for the finite difference scheme:
+Analyzing the finite difference scheme based on the nine-point stencil for the
+Laplacian, we find that the leading-order terms of the local truncation error
+is
 \bea
@@
 \eea
-Therefore, the optimal time step is given by $\dto = \dx^2/6D$ and the 
-defect correction term is $\dt^2 \left( D\nabla^2 f + f_t \right)$.
-This choice of time step and correction term yields a scheme that is 
-fourth-order accurate.  It is interesting to note that the optimal time step
-for the forward Euler scheme for the diffusion equation is \emph{independent} 
-of the number of spatial dimensions.
+Therefore, the optimal time step and defect correction terms are 
+$\dto = \dx^2/6D$ and $\dt^2 \left( D\nabla^2 f + f_t \right)$, respectively.
+The resulting OTS-NIDC scheme that fourth-order accurate.  It is interesting
+to note that the optimal time step for the forward Euler scheme for the
+diffusion equation is \emph{independent} of the number of spatial dimensions.
 
@@
 \begin{center}
-\scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_error_vs_N}} 
-\ \ 
-\scalebox{0.33}{\includegraphics{figures/diffusion_eqn_2d_src_comp_time}} 
-\caption{$L^\infty$ error as a function of number of grid points in
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/diffusion_eqn_2d_src_error_vs_N} \label{fig:diffusion_eqn_2d_src_error}}
+\ \ \ \ \ \ \ 
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/diffusion_eqn_2d_src_comp_time} \label{fig:diffusion_eqn_2d_src_perf}}
+\caption{(a) $L^\infty$ error as a function of number of grid points in
 each coordinate direction (left) and computation time as a function of 
@@
 resolution of the figures.  
-These results were obtained using MATLAB implementations of the 
-finite difference schemes run on a 2.4 GHz MacBook Pro.
+(b) Computation time as a function of $L^\infty$ error.  These results were
+obtained using MATLAB implementations of the finite difference schemes run on
+a dual-core 2.4 GHz MacBook Pro.
 }
-\label{fig:diffusion_eqn_2d_src_analysis}
 \end{center}
@@
 The improvement in the accuracy of the forward Euler scheme with OTS-NIDC
-is shown in Figure~\ref{fig:diffusion_eqn_2d_src_analysis}.
-Figure~\ref{fig:diffusion_eqn_2d_src_analysis} also shows that the 
+is shown in Figure~\ref{fig:diffusion_eqn_2d_src_error}.
+Figure~\ref{fig:diffusion_eqn_2d_src_perf} shows that the 
 scaling of the computational time with the error is in good agreement with the
@@
 could have been used to obtain solutions with similar or even higher accuracy.
-For example, a fourth-order accurate solution to the diffusion equation
-(in any number of space dimensions) could be obtained by using a 
-pseudospectral spatial 
-discretization~\cite{trefethen_spectral_book,boyd_spectral_book} 
-of the Laplacian and a fourth-order time integration scheme 
-(\eg Runge-Kutta).  When the domain is irregularly shaped, high-order methods 
-are harder to design and implement.  
+When the domain is irregularly shaped, high-order methods are harder to design
+and implement.  
 In this section, we extend the OTS-NIDC finite difference scheme for the 2D 
@@
 researchers have developed fourth-order accurate methods for the diffusion 
-equation on irregular domains~\cite{gibou_2005,ito_2005}.  However, these 
-earlier methods require a wider ghost cell region~\cite{gibou_2005}, use
-implicit time integration~\cite{gibou_2005,ito_2005}, or involve a more 
-complex calculation for the boundary conditions~\cite{ito_2005}.  As we shall 
+equation on irregular domains \cite{gibou_2005,ito_2005}.  However, these 
+earlier methods require a wider ghost cell region \cite{gibou_2005}, use
+implicit time integration \cite{gibou_2005,ito_2005}, or involve a more 
+complex calculation for the boundary conditions \cite{ito_2005}.  As we shall 
 see, it is possible to achieve fourth-order accuracy for the diffusion 
@@
 boundary, we adopt the `ghost cell' approach for imposing boundary 
-conditions~\cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.
+conditions \cite{gibou_2005,ito_2005,fedkiw_1999,osher_fedkiw_book}.
 The main challenge in using the ghost cell method is ensuring ghost cells 
@@
 nearest interior neighbor (see Figure~\ref{fig:ghost_cells})\footnote{If
-a level set representation of the boundary~\cite{osher_fedkiw_book} is used, 
+a level set representation of the boundary \cite{osher_fedkiw_book} is used, 
 the location of the point $u_B$ can be calculated by inverting the linear 
@@
 when the distance between the boundary point and the nearest interior grid 
-point is less than $\dx^2$~\cite{gibou_2005}.  For the numerical scheme 
+point is less than $\dx^2$ \cite{gibou_2005}.  For the numerical scheme 
 presented in this article, this threshold was too low.  Better stability was 
@@
 \scalebox{0.35}{\includegraphics{figures/diffusion_eqn_2d_starfish_domain_soln}}
+\ \ \ \ \ \ \ 
 \scalebox{0.35}{\includegraphics{figures/diffusion_eqn_2d_starfish_domain_error}} 
@@
 error in the solution is concentrated near the boundaries of the irregular
-domain.  The improved accuracy of the OTS-NIDC solution compared to a simple 
-forward Euler scheme with a suboptimal time step is illustrated in 
-Figure~\ref{fig:diffusion_eqn_2d_starfish_error}.
+domain.  The improved accuracy and performance of the OTS-NIDC solution
+compared to a simple forward Euler scheme with a suboptimal time step are
+illustrated in Figure~\ref{fig:diffusion_eqn_2d_starfish_error} and
+Figure~\ref{fig:diffusion_eqn_2d_starfish_perf}.
 
@@
 \begin{center}
-\scalebox{0.35}{
-  \includegraphics{figures/diffusion_eqn_2d_starfish_domain_error_vs_N}} 
-\caption{$L^\infty$ error as a function of number of grid points in each
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/diffusion_eqn_2d_starfish_domain_error_vs_N} \label{fig:diffusion_eqn_2d_starfish_error}}
+\ \ \ \ \ \ \ 
+\subfloat[]{\includegraphics[width=0.40\textwidth]{figures/diffusion_eqn_2d_starfish_domain_comp_time} \label{fig:diffusion_eqn_2d_starfish_perf}}
+\caption{(a) $L^\infty$ error as a function of number of grid points in each
 coordinate direction for two finite difference schemes that solve the 2D 
@@
 (circles) and forward Euler with suboptimal time step (squares).  
+(b) Computation time as a function of $L^\infty$ error.  These results were
+obtained using MATLAB implementations of the finite difference schemes run on
+a dual-core 2.4 GHz MacBook Pro.
 }
-\label{fig:diffusion_eqn_2d_starfish_error}
 \end{center}
@@
 discretization error.  For 3D problems, a catalog of several such stencils is 
-available in~\cite{patra_2005}.   For higher dimensional problems, one simple
+available in \cite{patra_2005}.   For higher dimensional problems, one simple
 approach for deriving an isotropic finite difference stencil for the Laplacian
@@
 approximations with more general or complex properties, symbolic algebra 
-software can be helpful~\cite{patra_2005,gupta_1998}.
+software can be helpful \cite{patra_2005,gupta_1998}.
 
@@
 
-\section{\label{sec:summary} Summary} 
+\section{\label{sec:conclusions} Conclusions} 
 In this article, we have presented OTS-NIDC, a simple approach for
@@
 \emph{without} requiring the use of high-order stencils or high-order time
-integration schemes.  The major idea underlying OTS-NIDC is that 
-spatial and temporal discretization errors can be simultaneously eliminated
-by (1) using the PDE to relate terms in the discretization error, (2)
-optimally selecting numerical parameters, and (3) adding a few defect 
-correction terms in a non-iterative manner.  The boost in the order of 
-accuracy achieved by using OTS-NIDC is well worth the extra effort required to 
-derive the optimal time step and defect correction terms.
-
-Through many examples, we have demonstrated the utility of OTS-NIDC
-to several types of finite difference schemes for a wide range of problems.  
-We showed how OTS-NIDC can be used to very easily obtain high-order accurate 
-solutions to many linear and semilinear PDEs in any number of space dimensions 
-on both regular and irregular domains.  The examples illustrate the 
-high-order of accuracy that can be achieved using only low-order 
-discretizations for spatial derivatives and simple forward Euler time 
-integration.  They also demonstrate the applicability of OTS-NIDC to more 
-exotic finite difference schemes, such as the Kreiss-Petersson-Ystr\"om 
-discretization of the second-order wave equation and the DuFort-Frankel scheme 
-for the diffusion equation.
-
-Optimal time step selection is an example of the more general technique
-of optimally selecting numerical parameters to boost the accuracy of 
-numerical methods.  Little research appears to have been done in this area,
-but as OTS-NIDC selection demonstrates, optimal selection of numerical 
-parameters can significantly impact the accuracy of numerical methods.  To 
-realize the full potential of existing and novel numerical schemes, 
-both optimal time step selection and optimal numerical parameter selection 
-deserve further investigation and attention.
+integration schemes.  The essential idea underlying OTS selection is that 
+leading order spatial and temporal discretization errors with the highest
+order derivatives can be simultaneously eliminated by optimally selecting
+numerical discretization parameters.  Any terms in the truncation error that
+remain are handled in a straightforward manner using non-iterative defect
+correction.  The boost in the order of accuracy achieved by using OTS-NIDC is
+well worth the extra effort required to derive the optimal time step and
+defect correction terms.  In fact, the amount of analysis required to compute
+the OTS and correction terms is comparable to traditional defect/deferred
+correction and modified equation methods.
+
+Through many examples, we have demonstrated the utility of OTS-NIDC to several
+types of finite difference schemes for a wide range of problems.  We showed
+how OTS-NIDC can be used to very easily obtain high-order accurate solutions to
+many linear and semilinear PDEs in any number of space dimensions on both
+regular and irregular domains.  The examples illustrate the high-order of
+accuracy that can be achieved using only low-order discretizations for spatial
+derivatives and simple forward Euler time integration.  They also demonstrate
+the applicability of OTS-NIDC to more exotic finite difference schemes, such as
+the Kreiss-Petersson-Ystr\"om discretization of the second-order wave equation
+and the DuFort-Frankel scheme for the diffusion equation.
+
+The work presented in this article suggest several potential areas for future
+research.  First, because explicit time integration schemes tend to be easier
+to implement and appear to be more suited to modern graphics processing unit
+(GPU) hardware architectures,
+it would be interesting to explore application of OTS-NIDC to GPU
+implementations of numerical algorithms for time dependent PDEs.  Second,
+while it is straightforward to derive of the optimal time step and defect
+correction terms by using Taylor series expansions, the process can be tedious
+and error-prone.  It would be valuable to use modern symbolic algebra software
+to develop tools to help automate the process.  Finally, the accuracy results
+for the fourth-order parabolic equation suggest that it may be prudent to 
+empirically revisit the question of accuracy of numerical algorithms based
+on implicit time integration for problems involving high-order spatial
+derivatives.
 
@@
 The would like to thank P. Fok and J. Lambers for many enlightening 
-discussions and P. Fok, M. Prodanovi\'c and A. Chiu for helpful suggestions 
-on the manuscript.
-
-\appendix
-\section{Formal vs. Practical Accuracy
-         \label{app:formal_vs_practical_accuracy} }
-For time dependent PDEs, evaluating the order of accuracy for a numerical
-method is subtle because there are formally two separate orders of accuracy 
-to consider:  temporal and spatial.  While it is theoretically
-interesting and important to understand how the error depends on both the 
-grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
-of a numerical scheme is always controlled by only one of the two numerical
-parameters.
-
-Even though spatial and temporal orders of accuracy for a numerical method
-are formally separate, one will always dominate for a given choice of 
-$\dx$ and $\dt$.  For example, when solving the diffusion equation 
-using the backward Euler method for time integration with a second-order 
-central difference stencil for the Laplacian, formal analysis shows that 
-the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
-stability constraints on the numerical parameters, the time
-step and grid spacing are free to vary independently.  In this situation, the 
-practical error for the method depends on the relative sizes of the time step 
-and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
-$O(\dt)$ which means that the error in the numerical solution is 
-primarily controlled by the time step.  Similarly, when 
-$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
-the error is controlled by the grid spacing.  Finally, when 
-$\dt  = O(\dx^2)$, the practical error is 
-$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
-primarily controlled by one of the two numerical parameters, and varying
-the subdominant parameter while holding the dominant parameter fixed does 
-not significantly affect the error.
-
-When there are constraints on the numerical parameters, there is less freedom 
-in choosing the controlling parameter.  For instance, if we solve the 
-diffusion equation using a forward Euler time integration scheme with a 
-second-order central difference stencil for the Laplacian, stability
-considerations require that we choose $\dt = O(\dx^2)$.  
-Combining this stability constraint with the formal error for the scheme
-shows that the practical error is 
-$O(\dx^2) + O(\dt) = O(\dx^2)$.  Therefore, the accuracy of the 
-method is completely controlled by the grid spacing; the temporal error can 
-never dominate the spatial error.
-
-\section{Importance of High-Accuracy for First Time Step for the KPY Scheme
-         \label{appendix:KPY_analysis} }
-The need for higher-order accuracy when taking the first time step of the KPY 
-scheme for the second-order wave equation can be understood by solving the 
-difference equation 
-\bea
-  \he^{n+1} - 2 \he^{n} + \he^{n-1} = \dt^2 \lambda \he^{n}
-  \label{eq:error_eqn_normal_mode},
-\eea
-for the normal modes of the error, where $\he$ is the coefficient of an 
-arbitrary normal mode of the spatial operator for the error 
-$e \equiv u - \tu$.  Using standard methods for solving linear difference
-equations~\cite{carrier_pearson_book} yields the solution
-\bea
-  \he^n = \he^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-}
-        + \he^0 \kappa_+ \kappa_-
-          \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-},
-  \label{eq:error_eqn_normal_mode_soln}
-\eea
-where $\kappa_\pm$ are the roots of the characteristic equation for
-(\ref{eq:error_eqn_normal_mode})~\cite{kreiss2002}
-\beq
-  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2
-             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
-\eeq
-Notice that the denominator of both terms in
-(\ref{eq:error_eqn_normal_mode_soln}) is $O(\dt)$, which implies that the
-error in the initial conditions is degraded by one temporal order of accuracy
-by the KPY scheme.  Therefore, KPY without OTS-NIDC requires the first time 
-step must be at least third-order accurate; KPY with OTS-NIDC requires the 
-first time step must be at least fifth-order accurate.
+discussions and P. Fok, M. Prodanovi\'c, A. Chiu and the manuscript's
+anonymous reviewers for helpful suggestions on the manuscript.
 
@@
 
-\bibitem{zhao_2006}
-F.~Zhao.
-\newblock Weighting {F}actors for {S}ingle-{S}tep {T}rapezoidal {M}ethod.
-\newblock {\em J. Heat Transf.-Trans. ASME}, 128:409--412, 2006.
-
 \bibitem{pereyra_1968}
@@
 
+\bibitem{cash_1996}
+J.~R. Cash and Y.~Psihoyios.
+\newblock {T}he {MOL} solution of time dependent partial differential
+  equations.
+\newblock {\em Comput. Math. Appl.}, 31:69--78, 1996.
+
+\bibitem{psihoyios_2007}
+G.~Psihoyios.
+\newblock Solving time dependent pdes via an improved modified extended bdf
+  scheme.
+\newblock {\em Appl. Math. Comput.}, 184:104--115, 2007.
+
 \bibitem{leveque_book_1992}
@@
 
-\bibitem{trefethen_spectral_book}
-L.~N. Trefethen.
-\newblock {\em Spectral Methods in MATLAB}.
-\newblock SIAM, Philadelphia, PA, 2000.
-
-\bibitem{boyd_spectral_book}
-J.~P. Boyd.
-\newblock {\em Chebyshev and Fourier Spectral Methods}.
-\newblock Dover Publications, Inc., Mineola, NY, 2nd edition, 2001.
-
 \bibitem{gupta_1998}
@@
 
-\bibitem{carrier_pearson_book}
-G.~F. Carrier and C.~E. Pearson.
-\newblock {\em Ordinary Differential Equations}.
-\newblock Society for Industrial and Applied Mathematics, Philadelphia, 1991.
-
 \end{thebibliography}

### 2011-07-27 papers/OTSPDE/OTSPDE.tex
Additional revisions to OTSPDE manuscript.

@@
 boosting the order of accuracy of finite difference schemes for time dependent
-partial differential equations by optimally selecting the time step used to
-advance the numerical solution and adding defect correction terms in a
+partial differential equations (PDEs) by optimally selecting the time step
+used to advance the numerical solution and adding defect correction terms in a
 non-iterative manner.
-The power of the technique is its ability to increase the accuracy of formally
+The power of the technique, which is applicable to time dependent, semilinear,
+scalar PDEs where the leading-order spatial derivative has a constant
+coefficient, is its ability to increase the accuracy of formally
 low-order finite difference schemes without major modification to the basic 
@@
 dependent PDEs by optimally choosing the time step and, when necessary, adding
-defect correction terms in a \emph{non-iterative} manner.  This technique is a
-systematic generalization of the ideas underlying the previously mentioned
-examples and specific examples that are scattered throughout the literature.
+defect correction terms in a \emph{non-iterative} manner.  This technique,
+which is applicable when the leading-order spatial derivative in a semilinear,
+scalar PDE has a constant coefficient, is a systematic generalization of the
+ideas underlying the previously mentioned examples and specific examples that
+are scattered throughout the literature.
 Optimal time step (OTS) selection is based on the observation that a carefully
 chosen time step can \emph{simultaneously} eliminate low-order terms in both
-the spatial and temporal the discretization errors.  At its essence, OTS
+the spatial and temporal discretization errors.  At its essence, OTS
 selection combines the powerful idea of optimally choosing numerical parameters
@@
 unstable when the dimension of space is greater than 3.  To make OTS-NIDC 
-useful for diffusion problems in 4 and more space dimensions, we can switch 
+useful for diffusion problems in four and more space dimensions, we can switch 
 to the DuFort-Frankel scheme, which is explicit and unconditionally 
@@
 scheme should do better asymptotically.  The origin of this discrepancy is the
-constant hidden by the big-$O$ notation.  In particular, the factor of $7/120$
+constants hidden by the big-$O$ notation.  In particular, the factor of $7/120$
 in the optimal time step for the forward Euler scheme with a fourth-order 
 approximation of the bilaplacian leads to a large hidden constant 
-for the compute time. 
-However, as Figures~\ref{fig:4th_order_parabolic_eqn_1d_perf} and
-\ref{fig:4th_order_parabolic_eqn_1d_error_roundoff} show, there is an important 
-limitation of the 2nd-order Crank-Nicholson scheme even if we ignore the 
-cross-over which will occur at sufficiently small values of the error -- 
-round-off error associated with matrix inversion limits the accuracy that is 
-achievable by the scheme.  This limitation also exists for the 4th-order 
-Crank-Nicholson scheme.
+for the compute time for the OTS-NIDC scheme, and the low cost of inverting
+the banded-diagonal biaplacian matrix in 1D leads to a relatively small hidden
+constant in the Crank-Nicholson schemes.
+
+As Figures~\ref{fig:4th_order_parabolic_eqn_1d_perf} and
+\ref{fig:4th_order_parabolic_eqn_1d_error_roundoff} show, there is an
+important advantage to explicit time integration schemes for the fourth-order
+parabolic equation -- round-off error associated with matrix inversion does 
+\emph{not} limit the achievable accuracy of the numerical solution.  In 
+constrast, errors introduced in matrix inversion affects both the 2nd- and
+4th-order Crank-Nicholson schemes.
 

### 2011-07-27 papers/OTSPDE/OTSPDE.tex
Ran OTSPDE manuscript through spell checker.

@@
 for the compute time for the OTS-NIDC scheme, and the low cost of inverting
-the banded-diagonal biaplacian matrix in 1D leads to a relatively small hidden
+the banded-diagonal bilaplacian matrix in 1D leads to a relatively small hidden
 constant in the Crank-Nicholson schemes.
@@
 \emph{not} limit the achievable accuracy of the numerical solution.  In 
-constrast, errors introduced in matrix inversion affects both the 2nd- and
+contrast, errors introduced in matrix inversion affects both the 2nd- and
 4th-order Crank-Nicholson schemes.
@@
 \newblock Error estimates for deferred correction methods in time.
-\newblock {\em Apll. Numer. Math}, 57:335--353, 2006.
+\newblock {\em Appl. Numer. Math}, 57:335--353, 2006.
 

### 2011-07-27 papers/OTSPDE/submissions/201107-Applied_Mathematics_and_Computation-REVISION/cover_letter.tex
Added OTSPDE revision submission cover letter, responses to reviewer, etc.

\documentclass[12pt]{letter}

% page settings
\textwidth 6.5in
\textheight 9in
\oddsidemargin -0.05in
\evensidemargin -0.05in
\topmargin -0.85in

\begin{document}

\address{}
\signature{Kevin T. Chu}

\begin{letter}
{
% NORMALLY ADDRESS GOES HERE
}

\opening{Dear Applied Mathematics and Computations Editor/Reviewer,}

I have revised my manuscript titled \emph{Boosting the Accuracy of Finite
Difference Schemes via Optimal Time Step Selection and Non-Iterative Defect
Correction} in response to the suggestions and comments of the anonymous
reviewers.  The manuscript discusses a simple, systematic technique for
boosting the order of accuracy of finite difference schemes for time dependent,
semilinear, scalar PDEs with constant coefficient leading-order spatial
derivative term by choosing an optimal, \emph{fixed} time step and, when 
necessary, adding defect correction terms in a \emph{non-iterative} manner. 
As discussed in the paper, a careful choice of time step and correction terms
can reduce the order of the numerical error of simple finite difference
schemes and make it possible to achieve high-order accuracy using only
formally low-order finite difference stencils and time integration schemes.

This paper describes the method for calculating the optimal time step and
defect correction terms and explains the observed boost in accuracy using 
straightforward numerical analysis arguments.  The technique of optimal time 
step selection with non-iterative defect correction (OTS-NIDC) is applicable
to several types of finite difference schemes for linear and semilinear PDEs
in any number of space dimensions and both regular and irregular domains.
Examples of each of these types of problems are discussed and analyzed in the
paper.

Thank you for your review of the paper.  I look forward to hearing your 
comments and suggestions for improvement of the manuscript.

\closing{Sincerely,}

\end{letter}

\end{document}

### 2017-07-22 papers/2011-OTSPDE/submissions/201107-Applied_Mathematics_and_Computation-REVISION/responses_to_reviewers.tex
Rename papers/OTSPDE to papers/2011-OTSPDE

\documentclass[12pt]{article}

% page settings
\textwidth 6.5in
\textheight 9in
\oddsidemargin -0.05in
\evensidemargin -0.05in
\topmargin -0.85in

\begin{document}

\section*{Responses to Reviewers}

\subsection*{List of Changes to Manuscript}

\begin{enumerate}

\item Throughout manuscript (including the title), clarified and emphasized the
      general need for OTS and non-iterative defect correction in order to
      achieve the boost in the order of accuracy.  Changed OTS-DC to OTS-NIDC
      throughout manuscript.

\item Streamlined discussion throughout the manuscript to "shorten the verbiage
      without altering the message". Reorganized figures a little bit to reduce
      length of manuscript.

\item Moved comparisons with other methods out of the introduction section into
      its own section titled "Theoretical Comparisons with Related Numerical
      Techniques"

\item Modified the discussion in Section 2.2 (previously Section 1.2) to
      emphasize that OTS-NIDC is an important special case of the modified
      equation method where regularization of the modified equation is
      automatic via parameter choice rather than via iterative procedures.

\item Removed claims that optimal selection of numerical parameters is not 
      common.  When I originally wrote this manuscript (2007), I did not find
      many examples of optimal numerical parameter selection.  Apparently the
      situation has changed.

\item Removed discussion about applicability to GPU computing except for a
      brief comment that investigation of benefits of OTS-NIDC to GPU computing
      is an area for future research.

\item Throughout the manuscript, added clarification that theoretical costs are
      asymptotic.

\item Added performance data for 1D Burgers equation and a brief discussion of
      its interpretation.

\item Added performance data for 2D diffusion equation on an irregular domain.

\item Added emphasis on the accuracy limitations of the Crank-Nicholson schemes
      for the fourth-order parabolic equation.

\item Renamed 'Summary' section to 'Conclusions' and revised second half of
      section.

\item Minor grammatical revisions throughout manuscript.

\end{enumerate}

\subsection*{Reviewer \#2}

\subsubsection*{General rebuttal}

On the point that OTS-NIDC is "technically a special case of the method of
Modified Equation," I offer the perspective that it is an important special
case because the singular perturbations in the modified equations are
regularized \emph{automatically} by the choice of optimal time step.  The
resulting modified equation is a \emph{regular} perturbation to the original
PDE, so there is no need for iterative refinement to handle the singular
perturbation.

\subsubsection*{Rebuttals to specific points}

\begin{enumerate}
\item Throughout the manuscript, acronyms have been defined before use.

\item Throughout the manuscript, the identified typos have been fixed.  The
      appendices have been removed to reduce the length of the manuscript.

\item The comparison with other numerical methods has been moved to its own 
      section titled "Theoretical Comparisons with Related Numerical
      Techniques".  All other sections and references have been renumbered.

\item The suggested references have been added.

\item The last section has been renamed "Conclusions" as suggested.
      Discussion of future directions for research has been added to the 
      conclusions section.

\item The entire manuscript has been rechecked again for missed mistakes and
      reorganized to improve clarity and reduce length of manuscript.

\end{enumerate}

\subsection*{Reviewer \#3}

\subsubsection*{General rebuttal}

I agree that the OTS-NIDC concept contains elements that are very old ideas. 
However, I have personally only seen the method used as a 'trick' in special
cases.  I have not seen a unified presentation that discusses its utility as
a general approach to boosting the order of accuracy of finite difference
schemes without requiring derivation and use of formally high-order
discretizations, etc.  This perspective could certainly be a result not having
seen \emph{all} of the extensive literature on numerical methods.  That said,
the approach I present does not appear to be employed as often as I believe it
could be.  Perhaps the ideas, while old, could use a refresher in the
collective consciousness of computational mathematicians and scientists.

I also agree that there is additional analytical and coding work
that needs to be done in order to use OTS-NIDC.  However, the extra work and
coding complication is not as onerous as it may seem at first glance.  For
example, in the process of implementing the original numerical scheme, it is
likely that several of the required differential operators have already been 
implemented (hopefully even coded as functions to make them easy to reuse).  
Also, almost any time one wants to improve the accuracy or computational
performance of a numerical scheme (e.g.  modified equation method, formally
higher-order discretizations, implicit time integration), additional work is
likely to be required.

On the issue of cost overhead, there is certainly a cost overhead with adding 
correction terms.  However, the boost in the order of accuracy results in an
overall savings in computational effort for a given desired level of accuracy.
Throughout the manuscript, I have attempted to emphasize this point by plotting
computational performance as a function of numerical error (instead of as a
function of the number of grid points).

On the comment that the results shown in Figure 4c (previously Figure 6)
indicate that the cost overhead of the defect correction terms make the method 
quite demanding, I have to disagree.  The actual reason for the poor
performance of OTS-NIDC scheme relative to the Crank-Nicholson schemes for the
fourth-order parabolic equation is \emph{not} the cost of the correction term. 
While the computational cost correction terms certainly widens the gap, 
the correction terms themselves ($f_xxxx$ and $f_t$) are not that complicated or
costly to compute.  Rather, it is the extreme smallness of the prefactor of
the optimal time step which is 7/120 combined with the low cost of inverting
the bilaplacian in 1D.  The small prefactor on the time step also afflicts the
non-OTS-NIDC forward Euler scheme which is constrained to have a time step 
smaller than $3 (\Delta x)^4 / 40 \kappa$.  This point is discussed in second
to last paragraph of Section 4.5 (previously Section 3.5).

When analyzing the performance of the various schemes for the fourth-order
parabolic equation, it is worth noting that it is important to take the
accuracy results into consideration.  As Figure 4b and 4c (previously Figures
5 and 6) illustrate, the Crank-Nicholson schemes suffer from a limitation in
the achievable accuracy as a result of round-off errors introduced by matrix
inversion.  The forward Euler schemes (including OTS-NIDC) do not suffer from
this problem.

On the point of applicability to systems of nonlinear PDEs, I am currently 
working on applying the method to systems of PDEs and the Navier-Stokes
equations.  I have completed work on the former and can confidently report that
with one additional simple idea, OTS-NIDC works for systems of PDEs.  My work
on the latter looks promising, but I have not fully completed the analysis or
the coding required to demonstrated the theoretical performance estimates.

\subsubsection*{Rebuttals to specific points}

\begin{enumerate}

\item I have added measured performance results for the 1D viscous Burgers
      equation and the 2D diffusion equation on an irregular domain.  The 
      measured performance results now include: 1D viscous Burgers equation,
      1D fourth-order parabolic equation, 2D diffusion equation on a regular 
      domain, and 2D diffusion equation on an irregular domain.  I did not 
      include performance results for the 1D second-order wave equation because
      that simulation ran so fast that the timing results contained too much 
      noise.

      I agree there are many alternative methods besides OTS-NIDC for boosting
      the order of accuracy of linear problems.  My manuscript does not claim
      to be better than those other methods.  It just presents an alternative
      method that has its own set of advantages and disadvantages (just like
      all other numerical methods and techniques).

      Regarding doubt about the performance results presented in Figure 6 
      (previous Figure 8).  The results are accurate.  While it is true that
      the computation time as a function of the number of grid points is 
      higher for OTS-NIDC schemes, the computation time as a function of the
      numerical erorr is lower for OTS-NIDC schemes.

      I have made the the MATLAB source code (and pre-generated data) for all of
      the plots in the manuscripts available on the web at

\begin{verbatim}
http://ktchu.serendipityresearch.org/download/research/applied_math/OTS-NIDC/
\end{verbatim}

      I will leave the source code in this location (or provide a redirect if it
      moved) until the review of the manuscript is complete.  I encourage the
      reviewers to look through and point out any errors that I may have
      accidentally overlooked.  I have made an effort to make the code readable
      but apologize in advance if the implementation is not completely
      transparent.

      For what it's worth, I am of the opinion of the opinion that source code
      for numerical mathematics and computational science results be part of the
      manuscript review process (and even publication).  Unfortunately, this is
      not currently a standard part of the review process.

\item I have added the requested result to the manuscript. The results show
      that for a desired accuracy level, OTS-NIDC does indeed outperform a
      conventional forward Euler schemea even with the additional cost of the
      defect correction terms.

\item Noted and reworded to acknowledge that there has been other work in the
      area of optimal numerical parameter selection in recent years.  If there
      are specific examples that Reviewer \#3 feels should be referenced, I
      would be happy to include them.

\item Fourth-order accuracy of the result in Figure 8 (previously Figure 11)
      results from the use of fourth-order accurate ghost cell values when
      imposing boundary conditions.  The details of how the ghost cell values
      are calculated is discussed in Section 5.3.1 (previously Section 4.3.1).

\end{enumerate}

\subsection*{Reviewer \#4}

\subsubsection*{Rebuttals to specific points}

\begin{enumerate}

\item I have reworked the entire manuscript to emphasize that both optimal time
      step selection \emph{and} non-iterative defect correction are required to 
      boost the order of accuracy.

      I have added comments in the abstract and the third paragraph of the
      introduction to clarify the class of PDEs that OTS-NIDC is applicable to.

      One point that I would like to emphasize is that, unlike traditional
      defect correction and modified equation methods, the defect correction
      method is non-iterative and easy to incorporate into the software
      implementation.  As far as I can tell, the non-iterative approach is
      somewhat novel or less common.

\item I have changed the comments applicability to GPU computing to be a 
      potential area of future research rather than presenting it as a
      conclusion drawn from existing work.

\item I have streamlined the entire manuscript to remove unnecessary detail,
      remove overly pedantic discussion and consolidate figures to reduce the
      length of the manuscript.  The new manuscript is about 4 pages shorter
      than the original manuscript but should still contain the same message
      as before.

\end{enumerate}

\end{document}

### 2017-07-22 papers/2011-OTSPDE/OTSPDE-supplemental.tex
Rename papers/OTSPDE to papers/2011-OTSPDE

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% First completed draft: 2011/09/01, Serendipity Research
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[fleqn,12pt,twoside]{article}
\usepackage[headings]{espcrc1}
\usepackage{amsmath}
\usepackage{graphicx}
\PassOptionsToPackage{ctagsplt,righttag}{amsmath}
\usepackage{multirow}

%******* MACROS *******

% MATH MACROS
\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}

\def\pt{\partial t}
\def\px{\partial x}
\def\py{\partial y}

% NUMERICS MACROS
\def\tu{\tilde{u}}
\def\he{\hat{e}}
\def\dt{\Delta t}
\def\dx{\Delta x}
\def\dy{\Delta y}
\def\dto{\dt_{opt}}

% MISC MACROS
\def\eg{\emph{e.g. }}
\def\ie{\emph{i.e. }}
\def\etal{\emph{et al.}}

%******************************SIZE DECLARATIONS********************************
\topmargin -0.7in

%*******************************DOCUMENT BEGIN**********************************

\begin{document}
\bibliographystyle{unsrt}

\title{Boosting the Accuracy of Finite Difference Schemes via Optimal Time
Step Selection and Non-Iterative Defect Correction \\
(Supplemental Material)}

\author{Kevin T. Chu\address{Serendipity Research, Mountain View, CA 94041, United States}
}

\maketitle

\noindent \rule{6.3in}{1pt}

\appendix
\section{Formal vs. Practical Accuracy
         \label{app:formal_vs_practical_accuracy} }
For time dependent PDEs, evaluating the order of accuracy for a numerical
method is subtle because there are formally two separate orders of accuracy 
to consider:  temporal and spatial.  While it is theoretically
interesting and important to understand how the error depends on both the 
grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
of a numerical scheme is always controlled by only one of the two numerical
parameters.

Even though spatial and temporal orders of accuracy for a numerical method
are formally separate, one will always dominate for a given choice of 
$\dx$ and $\dt$.  For example, when solving the diffusion equation 
using the backward Euler method for time integration with a second-order 
central difference stencil for the Laplacian, formal analysis shows that 
the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
stability constraints on the numerical parameters, the time
step and grid spacing are free to vary independently.  In this situation, the 
practical error for the method depends on the relative sizes of the time step 
and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
$O(\dt)$ which means that the error in the numerical solution is 
primarily controlled by the time step.  Similarly, when 
$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
the error is controlled by the grid spacing.  Finally, when 
$\dt  = O(\dx^2)$, the practical error is 
$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
primarily controlled by one of the two numerical parameters, and varying
the subdominant parameter while holding the dominant parameter fixed does 
not significantly affect the error.

When there are constraints on the numerical parameters, there is less freedom 
in choosing the controlling parameter.  For instance, if we solve the 
diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, stability
considerations require that we choose $\dt = O(\dx^2)$.  
Combining this stability constraint with the formal error for the scheme
shows that the practical error is 
$O(\dx^2) + O(\dt) = O(\dx^2)$.  Therefore, the accuracy of the 
method is completely controlled by the grid spacing; the temporal error can 
never dominate the spatial error.

\section{Importance of High-Accuracy for First Time Step for the KPY Scheme
         \label{appendix:KPY_analysis} }
The need for higher-order accuracy when taking the first time step of the KPY 
scheme for the second-order wave equation can be understood by solving the 
difference equation 
\bea
  \he^{n+1} - 2 \he^{n} + \he^{n-1} = \dt^2 \lambda \he^{n}
  \label{eq:error_eqn_normal_mode},
\eea
for the normal modes of the error, where $\he$ is the coefficient of an 
arbitrary normal mode of the spatial operator for the error 
$e \equiv u - \tu$.  Using standard methods for solving linear difference
equations~\cite{carrier_pearson_book} yields the solution
\bea
  \he^n = \he^1 \frac{\kappa_+^n - \kappa_-^n}{\kappa_+ - \kappa_-}
        + \he^0 \kappa_+ \kappa_-
          \frac{\kappa_+^{n-1} - \kappa_-^{n-1}}{\kappa_+ - \kappa_-},
  \label{eq:error_eqn_normal_mode_soln}
\eea
where $\kappa_\pm$ are the roots of the characteristic equation for
(\ref{eq:error_eqn_normal_mode})~\cite{kreiss2002}
\beq
  \kappa_\pm = 1 + \frac{1}{2} \lambda \dt^2
             \pm \dt \sqrt{\lambda + \frac{\lambda^2 \dt^2}{4}}.
\eeq
Notice that the denominator of both terms in
(\ref{eq:error_eqn_normal_mode_soln}) is $O(\dt)$, which implies that the
error in the initial conditions is degraded by one temporal order of accuracy
by the KPY scheme.  Therefore, KPY without OTS-DC requires the first time 
step must be at least third-order accurate; KPY with OTS-DC requires the 
first time step must be at least fifth-order accurate.

%*********************************BIBLIOGRAPHY**********************************

\begin{thebibliography}{10}

\bibitem{carrier_pearson_book}
G.~F. Carrier and C.~E. Pearson.
\newblock {\em Ordinary Differential Equations}.
\newblock Society for Industrial and Applied Mathematics, Philadelphia, 1991.

\bibitem{kreiss2002}
H.-O. Kreiss, N.~A. Petersson, and J.~Ystr\"om.
\newblock {D}ifference {A}pproximations for the {S}econd {O}rder {W}ave
  {E}quation.
\newblock {\em SIAM. J. Numer. Anal}, 40:1940--1967, 2002.

\end{thebibliography}
\end{document}

### 2017-07-22 papers/2011-OTSPDE/OTSPDE.tex
Rename papers/OTSPDE to papers/2011-OTSPDE

--- 2011-07-27 papers/OTSPDE/OTSPDE.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 % Manuscript started: 2007/09/01
@@
 % Resubmission draft: 2011/07/30, Serendipity Research
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Minor revisions to accepted manuscript: 2011/09/17, Serendipity Research
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 minimal additional work.  For instance, we will see how OTS-NIDC makes it
-possible to achieve 4th-order accuracy for the diffusion equation in any
+possible to achieve fourth-order accuracy for the diffusion equation in any
 number of space dimensions using only a second-order stencil for the Laplacian
@@
     & \multirow{2}{*}{$O\left( e^{-5/2} \right)$} \\
-  \ \ 2nd-Order Bilaplacian & & & & \\
+  \ \ second-order Bilaplacian & & & & \\
   Forward Euler with 
@@
     & \multirow{2}{*}{$O\left( e^{-{5/4}} \right)$} \\
-  \ \ 4th-Order Bilaplacian & & & & \\
+  \ \ fourth-order Bilaplacian & & & & \\
   Forward Euler with 
@@
     & \multirow{2}{*}{$O\left( e^{-5/6} \right)$} \\ 
-  \ \ 4th-Order Bilaplacian and OTS-NIDC & & & & \\
+  \ \ fourth-order Bilaplacian and OTS-NIDC & & & & \\
   Crank-Nicholson with 
@@
     & \multirow{2}{*}{$O\left( e^{-1} \right)$} \\
-  \ \ 2nd-Order Bilaplacian & & & & \\
+  \ \ second-order Bilaplacian & & & & \\
   Crank-Nicholson with 
@@
     & \multirow{2}{*}{$O\left( e^{-3/4} \right)$} \\
-  \ \ 4th-Order Bilaplacian & & & & \\
+  \ \ fourth-order Bilaplacian & & & & \\
   \hline
@@
 In Figure~\ref{fig:4th_order_parabolic_eqn_1d_perf}, observe that the 
-2nd-order Crank-Nicholson scheme actually outperforms the OTS-NIDC scheme over 
-the high (but still practically useful) range of errors even though 
+second-order Crank-Nicholson scheme actually outperforms the OTS-NIDC scheme
+over the high (but still practically useful) range of errors even though 
 Table~\ref{tab:comp_perf_vs_err_4th_order_parabolic} predicts that the OTS-NIDC
@@
 \emph{not} limit the achievable accuracy of the numerical solution.  In 
-contrast, errors introduced in matrix inversion affects both the 2nd- and
-4th-order Crank-Nicholson schemes.
+contrast, errors introduced in matrix inversion affects both the second- and
+fourth-order Crank-Nicholson schemes.
 
@@
 and the Institute for High-Performance Computing (IHPC) in Singapore. 
-The would like to thank P. Fok and J. Lambers for many enlightening 
+He would like to thank P. Fok and J. Lambers for many enlightening 
 discussions and P. Fok, M. Prodanovi\'c, A. Chiu and the manuscript's

### 2017-07-22 papers/2011-OTSPDE/submissions/200806-SINUM/cover_letter.tex
Rename papers/OTSPDE to papers/2011-OTSPDE

--- 2008-11-18 papers/OTSPDE/submissions/200811-JCP/cover_letter.tex
@@
 
-\opening{Dear J. Computational Physics Editor/Referee,}
+\opening{Dear SIAM J. of Numerical Analysis Editor/Referee,}
 
-My article titled \emph{Boosting the Accuracy of Finite Difference Schemes 
-via Optimal Time Step Selection} discusses a novel technique for boosting 
+My article titled \emph{High-Order Accurate Finite Difference Schemes 
+Via Optimal Time Step Selection} discusses a novel technique for boosting 
 the order of accuracy of finite difference schemes for time dependent 

### 2017-07-22 papers/2009-OTSPDE-WCE/OTSPDE-WCE2009.tex
Rename papers/OTSPDE-WCE2009 to papers/2009-OTSPDE-WCE

--- 2009-03-20 papers/OTSPDE-WCE2009/OTSPDE-WCE2009.tex
@@
        Boost the Accuracy of FD Schemes for \\
-       Variable Coefficient PDEs }}
+       Variable-Coefficient PDEs }}
 
@@
 analysis of the FD scheme.  We illustrate the utility of our method by 
-applying it to variable coefficient wave and diffusion equations.  In 
+applying it to variable-coefficient wave and diffusion equations.  In 
 addition, we demonstrate the viability of OTS selection for the two-step 
@@
 \em Keywords: optimal time step selection, optimal grid selection, finite 
-    difference schemes, time dependent PDEs, variable coefficient PDEs}}
+    difference schemes, time dependent PDEs, variable-coefficient PDEs}}
  }
@@
 We then demonstrate the use of variable-coefficient OTS selection on the 
-variable-coefficient wave and heat equations.  Our analysis of the wave is 
+variable-coefficient wave and diffusion equations.  Our analysis of the wave is 
 equation is particularly noteworthy because it is \emph{not} based on a one-step 
@@
 
-\section{OTS Selection for Variable Coefficient PDEs
+\section{OTS Selection for Variable-Coefficient PDEs
   \label{sec:ots_var_coef_pdes}}
 Transformation of the spatial domain is a natural way to extend OTS selection 
-to semilinear PDEs that have variable coefficient leading-order spatial 
+to semilinear PDEs that have variable-coefficient leading-order spatial 
 derivatives.  Consider a semilinear PDE of the form:
@@
 
-\subsection{Variable Coefficient Wave Equation}
+\subsection{Variable-Coefficient Wave Equation}
 In this section, we apply OTS and optimal grid selection to the
-variable coefficient wave equation.  Because we choose a FD scheme based on 
+variable-coefficient wave equation.  Because we choose a FD scheme based on 
 the Kreiss-Petersson-Ystr\"om discretization scheme for the wave 
@@
 
-\subsubsection*{OTS Selection for Variable Coefficient Wave Equation}
-We now consider the variable coefficient second-order wave equation
+\subsubsection*{OTS Selection for Variable-Coefficient Wave Equation}
+We now consider the variable-coefficient second-order wave equation
 \beq
@@
 Using the optimal time step and the correction term, we obtain a fourth-order
-accurate KPY scheme for the variable coefficient wave equation 
+accurate KPY scheme for the variable-coefficient wave equation 
 (see Figure~\ref{fig:var_coef_wave_eqn_1d_error}).  As mentioned earlier,
@@
 \caption{$L^\infty$ error as a function of number of grid points for various
-KPY discretizations of the variable coefficient second-order wave equation:
+KPY discretizations of the variable-coefficient second-order wave equation:
 KPY using a uniform grid on the original domain (red squares),
@@
 grid, optimal time step, and the correction terms, we are able to obtain a 
-fourth-order accurate scheme for the variable coefficient wave equation on
+fourth-order accurate scheme for the variable-coefficient wave equation on
 the original domain.  The error on the variable-spaced grid is almost 
@@
 
-\subsection{Application to Variable Coefficient Diffusion Equation}
+\subsection{Application to Variable-Coefficient Diffusion Equation}
 In this section, we apply OTS and optimal grid selection to the 
@@
 Since the optimal time step for the constant coefficient diffusion equation 
-is given by $\dto = \dx^2/6D$, the optimal time step for the variable
-coefficient diffusion equation is given by $\dto = \left(\dy/\dbar\right)^2/6$.
+is given by $\dto = \dx^2/6D$, the optimal time step for the 
+variable-coefficient diffusion equation is given by 
+$\dto = \left(\dy/\dbar\right)^2/6$.
 The correction term for this scheme is 
@@
   \begin{array}{c}
-  6 D D_{xx} u_{xx} + 4 D D_{xxx} u_x D D_{xxxx} u + D f_{xx} \\
+  6 D D_{xx} u_{xx} + 4 D D_{xxx} u_x + D D_{xxxx} u + D f_{xx} \\
   + \ 6 D_x D_{xx} u_x + 2 D_x D_{xxx} u + 2 D_x f_x \\
@@
 \scalebox{0.24}{\includegraphics{figures/var_coef_diffusion_eqn_1d_FE_soln}}
-\caption{Comparison of numerical solutions for the variable coefficient
+\caption{Comparison of numerical solutions for the variable-coefficient
 diffusion equation~(\ref{eq:var_coef_diffusion_eqn}) computed using 
@@
 computational grid.  In particular, we have presented a method to boost the 
-order of accuracy of FD schemes for PDEs where the leading-order spatial derivative has a 
-variable coefficient by using an optimal time step and an optimal grid.  We have 
-described the basic procedure for constructing the optimal grid, computing the optimal time step, 
-and deriving the required correction terms.  To demonstrate the power of these 
-techniques, we have applied them to the variable coefficient wave equation and 
-the Fokker-Planck form of the diffusion equation in an inhomogeneous medium.  
-Our analysis of the KPY scheme (used to solve the wave equation) also serves 
-as an example of OTS selection applied to multistep time integration schemes.
+order of accuracy of FD schemes for PDEs where the leading-order spatial 
+derivative has a variable coefficient by using an optimal time step and an 
+optimal grid.  We have described the basic procedure for constructing the 
+optimal grid, computing the optimal time step, and deriving the required 
+correction terms.  To demonstrate the power of these techniques, we have 
+applied them to the variable-coefficient wave equation and the Fokker-Planck 
+form of the diffusion equation in an inhomogeneous medium.  Our analysis of 
+the KPY scheme (used to solve the wave equation) also serves as an example of 
+OTS selection applied to multistep time integration schemes.
 
